### 2015-06-22 Manuscript/SW_early_life_telomeres_SOM_v9.Rmd
First upload of all data andscripts

---
output: word_document
---

###Supplementary information for Spurgin *et al.* "Telomeres reveal complex silver spoon effects in a wild population"

```{r load and clean, include = F}

#Set wd one level up
setwd('..')

#Load and clean data

source('Scripts/SW juvenile telomeres IPC load.r')
source('Scripts/SW juvenile telomeres IPC functions.r')
source('Scripts/SW juvenile telomeres IPC clean.r')

#Set some options
options(na.action='na.fail')
opts_knit$set(root.dir = getwd())
opts_chunk$set(echo=FALSE,
               warning=FALSE,
               message=FALSE,
               fig.path='../Figures/',
               fig.width= 4,
               fig.height= 8,
               fig.show='asis',
               dev='png')

```

```{r Figure S1,fig.width=4,fig.height=4}

temp <- dd

ddFigS1 <- ddply(dd,
                 .(Age),
                 summarize,
                 TLKBmean = mean(TLKB),
                 TLKBse = se(TLKB))

ddFigS1 <- ddFigS1[complete.cases(ddFigS1),]
ddFigS1 <- subset(ddFigS1,Age<14)

FigS1 <- ggplot(ddFigS1,
       aes(x = Age,
           y = TLKBmean)) +
  geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse)) +
  ylab('Telomere length (kb)') + 
  xlab('Age') + 
  theme_classic() +
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16))

FigS1

```

**Figure S1** Cross-sectional telomere length (mean $\pm$ s.e.) in relation to age in Seychelles warblers.

PAGEBREAK

```{r Figure S2,fig.width=4,fig.height=4}

temp <- FlSA

ddFigS2 <- ddply(temp,
                 .(LayYear),
                 summarize,
                 TLKBmean = mean(TLKB),
                 TLKBse = se(TLKB))

ddFigS2 <- ddFigS2[complete.cases(ddFigS2),]
ddFigS2$LayYear <- as.numeric(paste(ddFigS2$LayYear))

FigS2 <- ggplot(ddFigS2,
       aes(x = LayYear,
           y = TLKBmean)) +
  geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse)) +
  ylab('Telomere length (kb)') + 
  xlab('Lay Year') + 
  theme_classic() +
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16))

FigS2

```

**Figure S2** Cross-sectional juvenile telomere length (mean $\pm$ s.e.) across birth years in Seychelles warblers

PAGEBREAK

```{r Figure S3,fig.width=5}

temp <- subset(FlSA,LayYear %in% c(1999,2006,2007,2000))

FigS3A <- ggplot(temp,
      aes(x=GroupSize,
      y=TLKB,
      col=LayYear)) +
  stat_smooth(method=lm,fullrange=T,alpha=0.2, aes(fill=LayYear),lwd=1.2) + 
    ylab('Telomere length (kb)') + 
  xlab('Local Density') + 
  theme_classic() +
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16)) +
  scale_color_brewer(palette='Set1') + 
  scale_fill_brewer(palette='Set1') +
  annotate('text',x=2,y=10,label='A')

FigS3B <- ggplot(temp,
      aes(x=Condition,
      y=TLKB,
      col=LayYear)) +
  stat_smooth(method=lm,fullrange=T,alpha=0.2, aes(fill=LayYear),lwd=1.2) + 
    ylab('Telomere length (kb)') + 
  xlab('Body Condition') + 
  theme_classic() +
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16)) +
  scale_color_brewer(palette='Set1') + 
  scale_fill_brewer(palette='Set1') +
  annotate('text',x=-2,y=15,label='B')

multiplot(FigS3A,FigS3B)

```

**Figure S3** Within cohort relationships between telomere length and **A** local density and **B** body condition in juvenile Seychelles warblers. Only years with juvenile sample sizes >20 are included. Lines and shaded areas represent fitted values and 95% confidence limits from linear regressions.

```{r Figure S4,fig.width=5,fig.height=5}
glm1 <- glm(SurvivedNext ~ Sex + Condition + TLKB + Insect + Ageclass + TLKB + GroupSize,
            data = FlSAall,
            family = 'binomial')
glm1 <- standardize(glm1)
glm2 <- dredge(glm1)
glm2avg <- model.avg(glm2, subset = delta <=6)
myimp2 <- importance(glm2avg)

#Make into a nice table
table2 <- imptable(glm2avg,myimp2)
table2 <- signif(table2,2)

#Sort out row names
nams <- rownames(table2)
nams[nams == 'c.Ageclass'] <- 'Age class'
nams[nams == 'z.Condition'] <- 'Body condition'
nams[nams == 'c.Sex'] <- 'Sex'
nams[nams == 'z.Insect'] <- 'Food av.'
nams[nams == 'z.TQspace'] <- 'TQ'
nams[nams == 'z.TLKB'] <- 'TL'
nams[nams == 'z.GroupSize'] <- 'Local density'
nams[nams == 'z.Insect:z.TLKB'] <- 'TL x Food av.'

rownames(table2) <- nams
colnames(table2) <- c('Estimate','SE','2.5%','97.5%','RI')

table2 <- data.frame(table2,myx=row.names(table2))

forestplot(table2)

```

**Figure S4** Model averaged estimates and confidence intervals (points and error bars, respectively) for all explanatory terms used in a generalised linear model with a binomial error structure and survival to adulthood as the response. Numbers in brackets are the relative importance of each term in the top model set (see main text for details). Food av = annual food availability, TL = telomere length TQ = territory quality.

### 2015-06-22 Manuscript/SW_early_life_telomeres_v9.Rmd
First upload of all data andscripts

---
output:
  word_document:
    fig_caption: yes
    highlight: null
bibliography: mybib.bib
csl: ProceedingsB.csl
---
```{r load and clean, include = F}

#Set wd one level up
setwd('..')

#Load and clean data

source('Scripts/SW juvenile telomeres IPC load.r')
source('Scripts/SW juvenile telomeres IPC functions.r')
source('Scripts/SW juvenile telomeres IPC clean.r')

```

###Telomeres reveal complex silver spoon effects in a wild population

Lewis G. Spurgin^1^,  Kathryn L. Bebbington^1^, Eleanor A. Fairfield^1^, Jan Komdeur^2^, Terry Burke^3^, Hannah, L. Dugdale^3^,  and David S. Richardson^1,4^.

1. School of Biological Sciences, University of East Anglia, Norwich Research Park, NR4 7TJ, United Kingdom
2. Behavioural Ecology and Self-organization Group, Centre for Ecological and Evolutionary Studies, University of Groningen, Groningen, The Netherlands
3. Department of Animal and Plant Sciences, NERC Biomolecular Analysis Facility, University of Sheffield, Sheffield, UK
4. Nature Seychelles, Roche Caiman, Mahé, Republic of Seychelles

**Correspondence:** Lewis Spurgin: lewisspurgin@gmail.com; David Richardson: david.richardson@uea.ac.uk

**Keywords:** Ageing; Life-history; Seychelles warbler; Telomeres

PAGEBREAK

###Abstract
Favourable conditions early in life can confer delayed survival advantages. These 'silver spoon effects' are most likely common in nature and are central to our understanding of life history evolution and population ecology. However, how silver spoon effects vary among individuals, populations and species remains poorly understood. Telomeres can act as biomarkers of cost in wild populations, and may provide insight into how silver spoon effects occur. Using a dataset spanning over a decade, we studied how telomere length links spatiotemporal variation in early-life conditions to survival to in the Seychelles warbler (*Acrocephalus sechellensis*). We show that early-life telomere length varies markedly across birth years, and that this variation is positively related to annual variation in food availability. We then show that having short telomeres as a juvenile is associated with reduced late life survival, but that this relationship was subject to strong cohort effects. Having short telomeres was associated with reduced late-life survival among individuals born in years where food was limited, but not among individuals born in years with high food availability. Our results suggest that telomeres can reveal hidden costs and complex silver spoon effects in wild populations.

PAGEBREAK

###Introduction
Being subject to favourable environmental conditions during development and growth can confer fitness advantages later in life (so called 'silver spoon effects') [@Grafen1988; @Monaghan2008]. There is now evidence that adult fitness can be affected by a range of early-life experiences, including variation in resource availability [@Hayward2013], population density [@Nussey2007] and natural or anthropogenic environmental disturbance [@Reid2003]. Understanding the causes and consequences of these early-life experiences is key to understanding many ecological and evolutionary  processes, including  patterns of natural and sexual selection,  population growth rates and even local extinction [@Coulson2001; @Roach2014]. Understanding silver spoon effects is therefore of central interest to ecologists, evolutionary biologists and conservationists.

While it is clear that silver spoon effects can occur, we have little understanding of how and under what conditions early-life environments will affect adult fitness. Because the phenotypic consequences of an adverse environment can depend on an individual's genetic or epigenetic makeup [@Hoffman2000; @Richards2006], the later-life consequences of a good/poor start are expected to vary among individuals, populations and species. Moreover, it is not always possible to fully quantify what constitutes a good or bad environment, and any 'hidden' environmental variation may obscure relationships relationships between the early-life conditions that are measured and adult phenotypes. Indeed, the pervasiveness of silver spoon effects varies between species [@Drummond2011], cohorts [@Reid2003] and sexes [@Wilkin2009]. In order to better understand the later-life consequences of early-life experiences, we therefore need to understand, or at least be able to measure, how the environment differentially affects individuals within a population.

A key process that has been associated with adult survival is the rate at which telomere repeats are lost. Telomeres shorten with age [@Monaghan2006], and in response to oxidative damage that can occur due to environmental stress [@VonZglinicki2002]. When telomeres reach a critical shortness cells senesce  [@Campisi2003], resulting in organismal senescence and death  [@Wong2003]. These links between telomere length and senescence have inspired a great deal of recent research into telomere ecology [@Monaghan2014]. An emerging consensus from this research is that telomeres can act as biomarkers of cost in wild populations, retaining signatures of ecological stress that are otherwise difficult to detect [@Schultner2014; @Asghar2015].

There is evidence from a range of taxa that the greatest rate of telomere loss occurs in early life (e.g. [@Frenck1998; @Haussmann2003]), and that the extent of this early-life telomere shortening is influenced by the environment [@Boonekamp2014]. Moreover, a recent laboratory study showed that juvenile telomere dynamics are related to late-life survival [@Heidinger2012]. However, whether telomeres provide a link between early-life conditions and late-life survival in the wild is not known.

The longitudinal study (since 1986) of the Seychelles warbler (*Acrocephalus sechellensis*) population on Cousin Island provides an excellent opportunity to assess the relationship between early-life environmental variation, telomere length and late-life fitness. Firstly, due to the isolated nature of the study population and intensive field monitoring, we have excellent survival data and sampling material spanning many years (see methods, below). Secondly, telomere length predicts survival independently of age in adult Seychelles warblers, suggesting that telomeres act as a biomarker of cost in this species [@Barrett2013a]. Finally, ecological conditions on Cousin are variable over space and time due to spatiotemporal variation in foliage cover and food availability [@VandeCrommenacker2011a].

In Seychelles warblers, oxidative stress is linked to natal territory quality [@VandeCrommenacker2011a], but neither natal territory quality nor natal local density are linked to juvenile or adult survival [@Brouwer2006; @Hammers2013]. In this study we analyse patterns of early-life telomere length and subsequent survival to test the prediction that telomeres fill a missing link between early-life environmental variation and late-life fitness. Specifically we predict that: i) individuals raised in good quality conditions will have longer telomeres, and ii) the relationship between telomere length and early-life conditions influences short and long-term survival.

###Methods

#####Study species and sampling
The Seychelles warbler is a small (~15 g) passerine bird with a mean life expectancy of 5.5 years at fledging [@Hammers2013]. The population of *ca.* 320 adults on Cousin Island (04'20'S, 55'40'E) has been intensively studied since 1986 [@Richardson2003b; @Barrett2013a]. Individuals are usually ringed in their first year of life, and so are of known age. They are then followed throughout their lives, and as they are non-migratory endemics and are naturally confined to the island [@Komdeur2004a], a biannual census of birds on the island gives an accurate measure of levels of local density and individual survival rate [@Barrett2013a]. Full details of catching and census methods can be found in Brouwer *et al.* [@Brouwer2012].

All territories are mapped biannually based on detailed observational data of foraging and territorial defence behaviour, and surveyed for territory quality [@Richardson2003b]. Territory quality is calculated based on territory size, foliage cover and insect abundance [@Komdeur1992] and averaged within territories [@Hammers2013]. Cousin is also subject to inter-annual variation in rainfall and food availability, and such yearly effects may override the effects of absolute territory quality. As an estimate of annual variation in food availability, we calculated an index of the number of insects across the entire island during the main breeding season of each sampling year based on the insect abundance data (see above).

Each time a bird is caught on Cousin body mass and tarsus length are measured, and age is assessed on the basis of eye colour and previous captures [@Richardson2003b]. A blood sample (ca 25 $\mu$l) is taken from each bird via brachial venipuncture, and stored in 1 m1 absolute ethanol in a 1.5 ml screw-cap microfuge tube at room temperature.

#####Molecular methods
Genomic DNA was extracted from a ~2 mm^2^ flake of preserved blood using the DNeasy Blood and Tissue Kit (Qiagen), following the manufacturer's protocol, with the modification of overnight lysis at 37^o^C and a final DNA elution volume of 80 $\mu$l. DNA concentration and purity were quantified using a NanoDrop 8000 Spectrophotometer (ThermoScientific), and DNA integrity was validated by by visualization with ethidium bromide after electrophoresis on a 1.2% agarose gel. Sex was determined using the molecular method outlined by Griffiths *et al.* [-@Griffiths1998].

Telomere lengths were measured from `r nrow(FlSA)` samples taken cross-sectionally from birds aged between two and twelve months, between `r min(as.numeric(paste(juv$LayYear)))` and `r max(as.numeric(paste(juv$LayYear)))`. Of these birds, `r nrow(subset(FlSA,Ageclass=='FL'))` were aged between two and six months (juveniles), and `r nrow(subset(FlSA,Ageclass=='SA'))` were aged seven to twelve months (subadults). We chose to focus on juveniles and subadults rather than nestlings, because i) we have juvenile samples spanning a range of years, whereas nestling samples are predominantly from two early years, and ii) telomere length during the nestling phase will largely be a product of inheritance and growth, whereas in juveniles and subadults we expect telomere length to incorporate the impact of early-life experiences. We measured absolute telomere quantity per diploid genome using a quantitative PCR (qPCR) assay  (full details in [@Barrett2012]).

#####Statistical analyses
We performed all analyses using R version 3.0.1 [@RDevelopmentCoreTeam2011]. We used linear models (LMs) and generalised linear models (GLMs) with model averaging to assess the causes and consequences of early-life telomere length in the Seychelles warbler. For each analysis we created a full model with all terms and biologically relevant interactions, from which a top model set was defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011]. We report model averaged coefficients, confidence intervals, and 'relative importance', which reflects the relative weights of each predictor variable across the top model set. Model averaging was carried out using the MuMIn package (version 1.10.5) in R [@Barton2012].

We first tested whether early-life environmental variation influences telomere length using linear models, with telomere length as the response variable. As explanatory variables we included body condition (calculated as deviation from a regression line of mass on tarsus length; [@VandeCrommenacker2011a]), sex, age class (juvenile or subadult), territory quality, yearly food availability and local density (measured as the number of individuals in the juvenile's natal social group [@Brouwer2012]).

We then used generalised linear models to examine variation in telomere length in relation to survival. In these models we included the same predictor variables as the previous analysis, plus early-life telomere length. We first performed logistic regression to test whether telomere length predicts survival to adulthood (response = survived to the following year; yes/no). We then used Poisson regression, excluding individuals that died in their first year of life, to test whether telomere length predicts late-life survival (in years).

###Results

Mean ($\pm$ standard error) early-life telomere length was `r round2(mean(FlSAall$TLKB))` $\pm$ `r round2(se(FlSAall$TLKB))` kb. This is only marginally higher than mean telomere length for all adult birds (`r round2(mean(subset(dd,Age>0)$TLKB))` $\pm$ `r round2(se(subset(dd,Age>0)$TLKB))` kb), and lower than the mean for one year-old birds based on cross-sectional data (`r round2(mean(subset(dd,Age==1)$TLKB))` $\pm$ `r round2(se(subset(dd,Age==1)$TLKB))` kb; Figure S1). Using longitudinal data, however, we found that telomere length was shorter in young adult birds compared to when they were juveniles (mean loss between juvenile and first adult sample = `r round(mean(FlSA$TLloss,na.rm=T),1)*1000` bp). We found no significant difference in telomere length between age classes (juvenile vs subadult), or between sexes (Figure 1A).

Early-life telomere length differed substantially among birth years (Figure S2), with longer telomeres in years where insect abundance was high (Figure 1B). Model averaging confirmed this strong, positive effect of yearly food availability on telomere length (Figure 1A). The best supported model with early-life telomere length as the response variable contained annual food availability, body condition and local density as explanatory variables (R^2^ = `r round2(summary(lm(TLKB~YQ+GroupSize+Condition,data=FlSAall))$r.squared)`). Early-life telomere length  was positively related to body condition (Figure 1C), and negatively related to local density, although in the latter case model-averaged confidence intervals overlapped zero (Figure 1A).

To better visualise cohort effects we plotted the relationship between telomere length and both local density and body condition among cohort years (Figure S3). We found that the relationship between local density and telomere length varied among cohorts, from being negative to positive (Figure S3A). A similar, albeit less pronounced, pattern was observed for within-cohort relationships between telomere length and body condition (Figure S3B).

Annual food availability and local density were the most important factors predicting survival to adulthood in juvenile Seychelles warblers (Figure S4). The top model contained these two factors, but was very weak in terms of explanatory power (McFadden's pseudo R^2^ = `r round2(1-logLik(glm(SurvivedNext~TLKB*Insect+GroupSize,family='binomial',data=FlSAall))/logLik(glm(SurvivedNext~1,family='binomial',data=FlSAall)))`), and the confidence intervals for all explanatory terms overlapped zero (Figure S4), suggesting that none of the variables studied predict survival to adulthood.

Early-life telomere length and annual food availability, plus the interaction between these two variables, were the top predictors of late-life survival (Figure 2A). The interaction between telomere length and annual food availability was in the top model explaining late-life survival, and this model had good explanatory power (McFadden's pseudo R^2^ = `r round2(1-logLik(glm(RemainingLife~TLKB*Insect,family='poisson',data=subset(FlSAall,RemainingLife>0)))/logLik(glm(RemainingLife~1,family='poisson',data=FlSAall)))`). There was a positive relationship between telomere length and late-life survival in years of low insect abundance, with no such relationship when food availability was high (Figure 2B).

###Discussion
Previous studies have shown that in juvenile Seychelles warblers, neither territory quality nor annual variation in food availability predict survival to adulthood or late-life survival [@Brouwer2006; @Brouwer2012; @Hammers2013]. Here we show that telomeres link early-life conditions to later-life survival in the Seychelles warbler. These findings have a number of implications for research on life history evolution in wild populations.

Our finding that food availability is related to early-life telomere length adds to a growing body of correlative and experimental evidence showing that the natal environment can have pronounced impacts on early-life telomere length [@Monaghan2014].  However, very few studies have shown that temporal variation in environmental conditions affects telomere length [@Mizutani2013]. We suspect this is more due to a lack of available long-term datasets with telomere screening, rather than the Seychelles warbler being unique. Indeed, the environment on Cousin is benign in comparison with many regions outside the tropics, where populations undergo large fluctuations in size (e.g. [@Coulson2001]). We therefore expect that as more long-term studies generate telomere datasets, the effects of temporal changes in the environment on telomere dynamics will become increasingly clear.

We found no evidence that telomere length predicts survival to adulthood in Seychelles warblers. This is surprising, as we previously found that telomere length in adult Seychelles warblers predicts survival to the next year [@Barrett2013a]. Moreover, our cross-sectional data show that telomere length is longer, on average, in one year-old birds compared to juveniles (Figure S1), while there was no evidence of telomere lengthening from the longitudinal data. A pattern like this is expected when there is selective disappearance - a known phenomenon in studies of telomere dynamics whereby individuals with short telomeres die, resulting in an apparent increase in telomere length in cross-sectional data [@Haussmann2008a]. Thus, the raw data suggest that a survival effect may occur in early life, but this was not supported by our logistic regression analyses. This contradiction in results may be due to lack of power if a survival effect is present but weak and/or cohort specific. Increased sampling within years will help better elucidate the short-term survival impacts of early-life telomere length in this system.

In captive zebra finches, juvenile telomere length predicts late-life survival [@Heidinger2012]. Our results are the first to replicate this relationship in the wild. However, the effect varied among cohorts, suggesting that the relationship between early-life telomere length and late-life survival may not be a straightforward one in natural systems. Indeed, it appears that the interplay between the environment, telomeres and late-life survival at a given point in time is a snapshot of a fluid and dynamic evolutionary process. We therefore recommend that relationships involving telomeres from single cohorts should be interpreted with caution until we gain a better understanding of what drives spatiotemporal variation in telomere dynamics and survival.

Together, our results confirm that there is cohort-level variation in the relationship between the early-life environment and later-life survival, and suggest that telomeres go some to way to capturing this complexity. Thus, our results support the idea that telomeres can be used as a 'generic currency' with which we can measure variation in costs among individuals and populations [@Asghar2015]. The prospect of using telomeres as indicators of population health is a particularly interesting one, as such an approach would be useful to a broad range of conservation and animal health practitioners. Our findings suggest that telomere length at the population level does reflect environmental stress. However, our analyses are confined to a single island, and research across populations with different environments and genetic histories is required in order to provide further insight into the usefulness of telomeres as biomarkers of population health.

Despite the promise of telomeres as biomarkers of cost, ecologists interested in understanding variation within populations are still left with the problem of identifying what explains individual-level variation in telomere length. Telomere length in juveniles will be the product of a range of factors, including inherited telomere length at birth [@Horn2011], age at sampling [@Heidinger2012], the early-life environment [@Boonekamp2014], and, potentially, individual quality. In the Seychelles warbler we do not yet know how these different factors contribute to within-cohort variation in telomere length. However, a previous study on the Seychelles warbler showed that maternal heterozygosity influences offspring survival in bad years, but not good years [@Brouwer2007]. This raises the possibility that in the Seychelles warbler there is variation in individual quality, and high quality individuals are better able to cope with early-life stress [@Brouwer2007]. Thus in the near future it will become possible to use telomeres to link genetic variation, inheritance, environmental variation and senescence in the Seychelles warbler and other long-term systems.

###Ethics
All research activities were approved by the ethics committee at the University of East Anglia, the Seychelles Department of Environment and the Seychelles Bureau of Standards (approval reference A0347).

###Data accessibility
All data required to reproduce the analyses within this manuscript have been uploaded to Dryad (XXXX)

###Competing interests
There are no competing interests

###Authors' contributions
DSR developed the idea, obtained funding for the overall telomere project and managed the study; DSR, JK, HLD and TB supervised the overall running of the overall Seychelles warbler project and contributed resources; HLD managed the longitudinal fitness database. EAF and KLB generated and processed the telomere data; LGS performed the analyses and wrote the manuscript in discussion with DSR. All authors read and contributed to the manuscript.

###Acknowledgements
Nature Seychelles facilitate the long-term Seychelles warbler project, allowing us to stay on Cousin Island, and providing accommodation and facilities during field work. The Seychelles Bureau of Standards and Department of Environment gave permission for sampling and fieldwork. Emma Barrett laid the foundations for this study, generating much of the telomere  data. We thank everyone who has helped in the field, and the current Seychelles warbler research group for useful discussions.

###Funding
This work was funded by two Natural Environment Research Council (NERC) grants to DSR (NE/F02083X/1 and NE/K005502/1) on which JK, TB and HLD were project partners. HLD was funded by a NERC fellowship.

###References

### 2015-06-22 Manuscript/SW_early_life_telomeres_tables_and_figures v9.Rmd
First upload of all data and scripts

---
output: word_document
---
```{r setwd, include = F}

#Set working directory up one level
setwd('..')

#Load and clean data
source('Scripts/SW juvenile telomeres IPC load.r')
source('Scripts/SW juvenile telomeres IPC functions.r')
source('Scripts/SW juvenile telomeres IPC clean.r')

#Set some options
options(na.action='na.fail')
opts_knit$set(root.dir = getwd())
opts_chunk$set(echo=FALSE,
               warning=FALSE,
               message=FALSE,
               fig.path='../Figures/',
               fig.width= 5,
               fig.height= 5,
               fig.show='asis',
               dev='png')

```

**Figure Legends**

**Figure 1** Factors affecting telomere length in juvenile Seychelles warblers. **A** Model averaged estimates and 95% confidence intervals (points and error bars, respectively) for all explanatory terms used in a linear model with juvenile telomere length as the response variable. Numbers in brackets are the relative importance of each term in the top model set (see main text for details; Food av = annual food availability, TQ = territory quality). **B** Annual food availability and **C** body condition in relation to telomere length. Points and error bars in **A** are mean $\pm$ 95% confidence interval telomere lengths for each year or group size (means are used for plotting purposes only), and the lines represent fitted values and from linear regressions.

**Figure 2** Juvenile telomere length and lifespan in Seychelles warblers. **A** Model averaged estimates and 95% confidence intervals (points and error bars, respectively) for all explanatory terms used in a generalised linear model with a poisson error structure and lifespan as the response variable. Only juveniles that survived their first year of life were included in the model. Numbers in brackets are the relative importance of each term in the top model set (see main text for details). Food av = annual food availability, TL = telomere length TQ = territory quality. **B** Raw telomere length and annual food availability in relation to lifespan. Annual food availability was modeled as a covariate (see main text for details) but is plotted as a factor here for visualisation purposes. The lines represent fitted values from a Poisson regression.

PAGEBREAK

**Figure 1**

```{r Figure 1,fig.width=8,fig.height=9}

###########
#Fig 1A

#Run the model
lm1 <- lmer(TLKB~ TQspace + Insect + GroupSize + Sex + Condition + Helper + (1|LayYear),
            data = FlSAall,REML=FALSE)

lm1 <- standardize(lm1)

#Model averaging
lm2 <- dredge(lm1)
lmavg <- model.avg(lm2,subset = delta <=6)
myimp <- importance(lmavg) #Relative importance of each explanatory variable

#Put it all into a nice table
table1 <- imptable(lmavg,myimp)
table1 <- signif(table1,2)

#Sort out row/col names
nams <- rownames(table1)
nams[nams == 'c.Ageclass'] <- 'Age class'
nams[nams == 'z.Condition'] <- 'Body condition'
nams[nams == 'z.GroupSize'] <- 'Local density'
nams[nams == 'c.Sex'] <- 'Sex'
nams[nams == 'z.Insect'] <- 'Food av.'
nams[nams == 'z.TQspace'] <- 'TQ'
nams[nams == 'z.Helper'] <- 'Helpers'

rownames(table1) <- nams
colnames(table1) <- c('Estimate','SE','2.5%','97.5%','RI')

table1 <- data.frame(table1,myx=row.names(table1))

Fig1A <- forestplot(table1) +
  annotate('text',x=0.5,y=4,label='A')

######################################
temp <- FlSAall
temp$Insect <- factor(temp$Insect)

ddFig1B <- ddply(temp,
                 .(Insect),
                 summarize,
                 TLKBmean = mean(TLKB),
                 TLKBse = 1.96*se(TLKB))
ddFig1B <- ddFig1B[complete.cases(ddFig1B),]
ddFig1B$Insect <- as.numeric(paste(ddFig1B$Insect))

Fig1B <- ggplot(ddFig1B,
       aes(x = Insect,
           y = TLKBmean)) +
  ylab('Telomere length (kb)') + 
  xlab('Annual food availability') + 
  theme_classic() +
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16)) +
  stat_smooth(data=FlSAall,aes(x=Insect,y=TLKB),method=lm,col='black',se=F) +
  geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse)) +
  scale_colour_manual(values = wes_palette("Royal1")) +
  annotate('text',x=8,y=12,label='B')

Fig1C <- ggplot(FlSAall,
                aes(x = Condition,
                    y = TLKB)) +
    geom_point() +
  ylab('Telomere length (kb)') + 
  xlab('Body Condition') + 
  theme_classic() +
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16)) +
  stat_smooth(method=lm,col='black',se=F) +
  scale_colour_manual(values = wes_palette("Royal1")) +
  annotate('text',x=-2.5,y=15,label='C')

multiplot(Fig1A,Fig1B,Fig1C,layout=matrix(c(1,1,1,1,1,1,2,3,2,3), nrow=5, byrow=TRUE))

```

PAGEBREAK

**Figure 2**

```{r Figure 2}

x <- summary(lm(TLKB~Insect,data= FlSAall))$resid
xf <- ifelse(x>median(x),'Long','Short')

s1 <- survfit(Surv(time = Lifespan+1, event = Died)~xf, data = FlSAall)
Fig2 <- ggsurv(s1,plot.cens=F) + 
  theme_classic() +
  ylab('Proportion Survived') + 
  xlab('Age (years)') + 
  theme_classic() +
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16),
        legend.position = c(0.8,0.8),
        legend.title = element_text(size = 14),
        legend.text = element_text(size = 12)) +
  scale_colour_manual(name = 'Residual TL',
                      values = c('darkred','red')) +
  guides(linetype=F) +
  theme(legend.title = element_text('Telomere length'))

Fig2

```

### 2015-06-22 Talks/SW early life telomeres Glasgow.Rmd
First upload of all data and scripts

---
output:
  beamer_presentation:
    fig_caption: yes
    colortheme: 'dove'
---

```{r load clean, include = FALSE}
#Load and clean data
rm(list=ls())
setwd('..')
source('Scripts/SW juvenile telomeres IPC load.r')
source('Scripts/SW juvenile telomeres IPC functions.r')
source('Scripts/SW juvenile telomeres IPC clean.r')

```

## **In early life?**

\begin{center}
  \includegraphics{Pictures/SWfeedingcomp.png}
\end{center}

## **Questions**

####1. Is telomere loss greatest in the first year of life?

####2. What factors affect telomere length in early life?

####3. What are the consequences of early life telomere dynamics?

\begin{center}
  \includegraphics{Pictures/SWfeedingcomp.png}
\end{center}

## **Is telomere loss greatest in the first year of life?**

```{r, echo=FALSE,fig.height=5,fig.width=8,fig.cap='Telomere length and age'}
par(pty='s')
errorplot(y = dd$TLKB,
          x = dd$Age,
          myylab = 'Telomere Length (kb)',
          myxlab = 'Age')
polygon(x=c(-0.2,0.81,0.81,-0.2),
        y=c(5.5,5.5,6,6),
        col='white',
        border = NA)
```

## **Is telomere loss greatest in the first year of life?**

```{r, echo=FALSE,fig.height=5,fig.width=8, fig.cap='Telomere length and age'}
par(pty='s')
errorplot(y = dd$TLKB,
          x = dd$Age,
          myylab = 'Telomere Length (kb)',
          myxlab = 'Age')
```

## **Is telomere loss greatest in the first year of life?**

```{r, echo=FALSE,fig.height=5,fig.width=8, fig.cap='Telomere length and juvenile age'}
par(pty='s')
errorplot(y = dd$TLKB,
          x = dd$Ageclass,
          myylab = 'Telomere Length (kb)',
          myxlab = 'Age Class')
```

## **Questions**

####1. Is telomere loss greatest in the first year of life?  
*Yes, but it's lost at a very early stage*

####2. What factors affect telomere length in early life?

####3. What are the consequences of early life telomere dynamics?

\begin{center}
  \includegraphics{Pictures/SWfeedingcomp.png}
\end{center}

## **What factors affect telomere length in early life?**

####Model including only juvenile birds (*N* = `r nrow(juv)`)

```{r, eval = FALSE}
mymodel <- lm(Telomere_Length ~ Mass + 
                                Tarsus_Length + 
                                Age_Class + 
                                Sex + 
                                Territory_Quality + 
                                EPP + 
                                Helper,
              data = juveniles)
```

## **What factors affect telomere length in early life?**

####Key factors:

- Territory Quality
- Age class x Sex
- EPP

\begin{center}
  \includegraphics{Pictures/cousinandchicks.png}
  \end{center}

## **What factors affect telomere length in early life?**
```{r, echo=FALSE,fig.height=5,fig.width=8,fig.cap='Telomere length and territory quality',fig.pos='H'}
par(pty='s')
errorplot(juv$TLKB,juv$LogTQF,
          myxlab = 'Territory Quality',
          myylab = 'Telomere Length (kb)')

```

## **What factors affect telomere length in early life?**
```{r, echo=FALSE,fig.height=4,fig.width=8, fig.cap='Telomere length and Sex * Age'}
layout(matrix(1:2,1,2))
par(mar = c(5,5,2,0))
errorplot(y = m_juv$TLKB,
          x = m_juv$Ageclass,
          myylab = 'Telomere Length (kb)',
          myxlab = 'Age Class',
          mymain = 'Males',
          myylim = c(5,7),
          myxlim <- c(0,1))

par(mar=c(5,3,2,2))
errorplot(y = f_juv$TLKB,
          x = f_juv$Ageclass,
          myylab = '',
          myxlab = 'Age Class',
          mymain = 'Females',
          myylim = c(5,7),
          myxlim <- c(0,1))
```

## **What factors affect telomere length in early life?**
```{r, echo=FALSE,fig.height=4,fig.width=8, fig.cap='Telomere length and EPP * Age'}
layout(matrix(1:2,1,2))
par(mar = c(5,5,2,0))
errorplot(y = juvnotEPP$TLKB,
          x = juvnotEPP$Ageclass,
          myylab = 'Telomere Length (kb)',
          myxlab = 'Age Class',
          mymain = 'Within Pair',
          myylim = c(5,7),
          myxlim <- c(0,1))

par(mar=c(5,3,2,2))
errorplot(y = juvEPP$TLKB,
          x = juvEPP$Ageclass,
          myylab = '',
          myxlab = 'Age Class',
          mymain = 'Extra Pair',
          myylim = c(5,7),
          myxlim <- c(0,1))
```

## **Questions**

####1. Is telomere loss greatest in the first year of life?  
*Yes, but it's lost at a very early stage*

####2. What factors affect telomere length in early life?  
*Age, Sex, Territory Quality and EPP*

####3. What are the consequences of early life telomere dynamics?

\begin{center}
  \includegraphics{Pictures/SWfeedingcomp.png}
\end{center}

## **What are the consequences of early life telomere length?**
```{r, echo=FALSE,fig.height=4,fig.width=8, fig.cap='Telomere length and survival'}

layout(matrix(1:2,1,2))

######
#Survival to next year
######

errorplot(x=juv$SurvivedNextFig,y=juv$TLKB,
          myxlim = c(-0.5,1.5),
          myylim = c(5,6),
          myylab = 'Telomere Length (kb)',
          myxlab = 'Survived to adulthood?')

######
#Remaining Life
######

plot1(RemainingLife~TLKB,
      mydata=juv,
      myxlab = 'Telomere Length (kb)',
      myylab = 'Remaining Life',
      myylim=c(0,20))

#Run a model(for fit line)
glm1 <- zeroinfl(RemainingLife~TLKB,
      data=juv,
      dist='poisson')

#Get fitted values and fit the line
predProbs<-predict(glm1,data.frame(TLKB=seq(1, 12, length.out=100)), type="response")
lines(seq(1, 12, length.out=100), predProbs, col='red', lwd=2)

```

## **What are the consequences of early life telomere length?**
```{r, echo=FALSE,fig.height=4,fig.width=8, fig.cap='Telomere length and survival'}

layout(matrix(1:2,1,2))

######
#Survival to next year
######

errorplot(x=m_juv$SurvivedNextFig,y=m_juv$TLKB,
          myxlim = c(-0.5,1.5),
          myylim = c(4,7),
          myylab = 'Telomere Length (kb)',
          myxlab = 'Survived to adulthood?')
errorplot(x=f_juv$SurvivedNextFig,y=f_juv$TLKB,
          add=T,
          mylegend=T,
          legendlevels = c('Males','Females'))

######
#Remaining Life
######

plot1(RemainingLife~TLKB,
      mydata=juv,
      myxlab = 'Telomere Length (kb)',
      myylab = 'Remaining Life',
      myylim=c(0,20))

#Run a model(for fit line)
glmMale <- zeroinfl(RemainingLife~TLKB,
      data=m_juv,
      dist='poisson')
glmFemale <- zeroinfl(RemainingLife~TLKB,
      data=f_juv,
      dist='poisson')

#Get fitted values and fit the line
predProbs<-predict(glmMale,data.frame(TLKB=seq(1, 12, length.out=100)), type="response")
lines(seq(1, 12, length.out=100), predProbs, col='red', lwd=2)
predProbs<-predict(glmFemale,data.frame(TLKB=seq(1, 12, length.out=100)), type="response")
lines(seq(1, 12, length.out=100), predProbs, col='blue', lwd=2)

legend('topleft',
       pch='--',
       bty='n',
       col=c('red','blue'),
       legend=c('Males','Females'),
       cex=1.2)

```

## **What are the consequences of early life telomere length?**
```{r, echo=FALSE,fig.height=4,fig.width=8, fig.cap='Telomere length and survival'}

layout(matrix(1:2,1,2))

######
#Survival to next year
######

errorplot(x=juvEPP$SurvivedNextFig,y=juvEPP$TLKB,
          myxlim = c(-0.5,1.5),
          myylim = c(4.5,7.5),
          myylab = 'Telomere Length (kb)',
          myxlab = 'Survived to adulthood?')
errorplot(x=juvnotEPP$SurvivedNextFig,y=juvnotEPP$TLKB,
          add=T,
          mylegend=T,
          legendlevels = c('Extra Pair','Within Pair'))

######
#Remaining Life
######

plot1(RemainingLife~TLKB,
      mydata=juv,
      myxlab = 'Telomere Length (kb)',
      myylab = 'Remaining Life',
      myylim=c(0,20))

#Run a model(for fit line)
glmEPP <- zeroinfl(RemainingLife~TLKB,
      data=juvEPP,
      dist='poisson')
glmnotEPP <- zeroinfl(RemainingLife~TLKB,
      data=juvnotEPP,
      dist='poisson')

#Get fitted values and fit the line
predProbs<-predict(glmEPP,data.frame(TLKB=seq(1, 12, length.out=100)), type="response")
lines(seq(1, 12, length.out=100), predProbs, col='red', lwd=2)
predProbs<-predict(glmnotEPP,data.frame(TLKB=seq(1, 12, length.out=100)), type="response")
lines(seq(1, 12, length.out=100), predProbs, col='blue', lwd=2)

legend('topleft',
       pch='--',
       bty='n',
       col=c('red','blue'),
       legend=c('Extra Pair','Within Pair'),
       cex=1.2)

```

## **Questions**

####1. Is telomere loss greatest in the first year of life?  
*Yes, but it's lost at a very early stage*

####2. What factors affect telomere length in early life?  
*Age, Sex, Territory Quality and EPP*

####3. What are the consequences of early life telomere dynamics?
*Important for survival but dependent on early life conditions*

\begin{center}
  \includegraphics{Pictures/SWfeedingcomp.png}
\end{center}

## **Thanks!**

- Pat Monaghan
- Winnie Boner
- Simon Verhulst

**Seychelles warbler study group**  (*Emma Barrett*, Jan Komdeur, Terry Burke and all the other warbler researchers)

\begin{center}
  \includegraphics{Pictures/UEAlogo.png}
\end{center}

\begin{center}
  \includegraphics{Pictures/NERClogo.png}
\end{center}

### 2015-06-22 Talks/Spurgin_Rebellion_2015.Rmd
First upload of all data and scripts

---
output:
  beamer_presentation:
    fig_caption: false
    colortheme: 'seahorse'
---

```{r setwd, include = F}

#Set working directory up one level
setwd('..')

#Load and clean data
source('Scripts/SW juvenile telomeres IPC load.r')
source('Scripts/SW juvenile telomeres IPC functions.r')
source('Scripts/SW juvenile telomeres IPC clean.r')

#Set some options

opts_chunk$set(echo=FALSE,
               warning=FALSE,
               message=FALSE)

```

## Good Years, Bad Years and Telomeres in a Wild Bird Population

\begin{center}
  \includegraphics{Pictures/SWfeedingcomp2.png} \\~\\

Lewis Spurgin \\~\\

March 2015 \\~\\

  \includegraphics{Pictures/UEANERCwide.png}

\end{center}

## The benefits of a good start

\begin{center}
  \includegraphics{Pictures/Bullingdon.png}
\end{center}

## The benefits of a good start

\begin{center}
  \includegraphics{Pictures/Slytherin.png}
\end{center}

## A good/bad start doesn't affect everyone equally

\begin{center}
  \includegraphics{Pictures/powers.png}
\end{center}

## Silver spoons in ecology

\begin{center}
  \includegraphics{Pictures/DeerFig.png}
\end{center}

## Silver spoons in ecology

\begin{center}
  \includegraphics{Pictures/GreatTitFig.png}
\end{center}

## Can we quantify how badly an individual suffers from a bad start in life?

\begin{center}
  \includegraphics{Pictures/BirdOfParadise.png}
\end{center}

## Telomeres - (TTAGGG)~*n*~

\begin{center}
  \includegraphics{Pictures/Telomeres.png}
\end{center}

## Telomeres and oxidative stress

\begin{center}
  \includegraphics{Pictures/Oxidants.png}
\end{center}

## Telomeres - biomarkers of cost

\begin{center}
  \includegraphics{Pictures/Google.png}
\end{center}

## Telomeres and silver spoons

\begin{center}
  \includegraphics{Pictures/ShagFig.png}
\end{center}

## Telomeres and silver spoons

\begin{center}
  \includegraphics{Pictures/HeidingerFig.png}
\end{center}

## Seychelles warbler (*Acrocephalus sechellensis*)

\begin{center}
  \includegraphics{Pictures/CousinAndWarbler.png}
\end{center}

## Seychelles warbler (*Acrocephalus sechellensis*)

\begin{center}
  \includegraphics{Pictures/SWnest.png}
\end{center}

## It rains in the tropics

\begin{center}
  \includegraphics{Pictures/Rain.png}
\end{center}

## Food availiability varies over time

```{r food availability over time,fig.width=4,fig.height=4}
temp <- subset(FlSAall,LayYear!='1996')

temp$LayYear <- as.numeric(paste(temp$LayYear))
Fig1A <- ggplot(temp,
       aes(x = LayYear,
           y = Insect)) +
  xlab('Year') + 
  ylab('Insect availability') + 
  theme_classic() +
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14),
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16)) +
  geom_ribbon(aes(ymax=Insect+InsectSE,ymin=Insect-InsectSE),alpha=0.2) +
  geom_path()

Fig1A

```

## Food availiability in early life doesn't predict lifespan in the Seychelles warbler

```{r Lifespan and food availability,fig.width=4,fig.height=4}

temp <- FlSAall
temp$Insect <- factor(temp$Insect)

ddFig1A <- ddply(temp,
                 .(Insect),
                 summarize,
                 TLKBmean = mean(RemainingLife),
                 TLKBse = 1.96*se(RemainingLife))
ddFig1A <- ddFig1A[complete.cases(ddFig1A),]
ddFig1A$Insect <- as.numeric(paste(ddFig1A$Insect))

Fig1A <- ggplot(ddFig1A,
       aes(x = Insect,
           y = TLKBmean)) +
  ylab('Lifespan') + 
  xlab('Food availability in early life') + 
  theme_classic() +
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16)) +

  geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse))

Fig1A

```

## Predictions

**Telomeres link early life conditions to late life survival**

- Juvenile telomere length varies over years

- Telomere length in early life is related  to food availability

- Individuals who suffer greatest telomere shortening in early life have reduced late life survival

\begin{center}
  \includegraphics{Pictures/BirdOfParadiseSmall.png}
\end{center}

## Measuring warbler telomeres

\begin{center}
  \includegraphics{Pictures/Emma.png}
\end{center}

## Juvenile telomere length varies over years

```{r Telomere length over time,fig.width=4,fig.height=4}

temp <- FlSA

ddFigS2 <- ddply(temp,
                 .(LayYear),
                 summarize,
                 TLKBmean = mean(TLKB),
                 TLKBse = se(TLKB))

ddFigS2 <- ddFigS2[complete.cases(ddFigS2),]
ddFigS2$LayYear <- as.numeric(paste(ddFigS2$LayYear))

FigS2 <- ggplot(ddFigS2,
       aes(x = LayYear,
           y = TLKBmean)) +
  geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse)) +
  geom_line() +
  ylab('Telomere length (kb)') + 
  xlab('Lay Year') + 
  theme_classic() +
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16))

FigS2

```

## Telomere length in early life is related  to food availability

```{r Telomeres and food availability,fig.width=4,fig.height=4}

temp <- FlSAall
temp$Insect <- factor(temp$Insect)

ddFig1A <- ddply(temp,
                 .(Insect),
                 summarize,
                 TLKBmean = mean(TLKB),
                 TLKBse = 1.96*se(TLKB))
ddFig1A <- ddFig1A[complete.cases(ddFig1A),]
ddFig1A$Insect <- as.numeric(paste(ddFig1A$Insect))

Fig1A <- ggplot(ddFig1A,
       aes(x = Insect,
           y = TLKBmean)) +
  ylab('Telomere length (kb)') + 
  xlab('Food availability in early life') + 
  theme_classic() +
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16)) +
  stat_smooth(data=FlSAall,aes(x=Insect,y=TLKB),method=lm,col=grey(0.7)) +
  geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse))

Fig1A

```

## Individuals who suffer greatest telomere shortening in early life have reduced late life survival

```{r Telomeres and survival,fig.width=5.5,fig.height=4}

Fig2 <- ggplot(subset(subset(FlSAall,RemainingLife>1),RemainingLife<11),
                aes(y = RemainingLife,
                x = TLKB,
                col=YQ)) +
  theme_classic() +
  geom_point(alpha=0.3) +
  xlab('Telomere length (kb)') + 
  ylab('Remaining life') + 
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.x = element_text(size = 16), 
        axis.title.y = element_text(size = 16,vjust = 1), 
        legend.text = element_text(size = 13),
        legend.title = element_blank()) +
  scale_color_manual(values=c('#d95f02','#1b9e77'))+
  scale_fill_manual(values=c('#d95f02','#1b9e77'))+
  stat_smooth(method=glm,family=poisson,lwd=1.2,aes(fill=YQ,lty=YQ),fullrange=T)

Fig2

```

## Predictions

**Telomeres link early life conditions to late life survival**

- Juvenile telomere length varies over years

- Telomere length in early life is related  to food availability

- Individuals who suffer greatest telomere shortening in early life have reduced late life survival

\begin{center}
  \includegraphics{Pictures/BirdOfParadiseSmall.png}
\end{center}

## Predictions

**Telomeres link early life conditions to late life survival**

- Juvenile telomere length varies over years **YES**

- Telomere length in early life is related  to food availability

- Individuals who suffer greatest telomere shortening in early life have reduced late life survival

\begin{center}
  \includegraphics{Pictures/BirdOfParadiseSmall.png}
\end{center}

## Predictions

**Telomeres link early life conditions to late life survival**

- Juvenile telomere length varies over years **YES**

- Telomere length in early life is related  to food availability **YES**

- Individuals who suffer greatest telomere shortening in early life have reduced late life survival

\begin{center}
  \includegraphics{Pictures/BirdOfParadiseSmall.png}
\end{center}

## Predictions

**Telomeres link early life conditions to late life survival**

- Juvenile telomere length varies over years **YES**

- Telomere length in early life is related to food availability **YES**

- Individuals who suffer greatest telomere shortening in early life have reduced late life survival **YES**

\begin{center}
  \includegraphics{Pictures/BirdOfParadiseSmall.png}
\end{center}

## Thank you!

\begin{center}
  \includegraphics{Pictures/Acks.png}
\end{center}

### 2015-06-22 Talks/Warbler meeting December 2014.Rmd
First upload of all data and scripts

--- 2015-06-22 Talks/SW early life telomeres Glasgow.Rmd
@@
 ---
-output:
-  beamer_presentation:
-    fig_caption: yes
-    colortheme: 'dove'
+title: "Early life telomere dynamics"
+author: "Lewis Spurgin"
+date: "December 2014"
+output: beamer_presentation
 ---
@@
 source('Scripts/SW juvenile telomeres IPC functions.r')
-source('Scripts/SW juvenile telomeres IPC clean.r')
+source('Scripts/SW juvenile telomeres IPC clean v2.r')
 
@@
 
-## **In early life?**
+##Late life consequences of early life experiences
+\begin{center}
+  \includegraphics{Pictures/BrouwerFig.png}
+\end{center}
 
+##Telomeres as biomarkers of early life costs
 \begin{center}
-  \includegraphics{Pictures/SWfeedingcomp.png}
+  \includegraphics{Pictures/HeidingerFig.png}
 \end{center}
@@
 \end{center}
+
+## **Data**
+- Juvenile birds caught between 1997-2000 (plus a few later ones)
+
+- Mostly Dead
+
+- *N* = `r nrow(juv)`
 
@@
 
+tapply(juv$LogTL,list(juv$EPP,juv$Sex),length)
+
 ## **Questions**
@@
 \end{center}
-
-## **What factors affect telomere length in early life?**
-
-####Model including only juvenile birds (*N* = `r nrow(juv)`)
-
-```{r, eval = FALSE}
-mymodel <- lm(Telomere_Length ~ Mass + 
-                                Tarsus_Length + 
-                                Age_Class + 
-                                Sex + 
-                                Territory_Quality + 
-                                EPP + 
-                                Helper,
-              data = juveniles)
-```
 
@@
           myylim = c(5,7),
+          myxlim <- c(0,1))
+```
+
+## **What factors affect telomere length in early life?**
+```{r, echo=FALSE,fig.height=4,fig.width=8, fig.cap='Telomere length and EPP * Age'}
+layout(matrix(1:2,1,2))
+par(mar = c(5,5,2,0))
+errorplot(y = juvHelp$TLKB,
+          x = juvHelp$Ageclass,
+          myylab = 'Telomere Length (kb)',
+          myxlab = 'Age Class',
+          mymain = 'Helpers present',
+          myylim = c(4.5,7),
+          myxlim <- c(0,1))
+
+par(mar=c(5,3,2,2))
+errorplot(y = juvNotHelp$TLKB,
+          x = juvNotHelp$Ageclass,
+          myylab = '',
+          myxlab = 'Age Class',
+          mymain = 'Helpers absent',
+          myylim = c(4.5,7),
           myxlim <- c(0,1))
@@
 
+##Telomeres as biomarkers of early life costs
+\begin{center}
+  \includegraphics{Pictures/BoonekampFig.png}
+\end{center}
+
 ## **Questions**
@@
 \end{center}
-
-## **Thanks!**
-
-- Pat Monaghan
-- Winnie Boner
-- Simon Verhulst
-
-**Seychelles warbler study group**  (*Emma Barrett*, Jan Komdeur, Terry Burke and all the other warbler researchers)
-
-\begin{center}
-  \includegraphics{Pictures/UEAlogo.png}
-\end{center}
-
-\begin{center}
-  \includegraphics{Pictures/NERClogo.png}
-\end{center}

### 2015-06-22 Scripts/SW juvenile telomeres IPC functions.r

@@ -1,0 +1,426 @@
+#################################################################################*
+## Early life telomere dynamics and late-life fitness in a wild bird population
+## DEFINE FUNCTIONS
+#################################################################################*
+
+
+
+
+
+
+# Model and table functions -----------------------------------------------
+
+#Order interaction terms alphabetically (needed for function below)
+orderinteraction <- function(x)
+{terms <- unlist(strsplit(x,':'))
+ newterms <- terms[order(terms)]
+ paste(newterms,collapse=':')
+}
+
+
+
+#Randomise a model and get baseline relative importance
+baselineimp <- function(mymodel,n.permutations = 1000)
+{
+  my.exp.terms <- attr(terms(formula(mymodel)),'term.labels')
+  my.exp.terms <- unlist(lapply(my.exp.terms,orderinteraction))
+  myresponse <- paste(formula(mymodel)[[2]])
+  mydata <- mymodel$model         
+  output <- mat.or.vec(n.permutations,length(my.exp.terms))
+  colnames(output) <- my.exp.terms
+  
+  for(i in 1:n.permutations)
+  {
+    tempdata <- mydata
+    tempdata[,myresponse] <- sample(tempdata[,myresponse])
+    newmodel <- update(mymodel,data = tempdata)
+    allmodels <- dredge(newmodel)
+    mymodelavg <- model.avg(allmodels,subset = delta <7)
+    myimp <- importance(mymodelavg)
+    
+    for(j in 1:length(myimp))
+    {
+      currentterm <- names(myimp)[j]
+      currentcol <- which(colnames(output) == currentterm)
+      output[i,paste(currentterm)] <- myimp[j]
+    }
+  }
+  return(output)
+}
+
+
+
+
+
+#Correct relative importance according to baseline importance
+correctedimp <- function(importance,baselineimportance)
+{
+  bl <- baselineimportance[,colnames(baselineimportance) %in% names(importance)]
+  bl <- bl[,order(colnames(bl))]
+  imp <- importance[order(names(importance))]
+  blmean <- apply(bl,2,mean)
+  return(imp-blmean)
+}
+
+
+
+
+#Combine summary table from modelavereging with relative importance values
+#to produce a publication-ready table
[…]

### 2015-06-22 Scripts/SW juvenile telomeres IPC load.r

@@ -1,0 +1,30 @@
+#################################################################################*
+## Early life telomere dynamics and late-life fitness in a wild bird population
+## LOAD THE DATA AND PACKAGES
+#################################################################################*
+
+
+# Clear out R -------------------------------------------------------------
+rm(list=ls())
+
+
+# Load relevant libraries -------------------------------------------------
+library(knitr)
+library(MuMIn)
+library(arm)
+#library(beepr)
+#library(pscl)
+library(plyr)
+library(ggplot2)
+#library(car)
+library(gvlma)
+#library(scales)
+library(wesanderson)
+library(survival)
+library(flexsurv)
+
+# Load data ---------------------------------------------------------------
+dd <- read.csv('Data/SW TL main data for analysis.csv') #main dataset
+terr <- read.csv('Data/SW TL all territories.csv')
+status <- read.csv('Data/SW TL all breedstatus.csv')
+insects <- read.csv('Data/Insects.csv')

### 2015-06-22 Manuscript/SW_early_life_telomeres_SOM_v9.Rmd
Added model selection table, removed logistic regression table (as now doing survival analysis instead)

@@
 ```
+
+**Table S1** Model selection table for analysis of factors affecting early life telomere length in juvenile Seychelles warblers.
+Only models that remained in the top model set for model averaging (AICc $\leq$ 6 compared to the top model) are included here.
+i = intercept.
+
+```{r Table S1}
+
+#Run the model
+lm1 <- lmer(TLKB ~ TQspace + Insect + GroupSize + Sex + Condition + Helper + (1|LayYear),
+            data = FlSAall,REML=FALSE)
+
+lm1 <- standardize(lm1)
+
+#Model averaging
+lm2 <- dredge(lm1)
+t1 <- as.matrix(subset(lm2, delta<=6))
+colnames(t1)[2:7] <- c('Sex', 'Body condition','Group size','Helper','Food av.','TQ')
+getterms <- function(x)
+  {
+  temp <- c('i',names(which(!(is.na(x[2:7])))))
+  return(paste(temp,collapse = " + "))
+  }
+
+tableS1 <- data.frame(Model = apply(t1,1,getterms),t1[,c('df','AICc','delta','weight')])
+colnames(tableS1) <- c('Model','df','AICc', 'Delta AICc','Weight')
+rownames(tableS1) <- NULL
+tableS1$Model[tableS1$Model == ''] <- 'Null model'
+kable(tableS1,digits=3)
+
+```
+
+PAGEBREAK
 
@@
 Lines and shaded areas represent fitted values and 95% confidence limits from linear regressions.
-
-```{r Figure S4,fig.width=5,fig.height=5}
-glm1 <- glm(SurvivedNext ~ Sex + Condition + TLKB + Insect + Ageclass + TLKB + GroupSize,
-            data = FlSAall,
-            family = 'binomial')
-glm1 <- standardize(glm1)
-glm2 <- dredge(glm1)
-glm2avg <- model.avg(glm2, subset = delta <=6)
-myimp2 <- importance(glm2avg)
-
-#Make into a nice table
-table2 <- imptable(glm2avg,myimp2)
-table2 <- signif(table2,2)
-
-#Sort out row names
-nams <- rownames(table2)
-nams[nams == 'c.Ageclass'] <- 'Age class'
-nams[nams == 'z.Condition'] <- 'Body condition'
-nams[nams == 'c.Sex'] <- 'Sex'
-nams[nams == 'z.Insect'] <- 'Food av.'
-nams[nams == 'z.TQspace'] <- 'TQ'
-nams[nams == 'z.TLKB'] <- 'TL'
-nams[nams == 'z.GroupSize'] <- 'Local density'
-nams[nams == 'z.Insect:z.TLKB'] <- 'TL x Food av.'
-
-rownames(table2) <- nams
-colnames(table2) <- c('Estimate','SE','2.5%','97.5%','RI')
-
-table2 <- data.frame(table2,myx=row.names(table2))
-
-forestplot(table2)
-
-```
-
-**Figure S4** Model averaged estimates and confidence intervals (points and error bars, respectively) for all explanatory terms used in a generalised linear model with a binomial error structure and survival to adulthood as the response.
-Numbers in brackets are the relative importance of each term in the top model set (see main text for details).
-Food av = annual food availability, TL = telomere length TQ = territory quality.

### 2015-06-22 Manuscript/SW_early_life_telomeres_tables_and_figures v9.Rmd
Incorporating reviewers comments from Proc B, clarifying survival analysis

@@
                fig.path='../Figures/',
-               fig.width= 5,
-               fig.height= 5,
+               fig.width= 4,
+               fig.height= 8,
                fig.show='asis',
@@
 
-x <- summary(lm(TLKB~Insect,data= FlSAall))$resid
-xf <- ifelse(x>median(x),'Long','Short')
+t1 <- data.frame(with(FlSAall,flexsurvreg(Surv(Lifespan+1,Died)~Condition+TLKB+Insect, dist="lnorm"))$res)
+t1 <- t1[3:5,]
+rownames(t1) <- c('Body condition', 'Tel. length', 'Food av.')
 
-s1 <- survfit(Surv(time = Lifespan+1, event = Died)~xf, data = FlSAall)
-Fig2 <- ggsurv(s1,plot.cens=F) + 
+Fig2A <- ggplot(t1,aes(x = rownames(t1),y = est)) +
+  geom_point() + 
+  geom_pointrange(aes(ymin = L95.,ymax = U95.)) +
+    theme_classic() +
+  ylab('Estimate') + 
+  xlab('') + 
+  theme_classic() +
+  theme(axis.text.x = element_text(angle=90,size = 14), 
+        axis.text.y = element_text(size = 14), 
+        axis.title.y = element_text(size = 16,vjust=0.8),
+        axis.title.x = element_text(size = 16),
+        legend.position = c(0.8,0.8),
+        legend.title = element_text(size = 14),
+        legend.text = element_text(size = 12)) +
+  geom_abline(intercept = 0,slope = 0,lty=2)
+
+s1 <- survfit(Surv(time = Lifespan+1, event = Died)~TLF, data = FlSA)
+Fig2B <- ggsurv(s1,plot.cens=F) + 
   theme_classic() +
@@
         legend.text = element_text(size = 12)) +
-  scale_colour_manual(name = 'Residual TL',
+  scale_colour_manual(name = 'TL',
                       values = c('darkred','red')) +
   guides(linetype=F) +
-  theme(legend.title = element_text('Telomere length'))
+  theme(legend.title = element_blank())
 
-Fig2
+multiplot(Fig2A,Fig2B)
 

### 2015-06-22 Manuscript/SW_early_life_telomeres_v9.Rmd
Incorporating reviewers comments from Proc B, clarifying survival analysis

@@
 We performed all analyses using R version 3.0.1 [@RDevelopmentCoreTeam2011].
-We used linear models (LMs) and generalised linear models (GLMs) with model averaging to assess the causes and consequences of early-life telomere length in the Seychelles warbler.
-For each analysis we created a full model with all terms and biologically relevant interactions, from which a top model set was defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011].
+We first tested whether early-life environmental variation influences telomere length using linear mixed models, with telomere length as the response variable.
+As explanatory variables we included body condition (calculated as deviation from a regression line of mass on tarsus length; [@VandeCrommenacker2011a]), sex, age class (juvenile or subadult), territory quality, yearly food availability and local density (measured as the number of individuals in the juvenile's natal social group [@Brouwer2012]).
+Lay Year was included as a random factor, in order to account for any annual variation in telomere length not explained by yearly food availability.
+We created a full model containing all of the above terms, and a top model set was then defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011].
 We report model averaged coefficients, confidence intervals, and 'relative importance', which reflects the relative weights of each predictor variable across the top model set.
@@
 
-We first tested whether early-life environmental variation influences telomere length using linear models, with telomere length as the response variable.
-As explanatory variables we included body condition (calculated as deviation from a regression line of mass on tarsus length; [@VandeCrommenacker2011a]), sex, age class (juvenile or subadult), territory quality, yearly food availability and local density (measured as the number of individuals in the juvenile's natal social group [@Brouwer2012]).
-
-We then used generalised linear models to examine variation in telomere length in relation to survival.
-In these models we included the same predictor variables as the previous analysis, plus early-life telomere length.
-We first performed logistic regression to test whether telomere length predicts survival to adulthood (response = survived to the following year; yes/no).
-We then used Poisson regression, excluding individuals that died in their first year of life, to test whether telomere length predicts late-life survival (in years).
+We then used parametric survival analysis to examine variation in telomere length in relation to survival, implemented using the 'flexsurv' package in R.
+The survival time distribution was chosen by comparing the fit (using visual inspection of plots and a comparison of AIC values) of a range of distributions (exponential, Weibull, log-normal, gamma, gompertz and generalized F) against the observed survival data.
+As explanatory variables of survival we included juvenile telomere length, plus the factors identified as being related to telomere length in the previous analysis.
+Lay year was included as a random effect.
 
@@
 This is only marginally higher than mean telomere length for all adult birds (`r round2(mean(subset(dd,Age>0)$TLKB))` $\pm$ `r round2(se(subset(dd,Age>0)$TLKB))` kb), and lower than the mean for one year-old birds based on cross-sectional data (`r round2(mean(subset(dd,Age==1)$TLKB))` $\pm$ `r round2(se(subset(dd,Age==1)$TLKB))` kb; Figure S1).
-Using longitudinal data, however, we found that telomere length was shorter in young adult birds compared to when they were juveniles (mean loss between juvenile and first adult sample = `r round(mean(FlSA$TLloss,na.rm=T),1)*1000` bp).
 We found no significant difference in telomere length between age classes (juvenile vs subadult), or between sexes (Figure 1A).
@@
 Model averaging confirmed this strong, positive effect of yearly food availability on telomere length (Figure 1A).
-The best supported model with early-life telomere length as the response variable contained annual food availability, body condition and local density as explanatory variables (R^2^ = `r round2(summary(lm(TLKB~YQ+GroupSize+Condition,data=FlSAall))$r.squared)`).
-Early-life telomere length  was positively related to body condition (Figure 1C), and negatively related to local density, although in the latter case model-averaged confidence intervals overlapped zero (Figure 1A).
+The best supported model with early-life telomere length as the response variable contained annual food availability and body condition as explanatory variables (R^2^ = `r round2(summary(lm(TLKB~Insect+GroupSize+Condition,data=FlSAall))$r.squared)`).
+Early-life telomere length  was positively related to body condition (Figure 1C).
 
@@
 
-Annual food availability and local density were the most important factors predicting survival to adulthood in juvenile Seychelles warblers (Figure S4).
-The top model contained these two factors, but was very weak in terms of explanatory power (McFadden's pseudo R^2^ = `r round2(1-logLik(glm(SurvivedNext~TLKB*Insect+GroupSize,family='binomial',data=FlSAall))/logLik(glm(SurvivedNext~1,family='binomial',data=FlSAall)))`), and the confidence intervals for all explanatory terms overlapped zero (Figure S4), suggesting that none of the variables studied predict survival to adulthood.
-
-Early-life telomere length and annual food availability, plus the interaction between these two variables, were the top predictors of late-life survival (Figure 2A).
-The interaction between telomere length and annual food availability was in the top model explaining late-life survival, and this model had good explanatory power (McFadden's pseudo R^2^ = `r round2(1-logLik(glm(RemainingLife~TLKB*Insect,family='poisson',data=subset(FlSAall,RemainingLife>0)))/logLik(glm(RemainingLife~1,family='poisson',data=FlSAall)))`).
-There was a positive relationship between telomere length and late-life survival in years of low insect abundance, with no such relationship when food availability was high (Figure 2B).
+A log-normal survival model best fitted the Seychelles warbler survival data (AIC = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="lnorm")$AIC)`, AIC of next best distribution = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="weibull")$AIC)`).
+A survival model including telomere length, body condition and annual food availability showed that only telomere length had an on survival (Fig.
+2A).
+Because of the strong correlation between telomere length and
 
@@
 ###Data accessibility
-All data required to reproduce the analyses within this manuscript have been uploaded to Dryad (XXXX)
+All data and scripts required to reproduce the analyses within this manuscript are available on GitHub (https://github.com/lewisspurgin/SW_Early_life_telomeres)
 
@@
 ###Authors' contributions
-DSR developed the idea, obtained funding for the overall telomere project and managed the study; DSR, JK, HLD and TB supervised the overall running of the overall Seychelles warbler project and contributed resources; HLD managed the longitudinal fitness database.
+DSR developed the idea and obtained funding for the overall telomere project; DSR, JK, HLD and TB supervised the overall running of the overall Seychelles warbler project and contributed resources; HLD managed the longitudinal fitness database.
 EAF and KLB generated and processed the telomere data; LGS performed the analyses and wrote the manuscript in discussion with DSR.

### 2015-06-22 Scripts/SW juvenile telomeres IPC clean.R

@@ -150,7 +150,7 @@ dd <- subset(dd,BodyMass>11)
 
 
 mymed <- median(dd$TLKB,na.rm=T)
-dd$TLF <- ifelse(dd$TLKB > mymed,'Long','Short')
+dd$TLF <- ifelse(dd$TLKB > mymed,'Long telomeres','Short telomeres')
 
 # Subset juveniles --------------------------------------------------------------
 

### 2015-06-22 Scripts/SW juvenile telomeres IPC functions.r

@@ -402,7 +402,7 @@ ggsurv <- function(s, CI = 'def', plot.cens = T, surv.col = 'gg.def',
 }
 
 
-# In-tect functions -------------------------------------------------------
+# In-text functions -------------------------------------------------------
 
 
 #Round to 2 or 3 dp and keep trailing zeros

### 2015-08-14 Manuscript/SW_early_life_telomeres_SOM_v9.Rmd
Added new data, included chicks, updated figures

@@
 #Run the model
-lm1 <- lmer(TLKB ~ TQspace + Insect + GroupSize + Sex + Condition + Helper + (1|LayYear),
+lm1 <- lmer(TLKB ~ Agemonths + TQspace + Insect + GroupSize + Sex + Condition + Helper + (1|LayYear),
             data = FlSAall,REML=FALSE)
@@
 **Figure S2** Cross-sectional juvenile telomere length (mean $\pm$ s.e.) across birth years in Seychelles warblers
-
-PAGEBREAK
-
-```{r Figure S3,fig.width=5}
-
-temp <- subset(FlSA,LayYear %in% c(1999,2006,2007,2000))
-
-FigS3A <- ggplot(temp,
-      aes(x=GroupSize,
-      y=TLKB,
-      col=LayYear)) +
-  stat_smooth(method=lm,fullrange=T,alpha=0.2, aes(fill=LayYear),lwd=1.2) + 
-    ylab('Telomere length (kb)') + 
-  xlab('Local Density') + 
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 14), 
-        axis.text.y = element_text(size = 14), 
-        axis.title.y = element_text(size = 16,vjust=0.8),
-        axis.title.x = element_text(size = 16)) +
-  scale_color_brewer(palette='Set1') + 
-  scale_fill_brewer(palette='Set1') +
-  annotate('text',x=2,y=10,label='A')
-
-FigS3B <- ggplot(temp,
-      aes(x=Condition,
-      y=TLKB,
-      col=LayYear)) +
-  stat_smooth(method=lm,fullrange=T,alpha=0.2, aes(fill=LayYear),lwd=1.2) + 
-    ylab('Telomere length (kb)') + 
-  xlab('Body Condition') + 
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 14), 
-        axis.text.y = element_text(size = 14), 
-        axis.title.y = element_text(size = 16,vjust=0.8),
-        axis.title.x = element_text(size = 16)) +
-  scale_color_brewer(palette='Set1') + 
-  scale_fill_brewer(palette='Set1') +
-  annotate('text',x=-2,y=15,label='B')
-
-multiplot(FigS3A,FigS3B)
-
-```
-
-**Figure S3** Within cohort relationships between telomere length and **A** local density and **B** body condition in juvenile Seychelles warblers.
-Only years with juvenile sample sizes >20 are included.
-Lines and shaded areas represent fitted values and 95% confidence limits from linear regressions.

### 2015-08-14 Manuscript/SW_early_life_telomeres_tables_and_figures v9.Rmd
Added new data, included chicks, updated figures

--- 2015-06-22 Manuscript/SW_early_life_telomeres_tables_and_figures v9.Rmd
@@
 #Run the model
-lm1 <- lmer(TLKB~ TQspace + Insect + GroupSize + Sex + Condition + Helper + (1|LayYear),
+lm1 <- lmer(TLKB ~ TQspace + Insect + NonHelper + Sex + Tarsus + BodyMass + Helper + (1|LayYear),
+            data = chickall,REML=FALSE)
+
+lm1 <- standardize(lm1)
+
+#Model averaging
+lm2 <- dredge(lm1)
+lmavg <- model.avg(lm2,subset = delta <=6)
+myimp <- importance(lmavg) #Relative importance of each explanatory variable
+
+#Put it all into a nice table
+table1 <- imptable(lmavg,myimp)
+table1 <- signif(table1,2)
+
+#Sort out row/col names
+nams <- rownames(table1)
+nams[nams == 'c.Ageclass'] <- 'Age class'
+nams[nams == 'z.BodyMass'] <- 'Body mass'
+nams[nams == 'z.Tarsus'] <- 'Tarsus'
+nams[nams == 'z.NonHelper'] <- 'Non helpers'
+nams[nams == 'c.Sex'] <- 'Sex'
+nams[nams == 'z.Insect'] <- 'Food av.'
+nams[nams == 'z.TQspace'] <- 'TQ'
+nams[nams == 'z.Helper'] <- 'Helpers'
+
+rownames(table1) <- nams
+colnames(table1) <- c('Estimate','SE','2.5%','97.5%','RI')
+
+table1 <- data.frame(table1,myx=row.names(table1))
+
+Fig1A <- forestplot(table1) +
+  annotate('text',x=0.5,y=4,label='A')
+
+######################################
+temp <- chickall
+temp$Helper <- ifelse(temp$Helper>0,'Helpers present','Helpers absent')
+
+ddFig1B <- ddply(temp,
+                 .(Helper),
+                 summarize,
+                 TLKBmean = mean(TLKB),
+                 TLKBse = se(TLKB))
+
+Fig1B <- ggplot(ddFig1B,
+       aes(x = Helper,
+           y = TLKBmean)) +
+  ylab('Telomere length (kb)') + 
+  xlab('') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 14), 
+        axis.text.y = element_text(size = 14), 
+        axis.title.y = element_text(size = 16,vjust=0.8),
+        axis.title.x = element_text(size = 16)) +
+  geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse)) +
+  scale_colour_manual(values = wes_palette("Royal1")) +
+  annotate('text',x=0.5,y=7.5,label='B')
+
+Fig1C <- ggplot(chickall,
+       aes(x = BodyMass,
+           y = TLKB)) +
+  ylab('Telomere length (kb)') + 
+  xlab('Body mass (g)') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 14), 
+        axis.text.y = element_text(size = 14), 
+        axis.title.y = element_text(size = 16,vjust=0.8),
+        axis.title.x = element_text(size = 16)) +
+  geom_point() +
+  stat_smooth(method = 'lm',col = 'black') +
+  annotate('text',x=12,y=15,label='C')
+
+multiplot(Fig1A,Fig1B,Fig1C,layout=matrix(c(1,1,1,1,1,1,2,3,2,3), nrow=5, byrow=TRUE))
+
+```
+
+PAGEBREAK
+
+```{r Figure 2,fig.width=5,fig.height=9}
+###########
+#Fig 2A
+
+#Run the model
+lm1 <- lmer(TLKB ~ TQspace + Insect + NonHelper + Sex + Tarsus + BodyMass + Helper + (1|LayYear),
             data = FlSAall,REML=FALSE)
@@
 nams[nams == 'c.Ageclass'] <- 'Age class'
-nams[nams == 'z.Condition'] <- 'Body condition'
-nams[nams == 'z.GroupSize'] <- 'Local density'
+nams[nams == 'z.BodyMass'] <- 'Body mass'
+nams[nams == 'z.Tarsus'] <- 'Tarsus'
+nams[nams == 'z.NonHelper'] <- 'Non helpers'
 nams[nams == 'c.Sex'] <- 'Sex'
@@
 
-Fig1A <- forestplot(table1) +
+Fig2A <- forestplot(table1) +
   annotate('text',x=0.5,y=4,label='A')
@@
 
-ddFig1B <- ddply(temp,
+ddFig2B <- ddply(temp,
                  .(Insect),
@@
                  TLKBse = 1.96*se(TLKB))
-ddFig1B <- ddFig1B[complete.cases(ddFig1B),]
-ddFig1B$Insect <- as.numeric(paste(ddFig1B$Insect))
+ddFig2B <- droplevels(ddFig2B[complete.cases(ddFig2B),])
+ddFig2B$Insect <- as.numeric(paste(ddFig2B$Insect))
 
-Fig1B <- ggplot(ddFig1B,
+Fig2B <- ggplot(ddFig2B,
        aes(x = Insect,
@@
         axis.title.x = element_text(size = 16)) +
-  stat_smooth(data=FlSAall,aes(x=Insect,y=TLKB),method=lm,col='black',se=F) +
+  stat_smooth(data=FlSAall,aes(x=Insect,y=TLKB),method=lm,col='black',se=T) +
   geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse)) +
   scale_colour_manual(values = wes_palette("Royal1")) +
-  annotate('text',x=8,y=12,label='B')
+  annotate('text',x=2,y=10,label='B')
 
-Fig1C <- ggplot(FlSAall,
-                aes(x = Condition,
-                    y = TLKB)) +
-    geom_point() +
-  ylab('Telomere length (kb)') + 
-  xlab('Body Condition') + 
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 14), 
-        axis.text.y = element_text(size = 14), 
-        axis.title.y = element_text(size = 16,vjust=0.8),
-        axis.title.x = element_text(size = 16)) +
-  stat_smooth(method=lm,col='black',se=F) +
-  scale_colour_manual(values = wes_palette("Royal1")) +
-  annotate('text',x=-2.5,y=15,label='C')
-
-multiplot(Fig1A,Fig1B,Fig1C,layout=matrix(c(1,1,1,1,1,1,2,3,2,3), nrow=5, byrow=TRUE))
+multiplot(Fig2A,Fig2B)
 
@@
 
-```{r Figure 2}
+```{r Figure 3}
 
-x <- summary(lm(TLKB~Insect,data= FlSAall))$resid
-xf <- ifelse(x>median(x),'Long','Short')
+s1 <- with(FlSAall,flexsurvreg(Surv(Lifespan+1,Died)~TLKB+Insect, dist="lnorm"))
+t1 <- data.frame(s1$res)
+t1 <- t1[3:5,]
+t1 <- t1[order(rownames(t1)),]
+rownames(t1) <- c('Food av.','Tel. length')
 
-s1 <- survfit(Surv(time = Lifespan+1, event = Died)~xf, data = FlSAall)
-Fig2 <- ggsurv(s1,plot.cens=F) + 
+Fig2A <- ggplot(t1,aes(x = rownames(t1),y = est)) +
+  geom_point() + 
+  geom_pointrange(aes(ymin = L95.,ymax = U95.)) +
     theme_classic() +
-  ylab('Proportion Survived') + 
+  ylab('Estimate') + 
+  xlab('') + 
+  theme_classic() +
+  theme(axis.text.x = element_text(angle=90,size = 14), 
+        axis.text.y = element_text(size = 14), 
+        axis.title.y = element_text(size = 16,vjust=0.8),
+        axis.title.x = element_text(size = 16),
+        legend.position = c(0.8,0.8),
+        legend.title = element_text(size = 14),
+        legend.text = element_text(size = 12)) +
+  geom_abline(intercept = 0,slope = 0,lty=2) +
+  annotate('text',x=0.6,y=0.15,label='A')
+
+s1 <- survfit(Surv(time = Lifespan+1, event = Died)~TLF, data = FlSA)
+Fig2B <- ggsurv(s1,plot.cens=F) + 
+  theme_classic() +
+  ylab('Proportion survived') + 
   xlab('Age (years)') + 
@@
         legend.position = c(0.8,0.8),
-        legend.title = element_text(size = 14),
-        legend.text = element_text(size = 12)) +
-  scale_colour_manual(name = 'Residual TL',
+        legend.text = element_text(size = 12),
+        legend.title = element_blank()) +
+  scale_colour_manual(name = 'TL',
                       values = c('darkred','red')) +
   guides(linetype=F) +
-  theme(legend.title = element_text('Telomere length'))
+  annotate('text',x=0.5,y=1.1,label='B')
 
-Fig2
+multiplot(Fig2A,Fig2B)
 

### 2015-08-14 Manuscript/SW_early_life_telomeres_v9.Rmd
Added new data, included chicks, updated figures

@@
     highlight: null
-bibliography: mybib.bib
+bibliography: MyBib.bib
 csl: ProceedingsB.csl
@@
 In this study we analyse patterns of early-life telomere length and subsequent survival to test the prediction that telomeres fill a missing link between early-life environmental variation and late-life fitness.
-Specifically we predict that: i) individuals raised in good quality conditions will have longer telomeres, and ii) the relationship between telomere length and early-life conditions influences short and long-term survival.
+Specifically we predict that: i) individuals raised in good quality conditions will have longer telomeres, and ii) having longer telomeres in early life is associated with greater short and long-term survival.
 
@@
 
-We then used parametric survival analysis to examine variation in telomere length in relation to survival, implemented using the 'flexsurv' package in R.
-The survival time distribution was chosen by comparing the fit (using visual inspection of plots and a comparison of AIC values) of a range of distributions (exponential, Weibull, log-normal, gamma, gompertz and generalized F) against the observed survival data.
+We then used parametric survival analysis to examine variation in juvenile telomere length in relation to survival, implemented using the 'flexsurv' package in R.
+The survival time distribution was chosen by comparing the fit of a range of distributions (exponential, Weibull, log-normal, gamma, gompertz and generalized F) against the observed survival data, by visually inspecting plots and by comparing AIC values.
 As explanatory variables of survival we included juvenile telomere length, plus the factors identified as being related to telomere length in the previous analysis.
-Lay year was included as a random effect.
+With this approach we aimed to assess the effect of juvenile telomere length on survival while controlling for potentially confounding variables that may be correlated with telomere length.
 
@@
 Early-life telomere length differed substantially among birth years (Figure S2), with longer telomeres in years where insect abundance was high (Figure 1B).
-Model averaging confirmed this strong, positive effect of yearly food availability on telomere length (Figure 1A).
-The best supported model with early-life telomere length as the response variable contained annual food availability and body condition as explanatory variables (R^2^ = `r round2(summary(lm(TLKB~Insect+GroupSize+Condition,data=FlSAall))$r.squared)`).
+Model averaging confirmed this strong, positive effect of yearly food availability on telomere length (Figure 1A; Table S1).
+The best supported model with early-life telomere length as the response variable contained annual food availability and body condition as explanatory variables (Table S1; R^2^ = `r round2(summary(lm(TLKB~Insect+GroupSize+Condition,data=FlSAall))$r.squared)`).
 Early-life telomere length  was positively related to body condition (Figure 1C).
 
-To better visualise cohort effects we plotted the relationship between telomere length and both local density and body condition among cohort years (Figure S3).
-We found that the relationship between local density and telomere length varied among cohorts, from being negative to positive (Figure S3A).
-A similar, albeit less pronounced, pattern was observed for within-cohort relationships between telomere length and body condition (Figure S3B).
+A log-normal survival model best fitted the Seychelles warbler data (AIC = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="lnorm")$AIC)`, AIC of next best distribution = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="weibull")$AIC)`).
+A survival model including juvenile telomere length, body condition and early-life food availability showed that only juvenile telomere length had an effect on survival (estimate = `r round2(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB+Insect), dist="lnorm")$res[3,1])`, CI = `r round2(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB+Insect), dist="lnorm")$res[3,2])`-`r round2(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB+Insect), dist="lnorm")$res[3,3])`; Fig.
+2A).
+This survival effect was most pronounced in individuals between five and ten years of age (Fig.
+2B).
+Because telomere length and annual food availability are highly correlated (*R* = `r round2(cor(FlSAall$Insect,FlSAall$TLKB)) #confidence test`), we ran an analysis of early-life telomere length on survival excluding early-life food availability, and the telomere effect beacme marginally non-significant (estimate = `r round2(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,1])`, CI = `r round2(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,2])`-`r round2(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,3])`).
 
-A log-normal survival model best fitted the Seychelles warbler survival data (AIC = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="lnorm")$AIC)`, AIC of next best distribution = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="weibull")$AIC)`).
-A survival model including telomere length, body condition and annual food availability showed that only telomere length had an on survival (Fig.
-2A).
-Because of the strong correlation between telomere length and
+FlSAall$LayYear <- factor(FlSAall$LayYear)
+summary(glmer(Lifespan~TLKB+Insect+Condition+(1|LayYear),data=FlSAall,family=poisson))
 

### 2015-08-14 Scripts/SW juvenile telomeres IPC clean.R

@@ -3,33 +3,32 @@
 ## CLEAN THE DATA
 #################################################################################*
 
+#Average repeats of blood samples
 av <- ave(dd$TL,dd$BloodID)
 dd$TL <- av
 
 dd <- unique(dd)
 
+
+
+
 # Telomere data -----------------------------------------------------------
-
-
-#Add in some useful new TL variables
 
 dd$LogTL <- log(dd$TL)
 dd$TLKB <- dd$TL/1000
+mymed <- median(dd$TLKB,na.rm=T)
+dd$TLF <- ifelse(dd$TLKB > mymed,'Long telomeres','Short telomeres')
 
 
 # Weird variable names ----------------------------------------------------
 
-
-#Change names of weird database variables
 colnames(dd)[colnames(dd) == 'OccasionDate'] <- 'CatchDate'
-colnames(dd)[colnames(dd) == 'LastSighting'] <- 'DeathDate'
-
+colnames(dd)[colnames(dd) == 'MaxOfMaxOfSeenDate'] <- 'DeathDate'
+colnames(dd)[colnames(dd) == 'MinOfFieldPeriodID'] <- 'FieldPeriodID'
 
 
 # Catch Year, Catch month and death year ----------------------------------
 
-
-#Extract catch year, death year using substring
 dd$CatchYear <- as.numeric(substr(dd$CatchDate,7,10))
 dd$DeathYear <- as.numeric(substr(dd$DeathDate,7,10))
 
@@ -49,36 +48,39 @@ dd$Ageclass <- factor(dd$Ageclass,levels = c('CH','FL','FL','FL','SA','A'))
 
 dd$Age[dd$Ageclass != 'A'] <- 0
 
+dd$Agemonths <- ifelse(dd$Ageclass == 'CH',1,
+                       ifelse(dd$Ageclass == 'FL',6,
+                              ifelse(dd$Ageclass == 'SA',9,dd$Age*12)))
+
+#Delta age
+for(i in 1:nrow(dd))
+{
+  currentdata = subset(dd,BirdID == dd$BirdID[i])
+  dd$MeanAge[i] <- mean(currentdata$Age)
+  dd$DeltaAge[i] <- dd$Age[i]-dd$MeanAge[i]
+}
+
+
+
 # Survival and lifespan ---------------------------------------------------
 
-
-#Add some survival variables
[…]

### 2015-08-14 Scripts/SW juvenile telomeres IPC functions.r

@@ -126,6 +126,13 @@ se <- function(x) sd(x,na.rm=T)/sqrt(length(x))
 
 
 
+# Find telomere function --------------------------------------------------
+
+findTL <- function(x,TLfile = AvgTL)
+{
+  bird <- paste(x)
+  return(TLfile[names(TLfile) == bird])
+}
 
 
 # Knitr options -----------------------------------------------------------

### 2015-08-14 Scripts/SW juvenile telomeres IPC load.r

@@ -24,7 +24,7 @@ library(survival)
 library(flexsurv)
 
 # Load data ---------------------------------------------------------------
-dd <- read.csv('Data/SW TL main data for analysis.csv') #main dataset
+dd <- read.csv('Data/SW TL main data for analysis 2.csv') #main dataset
 terr <- read.csv('Data/SW TL all territories.csv')
 status <- read.csv('Data/SW TL all breedstatus.csv')
 insects <- read.csv('Data/Insects.csv')

### 2015-09-01 Manuscript/SW_early_life_telomeres_SOM_v9.Rmd
Added longitudinal data. General updates.

@@
 
-**Table S1** Model selection table for analysis of factors affecting early life telomere length in juvenile Seychelles warblers.
+**Table S1** Model selection table for analysis of factors affecting early life telomere length in Seychelles warbler chicks.
 Only models that remained in the top model set for model averaging (AICc $\leq$ 6 compared to the top model) are included here.
@@
 #Run the model
-lm1 <- lmer(TLKB ~ Agemonths + TQspace + Insect + GroupSize + Sex + Condition + Helper + (1|LayYear),
+lm1 <- lmer(TLKB ~ TQspace + Insect + NonHelper + Sex + BodyMass + Tarsus + Helper + (1|LayYear),
+            data = chickall,REML=FALSE)
+
+lm1 <- standardize(lm1)
+
+#Model averaging
+lm2 <- dredge(lm1)
+t1 <- as.matrix(subset(lm2, delta<=6))
+colnames(t1)[2:8] <- c('Sex', 'Body Mass','Helpers','Food av.','Non helpers','Tarsus length','TQ')
+
+tableS1 <- data.frame(Model = apply(t1,1,getterms,c(2:8)),t1[,c('df','AICc','delta','weight')])
+colnames(tableS1) <- c('Model','df','AICc', 'Delta AICc','Weight')
+rownames(tableS1) <- NULL
+tableS1$Model[tableS1$Model == ''] <- 'Null model'
+kable(tableS1,digits=3)
+
+```
+
+**Table S2** Model selection table for analysis of factors affecting early life telomere length in Seychelles warbler fledglings and subadults.
+Only models that remained in the top model set for model averaging (AICc $\leq$ 6 compared to the top model) are included here.
+i = intercept.
+
+```{r Table s2}
+
+#Run the model
+lm1 <- lmer(TLKB ~ TQspace + Insect + NonHelper + Sex + BodyMass + Tarsus + Helper + (1|LayYear),
             data = FlSAall,REML=FALSE)
@@
 t1 <- as.matrix(subset(lm2, delta<=6))
-colnames(t1)[2:7] <- c('Sex', 'Body condition','Group size','Helper','Food av.','TQ')
-getterms <- function(x)
-  {
-  temp <- c('i',names(which(!(is.na(x[2:7])))))
-  return(paste(temp,collapse = " + "))
-  }
+colnames(t1)[2:8] <- c('Sex', 'Body Mass','Helpers','Food av.','Non helpers','Tarsus length','TQ')
 
-tableS1 <- data.frame(Model = apply(t1,1,getterms),t1[,c('df','AICc','delta','weight')])
-colnames(tableS1) <- c('Model','df','AICc', 'Delta AICc','Weight')
-rownames(tableS1) <- NULL
-tableS1$Model[tableS1$Model == ''] <- 'Null model'
-kable(tableS1,digits=3)
+tables2 <- data.frame(Model = apply(t1,1,getterms,c(2:8)),t1[,c('df','AICc','delta','weight')])
+colnames(tables2) <- c('Model','df','AICc', 'Delta AICc','Weight')
+rownames(tables2) <- NULL
+tables2$Model[tables2$Model == ''] <- 'Null model'
+kable(tables2,digits=3)
 
@@
 
+temp <- Loss
+levels(temp$Ageclass) <- c('Chicks', 'Fledglings','Subadults')
+
+ddFigS2 <- ddply(temp,
+                 .(Ageclass),
+                 summarise,
+                 TROCmean = mean(TROC),
+                 TROCse = se(TROC))
+
+FigS2 <- ggplot(ddFigS2,
+       aes(x = Ageclass,
+           y = TROCmean)) +
+  geom_pointrange(aes(min=TROCmean-TROCse,ymax=TROCmean+TROCse)) +
+  ylab('Corrected telomere loss') + 
+  xlab('') + 
+  theme_classic() +
+  ylim(0,0.002) +
+  theme(axis.text.x = element_text(size = 14), 
+        axis.text.y = element_text(size = 14), 
+        axis.title.y = element_text(size = 16,vjust=0.8),
+        axis.title.x = element_text(size = 16))
+
+FigS2
+
+```
+
+**Figure S2** Telomere loss (mean $\pm$ s.e.) in juvenile Seychelles warblers.
+Telomere loss a standardised and corrected value (see methods for details) based on rates of lost between the sampling event of birds as a juvnile, and the first available sampling as an adult.
+Values above zero indicate telomere shortening in early life.
+
+```{r Figure S3,fig.width=4,fig.height=4}
+
 temp <- FlSA
 
-ddFigS2 <- ddply(temp,
+ddFigS3 <- ddply(temp,
                  .(LayYear),
@@
 
-ddFigS2 <- ddFigS2[complete.cases(ddFigS2),]
-ddFigS2$LayYear <- as.numeric(paste(ddFigS2$LayYear))
+ddFigS3 <- ddFigS3[complete.cases(ddFigS3),]
+ddFigS3$LayYear <- as.numeric(paste(ddFigS3$LayYear))
 
-FigS2 <- ggplot(ddFigS2,
+FigS3 <- ggplot(ddFigS3,
        aes(x = LayYear,
@@
 
-FigS2
+FigS3
 
@@
 
-**Figure S2** Cross-sectional juvenile telomere length (mean $\pm$ s.e.) across birth years in Seychelles warblers
+**Figure S3** Cross-sectional juvenile telomere length (mean $\pm$ s.e.) across birth years in Seychelles warblers

### 2015-09-01 Manuscript/SW_early_life_telomeres_tables_and_figures v9.Rmd
Added longitudinal data. General updates.

--- 2015-06-22 Manuscript/SW_early_life_telomeres_tables_and_figures v9.Rmd
@@
                fig.path='../Figures/',
-               fig.width= 5,
-               fig.height= 5,
                fig.show='asis',
@@
 
-**Figure 1** Factors affecting telomere length in juvenile Seychelles warblers.
+**Figure 3** Telomere length and age in juvenile Seychelles warblers. Points and error bars represent mean and 95% confidence intervals, respectively.
+
+**Figure 2** Factors affecting telomere length in Seychelles warbler chicks.
 **A** Model averaged estimates and 95% confidence intervals (points and error bars, respectively) for all explanatory terms used in a linear model with juvenile telomere length as the response variable.
-Numbers in brackets are the relative importance of each term in the top model set (see main text for details; Food av = annual food availability, TQ = territory quality).
-**B** Annual food availability and **C** body condition in relation to telomere length.
-Points and error bars in **A** are mean $\pm$ 95% confidence interval telomere lengths for each year or group size (means are used for plotting purposes only), and the lines represent fitted values and from linear regressions.
+Numbers in brackets are the relative importance of each term in the top model set (see main text for details; Food av.
+= annual food availability, TQ = territory quality).
+**B** Presence/absence of helpers and **C** body mass in relation to raw telomere length.
+Points and error bars in **B** are mean $\pm$ 95% confidence interval telomere lengths for each group, and the line and shaded areas from **C** represent fitted values and 95% confidence limits from a linear regression.
 
-**Figure 2** Juvenile telomere length and lifespan in Seychelles warblers.
-**A** Model averaged estimates and 95% confidence intervals (points and error bars, respectively) for all explanatory terms used in a generalised linear model with a poisson error structure and lifespan as the response variable.
-Only juveniles that survived their first year of life were included in the model.
-Numbers in brackets are the relative importance of each term in the top model set (see main text for details).
-Food av = annual food availability, TL = telomere length TQ = territory quality.
-**B** Raw telomere length and annual food availability in relation to lifespan.
-Annual food availability was modeled as a covariate (see main text for details) but is plotted as a factor here for visualisation purposes.
-The lines represent fitted values from a Poisson regression.
+**Figure 3** Factors affecting telomere length in juvenile Seychelles warblers.
+**A** Model averaged estimates and 95% confidence intervals (points and error bars, respectively) for all explanatory terms used in a linear model with juvenile telomere length as the response variable.
+Numbers in brackets are the relative importance of each term in the top model set (see main text for details; Food av.
+= annual food availability, TQ = territory quality).
+**B** Raw telomere length in relation to annual food availability Points and error bars in **B** are mean $\pm$ 95% confidence interval telomere lengths for each year (means are used for plotting purposes only), and the line and shaded areas represent fitted values and 95% confidence limits from a linear regression.
+
+**Figure 4** Juvenile telomere length and survival in chick (**A,B**) and juvenile (**C,D**) Seychelles warblers.
+**A,C** Model averaged estimates and 95% confidence intervals (points and error bars, respectively) for all explanatory terms used in a parametric survival model (see main text for details).
+Food av = annual food availability, Tel.
+length = telomere length.
+**B,D** Kaplan-Meier curves showing the relationship between telomere length and survival.
+Telomere length was modelled as a covariate, but is binned into groups here for visualisation purposes (long and short = greater than or less than median telomere length, repectively).
 
@@
 
-```{r Figure 1,fig.width=8,fig.height=9}
+```{r Figure 1,fig.width=5,fig.height=4}
+
+temp <- dd
+
+ddFig1 <- ddply(temp,
+                 .(AgemonthF),
+                 summarize,
+                 TLKBmean = mean(TLKB),
+                 TLKBse = se(TLKB)*1.96)
+
+Fig1 <- ggplot(ddFig1,
+       aes(x = AgemonthF,
+           y = TLKBmean)) +
+  ylab('Telomere length (kb)') + 
+  xlab('Age (months)') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 14), 
+        axis.text.y = element_text(size = 14), 
+        axis.title.y = element_text(size = 16,vjust=0.8),
+        axis.title.x = element_text(size = 16)) +
+  geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse))
+
+Fig1
+
+```
+
+PAGEBREAK
+
+**Figure 2**
+
+```{r Figure 2,fig.width=8,fig.height=9}
 
 ###########
-#Fig 1A
+#Fig 2A
+
+chickall <- subset(chickall,BodyMass>5)
+#Run the model
+lm2 <- lmer(TLKB ~ TQspace + Insect + BodyMass+ Tarsus + NonHelper + Sex + Helper + (1|LayYear),
+            data = chickall,REML=FALSE)
+
+lm2 <- standardize(lm2)
+
+#Model averaging
+lm3 <- dredge(lm2)
+lmavg <- model.avg(lm3,subset = delta <=6)
+myimp <- importance(lmavg) #Relative importance of each explanatory variable
+
+#Put it all into a nice table
+table2 <- imptable(lmavg,myimp)
+table2 <- signif(table2,2)
+
+#Sort out row/col names
+nams <- rownames(table2)
+nams[nams == 'c.Ageclass'] <- 'Age class'
+nams[nams == 'z.BodyMass'] <- 'Body mass'
+nams[nams == 'z.Tarsus'] <- 'Tarsus'
+nams[nams == 'z.NonHelper'] <- 'Non helpers'
+nams[nams == 'c.Sex'] <- 'Sex'
+nams[nams == 'z.Insect'] <- 'Food av.'
+nams[nams == 'z.TQspace'] <- 'TQ'
+nams[nams == 'z.Helper'] <- 'Helpers'
+
+rownames(table2) <- nams
+colnames(table2) <- c('Estimate','SE','2.5%','97.5%','RI')
+
+table2 <- data.frame(table2,myx=row.names(table2))
+
+Fig2A <- forestplot(table2) +
+  annotate('text',x=0.75,y=4,label='A')
+
+######################################
+temp <- chickall
+temp$Helper <- ifelse(temp$Helper>0,'Present','Absent')
+
+ddFig2B <- ddply(temp,
+                 .(Helper),
+                 summarize,
+                 TLKBmean = mean(TLKB),
+                 TLKBse = 1.96*se(TLKB))
+
+Fig2B <- ggplot(ddFig2B,
+       aes(x = Helper,
+           y = TLKBmean)) +
+  ylab('Telomere length (kb)') + 
+  xlab('Helpers') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 14), 
+        axis.text.y = element_text(size = 14), 
+        axis.title.y = element_text(size = 16,vjust=0.8),
+        axis.title.x = element_text(size = 16)) +
+  geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse)) +
+  scale_colour_manual(values = wes_palette("Royal1")) +
+  annotate('text',x=0.65,y=9,label='B')
+
+Fig2C <- ggplot(chickall,
+       aes(x = BodyMass,
+           y = TLKB)) +
+  ylab('Telomere length (kb)') + 
+  xlab('Body mass (g)') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 14), 
+        axis.text.y = element_text(size = 14), 
+        axis.title.y = element_text(size = 16,vjust=0.8),
+        axis.title.x = element_text(size = 16)) +
+  geom_point() +
+  stat_smooth(method = 'lm',col = 'black') +
+  annotate('text',x=10.5,y=15,label='C')
+
+multiplot(Fig2A,Fig2B,Fig2C,layout=matrix(c(1,1,1,1,1,1,2,3,2,3), nrow=5, byrow=TRUE))
+
+```
+
+PAGEBREAK
+
+**Figure 3**
+
+```{r Figure 3,fig.width=5,fig.height=9}
+###########
+#Fig 3A
 
 #Run the model
-lm1 <- lmer(TLKB~ TQspace + Insect + GroupSize + Sex + Condition + Helper + (1|LayYear),
+lm3 <- lmer(TLKB ~ TQspace + Insect + NonHelper + Sex + Tarsus + BodyMass + Helper + (1|LayYear),
             data = FlSAall,REML=FALSE)
 
-lm1 <- standardize(lm1)
+lm3 <- standardize(lm3)
 
 #Model averaging
-lm2 <- dredge(lm1)
-lmavg <- model.avg(lm2,subset = delta <=6)
+lm4 <- dredge(lm3)
+lmavg <- model.avg(lm4,subset = delta <=6)
 myimp <- importance(lmavg) #Relative importance of each explanatory variable
@@
 
-Fig1A <- forestplot(table1) +
-  annotate('text',x=0.5,y=4,label='A')
+Fig3A <- forestplot(table1) +
+  annotate('text',x=0.75,y=2,label='A')
 
@@
 
-ddFig1B <- ddply(temp,
+ddFig3B <- ddply(temp,
                  .(Insect),
@@
                  TLKBse = 1.96*se(TLKB))
-ddFig1B <- ddFig1B[complete.cases(ddFig1B),]
-ddFig1B$Insect <- as.numeric(paste(ddFig1B$Insect))
+ddFig3B <- droplevels(ddFig3B[complete.cases(ddFig3B),])
+ddFig3B$Insect <- as.numeric(paste(ddFig3B$Insect))
 
-Fig1B <- ggplot(ddFig1B,
+Fig3B <- ggplot(ddFig3B,
        aes(x = Insect,
@@
         axis.title.x = element_text(size = 16)) +
-  stat_smooth(data=FlSAall,aes(x=Insect,y=TLKB),method=lm,col='black',se=F) +
+  stat_smooth(data=FlSAall,aes(x=Insect,y=TLKB),method=lm,col='black',se=T) +
   geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse)) +
   scale_colour_manual(values = wes_palette("Royal1")) +
-  annotate('text',x=8,y=12,label='B')
+  annotate('text',x=2,y=10,label='B')
 
-Fig1C <- ggplot(FlSAall,
-                aes(x = Condition,
-                    y = TLKB)) +
-    geom_point() +
-  ylab('Telomere length (kb)') + 
-  xlab('Body Condition') + 
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 14), 
-        axis.text.y = element_text(size = 14), 
-        axis.title.y = element_text(size = 16,vjust=0.8),
-        axis.title.x = element_text(size = 16)) +
-  stat_smooth(method=lm,col='black',se=F) +
-  scale_colour_manual(values = wes_palette("Royal1")) +
-  annotate('text',x=-2.5,y=15,label='C')
-
-multiplot(Fig1A,Fig1B,Fig1C,layout=matrix(c(1,1,1,1,1,1,2,3,2,3), nrow=5, byrow=TRUE))
+multiplot(Fig3A,Fig3B)
 
@@
 
-**Figure 2**
+**Figure 4**
 
-```{r Figure 2}
+```{r Figure 4,fig.height=8,fig.width=8}
+s1 <- with(subset(dd,Ageclass == 'FL'),flexsurvreg(Surv(Lifespan+1,Died)~TLKB, dist="lnorm"))
 
-x <- summary(lm(TLKB~Insect,data= FlSAall))$resid
-xf <- ifelse(x>median(x),'Long','Short')
+s1 <- with(chicks,flexsurvreg(Surv(Lifespan+1,Died)~TLKB+BodyMass+TLKB+Helper, dist="lnorm"))
+t1 <- data.frame(s1$res)
+t1 <- t1[3:5,]
+t1 <- t1[order(rownames(t1)),]
+rownames(t1) <- c('Body mass','Helper','Tel. length')
 
-s1 <- survfit(Surv(time = Lifespan+1, event = Died)~xf, data = FlSAall)
-Fig2 <- ggsurv(s1,plot.cens=F) + 
+Fig4A <- ggplot(t1,aes(x = rownames(t1),y = est)) +
+  geom_point() + 
+  geom_pointrange(aes(ymin = L95.,ymax = U95.)) +
     theme_classic() +
-  ylab('Proportion Survived') + 
+  ylab('Estimate') + 
+  xlab('') + 
+  theme_classic() +
+  theme(axis.text.x = element_text(angle=90,size = 14), 
+        axis.text.y = element_text(size = 14), 
+        axis.title.y = element_text(size = 16,vjust=0.8),
+        axis.title.x = element_text(size = 16),
+        legend.position = c(0.8,0.8),
+        legend.title = element_text(size = 14),
+        legend.text = element_text(size = 12)) +
+  geom_abline(intercept = 0,slope = 0,lty=2) +
+  annotate('text',x=0.6,y=0.4,label='A')
+
+s2 <- survfit(Surv(time = Lifespan+1, event = Died)~TLF, data = chicks)
+Fig4B <- ggsurv(s2,plot.cens=F) + 
+  theme_classic() +
+  ylab('Proportion survived') + 
   xlab('Age (years)') + 
@@
         legend.position = c(0.8,0.8),
+        legend.text = element_text(size = 12),
+        legend.title = element_blank()) +
+  scale_colour_manual(name = 'TL',
+                      values = c('darkred','red')) +
+  guides(linetype=F) +
+  annotate('text',x=0.5,y=1.1,label='B')
+
+###########################################
+
+s2 <- with(FlSAall,flexsurvreg(Surv(Lifespan+1,Died)~TLKB+Insect, dist="lnorm"))
+t4 <- data.frame(s2$res)
+t4 <- t4[3:4,]
+t4 <- t4[order(rownames(t4)),]
+rownames(t4) <- c('Food av.','Tel. length')
+
+Fig4C <- ggplot(t4,aes(x = rownames(t4),y = est)) +
+  geom_point() + 
+  geom_pointrange(aes(ymin = L95.,ymax = U95.)) +
+    theme_classic() +
+  ylab('Estimate') + 
+  xlab('') + 
+  theme_classic() +
+  theme(axis.text.x = element_text(angle=90,size = 14), 
+        axis.text.y = element_text(size = 14), 
+        axis.title.y = element_text(size = 16,vjust=0.8),
+        axis.title.x = element_text(size = 16),
+        legend.position = c(0.8,0.8),
         legend.title = element_text(size = 14),
         legend.text = element_text(size = 12)) +
-  scale_colour_manual(name = 'Residual TL',
+  geom_abline(intercept = 0,slope = 0,lty=2) +
+  annotate('text',x=0.6,y=0.15,label='C')
+
+s4 <- survfit(Surv(time = Lifespan+1, event = Died)~TLF, data = FlSA)
+Fig4D <- ggsurv(s4,plot.cens=F) + 
+  theme_classic() +
+  ylab('Proportion survived') + 
+  xlab('Age (years)') + 
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 14), 
+        axis.text.y = element_text(size = 14), 
+        axis.title.y = element_text(size = 16,vjust=0.8),
+        axis.title.x = element_text(size = 16),
+        legend.position = c(0.8,0.8),
+        legend.text = element_text(size = 12),
+        legend.title = element_blank()) +
+  scale_colour_manual(name = 'TL',
                       values = c('darkred','red')) +
   guides(linetype=F) +
-  theme(legend.title = element_text('Telomere length'))
+  annotate('text',x=0.5,y=1.1,label='D')
 
-Fig2
+multiplot(Fig4A,Fig4B,Fig4C,Fig4D,layout=matrix(c(1:4), nrow=2, byrow=FALSE))
 

### 2015-09-01 Manuscript/SW_early_life_telomeres_v9.Rmd
Added longitudinal data. General updates.

@@
 bibliography: MyBib.bib
-csl: ProceedingsB.csl
+csl: JAE.csl
 ---
@@
 
-Lewis G. Spurgin^1^,  Kathryn L. Bebbington^1^, Eleanor A. Fairfield^1^, Jan Komdeur^2^, Terry Burke^3^, Hannah, L. Dugdale^3^,  and David S. Richardson^1,4^.
+Lewis G. Spurgin^1,2^,  Kat Bebbington^1^, Eleanor A. Fairfield^1^, Jan Komdeur^3^, Terry Burke^4^, Hannah, L. Dugdale^2^,  and David S. Richardson^1,5^.
 
 1. School of Biological Sciences, University of East Anglia, Norwich Research Park, NR4 7TJ, United Kingdom
-2. Behavioural Ecology and Self-organization Group, Centre for Ecological and Evolutionary Studies, University of Groningen, Groningen, The Netherlands
-3. Department of Animal and Plant Sciences, NERC Biomolecular Analysis Facility, University of Sheffield, Sheffield, UK
-4. Nature Seychelles, Roche Caiman, Mahé, Republic of Seychelles
+2. Department of Zoology, Edward Grey Institute, University of Oxford, Oxford, UK
+3. Behavioural Ecology and Self-organization Group, Centre for Ecological and Evolutionary Studies, University of Groningen, Groningen, The Netherlands
+4. Department of Animal and Plant Sciences, NERC Biomolecular Analysis Facility, University of Sheffield, Sheffield, UK
+5. Nature Seychelles, Roche Caiman, Mahé, Republic of Seychelles
 
@@
 
-Telomere lengths were measured from `r nrow(FlSA)` samples taken cross-sectionally from birds aged between two and twelve months, between `r min(as.numeric(paste(juv$LayYear)))` and `r max(as.numeric(paste(juv$LayYear)))`.
-Of these birds, `r nrow(subset(FlSA,Ageclass=='FL'))` were aged between two and six months (juveniles), and `r nrow(subset(FlSA,Ageclass=='SA'))` were aged seven to twelve months (subadults).
-We chose to focus on juveniles and subadults rather than nestlings, because i) we have juvenile samples spanning a range of years, whereas nestling samples are predominantly from two early years, and ii) telomere length during the nestling phase will largely be a product of inheritance and growth, whereas in juveniles and subadults we expect telomere length to incorporate the impact of early-life experiences.
+Telomere lengths were measured from `r nrow(FlSA)+nrow(chicks)` samples taken cross-sectionally from birds aged less than twelve months, between `r min(as.numeric(paste(juv$LayYear)))` and `r max(as.numeric(paste(juv$LayYear)))`.
+Of these birds, `r nrow(chicks)` were nestlings less than two months old, `r nrow(subset(FlSA,Ageclass=='FL'))` were aged between two and six months (juveniles), and `r nrow(subset(FlSA,Ageclass=='SA'))` were aged seven to twelve months (subadults).
 We measured absolute telomere quantity per diploid genome using a quantitative PCR (qPCR) assay  (full details in [@Barrett2012]).
@@
 We performed all analyses using R version 3.0.1 [@RDevelopmentCoreTeam2011].
-We first tested whether early-life environmental variation influences telomere length using linear mixed models, with telomere length as the response variable.
-As explanatory variables we included body condition (calculated as deviation from a regression line of mass on tarsus length; [@VandeCrommenacker2011a]), sex, age class (juvenile or subadult), territory quality, yearly food availability and local density (measured as the number of individuals in the juvenile's natal social group [@Brouwer2012]).
+We tested whether telomere length varied within the first year of life using one-way ANOVA, with telomere length as the response variable and age class (chick, juvenile or subadult) as a fixed factor.
+As telomere length varied substantially among age classes (see results) we carried out all subsequent analyses seperately for each age class.
+
+We tested tested whether early-life environmental variation influences telomere length using linear mixed models, with telomere length as the response variable.
+As explanatory variables we included body condition (calculated as deviation from a regression line of mass on tarsus length; [@VandeCrommenacker2011a]), sex, territory quality, yearly food availability, and the number of helping and non-helping subordinates present in the natal territory  ([@Brouwer2012]).
 Lay Year was included as a random factor, in order to account for any annual variation in telomere length not explained by yearly food availability.
@@
 
+For a subset of juveniles (n = `r nrow(Loss)` individuals) we had longitudinal data (i.e samples from individuals caught both as a juvenile an adult).
+For these individuals we calculated a rate of telomere loss between the juvenile sample and earliest adult sample by subtracting the adult telomere length from the juvenile telomere length and dividing this difference by the length of time (in days) between sampling events.
+To account for regression to the mean effects, we applied a correction based on correlations among samples within individuals following Verhulst *et al.* [-@Verhulst2013].
+
 ###Results
@@
 This is only marginally higher than mean telomere length for all adult birds (`r round2(mean(subset(dd,Age>0)$TLKB))` $\pm$ `r round2(se(subset(dd,Age>0)$TLKB))` kb), and lower than the mean for one year-old birds based on cross-sectional data (`r round2(mean(subset(dd,Age==1)$TLKB))` $\pm$ `r round2(se(subset(dd,Age==1)$TLKB))` kb; Figure S1).
-We found no significant difference in telomere length between age classes (juvenile vs subadult), or between sexes (Figure 1A).
+However, when we considered ageclasses seperately, we found substantial age-related variation in telomere length within the first year of life (F = `r round2(anova(lm(TLKB~Ageclass,data = juv))[1,4])`; *P* = `r round2(anova(lm(TLKB~Ageclass,data = juv))[1,5])`; Fig.
+1).
+This cross-sectional data revealed that chicks have the longest telomeres, but that subsadults and adults had longer telomeres than fledglings (Fig.
+1).
+However, a longitudinal analysis showed that chicks, fleglings nd subadults all lost telomeres early in life, with no differences in rates of telomere shortening among age classes (F = `r round2(anova(lm(TROC~Ageclass,data = Loss))[1,4])`; *P* = `r round2(anova(lm(TROC~Ageclass,data = Loss))[1,5])`; Fig.
+S2).
+This suggests that the apparent increase in telomere length between juveniles and subadults in the corss-secitonal data was a result of the sampling, rather than telomere elongation.
 
-Early-life telomere length differed substantially among birth years (Figure S2), with longer telomeres in years where insect abundance was high (Figure 1B).
-Model averaging confirmed this strong, positive effect of yearly food availability on telomere length (Figure 1A; Table S1).
-The best supported model with early-life telomere length as the response variable contained annual food availability and body condition as explanatory variables (Table S1; R^2^ = `r round2(summary(lm(TLKB~Insect+GroupSize+Condition,data=FlSAall))$r.squared)`).
-Early-life telomere length  was positively related to body condition (Figure 1C).
+In chicks, telomere length did not vary significantly among birth years (Fig. 2A, Fig. S2A).
+
+Juvenile telomere length differed substantially among birth years (Figure S2), with longer telomeres in years where insect abundance was high (Figure 2B).
+Model averaging confirmed this strong, positive effect of yearly food availability on telomere length (Figure 2A; Table S1).
+The best supported model with juvenile telomere length as the response variable contained annual food availability as the only explanatory variable (Table S1; R^2^ = `r round2(summary(lm(TLKB~Insect,data=FlSAall))$r.squared)`).
 
@@
 Because telomere length and annual food availability are highly correlated (*R* = `r round2(cor(FlSAall$Insect,FlSAall$TLKB)) #confidence test`), we ran an analysis of early-life telomere length on survival excluding early-life food availability, and the telomere effect beacme marginally non-significant (estimate = `r round2(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,1])`, CI = `r round2(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,2])`-`r round2(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,3])`).
-
-FlSAall$LayYear <- factor(FlSAall$LayYear)
-summary(glmer(Lifespan~TLKB+Insect+Condition+(1|LayYear),data=FlSAall,family=poisson))
 

### 2015-09-01 Scripts/SW juvenile telomeres IPC clean.R

@@ -3,10 +3,10 @@
 ## CLEAN THE DATA
 #################################################################################*
 
-#Average repeats of blood samples
-av <- ave(dd$TL,dd$BloodID)
-dd$TL <- av
 
+#Average repeats of blood samples
+av <- ave(dd$TL,c(dd$BloodID,dd$Status))
+dd$TL <- av
 dd <- unique(dd)
 
 
@@ -27,11 +27,12 @@ colnames(dd)[colnames(dd) == 'MaxOfMaxOfSeenDate'] <- 'DeathDate'
 colnames(dd)[colnames(dd) == 'MinOfFieldPeriodID'] <- 'FieldPeriodID'
 
 
-# Catch Year, Catch month and death year ----------------------------------
+# Catch Year, Catch date and death year ----------------------------------
+
 
 dd$CatchYear <- as.numeric(substr(dd$CatchDate,7,10))
 dd$DeathYear <- as.numeric(substr(dd$DeathDate,7,10))
-
+dd$CatchDate <- as.Date(dd$CatchDate,"%d/%m/%Y")
 
 
 
@@ -39,18 +40,24 @@ dd$DeathYear <- as.numeric(substr(dd$DeathDate,7,10))
 # Age data ----------------------------------------------------------------
 
 dd$Age <- dd$CatchYear-dd$LayYear
-dd$Age[dd$Ageclass]
+
 
 #Sort out and order ageclass levels
 dd <- droplevels(dd[dd$Ageclass != '',])
 levels(dd$Ageclass) <- c('A','CH','FL','FL','FL','SA')
 dd$Ageclass <- factor(dd$Ageclass,levels = c('CH','FL','FL','FL','SA','A'))
 
-dd$Age[dd$Ageclass != 'A'] <- 0
+
 
 dd$Agemonths <- ifelse(dd$Ageclass == 'CH',1,
-                       ifelse(dd$Ageclass == 'FL',6,
+                       ifelse(dd$Ageclass == 'FL',3,
                               ifelse(dd$Ageclass == 'SA',9,dd$Age*12)))
+dd$AgemonthF <- ifelse(dd$Ageclass == 'CH','<1',
+                       ifelse(dd$Ageclass == 'FL','1-9',
+                              ifelse(dd$Ageclass == 'SA','9-12','>12')))
[…]

### 2015-09-01 Scripts/SW juvenile telomeres IPC functions.r

@@ -52,6 +52,23 @@ baselineimp <- function(mymodel,n.permutations = 1000)
 
 
 
+
+
+
+#Get terms
+getterms <- function(x,cols)
+{
+  temp <- c('i',names(which(!(is.na(x[cols])))))
+  return(paste(temp,collapse = " + "))
+}
+
+
+
+
+
+
+
+
 #Correct relative importance according to baseline importance
 correctedimp <- function(importance,baselineimportance)
 {
@@ -278,7 +295,7 @@ forestplot <- function(mytable)
           axis.title.y = element_text(size = 16,vjust=0.8),
           axis.title.x = element_text(size = 16)) +
     geom_abline(slope=0,intercept=0,lty=2) +
-    annotate('text',x=mytable$myx,y=min(mytable$X2.5)-0.2,label=paste('(',mytable$RI,')',sep=''))
+    annotate('text',x=mytable$myx,y=min(mytable$X2.5)*1.1,label=paste('(',mytable$RI,')',sep=''))
     
 }
 

### 2015-09-14 Manuscript/SW_early_life_telomeres_SOM_v9.Rmd
Updated figures and manuscript - standardised all response variables for within year analyses. Added markdown docx template (myref.docx)

@@
 
-```{r load and clean, include = F}
+```{r setup, include = F}
 
-#Set wd one level up
-setwd('..')
+rm(list=ls())
 
-#Load and clean data
-
-source('Scripts/SW juvenile telomeres IPC load.r')
-source('Scripts/SW juvenile telomeres IPC functions.r')
-source('Scripts/SW juvenile telomeres IPC clean.r')
+#Load knitr
+library(knitr)
 
@@
 options(na.action='na.fail')
-opts_knit$set(root.dir = getwd())
+opts_knit$set(root.dir = '..')
 opts_chunk$set(echo=FALSE,
@@
                fig.show='asis',
-               dev='png')
+               dev=c('png','pdf'))
+
+```
+
+```{r load and clean, include = F}
+
+#Load and clean data
+source('Scripts/SW juvenile telomeres IPC load.r')
+source('Scripts/SW juvenile telomeres IPC functions.r')
+source('Scripts/SW juvenile telomeres IPC clean.r')
 
@@
 ```
+
+PAGEBREAK
 
@@
   theme_classic() +
-  ylim(0,0.002) +
   theme(axis.text.x = element_text(size = 14), 
@@
 
-**Figure S2** Telomere loss (mean $\pm$ s.e.) in juvenile Seychelles warblers.
+**Figure S2** Longitudinal telomere loss (mean $\pm$ s.e.) in juvenile Seychelles warblers.
 Telomere loss a standardised and corrected value (see methods for details) based on rates of lost between the sampling event of birds as a juvnile, and the first available sampling as an adult.
@@
 
-```{r Figure S3,fig.width=4,fig.height=4}
+PAGEBREAK
 
-temp <- FlSA
+```{r Figure S3,fig.width=5,fig.height=9}
 
-ddFigS3 <- ddply(temp,
-                 .(LayYear),
-                 summarize,
-                 TLKBmean = mean(TLKB),
-                 TLKBse = se(TLKB))
+###########
+#Fig S3A
 
-ddFigS3 <- ddFigS3[complete.cases(ddFigS3),]
-ddFigS3$LayYear <- as.numeric(paste(ddFigS3$LayYear))
+temp <- subset(FlSAall,Ageclass == 'FL')
 
-FigS3 <- ggplot(ddFigS3,
-       aes(x = LayYear,
-           y = TLKBmean)) +
-  geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse)) +
-  ylab('Telomere length (kb)') + 
-  xlab('Lay Year') + 
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 14), 
-        axis.text.y = element_text(size = 14), 
-        axis.title.y = element_text(size = 16,vjust=0.8),
-        axis.title.x = element_text(size = 16))
+#Run the model
+lm3 <- lmer(TLKB ~ TQspace + Insect + Sex + Tarsus + BodyMass + (1|LayYear),
+            data = temp,REML=FALSE)
 
-FigS3
+lm3 <- standardize(lm3)
+
+#Model averaging
+lm4 <- dredge(lm3)
+lmavg <- model.avg(lm4,subset = delta <=6)
+myimp <- importance(lmavg) #Relative importance of each explanatory variable
+
+#Put it all into a nice table
+table1 <- imptable(lmavg,myimp)
+table1 <- signif(table1,2)
+
+#Sort out row/col names
+nams <- rownames(table1)
+nams[nams == 'c.Ageclass'] <- 'Age class'
+nams[nams == 'z.BodyMass'] <- 'Body mass'
+nams[nams == 'z.Tarsus'] <- 'Tarsus'
+nams[nams == 'z.NonHelper'] <- 'Non helpers'
+nams[nams == 'c.Sex'] <- 'Sex'
+nams[nams == 'z.Insect'] <- 'Food av.'
+nams[nams == 'z.TQspace'] <- 'TQ'
+nams[nams == 'z.Helper'] <- 'Helpers'
+
+rownames(table1) <- nams
+colnames(table1) <- c('Estimate','SE','2.5%','97.5%','RI')
+
+table1 <- data.frame(table1,myx=row.names(table1))
+
+FigS3A <- forestplot(table1) +
+  annotate('text',x=0.75,y=3,label='A')
+
+###########
+#Fig S3B
+
+temp <- subset(FlSAall,Ageclass == 'SA')
+
+#Run the model
+lm3 <- lmer(TLKB ~ TQspace + Insect + Sex + Tarsus + BodyMass + (1|LayYear),
+            data = temp,REML=FALSE)
+
+#lm3 <- standardize(lm3)
+
+#Model averaging
+lm4 <- dredge(lm3)
+lmavg <- model.avg(lm4,subset = delta <=6)
+myimp <- importance(lmavg) #Relative importance of each explanatory variable
+
+#Put it all into a nice table
+table1 <- imptable(lmavg,myimp)
+table1 <- signif(table1,2)
+
+#Sort out row/col names
+nams <- rownames(table1)
+nams[nams == 'c.Ageclass'] <- 'Age class'
+nams[nams == 'z.BodyMass'] <- 'Body mass'
+nams[nams == 'z.Tarsus'] <- 'Tarsus'
+nams[nams == 'z.NonHelper'] <- 'Non helpers'
+nams[nams == 'c.Sex'] <- 'Sex'
+nams[nams == 'z.Insect'] <- 'Food av.'
+nams[nams == 'z.TQspace'] <- 'TQ'
+nams[nams == 'z.Helper'] <- 'Helpers'
+
+rownames(table1) <- nams
+colnames(table1) <- c('Estimate','SE','2.5%','97.5%','RI')
+
+table1 <- data.frame(table1,myx=row.names(table1))
+
+FigS3B <- forestplot(table1) +
+  annotate('text',x=0.75,y=3,label='B')
+
+multiplot(FigS3A,FigS3B)
 
@@
 
-**Figure S3** Cross-sectional juvenile telomere length (mean $\pm$ s.e.) across birth years in Seychelles warblers
+**Figure S3** Factors affecting telomere length in Seychelles warbler fledglings **(A)** and subadults **(B)**.
+Points and error bars represent model averaged estimates and 95% confidence intervals for all explanatory terms used in a linear model with telomere length as the response variable.
+Numbers in brackets are the relative importance of each term in the top model set (see main text for details; Food av.
+= annual food availability, TQ = territory quality).

### 2015-09-14 Manuscript/SW_early_life_telomeres_tables_and_figures v9.Rmd
Updated figures and manuscript - standardised all response variables for within year analyses. Added markdown docx template (myref.docx)

--- 2015-06-22 Manuscript/SW_early_life_telomeres_tables_and_figures v9.Rmd
@@
 ---
-output: word_document
+output:
+  word_document:
+    reference_docx: myref.docx
 ---
-```{r setwd, include = F}
+```{r setup, include = F}
 
-#Set working directory up one level
-setwd('..')
+rm(list=ls())
 
-#Load and clean data
-source('Scripts/SW juvenile telomeres IPC load.r')
-source('Scripts/SW juvenile telomeres IPC functions.r')
-source('Scripts/SW juvenile telomeres IPC clean.r')
+#Load knitr
+library(knitr)
 
@@
 
-**Figure 1** Factors affecting telomere length in juvenile Seychelles warblers.
-**A** Model averaged estimates and 95% confidence intervals (points and error bars, respectively) for all explanatory terms used in a linear model with juvenile telomere length as the response variable.
-Numbers in brackets are the relative importance of each term in the top model set (see main text for details; Food av = annual food availability, TQ = territory quality).
-**B** Annual food availability and **C** body condition in relation to telomere length.
-Points and error bars in **A** are mean $\pm$ 95% confidence interval telomere lengths for each year or group size (means are used for plotting purposes only), and the lines represent fitted values and from linear regressions.
+**Figure 1** Telomere length and age Seychelles warblers, based on cross-sectional **(A)** and longitudinal **(B)** data.
+Points and error bars represent mean and 95% confidence intervals, respectively.
+In **B**, telomere loss is a standardised and corrected rate (see methods for details) based on differences in telomere length between the sampling event of birds as a juvnile, and the first available sampling as an adult.
+Values above zero indicate telomere shortening in early life.
 
-**Figure 2** Juvenile telomere length and lifespan in Seychelles warblers.
-**A** Model averaged estimates and 95% confidence intervals (points and error bars, respectively) for all explanatory terms used in a generalised linear model with a poisson error structure and lifespan as the response variable.
-Only juveniles that survived their first year of life were included in the model.
-Numbers in brackets are the relative importance of each term in the top model set (see main text for details).
-Food av = annual food availability, TL = telomere length TQ = territory quality.
-**B** Raw telomere length and annual food availability in relation to lifespan.
-Annual food availability was modeled as a covariate (see main text for details) but is plotted as a factor here for visualisation purposes.
-The lines represent fitted values from a Poisson regression.
+**Figure 2** Temporal variation in early-life telomere length in the Seychelles warbler based on cross-sectional **(A,B)** and longitudinal **(C,D)** data (see Figure 1 and main text for details).
+Points and error bars represent mean and standard error telomere length for each summer (blue) or winter (red) field season.
+In **A** and **C** points are ordered temporally, while in **B** and **D** telomere length/loss for each season is plotted against island-wide food availability.
+Lines and shaded areas represent fitted values and 95% confidence limits from a linear regression.
+
+**Figure 3** Factors affecting telomere length in Seychelles warbler chicks.
+**A** Model averaged estimates and 95% confidence intervals for all explanatory terms used in a linear model with telomere length as the response variable.
+Numbers in brackets are the relative importance of each term in the top model set (see main text for details; Food av.
+= annual food availability, TQ = territory quality).
+**B** Presence/absence of helpers, **C** sex and **D** tarsus length in relation to raw telomere length.
+Points and error bars in **B and C** are mean $\pm$ standard error telomere lengths for each group, and the line and shaded areas from **D** represent fitted values and 95% confidence limits from a linear regression.
+
+**Figure 4** Factors affecting telomere length in Seychelles warblers fledglings **(A)** and subadults **(B)**.
+Model averaged estimates and 95% confidence intervals for all explanatory terms used in a linear model with  telomere length as the response variable.
+Numbers in brackets are the relative importance of each term in the top model set (see main text for details; Food av.
+= annual food availability, TQ = territory quality).
+
+**Figure 5** Kaplan-Meier curves showing the relationship between telomere length and survival in chick (**A**) and juvenile (**B**) Seychelles warblers.
+Telomere length is binned into groups here for visualisation purposes only (long and short = greater than or less than median telomere length, repectively).
 
@@
 ```{r Figure 1,fig.width=8,fig.height=9}
+######################
+#1A
+######################
 
-###########
-#Fig 1A
+temp <- dd
+levels(temp$Ageclass) <- c('Chicks',rep('Fledglings',3),'Sub-adults','Adults')
 
-#Run the model
-lm1 <- lmer(TLKB~ TQspace + Insect + GroupSize + Sex + Condition + Helper + (1|LayYear),
-            data = FlSAall,REML=FALSE)
-
-lm1 <- standardize(lm1)
-
-#Model averaging
-lm2 <- dredge(lm1)
-lmavg <- model.avg(lm2,subset = delta <=6)
-myimp <- importance(lmavg) #Relative importance of each explanatory variable
-
-#Put it all into a nice table
-table1 <- imptable(lmavg,myimp)
-table1 <- signif(table1,2)
-
-#Sort out row/col names
-nams <- rownames(table1)
-nams[nams == 'c.Ageclass'] <- 'Age class'
-nams[nams == 'z.Condition'] <- 'Body condition'
-nams[nams == 'z.GroupSize'] <- 'Local density'
-nams[nams == 'c.Sex'] <- 'Sex'
-nams[nams == 'z.Insect'] <- 'Food av.'
-nams[nams == 'z.TQspace'] <- 'TQ'
-nams[nams == 'z.Helper'] <- 'Helpers'
-
-rownames(table1) <- nams
-colnames(table1) <- c('Estimate','SE','2.5%','97.5%','RI')
-
-table1 <- data.frame(table1,myx=row.names(table1))
-
-Fig1A <- forestplot(table1) +
-  annotate('text',x=0.5,y=4,label='A')
-
-######################################
-temp <- FlSAall
-temp$Insect <- factor(temp$Insect)
-
-ddFig1B <- ddply(temp,
-                 .(Insect),
+ddFig1A <- ddply(temp,
+                 .(Ageclass),
                  summarize,
                  TLKBmean = mean(TLKB),
-                 TLKBse = 1.96*se(TLKB))
-ddFig1B <- ddFig1B[complete.cases(ddFig1B),]
-ddFig1B$Insect <- as.numeric(paste(ddFig1B$Insect))
+                 TLKBse = se(TLKB))
 
-Fig1B <- ggplot(ddFig1B,
-       aes(x = Insect,
+Fig1A <- ggplot(ddFig1A,
+       aes(x = Ageclass,
            y = TLKBmean)) +
-  ylab('Telomere length (kb)') + 
-  xlab('Annual food availability') + 
+  ylab('Corrected telomere loss') + 
+  xlab('') +
   theme_classic() +
@@
         axis.title.y = element_text(size = 16,vjust=0.8),
-        axis.title.x = element_text(size = 16)) +
-  stat_smooth(data=FlSAall,aes(x=Insect,y=TLKB),method=lm,col='black',se=F) +
+        axis.title.x = element_text(size = 16),
+        legend.position = c(0.1,0.9),
+        legend.title = element_blank(),
+        legend.text = element_text(size = 15)) +
   geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse)) +
-  scale_colour_manual(values = wes_palette("Royal1")) +
-  annotate('text',x=8,y=12,label='B')
+  annotate('text',label = 'A',x=0.5,y=6.2)
 
-Fig1C <- ggplot(FlSAall,
-                aes(x = Condition,
-                    y = TLKB)) +
-    geom_point() +
+######################
+#1B
+######################
+
+temp <- Loss
+levels(temp$Ageclass) <- c('Chicks','Fledglings','Sub-adults')
+
+ddFig1B <- ddply(temp,
+                 .(Ageclass),
+                 summarize,
+                 TROCmean = mean(TROC),
+                 TROCse = se(TROC))
+
+Fig1B <- ggplot(ddFig1B,
+       aes(x = Ageclass,
+           y = TROCmean)) +
   ylab('Telomere length (kb)') + 
-  xlab('Body Condition') + 
+  xlab('') +
   theme_classic() +
@@
         axis.title.y = element_text(size = 16,vjust=0.8),
-        axis.title.x = element_text(size = 16)) +
-  stat_smooth(method=lm,col='black',se=F) +
-  scale_colour_manual(values = wes_palette("Royal1")) +
-  annotate('text',x=-2.5,y=15,label='C')
+        axis.title.x = element_text(size = 16),
+        legend.position = c(0.1,0.9),
+        legend.title = element_blank(),
+        legend.text = element_text(size = 15)) +
+  geom_pointrange(aes(ymin=TROCmean-TROCse,ymax=TROCmean+TROCse)) +
+  annotate('text',label = 'B',x=0.5,y=0.0018)
 
-multiplot(Fig1A,Fig1B,Fig1C,layout=matrix(c(1,1,1,1,1,1,2,3,2,3), nrow=5, byrow=TRUE))
+multiplot(Fig1A,Fig1B)
 
@@
 
-```{r Figure 2}
+```{r Figure 2,fig.width=9,fig.height=9}
 
-t1 <- data.frame(with(FlSAall,flexsurvreg(Surv(Lifespan+1,Died)~Condition+TLKB+Insect, dist="lnorm"))$res)
-t1 <- t1[3:5,]
-rownames(t1) <- c('Body condition', 'Tel. length', 'Food av.')
+ddFig2A <- ddply(juv,
+                 .(CatchYear,Season),
+                 summarize,
+                 TLKBmean = mean(TLKB),
+                 TLKBse = se(TLKB))
+ddFig2A <- ddFig2A[complete.cases(ddFig2A),]
+ddFig2A <- subset(ddFig2A,TLKBse < 4)
 
-Fig2A <- ggplot(t1,aes(x = rownames(t1),y = est)) +
-  geom_point() + 
-  geom_pointrange(aes(ymin = L95.,ymax = U95.)) +
+Fig2A <- ggplot(ddFig2A,
+       aes(x = CatchYear,
+           y = TLKBmean,
+           col = Season)) +
+  ylab('Telomere length') + 
+  xlab('Year') +
   theme_classic() +
-  ylab('Estimate') + 
-  xlab('') + 
-  theme_classic() +
-  theme(axis.text.x = element_text(angle=90,size = 14), 
+  theme(axis.text.x = element_text(size = 14), 
         axis.text.y = element_text(size = 14), 
@@
         axis.title.x = element_text(size = 16),
-        legend.position = c(0.8,0.8),
-        legend.title = element_text(size = 14),
-        legend.text = element_text(size = 12)) +
-  geom_abline(intercept = 0,slope = 0,lty=2)
+        legend.position = 'none') +
+  geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse),
+                  position = position_dodge(width = 1)) +
+  scale_color_manual(values = wes_palette("Royal1")) +
+  annotate('text',label = 'A',x=1995,y=10)
 
-s1 <- survfit(Surv(time = Lifespan+1, event = Died)~TLF, data = FlSA)
-Fig2B <- ggsurv(s1,plot.cens=F) + 
+####################
+
+temp <- juv
+temp$Insect <- factor(temp$Insect)
+
+ddFig2B <- ddply(temp,
+                 .(Insect,Season),
+                 summarize,
+                 TLKBmean = mean(TLKB),
+                 TLKBse = se(TLKB))
+ddFig2B$Insect <- as.numeric(paste(ddFig2B$Insect))
+ddFig2B <- ddFig2B[complete.cases(ddFig2B),]
+
+Fig2B <- ggplot(ddFig2B,
+       aes(x = Insect,
+           y = TLKBmean,
+           col = Season)) +
+  ylab('Telomere length') + 
+  xlab('Food availability') +
   theme_classic() +
-  ylab('Proportion Survived') + 
+  theme(axis.text.x = element_text(size = 14), 
+        axis.text.y = element_text(size = 14), 
+        axis.title.y = element_text(size = 16,vjust=0.8),
+        axis.title.x = element_text(size = 16),
+        legend.position = 'none') +
+  geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse)) +
+  stat_smooth(method = 'lm',se=T,col = 'darkgrey') +
+  scale_color_manual(values = wes_palette("Royal1")) +
+  annotate('text',label = 'B',x=-1.5,y=8)
+
+################################
+
+ddFig2C <- ddply(Loss,
+                 .(CatchYear,Season),
+                 summarize,
+                 TROCmean = mean(TROC),
+                 TROCse = se(TROC))
+ddFig2C <- ddFig2C[complete.cases(ddFig2C),]
+ddFig2C <- subset(ddFig2C,TROCse < 4)
+
+Fig2C <- ggplot(ddFig2C,
+       aes(x = CatchYear,
+           y = TROCmean,
+           col = Season)) +
+  ylab('Corrected telomere loss') + 
+  xlab('Year') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 14), 
+        axis.text.y = element_text(size = 14), 
+        axis.title.y = element_text(size = 16,vjust=0.8),
+        axis.title.x = element_text(size = 16),
+        legend.position = 'none') +
+  geom_pointrange(aes(ymin=TROCmean-TROCse,ymax=TROCmean+TROCse),
+                  position = position_dodge(width = 1)) +
+  scale_color_manual(values = wes_palette("Royal1")) +
+  annotate('text',label = 'C',x=1995,y=0.009)
+
+####################################
+temp <- Loss
+temp$Insect <- factor(temp$Insect)
+
+ddFig2D <- ddply(temp,
+                 .(Insect,Season),
+                 summarise,
+                 TROCmean = mean(TROC),
+                 TROCse = se(TROC))
+ddFig2D$Insect <- as.numeric(paste(ddFig2D$Insect))
+ddFig2D <- ddFig2D[complete.cases(ddFig2D),]
+
+Fig2D <- ggplot(ddFig2D,
+       aes(x = Insect,
+           y = TROCmean,
+           col = Season)) +
+  geom_pointrange(aes(min=TROCmean-TROCse,ymax=TROCmean+TROCse)) +
+  ylab('Corrected telomere loss') + 
+  xlab('Food availability') + 
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 14), 
+        axis.text.y = element_text(size = 14), 
+        axis.title.y = element_text(size = 16,vjust=0.8),
+        axis.title.x = element_text(size = 16),
+        legend.position = 'none') +
+  stat_smooth(method = 'lm', se=T, col = 'darkgrey') +
+  scale_color_manual(values = wes_palette("Royal1")) +
+  annotate('text',label = 'D',x=-1.5,y=0.0075)
+
+multiplot(Fig2A,Fig2B,Fig2C,Fig2D,layout=matrix(1:4,2,2))
+
+```
+
+PAGEBREAK
+
+**Figure 3**
+
+```{r Figure 3,fig.width=8,fig.height=9}
+
+######################
+#3A
+######################
+
+#Run the model
+lm2 <- lm(cenTL ~ cenHelper + cenNonHelper + Sex + cenTarsus  + cenTQ,
+            data = chickall)
+
+#Model averaging
+
+lm3 <- dredge(lm2)
+lmavg <- model.avg(lm3,subset = delta <=6)
+myimp <- importance(lmavg) #Relative importance of each explanatory variable
+
+#Put it all into a nice table
+table2 <- imptable(lmavg,myimp)
+table2 <- signif(table2,2)
+
+#Sort out row/col names
+nams <- rownames(table2)
+nams[nams == 'cenTarsus'] <- 'Tarsus'
+nams[nams == 'cenNonHelper'] <- 'Non helpers'
+nams[nams == 'SexMales'] <- 'Sex'
+nams[nams == 'cenTQ'] <- 'TQ'
+nams[nams == 'cenHelper'] <- 'Helpers'
+
+rownames(table2) <- nams
+colnames(table2) <- c('Estimate','SE','2.5%','97.5%','RI')
+
+table2 <- data.frame(table2,myx=row.names(table2))
+
+Fig3A <- forestplot(table2) +
+  annotate('text',label = 'A',x=0.6,y=1.2)
+
+######################
+#3B
+######################
+
+temp <- chicks
+
+ddFig3B <- ddply(temp,
+                 .(Sex),
+                 summarize,
+                 TLKBmean = mean(TLKB),
+                 TLKBse = se(TLKB))
+
+Fig3B <- ggplot(ddFig3B,
+       aes(x = Sex,
+           y = TLKBmean)) +
+  ylab('Telomere length (kb)') + 
+  xlab('') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 14), 
+        axis.text.y = element_text(size = 14), 
+        axis.title.y = element_text(size = 16,vjust=0.8),
+        axis.title.x = element_text(size = 16),
+        legend.position = c(0.1,0.9),
+        legend.title = element_blank(),
+        legend.text = element_text(size = 15)) +
+  geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse)) +
+  annotate('text',label = 'B',x=0.5,y=7)
+
+######################
+#3C
+######################
+
+temp <- chicks
+temp$HelperF <- ifelse(chicks$Helper>0,'Helpers present','Helpers absent')
+
+ddFig3C <- ddply(temp,
+                 .(HelperF),
+                 summarize,
+                 TLKBmean = mean(TLKB),
+                 TLKBse = se(TLKB))
+
+Fig3C <- ggplot(ddFig3C,
+       aes(x = HelperF,
+           y = TLKBmean)) +
+  ylab('Telomere length (kb)') + 
+  xlab('') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 14), 
+        axis.text.y = element_text(size = 14), 
+        axis.title.y = element_text(size = 16,vjust=0.8),
+        axis.title.x = element_text(size = 16),
+        legend.position = c(0.1,0.9),
+        legend.title = element_blank(),
+        legend.text = element_text(size = 15)) +
+  geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse)) +
+  annotate('text',label = 'C',x=0.5,y=8)
+
+######################
+#3D
+######################
+
+Fig3D <- ggplot(chicks,
+                aes(x = Tarsus,
+                    y = TLKB)) +
+  ylab('Telomere length (kb)') + 
+  xlab('Tarsus length (mm)') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 14), 
+        axis.text.y = element_text(size = 14), 
+        axis.title.y = element_text(size = 16,vjust=0.8),
+        axis.title.x = element_text(size = 16)) +
+  geom_point(size=3) +
+  stat_smooth(method = 'lm',col = 'black') +
+  scale_color_manual(values = wes_palette("Royal1")) +
+  annotate('text',label = 'D',x=19,y=15)
+
+multiplot(Fig3A,Fig3B,Fig3C,Fig3D,layout=matrix(c(1:4), nrow=2, byrow=FALSE))
+
+```
+
+PAGEBREAK
+
+**Figure 4**
+
+```{r Figure 4,fig.width=5,fig.height=9}
+
+######################
+#3A
+######################
+Fl <- subset(FlSAall,Ageclass == 'FL')
+
+#Run the model
+lm2 <- lm(cenTL ~ cenHelper + cenNonHelper + Sex + cenTarsus  + cenTQ,
+            data = Fl)
+
+#Model averaging
+lm3 <- dredge(lm2)
+lmavg <- model.avg(lm3,subset = delta <=6)
+myimp <- importance(lmavg) #Relative importance of each explanatory variable
+
+#Put it all into a nice table
+table2 <- imptable(lmavg,myimp)
+table2 <- signif(table2,2)
+
+nams <- rownames(table2)
+nams[nams == 'cenTarsus'] <- 'Tarsus'
+nams[nams == 'cenNonHelper'] <- 'Non helpers'
+nams[nams == 'SexMales'] <- 'Sex'
+nams[nams == 'cenTQ'] <- 'TQ'
+nams[nams == 'cenHelper'] <- 'Helpers'
+
+rownames(table2) <- nams
+colnames(table2) <- c('Estimate','SE','2.5%','97.5%','RI')
+
+table2 <- data.frame(table2,myx=row.names(table2))
+
+Fig4A <- forestplot(table2) +
+  annotate('text',label = 'A',x=0.6,y=0.8)
+
+######################
+#4B
+######################
+SA <- subset(FlSAall,Ageclass == 'SA')
+
+#Run the model
+lm2 <- lm(cenTL ~ cenHelper + cenNonHelper + Sex + cenTarsus  + cenTQ,
+            data = SA)
+
+#Model averaging
+lm4 <- dredge(lm2)
+lmavg <- model.avg(lm4,subset = delta <=6)
+myimp <- importance(lmavg) #Relative importance of each explanatory variable
+
+#Put it all into a nice table
+table2 <- imptable(lmavg,myimp)
+table2 <- signif(table2,2)
+
+nams <- rownames(table2)
+nams[nams == 'cenTarsus'] <- 'Tarsus'
+nams[nams == 'cenNonHelper'] <- 'Non helpers'
+nams[nams == 'SexMales'] <- 'Sex'
+nams[nams == 'cenTQ'] <- 'TQ'
+nams[nams == 'cenHelper'] <- 'Helpers'
+
+rownames(table2) <- nams
+colnames(table2) <- c('Estimate','SE','2.5%','97.5%','RI')
+
+table2 <- data.frame(table2,myx=row.names(table2))
+
+Fig4B <- forestplot(table2) +
+  annotate('text',label = 'B',x=0.6,y=0.6)
+
+multiplot(Fig4A,Fig4B)
+
+```
+
+PAGEBREAK
+
+**Figure 5**
+
+```{r Figure 5,fig.height=8,fig.width=5}
+
+s1 <- survfit(Surv(time = Lifespan+1, event = Died)~TLF, data = chicks)
+Fig5A <- ggsurv(s1,plot.cens=F) + 
+  theme_classic() +
+  ylab('Proportion survived') + 
   xlab('Age (years)') + 
@@
         legend.position = c(0.8,0.8),
-        legend.title = element_text(size = 14),
-        legend.text = element_text(size = 12)) +
+        legend.text = element_text(size = 12),
+        legend.title = element_blank()) +
   scale_colour_manual(name = 'TL',
@@
   guides(linetype=F) +
-  theme(legend.title = element_blank())
+  annotate('text',x=0.5,y=1.1,label='A')
 
-multiplot(Fig2A,Fig2B)
+###########################################
+
+s2 <- survfit(Surv(time = Lifespan+1, event = Died)~TLF, data = FlSAall)
+Fig5B <- ggsurv(s2,plot.cens=F) + 
+  theme_classic() +
+  ylab('Proportion survived') + 
+  xlab('Age (years)') + 
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 14), 
+        axis.text.y = element_text(size = 14), 
+        axis.title.y = element_text(size = 16,vjust=0.8),
+        axis.title.x = element_text(size = 16),
+        legend.position = c(0.8,0.8),
+        legend.text = element_text(size = 12),
+        legend.title = element_blank()) +
+  scale_colour_manual(name = 'TL',
+                      values = c('darkred','red')) +
+  guides(linetype=F) +
+  annotate('text',x=0.5,y=1.1,label='B')
+
+multiplot(Fig5A,Fig5B)
 

### 2015-09-14 Manuscript/SW_early_life_telomeres_v9.Rmd
Updated figures and manuscript - standardised all response variables for within year analyses. Added markdown docx template (myref.docx)

@@
 ---
+```{r setup, include = F}
+
+rm(list=ls())
+
+#Load knitr
+library(knitr)
+
+#Set some options
+options(na.action='na.fail')
+opts_knit$set(root.dir = '..')
+opts_chunk$set(echo=FALSE,
+               warning=FALSE,
+               message=FALSE,
+               fig.path='../Figures/',
+               fig.width= 4,
+               fig.height= 8,
+               fig.show='asis',
+               dev=c('png','pdf'))
+
+```
+
 ```{r load and clean, include = F}
 
-#Set wd one level up
-setwd('..')
-
 #Load and clean data
-
 source('Scripts/SW juvenile telomeres IPC load.r')
@@
 [@Frenck1998; @Haussmann2003]), and that the extent of this early-life telomere shortening is influenced by the environment [@Boonekamp2014].
-Moreover, a recent laboratory study showed that juvenile telomere dynamics are related to late-life survival [@Heidinger2012].
+Moreover, a recent laboratory study showed that early-life telomere dynamics are related to late-life survival [@Heidinger2012].
 However, whether telomeres provide a link between early-life conditions and late-life survival in the wild is not known.
@@
 
-In Seychelles warblers, oxidative stress is linked to natal territory quality [@VandeCrommenacker2011a], but neither natal territory quality nor natal local density are linked to juvenile or adult survival [@Brouwer2006; @Hammers2013].
+In Seychelles warblers, oxidative stress is linked to natal territory quality [@VandeCrommenacker2011a], but neither natal territory quality nor natal local density are linked to early-life or adult survival [@Brouwer2006; @Hammers2013].
 In this study we analyse patterns of early-life telomere length and subsequent survival to test the prediction that telomeres fill a missing link between early-life environmental variation and late-life fitness.
@@
 #####Study species and sampling
-The Seychelles warbler is a small (~15 g) passerine bird with a mean life expectancy of 5.5 years at fledging [@Hammers2013].
-The population of *ca.* 320 adults on Cousin Island (04'20'S, 55'40'E) has been intensively studied since 1986 [@Richardson2003b; @Barrett2013a].
+The Seychelles warbler is a small (~15 g), insectivorous passerine bird with a mean life expectancy of 5.5 years at fledging [@Hammers2013].
+The population on Cousin Island (04'20'S, 55'40'E) has been intensively studied since 1986 [@Richardson2003b; @Barrett2013a].
 Individuals are usually ringed in their first year of life, and so are of known age.
-They are then followed throughout their lives, and as they are non-migratory endemics and are naturally confined to the island [@Komdeur2004a], a biannual census of birds on the island gives an accurate measure of levels of local density and individual survival rate [@Barrett2013a].
+They are then followed throughout their lives, and as they are non-migratory endemics and are naturally confined to the island [@Komdeur2004a], a biannual census of birds on Cousin during the main (June-August) and minor (December-February) breeding seasons gives an accurate measure of levels of local density and individual survival rate [@Barrett2013a].
 Full details of catching and census methods can be found in Brouwer *et al.* [@Brouwer2012].
 
-All territories are mapped biannually based on detailed observational data of foraging and territorial defence behaviour, and surveyed for territory quality [@Richardson2003b].
+All territories are mapped during the main and minor breeding seasons based on detailed observational data of foraging and territorial defence behaviour, and surveyed for territory quality [@Richardson2003b].
 Territory quality is calculated based on territory size, foliage cover and insect abundance [@Komdeur1992] and averaged within territories [@Hammers2013].
-Cousin is also subject to inter-annual variation in rainfall and food availability, and such yearly effects may override the effects of absolute territory quality.
-As an estimate of annual variation in food availability, we calculated an index of the number of insects across the entire island during the main breeding season of each sampling year based on the insect abundance data (see above).
+Cousin is also subject to intra- and inter-annual variation in rainfall and food availability, and such island-wide temporal variation may override the effects of absolute territory quality.
+As an estimate of variation in food availability, we calculated an index of the number of insects across the entire island during.
+This  index represents the average number of insects found per unit leaf area over all surveys carried out on the island in a breeding season.
 
@@
 Telomere lengths were measured from `r nrow(FlSA)+nrow(chicks)` samples taken cross-sectionally from birds aged less than twelve months, between `r min(as.numeric(paste(juv$LayYear)))` and `r max(as.numeric(paste(juv$LayYear)))`.
-Of these birds, `r nrow(chicks)` were nestlings less than two months old, `r nrow(subset(FlSA,Ageclass=='FL'))` were aged between two and six months (juveniles), and `r nrow(subset(FlSA,Ageclass=='SA'))` were aged seven to twelve months (subadults).
+Of these birds, `r nrow(chicks)` were nestlings less than two months old (chicks), `r nrow(subset(FlSA,Ageclass=='FL'))` were aged between two and six months (fledglings), and `r nrow(subset(FlSA,Ageclass=='SA'))` were aged seven to twelve months (subadults).
 We measured absolute telomere quantity per diploid genome using a quantitative PCR (qPCR) assay  (full details in [@Barrett2012]).
@@
 We performed all analyses using R version 3.0.1 [@RDevelopmentCoreTeam2011].
-We tested whether telomere length varied within the first year of life using one-way ANOVA, with telomere length as the response variable and age class (chick, juvenile or subadult) as a fixed factor.
-As telomere length varied substantially among age classes (see results) we carried out all subsequent analyses seperately for each age class.
+Our sampling regime covers a long time-period, and temporal variation in rainfall and food availability on Cousin may cause variation  in a range of social and evironmental variables [@Brouwer2012].
+This makes it difficult to disentangle how spatial and temporal social and environmental variation affects early-life telomere dynamics.
+we seperately analysed how variation in early life conditions among and within breeding seasons was related to variation in early life telomere length.
+First, we tested whether there was a temporal effect on telomere lengthby calculating average telomere length among all birds born in a given breeding season.
+We tested for a difference in amomg breeding seasons using a one-way ANOVA, and then used linear regression to test whether variation in mean telomere length among breeding seasons is explained by temporal variation in insect abundance.
 
-We tested tested whether early-life environmental variation influences telomere length using linear mixed models, with telomere length as the response variable.
-As explanatory variables we included body condition (calculated as deviation from a regression line of mass on tarsus length; [@VandeCrommenacker2011a]), sex, territory quality, yearly food availability, and the number of helping and non-helping subordinates present in the natal territory  ([@Brouwer2012]).
-Lay Year was included as a random factor, in order to account for any annual variation in telomere length not explained by yearly food availability.
-We created a full model containing all of the above terms, and a top model set was then defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011].
+We tested tested how spatial early-life environmental and social conditions influence telomere length within cohorts using linear mixed models.
+Telomere length was standarsised by mean-centering within cohorts (i.e.
+using all birds born in a given breeding season).
+As explanatory variables we included tarsus, length, sex, territory quality, yearly food availability, and the number of helping and non-helping subordinates present in the natal territory  ([@Brouwer2012]).
+
+We used model averaging We created a full model containing all of the above terms, and a top model set was then defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011].
 We report model averaged coefficients, confidence intervals, and 'relative importance', which reflects the relative weights of each predictor variable across the top model set.
@@
 
-We then used parametric survival analysis to examine variation in juvenile telomere length in relation to survival, implemented using the 'flexsurv' package in R.
+We used parametric survival analysis to examine variation in early-life telomere length in relation to survival, implemented using the 'flexsurv' package in R.
 The survival time distribution was chosen by comparing the fit of a range of distributions (exponential, Weibull, log-normal, gamma, gompertz and generalized F) against the observed survival data, by visually inspecting plots and by comparing AIC values.
-As explanatory variables of survival we included juvenile telomere length, plus the factors identified as being related to telomere length in the previous analysis.
-With this approach we aimed to assess the effect of juvenile telomere length on survival while controlling for potentially confounding variables that may be correlated with telomere length.
+As explanatory variables of survival we included early-life telomere length, plus the factors identified as being related to telomere length in the previous analysis.
+With this approach we aimed to assess the effect of early-life telomere length on survival while controlling for potentially confounding variables that may be correlated with telomere length.
 
-For a subset of juveniles (n = `r nrow(Loss)` individuals) we had longitudinal data (i.e samples from individuals caught both as a juvenile an adult).
-For these individuals we calculated a rate of telomere loss between the juvenile sample and earliest adult sample by subtracting the adult telomere length from the juvenile telomere length and dividing this difference by the length of time (in days) between sampling events.
+For a subset of first-year birds (n = `r nrow(Loss)` individuals) we had longitudinal data (i.e samples from individuals caught both in their first year and as an adult).
+For these individuals we calculated a rate of telomere loss between the first-year sample and earliest adult sample by subtracting the adult telomere length from the early-life telomere length and dividing this difference by the length of time (in days) between sampling events.
 To account for regression to the mean effects, we applied a correction based on correlations among samples within individuals following Verhulst *et al.* [-@Verhulst2013].
@@
 
-Mean ($\pm$ standard error) early-life telomere length was `r round2(mean(FlSAall$TLKB))` $\pm$ `r round2(se(FlSAall$TLKB))` kb.
-This is only marginally higher than mean telomere length for all adult birds (`r round2(mean(subset(dd,Age>0)$TLKB))` $\pm$ `r round2(se(subset(dd,Age>0)$TLKB))` kb), and lower than the mean for one year-old birds based on cross-sectional data (`r round2(mean(subset(dd,Age==1)$TLKB))` $\pm$ `r round2(se(subset(dd,Age==1)$TLKB))` kb; Figure S1).
-However, when we considered ageclasses seperately, we found substantial age-related variation in telomere length within the first year of life (F = `r round2(anova(lm(TLKB~Ageclass,data = juv))[1,4])`; *P* = `r round2(anova(lm(TLKB~Ageclass,data = juv))[1,5])`; Fig.
-1).
-This cross-sectional data revealed that chicks have the longest telomeres, but that subsadults and adults had longer telomeres than fledglings (Fig.
-1).
-However, a longitudinal analysis showed that chicks, fleglings nd subadults all lost telomeres early in life, with no differences in rates of telomere shortening among age classes (F = `r round2(anova(lm(TROC~Ageclass,data = Loss))[1,4])`; *P* = `r round2(anova(lm(TROC~Ageclass,data = Loss))[1,5])`; Fig.
-S2).
-This suggests that the apparent increase in telomere length between juveniles and subadults in the corss-secitonal data was a result of the sampling, rather than telomere elongation.
+#####Early-life telomere length varies with age and among breeding seasons
+Mean ($\pm$ standard error) telomere length in first-year birds was `r round2(mean(FlSAall$TLKB))` $\pm$ `r round2(se(FlSAall$TLKB))` kb.
+This is lower than the mean for one year-old birds based on cross-sectional data (`r round2(mean(subset(dd,Age==1)$TLKB))` $\pm$ `r round2(se(subset(dd,Age==1)$TLKB))` kb; Figure S1).
+However, when we considered age classes seperately, we found substantial age-related variation in telomere length within the first year of life (F = `r round2(anova(lm(TLKB~Ageclass,data = juv))[1,4])`; *P* = `r round2(anova(lm(TLKB~Ageclass,data = juv))[1,5])`).
+This cross-sectional data revealed that chicks have longer telomeres than any other age group, but that subsadults and adults had longer telomeres than fledglings (Fig.
+1A).
+A longitudinal analysis of telomere loss showed that chicks, fleglings and subadults all lost telomeres early in life, with highest and lowest rates of loss in chicks and subaldults, respectively, although the difference among age classes was not significant (F = `r round2(anova(lm(TROC~Ageclass,data = Loss))[1,4])`; *P* = `r round2(anova(lm(TROC~Ageclass,data = Loss))[1,5])`; Fig.
+1B).
+This suggests that the apparent increase in telomere length between fledglings and subadults in the cross-sectional data was a result of the sampling, rather than telomere elongation.
 
-In chicks, telomere length did not vary significantly among birth years (Fig. 2A, Fig. S2A).
+Early-life telomere length varied significantly among breeding seasons (one-way ANOVA, F = `r round2(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,4])`; *P* = `r round2(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,5])`; Fig.
+2A), and this variation in average telomere length over breeding seasons was positively related to island-wide food aviilability (linear regression, R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=juvseason))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data = juvseason))[1,4])`; *P* = `r round2(anova(lm(Insect~TLKBmean,data = juvseason))[1,5])`; Fig.
+2B).
+Similarly, the amount of telomeres lost in early life varied among breeding seasons (F = `r round2(anova(lm(TROC~factor(FieldPeriodID),data = Loss))[1,4])`; *P* = `r round2(anova(lm(TROC~factor(FieldPeriodID),data = Loss))[1,5])`; Fig.
+2C), and individuals born in seasons with low food availability lost telomeres, on average, at a faster rate than those born in years with high food availability (linear regression, R^2^ = `r round2(summary(lm(Insect~TROCmean,data=loss.season))$r.squared)`; F = `r round2(anova(lm(Insect~TROCmean,data = loss.season))[1,4])`; *P* = `r round2(anova(lm(Insect~TROCmean,data = loss.season))[1,5])`; Fig.
+2D).
 
-Juvenile telomere length differed substantially among birth years (Figure S2), with longer telomeres in years where insect abundance was high (Figure 2B).
-Model averaging confirmed this strong, positive effect of yearly food availability on telomere length (Figure 2A; Table S1).
-The best supported model with juvenile telomere length as the response variable contained annual food availability as the only explanatory variable (Table S1; R^2^ = `r round2(summary(lm(TLKB~Insect,data=FlSAall))$r.squared)`).
+#####Factors affecting early-life telomere dynamics within breeding seasons
+The top explaining within-season variation in chick telomere length contained tarsus length, sex and the number of helpers present in the natal territory (Fig.
+3A; Table S1).
+All three variables had high relative importance int he top model set (Fig.
+3A), and the model containing these variables explained a reasonable amount of variation in telomere length (R^2^ = `r round2(summary(lm(TLKB~Tarsus+Helper+Sex,data=chickall))$r.squared)`).
+Female chicks had longer telomeres than males (Fig.
+3B), while telomere length was positively related to the number of helpers (Fig.
+3C), and negatively related to tarus to length (Fig.
+3D).
+However, it should be noted that model-averaged confidence intervals for sex and the number of helpers (but not tarsus length) marginally overlapped zero.
+For both juveniles and sub-adults the null model (intercept only) was the top model (Table S1), and no explanatory terms had high relative importance in the top model set, with all estimates close to zero (Fig.
+4).
 
+For the subset of individuals with longitudinal data, we tested whether the factors previously identified as being related early-life telomere length (i.e.
+tarsus length, sex and number of helpers) were also associated with differences in rates of early life telomere loss.
+Due to a limited sample size, we considered all age classes together for this analysis.
+We found that none of these variables were associated with differences in telomere loss (all *P* > 0.05).
+
+#####Early life telomere dynamics and survival
 A log-normal survival model best fitted the Seychelles warbler data (AIC = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="lnorm")$AIC)`, AIC of next best distribution = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="weibull")$AIC)`).
@@
 DSR developed the idea and obtained funding for the overall telomere project; DSR, JK, HLD and TB supervised the overall running of the overall Seychelles warbler project and contributed resources; HLD managed the longitudinal fitness database.
-EAF and KLB generated and processed the telomere data; LGS performed the analyses and wrote the manuscript in discussion with DSR.
+EAF, KLB and LGS generated and processed the telomere data; LGS performed the analyses and wrote the manuscript, in discussion with DSR.
 All authors read and contributed to the manuscript.

### 2015-09-14 Scripts/SW juvenile telomeres IPC functions.r

@@ -432,7 +432,13 @@ ggsurv <- function(s, CI = 'def', plot.cens = T, surv.col = 'gg.def',
 #Round to 2 or 3 dp and keep trailing zeros
 round2 <- function(x)
   {
+  if(x < 0.01)
+  {
+    return('< 0.01')
+  } else
+  {
   sprintf(round(x,2), fmt="%.2f")
+  }
   }
 
 round3 <- function(x)

### 2015-09-14 Manuscript/SW_early_life_telomeres_v9.Rmd
Test from home laptop

@@
 We performed all analyses using R version 3.0.1 [@RDevelopmentCoreTeam2011].
-Our sampling regime covers a long time-period, and temporal variation in rainfall and food availability on Cousin may cause variation  in a range of social and evironmental variables [@Brouwer2012].
+Our sampling regime covers `r max(juv$LayYear)-min(juv$LayYear)` years, and temporal variation in rainfall and food availability on Cousin may cause variation  in a range of social and evironmental variables [@Brouwer2012].
 This makes it difficult to disentangle how spatial and temporal social and environmental variation affects early-life telomere dynamics.

### 2015-09-14 Manuscript/SW_early_life_telomeres_v9.Rmd
Manuscript method updates

--- 2015-09-14 Manuscript/SW_early_life_telomeres_v9.Rmd
@@
 We performed all analyses using R version 3.0.1 [@RDevelopmentCoreTeam2011].
-Our sampling regime covers a long time-period, and temporal variation in rainfall and food availability on Cousin may cause variation  in a range of social and evironmental variables [@Brouwer2012].
-This makes it difficult to disentangle how spatial and temporal social and environmental variation affects early-life telomere dynamics.
-we seperately analysed how variation in early life conditions among and within breeding seasons was related to variation in early life telomere length.
-First, we tested whether there was a temporal effect on telomere lengthby calculating average telomere length among all birds born in a given breeding season.
+Our sampling regime covers `r max(juv$LayYear)-min(juv$LayYear)` years, and temporal variation in rainfall and food availability on Cousin over this period is expected to result in variation  in a range of social and evironmental variables [@Brouwer2012].
+This makes it difficult to disentangle how spatial and temporal processes differentially affect early-life telomere dynamics.
+To get around this, we seperately analysed how variation in early life conditions among and within breeding seasons was related to variation in early life telomere length.
+
+First, to analyse whether temporal variation in food availability drives population-level variation in telomere dynamics, we calculated average telomere length among all birds born in a given breeding season.
 We tested for a difference in amomg breeding seasons using a one-way ANOVA, and then used linear regression to test whether variation in mean telomere length among breeding seasons is explained by temporal variation in insect abundance.
 
-We tested tested how spatial early-life environmental and social conditions influence telomere length within cohorts using linear mixed models.
-Telomere length was standarsised by mean-centering within cohorts (i.e.
-using all birds born in a given breeding season).
-As explanatory variables we included tarsus, length, sex, territory quality, yearly food availability, and the number of helping and non-helping subordinates present in the natal territory  ([@Brouwer2012]).
-
-We used model averaging We created a full model containing all of the above terms, and a top model set was then defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011].
+We then tested tested how spatial early-life environmental and social conditions influence telomere length within cohorts using linear models.
+For these analyses, telomere length was standarsised by mean-centering within cohorts (i.e.
+using all birds born in a given breeding season), thus controlling for any temporal variation in telomere length.
+As explanatory variables we included tarsus length, sex, territory quality and the number of helping and non-helping subordinates present in the natal territory  ([@Brouwer2012]).
+We used model averaging to assess how these variables shape telomere length and dynamics in early life.
+We created a full model containing all of the above terms, and a top model set was then defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011].
 We report model averaged coefficients, confidence intervals, and 'relative importance', which reflects the relative weights of each predictor variable across the top model set.
 Model averaging was carried out using the MuMIn package (version 1.10.5) in R [@Barton2012].
-
-We used parametric survival analysis to examine variation in early-life telomere length in relation to survival, implemented using the 'flexsurv' package in R.
-The survival time distribution was chosen by comparing the fit of a range of distributions (exponential, Weibull, log-normal, gamma, gompertz and generalized F) against the observed survival data, by visually inspecting plots and by comparing AIC values.
-As explanatory variables of survival we included early-life telomere length, plus the factors identified as being related to telomere length in the previous analysis.
-With this approach we aimed to assess the effect of early-life telomere length on survival while controlling for potentially confounding variables that may be correlated with telomere length.
 
@@
 To account for regression to the mean effects, we applied a correction based on correlations among samples within individuals following Verhulst *et al.* [-@Verhulst2013].
+Using this dataset we repeated the analyses of among and within cohort telomere dynamics, replacing telomere length with telomere loss as the response variable.
+
+Finally, we used parametric survival analysis to examine variation in early-life telomere length and loss in relation to survival, implemented using the 'flexsurv' package in R.
+The survival time distribution was chosen by comparing the fit of a range of distributions (exponential, Weibull, log-normal, gamma, gompertz and generalized F) against the observed survival data, by visually inspecting plots and by comparing AIC values.
+As explanatory variables of survival we included early-life telomere length/loss, plus the factors identified as being related to telomere length in the previous analysis.
+With this approach we aimed to assess the effect of early-life telomere length on survival while controlling for potentially confounding variables that may be correlated with telomere dynamics.
 

### 2015-09-15 Manuscript/SW_early_life_telomeres_SOM_v9.Rmd
More manuscript updates - mainly results

--- 2015-08-14 Manuscript/SW_early_life_telomeres_SOM_v9.Rmd
@@
                fig.show='asis',
-               dev='png')
+               dev=c('png','pdf'))
 
@@
 
-**Table S1** Model selection table for analysis of factors affecting early life telomere length in juvenile Seychelles warblers.
+```{r load and clean, include = F}
+
+#Load and clean data
+source('Scripts/SW juvenile telomeres IPC load.r')
+source('Scripts/SW juvenile telomeres IPC functions.r')
+source('Scripts/SW juvenile telomeres IPC clean.r')
+
+```
+
+**Table S1** Model selection table for analysis of factors affecting early life telomere length in Seychelles warbler chicks.
 Only models that remained in the top model set for model averaging (AICc $\leq$ 6 compared to the top model) are included here.
-i = intercept.
 
@@
 #Run the model
-lm1 <- lmer(TLKB ~ Agemonths + TQspace + Insect + GroupSize + Sex + Condition + Helper + (1|LayYear),
-            data = FlSAall,REML=FALSE)
-
-lm1 <- standardize(lm1)
+lm1 <- lm(cenTL ~ cenHelper + cenNonHelper + Sex + cenTarsus  + cenTQ,
+            data = chickall)
 
@@
 t1 <- as.matrix(subset(lm2, delta<=6))
-colnames(t1)[2:7] <- c('Sex', 'Body condition','Group size','Helper','Food av.','TQ')
-getterms <- function(x)
-  {
-  temp <- c('i',names(which(!(is.na(x[2:7])))))
-  return(paste(temp,collapse = " + "))
-  }
 
-tableS1 <- data.frame(Model = apply(t1,1,getterms),t1[,c('df','AICc','delta','weight')])
+#Sort out row/col names
+nams <- colnames(t1)
+nams[nams == 'cenTarsus'] <- 'Tarsus'
+nams[nams == 'cenNonHelper'] <- 'Non helpers'
+nams[nams == 'SexMales'] <- 'Sex'
+nams[nams == 'cenTQ'] <- 'TQ'
+nams[nams == 'cenHelper'] <- 'Helpers'
+
+colnames(t1) <- nams
+
+tableS1 <- data.frame(Model = apply(t1,1,getterms,c(2:6)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
 colnames(tableS1) <- c('Model','df','AICc', 'Delta AICc','Weight')
@@
 tableS1$Model[tableS1$Model == ''] <- 'Null model'
-kable(tableS1,digits=3)
+tableS1[,3:5] <- lapply(tableS1[,3:5],as.numeric)
+kable(tableS1,digits=2)
 
@@
 
-```{r Figure S1,fig.width=4,fig.height=4}
+**Table S2** Model selection table for analysis of factors affecting early life telomere length in Seychelles warbler fledglings.
+Only models that remained in the top model set for model averaging (AICc $\leq$ 6 compared to the top model) are included here.
 
-temp <- dd
+```{r Table S2}
 
-ddFigS1 <- ddply(dd,
-                 .(Age),
-                 summarize,
-                 TLKBmean = mean(TLKB),
-                 TLKBse = se(TLKB))
+temp <- subset(FlSAall,Ageclass == 'FL')
 
-ddFigS1 <- ddFigS1[complete.cases(ddFigS1),]
-ddFigS1 <- subset(ddFigS1,Age<14)
+#Run the model
+lm1 <- lm(cenTL ~ cenHelper + cenNonHelper + Sex + cenTarsus  + cenTQ,
+            data = temp)
 
-FigS1 <- ggplot(ddFigS1,
-       aes(x = Age,
-           y = TLKBmean)) +
-  geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse)) +
-  ylab('Telomere length (kb)') + 
-  xlab('Age') + 
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 14), 
-        axis.text.y = element_text(size = 14), 
-        axis.title.y = element_text(size = 16,vjust=0.8),
-        axis.title.x = element_text(size = 16))
+#Model averaging
+lm2 <- dredge(lm1)
+t1 <- as.matrix(subset(lm2, delta<=6))
 
-FigS1
+#Sort out row/col names
+nams <- colnames(t1)
+nams[nams == 'cenTarsus'] <- 'Tarsus'
+nams[nams == 'cenNonHelper'] <- 'Non helpers'
+nams[nams == 'SexMales'] <- 'Sex'
+nams[nams == 'cenTQ'] <- 'TQ'
+nams[nams == 'cenHelper'] <- 'Helpers'
+
+colnames(t1) <- nams
+
+tableS2 <- data.frame(Model = apply(t1,1,getterms,c(2:6)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
+colnames(tableS2) <- c('Model','df','AICc', 'Delta AICc','Weight')
+rownames(tableS2) <- NULL
+tableS2$Model[tableS2$Model == ''] <- 'Null model'
+tableS2[,3:5] <- lapply(tableS2[,3:5],as.numeric)
+kable(tableS2,digits=2)
 
@@
 
-**Figure S1** Cross-sectional telomere length (mean $\pm$ s.e.) in relation to age in Seychelles warblers.
+**Table S3** Model selection table for analysis of factors affecting early life telomere length in Seychelles warbler sub-adults Only models that remained in the top model set for model averaging (AICc $\leq$ 6 compared to the top model) are included here.
 
-PAGEBREAK
+```{r Table S3}
 
-```{r Figure S2,fig.width=4,fig.height=4}
+temp <- subset(FlSAall,Ageclass == 'SA')
 
-temp <- FlSA
+#Run the model
+lm1 <- lm(cenTL ~ cenHelper + cenNonHelper + Sex + cenTarsus  + cenTQ,
+            data = temp)
 
-ddFigS2 <- ddply(temp,
-                 .(LayYear),
-                 summarize,
-                 TLKBmean = mean(TLKB),
-                 TLKBse = se(TLKB))
+#Model averaging
+lm2 <- dredge(lm1)
+t1 <- as.matrix(subset(lm2, delta<=6))
 
-ddFigS2 <- ddFigS2[complete.cases(ddFigS2),]
-ddFigS2$LayYear <- as.numeric(paste(ddFigS2$LayYear))
+#Sort out row/col names
+nams <- colnames(t1)
+nams[nams == 'cenTarsus'] <- 'Tarsus'
+nams[nams == 'cenNonHelper'] <- 'Non helpers'
+nams[nams == 'SexMales'] <- 'Sex'
+nams[nams == 'cenTQ'] <- 'TQ'
+nams[nams == 'cenHelper'] <- 'Helpers'
 
-FigS2 <- ggplot(ddFigS2,
-       aes(x = LayYear,
-           y = TLKBmean)) +
-  geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse)) +
-  ylab('Telomere length (kb)') + 
-  xlab('Lay Year') + 
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 14), 
-        axis.text.y = element_text(size = 14), 
-        axis.title.y = element_text(size = 16,vjust=0.8),
-        axis.title.x = element_text(size = 16))
+colnames(t1) <- nams
 
-FigS2
+tableS3 <- data.frame(Model = apply(t1,1,getterms,c(2:6)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
+colnames(tableS3) <- c('Model','df','AICc', 'Delta AICc','Weight')
+rownames(tableS3) <- NULL
+tableS3$Model[tableS3$Model == ''] <- 'Null model'
+tableS3[,3:5] <- lapply(tableS3[,3:5],as.numeric)
+kable(tableS3,digits=2)
 
 ```
-
-**Figure S2** Cross-sectional juvenile telomere length (mean $\pm$ s.e.) across birth years in Seychelles warblers

### 2015-09-15 Manuscript/SW_early_life_telomeres_tables_and_figures v9.Rmd
More manuscript updates - mainly results

@@
 
-**Figure 1** Telomere length and age Seychelles warblers, based on cross-sectional **(A)** and longitudinal **(B)** data.
-Points and error bars represent mean and 95% confidence intervals, respectively.
-In **B**, telomere loss is a standardised and corrected rate (see methods for details) based on differences in telomere length between the sampling event of birds as a juvnile, and the first available sampling as an adult.
+**Figure 1** Telomere length (mean $\pm$ s.e.) and age Seychelles warblers, based on cross-sectional **(A)** and longitudinal **(B)** data.
+In **B**, telomere loss is a standardised and corrected rate (see methods for details) based on the difference in telomere length from the sample taken from birds in their first year of life, and the first available sampling as an adult.
 Values above zero indicate telomere shortening in early life.
 
-**Figure 2** Temporal variation in early-life telomere length in the Seychelles warbler based on cross-sectional **(A,B)** and longitudinal **(C,D)** data (see Figure 1 and main text for details).
-Points and error bars represent mean and standard error telomere length for each summer (blue) or winter (red) field season.
+**Figure 2** Temporal variation in early-life telomere dynamics in the Seychelles warbler based on cross-sectional **(A,B)** and longitudinal **(C,D)** data (see Fig.
+1 and main text for details).
+Points and error bars represent mean and standard error telomere length for all birds born in each summer (blue) or winter (red) breeding season.
 In **A** and **C** points are ordered temporally, while in **B** and **D** telomere length/loss for each season is plotted against island-wide food availability.
@@
 **A** Model averaged estimates and 95% confidence intervals for all explanatory terms used in a linear model with telomere length as the response variable.
-Numbers in brackets are the relative importance of each term in the top model set (see main text for details; Food av.
-= annual food availability, TQ = territory quality).
-**B** Presence/absence of helpers, **C** sex and **D** tarsus length in relation to raw telomere length.
-Points and error bars in **B and C** are mean $\pm$ standard error telomere lengths for each group, and the line and shaded areas from **D** represent fitted values and 95% confidence limits from a linear regression.
+Numbers in brackets are the relative importance of each term in the top model set (see main text for details; TQ = territory quality).
+**B** Presence/absence of helpers and **c** tarsus length in relation to raw telomere length.
+Points and error bars in **B** are mean $\pm$ s.e., and the line and shaded areas from **C** represent fitted values and 95% confidence limits from a linear regression.
 
 **Figure 4** Factors affecting telomere length in Seychelles warblers fledglings **(A)** and subadults **(B)**.
-Model averaged estimates and 95% confidence intervals for all explanatory terms used in a linear model with  telomere length as the response variable.
-Numbers in brackets are the relative importance of each term in the top model set (see main text for details; Food av.
-= annual food availability, TQ = territory quality).
+Points and error bars represent model averaged estimates and 95% confidence intervals for all explanatory terms used in a linear model with  telomere length as the response variable.
+Numbers in brackets are the relative importance of each term in the top model set (see main text for details; TQ = territory quality).
 
-**Figure 5** Kaplan-Meier curves showing the relationship between telomere length and survival in chick (**A**) and juvenile (**B**) Seychelles warblers.
+**Figure 5** Temporal variation in early-life telomere dynamics and lifespan in the Seychelles warbler, based on cross-sectional **(A)** and longitudinal **(B)** data (see Fig.
+1 and main text for details).
+Points and error bars represent means and $\pm$ s.e., respectively.
+
+**Figure 6** Kaplan-Meier curves showing the relationship between telomere length and survival in chick (**A**), fledgling  (**B**) and sub-adult (**C**) Seychelles warblers.
 Telomere length is binned into groups here for visualisation purposes only (long and short = greater than or less than median telomere length, repectively).
@@
 
-```{r Figure 1,fig.width=8,fig.height=9}
+```{r Figure 1,fig.width=6,fig.height=9}
 ######################
@@
            y = TLKBmean)) +
-  ylab('Corrected telomere loss') + 
+  ylab('Telomere length (kb)') + 
   xlab('') +
@@
            y = TROCmean)) +
-  ylab('Telomere length (kb)') + 
+  ylab('Corrected telomere loss') + 
   xlab('') +
@@
 
-ddFig2A <- ddply(juv,
-                 .(CatchYear,Season),
-                 summarize,
-                 TLKBmean = mean(TLKB),
-                 TLKBse = se(TLKB))
-ddFig2A <- ddFig2A[complete.cases(ddFig2A),]
-ddFig2A <- subset(ddFig2A,TLKBse < 4)
-
-Fig2A <- ggplot(ddFig2A,
+Fig2A <- ggplot(juvseason,
        aes(x = CatchYear,
@@
   geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse),
-                  position = position_dodge(width = 1)) +
+                  position = position_dodge(width = 0.2)) +
   scale_color_manual(values = wes_palette("Royal1")) +
@@
 
-temp <- juv
-temp$Insect <- factor(temp$Insect)
-
-ddFig2B <- ddply(temp,
-                 .(Insect,Season),
-                 summarize,
-                 TLKBmean = mean(TLKB),
-                 TLKBse = se(TLKB))
-ddFig2B$Insect <- as.numeric(paste(ddFig2B$Insect))
-ddFig2B <- ddFig2B[complete.cases(ddFig2B),]
-
-Fig2B <- ggplot(ddFig2B,
+Fig2B <- ggplot(juvseason,
        aes(x = Insect,
@@
 
-ddFig2C <- ddply(Loss,
-                 .(CatchYear,Season),
-                 summarize,
-                 TROCmean = mean(TROC),
-                 TROCse = se(TROC))
-ddFig2C <- ddFig2C[complete.cases(ddFig2C),]
-ddFig2C <- subset(ddFig2C,TROCse < 4)
+temp <- loss.season[complete.cases(loss.season),]
 
-Fig2C <- ggplot(ddFig2C,
+Fig2C <- ggplot(temp,
        aes(x = CatchYear,
@@
   geom_pointrange(aes(ymin=TROCmean-TROCse,ymax=TROCmean+TROCse),
-                  position = position_dodge(width = 1)) +
+                  position = position_dodge(width = 0.2)) +
   scale_color_manual(values = wes_palette("Royal1")) +
-  annotate('text',label = 'C',x=1995,y=0.009)
+  annotate('text',label = 'C',x=1995,y=0.005)
 
 ####################################
-temp <- Loss
-temp$Insect <- factor(temp$Insect)
-
-ddFig2D <- ddply(temp,
-                 .(Insect,Season),
-                 summarise,
-                 TROCmean = mean(TROC),
-                 TROCse = se(TROC))
-ddFig2D$Insect <- as.numeric(paste(ddFig2D$Insect))
-ddFig2D <- ddFig2D[complete.cases(ddFig2D),]
-
-Fig2D <- ggplot(ddFig2D,
+Fig2D <- ggplot(temp,
        aes(x = Insect,
@@
 
-```{r Figure 3,fig.width=8,fig.height=9}
+```{r Figure 3,fig.width=9,fig.height=9}
 
@@
 Fig3A <- forestplot(table2) +
-  annotate('text',label = 'A',x=0.6,y=1.2)
+  annotate('text',label = 'A',x=0.6,y=0.6)
 
@@
 temp <- chicks
+temp$HelperF <- ifelse(chicks$Helper>0,'Helpers present','Helpers absent')
 
 ddFig3B <- ddply(temp,
-                 .(Sex),
+                 .(HelperF),
                  summarize,
@@
 Fig3B <- ggplot(ddFig3B,
-       aes(x = Sex,
-           y = TLKBmean)) +
-  ylab('Telomere length (kb)') + 
-  xlab('') +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 14), 
-        axis.text.y = element_text(size = 14), 
-        axis.title.y = element_text(size = 16,vjust=0.8),
-        axis.title.x = element_text(size = 16),
-        legend.position = c(0.1,0.9),
-        legend.title = element_blank(),
-        legend.text = element_text(size = 15)) +
-  geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse)) +
-  annotate('text',label = 'B',x=0.5,y=7)
-
-######################
-#3C
-######################
-
-temp <- chicks
-temp$HelperF <- ifelse(chicks$Helper>0,'Helpers present','Helpers absent')
-
-ddFig3C <- ddply(temp,
-                 .(HelperF),
-                 summarize,
-                 TLKBmean = mean(TLKB),
-                 TLKBse = se(TLKB))
-
-Fig3C <- ggplot(ddFig3C,
        aes(x = HelperF,
@@
   geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse)) +
-  annotate('text',label = 'C',x=0.5,y=8)
+  annotate('text',label = 'B',x=0.5,y=8)
 
 ######################
-#3D
+#3C
 ######################
 
-Fig3D <- ggplot(chicks,
+Fig3C <- ggplot(chicks,
                 aes(x = Tarsus,
@@
   scale_color_manual(values = wes_palette("Royal1")) +
-  annotate('text',label = 'D',x=19,y=15)
+  annotate('text',label = 'C',x=19,y=15)
 
-multiplot(Fig3A,Fig3B,Fig3C,Fig3D,layout=matrix(c(1:4), nrow=2, byrow=FALSE))
+multiplot(Fig3A,Fig3B,Fig3C,layout=matrix(c(1,1,2,3),2,2,byrow=TRUE))
 
@@
 
-```{r Figure 5,fig.height=8,fig.width=5}
+```{r Figure 5,fig.height=8,fig.width=4}
+
+Fig5A <- ggplot(juvseason,
+                aes(x = TLKBmean,
+                    y = Lifespan,
+                    col = Season)) +
+  geom_pointrange(aes(ymin=Lifespan-Lifespanse,ymax=Lifespan+Lifespanse)) +
+  geom_errorbarh(aes(xmin = TLKBmean-TLKBse,xmax = TLKBmean+TLKBse)) +
+  theme_classic() +
+  ylab('Remaining life') + 
+  xlab('Telomere length (kb)') + 
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        legend.position = 'none') +
+  scale_color_manual(values = wes_palette("Royal1")) +
+  annotate('text',x=2.5,y=12.5,label='A')
+
+Fig5B <- ggplot(loss.season,
+                aes(x = TROCmean,
+                    y = Lifespan,
+                    col = Season)) +
+  geom_pointrange(aes(ymin=Lifespan-Lifespanse,ymax=Lifespan+Lifespanse)) +
+  geom_errorbarh(aes(xmin = TROCmean-TROCse,xmax = TROCmean+TROCse)) +
+  theme_classic() +
+  ylab('Remaining life') + 
+  xlab('Corrected telomere loss') + 
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 11), 
+        axis.text.y = element_text(size = 11), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        legend.position = 'none') +
+  scale_color_manual(values = wes_palette("Royal1")) +
+  stat_smooth(method = 'lm',col = 'darkgrey',fullrange = T) +
+  annotate('text',x=-0.004,y=14,label='B')
+
+multiplot(Fig5A,Fig5B)
+
+```
+
+PAGEBREAK
+
+**Figure 6**
+
+```{r Figure 6,fig.height=9,fig.width=4}
 
 s1 <- survfit(Surv(time = Lifespan+1, event = Died)~TLF, data = chicks)
-Fig5A <- ggsurv(s1,plot.cens=F) + 
+Fig6A <- ggsurv(s1,plot.cens=F) + 
   theme_classic() +
@@
 ###########################################
-
-s2 <- survfit(Surv(time = Lifespan+1, event = Died)~TLF, data = FlSAall)
-Fig5B <- ggsurv(s2,plot.cens=F) + 
+s2 <- survfit(Surv(time = Lifespan+1, event = Died)~TLF, data = subset(FlSAall,Ageclass == 'FL'))
+Fig6B <- ggsurv(s2,plot.cens=F) + 
   theme_classic() +
@@
         axis.title.x = element_text(size = 16),
-        legend.position = c(0.8,0.8),
+        legend.position = 'none',
         legend.text = element_text(size = 12),
@@
 
-multiplot(Fig5A,Fig5B)
+###########################################
+
+s2 <- survfit(Surv(time = Lifespan+1, event = Died)~TLF, data = subset(FlSAall,Ageclass == 'SA'))
+Fig6C <- ggsurv(s2,plot.cens=F) + 
+  theme_classic() +
+  ylab('Proportion survived') + 
+  xlab('Age (years)') + 
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 14), 
+        axis.text.y = element_text(size = 14), 
+        axis.title.y = element_text(size = 16,vjust=0.8),
+        axis.title.x = element_text(size = 16),
+        legend.position = 'none',
+        legend.text = element_text(size = 12),
+        legend.title = element_blank()) +
+  scale_colour_manual(name = 'TL',
+                      values = c('darkred','red')) +
+  guides(linetype=F) +
+  annotate('text',x=0.5,y=1.1,label='C')
+
+multiplot(Fig6A,Fig6B,Fig6C)
 

### 2015-09-15 Manuscript/SW_early_life_telomeres_v9.Rmd
More manuscript updates - mainly results

@@
     highlight: null
+    reference_docx: myref.docx
 bibliography: MyBib.bib
@@
 There is evidence from a range of taxa that the greatest rate of telomere loss occurs in early life (e.g.
-[@Frenck1998; @Haussmann2003]), and that the extent of this early-life telomere shortening is influenced by the environment [@Boonekamp2014].
-Moreover, a recent laboratory study showed that early-life telomere dynamics are related to late-life survival [@Heidinger2012].
-However, whether telomeres provide a link between early-life conditions and late-life survival in the wild is not known.
+[@Frenck1998; @Haussmann2003]), and that the extent of this telomere shortening is influenced by the conditions experienced early in life [@Nettle2015; @Reichert2015].
+Importantly, early-life telomere dynamics are related to both short-term and late-life survival [@Heidinger2012; @Boonekamp2014], and to other parameters such as cognition [@Nettle2015a].
+However, the extent to which telomeres provide a link between early-life conditions and late-life survival in the wild is very poorly understood.
+Moreover, how early life telomere dynamics vary over spatial and temporal scales is not known.
 
@@
 
-First, to analyse whether temporal variation in food availability drives population-level variation in telomere dynamics, we calculated average telomere length among all birds born in a given breeding season.
-We tested for a difference in amomg breeding seasons using a one-way ANOVA, and then used linear regression to test whether variation in mean telomere length among breeding seasons is explained by temporal variation in insect abundance.
+First, to analyse whether temporal variation in food availability drives population-level variation in telomere dynamics, we calculated average telomere length among all birds born in a given breeding season (i.e.
+a cohort).
+We tested for a difference in telomere length amomg cohorts using a one-way ANOVA, and then used linear regression to test whether variation in mean telomere length among cohorts is explained by temporal variation in insect abundance.
 
-We then tested tested how spatial early-life environmental and social conditions influence telomere length within cohorts using linear models.
-For these analyses, telomere length was standarsised by mean-centering within cohorts (i.e.
-using all birds born in a given breeding season), thus controlling for any temporal variation in telomere length.
+We then tested tested how spatially varying early-life environmental and social conditions influence telomere length within cohorts using linear models.
+For these analyses, telomere length was standarsised by mean-centering within cohorts, thus controlling for any temporal variation in telomere length.
 As explanatory variables we included tarsus length, sex, territory quality and the number of helping and non-helping subordinates present in the natal territory  ([@Brouwer2012]).
@@
 
-Finally, we used parametric survival analysis to examine variation in early-life telomere length and loss in relation to survival, implemented using the 'flexsurv' package in R.
+Finally, we analysed the effects of telomere dynamics on survival, again by looking at variation both within and among cohorts.
+First, we tested the prediction that cohort-level telomere dynamics are associated with survival by comparing the mean telomere length and loss of each cohort with the mean lifespan of all individuals, using linear regression.
+We then  used parametric survival analysis to test whether individual-level telomere length and loss were related to to survival.
+For this analysis, implemented in the FlexSurv package in R, we mean-centred telomere length within cohorts to eliminate cohort-level variation in telomeres and survival.
 The survival time distribution was chosen by comparing the fit of a range of distributions (exponential, Weibull, log-normal, gamma, gompertz and generalized F) against the observed survival data, by visually inspecting plots and by comparing AIC values.
-As explanatory variables of survival we included early-life telomere length/loss, plus the factors identified as being related to telomere length in the previous analysis.
+As explanatory variables of survival we included early-life telomere length/loss (we ran a seperate analysis for each, using the cross-seciotna and longitudinal data, respectively), plus the factors identified as being related to telomere length in the previous analysis.
 With this approach we aimed to assess the effect of early-life telomere length on survival while controlling for potentially confounding variables that may be correlated with telomere dynamics.
@@
 
-#####Early-life telomere length varies with age and among breeding seasons
+#####Early-life telomere length and age
 Mean ($\pm$ standard error) telomere length in first-year birds was `r round2(mean(FlSAall$TLKB))` $\pm$ `r round2(se(FlSAall$TLKB))` kb.
-This is lower than the mean for one year-old birds based on cross-sectional data (`r round2(mean(subset(dd,Age==1)$TLKB))` $\pm$ `r round2(se(subset(dd,Age==1)$TLKB))` kb; Figure S1).
+This is lower than the mean for one year-old birds based on cross-sectional data (`r round2(mean(subset(dd,Age==1)$TLKB))` $\pm$ `r round2(se(subset(dd,Age==1)$TLKB))` kb).
 However, when we considered age classes seperately, we found substantial age-related variation in telomere length within the first year of life (F = `r round2(anova(lm(TLKB~Ageclass,data = juv))[1,4])`; *P* = `r round2(anova(lm(TLKB~Ageclass,data = juv))[1,5])`).
-This cross-sectional data revealed that chicks have longer telomeres than any other age group, but that subsadults and adults had longer telomeres than fledglings (Fig.
+This revealed that chicks have longer telomeres than any other age group, but also that there was an apparent increase in average telomere length from the fledgling to subadult stages (Fig.
 1A).
-A longitudinal analysis of telomere loss showed that chicks, fleglings and subadults all lost telomeres early in life, with highest and lowest rates of loss in chicks and subaldults, respectively, although the difference among age classes was not significant (F = `r round2(anova(lm(TROC~Ageclass,data = Loss))[1,4])`; *P* = `r round2(anova(lm(TROC~Ageclass,data = Loss))[1,5])`; Fig.
+However, a longitudinal analysis of telomere loss showed that chicks, fleglings and subadults all lost telomeres early in life (intercept of model of telomere loss and age class significantly greater than zero; *P* = `r round2(summary(lm(TROC~Ageclass,data = Loss))$coef[1,4])`), but that there was no difference in the rate of telomere loss among age classes (F = `r round2(anova(lm(TROC~Ageclass,data = Loss))[1,4])`; *P* = `r round2(anova(lm(TROC~Ageclass,data = Loss))[1,5])`; Fig.
 1B).
@@
 
+#####Temporal variation in early life telomere dynamics
 Early-life telomere length varied significantly among breeding seasons (one-way ANOVA, F = `r round2(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,4])`; *P* = `r round2(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,5])`; Fig.
-2A), and this variation in average telomere length over breeding seasons was positively related to island-wide food aviilability (linear regression, R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=juvseason))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data = juvseason))[1,4])`; *P* = `r round2(anova(lm(Insect~TLKBmean,data = juvseason))[1,5])`; Fig.
+2A), and this variation in average telomere length over breeding seasons was positively related to island-wide food aviilability (linear regression, R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,5])`; Fig.
 2B).
 Similarly, the amount of telomeres lost in early life varied among breeding seasons (F = `r round2(anova(lm(TROC~factor(FieldPeriodID),data = Loss))[1,4])`; *P* = `r round2(anova(lm(TROC~factor(FieldPeriodID),data = Loss))[1,5])`; Fig.
-2C), and individuals born in seasons with low food availability lost telomeres, on average, at a faster rate than those born in years with high food availability (linear regression, R^2^ = `r round2(summary(lm(Insect~TROCmean,data=loss.season))$r.squared)`; F = `r round2(anova(lm(Insect~TROCmean,data = loss.season))[1,4])`; *P* = `r round2(anova(lm(Insect~TROCmean,data = loss.season))[1,5])`; Fig.
+2C), and individuals born in seasons with low food availability lost telomeres, on average, at a faster rate than those born in years with high food availability (linear regression, R^2^ = `r round2(summary(lm(Insect~TROCmean,data=loss.season,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TROCmean,data = loss.season,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Insect~TROCmean,data = loss.season,na.action=na.exclude))[1,5])`; Fig.
 2D).
 
-#####Factors affecting early-life telomere dynamics within breeding seasons
-The top explaining within-season variation in chick telomere length contained tarsus length, sex and the number of helpers present in the natal territory (Fig.
-3A; Table S1).
-All three variables had high relative importance int he top model set (Fig.
-3A), and the model containing these variables explained a reasonable amount of variation in telomere length (R^2^ = `r round2(summary(lm(TLKB~Tarsus+Helper+Sex,data=chickall))$r.squared)`).
-Female chicks had longer telomeres than males (Fig.
-3B), while telomere length was positively related to the number of helpers (Fig.
-3C), and negatively related to tarus to length (Fig.
+#####Spatial variation in early life telomere dynamics
+The top model explaining within-season variation in chick telomere length contained tarsus length and the number of helpers present in the natal territory (Table S1), and this model was a much better fit than the null model (intercept only; $\Delta$AICc = `r round2(AICc(lm(cenTL~1,data=chickall))-AICc(lm(cenTL~Tarsus+Helper,data=chickall)))`).
+Both tarsus length and the number of helpers had high relative importance in the top model set (Fig.
+3A), and the model containing these variables explained a reasonable amount of variation in telomere length (R^2^ = `r round2(summary(lm(cenTL~cenTarsus+cenHelper,data=chickall))$r.squared)`).
+Telomere length was positively related to the number of helpers (Fig.
+3C), and negatively related to tarus length (Fig.
 3D).
-However, it should be noted that model-averaged confidence intervals for sex and the number of helpers (but not tarsus length) marginally overlapped zero.
-For both juveniles and sub-adults the null model (intercept only) was the top model (Table S1), and no explanatory terms had high relative importance in the top model set, with all estimates close to zero (Fig.
-4).
+
+For fledglings, the null model was the top model explaining early-life telomere length (Table S1).
+Moreover, none of the explanatory terms had high importance in the top model set, and confidence intervals of all estimates close to zero (Fig.
+4A), suggesting that none of the variables analysed had a significant effect on fledgling telomere length.
+For sub-adults, the top model contained territory quality (Table S1), which was negatively related to sub-adult telomere length (Fig.
+4B).
+However, there was only a small difference in AIC between this model and the null model ($\Delta$AICc = `r round2(AICc(lm(cenTL~1,data=subset(FlSAall,Ageclass == 'SA')))-AICc(lm(cenTL~cenTQ,data=subset(FlSAall,Ageclass == 'SA'))))`).
+Moreover, territory quality explained a tiny amount of variation in subadult telomere length (R^2^ = `r round2(summary(lm(cenTL~cenTQ,data=subset(FlSAall,Ageclass == 'SA')))$r.squared)`), and the confidence intervals of the territory quality effect overlapped with zero (Fig.
+4B).
 
@@
 Due to a limited sample size, we considered all age classes together for this analysis.
-We found that none of these variables were associated with differences in telomere loss (all *P* > 0.05).
+We found that none of these variables were associated with differences in telomere shortening (all *P* > 0.05).
 
 #####Early life telomere dynamics and survival
-A log-normal survival model best fitted the Seychelles warbler data (AIC = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="lnorm")$AIC)`, AIC of next best distribution = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="weibull")$AIC)`).
-A survival model including juvenile telomere length, body condition and early-life food availability showed that only juvenile telomere length had an effect on survival (estimate = `r round2(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB+Insect), dist="lnorm")$res[3,1])`, CI = `r round2(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB+Insect), dist="lnorm")$res[3,2])`-`r round2(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB+Insect), dist="lnorm")$res[3,3])`; Fig.
-2A).
-This survival effect was most pronounced in individuals between five and ten years of age (Fig.
-2B).
-Because telomere length and annual food availability are highly correlated (*R* = `r round2(cor(FlSAall$Insect,FlSAall$TLKB)) #confidence test`), we ran an analysis of early-life telomere length on survival excluding early-life food availability, and the telomere effect beacme marginally non-significant (estimate = `r round2(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,1])`, CI = `r round2(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,2])`-`r round2(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,3])`).
+At the population-level, we found no evidence that individuals born in breeding seasons with high food avilability had longer lifespans (R^2^ = `r round2(summary(lm(Lifespan~Insect,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Lifespan~Insect,data=juvseason,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Lifespan~Insect,data=juvseason,na.action=na.exclude))[1,5])`), nor that individuals born in years where average telomere length was longer had longer lifespans (R^2^ = `r round3(summary(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))[1,5])`; Fig.
+5A).
+Interestingly, however, we found a strong, negative relationship between population-level telomere loss and lifespan (R^2^ = `r round3(summary(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))[1,5])`; Fig.
+5B), which suggests that individuals born in periods where early-life telomere shortening is lowest live, on average, longer overall.
+
+For the individual-based survival analysis, a log-normal survival model best fitted the Seychelles warbler data (AIC = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="lnorm")$AIC)`, AIC of next best distribution (Weibull) = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="weibull")$AIC)`).
+A survival model including all first year birds, early-life telomere length was not associated with survival (estimate = `r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1])`, CI = `r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2])`-`r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3])`).
+However, we did find differential survival effects among age classes.
+Chick telomere length had no effect on survival  (estimate = `r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1])`, CI = `r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2])`-`r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3])`; Fig.
+6A).
+However, telomere length at both the fledgling and subadult stage was significantly related to survival (fledglings: estimate = `r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'FL'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1])`, CI = `r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'FL'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2])`-`r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'FL'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3])`; subadults: estimate = `r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'SA'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1])`, CI = `r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'SA'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2])`-`r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'SA'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3])`).
+In both cases this effect was positive, with longer telomeres in early life associated with increased survival later in life (Figs 6B, 6C).
+Finally, using the longitudinal data, we found no effect of early-life telomere shortening on subsequent survival (estimate = `r round2(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,1])`, CI = `r round3(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,2])`-`r round3(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,3])`).
 

### 2015-09-15 Scripts/SW juvenile telomeres IPC functions.r

@@ -58,11 +58,9 @@ baselineimp <- function(mymodel,n.permutations = 1000)
 #Get terms
 getterms <- function(x,cols)
 {
-  temp <- c('i',names(which(!(is.na(x[cols])))))
-  return(paste(temp,collapse = " + "))
+  temp <- which(!(is.na(x[cols])))
+  return(paste(names(temp),collapse = " + "))
 }
-
-
 
 
 

### 2015-09-15 Manuscript/SW_early_life_telomeres_v9.Rmd
JAE formatting, abstract and intro edited

@@
 
-Lewis G. Spurgin^1,2^,  Kat Bebbington^1^, Eleanor A. Fairfield^1^, Jan Komdeur^3^, Terry Burke^4^, Hannah, L. Dugdale^2^,  and David S. Richardson^1,5^.
+Lewis G. Spurgin^a,b^,  Kat Bebbington^a^, Eleanor A. Fairfield^a^, Jan Komdeur^c^, Terry Burke^d^, Hannah, L. Dugdale^c,d^,  and David S. Richardson^a,e,^.
 
-1. School of Biological Sciences, University of East Anglia, Norwich Research Park, NR4 7TJ, United Kingdom
-2. Department of Zoology, Edward Grey Institute, University of Oxford, Oxford, UK
-3. Behavioural Ecology and Self-organization Group, Centre for Ecological and Evolutionary Studies, University of Groningen, Groningen, The Netherlands
-4. Department of Animal and Plant Sciences, NERC Biomolecular Analysis Facility, University of Sheffield, Sheffield, UK
-5. Nature Seychelles, Roche Caiman, Mahé, Republic of Seychelles
+a. School of Biological Sciences, University of East Anglia, Norwich Research Park, NR4 7TJ, United Kingdom
+b. Department of Zoology, Edward Grey Institute, University of Oxford, Oxford, UK
+c. Behavioural Ecology and Self-organization Group, Centre for Ecological and Evolutionary Studies, University of Groningen, Groningen, The Netherlands
+d. Department of Animal and Plant Sciences, NERC Biomolecular Analysis Facility, University of Sheffield, Sheffield, UK
+e. Nature Seychelles, Roche Caiman, Mahé, Republic of Seychelles
 
 **Correspondence:** Lewis Spurgin: lewisspurgin@gmail.com; David Richardson: david.richardson@uea.ac.uk
+
+###Abstract
+1.
+Favourable conditions early in life can confer delayed fitness advantages.
+However, the mechanistic basis of these 'silver spoon effects' remains poorly understood.
+Telomeres - protective caps on the ends of chromosomes - shorten in response to early life stress, and early life telomere shortening can cause reduced later life survival.
+It is therefore possible that telomeres provide a 'missing link' between early life environmental variation and later life fitness.
+
+2.
+The prevalence and strength of silver spoon effects has pronounced consequences for individual fitness and population growth rates.
+Understanding the physiological and ecological links between early life experiences and later life survival is therefore fundamental to our understanding of life-history evolution and population ecology alike.
+
+3.
+We studied how telomere length links spatiotemporal variation in early-life conditions to survival in the Seychelles warbler (*Acrocephalus sechellensis*).
+The Seychelles warbler system is excellently suited to addressing this question, due to the excellent, long-term lonigtudinal data on survival and senescence.
+We combined this with cross-seciotnal and longitudinal telomere measurements taken from over 300 birds, and analyse how the variation in the early life environment within and across cohorts affects telomere length and survival.
+
+4.
+We show that early-life telomere length varies markedly across cohorts, and that this variation is positively related to annual variation in food availability.
+We then show that having short telomeres as a juvenile is associated with reduced late life survival, but that this relationship was subject to strong cohort effects.
+Having short telomeres was associated with reduced late-life survival among individuals born in years where food was limited, but not among individuals born in years with high food availability.
+
+5.
+Our results show that telomeres can reveal hidden costs at the individual and cohort levels, and suggest that they may provide a link between early-life conditions and late-life survival in wild populations.
 
@@
 
-###Abstract
-Favourable conditions early in life can confer delayed survival advantages.
-These 'silver spoon effects' are most likely common in nature and are central to our understanding of life history evolution and population ecology.
-However, how silver spoon effects vary among individuals, populations and species remains poorly understood.
-Telomeres can act as biomarkers of cost in wild populations, and may provide insight into how silver spoon effects occur.
-Using a dataset spanning over a decade, we studied how telomere length links spatiotemporal variation in early-life conditions to survival to in the Seychelles warbler (*Acrocephalus sechellensis*).
-We show that early-life telomere length varies markedly across birth years, and that this variation is positively related to annual variation in food availability.
-We then show that having short telomeres as a juvenile is associated with reduced late life survival, but that this relationship was subject to strong cohort effects.
-Having short telomeres was associated with reduced late-life survival among individuals born in years where food was limited, but not among individuals born in years with high food availability.
-Our results suggest that telomeres can reveal hidden costs and complex silver spoon effects in wild populations.
-
-PAGEBREAK
-
 ###Introduction
 Being subject to favourable environmental conditions during development and growth can confer fitness advantages later in life (so called 'silver spoon effects') [@Grafen1988; @Monaghan2008].
-There is now evidence that adult fitness can be affected by a range of early-life experiences, including variation in resource availability [@Hayward2013], population density [@Nussey2007] and natural or anthropogenic environmental disturbance [@Reid2003].
+There is now evidence that adult fitness can be affected by a range of early-life experiences, including variation in habitat quality and resource availability [@Madsen2000; @VandePol2006; @Hayward2013], population density [@Nussey2007;@Douhard2013] and natural or anthropogenic environmental disturbance [@Reid2003; @Cartwright2014].
 Understanding the causes and consequences of these early-life experiences is key to understanding many ecological and evolutionary  processes, including  patterns of natural and sexual selection,  population growth rates and even local extinction [@Coulson2001; @Roach2014].
@@
 
-The longitudinal study (since 1986) of the Seychelles warbler (*Acrocephalus sechellensis*) population on Cousin Island provides an excellent opportunity to assess the relationship between early-life environmental variation, telomere length and late-life fitness.
+The longitudinal study (since 1986) of the Seychelles warbler (*Acrocephalus sechellensis*) population on Cousin Island provides an excellent model system for studying senescnece in the wild [@Hammers2015].
+Specifically, it is ideally suited for assessing the relationship between early-life environmental variation, telomere length and late-life fitness.
 Firstly, due to the isolated nature of the study population and intensive field monitoring, we have excellent survival data and sampling material spanning many years (see methods, below).
@@
 
-###Ethics
-All research activities were approved by the ethics committee at the University of East Anglia, the Seychelles Department of Environment and the Seychelles Bureau of Standards (approval reference A0347).
-
 ###Data accessibility
 All data and scripts required to reproduce the analyses within this manuscript are available on GitHub (https://github.com/lewisspurgin/SW_Early_life_telomeres)
-
-###Competing interests
-There are no competing interests
-
-###Authors' contributions
-DSR developed the idea and obtained funding for the overall telomere project; DSR, JK, HLD and TB supervised the overall running of the overall Seychelles warbler project and contributed resources; HLD managed the longitudinal fitness database.
-EAF, KLB and LGS generated and processed the telomere data; LGS performed the analyses and wrote the manuscript, in discussion with DSR.
-All authors read and contributed to the manuscript.
 
@@
 We thank everyone who has helped in the field, and the current Seychelles warbler research group for useful discussions.
-
-###Funding
 This work was funded by two Natural Environment Research Council (NERC) grants to DSR (NE/F02083X/1 and NE/K005502/1) on which JK, TB and HLD were project partners.

### 2015-09-16 Manuscript/SW_early_life_telomeres_tables_and_figures v9.Rmd
Discussion and results work

@@
 
-**Figure 2** Temporal variation in early-life telomere dynamics in the Seychelles warbler based on cross-sectional **(A,B)** and longitudinal **(C,D)** data (see Fig.
+**Figure 2** Temporal variation in early-life telomere dynamics in the Seychelles warbler based on cross-sectional **(A)** and longitudinal **(B)** data (see Fig.
 1 and main text for details).
-Points and error bars represent mean and standard error telomere length for all birds born in each summer (blue) or winter (red) breeding season.
-In **A** and **C** points are ordered temporally, while in **B** and **D** telomere length/loss for each season is plotted against island-wide food availability.
-Lines and shaded areas represent fitted values and 95% confidence limits from a linear regression.
+Points and error bars represent mean and standard error telomere length for all birds born in each summer (black) or winter (grey) breeding season, in relation to island-wide food availability.
+Trend lines represent fitted values from a linear regression.
 
@@
 
-**Figure 5** Temporal variation in early-life telomere dynamics and lifespan in the Seychelles warbler, based on cross-sectional **(A)** and longitudinal **(B)** data (see Fig.
+**Figure 5** Temporal, cohort-level variation in early-life telomere dynamics and lifespan in the Seychelles warbler, based on cross-sectional **(A)** and longitudinal **(B)** data (see Fig.
 1 and main text for details).
-Points and error bars represent means and $\pm$ s.e., respectively.
+Points and error bars represent means and $\pm$ s.e., respectively, of all birds born in each summer (black) or winter (grey) breeding season.
 
@@
 
-```{r Figure 1,fig.width=6,fig.height=9}
+```{r Figure 1,fig.width=5,fig.height=9}
 ######################
@@
   geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse)) +
-  annotate('text',label = 'A',x=0.5,y=6.2)
+  annotate('text',label = 'A',x=0.6,y=6.2)
 
@@
   geom_pointrange(aes(ymin=TROCmean-TROCse,ymax=TROCmean+TROCse)) +
-  annotate('text',label = 'B',x=0.5,y=0.0018)
+  annotate('text',label = 'B',x=0.6,y=0.0018)
 
@@
 
-```{r Figure 2,fig.width=9,fig.height=9}
-
-Fig2A <- ggplot(juvseason,
-       aes(x = CatchYear,
-           y = TLKBmean,
-           col = Season)) +
-  ylab('Telomere length') + 
-  xlab('Year') +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 14), 
-        axis.text.y = element_text(size = 14), 
-        axis.title.y = element_text(size = 16,vjust=0.8),
-        axis.title.x = element_text(size = 16),
-        legend.position = 'none') +
-  geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse),
-                  position = position_dodge(width = 0.2)) +
-  scale_color_manual(values = wes_palette("Royal1")) +
-  annotate('text',label = 'A',x=1995,y=10)
+```{r Figure 2,fig.width=5,fig.height=9}
 
@@
 
-Fig2B <- ggplot(juvseason,
+Fig2A <- ggplot(juvseason,
        aes(x = Insect,
@@
   geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse)) +
-  stat_smooth(method = 'lm',se=T,col = 'darkgrey') +
-  scale_color_manual(values = wes_palette("Royal1")) +
-  annotate('text',label = 'B',x=-1.5,y=8)
+  stat_smooth(method = 'lm',se = F,col = 'black') +
+  scale_color_manual(values = c('black',grey(0.4))) +
+  annotate('text',label = 'A',x=-1.5,y=8)
 
@@
 
-Fig2C <- ggplot(temp,
-       aes(x = CatchYear,
-           y = TROCmean,
-           col = Season)) +
-  ylab('Corrected telomere loss') + 
-  xlab('Year') +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 14), 
-        axis.text.y = element_text(size = 14), 
-        axis.title.y = element_text(size = 16,vjust=0.8),
-        axis.title.x = element_text(size = 16),
-        legend.position = 'none') +
-  geom_pointrange(aes(ymin=TROCmean-TROCse,ymax=TROCmean+TROCse),
-                  position = position_dodge(width = 0.2)) +
-  scale_color_manual(values = wes_palette("Royal1")) +
-  annotate('text',label = 'C',x=1995,y=0.005)
-
 ####################################
-Fig2D <- ggplot(temp,
+Fig2B <- ggplot(temp,
        aes(x = Insect,
@@
         legend.position = 'none') +
-  stat_smooth(method = 'lm', se=T, col = 'darkgrey') +
-  scale_color_manual(values = wes_palette("Royal1")) +
-  annotate('text',label = 'D',x=-1.5,y=0.0075)
+  stat_smooth(method = 'lm',se = F,  col = 'black') +
+  scale_color_manual(values = c('black',grey(0.4))) +
+  annotate('text',label = 'B',x=-1.5,y=0.0075)
 
-multiplot(Fig2A,Fig2B,Fig2C,Fig2D,layout=matrix(1:4,2,2))
+multiplot(Fig2A,Fig2B)
 
@@
   geom_point(size=3) +
-  stat_smooth(method = 'lm',col = 'black') +
-  scale_color_manual(values = wes_palette("Royal1")) +
+  stat_smooth(method = 'lm',se = F,col = 'black') +
+  scale_color_manual(values = c('black',grey(0.4))) +
   annotate('text',label = 'C',x=19,y=15)
@@
         legend.position = 'none') +
-  scale_color_manual(values = wes_palette("Royal1")) +
+  scale_color_manual(values = c('black',grey(0.4))) +
   annotate('text',x=2.5,y=12.5,label='A')
@@
         legend.position = 'none') +
-  scale_color_manual(values = wes_palette("Royal1")) +
-  stat_smooth(method = 'lm',col = 'darkgrey',fullrange = T) +
+  scale_color_manual(values = c('black',grey(0.4))) +
+  stat_smooth(method = 'lm',se = F,col = 'black',fullrange = T) +
   annotate('text',x=-0.004,y=14,label='B')
@@
   scale_colour_manual(name = 'TL',
-                      values = c('darkred','red')) +
+                      values = c('black',grey(0.4))) +
   guides(linetype=F) +

### 2015-09-16 Manuscript/SW_early_life_telomeres_v9.Rmd
Discussion and results work

@@
 
-###Telomeres reveal complex silver spoon effects in a wild population
+###Telomeres reveal silver spoon effects in a wild population
 
@@
 Telomeres - protective caps on the ends of chromosomes - shorten in response to early life stress, and early life telomere shortening can cause reduced later life survival.
-It is therefore possible that telomeres provide a 'missing link' between early life environmental variation and later life fitness.
+It is thus possible that telomeres provide a 'missing link' between early-life environmental variation and later-life fitness.
 
 2.
-The prevalence and strength of silver spoon effects has pronounced consequences for individual fitness and population growth rates.
+The prevalence and strength of silver spoon effects has consequences for individual fitness and population growth rates.
 Understanding the physiological and ecological links between early life experiences and later life survival is therefore fundamental to our understanding of life-history evolution and population ecology alike.
@@
 3.
-We studied how telomere length links spatiotemporal variation in early-life conditions to survival in the Seychelles warbler (*Acrocephalus sechellensis*).
-The Seychelles warbler system is excellently suited to addressing this question, due to the excellent, long-term lonigtudinal data on survival and senescence.
-We combined this with cross-seciotnal and longitudinal telomere measurements taken from over 300 birds, and analyse how the variation in the early life environment within and across cohorts affects telomere length and survival.
+We studied how telomere length and shortening link spatiotemporal variation in early-life conditions to survival in the Seychelles warbler (*Acrocephalus sechellensis*).
+The Seychelles warbler system is excellently suited to addressing this question, due to the long-term lonigtudinal data on survival and senescence.
+We combined this with cross-sectional and longitudinal telomere measurements taken from over 300 birds from multiple cohorts, and analyse how the variation in the early life environment within and across cohorts affects telomere dynamics and survival.
 
 4.
-We show that early-life telomere length varies markedly across cohorts, and that this variation is positively related to annual variation in food availability.
-We then show that having short telomeres as a juvenile is associated with reduced late life survival, but that this relationship was subject to strong cohort effects.
-Having short telomeres was associated with reduced late-life survival among individuals born in years where food was limited, but not among individuals born in years with high food availability.
+We show that early-life telomere length and shortening rates vary markedly across cohorts, and that this variation is related to temporal fluctuations variation in food availability.
+Individuals born in breeding seasons with high food availability have longer telomeres and reduced telomere shortening compared to individuals born in years with low food availability.
+Within cohorts, we found telomere length in chicks was related to growth, and to the number of helpers present in the natal territory.
+However, the number of helpers was not related to telomere shortening rates, nor to telomere length in fledgling or subadult birds.
 
 5.
-Our results show that telomeres can reveal hidden costs at the individual and cohort levels, and suggest that they may provide a link between early-life conditions and late-life survival in wild populations.
+Our results show that telomeres can reveal complex hidden costs at the individual and cohort levels, and suggest that they may provide a link between early-life conditions and late-life survival in wild populations.
 
@@
 A key process that has been associated with adult survival is the rate at which telomere repeats are lost.
-Telomeres shorten with age [@Monaghan2006], and in response to oxidative damage that can occur due to environmental stress [@VonZglinicki2002].
+Telomeres are protective caps on the ends of chromosomes which shorten with age [@Monaghan2006], and in response to oxidative damage that can occur due to environmental stress [@VonZglinicki2002].
 When telomeres reach a critical shortness cells senesce  [@Campisi2003], resulting in organismal senescence and death  [@Wong2003].
-These links between telomere length and senescence have inspired a great deal of recent research into telomere ecology [@Monaghan2014].
-An emerging consensus from this research is that telomeres can act as biomarkers of cost in wild populations, retaining signatures of ecological stress that are otherwise difficult to detect [@Schultner2014; @Asghar2015].
+These links between telomere length and senescence have inspired a great deal of recent research into telomere ecology [reviewed in @Horn2010; @Haussmann2010; @Monaghan2014].
+While there is little evidence the relationship between telomere dynamics and survival is causal [@Simons2015], there is now excellent evidence that telomeres can act as biomarkers of cost in wild populations, retaining signatures of ecological stress that are otherwise difficult to detect [@Monaghan2014; @Schultner2014; @Asghar2015].
 
 There is evidence from a range of taxa that the greatest rate of telomere loss occurs in early life (e.g.
-[@Frenck1998; @Haussmann2003]), and that the extent of this telomere shortening is influenced by the conditions experienced early in life [@Nettle2015; @Reichert2015].
+[@Frenck1998; @Haussmann2003]), and that the extent of this telomere shortening is influenced by the conditions experienced early in life [@Price2013; @Monaghan2014; @Nettle2015; @Reichert2015].
 Importantly, early-life telomere dynamics are related to both short-term and late-life survival [@Heidinger2012; @Boonekamp2014], and to other parameters such as cognition [@Nettle2015a].
@@
 
-The longitudinal study (since 1986) of the Seychelles warbler (*Acrocephalus sechellensis*) population on Cousin Island provides an excellent model system for studying senescnece in the wild [@Hammers2015].
+The longitudinal study (since 1986) of the Seychelles warbler (*Acrocephalus sechellensis*) population on Cousin Island provides an excellent model system for studying senescence in the wild [@Hammers2015].
 Specifically, it is ideally suited for assessing the relationship between early-life environmental variation, telomere length and late-life fitness.
@@
 For these analyses, telomere length was standarsised by mean-centering within cohorts, thus controlling for any temporal variation in telomere length.
-As explanatory variables we included tarsus length, sex, territory quality and the number of helping and non-helping subordinates present in the natal territory  ([@Brouwer2012]).
+As explanatory variables we included tarsus length, sex, territory quality and the number of helping and non-helping subordinates present in the natal territory  [@Brouwer2012].
 We used model averaging to assess how these variables shape telomere length and dynamics in early life.
@@
 #####Temporal variation in early life telomere dynamics
-Early-life telomere length varied significantly among breeding seasons (one-way ANOVA, F = `r round2(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,4])`; *P* = `r round2(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,5])`; Fig.
-2A), and this variation in average telomere length over breeding seasons was positively related to island-wide food aviilability (linear regression, R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,5])`; Fig.
+Early-life telomere length varied significantly among breeding seasons (one-way ANOVA, F = `r round2(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,4])`; *P* = `r round3(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,5])`), and this variation in average telomere length over breeding seasons was positively related to island-wide food aviilability (linear regression, R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,5])`; Fig.
+2A).
+Running the regression of cohort-level telomere length and food availability seperately for each age class revealed a postive and significant relationship in fledglings (R^2^ = `r round2(summary(lm(Insect~TLKB,data=flseason))$r.squared)`; F = `r round2(anova(lm(Insect~TLKB,data=flseason))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKB,data=flseason))[1,5])`) and subadults (R^2^ = `r round2(summary(lm(Insect~TLKB,data=saseason))$r.squared)`; F = `r round2(anova(lm(Insect~TLKB,data=saseason))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKB,data=saseason))[1,5])`), but a marginally non-significant *negative* relationship was found for chicks (R^2^ = `r round2(summary(lm(Insect~TLKB,data=chickseason))$r.squared)`; F = `r round2(anova(lm(Insect~TLKB,data=chickseason))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKB,data=chickseason))[1,5])`).
+
+The rate of telomere shortening in early life did not vary significantly among breeding seasons (F = `r round2(anova(lm(TROC~factor(FieldPeriodID),data = Loss))[1,4])`; *P* = `r round2(anova(lm(TROC~factor(FieldPeriodID),data = Loss))[1,5])`); however, what variation there was in rates of telomere shortening could be explained by temporal variation in food availability (R^2^ = `r round2(summary(lm(Insect~TROCmean,data=loss.season,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TROCmean,data = loss.season,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Insect~TROCmean,data = loss.season,na.action=na.exclude))[1,5])`).
+This pattern corresponded with the cross-sectional data: individuals born in seasons with low food availability lost telomeres, on average, at a faster rate than those born in years with high food availability (Fig.
 2B).
-Similarly, the amount of telomeres lost in early life varied among breeding seasons (F = `r round2(anova(lm(TROC~factor(FieldPeriodID),data = Loss))[1,4])`; *P* = `r round2(anova(lm(TROC~factor(FieldPeriodID),data = Loss))[1,5])`; Fig.
-2C), and individuals born in seasons with low food availability lost telomeres, on average, at a faster rate than those born in years with high food availability (linear regression, R^2^ = `r round2(summary(lm(Insect~TROCmean,data=loss.season,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TROCmean,data = loss.season,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Insect~TROCmean,data = loss.season,na.action=na.exclude))[1,5])`; Fig.
-2D).
+Unforutunately our limited sample size within the longitudinal dataset precluded us from running this analysis seperately for each age class.
 
 #####Spatial variation in early life telomere dynamics
-The top model explaining within-season variation in chick telomere length contained tarsus length and the number of helpers present in the natal territory (Table S1), and this model was a much better fit than the null model (intercept only; $\Delta$AICc = `r round2(AICc(lm(cenTL~1,data=chickall))-AICc(lm(cenTL~Tarsus+Helper,data=chickall)))`).
+The top model explaining within-season variation in chick telomere length contained tarsus length and the number of helpers present in the natal territory (Table S1), and this model was a much better fit than the null model ($\Delta$AICc = `r round2(AICc(lm(cenTL~1,data=chickall))-AICc(lm(cenTL~Tarsus+Helper,data=chickall)))`).
 Both tarsus length and the number of helpers had high relative importance in the top model set (Fig.
@@
 ###Discussion
-Previous studies have shown that in juvenile Seychelles warblers, neither territory quality nor annual variation in food availability predict survival to adulthood or late-life survival [@Brouwer2006; @Brouwer2012; @Hammers2013].
 Here we show that telomeres link early-life conditions to later-life survival in the Seychelles warbler.
-These findings have a number of implications for research on life history evolution in wild populations.
+We first show that telomere length and rates of shortening in early life are subject to strong cohort effects, and that cohort-level telomere dynamics can be explained by island-wide temporal variation in food availability.
+We then show that within cohorts, the social environemnt (i.e.
+number of helpers) affects telomere length, but that this effect is age-dependent.
+Finally, we show that telomere length in early life predicts later-life survival.
+These findings have important implications for research on life history evolution in wild populations.
 
-Our finding that food availability is related to early-life telomere length adds to a growing body of correlative and experimental evidence showing that the natal environment can have pronounced impacts on early-life telomere length [@Monaghan2014].
-However, very few studies have shown that temporal variation in environmental conditions affects telomere length [@Mizutani2013].
+Our finding that food availability is related to early-life telomere length adds to a growing body of evidence from humans and wild animals, showing that the natal environment can have pronounced impacts on early-life telomere length [reviewed in @Price2013; @Monaghan2014].
+However, very few studies have shown that temporal variation in environmental conditions affects telomere dynamics in natural populations [@Mizutani2013; @Watson2015].
 We suspect this is more due to a lack of available long-term datasets with telomere screening, rather than the Seychelles warbler being unique.
@@
 
-We found no evidence that telomere length predicts survival to adulthood in Seychelles warblers.
-This is surprising, as we previously found that telomere length in adult Seychelles warblers predicts survival to the next year [@Barrett2013a].
-Moreover, our cross-sectional data show that telomere length is longer, on average, in one year-old birds compared to juveniles (Figure S1), while there was no evidence of telomere lengthening from the longitudinal data.
-A pattern like this is expected when there is selective disappearance - a known phenomenon in studies of telomere dynamics whereby individuals with short telomeres die, resulting in an apparent increase in telomere length in cross-sectional data [@Haussmann2008a].
-Thus, the raw data suggest that a survival effect may occur in early life, but this was not supported by our logistic regression analyses.
-This contradiction in results may be due to lack of power if a survival effect is present but weak and/or cohort specific.
-Increased sampling within years will help better elucidate the short-term survival impacts of early-life telomere length in this system.
+If temporal variation in early-life telomere dynamics is common in nature, this has a number of ramifications for our understanding of telomere dynamics in natural populations.
+First, our findings suggest that telomere dynamics
 
-In captive zebra finches, juvenile telomere length predicts late-life survival [@Heidinger2012].
-Our results are the first to replicate this relationship in the wild.
-However, the effect varied among cohorts, suggesting that the relationship between early-life telomere length and late-life survival may not be a straightforward one in natural systems.
-Indeed, it appears that the interplay between the environment, telomeres and late-life survival at a given point in time is a snapshot of a fluid and dynamic evolutionary process.
-We therefore recommend that relationships involving telomeres from single cohorts should be interpreted with caution until we gain a better understanding of what drives spatiotemporal variation in telomere dynamics and survival.
+We found that after controlling for among-cohort variation in telomere length, tarsus length and the number of helpers were the most important predictors of telomere length in Seychelles warbler chicks.
 
-Together, our results confirm that there is cohort-level variation in the relationship between the early-life environment and later-life survival, and suggest that telomeres go some to way to capturing this complexity.
+In captive zebra finches, juvenile telomere length predicts late-life survival [@Heidinger2012], and in wild bird populations early-life telomere length has recently been linked to survival during the nestling-phase [@Watson2015], and survival to adulthood [@Boonekamp2014].
+Our results are, to our knowledge, the first to demonstrate that early-life telomere length is associated with late-life survival in a wild population.
+Previous studies have shown that in juvenile Seychelles warblers, the number of helpers, but not temporal variation in food availability, predicts later-life survival [@Brouwer2006; @Brouwer2012; @Hammers2013].
+
+That we found a relationship between telomere length, but not loss, and survival is surprising as other studies have found the opposite [@Boonekamp2014].
+
+There are numerous promising avenues for future research into telomere molecular ecology in this system and others.
+Here we have considered survival, but telomere length and shortening in early life may also be linked to recruitment into the breeding population and/or reproductive senescence.
+
+Together with previous research on the Seychelles warbler, our results confirm that both temporal and spatial variation in the early-life environment is linked to later-life survival, and our results suggest that telomeres go some to way to capturing this complexity.
 Thus, our results support the idea that telomeres can be used as a 'generic currency' with which we can measure variation in costs among individuals and populations [@Asghar2015].
@@
 
-Despite the promise of telomeres as biomarkers of cost, ecologists interested in understanding variation within populations are still left with the problem of identifying what explains individual-level variation in telomere length.
-Telomere length in juveniles will be the product of a range of factors, including inherited telomere length at birth [@Horn2011], age at sampling [@Heidinger2012], the early-life environment [@Boonekamp2014], and, potentially, individual quality.
-In the Seychelles warbler we do not yet know how these different factors contribute to within-cohort variation in telomere length.
-However, a previous study on the Seychelles warbler showed that maternal heterozygosity influences offspring survival in bad years, but not good years [@Brouwer2007].
-This raises the possibility that in the Seychelles warbler there is variation in individual quality, and high quality individuals are better able to cope with early-life stress [@Brouwer2007].
-Thus in the near future it will become possible to use telomeres to link genetic variation, inheritance, environmental variation and senescence in the Seychelles warbler and other long-term systems.
-
 ###Data accessibility
-All data and scripts required to reproduce the analyses within this manuscript are available on GitHub (https://github.com/lewisspurgin/SW_Early_life_telomeres)
+This manuscript was written in R Markdown (http://rmarkdown.rstudio.com/).
+All data and scripts required to reproduce the manuscript, and all figures and analyses contained within it, will be made available on GitHub.
 

### 2015-09-16 Scripts/SW juvenile telomeres IPC clean.R

@@ -279,6 +279,11 @@ juvseason <- ddply(juv,
 juvseason$cenTL <- juvseason$TLKBmean-mean(juvseason$TLKBmean)
 juvseason <- subset(juvseason,n>2)
 
+chickseason <- aggregate(TLKB~Insect,mean,data=chickall)
+flseason <- aggregate(TLKB~Insect,mean,data=subset(FlSAall,Ageclass == 'FL'))
+saseason <- aggregate(TLKB~Insect,mean,data=subset(FlSAall,Ageclass == 'SA')) 
+                      
+
 loss.season <- ddply(Loss,
                    .(FieldPeriodID,Season),
                    summarize,

### 2015-09-16 Scripts/SW juvenile telomeres IPC functions.r

@@ -440,9 +440,15 @@ round2 <- function(x)
   }
 
 round3 <- function(x)
+{
+  if(x < 0.001)
   {
-  sprintf(round(x,3), fmt="%.3f")
+    return('< 0.001')
+  } else
+  {
+    sprintf(round(x,3), fmt="%.3f")
   }
+}
 
 
 

### 2015-09-16 Manuscript/SW_early_life_telomeres_v9.Rmd
Discussion edits

@@
 We studied how telomere length and shortening link spatiotemporal variation in early-life conditions to survival in the Seychelles warbler (*Acrocephalus sechellensis*).
-The Seychelles warbler system is excellently suited to addressing this question, due to the long-term lonigtudinal data on survival and senescence.
+The Seychelles warbler system is excellently suited to addressing this question, due to the long-term longitudinal data on survival and senescence.
 We combined this with cross-sectional and longitudinal telomere measurements taken from over 300 birds from multiple cohorts, and analyse how the variation in the early life environment within and across cohorts affects telomere dynamics and survival.
@@
 We performed all analyses using R version 3.0.1 [@RDevelopmentCoreTeam2011].
-Our sampling regime covers `r max(juv$LayYear)-min(juv$LayYear)` years, and temporal variation in rainfall and food availability on Cousin over this period is expected to result in variation  in a range of social and evironmental variables [@Brouwer2012].
+Our sampling regime covers `r max(juv$LayYear)-min(juv$LayYear)` years, and temporal variation in rainfall and food availability on Cousin over this period is expected to result in variation  in a range of social and environmental variables [@Brouwer2012].
 This makes it difficult to disentangle how spatial and temporal processes differentially affect early-life telomere dynamics.
-To get around this, we seperately analysed how variation in early life conditions among and within breeding seasons was related to variation in early life telomere length.
+To get around this, we separately analysed how variation in early life conditions among and within breeding seasons was related to variation in early life telomere length.
 
@@
 a cohort).
-We tested for a difference in telomere length amomg cohorts using a one-way ANOVA, and then used linear regression to test whether variation in mean telomere length among cohorts is explained by temporal variation in insect abundance.
+We tested for a difference in telomere length among cohorts using a one-way ANOVA, and then used linear regression to test whether variation in mean telomere length among cohorts is explained by temporal variation in insect abundance.
 
 We then tested tested how spatially varying early-life environmental and social conditions influence telomere length within cohorts using linear models.
-For these analyses, telomere length was standarsised by mean-centering within cohorts, thus controlling for any temporal variation in telomere length.
-As explanatory variables we included tarsus length, sex, territory quality and the number of helping and non-helping subordinates present in the natal territory  [@Brouwer2012].
+For these analyses, both response and explanatory variables were standardised by mean-centering within cohorts, thus controlling for any temporal variation in telomere length and the social and ecological environment.
+As explanatory variables we included tarsus length, sex, territory quality and the number of helping and non-helping subordinates present in the natal territory.
 We used model averaging to assess how these variables shape telomere length and dynamics in early life.
@@
 The survival time distribution was chosen by comparing the fit of a range of distributions (exponential, Weibull, log-normal, gamma, gompertz and generalized F) against the observed survival data, by visually inspecting plots and by comparing AIC values.
-As explanatory variables of survival we included early-life telomere length/loss (we ran a seperate analysis for each, using the cross-seciotna and longitudinal data, respectively), plus the factors identified as being related to telomere length in the previous analysis.
+As explanatory variables of survival we included early-life telomere length/loss (we ran a separate analysis for each, using the cross-sectional and longitudinal data, respectively), plus the factors identified as being related to telomere length in the previous analysis.
 With this approach we aimed to assess the effect of early-life telomere length on survival while controlling for potentially confounding variables that may be correlated with telomere dynamics.
@@
 This is lower than the mean for one year-old birds based on cross-sectional data (`r round2(mean(subset(dd,Age==1)$TLKB))` $\pm$ `r round2(se(subset(dd,Age==1)$TLKB))` kb).
-However, when we considered age classes seperately, we found substantial age-related variation in telomere length within the first year of life (F = `r round2(anova(lm(TLKB~Ageclass,data = juv))[1,4])`; *P* = `r round2(anova(lm(TLKB~Ageclass,data = juv))[1,5])`).
+However, when we considered age classes separately, we found substantial age-related variation in telomere length within the first year of life (F = `r round2(anova(lm(TLKB~Ageclass,data = juv))[1,4])`; *P* = `r round2(anova(lm(TLKB~Ageclass,data = juv))[1,5])`).
 This revealed that chicks have longer telomeres than any other age group, but also that there was an apparent increase in average telomere length from the fledgling to subadult stages (Fig.
 1A).
-However, a longitudinal analysis of telomere loss showed that chicks, fleglings and subadults all lost telomeres early in life (intercept of model of telomere loss and age class significantly greater than zero; *P* = `r round2(summary(lm(TROC~Ageclass,data = Loss))$coef[1,4])`), but that there was no difference in the rate of telomere loss among age classes (F = `r round2(anova(lm(TROC~Ageclass,data = Loss))[1,4])`; *P* = `r round2(anova(lm(TROC~Ageclass,data = Loss))[1,5])`; Fig.
+However, a longitudinal analysis of telomere loss showed that chicks, fledglings and subadults all lost telomeres early in life (intercept of model of telomere loss and age class significantly greater than zero; *P* = `r round2(summary(lm(TROC~Ageclass,data = Loss))$coef[1,4])`), but that there was no difference in the rate of telomere loss among age classes (F = `r round2(anova(lm(TROC~Ageclass,data = Loss))[1,4])`; *P* = `r round2(anova(lm(TROC~Ageclass,data = Loss))[1,5])`; Fig.
 1B).
@@
 #####Temporal variation in early life telomere dynamics
-Early-life telomere length varied significantly among breeding seasons (one-way ANOVA, F = `r round2(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,4])`; *P* = `r round3(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,5])`), and this variation in average telomere length over breeding seasons was positively related to island-wide food aviilability (linear regression, R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,5])`; Fig.
+Early-life telomere length varied significantly among breeding seasons (one-way ANOVA, F = `r round2(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,4])`; *P* = `r round3(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,5])`), and this variation in average telomere length over breeding seasons was positively related to island-wide food availability (linear regression, R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,5])`; Fig.
 2A).
-Running the regression of cohort-level telomere length and food availability seperately for each age class revealed a postive and significant relationship in fledglings (R^2^ = `r round2(summary(lm(Insect~TLKB,data=flseason))$r.squared)`; F = `r round2(anova(lm(Insect~TLKB,data=flseason))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKB,data=flseason))[1,5])`) and subadults (R^2^ = `r round2(summary(lm(Insect~TLKB,data=saseason))$r.squared)`; F = `r round2(anova(lm(Insect~TLKB,data=saseason))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKB,data=saseason))[1,5])`), but a marginally non-significant *negative* relationship was found for chicks (R^2^ = `r round2(summary(lm(Insect~TLKB,data=chickseason))$r.squared)`; F = `r round2(anova(lm(Insect~TLKB,data=chickseason))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKB,data=chickseason))[1,5])`).
+Running the regression of cohort-level telomere length and food availability separately for each age class revealed a positive and significant relationship in fledglings (R^2^ = `r round2(summary(lm(Insect~TLKB,data=flseason))$r.squared)`; F = `r round2(anova(lm(Insect~TLKB,data=flseason))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKB,data=flseason))[1,5])`) and subadults (R^2^ = `r round2(summary(lm(Insect~TLKB,data=saseason))$r.squared)`; F = `r round2(anova(lm(Insect~TLKB,data=saseason))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKB,data=saseason))[1,5])`), but a marginally non-significant *negative* relationship was found for chicks (R^2^ = `r round2(summary(lm(Insect~TLKB,data=chickseason))$r.squared)`; F = `r round2(anova(lm(Insect~TLKB,data=chickseason))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKB,data=chickseason))[1,5])`).
 
@@
 2B).
-Unforutunately our limited sample size within the longitudinal dataset precluded us from running this analysis seperately for each age class.
+Unfortunately our limited sample size within the longitudinal dataset precluded us from running this analysis separately for each age class.
 
@@
 For fledglings, the null model was the top model explaining early-life telomere length (Table S1).
-Moreover, none of the explanatory terms had high importance in the top model set, and confidence intervals of all estimates close to zero (Fig.
+None of the explanatory terms had high importance in the top model set, and confidence intervals of all estimates were close to zero (Fig.
 4A), suggesting that none of the variables analysed had a significant effect on fledgling telomere length.
@@
 4B).
-However, there was only a small difference in AIC between this model and the null model ($\Delta$AICc = `r round2(AICc(lm(cenTL~1,data=subset(FlSAall,Ageclass == 'SA')))-AICc(lm(cenTL~cenTQ,data=subset(FlSAall,Ageclass == 'SA'))))`).
-Moreover, territory quality explained a tiny amount of variation in subadult telomere length (R^2^ = `r round2(summary(lm(cenTL~cenTQ,data=subset(FlSAall,Ageclass == 'SA')))$r.squared)`), and the confidence intervals of the territory quality effect overlapped with zero (Fig.
-4B).
+However, territory quality explained a tiny amount of variation in subadult telomere length (R^2^ = `r round2(summary(lm(cenTL~cenTQ,data=subset(FlSAall,Ageclass == 'SA')))$r.squared)`), and there was only a small difference in AIC between the model containing territory quality and the null model ($\Delta$AICc = `r round2(AICc(lm(cenTL~1,data=subset(FlSAall,Ageclass == 'SA')))-AICc(lm(cenTL~cenTQ,data=subset(FlSAall,Ageclass == 'SA'))))`).
+Moreover, the confidence intervals of the territory quality effect overlapped with zero (Fig.
+4B), suggesting that any negative effect of territory quality on subadult telomere length was negligible.
 
@@
 #####Early life telomere dynamics and survival
-At the population-level, we found no evidence that individuals born in breeding seasons with high food avilability had longer lifespans (R^2^ = `r round2(summary(lm(Lifespan~Insect,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Lifespan~Insect,data=juvseason,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Lifespan~Insect,data=juvseason,na.action=na.exclude))[1,5])`), nor that individuals born in years where average telomere length was longer had longer lifespans (R^2^ = `r round3(summary(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))[1,5])`; Fig.
+At the population-level, we found no evidence that individuals born in breeding seasons with high food availability had longer lifespans (R^2^ = `r round2(summary(lm(Lifespan~Insect,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Lifespan~Insect,data=juvseason,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Lifespan~Insect,data=juvseason,na.action=na.exclude))[1,5])`), nor that individuals born in years where average telomere length was longer had longer lifespans (R^2^ = `r round3(summary(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))[1,5])`; Fig.
 5A).
@@
 We first show that telomere length and rates of shortening in early life are subject to strong cohort effects, and that cohort-level telomere dynamics can be explained by island-wide temporal variation in food availability.
-We then show that within cohorts, the social environemnt (i.e.
+We then show that within cohorts, the social environment (i.e.
 number of helpers) affects telomere length, but that this effect is age-dependent.
@@
 
-Our finding that food availability is related to early-life telomere length adds to a growing body of evidence from humans and wild animals, showing that the natal environment can have pronounced impacts on early-life telomere length [reviewed in @Price2013; @Monaghan2014].
-However, very few studies have shown that temporal variation in environmental conditions affects telomere dynamics in natural populations [@Mizutani2013; @Watson2015].
-We suspect this is more due to a lack of available long-term datasets with telomere screening, rather than the Seychelles warbler being unique.
+Our finding that food availability is related to early-life telomere length and rate of shortening adds to a growing body of evidence from humans and wild animals showing that the natal environment can have pronounced impacts on early-life telomere dynamics [reviewed in @Price2013; @Monaghan2014].
+However, very few studies have shown that temporal variation in environmental conditions affects telomere dynamics in natural populations.
+In a recent study on black-tailed gulls (*Larus crassirostris*), Mizutani *et al.* [-@Mizutani2013] showed that temporal variation in telomere shortening occurred over a two-year period, which they suggest was the result of changing climatic conditions.
+Similarly, Watson *et al.* [-@Watson2015] studied telomere shortening in two cohorts of European storm petrel (*Hydrobates pelagicus*), and found that the cohort raised in more favourable conditions had, on average, longer telomeres.
+The long-term Seychelles warbler dataset has allowed us, using many more cohorts than has been possible in other studies, to show that temporal variation in environmental conditions does indeed affect telomere dynamics, and that this effect can be seen at the population level.
+
+We suspect the current lack of information on temporal variation in telomere dynamics is more due to a lack of available long-term datasets with telomere screening, rather than the Seychelles warbler being unique.
 Indeed, the environment on Cousin is benign in comparison with many regions outside the tropics, where populations undergo large fluctuations in size (e.g.
@@
 We therefore expect that as more long-term studies generate telomere datasets, the effects of temporal changes in the environment on telomere dynamics will become increasingly clear.
+If our findings are replicated in other systems and temporal variation in early-life telomere dynamics is common in nature, this has a number of ramifications for our understanding of telomere dynamics in natural populations.
+First, our findings suggest that telomere dynamics of a population at a given point in time represent a snapshot of a temporally dynamic process.
+More research is now needed within and across multiple cohorts and populations to better understand the complexity of telomere evolutionary ecology.
+Secondly, our findings raise the interesting prospect of using telomeres as indicators of population health, an approach that would be useful to a broad range of conservation and animal health practitioners.
+However, our analyses are confined to a single island, and research across populations with different environments and genetic histories is required in order to provide further insight into the usefulness of telomeres as biomarkers of population health.
 
-If temporal variation in early-life telomere dynamics is common in nature, this has a number of ramifications for our understanding of telomere dynamics in natural populations.
-First, our findings suggest that telomere dynamics
+We found that when looking within cohorts, tarsus length and the number of helpers were the most important predictors of telomere length in Seychelles warbler chicks.
+The tarsus effect most likely due to the fact that in passerine birds tarsus length is correlated with age during the nestling stage (DAVE DO YOU HAVE A REF??).
+Telomere loss is most rapid early in life due to ongoing cell replication, and a negative correlation between telomere length and body size in early life is therefore expectec (REF).
+However, ours is the first study to show that the presence of helping co-breeders has a beneficial effect on offspring in terms of their telomeres.
+In the Seychelles warbler the survival benefts of having helpers in the natal territory is well established [@Komdeur1994;@Brouwer2012], and our confirm that telomeres can act as biomarkers of cost in wild populations [see also @Asghar 2015].
 
-We found that after controlling for among-cohort variation in telomere length, tarsus length and the number of helpers were the most important predictors of telomere length in Seychelles warbler chicks.
+While the link between telomeres and mortality in adults has been established for some time [@Cawthon2003], only recently has the link between early-life telomere dynamics and later-life survival been studied.
+In captive zebra finches, juvenile telomere length predicts late-life survival [@Heidinger2012], and in wild bird populations early-life telomere dynamics have been linked to survival during the nestling-phase [@Watson2015], and survival to adulthood [@Boonekamp2014].
+Our results are, to our knowledge, the first to demonstrate that early-life telomere length is associated with later-life survival in a wild population.
+As with variation in telomere dynamics, the link between telomeres and survival can be seen at both the individual and cohort level, although these effects vary according to to age class, and depending on whether cross-sectional or longitudinal data are used (Figs 5 &6).
 
-In captive zebra finches, juvenile telomere length predicts late-life survival [@Heidinger2012], and in wild bird populations early-life telomere length has recently been linked to survival during the nestling-phase [@Watson2015], and survival to adulthood [@Boonekamp2014].
-Our results are, to our knowledge, the first to demonstrate that early-life telomere length is associated with late-life survival in a wild population.
 Previous studies have shown that in juvenile Seychelles warblers, the number of helpers, but not temporal variation in food availability, predicts later-life survival [@Brouwer2006; @Brouwer2012; @Hammers2013].
 
-That we found a relationship between telomere length, but not loss, and survival is surprising as other studies have found the opposite [@Boonekamp2014].
+Length vs loss and survival
 
-There are numerous promising avenues for future research into telomere molecular ecology in this system and others.
+There are numerous promising avenues for future research into telomere evolutionary ecology in this system and others.
 Here we have considered survival, but telomere length and shortening in early life may also be linked to recruitment into the breeding population and/or reproductive senescence.
-
-Together with previous research on the Seychelles warbler, our results confirm that both temporal and spatial variation in the early-life environment is linked to later-life survival, and our results suggest that telomeres go some to way to capturing this complexity.
-Thus, our results support the idea that telomeres can be used as a 'generic currency' with which we can measure variation in costs among individuals and populations [@Asghar2015].
-The prospect of using telomeres as indicators of population health is a particularly interesting one, as such an approach would be useful to a broad range of conservation and animal health practitioners.
-Our findings suggest that telomere length at the population level does reflect environmental stress.
-However, our analyses are confined to a single island, and research across populations with different environments and genetic histories is required in order to provide further insight into the usefulness of telomeres as biomarkers of population health.
 

### 2015-09-17 Manuscript/SW_early_life_telomeres_v9.Rmd
Changes before sending to DSR

@@
 However, the mechanistic basis of these 'silver spoon effects' remains poorly understood.
-Telomeres - protective caps on the ends of chromosomes - shorten in response to early life stress, and early life telomere shortening can cause reduced later life survival.
+Telomeres - protective caps on the ends of chromosomes - shorten in response to early life stress, and early life telomere shortening has been associated with reduced later-life survival.
 It is thus possible that telomeres provide a 'missing link' between early-life environmental variation and later-life fitness.
@@
 The prevalence and strength of silver spoon effects has consequences for individual fitness and population growth rates.
-Understanding the physiological and ecological links between early life experiences and later life survival is therefore fundamental to our understanding of life-history evolution and population ecology alike.
+Understanding the physiological and ecological links between early life experiences and later life survival is therefore fundamental to our understanding of life-history evolution and population ecology.
 
@@
 The Seychelles warbler system is excellently suited to addressing this question, due to the long-term longitudinal data on survival and senescence.
-We combined this with cross-sectional and longitudinal telomere measurements taken from over 300 birds from multiple cohorts, and analyse how the variation in the early life environment within and across cohorts affects telomere dynamics and survival.
+We combined this with cross-sectional and longitudinal telomere measurements taken from over 300 birds from multiple cohorts, and analyse how variation in the early life environment within and across cohorts affects telomere dynamics and survival.
 
 4.
-We show that early-life telomere length and shortening rates vary markedly across cohorts, and that this variation is related to temporal fluctuations variation in food availability.
+We show that early-life telomere length and shortening rates vary markedly across cohorts, and that this variation is related to temporal fluctuations in food availability.
 Individuals born in breeding seasons with high food availability have longer telomeres and reduced telomere shortening compared to individuals born in years with low food availability.
@@
 However, the number of helpers was not related to telomere shortening rates, nor to telomere length in fledgling or subadult birds.
+Finally, we found that longer telomeres and reduced telomere shortening rates in early life were associated with increased survival later in life.
 
@@
 They are then followed throughout their lives, and as they are non-migratory endemics and are naturally confined to the island [@Komdeur2004a], a biannual census of birds on Cousin during the main (June-August) and minor (December-February) breeding seasons gives an accurate measure of levels of local density and individual survival rate [@Barrett2013a].
-Full details of catching and census methods can be found in Brouwer *et al.* [@Brouwer2012].
+Full details of catching and census methods can be found in Brouwer *et al.* [-@Brouwer2012].
 
@@
 Cousin is also subject to intra- and inter-annual variation in rainfall and food availability, and such island-wide temporal variation may override the effects of absolute territory quality.
-As an estimate of variation in food availability, we calculated an index of the number of insects across the entire island during.
+As an estimate of variation in food availability, we calculated an index of the number of insects across the entire island during each breeding season.
 This  index represents the average number of insects found per unit leaf area over all surveys carried out on the island in a breeding season.
@@
 Each time a bird is caught on Cousin body mass and tarsus length are measured, and age is assessed on the basis of eye colour and previous captures [@Richardson2003b].
-A blood sample (ca 25 $\mu$l) is taken from each bird via brachial venipuncture, and stored in 1 m1 absolute ethanol in a 1.5 ml screw-cap microfuge tube at room temperature.
+A blood sample (ca 25 $\mu$l) is taken from each bird via brachial venipuncture, and stored in 1 ml absolute ethanol in a 1.5 ml screw-cap microfuge tube at room temperature.
 
@@
 Genomic DNA was extracted from a ~2 mm^2^ flake of preserved blood using the DNeasy Blood and Tissue Kit (Qiagen), following the manufacturer's protocol, with the modification of overnight lysis at 37^o^C and a final DNA elution volume of 80 $\mu$l.
-DNA concentration and purity were quantified using a NanoDrop 8000 Spectrophotometer (ThermoScientific), and DNA integrity was validated by by visualization with ethidium bromide after electrophoresis on a 1.2% agarose gel.
+DNA concentration and purity were quantified using a NanoDrop 8000 Spectrophotometer (ThermoScientific), and DNA integrity was validated by visualization with ethidium bromide after electrophoresis on a 1.2% agarose gel.
 Sex was determined using the molecular method outlined by Griffiths *et al.* [-@Griffiths1998].
@@
 Of these birds, `r nrow(chicks)` were nestlings less than two months old (chicks), `r nrow(subset(FlSA,Ageclass=='FL'))` were aged between two and six months (fledglings), and `r nrow(subset(FlSA,Ageclass=='SA'))` were aged seven to twelve months (subadults).
-We measured absolute telomere quantity per diploid genome using a quantitative PCR (qPCR) assay  (full details in [@Barrett2012]).
+For a subset of first-year birds (n = `r nrow(Loss)` individuals) we had longitudinal data, with samples from individuals caught both in their first year and then again within two years.
+We measured absolute telomere quantity per diploid genome for all samples using a quantitative PCR (qPCR) assay  [full details in @Barrett2012].
 
@@
 
-For a subset of first-year birds (n = `r nrow(Loss)` individuals) we had longitudinal data (i.e samples from individuals caught both in their first year and as an adult).
-For these individuals we calculated a rate of telomere loss between the first-year sample and earliest adult sample by subtracting the adult telomere length from the early-life telomere length and dividing this difference by the length of time (in days) between sampling events.
+For the individuals with longitudinal data we calculated a rate of telomere loss between the first-year sample and earliest adult sample by subtracting the adult telomere length from the early-life telomere length and dividing this difference by the length of time (in days) between sampling events.
 To account for regression to the mean effects, we applied a correction based on correlations among samples within individuals following Verhulst *et al.* [-@Verhulst2013].
@@
 First, we tested the prediction that cohort-level telomere dynamics are associated with survival by comparing the mean telomere length and loss of each cohort with the mean lifespan of all individuals, using linear regression.
-We then  used parametric survival analysis to test whether individual-level telomere length and loss were related to to survival.
-For this analysis, implemented in the FlexSurv package in R, we mean-centred telomere length within cohorts to eliminate cohort-level variation in telomeres and survival.
+For this analysis we excluded any recent breeding seasons where >10% of individuals in a cohort were still alive.
+We then  used parametric survival analysis,implemented in the FlexSurv package in R, to test whether individual-level telomere length and loss were related to to survival.
+Again, we mean-centred variables within cohorts to eliminate temporal variation in telomeres and survival.
 The survival time distribution was chosen by comparing the fit of a range of distributions (exponential, Weibull, log-normal, gamma, gompertz and generalized F) against the observed survival data, by visually inspecting plots and by comparing AIC values.
@@
 However, when we considered age classes separately, we found substantial age-related variation in telomere length within the first year of life (F = `r round2(anova(lm(TLKB~Ageclass,data = juv))[1,4])`; *P* = `r round2(anova(lm(TLKB~Ageclass,data = juv))[1,5])`).
-This revealed that chicks have longer telomeres than any other age group, but also that there was an apparent increase in average telomere length from the fledgling to subadult stages (Fig.
+Chicks had longer telomeres than any other age group, but there was an apparent increase in average telomere length from the fledgling to subadult stages (Fig.
 1A).
-However, a longitudinal analysis of telomere loss showed that chicks, fledglings and subadults all lost telomeres early in life (intercept of model of telomere loss and age class significantly greater than zero; *P* = `r round2(summary(lm(TROC~Ageclass,data = Loss))$coef[1,4])`), but that there was no difference in the rate of telomere loss among age classes (F = `r round2(anova(lm(TROC~Ageclass,data = Loss))[1,4])`; *P* = `r round2(anova(lm(TROC~Ageclass,data = Loss))[1,5])`; Fig.
+However, a longitudinal analysis of telomere loss showed that chicks, fledglings and subadults all lost telomeres early in life, but that there was no difference in the rate of telomere loss among age classes (F = `r round2(anova(lm(TROC~Ageclass,data = Loss))[1,4])`; *P* = `r round2(anova(lm(TROC~Ageclass,data = Loss))[1,5])`; Fig.
 1B).
@@
 2A).
-Running the regression of cohort-level telomere length and food availability separately for each age class revealed a positive and significant relationship in fledglings (R^2^ = `r round2(summary(lm(Insect~TLKB,data=flseason))$r.squared)`; F = `r round2(anova(lm(Insect~TLKB,data=flseason))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKB,data=flseason))[1,5])`) and subadults (R^2^ = `r round2(summary(lm(Insect~TLKB,data=saseason))$r.squared)`; F = `r round2(anova(lm(Insect~TLKB,data=saseason))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKB,data=saseason))[1,5])`), but a marginally non-significant *negative* relationship was found for chicks (R^2^ = `r round2(summary(lm(Insect~TLKB,data=chickseason))$r.squared)`; F = `r round2(anova(lm(Insect~TLKB,data=chickseason))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKB,data=chickseason))[1,5])`).
+Running the regression of cohort-level telomere length and food availability separately for each age class revealed positive but non-significant relationships in fledglings (R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=flseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data=flseason,na.action=na.exclude))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKBmean,data=flseason,na.action=na.exclude))[1,5])`) and subadults (R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=saseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data=saseason,na.action=na.exclude))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKBmean,data=saseason,na.action=na.exclude))[1,5])`), but a non-significant *negative* relationship in chicks (R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=chickseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data=chickseason,na.action=na.exclude))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKBmean,data=chickseason,na.action=na.exclude))[1,5])`).
 
@@
 Telomere length was positively related to the number of helpers (Fig.
-3C), and negatively related to tarus length (Fig.
-3D).
+3B), and negatively related to tarus length (Fig.
+3C).
 
@@
 4A), suggesting that none of the variables analysed had a significant effect on fledgling telomere length.
-For sub-adults, the top model contained territory quality (Table S1), which was negatively related to sub-adult telomere length (Fig.
+For subadults, the top model contained territory quality (Table S1), which was negatively related to subadult telomere length (Fig.
 4B).
@@
 
-For the subset of individuals with longitudinal data, we tested whether the factors previously identified as being related early-life telomere length (i.e.
+For the subset of individuals with longitudinal data, we tested whether the factors previously identified as being related to early-life telomere length (i.e.
 tarsus length, sex and number of helpers) were also associated with differences in rates of early life telomere loss.
 Due to a limited sample size, we considered all age classes together for this analysis.
-We found that none of these variables were associated with differences in telomere shortening (all *P* > 0.05).
+We found that none of these variables were associated with differences in telomere loss (all *P* > 0.05).
 
@@
 5A).
-Interestingly, however, we found a strong, negative relationship between population-level telomere loss and lifespan (R^2^ = `r round3(summary(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))[1,5])`; Fig.
-5B), which suggests that individuals born in periods where early-life telomere shortening is lowest live, on average, longer overall.
+Testing the latter relationship separately for each age class revealed that there was no age-specific relationship between cohort-level telomere length and lifespan (all *P* > 0.2).
+There was a negative relationship between population-level telomere loss and lifespan, although this was marginally non-significant (R^2^ = `r round3(summary(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))[1,5])`; Fig.
+5B).
 
 For the individual-based survival analysis, a log-normal survival model best fitted the Seychelles warbler data (AIC = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="lnorm")$AIC)`, AIC of next best distribution (Weibull) = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="weibull")$AIC)`).
-A survival model including all first year birds, early-life telomere length was not associated with survival (estimate = `r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1])`, CI = `r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2])`-`r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3])`).
+A survival model including all first year birds, early-life telomere length was not associated with survival (estimate = `r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`).
 However, we did find differential survival effects among age classes.
-Chick telomere length had no effect on survival  (estimate = `r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1])`, CI = `r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2])`-`r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3])`; Fig.
+Chick telomere length had no effect on survival  (estimate = `r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`; Fig.
 6A).
-However, telomere length at both the fledgling and subadult stage was significantly related to survival (fledglings: estimate = `r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'FL'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1])`, CI = `r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'FL'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2])`-`r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'FL'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3])`; subadults: estimate = `r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'SA'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1])`, CI = `r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'SA'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2])`-`r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'SA'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3])`).
+However, telomere length at both the fledgling and subadult stage was significantly related to survival (fledglings: estimate = `r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'FL'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'FL'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'FL'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`; subadults: estimate = `r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'SA'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'SA'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'SA'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`).
 In both cases this effect was positive, with longer telomeres in early life associated with increased survival later in life (Figs 6B, 6C).
-Finally, using the longitudinal data, we found no effect of early-life telomere shortening on subsequent survival (estimate = `r round2(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,1])`, CI = `r round3(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,2])`-`r round3(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,3])`).
+Finally, using the longitudinal data, we found no effect of early-life telomere shortening on subsequent survival (estimate = `r round2(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,1])`, CI = `r round3(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,3], lessthan = F)`).
 
@@
 The long-term Seychelles warbler dataset has allowed us, using many more cohorts than has been possible in other studies, to show that temporal variation in environmental conditions does indeed affect telomere dynamics, and that this effect can be seen at the population level.
+Interestingly, in the Seychelles warbler annual variation in food availability is not directly linked to survival [@Brouwer2006; @Hammers2013].
+Our results therefore suggest that telomeres can detect hidden costs in natural populations that are not detectable using life-history data alone [@Asghar2015]
 
 We suspect the current lack of information on temporal variation in telomere dynamics is more due to a lack of available long-term datasets with telomere screening, rather than the Seychelles warbler being unique.
-Indeed, the environment on Cousin is benign in comparison with many regions outside the tropics, where populations undergo large fluctuations in size (e.g.
-[@Coulson2001]).
+Indeed, the environment on Cousin is benign in comparison with many regions outside the tropics, where populations undergo large fluctuations in size [e.g.
+@Coulson2001].
 We therefore expect that as more long-term studies generate telomere datasets, the effects of temporal changes in the environment on telomere dynamics will become increasingly clear.
-If our findings are replicated in other systems and temporal variation in early-life telomere dynamics is common in nature, this has a number of ramifications for our understanding of telomere dynamics in natural populations.
-First, our findings suggest that telomere dynamics of a population at a given point in time represent a snapshot of a temporally dynamic process.
+If our findings are replicated in other systems and temporal variation in early-life telomere dynamics is common in nature, this has a number of ramifications for our understanding of senescence in natural populations.
+First, our findings suggest that telomere dynamics of a population at a given point in time represent a snapshot of a temporally varying process.
 More research is now needed within and across multiple cohorts and populations to better understand the complexity of telomere evolutionary ecology.
-Secondly, our findings raise the interesting prospect of using telomeres as indicators of population health, an approach that would be useful to a broad range of conservation and animal health practitioners.
+Secondly, our findings raise the interesting prospect of using telomeres as indicators of population health - an approach that would be useful to a broad range of conservation and animal health practitioners.
 However, our analyses are confined to a single island, and research across populations with different environments and genetic histories is required in order to provide further insight into the usefulness of telomeres as biomarkers of population health.
@@
 We found that when looking within cohorts, tarsus length and the number of helpers were the most important predictors of telomere length in Seychelles warbler chicks.
-The tarsus effect most likely due to the fact that in passerine birds tarsus length is correlated with age during the nestling stage (DAVE DO YOU HAVE A REF??).
-Telomere loss is most rapid early in life due to ongoing cell replication, and a negative correlation between telomere length and body size in early life is therefore expectec (REF).
+The tarsus effect most likely reflects the fact that in passerine birds tarsus length is correlated with age during the nestling stage (DAVE DO YOU HAVE A REF??).
+Telomere loss is most rapid early in life due to ongoing cell replication, and a negative correlation between telomere length and body size in early life is therefore expected (REF).
 However, ours is the first study to show that the presence of helping co-breeders has a beneficial effect on offspring in terms of their telomeres.
-In the Seychelles warbler the survival benefts of having helpers in the natal territory is well established [@Komdeur1994;@Brouwer2012], and our confirm that telomeres can act as biomarkers of cost in wild populations [see also @Asghar 2015].
+In the Seychelles warbler the survival benefits of having helpers in the natal territory is well established [@Komdeur1994;@Brouwer2012], and that this effect is detected in the telomere data is therefore encouraging.
+However, we found no evidence that tarsus length or the number of helpers was related to telomere loss - this is perhaps surprising as telomere loss is expected to be a better indicator of stress than telomere length *per se* [@Boonekamp2014].
+Our (lack of) finding here is probably due to the limited nature of our longitudinal dataset, both in terms of sample size and resolution (i.e.
+time between sampling events).
+Seychelles warblers have not rarely been sampled multiple times within their first year of life, so much of the telomere shortening that occurs in early life will be missed with our sampling regime.
+It is likely, therefore that in this system the effects of environmental variation on telomere loss will only be detected when they are very strong, and that telomere length constitutes a better indicator of early life stress.
 
-While the link between telomeres and mortality in adults has been established for some time [@Cawthon2003], only recently has the link between early-life telomere dynamics and later-life survival been studied.
+While the relationship between telomeres and mortality in adults has been established for some time [@Cawthon2003], only recently has the link between early-life telomere dynamics and later-life survival been studied.
 In captive zebra finches, juvenile telomere length predicts late-life survival [@Heidinger2012], and in wild bird populations early-life telomere dynamics have been linked to survival during the nestling-phase [@Watson2015], and survival to adulthood [@Boonekamp2014].
 Our results are, to our knowledge, the first to demonstrate that early-life telomere length is associated with later-life survival in a wild population.
-As with variation in telomere dynamics, the link between telomeres and survival can be seen at both the individual and cohort level, although these effects vary according to to age class, and depending on whether cross-sectional or longitudinal data are used (Figs 5 &6).
-
-Previous studies have shown that in juvenile Seychelles warblers, the number of helpers, but not temporal variation in food availability, predicts later-life survival [@Brouwer2006; @Brouwer2012; @Hammers2013].
-
-Length vs loss and survival
+This effect can be seen at the individual level, and (to a lesser extent) at the cohort level, although the nature of the relationship varies according to to age class, and depending on whether cross-sectional or longitudinal data are used (Figs 5,6).
+At the individual level, we found that telomere length in fledglings and subadults, but not chicks, was related to survival.
+This is unsurprising, as at the chick stage telomere length at the fledgling and subadult stage is likely to be the best indicator of early life stress (see also previous paragraph).
+The lack of correlation between population-level telomere length and lifespan is more difficult to explain, and given the sample size and resolution of the data, the replationship between population-level telomere loss and lifespan should be cautiously interpreted.
+Nonetheless, our data do support the general conclusion that early-life telomere dynamics are important indicators of early-life stress.
 
 There are numerous promising avenues for future research into telomere evolutionary ecology in this system and others.
-Here we have considered survival, but telomere length and shortening in early life may also be linked to recruitment into the breeding population and/or reproductive senescence.
+Here we have considered survival, but telomere length and shortening in early life may also be linked to other components of fitness [@Monaghan2014].
+Reproductive senescence occurs in the Seychelles warbler [reviewed in @Hammers2015], making this system well suited to examining how telomere length affects lifetime reproductive success.
+A further avenue for future research will be examining, in a quantitative genetic framework, how genetic and environmental components, and their interactions, affect telomere dynamics and senescence in natural populations [@Becker2015; @Asghar2015a].
+We expect that by gaining a fuller understanding of telomere dynamics in natural populations, the fields of life-history evolution and evolutionary ecology will be greatly enhanced.
 

### 2015-09-21 Manuscript/SW_early_life_telomeres_v9.Rmd
First edits after DSR comments

@@
 However, the mechanistic basis of these 'silver spoon effects' remains poorly understood.
-Telomeres - protective caps on the ends of chromosomes - shorten in response to early life stress, and early life telomere shortening has been associated with reduced later-life survival.
-It is thus possible that telomeres provide a 'missing link' between early-life environmental variation and later-life fitness.
+Telomeres - protective caps on the ends of chromosomes - shorten in response to early life stress, and early life telomere shortening is correlated with reduced later-life survival.
+It is thus possible that as telomeres are able to measure the impact of early life costs, they may provide a link between early-life environmental variation and later-life fitness.
 
@@
 We studied how telomere length and shortening link spatiotemporal variation in early-life conditions to survival in the Seychelles warbler (*Acrocephalus sechellensis*).
-The Seychelles warbler system is excellently suited to addressing this question, due to the long-term longitudinal data on survival and senescence.
+The Seychelles warbler system is excellently suited to addressing this question, due to the availability of long-term longitudinal survival and senescence data unconfounded by dispersal.
 We combined this with cross-sectional and longitudinal telomere measurements taken from over 300 birds from multiple cohorts, and analyse how variation in the early life environment within and across cohorts affects telomere dynamics and survival.
@@
 We show that early-life telomere length and shortening rates vary markedly across cohorts, and that this variation is related to temporal fluctuations in food availability.
-Individuals born in breeding seasons with high food availability have longer telomeres and reduced telomere shortening compared to individuals born in years with low food availability.
-Within cohorts, we found telomere length in chicks was related to growth, and to the number of helpers present in the natal territory.
-However, the number of helpers was not related to telomere shortening rates, nor to telomere length in fledgling or subadult birds.
+Individuals born in breeding seasons with high food availability have reduced telomere shortening compared to individuals born in years with low food availability.
+Within cohorts, we found telomere length in chicks was related to tarsus length (which likely represents chick ), and to the number of helpers present in the natal territory.
+However, the number of helpers was not related to telomere shortening rates, nor to telomere length in fledged birds.
 Finally, we found that longer telomeres and reduced telomere shortening rates in early life were associated with increased survival later in life.
@@
 While it is clear that silver spoon effects can occur, we have little understanding of how and under what conditions early-life environments will affect adult fitness.
-Because the phenotypic consequences of an adverse environment can depend on an individual's genetic or epigenetic makeup [@Hoffman2000; @Richards2006], the later-life consequences of a good/poor start are expected to vary among individuals, populations and species.
-Moreover, it is not always possible to fully quantify what constitutes a good or bad environment, and any 'hidden' environmental variation may obscure relationships relationships between the early-life conditions that are measured and adult phenotypes.
+Because the phenotypic consequences of an adverse environment can depend on an individual's initial  condition, and genetic or epigenetic makeup [@Hoffman2000; @Richards2006], the later-life consequences of a good or poor start are expected to vary among individuals, populations and species.
+Moreover, it is not always possible to fully quantify what constitutes a good or bad environment, and any 'hidden' environmental variation may obscure relationships between the early-life conditions that are measured and adult phenotypes.
 Indeed, the pervasiveness of silver spoon effects varies between species [@Drummond2011], cohorts [@Reid2003] and sexes [@Wilkin2009].
@@
 A key process that has been associated with adult survival is the rate at which telomere repeats are lost.
-Telomeres are protective caps on the ends of chromosomes which shorten with age [@Monaghan2006], and in response to oxidative damage that can occur due to environmental stress [@VonZglinicki2002].
+Telomeres are protective caps on the ends of chromosomes which shorten with age [@Monaghan2006], and in response to oxidative stress levels which can be elevated  due to environmental factors [@VonZglinicki2002].
 When telomeres reach a critical shortness cells senesce  [@Campisi2003], resulting in organismal senescence and death  [@Wong2003].
@@
 There are numerous promising avenues for future research into telomere evolutionary ecology in this system and others.
-Here we have considered survival, but telomere length and shortening in early life may also be linked to other components of fitness [@Monaghan2014].
+Here we have considered survival, but telomere length and shortening in early life may also be linked to other components of fitness [reviewed in @Monaghan2014].
 Reproductive senescence occurs in the Seychelles warbler [reviewed in @Hammers2015], making this system well suited to examining how telomere length affects lifetime reproductive success.

### 2015-09-22 Manuscript/SW_early_life_telomeres_tables_and_figures v9.Rmd
Dave's suggestions

@@
 1 and main text for details).
-Points and error bars represent mean and standard error telomere length for all birds born in each summer (black) or winter (grey) breeding season, in relation to island-wide food availability.
+Points and error bars represent mean and standard error telomere length for all birds born in each main (black) or minor (grey) breeding season, in relation to island-wide food availability.
 Trend lines represent fitted values from a linear regression.
@@
 1 and main text for details).
-Points and error bars represent means and $\pm$ s.e., respectively, of all birds born in each summer (black) or winter (grey) breeding season.
+Points and error bars represent means and $\pm$ s.e., respectively, of all birds born in each main (black) or minor (grey) breeding season.
 

### 2015-09-22 Manuscript/SW_early_life_telomeres_v9.Rmd
Dave's suggestions

@@
 However, the mechanistic basis of these 'silver spoon effects' remains poorly understood.
-Telomeres - protective caps on the ends of chromosomes - shorten in response to early life stress, and early life telomere shortening is correlated with reduced later-life survival.
+Telomeres - protective caps on the ends of chromosomes - shorten in response to early life stress, and early-life telomere shortening is correlated with reduced later-life survival.
 It is thus possible that as telomeres are able to measure the impact of early life costs, they may provide a link between early-life environmental variation and later-life fitness.
@@
 Telomeres are protective caps on the ends of chromosomes which shorten with age [@Monaghan2006], and in response to oxidative stress levels which can be elevated  due to environmental factors [@VonZglinicki2002].
-When telomeres reach a critical shortness cells senesce  [@Campisi2003], resulting in organismal senescence and death  [@Wong2003].
+When telomeres reach a critical shortness cells senesce  [@Campisi2003], and the accumulation of these cells can result in organismal senescence and death  [@Wong2003].
 These links between telomere length and senescence have inspired a great deal of recent research into telomere ecology [reviewed in @Horn2010; @Haussmann2010; @Monaghan2014].
-While there is little evidence the relationship between telomere dynamics and survival is causal [@Simons2015], there is now excellent evidence that telomeres can act as biomarkers of cost in wild populations, retaining signatures of ecological stress that are otherwise difficult to detect [@Monaghan2014; @Schultner2014; @Asghar2015].
+While there is little direct evidence the relationship between telomere dynamics and survival is causal [@Barrett2011; @Simons2015], there is now excellent evidence that telomeres can act as biomarkers of cost in wild populations, retaining signatures of ecological stress that are otherwise difficult to detect [@Monaghan2014; @Schultner2014; @Asghar2015].
 
-There is evidence from a range of taxa that the greatest rate of telomere loss occurs in early life (e.g.
-[@Frenck1998; @Haussmann2003]), and that the extent of this telomere shortening is influenced by the conditions experienced early in life [@Price2013; @Monaghan2014; @Nettle2015; @Reichert2015].
+There is evidence from a range of taxa that the greatest rate of telomere loss occurs in early life [e.g.
+@Frenck1998; @Haussmann2003], and that the extent of this telomere shortening is influenced by the conditions experienced early in life [@Price2013; @Monaghan2014; @Nettle2015; @Reichert2015].
 Importantly, early-life telomere dynamics are related to both short-term and late-life survival [@Heidinger2012; @Boonekamp2014], and to other parameters such as cognition [@Nettle2015a].
@@
 
-The longitudinal study (since 1986) of the Seychelles warbler (*Acrocephalus sechellensis*) population on Cousin Island provides an excellent model system for studying senescence in the wild [@Hammers2015].
+The longitudinal study (since 1986) of the Seychelles warbler (*Acrocephalus sechellensis*) population on Cousin Island provides an excellent model system for studying senescence in the wild [reviewed in @Hammers2015].
 Specifically, it is ideally suited for assessing the relationship between early-life environmental variation, telomere length and late-life fitness.
@@
 Secondly, telomere length predicts survival independently of age in adult Seychelles warblers, suggesting that telomeres act as a biomarker of cost in this species [@Barrett2013a].
-Finally, ecological conditions on Cousin are variable over space and time due to spatiotemporal variation in foliage cover and food availability [@VandeCrommenacker2011a].
+Finally, ecological conditions on Cousin are variable over space and time due to weather-induced changes in foliage cover and food availability [@VandeCrommenacker2011a].
 
-In Seychelles warblers, oxidative stress is linked to natal territory quality [@VandeCrommenacker2011a], but neither natal territory quality nor natal local density are linked to early-life or adult survival [@Brouwer2006; @Hammers2013].
-In this study we analyse patterns of early-life telomere length and subsequent survival to test the prediction that telomeres fill a missing link between early-life environmental variation and late-life fitness.
-Specifically we predict that: i) individuals raised in good quality conditions will have longer telomeres, and ii) having longer telomeres in early life is associated with greater short and long-term survival.
+In Seychelles warblers, oxidative stress is linked to natal territory quality [@VandeCrommenacker2011a], but neither natal territory quality nor natal local density appear to be linked to early-life or adult survival [@Brouwer2006; @Hammers2013].
+Importantly, cooperative breeding occurs in the Seychelles warbler [@Komdeur1994a; @Richardson2003a], and the presence of helping subordinates (but not non-helping subordiantes) in the natal territory has been associated with increased survival later in life [@Brouwer2012].
+In this study we test the prediction that telomeres link early-life environmental variation to late-life fitness.
+Because telomere length and shortening will reflect individual-level variation in the costs of early-life experiences, they may allow a more sensitive analysis of the effects of early early-life environmental variation than would be possible with a direct comparison of the early-life environment and survival.
+With this in mind, we predict that: i) individuals raised in good quality conditions will have longer telomeres, and ii) having longer telomeres in early life is associated with greater short and long-term survival.
 
@@
 The Seychelles warbler is a small (~15 g), insectivorous passerine bird with a mean life expectancy of 5.5 years at fledging [@Hammers2013].
-The population on Cousin Island (04'20'S, 55'40'E) has been intensively studied since 1986 [@Richardson2003b; @Barrett2013a].
+The population of *ca*.
+320 birds on Cousin Island (04'20'S, 55'40'E) has been intensively studied since 1986 [@Richardson2003b; @Spurgin2014a].
+This species has two breeding seasons, running from June-August (main breeding season) and December-February (minor breeding season), when females on one of the *ca*.
+115 breeding territories lay one or, rarely, two or three eggs [@Komdeur1991].
+As a result of this low reproductive output, combined with a higher mortality rate among first-year birds [61% survival; @Brouwer2006], cohort sizes in the Seychelles warbler are small (<50).
+
 Individuals are usually ringed in their first year of life, and so are of known age.
-They are then followed throughout their lives, and as they are non-migratory endemics and are naturally confined to the island [@Komdeur2004a], a biannual census of birds on Cousin during the main (June-August) and minor (December-February) breeding seasons gives an accurate measure of levels of local density and individual survival rate [@Barrett2013a].
+They are then followed throughout their lives, and as they are non-migratory endemics naturally confined to the island [@Komdeur2004a], a biannual census of birds on Cousin during each breeding season gives accurate measures of local density and individual survival [@Barrett2013a].
+The isolated nature of the Cousin population is a key advantage of the system for analyses involving survival, which are often confounded by emigration [see @Ergon2014 for a recent discussion].
 Full details of catching and census methods can be found in Brouwer *et al.* [-@Brouwer2012].
 
-All territories are mapped during the main and minor breeding seasons based on detailed observational data of foraging and territorial defence behaviour, and surveyed for territory quality [@Richardson2003b].
-Territory quality is calculated based on territory size, foliage cover and insect abundance [@Komdeur1992] and averaged within territories [@Hammers2013].
-Cousin is also subject to intra- and inter-annual variation in rainfall and food availability, and such island-wide temporal variation may override the effects of absolute territory quality.
-As an estimate of variation in food availability, we calculated an index of the number of insects across the entire island during each breeding season.
+All territories are mapped during the breeding seasons using detailed observational data of foraging and territorial defence behaviour, and surveyed for territory quality [@Richardson2003b].
+Territory quality is calculated based on territory size, foliage cover and insect abundance [@Komdeur1992], and territory quality estimates obtained across years are averaged to obtain a single value for each territory [@Hammers2013].
+Cousin is subject to intra- and inter-annual variation in rainfall and food availability, and such island-wide temporal variation may override the effects of absolute territory quality.
+As an estimate of seasonal variation in food availability, we calculated an index of the number of insects across the entire island during each breeding season.
 This  index represents the average number of insects found per unit leaf area over all surveys carried out on the island in a breeding season.
 
-Each time a bird is caught on Cousin body mass and tarsus length are measured, and age is assessed on the basis of eye colour and previous captures [@Richardson2003b].
-A blood sample (ca 25 $\mu$l) is taken from each bird via brachial venipuncture, and stored in 1 ml absolute ethanol in a 1.5 ml screw-cap microfuge tube at room temperature.
+Each time a bird is caught on Cousin body mass and tarsus length are measured (to the nearest 0.1g and 0.1mm, respectively), and age is assessed on the basis of eye colour and previous captures [@Richardson2003b].
+A blood sample (ca 25 $\mu$l) is taken from each bird via brachial venipuncture, and stored in 1 ml of absolute ethanol in a 1.5 ml screw-cap microfuge tube at room temperature.
 
@@
 
-Telomere lengths were measured from `r nrow(FlSA)+nrow(chicks)` samples taken cross-sectionally from birds aged less than twelve months, between `r min(as.numeric(paste(juv$LayYear)))` and `r max(as.numeric(paste(juv$LayYear)))`.
+Telomere lengths were measured from a total of `r nrow(FlSA)+nrow(chicks)+nrow(Loss)` samples from `r nrow(FlSA)+nrow(chicks)` birds caught within their first year of life, between `r min(as.numeric(paste(juv$LayYear)))` and `r max(as.numeric(paste(juv$LayYear)))`.
 Of these birds, `r nrow(chicks)` were nestlings less than two months old (chicks), `r nrow(subset(FlSA,Ageclass=='FL'))` were aged between two and six months (fledglings), and `r nrow(subset(FlSA,Ageclass=='SA'))` were aged seven to twelve months (subadults).
@@
 We then tested tested how spatially varying early-life environmental and social conditions influence telomere length within cohorts using linear models.
-For these analyses, both response and explanatory variables were standardised by mean-centering within cohorts, thus controlling for any temporal variation in telomere length and the social and ecological environment.
-As explanatory variables we included tarsus length, sex, territory quality and the number of helping and non-helping subordinates present in the natal territory.
+For these analyses, both response and explanatory variables were standardised by mean-centering within cohorts, thus controlling for any between-cohort temporal variation in telomere length and the social and ecological environment.
+As explanatory variables we included tarsus length, sex, territory quality and the number of helping and non-helping subordinate birds present in the natal territory.
 We used model averaging to assess how these variables shape telomere length and dynamics in early life.
@@
 
-For the individuals with longitudinal data we calculated a rate of telomere loss between the first-year sample and earliest adult sample by subtracting the adult telomere length from the early-life telomere length and dividing this difference by the length of time (in days) between sampling events.
+For individuals with longitudinal data we calculated a rate of telomere loss between the first-year sample and earliest adult sample by subtracting the adult telomere length from the early-life telomere length and dividing this difference by the length of time (in days) between sampling events.
 To account for regression to the mean effects, we applied a correction based on correlations among samples within individuals following Verhulst *et al.* [-@Verhulst2013].
@@
 Finally, we analysed the effects of telomere dynamics on survival, again by looking at variation both within and among cohorts.
-First, we tested the prediction that cohort-level telomere dynamics are associated with survival by comparing the mean telomere length and loss of each cohort with the mean lifespan of all individuals, using linear regression.
+First, we tested the prediction that cohort-level telomere dynamics are associated with survival by comparing the mean telomere length and rate of loss of each cohort with the mean lifespan of all individuals, using linear regression.
 For this analysis we excluded any recent breeding seasons where >10% of individuals in a cohort were still alive.
-We then  used parametric survival analysis,implemented in the FlexSurv package in R, to test whether individual-level telomere length and loss were related to to survival.
-Again, we mean-centred variables within cohorts to eliminate temporal variation in telomeres and survival.
+We then  used parametric survival analysis to test whether individual-level telomere length and rate of loss were related to to survival.
+Again, we mean-centred variables within cohorts to eliminate between-season temporal variation in telomeres and survival.
 The survival time distribution was chosen by comparing the fit of a range of distributions (exponential, Weibull, log-normal, gamma, gompertz and generalized F) against the observed survival data, by visually inspecting plots and by comparing AIC values.
@@
 This is lower than the mean for one year-old birds based on cross-sectional data (`r round2(mean(subset(dd,Age==1)$TLKB))` $\pm$ `r round2(se(subset(dd,Age==1)$TLKB))` kb).
-However, when we considered age classes separately, we found substantial age-related variation in telomere length within the first year of life (F = `r round2(anova(lm(TLKB~Ageclass,data = juv))[1,4])`; *P* = `r round2(anova(lm(TLKB~Ageclass,data = juv))[1,5])`).
+However, when we considered age classes within the first year of life separately, we found substantial age-related variation in telomere length (F = `r round2(anova(lm(TLKB~Ageclass,data = juv))[1,4])`; *P* = `r round3(anova(lm(TLKB~Ageclass,data = juv))[1,5])`).
 Chicks had longer telomeres than any other age group, but there was an apparent increase in average telomere length from the fledgling to subadult stages (Fig.
@@
 #####Temporal variation in early life telomere dynamics
-Early-life telomere length varied significantly among breeding seasons (one-way ANOVA, F = `r round2(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,4])`; *P* = `r round3(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,5])`), and this variation in average telomere length over breeding seasons was positively related to island-wide food availability (linear regression, R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,5])`; Fig.
+Early-life telomere length varied significantly among breeding seasons (one-way ANOVA, F = `r round2(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,4])`; *P* `r round3(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,5])`), and this variation in average telomere length over breeding seasons was positively related to island-wide food availability (linear regression, R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,5])`; Fig.
 2A).
@@
 Reproductive senescence occurs in the Seychelles warbler [reviewed in @Hammers2015], making this system well suited to examining how telomere length affects lifetime reproductive success.
-A further avenue for future research will be examining, in a quantitative genetic framework, how genetic and environmental components, and their interactions, affect telomere dynamics and senescence in natural populations [@Becker2015; @Asghar2015a].
+A further avenue for future research will be examining, in a quantitative genetic framework, how genetic and environmental components, and their interactions, affect telomere dynamics and senescence in natural populations [@Becker2015; @Asghar2014].
 We expect that by gaining a fuller understanding of telomere dynamics in natural populations, the fields of life-history evolution and evolutionary ecology will be greatly enhanced.

### 2015-09-22 Scripts/SW juvenile telomeres IPC clean.R

@@ -252,8 +252,8 @@ for(i in 1:nrow(Loss))
 }
 
 Loss <- Loss[!(is.na(Loss$cenTROC)),]
-Loss <- subset(subset(Loss,TROC > -0.01),TROC < 0.01)
-Loss <- subset(Loss,TimeDiff<1500)
+Loss <- subset(subset(Loss,TROC > -0.01),TROC < 0.02)
+Loss <- subset(subset(Loss,TimeDiff<1500),TimeDiff>365)
 
 
 # Get rid of stuff not to be used -----------------------------------------
@@ -279,10 +279,49 @@ juvseason <- ddply(juv,
 juvseason$cenTL <- juvseason$TLKBmean-mean(juvseason$TLKBmean)
 juvseason <- subset(juvseason,n>2)
 
-chickseason <- aggregate(TLKB~Insect,mean,data=chickall)
-flseason <- aggregate(TLKB~Insect,mean,data=subset(FlSAall,Ageclass == 'FL'))
-saseason <- aggregate(TLKB~Insect,mean,data=subset(FlSAall,Ageclass == 'SA')) 
-                      
+chickseason <- ddply(chicks,
+                   .(FieldPeriodID,Season),
+                   summarize,
+                   TLKBmean = mean(TLKB),
+                   TLKBse = se(TLKB),
+                   Tarsus = mean(Tarsus),
+                   Insect = mean(Insect),
+                   Lifespan = mean(RemainingLife),
+                   Lifespanse = se(RemainingLife),
+                   CatchYear = mean(CatchYear),
+                   n = length(TLKB))
+chickseason$cenTL <- chickseason$TLKBmean-mean(chickseason$TLKBmean)
+chickseason <- subset(chickseason,n>2)
+
+
+flseason <- ddply(subset(FlSA,Ageclass == 'FL'),
+                     .(FieldPeriodID,Season),
+                     summarize,
+                     TLKBmean = mean(TLKB),
+                     TLKBse = se(TLKB),
+                     Tarsus = mean(Tarsus),
+                     Insect = mean(Insect),
+                     Lifespan = mean(RemainingLife),
+                     Lifespanse = se(RemainingLife),
+                     CatchYear = mean(CatchYear),
+                     n = length(TLKB))
+flseason$cenTL <- flseason$TLKBmean-mean(flseason$TLKBmean)
+flseason <- subset(flseason,n>2)
+
+saseason <- ddply(subset(FlSA,Ageclass == 'SA'),
+                  .(FieldPeriodID,Season),
[…]

### 2015-09-22 Scripts/SW juvenile telomeres IPC functions.r

@@ -79,11 +79,10 @@ correctedimp <- function(importance,baselineimportance)
 
 
 
-
 #Combine summary table from modelavereging with relative importance values
 #to produce a publication-ready table
 imptable <- function(modelavg,imp)
-{  temp <- cbind(summary(modelavg)$coefmat[,1:2],confint(modelavg))
+{  temp <- cbind(summary(modelavg)$coefmat.subset[,1:2],confint(modelavg))
    temp <- temp[rownames(temp) != '(Intercept)',]
    temp <- cbind(temp,rep(NA,nrow(temp)))
    colnames(temp)[5] <- 'imp'
@@ -192,25 +191,25 @@ pganchor <- local({
     }
   }
 })
-
-knit_hooks$set( anchor = function(before, options, envir) {
-  if ( before ) {
-    sprintf('<a name="%s"/>\n', options$label )
-  }
-})
-
-knit_hooks$set( echo.label = function(before, options, envir) {
-  if ( before ) {
-    sprintf('> %s', options$label )
-  }
-})
-
-knit_hooks$set( pgbreak = function(before, options, envir) {
-  if ( !before ) {
-    pganchor();
-  }
-})
-
+# 
+# knit_hooks$set( anchor = function(before, options, envir) {
+#   if ( before ) {
+#     sprintf('<a name="%s"/>\n', options$label )
+#   }
+# })
+# 
+# knit_hooks$set( echo.label = function(before, options, envir) {
+#   if ( before ) {
+#     sprintf('> %s', options$label )
+#   }
+# })
+# 
+# knit_hooks$set( pgbreak = function(before, options, envir) {
+#   if ( !before ) {
+#     pganchor();
+#   }
+# })
+# 
 
 
 
@@ -439,9 +438,9 @@ round2 <- function(x)
   }
   }
 
-round3 <- function(x)
+round3 <- function(x,lessthan = T)
 {
-  if(x < 0.001)
+  if(lessthan == T && x < 0.001)
   {
     return('< 0.001')
   } else

### 2015-09-22 Manuscript/SW_early_life_telomeres_v9.Rmd
More updates

@@
 1B).
-This suggests that the apparent increase in telomere length between fledglings and subadults in the cross-sectional data was a result of the sampling, rather than telomere elongation.
+This suggests that the apparent increase in telomere length between fledglings and subadults in the cross-sectional data was a result of the sampling and selective mortality, rather than telomere elongation.
 
@@
 
-The rate of telomere shortening in early life did not vary significantly among breeding seasons (F = `r round2(anova(lm(TROC~factor(FieldPeriodID),data = Loss))[1,4])`; *P* = `r round2(anova(lm(TROC~factor(FieldPeriodID),data = Loss))[1,5])`); however, what variation there was in rates of telomere shortening could be explained by temporal variation in food availability (R^2^ = `r round2(summary(lm(Insect~TROCmean,data=loss.season,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TROCmean,data = loss.season,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Insect~TROCmean,data = loss.season,na.action=na.exclude))[1,5])`).
+The rate of telomere shortening in early life (taken from the smaller subsert of birds with longitudinal samples) did not vary significantly among breeding seasons (F = `r round2(anova(lm(TROC~factor(FieldPeriodID),data = Loss))[1,4])`; *P* = `r round2(anova(lm(TROC~factor(FieldPeriodID),data = Loss))[1,5])`); however, what variation there was in rates of telomere shortening could be explained by temporal variation in food availability (R^2^ = `r round2(summary(lm(Insect~TROCmean,data=loss.season,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TROCmean,data = loss.season,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Insect~TROCmean,data = loss.season,na.action=na.exclude))[1,5])`).
 This pattern corresponded with the cross-sectional data: individuals born in seasons with low food availability lost telomeres, on average, at a faster rate than those born in years with high food availability (Fig.
@@
 For the subset of individuals with longitudinal data, we tested whether the factors previously identified as being related to early-life telomere length (i.e.
-tarsus length, sex and number of helpers) were also associated with differences in rates of early life telomere loss.
+tarsus length and number of helpers) were also associated with differences in rates of early life telomere loss.
 Due to a limited sample size, we considered all age classes together for this analysis.
@@
 Testing the latter relationship separately for each age class revealed that there was no age-specific relationship between cohort-level telomere length and lifespan (all *P* > 0.2).
-There was a negative relationship between population-level telomere loss and lifespan, although this was marginally non-significant (R^2^ = `r round3(summary(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))[1,5])`; Fig.
+There was, however, a negative relationship between population-level telomere loss and lifespan, although this was marginally non-significant (R^2^ = `r round3(summary(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))[1,5])`; Fig.
 5B).
@@
 For the individual-based survival analysis, a log-normal survival model best fitted the Seychelles warbler data (AIC = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="lnorm")$AIC)`, AIC of next best distribution (Weibull) = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="weibull")$AIC)`).
-A survival model including all first year birds, early-life telomere length was not associated with survival (estimate = `r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`).
+A survival model including all first year birds showed that early-life telomere length was not associated with survival (estimate = `r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`).
 However, we did find differential survival effects among age classes.
 Chick telomere length had no effect on survival  (estimate = `r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`; Fig.
-6A).
-However, telomere length at both the fledgling and subadult stage was significantly related to survival (fledglings: estimate = `r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'FL'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'FL'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'FL'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`; subadults: estimate = `r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'SA'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'SA'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'SA'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`).
+6A), but telomere length at both the fledgling and subadult stage was significantly related to survival (fledglings: estimate = `r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'FL'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'FL'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'FL'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`; subadults: estimate = `r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'SA'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'SA'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'SA'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`).
 In both cases this effect was positive, with longer telomeres in early life associated with increased survival later in life (Figs 6B, 6C).

### 2015-09-23 Manuscript/SW_early_life_telomeres_v9.Rmd
Intro edits

@@
 The longitudinal study (since 1986) of the Seychelles warbler (*Acrocephalus sechellensis*) population on Cousin Island provides an excellent model system for studying senescence in the wild [reviewed in @Hammers2015].
-Specifically, it is ideally suited for assessing the relationship between early-life environmental variation, telomere length and late-life fitness.
-Firstly, due to the isolated nature of the study population and intensive field monitoring, we have excellent survival data and sampling material spanning many years (see methods, below).
-Secondly, telomere length predicts survival independently of age in adult Seychelles warblers, suggesting that telomeres act as a biomarker of cost in this species [@Barrett2013a].
-Finally, ecological conditions on Cousin are variable over space and time due to weather-induced changes in foliage cover and food availability [@VandeCrommenacker2011a].
+Due to the isolated nature of the study population and intensive field monitoring, we have excellent survival data and sampling material spanning many years (see methods, below).
+Ecological conditions an population density on Cousin are variable over space and time due to weather-induced changes in foliage cover and food availability [@VandeCrommenacker2011a].
+Variation in natal territory quality has been linked to changes in levels of oxidative stress [@VandeCrommenacker2011a], but neither natal territory quality nor natal local density appear to be linked to early-life or adult survival [@Brouwer2006; @Hammers2013].
+Cooperative breeding occurs in the Seychelles warbler [@Komdeur1994a; @Richardson2003a], and the presence of helping subordinates (but not non-helping subordiantes) in the natal territory has been associated with increased survival later in life [@Brouwer2012].Finally, we have an established protocol for assessing absoltute telomere length in this species [@Barrett2012], and telomere length predicts survival independently of age in adult Seychelles warblers, suggesting that telomeres act as a biomarker of cost [@Barrett2013a].
+Thus, we have an excellent system in which to assess the costs of different social and environmental conditions experienced early in life, and to assess the later life consequences of early-life conditions.
 
-In Seychelles warblers, oxidative stress is linked to natal territory quality [@VandeCrommenacker2011a], but neither natal territory quality nor natal local density appear to be linked to early-life or adult survival [@Brouwer2006; @Hammers2013].
-Importantly, cooperative breeding occurs in the Seychelles warbler [@Komdeur1994a; @Richardson2003a], and the presence of helping subordinates (but not non-helping subordiantes) in the natal territory has been associated with increased survival later in life [@Brouwer2012].
 In this study we test the prediction that telomeres link early-life environmental variation to late-life fitness.
-Because telomere length and shortening will reflect individual-level variation in the costs of early-life experiences, they may allow a more sensitive analysis of the effects of early early-life environmental variation than would be possible with a direct comparison of the early-life environment and survival.
-With this in mind, we predict that: i) individuals raised in good quality conditions will have longer telomeres, and ii) having longer telomeres in early life is associated with greater short and long-term survival.
+Because telomere dy6naimcs are expected to reflect individual-level variation in the costs of early-life experiences, they may allow a more sensitive analysis of the effects of early early-life environmental variation than would be possible with a direct comparison of the early-life environment and survival.
+With this in mind, we predict that: i) individuals raised in good quality conditions will have longer telomeres, and ii) having longer telomeres in early life is associated with greater survival.
 
@@
 Telomere length was positively related to the number of helpers (Fig.
-3B), and negatively related to tarus length (Fig.
+3B), and negatively related to tarsus length (Fig.
 3C).

### 2015-09-23 Manuscript/SW_early_life_telomeres_SOM_v9.Rmd
Changes before sending to DSR
Updates before sending to Kat

@@
 ---
-output: word_document
+output:
+  word_document:
+    reference_docx: myref.docx
 ---
@@
 #Run the model
-lm1 <- lm(cenTL ~ cenHelper + cenNonHelper + Sex + cenTarsus  + cenTQ,
-            data = chickall)
+lm1 <- lm(cenTL ~ Fledged*(cenHelper + cenNonHelper + Sex + cenTarsus  + cenTQ),
+            data = juvall)
 
@@
 nams[nams == 'cenHelper'] <- 'Helpers'
+nams[nams == 'cenHelper:Fledged'] <- 'Ageclass x Helpers'
+nams[nams == 'cenNonHelper:Fledged'] <- 'Ageclass x Non helpers'
+nams[nams == 'cenTarsus:Fledged'] <- 'Ageclass x Tarsus'
+nams[nams == 'cenTQ:Fledged'] <- 'Ageclass x TQ'
+nams[nams == 'Fledged:SexMales'] <- 'Ageclass x Sex'
+nams[nams == 'Fledged'] <- 'Ageclass'
 
@@
 
-tableS1 <- data.frame(Model = apply(t1,1,getterms,c(2:6)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
+tableS1 <- data.frame(Model = apply(t1,1,getterms,c(2:12)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
 colnames(tableS1) <- c('Model','df','AICc', 'Delta AICc','Weight')
@@
 ```
-
-PAGEBREAK
-
-**Table S2** Model selection table for analysis of factors affecting early life telomere length in Seychelles warbler fledglings.
-Only models that remained in the top model set for model averaging (AICc $\leq$ 6 compared to the top model) are included here.
-
-```{r Table S2}
-
-temp <- subset(FlSAall,Ageclass == 'FL')
-
-#Run the model
-lm1 <- lm(cenTL ~ cenHelper + cenNonHelper + Sex + cenTarsus  + cenTQ,
-            data = temp)
-
-#Model averaging
-lm2 <- dredge(lm1)
-t1 <- as.matrix(subset(lm2, delta<=6))
-
-#Sort out row/col names
-nams <- colnames(t1)
-nams[nams == 'cenTarsus'] <- 'Tarsus'
-nams[nams == 'cenNonHelper'] <- 'Non helpers'
-nams[nams == 'SexMales'] <- 'Sex'
-nams[nams == 'cenTQ'] <- 'TQ'
-nams[nams == 'cenHelper'] <- 'Helpers'
-
-colnames(t1) <- nams
-
-tableS2 <- data.frame(Model = apply(t1,1,getterms,c(2:6)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
-colnames(tableS2) <- c('Model','df','AICc', 'Delta AICc','Weight')
-rownames(tableS2) <- NULL
-tableS2$Model[tableS2$Model == ''] <- 'Null model'
-tableS2[,3:5] <- lapply(tableS2[,3:5],as.numeric)
-kable(tableS2,digits=2)
-
-```
-
-**Table S3** Model selection table for analysis of factors affecting early life telomere length in Seychelles warbler sub-adults Only models that remained in the top model set for model averaging (AICc $\leq$ 6 compared to the top model) are included here.
-
-```{r Table S3}
-
-temp <- subset(FlSAall,Ageclass == 'SA')
-
-#Run the model
-lm1 <- lm(cenTL ~ cenHelper + cenNonHelper + Sex + cenTarsus  + cenTQ,
-            data = temp)
-
-#Model averaging
-lm2 <- dredge(lm1)
-t1 <- as.matrix(subset(lm2, delta<=6))
-
-#Sort out row/col names
-nams <- colnames(t1)
-nams[nams == 'cenTarsus'] <- 'Tarsus'
-nams[nams == 'cenNonHelper'] <- 'Non helpers'
-nams[nams == 'SexMales'] <- 'Sex'
-nams[nams == 'cenTQ'] <- 'TQ'
-nams[nams == 'cenHelper'] <- 'Helpers'
-
-colnames(t1) <- nams
-
-tableS3 <- data.frame(Model = apply(t1,1,getterms,c(2:6)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
-colnames(tableS3) <- c('Model','df','AICc', 'Delta AICc','Weight')
-rownames(tableS3) <- NULL
-tableS3$Model[tableS3$Model == ''] <- 'Null model'
-tableS3[,3:5] <- lapply(tableS3[,3:5],as.numeric)
-kable(tableS3,digits=2)
-
-```

### 2015-09-23 Manuscript/SW_early_life_telomeres_tables_and_figures v9.Rmd
Updates before sending to Kat

@@
 
-**Figure 1** Telomere length (mean $\pm$ s.e.) and age Seychelles warblers, based on cross-sectional **(A)** and longitudinal **(B)** data.
-In **B**, telomere loss is a standardised and corrected rate (see methods for details) based on the difference in telomere length from the sample taken from birds in their first year of life, and the first available sampling as an adult.
+**Figure 1** Telomere length (mean $\pm$ s.e.) and age in Seychelles warblers, based on cross-sectional **(A)** and longitudinal **(B)** data.
+In **B**, telomere loss is a standardised and corrected rate (see methods for details) based on the difference in telomere length from the sample taken from birds in their first year of life, and the first available sample as an adult.
 Values above zero indicate telomere shortening in early life.
@@
 Points and error bars represent mean and standard error telomere length for all birds born in each main (black) or minor (grey) breeding season, in relation to island-wide food availability.
-Trend lines represent fitted values from a linear regression.
+Lines and shaded areas represent fitted values and 95% confidence limits from a linear regression.
 
-**Figure 3** Factors affecting telomere length in Seychelles warbler chicks.
-**A** Model averaged estimates and 95% confidence intervals for all explanatory terms used in a linear model with telomere length as the response variable.
+**Figure 3** Factors affecting early-life telomere length in Seychelles warblers at the individual level, within seasons.
+**A** Model averaged estimates and 95% confidence intervals for alxplanatory terms used in a linear model with telomere length as the response variable.
 Numbers in brackets are the relative importance of each term in the top model set (see main text for details; TQ = territory quality).
-**B** Presence/absence of helpers and **c** tarsus length in relation to raw telomere length.
+**B** Presence/absence of helpers and **c** tarsus length in relation to raw telomere length in nestlings (black, solid lines) and fledglings (grey, dashed lines).
 Points and error bars in **B** are mean $\pm$ s.e., and the line and shaded areas from **C** represent fitted values and 95% confidence limits from a linear regression.
 
-**Figure 4** Factors affecting telomere length in Seychelles warblers fledglings **(A)** and subadults **(B)**.
-Points and error bars represent model averaged estimates and 95% confidence intervals for all explanatory terms used in a linear model with  telomere length as the response variable.
-Numbers in brackets are the relative importance of each term in the top model set (see main text for details; TQ = territory quality).
-
-**Figure 5** Temporal, cohort-level variation in early-life telomere dynamics and lifespan in the Seychelles warbler, based on cross-sectional **(A)** and longitudinal **(B)** data (see Fig.
+**Figure 4** Temporal, cohort-level variation in early-life telomere dynamics and lifespan in the Seychelles warbler, based on cross-sectional **(A)** and longitudinal **(B)** data (see Fig.
 1 and main text for details).
 Points and error bars represent means and $\pm$ s.e., respectively, of all birds born in each main (black) or minor (grey) breeding season.
+Lines and shaded areas represent fitted values and 95% confidence limits from a linear regression.
 
-**Figure 6** Kaplan-Meier curves showing the relationship between telomere length and survival in chick (**A**), fledgling  (**B**) and sub-adult (**C**) Seychelles warblers.
+**Figure 5** Kaplan-Meier curves showing the relationship between telomere length and survival for individual Seychelles warblers measured as chicks (**A**), and fledglings  (**B**).
 Telomere length is binned into groups here for visualisation purposes only (long and short = greater than or less than median telomere length, repectively).
@@
 
-temp <- dd
-levels(temp$Ageclass) <- c('Chicks',rep('Fledglings',3),'Sub-adults','Adults')
-
-ddFig1A <- ddply(temp,
-                 .(Ageclass),
+ddFig1A <- ddply(dd,
+                 .(Fledged),
                  summarize,
@@
 Fig1A <- ggplot(ddFig1A,
-       aes(x = Ageclass,
+       aes(x = Fledged,
            y = TLKBmean)) +
@@
 
-temp <- Loss
-levels(temp$Ageclass) <- c('Chicks','Fledglings','Sub-adults')
-
-ddFig1B <- ddply(temp,
-                 .(Ageclass),
+ddFig1B <- ddply(Loss,
+                 .(Fledged),
                  summarize,
@@
 Fig1B <- ggplot(ddFig1B,
-       aes(x = Ageclass,
+       aes(x = Fledged,
            y = TROCmean)) +
@@
   geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse)) +
-  stat_smooth(method = 'lm',se = F,col = 'black') +
+  stat_smooth(method = 'lm',se = T,col = 'black') +
   scale_color_manual(values = c('black',grey(0.4))) +
@@
         legend.position = 'none') +
-  stat_smooth(method = 'lm',se = F,  col = 'black') +
+  stat_smooth(method = 'lm',se = T,  col = 'black') +
   scale_color_manual(values = c('black',grey(0.4))) +
@@
 #Run the model
-lm2 <- lm(cenTL ~ cenHelper + cenNonHelper + Sex + cenTarsus  + cenTQ,
-            data = chickall)
+lm2 <- lm(cenTL ~ Fledged*(cenHelper + cenNonHelper + Sex + cenTarsus  + cenTQ),
+            data = juvall)
 
@@
 nams[nams == 'cenHelper'] <- 'Helpers'
+nams[nams == 'cenHelper:FledgedFledglings'] <- 'Ageclass x Helpers'
+nams[nams == 'cenNonHelper:FledgedFledglings'] <- 'Ageclass x Non helpers'
+nams[nams == 'cenTarsus:FledgedFledglings'] <- 'Ageclass x Tarsus'
+nams[nams == 'cenTQ:FledgedFledglings'] <- 'Ageclass x TQ'
+nams[nams == 'FledgedFledglings:SexMales'] <- 'Ageclass x Sex'
+nams[nams == 'FledgedFledglings'] <- 'Ageclass'
 
@@
 Fig3A <- forestplot(table2) +
-  annotate('text',label = 'A',x=0.6,y=0.6)
+  ylim(c(-1,1)) +
+  annotate('text',label = 'A',x=0.6,y=1)
 
@@
 
-temp <- chicks
-temp$HelperF <- ifelse(chicks$Helper>0,'Helpers present','Helpers absent')
+temp <- juv
+temp$HelperF <- ifelse(juv$Helper>0,'Helpers present','Helpers absent')
 
 ddFig3B <- ddply(temp,
-                 .(HelperF),
+                 .(HelperF,Fledged),
                  summarize,
@@
        aes(x = HelperF,
-           y = TLKBmean)) +
+           y = TLKBmean,
+           col = Fledged,
+           lty = Fledged)) +
   ylab('Telomere length (kb)') + 
@@
         axis.title.x = element_text(size = 16),
-        legend.position = c(0.1,0.9),
+        legend.position = 'none',
         legend.title = element_blank(),
         legend.text = element_text(size = 15)) +
-  geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse)) +
+  geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse),
+                  position = position_dodge(width = 0.2)) +
+  scale_color_manual(values = c('black',grey(0.4))) +
   annotate('text',label = 'B',x=0.5,y=8)
@@
 
-Fig3C <- ggplot(chicks,
+Fig3C <- ggplot(juv,
                 aes(x = Tarsus,
-                    y = TLKB)) +
+                    y = TLKB,
+                    col=Fledged,
+                    lty = Fledged)) +
   ylab('Telomere length (kb)') + 
@@
         axis.title.y = element_text(size = 16,vjust=0.8),
-        axis.title.x = element_text(size = 16)) +
+        axis.title.x = element_text(size = 16),
+        legend.position = 'none') +
   geom_point(size=3) +
-  stat_smooth(method = 'lm',se = F,col = 'black') +
+  stat_smooth(method = 'lm',se = T) +
   scale_color_manual(values = c('black',grey(0.4))) +
@@
 
-```{r Figure 4,fig.width=5,fig.height=9}
+```{r Figure 4,fig.height=8,fig.width=4}
 
-######################
-#3A
-######################
-Fl <- subset(FlSAall,Ageclass == 'FL')
-
-#Run the model
-lm2 <- lm(cenTL ~ cenHelper + cenNonHelper + Sex + cenTarsus  + cenTQ,
-            data = Fl)
-
-#Model averaging
-lm3 <- dredge(lm2)
-lmavg <- model.avg(lm3,subset = delta <=6)
-myimp <- importance(lmavg) #Relative importance of each explanatory variable
-
-#Put it all into a nice table
-table2 <- imptable(lmavg,myimp)
-table2 <- signif(table2,2)
-
-nams <- rownames(table2)
-nams[nams == 'cenTarsus'] <- 'Tarsus'
-nams[nams == 'cenNonHelper'] <- 'Non helpers'
-nams[nams == 'SexMales'] <- 'Sex'
-nams[nams == 'cenTQ'] <- 'TQ'
-nams[nams == 'cenHelper'] <- 'Helpers'
-
-rownames(table2) <- nams
-colnames(table2) <- c('Estimate','SE','2.5%','97.5%','RI')
-
-table2 <- data.frame(table2,myx=row.names(table2))
-
-Fig4A <- forestplot(table2) +
-  annotate('text',label = 'A',x=0.6,y=0.8)
-
-######################
-#4B
-######################
-SA <- subset(FlSAall,Ageclass == 'SA')
-
-#Run the model
-lm2 <- lm(cenTL ~ cenHelper + cenNonHelper + Sex + cenTarsus  + cenTQ,
-            data = SA)
-
-#Model averaging
-lm4 <- dredge(lm2)
-lmavg <- model.avg(lm4,subset = delta <=6)
-myimp <- importance(lmavg) #Relative importance of each explanatory variable
-
-#Put it all into a nice table
-table2 <- imptable(lmavg,myimp)
-table2 <- signif(table2,2)
-
-nams <- rownames(table2)
-nams[nams == 'cenTarsus'] <- 'Tarsus'
-nams[nams == 'cenNonHelper'] <- 'Non helpers'
-nams[nams == 'SexMales'] <- 'Sex'
-nams[nams == 'cenTQ'] <- 'TQ'
-nams[nams == 'cenHelper'] <- 'Helpers'
-
-rownames(table2) <- nams
-colnames(table2) <- c('Estimate','SE','2.5%','97.5%','RI')
-
-table2 <- data.frame(table2,myx=row.names(table2))
-
-Fig4B <- forestplot(table2) +
-  annotate('text',label = 'B',x=0.6,y=0.6)
-
-multiplot(Fig4A,Fig4B)
-
-```
-
-PAGEBREAK
-
-**Figure 5**
-
-```{r Figure 5,fig.height=8,fig.width=4}
-
-Fig5A <- ggplot(juvseason,
+Fig4A <- ggplot(juvseason,
                 aes(x = TLKBmean,
@@
   scale_color_manual(values = c('black',grey(0.4))) +
-  annotate('text',x=2.5,y=12.5,label='A')
+  annotate('text',x=2.5,y=10,label='A')
 
-Fig5B <- ggplot(loss.season,
+Fig4B <- ggplot(loss.season,
                 aes(x = TROCmean,
@@
   scale_color_manual(values = c('black',grey(0.4))) +
-  stat_smooth(method = 'lm',se = F,col = 'black',fullrange = T) +
-  annotate('text',x=-0.004,y=14,label='B')
+  stat_smooth(method = 'lm',se = T,col = 'black',fullrange = T) +
+  annotate('text',x=-0.004,y=12.5,label='B')
 
-multiplot(Fig5A,Fig5B)
+multiplot(Fig4A,Fig4B)
 
@@
 
-**Figure 6**
+**Figure 5**
 
-```{r Figure 6,fig.height=9,fig.width=4}
+```{r Figure 5,fig.height=8,fig.width=4}
 
 s1 <- survfit(Surv(time = Lifespan+1, event = Died)~TLF, data = chicks)
-Fig6A <- ggsurv(s1,plot.cens=F) + 
+Fig5A <- ggsurv(s1,plot.cens=F) + 
   theme_classic() +
@@
   xlab('Age (years)') + 
+  xlim(c(0,15)) +
   theme_classic() +
@@
 ###########################################
-s2 <- survfit(Surv(time = Lifespan+1, event = Died)~TLF, data = subset(FlSAall,Ageclass == 'FL'))
-Fig6B <- ggsurv(s2,plot.cens=F) + 
+s2 <- survfit(Surv(time = Lifespan+1, event = Died)~TLF, data = FlSAall)
+Fig5B <- ggsurv(s2,plot.cens=F) + 
   theme_classic() +
@@
 
-###########################################
-
-s2 <- survfit(Surv(time = Lifespan+1, event = Died)~TLF, data = subset(FlSAall,Ageclass == 'SA'))
-Fig6C <- ggsurv(s2,plot.cens=F) + 
-  theme_classic() +
-  ylab('Proportion survived') + 
-  xlab('Age (years)') + 
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 14), 
-        axis.text.y = element_text(size = 14), 
-        axis.title.y = element_text(size = 16,vjust=0.8),
-        axis.title.x = element_text(size = 16),
-        legend.position = 'none',
-        legend.text = element_text(size = 12),
-        legend.title = element_blank()) +
-  scale_colour_manual(name = 'TL',
-                      values = c('black',grey(0.4))) +
-  guides(linetype=F) +
-  annotate('text',x=0.5,y=1.1,label='C')
-
-multiplot(Fig6A,Fig6B,Fig6C)
+multiplot(Fig5A,Fig5B)
 

### 2015-09-23 Manuscript/SW_early_life_telomeres_v9.Rmd
Updates before sending to Kat

@@
   word_document:
-    fig_caption: yes
-    highlight: null
     reference_docx: myref.docx
@@
 csl: JAE.csl
+
 ---
@@
 Individuals born in breeding seasons with high food availability have reduced telomere shortening compared to individuals born in years with low food availability.
-Within cohorts, we found telomere length in chicks was related to tarsus length (which likely represents chick ), and to the number of helpers present in the natal territory.
-However, the number of helpers was not related to telomere shortening rates, nor to telomere length in fledged birds.
+Within cohorts, we found telomere length in nestlings was nrgatively related to tarsus length (which likely represents nestling crowth), and positvely related to the number of helpers present in the natal territory.
+However, neither tarsus length nor the number of helpers was related to telomere length in fledged birds.
 Finally, we found that longer telomeres and reduced telomere shortening rates in early life were associated with increased survival later in life.
@@
 Telomere lengths were measured from a total of `r nrow(FlSA)+nrow(chicks)+nrow(Loss)` samples from `r nrow(FlSA)+nrow(chicks)` birds caught within their first year of life, between `r min(as.numeric(paste(juv$LayYear)))` and `r max(as.numeric(paste(juv$LayYear)))`.
-Of these birds, `r nrow(chicks)` were nestlings less than two months old (chicks), `r nrow(subset(FlSA,Ageclass=='FL'))` were aged between two and six months (fledglings), and `r nrow(subset(FlSA,Ageclass=='SA'))` were aged seven to twelve months (subadults).
+Of these birds, `r nrow(chicks)` were nestlings less than two months old (chicks), while the remaining `r nrow(FlSA)` were fledglings and subadults aged between two and twelve months (hereafter reffered to as fledglings).
 For a subset of first-year birds (n = `r nrow(Loss)` individuals) we had longitudinal data, with samples from individuals caught both in their first year and then again within two years.
@@
 First, to analyse whether temporal variation in food availability drives population-level variation in telomere dynamics, we calculated average telomere length among all birds born in a given breeding season (i.e.
-a cohort).
+a cohort), excluding cohorts where we had samples from $\leq$ 5 birds.
 We tested for a difference in telomere length among cohorts using a one-way ANOVA, and then used linear regression to test whether variation in mean telomere length among cohorts is explained by temporal variation in insect abundance.
 
-We then tested tested how spatially varying early-life environmental and social conditions influence telomere length within cohorts using linear models.
+We then individual-level analyses to test how spatially varying early-life environmental and social conditions influence telomere length within cohorts using linear models.
 For these analyses, both response and explanatory variables were standardised by mean-centering within cohorts, thus controlling for any between-cohort temporal variation in telomere length and the social and ecological environment.
-As explanatory variables we included tarsus length, sex, territory quality and the number of helping and non-helping subordinate birds present in the natal territory.
+As explanatory variables we included age class (nestling or fledgling), tarsus length, sex, territory quality and the number of helping and non-helping subordinate birds present in the natal territory.
+We also included interactions between age class and all the other variables.
 We used model averaging to assess how these variables shape telomere length and dynamics in early life.
@@
 To account for regression to the mean effects, we applied a correction based on correlations among samples within individuals following Verhulst *et al.* [-@Verhulst2013].
-Using this dataset we repeated the analyses of among and within cohort telomere dynamics, replacing telomere length with telomere loss as the response variable.
+Using this dataset we repeated the above analyses of among and within cohort telomere dynamics, replacing telomere length with telomere loss as the response variable.
 
@@
 This is lower than the mean for one year-old birds based on cross-sectional data (`r round2(mean(subset(dd,Age==1)$TLKB))` $\pm$ `r round2(se(subset(dd,Age==1)$TLKB))` kb).
-However, when we considered age classes within the first year of life separately, we found substantial age-related variation in telomere length (F = `r round2(anova(lm(TLKB~Ageclass,data = juv))[1,4])`; *P* = `r round3(anova(lm(TLKB~Ageclass,data = juv))[1,5])`).
-Chicks had longer telomeres than any other age group, but there was an apparent increase in average telomere length from the fledgling to subadult stages (Fig.
+However, when we considered age classes within the first year of life separately, we found substantial age-related variation in telomere length (F = `r round2(anova(lm(TLKB~Fledged,data = juv))[1,4])`; *P* = `r round3(anova(lm(TLKB~Fledged,data = juv))[1,5])`).
+Nestlings had longer telomeres than any other age group, but there was an apparent increase in average telomere length from the fledgling to adult stages (Fig.
 1A).
-However, a longitudinal analysis of telomere loss showed that chicks, fledglings and subadults all lost telomeres early in life, but that there was no difference in the rate of telomere loss among age classes (F = `r round2(anova(lm(TROC~Ageclass,data = Loss))[1,4])`; *P* = `r round2(anova(lm(TROC~Ageclass,data = Loss))[1,5])`; Fig.
+A longitudinal analysis of telomere loss showed that nestlings and fledglings both telomeres early in life, but that there was no difference in the rate of telomere loss among age classes (F = `r round2(anova(lm(TROC~Fledged,data = Loss))[1,4])`; *P* = `r round2(anova(lm(TROC~Fledged,data = Loss))[1,5])`; Fig.
 1B).
-This suggests that the apparent increase in telomere length between fledglings and subadults in the cross-sectional data was a result of the sampling and selective mortality, rather than telomere elongation.
 
@@
 2A).
-Running the regression of cohort-level telomere length and food availability separately for each age class revealed positive but non-significant relationships in fledglings (R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=flseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data=flseason,na.action=na.exclude))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKBmean,data=flseason,na.action=na.exclude))[1,5])`) and subadults (R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=saseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data=saseason,na.action=na.exclude))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKBmean,data=saseason,na.action=na.exclude))[1,5])`), but a non-significant *negative* relationship in chicks (R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=chickseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data=chickseason,na.action=na.exclude))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKBmean,data=chickseason,na.action=na.exclude))[1,5])`).
+Running the regression of cohort-level telomere length and food availability separately for each age class revealed a positive, significant relationship in fledglings (R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=flseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data=flseason,na.action=na.exclude))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKBmean,data=flseason,na.action=na.exclude))[1,5])`), but no relationship in nestlings (R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=chickseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data=chickseason,na.action=na.exclude))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKBmean,data=chickseason,na.action=na.exclude))[1,5])`).
 
@@
 2B).
-Unfortunately our limited sample size within the longitudinal dataset precluded us from running this analysis separately for each age class.
 
 #####Spatial variation in early life telomere dynamics
-The top model explaining within-season variation in chick telomere length contained tarsus length and the number of helpers present in the natal territory (Table S1), and this model was a much better fit than the null model ($\Delta$AICc = `r round2(AICc(lm(cenTL~1,data=chickall))-AICc(lm(cenTL~Tarsus+Helper,data=chickall)))`).
+The top model explaining within-season variation in early-life telomere length contained age class, tarsus length and the number of helpers present in the natal territory, as well as interactions between age class and both tarsus length and the number of helpers (Table S1).
+This model was much better supported than the null model ($\Delta$AICc = `r round2(AICc(lm(cenTL~1,data=juvall))-AICc(lm(cenTL~Fledged*(cenTarsus+cenHelper),data=juvall)))`).
 Both tarsus length and the number of helpers had high relative importance in the top model set (Fig.
-3A), and the model containing these variables explained a reasonable amount of variation in telomere length (R^2^ = `r round2(summary(lm(cenTL~cenTarsus+cenHelper,data=chickall))$r.squared)`).
-Telomere length was positively related to the number of helpers (Fig.
-3B), and negatively related to tarsus length (Fig.
+3A), and the model containing these variables explained a reasonable amount of variation in telomere length (R^2^ = `r round2(summary(lm(cenTL~Ageclass*(cenTarsus+cenHelper),data=juvall))$r.squared)`).
+Telomere length was positively related to the number of helpers in nestlings, but not in fledglings (Fig.
+3B).
+Tarsus length was negatively related to telomere length in nestlings, but not in fledglings (Fig.
 3C).
-
-For fledglings, the null model was the top model explaining early-life telomere length (Table S1).
-None of the explanatory terms had high importance in the top model set, and confidence intervals of all estimates were close to zero (Fig.
-4A), suggesting that none of the variables analysed had a significant effect on fledgling telomere length.
-For subadults, the top model contained territory quality (Table S1), which was negatively related to subadult telomere length (Fig.
-4B).
-However, territory quality explained a tiny amount of variation in subadult telomere length (R^2^ = `r round2(summary(lm(cenTL~cenTQ,data=subset(FlSAall,Ageclass == 'SA')))$r.squared)`), and there was only a small difference in AIC between the model containing territory quality and the null model ($\Delta$AICc = `r round2(AICc(lm(cenTL~1,data=subset(FlSAall,Ageclass == 'SA')))-AICc(lm(cenTL~cenTQ,data=subset(FlSAall,Ageclass == 'SA'))))`).
-Moreover, the confidence intervals of the territory quality effect overlapped with zero (Fig.
-4B), suggesting that any negative effect of territory quality on subadult telomere length was negligible.
 
@@
 #####Early life telomere dynamics and survival
-At the population-level, we found no evidence that individuals born in breeding seasons with high food availability had longer lifespans (R^2^ = `r round2(summary(lm(Lifespan~Insect,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Lifespan~Insect,data=juvseason,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Lifespan~Insect,data=juvseason,na.action=na.exclude))[1,5])`), nor that individuals born in years where average telomere length was longer had longer lifespans (R^2^ = `r round3(summary(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))[1,5])`; Fig.
-5A).
-Testing the latter relationship separately for each age class revealed that there was no age-specific relationship between cohort-level telomere length and lifespan (all *P* > 0.2).
-There was, however, a negative relationship between population-level telomere loss and lifespan, although this was marginally non-significant (R^2^ = `r round3(summary(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))[1,5])`; Fig.
-5B).
+At the population-level, there was a positive but non-significant relationship between telomere length and lifespan (R^2^ = `r round3(summary(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))[1,5])`; Fig.
+4A).
+Testing this relationship separately for each age class revealed that there was no age-specific relationship between cohort-level telomere length and lifespan (both *P* > 0.2).
+There was a negative relationship between population-level telomere loss and lifespan, although this was marginally non-significant (R^2^ = `r round3(summary(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))[1,5])`; Fig.
+4B).
 
 For the individual-based survival analysis, a log-normal survival model best fitted the Seychelles warbler data (AIC = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="lnorm")$AIC)`, AIC of next best distribution (Weibull) = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="weibull")$AIC)`).
-A survival model including all first year birds showed that early-life telomere length was not associated with survival (estimate = `r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`).
-However, we did find differential survival effects among age classes.
-Chick telomere length had no effect on survival  (estimate = `r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`; Fig.
-6A), but telomere length at both the fledgling and subadult stage was significantly related to survival (fledglings: estimate = `r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'FL'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'FL'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'FL'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`; subadults: estimate = `r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'SA'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'SA'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'SA'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`).
-In both cases this effect was positive, with longer telomeres in early life associated with increased survival later in life (Figs 6B, 6C).
+A survival model including all first year birds showed that early-life telomere length had a positive effect on survival, but this relationship was marginally non-significant (estimate = `r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`).
+Running this analysis among age classes revealed differential survival effects.
+Nestling telomere length had no effect on survival  (estimate = `r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`; Fig.
+5A), but telomere length at the fledgling stage was significantly related to survival (estimate = `r round3(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`).
+This effect was positive, with longer telomeres in early life associated with increased survival later in life (Fig.
+6B).
 Finally, using the longitudinal data, we found no effect of early-life telomere shortening on subsequent survival (estimate = `r round2(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,1])`, CI = `r round3(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,3], lessthan = F)`).
@@
 We then show that within cohorts, the social environment (i.e.
-number of helpers) affects telomere length, but that this effect is age-dependent.
+number of helpers) affects telomere length specifically at the nestling stage.
 Finally, we show that telomere length in early life predicts later-life survival.
@@
 Our finding that food availability is related to early-life telomere length and rate of shortening adds to a growing body of evidence from humans and wild animals showing that the natal environment can have pronounced impacts on early-life telomere dynamics [reviewed in @Price2013; @Monaghan2014].
-However, very few studies have shown that temporal variation in environmental conditions affects telomere dynamics in natural populations.
-In a recent study on black-tailed gulls (*Larus crassirostris*), Mizutani *et al.* [-@Mizutani2013] showed that temporal variation in telomere shortening occurred over a two-year period, which they suggest was the result of changing climatic conditions.
-Similarly, Watson *et al.* [-@Watson2015] studied telomere shortening in two cohorts of European storm petrel (*Hydrobates pelagicus*), and found that the cohort raised in more favourable conditions had, on average, longer telomeres.
+However, very few studies have shown that temporal variation in environmental conditions affects telomere dynamics in natural populations, and the studies that have done so are limited to just two seasons [@Mizutani2013; @Watson2015].
 The long-term Seychelles warbler dataset has allowed us, using many more cohorts than has been possible in other studies, to show that temporal variation in environmental conditions does indeed affect telomere dynamics, and that this effect can be seen at the population level.
 Interestingly, in the Seychelles warbler annual variation in food availability is not directly linked to survival [@Brouwer2006; @Hammers2013].
-Our results therefore suggest that telomeres can detect hidden costs in natural populations that are not detectable using life-history data alone [@Asghar2015]
+It has been suggested that telomeres act as biomarkers of cost in natural populations, as they are able to capture individual and cohort-level variability in the costs experienced as a result of environmental pressures [@Asghar2015].
+Our results confirm that telomeres can detect hidden costs in natural populations that are not detectable using life-history and environmental data alone
 
-We suspect the current lack of information on temporal variation in telomere dynamics is more due to a lack of available long-term datasets with telomere screening, rather than the Seychelles warbler being unique.
+We suspect the novelty of our findings of temporal, environmentally-induced variation in telomere dynamics within a population is more due to a lack of available long-term datasets with telomere screening, rather than the Seychelles warbler being unique.
 Indeed, the environment on Cousin is benign in comparison with many regions outside the tropics, where populations undergo large fluctuations in size [e.g.
 @Coulson2001].
-We therefore expect that as more long-term studies generate telomere datasets, the effects of temporal changes in the environment on telomere dynamics will become increasingly clear.
-If our findings are replicated in other systems and temporal variation in early-life telomere dynamics is common in nature, this has a number of ramifications for our understanding of senescence in natural populations.
-First, our findings suggest that telomere dynamics of a population at a given point in time represent a snapshot of a temporally varying process.
+We therefore expect that as more long-term studies generate telomere datasets, the effects of temporal changes in the environment on population-level telomere dynamics will become increasingly clear.
+If our findings are replicated in other systems and population-level variation in early-life telomere dynamics is common in nature, this has a number of ramifications for our understanding of senescence in natural populations.
+First, our findings suggest that the telomere dynamics of a population at a given point in time represent a snapshot of a temporally varying process.
 More research is now needed within and across multiple cohorts and populations to better understand the complexity of telomere evolutionary ecology.
@@
 
-We found that when looking within cohorts, tarsus length and the number of helpers were the most important predictors of telomere length in Seychelles warbler chicks.
-The tarsus effect most likely reflects the fact that in passerine birds tarsus length is correlated with age during the nestling stage (DAVE DO YOU HAVE A REF??).
-Telomere loss is most rapid early in life due to ongoing cell replication, and a negative correlation between telomere length and body size in early life is therefore expected (REF).
-However, ours is the first study to show that the presence of helping co-breeders has a beneficial effect on offspring in terms of their telomeres.
-In the Seychelles warbler the survival benefits of having helpers in the natal territory is well established [@Komdeur1994;@Brouwer2012], and that this effect is detected in the telomere data is therefore encouraging.
-However, we found no evidence that tarsus length or the number of helpers was related to telomere loss - this is perhaps surprising as telomere loss is expected to be a better indicator of stress than telomere length *per se* [@Boonekamp2014].
+We found that within cohorts, tarsus length and the number of helpers were the most important predictors of telomere length in Seychelles warblers, and that these effects occurred specifically at the nestling stage.
+The tarsus effect most likely reflects the fact that in passerine birds tarsus length is correlated with age during the nestling stage **WHY CAN'T I FIND A BLOODY REFERENCE FOR THIS!!!**.
+Telomere loss is most rapid early in life due to ongoing cell replication [@Fenck1998], and a negative correlation between telomere length and body size in early life is therefore expected.
+
+Ours is the first study to show that the presence of helping co-breeders has a beneficial effect on offspring in terms of their telomeres.
+In the Seychelles warbler the survival benefits to a nestling of their being helpers in the natal territory is well established [@Komdeur1994;@Brouwer2012], and that this effect is detected in terms of differential telomere length, is therefore encouraging.
+However, we found no evidence that the number of helpers was related to telomere loss - this is perhaps surprising as telomere loss is expected to be a better indicator of stress than telomere length *per se* [@Boonekamp2014].
 Our (lack of) finding here is probably due to the limited nature of our longitudinal dataset, both in terms of sample size and resolution (i.e.
 time between sampling events).
-Seychelles warblers have not rarely been sampled multiple times within their first year of life, so much of the telomere shortening that occurs in early life will be missed with our sampling regime.
-It is likely, therefore that in this system the effects of environmental variation on telomere loss will only be detected when they are very strong, and that telomere length constitutes a better indicator of early life stress.
+Seychelles warblers are rarely been sampled multiple times within their first year of life, so much of the telomere shortening that occurs in early life will be missed with our sampling regime.
+It is likely, therefore that the effects of environmental variation on telomere loss will only be detected when they are very strong using this dataset.
+Thus telomere length constitutes a better indicator of early life stress.
 
 While the relationship between telomeres and mortality in adults has been established for some time [@Cawthon2003], only recently has the link between early-life telomere dynamics and later-life survival been studied.
-In captive zebra finches, juvenile telomere length predicts late-life survival [@Heidinger2012], and in wild bird populations early-life telomere dynamics have been linked to survival during the nestling-phase [@Watson2015], and survival to adulthood [@Boonekamp2014].
+In captive zebra finches, juvenile telomere length predicts late-life survival [@Heidinger2012], and in wild bird populations early-life telomere dynamics have been linked to survival during the nestling phase [@Watson2015], and survival to adulthood [@Boonekamp2014].
 Our results are, to our knowledge, the first to demonstrate that early-life telomere length is associated with later-life survival in a wild population.
-This effect can be seen at the individual level, and (to a lesser extent) at the cohort level, although the nature of the relationship varies according to to age class, and depending on whether cross-sectional or longitudinal data are used (Figs 5,6).
-At the individual level, we found that telomere length in fledglings and subadults, but not chicks, was related to survival.
-This is unsurprising, as at the chick stage telomere length at the fledgling and subadult stage is likely to be the best indicator of early life stress (see also previous paragraph).
-The lack of correlation between population-level telomere length and lifespan is more difficult to explain, and given the sample size and resolution of the data, the replationship between population-level telomere loss and lifespan should be cautiously interpreted.
-Nonetheless, our data do support the general conclusion that early-life telomere dynamics are important indicators of early-life stress.
+This effect can be seen at the individual level, and (to a lesser extent) at the cohort level, although the nature of the relationship varies according to the age class at which chicks are measured.
+At the individual level, we found that telomere length in fledglings, but not nestlings, was related to survival.
+This is unsurprising, as only by the fledgling stage is telomere length likely to reflect the accumulation of early life stress (see also previous paragraph).
+The relationships between population-level telomere length and lifespan in the expected direction (Fig.
+4), but not significant, which is perhaps not surprising given that even though we have a lon-term dataset, we are still restricted in termas of sample size when conducting cohort-level analyses.
+Nonetheless, our data support the general conclusion that early-life telomere dynamics are important indicators of early-life stress.
 
@@
 ###Data accessibility
-This manuscript was written in R Markdown (http://rmarkdown.rstudio.com/).
-All data and scripts required to reproduce the manuscript, and all figures and analyses contained within it, will be made available on GitHub.
+This manuscript was written in R Markdown (http://rmarkdown.rstudio.com/). All data and scripts required to reproduce the manuscript, figures and analyses will be made available on GitHub.
 

### 2015-09-23 Scripts/SW juvenile telomeres IPC clean.R

@@ -70,7 +70,10 @@ dd$Age <- dd$CatchYear-dd$LayYear
 dd <- droplevels(dd[dd$Ageclass != '',])
 levels(dd$Ageclass) <- c('A','CH','FL','FL','FL','SA')
 dd$Ageclass <- factor(dd$Ageclass,levels = c('CH','FL','FL','FL','SA','A'))
-
+dd$Fledged <- ifelse(dd$Ageclass == 'CH','Nestlings',
+                     ifelse(dd$Ageclass == 'A','Adults',
+                            'Fledglings'))
+dd$Fledged <- factor(dd$Fledged,levels = c('Nestlings','Fledglings','Adults'))
 
 
 dd$Agemonths <- ifelse(dd$Ageclass == 'CH',1,
@@ -206,6 +209,8 @@ for(i in 1:nrow(juv))
   juv$cenNonHelper[i] <- (juv$NonHelper[i] -  mean(currentdata$NonHelper))/sd(currentdata$NonHelper)
 }
 
+juv <- subset(juv,cenTQ<4)
+
 
 FlSA <- subset(juv,Ageclass!='CH')
 chicks <- subset(juv,Ageclass == 'CH')
@@ -277,7 +282,7 @@ juvseason <- ddply(juv,
                    CatchYear = mean(CatchYear),
                    n = length(TLKB))
 juvseason$cenTL <- juvseason$TLKBmean-mean(juvseason$TLKBmean)
-juvseason <- subset(juvseason,n>2)
+juvseason <- subset(juvseason,n>5)
 
 chickseason <- ddply(chicks,
                    .(FieldPeriodID,Season),
@@ -291,10 +296,10 @@ chickseason <- ddply(chicks,
                    CatchYear = mean(CatchYear),
                    n = length(TLKB))
 chickseason$cenTL <- chickseason$TLKBmean-mean(chickseason$TLKBmean)
-chickseason <- subset(chickseason,n>2)
+chickseason <- subset(chickseason,n>5)
 
 
-flseason <- ddply(subset(FlSA,Ageclass == 'FL'),
+flseason <- ddply(FlSA,
                      .(FieldPeriodID,Season),
                      summarize,
                      TLKBmean = mean(TLKB),
@@ -306,22 +311,7 @@ flseason <- ddply(subset(FlSA,Ageclass == 'FL'),
                      CatchYear = mean(CatchYear),
                      n = length(TLKB))
[…]

### 2015-09-24 Manuscript/SW_early_life_telomeres.Rmd
Kat B edits

--- 2015-09-23 Manuscript/SW_early_life_telomeres_v9.Rmd
@@
 Telomeres - protective caps on the ends of chromosomes - shorten in response to early life stress, and early-life telomere shortening is correlated with reduced later-life survival.
-It is thus possible that as telomeres are able to measure the impact of early life costs, they may provide a link between early-life environmental variation and later-life fitness.
+As telomeres are able to measure the impact of early life costs, they may provide a link between early-life environmental variation and later-life fitness.
 
@@
 The Seychelles warbler system is excellently suited to addressing this question, due to the availability of long-term longitudinal survival and senescence data unconfounded by dispersal.
-We combined this with cross-sectional and longitudinal telomere measurements taken from over 300 birds from multiple cohorts, and analyse how variation in the early life environment within and across cohorts affects telomere dynamics and survival.
+We combined this with cross-sectional and longitudinal telomere measurements taken from over 300 birds from multiple cohorts, and analysed how variation in the early life environment within and across cohorts affects telomere dynamics and survival.
 
@@
 Individuals born in breeding seasons with high food availability have reduced telomere shortening compared to individuals born in years with low food availability.
-Within cohorts, we found telomere length in nestlings was nrgatively related to tarsus length (which likely represents nestling crowth), and positvely related to the number of helpers present in the natal territory.
+Within cohorts, we found that telomere length in nestlings was negatively related to tarsus length (which likely represents nestling growth), and positively related to the number of helpers present in the natal territory.
 However, neither tarsus length nor the number of helpers was related to telomere length in fledged birds.
@@
 Telomeres are protective caps on the ends of chromosomes which shorten with age [@Monaghan2006], and in response to oxidative stress levels which can be elevated  due to environmental factors [@VonZglinicki2002].
-When telomeres reach a critical shortness cells senesce  [@Campisi2003], and the accumulation of these cells can result in organismal senescence and death  [@Wong2003].
+When telomeres reach a critical shortness cells senescence  [@Campisi2003], and the accumulation of these cells can result in organismal senescence and death  [@Wong2003].
 These links between telomere length and senescence have inspired a great deal of recent research into telomere ecology [reviewed in @Horn2010; @Haussmann2010; @Monaghan2014].
@@
 Importantly, early-life telomere dynamics are related to both short-term and late-life survival [@Heidinger2012; @Boonekamp2014], and to other parameters such as cognition [@Nettle2015a].
-However, the extent to which telomeres provide a link between early-life conditions and late-life survival in the wild is very poorly understood.
+However, the link between early-life conditions and late-life survival is limited to a few species, and thus poorly understood.
 Moreover, how early life telomere dynamics vary over spatial and temporal scales is not known.
@@
 Ecological conditions an population density on Cousin are variable over space and time due to weather-induced changes in foliage cover and food availability [@VandeCrommenacker2011a].
-Variation in natal territory quality has been linked to changes in levels of oxidative stress [@VandeCrommenacker2011a], but neither natal territory quality nor natal local density appear to be linked to early-life or adult survival [@Brouwer2006; @Hammers2013].
-Cooperative breeding occurs in the Seychelles warbler [@Komdeur1994a; @Richardson2003a], and the presence of helping subordinates (but not non-helping subordiantes) in the natal territory has been associated with increased survival later in life [@Brouwer2012].Finally, we have an established protocol for assessing absoltute telomere length in this species [@Barrett2012], and telomere length predicts survival independently of age in adult Seychelles warblers, suggesting that telomeres act as a biomarker of cost [@Barrett2013a].
+Seychelles warblers remain on their natal territories for at least six months, and variation in natal territory quality has been linked to changes in levels of oxidative stress [@VandeCrommenacker2011a].
+However, neither natal territory quality nor natal local density appear to be linked to early-life or adult survival [@Brouwer2006; @Hammers2013].
+Facultative cooperative breeding occurs in the Seychelles warbler [@Komdeur1994a; @Richardson2003a], and the presence of helping subordinates (but not non-helping subordinates) in the natal territory has been associated with increased survival later in life [@Brouwer2012].
+Lastly, we have an established protocol for assessing absolute telomere length in this species [@Barrett2012], and telomere length predicts survival independently of age in adult Seychelles warblers, suggesting that telomeres act as a biomarker of cost [@Barrett2013a].
 Thus, we have an excellent system in which to assess the costs of different social and environmental conditions experienced early in life, and to assess the later life consequences of early-life conditions.
@@
 In this study we test the prediction that telomeres link early-life environmental variation to late-life fitness.
-Because telomere dy6naimcs are expected to reflect individual-level variation in the costs of early-life experiences, they may allow a more sensitive analysis of the effects of early early-life environmental variation than would be possible with a direct comparison of the early-life environment and survival.
+Because telomere dynamics are expected to reflect individual-level variation in the costs of early-life experiences, they may allow a more sensitive analysis of the effects of early-life environmental variation than would be possible with a direct comparison of the early-life environment and survival.
 With this in mind, we predict that: i) individuals raised in good quality conditions will have longer telomeres, and ii) having longer telomeres in early life is associated with greater survival.
@@
 320 birds on Cousin Island (04'20'S, 55'40'E) has been intensively studied since 1986 [@Richardson2003b; @Spurgin2014a].
-This species has two breeding seasons, running from June-August (main breeding season) and December-February (minor breeding season), when females on one of the *ca*.
-115 breeding territories lay one or, rarely, two or three eggs [@Komdeur1991].
+This species has two breeding seasons, running from June-August (main breeding season) and December-February (minor breeding season), when the breeding females on each of the *ca*.
+115 territories lay one or, rarely, two or three eggs [@Komdeur1991].
 As a result of this low reproductive output, combined with a higher mortality rate among first-year birds [61% survival; @Brouwer2006], cohort sizes in the Seychelles warbler are small (<50).
@@
 They are then followed throughout their lives, and as they are non-migratory endemics naturally confined to the island [@Komdeur2004a], a biannual census of birds on Cousin during each breeding season gives accurate measures of local density and individual survival [@Barrett2013a].
-The isolated nature of the Cousin population is a key advantage of the system for analyses involving survival, which are often confounded by emigration [see @Ergon2014 for a recent discussion].
+The isolated nature of the Cousin population is a key advantage of the system for analyses involving survival, which in other systems are often confounded by emigration [see @Ergon2014 for a recent discussion].
 Full details of catching and census methods can be found in Brouwer *et al.* [-@Brouwer2012].
 
-All territories are mapped during the breeding seasons using detailed observational data of foraging and territorial defence behaviour, and surveyed for territory quality [@Richardson2003b].
+Seychelles warblers are highly territorial and all territories are mapped during the breeding seasons using detailed observational data of foraging and territorial defence behaviour, and surveyed for territory quality [@Richardson2003b].
 Territory quality is calculated based on territory size, foliage cover and insect abundance [@Komdeur1992], and territory quality estimates obtained across years are averaged to obtain a single value for each territory [@Hammers2013].
-Cousin is subject to intra- and inter-annual variation in rainfall and food availability, and such island-wide temporal variation may override the effects of absolute territory quality.
+Cousin is subject to intra- and inter-annual variation in rainfall and food availability, and such island-wide temporal variation may override the effects of individual territory quality.
 As an estimate of seasonal variation in food availability, we calculated an index of the number of insects across the entire island during each breeding season.
@@
 
-Each time a bird is caught on Cousin body mass and tarsus length are measured (to the nearest 0.1g and 0.1mm, respectively), and age is assessed on the basis of eye colour and previous captures [@Richardson2003b].
+Each time a bird is caught on Cousin body mass and tarsus length are measured (to the nearest 0.1g and 0.1mm, respectively), and age is confirmed on the basis of eye colour and previous captures [@Richardson2003b].
 A blood sample (ca 25 $\mu$l) is taken from each bird via brachial venipuncture, and stored in 1 ml of absolute ethanol in a 1.5 ml screw-cap microfuge tube at room temperature.
@@
 Telomere lengths were measured from a total of `r nrow(FlSA)+nrow(chicks)+nrow(Loss)` samples from `r nrow(FlSA)+nrow(chicks)` birds caught within their first year of life, between `r min(as.numeric(paste(juv$LayYear)))` and `r max(as.numeric(paste(juv$LayYear)))`.
-Of these birds, `r nrow(chicks)` were nestlings less than two months old (chicks), while the remaining `r nrow(FlSA)` were fledglings and subadults aged between two and twelve months (hereafter reffered to as fledglings).
+Of these birds, `r nrow(chicks)` were nestlings less than one month old (chicks), while the remaining `r nrow(FlSA)` were fledglings and subadults aged between two and twelve months (hereafter referred to as fledglings).
 For a subset of first-year birds (n = `r nrow(Loss)` individuals) we had longitudinal data, with samples from individuals caught both in their first year and then again within two years.
@@
 This makes it difficult to disentangle how spatial and temporal processes differentially affect early-life telomere dynamics.
-To get around this, we separately analysed how variation in early life conditions among and within breeding seasons was related to variation in early life telomere length.
+To overcome this, we analysed how variation in early life conditions was related to variation in early-life telomere length separately among and within breeding seasons.
 
@@
 
-We then individual-level analyses to test how spatially varying early-life environmental and social conditions influence telomere length within cohorts using linear models.
+We then conducted individual-level analyses to test how spatially varying early-life environmental and social conditions influence telomere length within cohorts, using general linear models.
 For these analyses, both response and explanatory variables were standardised by mean-centering within cohorts, thus controlling for any between-cohort temporal variation in telomere length and the social and ecological environment.
@@
 Finally, we analysed the effects of telomere dynamics on survival, again by looking at variation both within and among cohorts.
-First, we tested the prediction that cohort-level telomere dynamics are associated with survival by comparing the mean telomere length and rate of loss of each cohort with the mean lifespan of all individuals, using linear regression.
+First, we tested the prediction that cohort-level telomere dynamics are associated with survival by comparing the mean telomere length and rate of loss of each cohort with the mean lifespan of all individuals in that cohort, using linear regression.
 For this analysis we excluded any recent breeding seasons where >10% of individuals in a cohort were still alive.
@@
 1A).
-A longitudinal analysis of telomere loss showed that nestlings and fledglings both telomeres early in life, but that there was no difference in the rate of telomere loss among age classes (F = `r round2(anova(lm(TROC~Fledged,data = Loss))[1,4])`; *P* = `r round2(anova(lm(TROC~Fledged,data = Loss))[1,5])`; Fig.
+A longitudinal analysis of telomere loss showed that nestlings and fledglings both lost telomeres early in life, but that there was no difference in the rate of telomere loss among age classes (F = `r round2(anova(lm(TROC~Fledged,data = Loss))[1,4])`; *P* = `r round2(anova(lm(TROC~Fledged,data = Loss))[1,5])`; Fig.
 1B).
@@
 #####Temporal variation in early life telomere dynamics
-Early-life telomere length varied significantly among breeding seasons (one-way ANOVA, F = `r round2(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,4])`; *P* `r round3(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,5])`), and this variation in average telomere length over breeding seasons was positively related to island-wide food availability (linear regression, R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,5])`; Fig.
+Early-life telomere length varied significantly among breeding seasons (one-way ANOVA, F = `r round2(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,4])`; *P* = `r round3(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,5])`), and this variation in average telomere length over breeding seasons was positively related to island-wide food availability (linear regression, R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,5])`; Fig.
 2A).
@@
 
-The rate of telomere shortening in early life (taken from the smaller subsert of birds with longitudinal samples) did not vary significantly among breeding seasons (F = `r round2(anova(lm(TROC~factor(FieldPeriodID),data = Loss))[1,4])`; *P* = `r round2(anova(lm(TROC~factor(FieldPeriodID),data = Loss))[1,5])`); however, what variation there was in rates of telomere shortening could be explained by temporal variation in food availability (R^2^ = `r round2(summary(lm(Insect~TROCmean,data=loss.season,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TROCmean,data = loss.season,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Insect~TROCmean,data = loss.season,na.action=na.exclude))[1,5])`).
+The rate of telomere shortening in early life (taken from the smaller subset of birds with longitudinal samples) did not vary significantly among breeding seasons (F = `r round2(anova(lm(TROC~factor(FieldPeriodID),data = Loss))[1,4])`; *P* = `r round2(anova(lm(TROC~factor(FieldPeriodID),data = Loss))[1,5])`); however, what variation there was in rates of telomere shortening could be explained by temporal variation in food availability (R^2^ = `r round2(summary(lm(Insect~TROCmean,data=loss.season,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TROCmean,data = loss.season,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Insect~TROCmean,data = loss.season,na.action=na.exclude))[1,5])`).
 This pattern corresponded with the cross-sectional data: individuals born in seasons with low food availability lost telomeres, on average, at a faster rate than those born in years with high food availability (Fig.
 2B).
+Unfortunately we did not have sufficient sample sizes within cohorts to run this analysis separately among age classes.
 
@@
 @Coulson2001].
-We therefore expect that as more long-term studies generate telomere datasets, the effects of temporal changes in the environment on population-level telomere dynamics will become increasingly clear.
 If our findings are replicated in other systems and population-level variation in early-life telomere dynamics is common in nature, this has a number of ramifications for our understanding of senescence in natural populations.
@@
 Secondly, our findings raise the interesting prospect of using telomeres as indicators of population health - an approach that would be useful to a broad range of conservation and animal health practitioners.
-However, our analyses are confined to a single island, and research across populations with different environments and genetic histories is required in order to provide further insight into the usefulness of telomeres as biomarkers of population health.
+However, further research across populations with different environments and genetic histories is required in order to provide further insight into the usefulness of telomeres as biomarkers of population health.
 
 We found that within cohorts, tarsus length and the number of helpers were the most important predictors of telomere length in Seychelles warblers, and that these effects occurred specifically at the nestling stage.
-The tarsus effect most likely reflects the fact that in passerine birds tarsus length is correlated with age during the nestling stage **WHY CAN'T I FIND A BLOODY REFERENCE FOR THIS!!!**.
-Telomere loss is most rapid early in life due to ongoing cell replication [@Fenck1998], and a negative correlation between telomere length and body size in early life is therefore expected.
+The tarsus effect most likely reflects the fact that in passerine birds tarsus length is correlated with age during the nestling stage [@Ricklefs1976].
+Telomere loss is most rapid early in life due to ongoing cell replication [@Frenck1998], and a negative correlation between telomere length and body size in early life is therefore expected.
 
@@
 Our results are, to our knowledge, the first to demonstrate that early-life telomere length is associated with later-life survival in a wild population.
-This effect can be seen at the individual level, and (to a lesser extent) at the cohort level, although the nature of the relationship varies according to the age class at which chicks are measured.
+This effect can be seen at the individual level, and (to a lesser extent) at the cohort level, although the nature of the relationship varies according to the age class at which telomeres are measured.
 At the individual level, we found that telomere length in fledglings, but not nestlings, was related to survival.
@@
 The relationships between population-level telomere length and lifespan in the expected direction (Fig.
-4), but not significant, which is perhaps not surprising given that even though we have a lon-term dataset, we are still restricted in termas of sample size when conducting cohort-level analyses.
+4), but not significant, which is perhaps not surprising given that even though we have a long-term dataset, we are still restricted in terms of sample size when conducting cohort-level analyses.
 Nonetheless, our data support the general conclusion that early-life telomere dynamics are important indicators of early-life stress.

### 2015-09-24 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Kat B edits

--- 2015-09-23 Manuscript/SW_early_life_telomeres_tables_and_figures v9.Rmd
@@
                fig.path='../Figures/',
-               fig.width= 4,
-               fig.height= 8,
                fig.show='asis',
@@
 **Figure 3** Factors affecting early-life telomere length in Seychelles warblers at the individual level, within seasons.
-**A** Model averaged estimates and 95% confidence intervals for alxplanatory terms used in a linear model with telomere length as the response variable.
+**A** Model averaged estimates and 95% confidence intervals for explanatory terms used in a linear model with telomere length as the response variable.
 Numbers in brackets are the relative importance of each term in the top model set (see main text for details; TQ = territory quality).
@@
 **Figure 5** Kaplan-Meier curves showing the relationship between telomere length and survival for individual Seychelles warblers measured as chicks (**A**), and fledglings  (**B**).
-Telomere length is binned into groups here for visualisation purposes only (long and short = greater than or less than median telomere length, repectively).
+Telomere length is binned into groups here for visualisation purposes only (long and short = greater than or less than median telomere length, respectively).
 
@@
 
-```{r Figure 1,fig.width=5,fig.height=9}
+```{r Figure 1,fig.width=4,fig.height=8}
 ######################
@@
 
-```{r Figure 2,fig.width=5,fig.height=9}
+```{r Figure 2,fig.width=4,fig.height=8}
 
@@
            col = Season)) +
-  ylab('Telomere length') + 
+  ylab('Telomere length (kb)') + 
   xlab('Food availability') +
@@
                     y = TLKB,
-                    col=Fledged,
+                    col = Fledged,
                     lty = Fledged)) +

### 2015-10-05 Manuscript/SW_early_life_telomeres.Rmd
HLD edits. Abstract and citation style changed to Evolution

@@
 bibliography: MyBib.bib
-csl: JAE.csl
+csl: Evolution.csl
 
@@
 b. Department of Zoology, Edward Grey Institute, University of Oxford, Oxford, UK
-c. Behavioural Ecology and Self-organization Group, Centre for Ecological and Evolutionary Studies, University of Groningen, Groningen, The Netherlands
+c. Groningen Institute for Evolutionary Life Sciences, University of Groningen, Groningen, The Netherlands
 d. Department of Animal and Plant Sciences, NERC Biomolecular Analysis Facility, University of Sheffield, Sheffield, UK
@@
 ###Abstract
-1.
-Favourable conditions early in life can confer delayed fitness advantages.
-However, the mechanistic basis of these 'silver spoon effects' remains poorly understood.
-Telomeres - protective caps on the ends of chromosomes - shorten in response to early life stress, and early-life telomere shortening is correlated with reduced later-life survival.
-As telomeres are able to measure the impact of early life costs, they may provide a link between early-life environmental variation and later-life fitness.
+Understanding the links between early-life experiences and later life survival is fundamental to our understanding of life-history evolution.
+Telomeres - protective caps on the ends of chromosomes - shorten in response to early-life stress, and early-life telomere shortening is correlated with reduced later-life survival.
+As telomeres are a biomarker of the impact of early-life costs, they may provide a link between early-life environmental variation and later-life fitness.
+We studied how telomere dynamics link spatiotemporal variation in early-life conditions to survival using a long-term study of the Seychelles warbler (*Acrocephalus sechellensis*).
+We show that early-life telomere dynamics vary markedly across cohorts, and that individuals born in breeding seasons with higher food availability have longer telomeres and reduced telomere shortening.
+Within cohorts, we found that telomere length in nestlings was negatively related to tarsus length (which likely represents nestling growth), and positively related to the number of helpers present in the natal territory.
+Finally, we found that longer telomeres and reduced telomere shortening rates in early life were associated with increased survival later in life.
+Our results show that telomeres may provide a link between early-life conditions and late-life survival in wild populations.
 
-2.
-The prevalence and strength of silver spoon effects has consequences for individual fitness and population growth rates.
-Understanding the physiological and ecological links between early life experiences and later life survival is therefore fundamental to our understanding of life-history evolution and population ecology.
-
-3.
-We studied how telomere length and shortening link spatiotemporal variation in early-life conditions to survival in the Seychelles warbler (*Acrocephalus sechellensis*).
-The Seychelles warbler system is excellently suited to addressing this question, due to the availability of long-term longitudinal survival and senescence data unconfounded by dispersal.
-We combined this with cross-sectional and longitudinal telomere measurements taken from over 300 birds from multiple cohorts, and analysed how variation in the early life environment within and across cohorts affects telomere dynamics and survival.
-
-4.
-We show that early-life telomere length and shortening rates vary markedly across cohorts, and that this variation is related to temporal fluctuations in food availability.
-Individuals born in breeding seasons with high food availability have reduced telomere shortening compared to individuals born in years with low food availability.
-Within cohorts, we found that telomere length in nestlings was negatively related to tarsus length (which likely represents nestling growth), and positively related to the number of helpers present in the natal territory.
-However, neither tarsus length nor the number of helpers was related to telomere length in fledged birds.
-Finally, we found that longer telomeres and reduced telomere shortening rates in early life were associated with increased survival later in life.
-
-5.
-Our results show that telomeres can reveal complex hidden costs at the individual and cohort levels, and suggest that they may provide a link between early-life conditions and late-life survival in wild populations.
-
-**Keywords:** Ageing; Life-history; Seychelles warbler; Telomeres
+**Keywords:** Life-history; Seychelles warbler; Senescence; Survival; Telomeres
 
@@
 Being subject to favourable environmental conditions during development and growth can confer fitness advantages later in life (so called 'silver spoon effects') [@Grafen1988; @Monaghan2008].
-There is now evidence that adult fitness can be affected by a range of early-life experiences, including variation in habitat quality and resource availability [@Madsen2000; @VandePol2006; @Hayward2013], population density [@Nussey2007;@Douhard2013] and natural or anthropogenic environmental disturbance [@Reid2003; @Cartwright2014].
+Adult fitness can be affected by a range of early-life experiences, including variation in habitat quality and resource availability [@Madsen2000; @VandePol2006; @Hayward2013], population density [@Nussey2007;@Douhard2013] and natural or anthropogenic environmental disturbance [@Reid2003; @Cartwright2014].
 Understanding the causes and consequences of these early-life experiences is key to understanding many ecological and evolutionary  processes, including  patterns of natural and sexual selection,  population growth rates and even local extinction [@Coulson2001; @Roach2014].
@@
 Indeed, the pervasiveness of silver spoon effects varies between species [@Drummond2011], cohorts [@Reid2003] and sexes [@Wilkin2009].
-In order to better understand the later-life consequences of early-life experiences, we therefore need to understand, or at least be able to measure, how the environment differentially affects individuals within a population.
+To better understand the later-life consequences of early-life experiences, we therefore need to understand, or at least be able to measure, how the environment differentially affects individuals within a population.
 
@@
 Telomeres are protective caps on the ends of chromosomes which shorten with age [@Monaghan2006], and in response to oxidative stress levels which can be elevated  due to environmental factors [@VonZglinicki2002].
-When telomeres reach a critical shortness cells senescence  [@Campisi2003], and the accumulation of these cells can result in organismal senescence and death  [@Wong2003].
+When telomeres reach a critical shortness cells senesce [@Campisi2003], and the accumulation of these cells can result in organismal senescence and death  [@Wong2003].
 These links between telomere length and senescence have inspired a great deal of recent research into telomere ecology [reviewed in @Horn2010; @Haussmann2010; @Monaghan2014].
-While there is little direct evidence the relationship between telomere dynamics and survival is causal [@Barrett2011; @Simons2015], there is now excellent evidence that telomeres can act as biomarkers of cost in wild populations, retaining signatures of ecological stress that are otherwise difficult to detect [@Monaghan2014; @Schultner2014; @Asghar2015].
+While there is little direct evidence that the relationship between telomere dynamics and survival is causal [@Barrett2011; @Simons2015], there is now excellent evidence that telomeres can act as biomarkers of cost in wild populations, retaining signatures of ecological stress that are otherwise difficult to detect [@Monaghan2014; @Schultner2014; @Asghar2015].
 
@@
 However, the link between early-life conditions and late-life survival is limited to a few species, and thus poorly understood.
-Moreover, how early life telomere dynamics vary over spatial and temporal scales is not known.
+Moreover, how early-life telomere dynamics vary over spatial and temporal scales is not known.
 
@@
 Due to the isolated nature of the study population and intensive field monitoring, we have excellent survival data and sampling material spanning many years (see methods, below).
-Ecological conditions an population density on Cousin are variable over space and time due to weather-induced changes in foliage cover and food availability [@VandeCrommenacker2011a].
+Ecological conditions and population density on Cousin are variable over space and time due to weather-induced changes in foliage cover and food availability [@VandeCrommenacker2011a].
 Seychelles warblers remain on their natal territories for at least six months, and variation in natal territory quality has been linked to changes in levels of oxidative stress [@VandeCrommenacker2011a].
@@
 
-Telomere lengths were measured from a total of `r nrow(FlSA)+nrow(chicks)+nrow(Loss)` samples from `r nrow(FlSA)+nrow(chicks)` birds caught within their first year of life, between `r min(as.numeric(paste(juv$LayYear)))` and `r max(as.numeric(paste(juv$LayYear)))`.
-Of these birds, `r nrow(chicks)` were nestlings less than one month old (chicks), while the remaining `r nrow(FlSA)` were fledglings and subadults aged between two and twelve months (hereafter referred to as fledglings).
-For a subset of first-year birds (n = `r nrow(Loss)` individuals) we had longitudinal data, with samples from individuals caught both in their first year and then again within two years.
+Telomere lengths were measured for `r nrow(FlSA)+nrow(chicks)+nrow(Loss)` samples.
+Of these, `r nrow(FlSA)+nrow(chicks)` were taken cross-sectionally from birds caught within their first year of life, between `r min(as.numeric(paste(juv$LayYear)))` and `r max(as.numeric(paste(juv$LayYear)))`; `r nrow(chicks)` were nestlings less than one month old (chicks), while the remaining `r nrow(FlSA)` were fledglings and subadults aged between two and twelve months (hereafter referred to as fledglings).
+For a subset of first-year birds (n = `r nrow(Loss)` individuals) we had longitudinal data, with an additional sample taken within two years of the original catch.
 We measured absolute telomere quantity per diploid genome for all samples using a quantitative PCR (qPCR) assay  [full details in @Barrett2012].
@@
 This makes it difficult to disentangle how spatial and temporal processes differentially affect early-life telomere dynamics.
-To overcome this, we analysed how variation in early life conditions was related to variation in early-life telomere length separately among and within breeding seasons.
+To overcome this, we analysed how variation in early-life conditions was related to variation in early-life telomere length separately among and within breeding seasons.
 
 First, to analyse whether temporal variation in food availability drives population-level variation in telomere dynamics, we calculated average telomere length among all birds born in a given breeding season (i.e.
-a cohort), excluding cohorts where we had samples from $\leq$ 5 birds.
+a cohort), excluding cohorts where we had samples from $\leq$ 5 birds (n = `r nrow(juvseason)` cohorts).
 We tested for a difference in telomere length among cohorts using a one-way ANOVA, and then used linear regression to test whether variation in mean telomere length among cohorts is explained by temporal variation in insect abundance.
@@
 
-For individuals with longitudinal data we calculated a rate of telomere loss between the first-year sample and earliest adult sample by subtracting the adult telomere length from the early-life telomere length and dividing this difference by the length of time (in days) between sampling events.
+For individuals with longitudinal data (n = `r nrow(Loss)`) we calculated a rate of telomere loss between the first-year sample and earliest adult sample by subtracting the adult telomere length from the early-life telomere length and dividing this difference by the length of time (in days) between sampling events.
 To account for regression to the mean effects, we applied a correction based on correlations among samples within individuals following Verhulst *et al.* [-@Verhulst2013].
@@
 First, we tested the prediction that cohort-level telomere dynamics are associated with survival by comparing the mean telomere length and rate of loss of each cohort with the mean lifespan of all individuals in that cohort, using linear regression.
-For this analysis we excluded any recent breeding seasons where >10% of individuals in a cohort were still alive.
 We then used parametric survival analysis to test whether individual-level telomere length and rate of loss were related to to survival.
@@
 
-#####Temporal variation in early life telomere dynamics
+#####Temporal variation in early-life telomere dynamics
 Early-life telomere length varied significantly among breeding seasons (one-way ANOVA, F = `r round2(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,4])`; *P* = `r round3(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,5])`), and this variation in average telomere length over breeding seasons was positively related to island-wide food availability (linear regression, R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,5])`; Fig.
 2A).
-Running the regression of cohort-level telomere length and food availability separately for each age class revealed a positive, significant relationship in fledglings (R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=flseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data=flseason,na.action=na.exclude))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKBmean,data=flseason,na.action=na.exclude))[1,5])`), but no relationship in nestlings (R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=chickseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data=chickseason,na.action=na.exclude))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKBmean,data=chickseason,na.action=na.exclude))[1,5])`).
+Running the regression of cohort-level telomere length and food availability separately for each age class revealed a positive, significant relationship in fledglings (R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=flseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data=flseason,na.action=na.exclude))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKBmean,data=flseason,na.action=na.exclude))[1,5])`), but no relationship in nestlings (R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=chickseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data=chickseason,na.action=na.exclude))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKBmean,data=chickseason,na.action=na.exclude))[1,5])`), indicating selective disappearance.
 
@@
 
-#####Spatial variation in early life telomere dynamics
+#####Spatial variation in early-life telomere dynamics
 The top model explaining within-season variation in early-life telomere length contained age class, tarsus length and the number of helpers present in the natal territory, as well as interactions between age class and both tarsus length and the number of helpers (Table S1).
@@
 For the subset of individuals with longitudinal data, we tested whether the factors previously identified as being related to early-life telomere length (i.e.
-tarsus length and number of helpers) were also associated with differences in rates of early life telomere loss.
+tarsus length and number of helpers) were also associated with differences in rates of early-life telomere loss.
 Due to a limited sample size, we considered all age classes together for this analysis.
@@
 
-#####Early life telomere dynamics and survival
-At the population-level, there was a positive but non-significant relationship between telomere length and lifespan (R^2^ = `r round3(summary(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))[1,5])`; Fig.
+#####early-life telomere dynamics and survival
+At the population-level, there was no relationship between telomere length and lifespan (R^2^ = `r round3(summary(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))[1,5])`; Fig.
 4A).
@@
 5A), but telomere length at the fledgling stage was significantly related to survival (estimate = `r round3(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`).
-This effect was positive, with longer telomeres in early life associated with increased survival later in life (Fig.
+This effect was positive, with longer telomeres at the fledgling stage associated with increased survival later in life (Fig.
 6B).
@@
 However, very few studies have shown that temporal variation in environmental conditions affects telomere dynamics in natural populations, and the studies that have done so are limited to just two seasons [@Mizutani2013; @Watson2015].
-The long-term Seychelles warbler dataset has allowed us, using many more cohorts than has been possible in other studies, to show that temporal variation in environmental conditions does indeed affect telomere dynamics, and that this effect can be seen at the population level.
+The long-term Seychelles warbler dataset has allowed us, using many more cohorts than has been used in other studies, to show that temporal variation in environmental conditions does indeed affect telomere dynamics, and that this effect can be seen at the population level.
 Interestingly, in the Seychelles warbler annual variation in food availability is not directly linked to survival [@Brouwer2006; @Hammers2013].
@@
 It is likely, therefore that the effects of environmental variation on telomere loss will only be detected when they are very strong using this dataset.
-Thus telomere length constitutes a better indicator of early life stress.
+Thus telomere length constitutes a better indicator of early-life stress.
 
@@
 At the individual level, we found that telomere length in fledglings, but not nestlings, was related to survival.
-This is unsurprising, as only by the fledgling stage is telomere length likely to reflect the accumulation of early life stress (see also previous paragraph).
-The relationships between population-level telomere length and lifespan in the expected direction (Fig.
+This is unsurprising, as only by the fledgling stage is telomere length likely to reflect the accumulation of early-life stress (see also previous paragraph).
+The relationships between population-level telomere length and lifespan were in the expected direction (Fig.
 4), but not significant, which is perhaps not surprising given that even though we have a long-term dataset, we are still restricted in terms of sample size when conducting cohort-level analyses.

### 2015-10-05 Manuscript/SW_early_life_telomeres_SOM.Rmd
Kat B edits
HLD edits. Abstract and citation style changed to Evolution

--- 2015-09-23 Manuscript/SW_early_life_telomeres_SOM_v9.Rmd
@@
 
-###Supplementary information for Spurgin *et al.* "Telomeres reveal complex silver spoon effects in a wild population"
+###Supplementary information for Spurgin *et al.* "Telomeres reveal silver spoon effects in a wild population"
 
@@
 
-**Table S1** Model selection table for analysis of factors affecting early life telomere length in Seychelles warbler chicks.
+**Table S1** Model selection table for analysis of factors affecting early life telomere length in Seychelles warblers.
 Only models that remained in the top model set for model averaging (AICc $\leq$ 6 compared to the top model) are included here.
+TQ = territory quality.
 

### 2015-10-05 Manuscript/SW_early_life_telomeres.Rmd
JK comments

@@
 
-A key process that has been associated with adult survival is the rate at which telomere repeats are lost.
+A key process that has been associated with adult survival is the rate at which telomeres are shortened.
 Telomeres are protective caps on the ends of chromosomes which shorten with age [@Monaghan2006], and in response to oxidative stress levels which can be elevated  due to environmental factors [@VonZglinicki2002].
@@
 115 territories lay one or, rarely, two or three eggs [@Komdeur1991].
-As a result of this low reproductive output, combined with a higher mortality rate among first-year birds [61% survival; @Brouwer2006], cohort sizes in the Seychelles warbler are small (<50).
+As a result of this low reproductive output, combined with a higher mortality rate among first-year birds [39%; @Brouwer2006], cohort sizes in the Seychelles warbler are small (<50).
 
@@
 The isolated nature of the Cousin population is a key advantage of the system for analyses involving survival, which in other systems are often confounded by emigration [see @Ergon2014 for a recent discussion].
-Full details of catching and census methods can be found in Brouwer *et al.* [-@Brouwer2012].
+Full details of catching and monitoring methods can be found in Brouwer *et al.* [-@Brouwer2012].
 
@@
 
-Each time a bird is caught on Cousin body mass and tarsus length are measured (to the nearest 0.1g and 0.1mm, respectively), and age is confirmed on the basis of eye colour and previous captures [@Richardson2003b].
-A blood sample (ca 25 $\mu$l) is taken from each bird via brachial venipuncture, and stored in 1 ml of absolute ethanol in a 1.5 ml screw-cap microfuge tube at room temperature.
+Each time a bird is caught on Cousin body mass and tarsus length are measured (to the nearest 0.1g and 0.1mm, respectively), and age is confirmed on the basis of eye colour [@Komdeur1991a] and previous captures [@Richardson2003b].
+A blood sample (ca 25 $\mu$l) is taken from each caught bird via brachial venipuncture, and stored in 1 ml of absolute ethanol in a 1.5 ml screw-cap microfuge tube at room temperature.
 
@@
 
-First, to analyse whether temporal variation in food availability drives population-level variation in telomere dynamics, we calculated average telomere length among all birds born in a given breeding season (i.e.
+First, to analyse whether temporal variation in food availability drives population-level variation in telomere dynamics, we calculated average telomere length among all birds born in a given (major or minor) breeding season (i.e.
 a cohort), excluding cohorts where we had samples from $\leq$ 5 birds (n = `r nrow(juvseason)` cohorts).
@@
 
-Ours is the first study to show that the presence of helping co-breeders has a beneficial effect on offspring in terms of their telomeres.
+Ours is the first study to show that the presence of helpers has a beneficial effect on offspring in terms of their telomeres.
 In the Seychelles warbler the survival benefits to a nestling of their being helpers in the natal territory is well established [@Komdeur1994;@Brouwer2012], and that this effect is detected in terms of differential telomere length, is therefore encouraging.

### 2015-10-14 Manuscript/SW_early_life_telomeres.Rmd
Terry edits, intro and methods

@@
 
-Lewis G. Spurgin^a,b^,  Kat Bebbington^a^, Eleanor A. Fairfield^a^, Jan Komdeur^c^, Terry Burke^d^, Hannah, L. Dugdale^c,d^,  and David S. Richardson^a,e,^.
+Lewis G. Spurgin^1,2^,  Kat Bebbington^1^, Eleanor A. Fairfield^1^, Jan Komdeur^3^, Terry Burke^4^, Hannah, L. Dugdale^3,4^,  and David S. Richardson^1,5,^.
 
@@
 c. Groningen Institute for Evolutionary Life Sciences, University of Groningen, Groningen, The Netherlands
-d. Department of Animal and Plant Sciences, NERC Biomolecular Analysis Facility, University of Sheffield, Sheffield, UK
+d. Department of Animal and Plant Sciences, University of Sheffield, Sheffield, UK
 e. Nature Seychelles, Roche Caiman, Mahé, Republic of Seychelles
@@
 ###Abstract
-Understanding the links between early-life experiences and later life survival is fundamental to our understanding of life-history evolution.
-Telomeres - protective caps on the ends of chromosomes - shorten in response to early-life stress, and early-life telomere shortening is correlated with reduced later-life survival.
-As telomeres are a biomarker of the impact of early-life costs, they may provide a link between early-life environmental variation and later-life fitness.
-We studied how telomere dynamics link spatiotemporal variation in early-life conditions to survival using a long-term study of the Seychelles warbler (*Acrocephalus sechellensis*).
-We show that early-life telomere dynamics vary markedly across cohorts, and that individuals born in breeding seasons with higher food availability have longer telomeres and reduced telomere shortening.
-Within cohorts, we found that telomere length in nestlings was negatively related to tarsus length (which likely represents nestling growth), and positively related to the number of helpers present in the natal territory.
-Finally, we found that longer telomeres and reduced telomere shortening rates in early life were associated with increased survival later in life.
-Our results show that telomeres may provide a link between early-life conditions and late-life survival in wild populations.
+Understanding the links between early-life experiences and later-life survival (so-called "silver spoon effects") is fundamental to understanding life-history evolution.
+Telomeres – the protective caps at the ends of chromosomes – shorten  in response to stress, and telomere shortening in early life is correlated with reduced later-life survival.
+Thus, telomere dyamics may help us quantify individual variation in early-life costs and silver spoon effects.
+We tested how telomere dynamics are related to spatiotemporal variation in early-life conditions and later-life survival  using a long-term study of the Seychelles warbler (*Acrocephalus sechellensis*).
+We show that telomere lengths and rates of shortening vary across cohorts, and that individuals born in breeding seasons with higher food availability had longer telomeres and reduced telomere shortening.
+Within cohorts, we found that telomere length in nestlings was negatively related to tarsus length (which probably  represents nestling growth), and positively related to the number of helpers present in the natal territory.
+Finally, we found that increased survival later in life was associated with longer telomeres and reduced telomere shortening rates in early life.
+Our results show that telomeres can both measure variation in early-costs and predict late-life survival in wild populations.
 
@@
 ###Introduction
-Being subject to favourable environmental conditions during development and growth can confer fitness advantages later in life (so called 'silver spoon effects') [@Grafen1988; @Monaghan2008].
-Adult fitness can be affected by a range of early-life experiences, including variation in habitat quality and resource availability [@Madsen2000; @VandePol2006; @Hayward2013], population density [@Nussey2007;@Douhard2013] and natural or anthropogenic environmental disturbance [@Reid2003; @Cartwright2014].
+Exposure to favourable environmental conditions during development and growth can confer fitness advantages later in life (so called 'silver spoon effects') [@Grafen1988; @Monaghan2008].
+Adult fitness can be affected by a range of early-life experiences, including the quality of the habitat and available resources [@Madsen2000; @VandePol2006; @Hayward2013], population density [@Nussey2007;@Douhard2013] and natural or anthropogenic environmental disturbance [@Reid2003; @Cartwright2014].
 Understanding the causes and consequences of these early-life experiences is key to understanding many ecological and evolutionary  processes, including  patterns of natural and sexual selection,  population growth rates and even local extinction [@Coulson2001; @Roach2014].
@@
 While it is clear that silver spoon effects can occur, we have little understanding of how and under what conditions early-life environments will affect adult fitness.
-Because the phenotypic consequences of an adverse environment can depend on an individual's initial  condition, and genetic or epigenetic makeup [@Hoffman2000; @Richards2006], the later-life consequences of a good or poor start are expected to vary among individuals, populations and species.
-Moreover, it is not always possible to fully quantify what constitutes a good or bad environment, and any 'hidden' environmental variation may obscure relationships between the early-life conditions that are measured and adult phenotypes.
+However , we do expect the later-life consequences of a good or poor start to vary among individuals, populations and species because the phenotypic consequences of an adverse environment can depend on an individual's initial condition, and genetic or epigenetic makeup [@Hoffman2000; @Richards2006].
+Moreover, it is not always possible to fully quantify what constitutes a good or bad environment, and any 'hidden' environmental variation may obscure relationships between the early-life conditions that are measured and the resulting adult phenotypes.
 Indeed, the pervasiveness of silver spoon effects varies between species [@Drummond2011], cohorts [@Reid2003] and sexes [@Wilkin2009].
-To better understand the later-life consequences of early-life experiences, we therefore need to understand, or at least be able to measure, how the environment differentially affects individuals within a population.
+We therefore need to understand, or at least be able to measure, how the environment affects individuals differentially within a population in order to elucidate the later-life consequences of early-life experience.
 
-A key process that has been associated with adult survival is the rate at which telomeres are shortened.
-Telomeres are protective caps on the ends of chromosomes which shorten with age [@Monaghan2006], and in response to oxidative stress levels which can be elevated  due to environmental factors [@VonZglinicki2002].
-When telomeres reach a critical shortness cells senesce [@Campisi2003], and the accumulation of these cells can result in organismal senescence and death  [@Wong2003].
-These links between telomere length and senescence have inspired a great deal of recent research into telomere ecology [reviewed in @Horn2010; @Haussmann2010; @Monaghan2014].
-While there is little direct evidence that the relationship between telomere dynamics and survival is causal [@Barrett2011; @Simons2015], there is now excellent evidence that telomeres can act as biomarkers of cost in wild populations, retaining signatures of ecological stress that are otherwise difficult to detect [@Monaghan2014; @Schultner2014; @Asghar2015].
+Telomeres - the protective caps on the ends of chromosomes - may provide a solution to this problem.
+Telomeres shorten with age [@Monaghan2006], and in response to oxidative stress, which can be elevated  due to environmental factors [@VonZglinicki2002].
+When telomeres become critically short cells senesce [@Campisi2003], and the accumulation of these cells can result in organismal senescence and death  [@Wong2003].
+This association between senescence and telomere length has inspired a great deal of recent research into telomere ecology [reviewed in @Horn2010; @Haussmann2010; @Monaghan2014].
+While there is little direct evidence that the relationship between telomere dynamics and survival is causal [@Barrett2011; @Simons2015], there is now excellent evidence that telomeres can act as biomarkers of cost in wild populations, providing a signature of the ecological stress that has been experienced and is otherwise difficult to detect [@Monaghan2014; @Schultner2014; @Asghar2015].
 
 There is evidence from a range of taxa that the greatest rate of telomere loss occurs in early life [e.g.
-@Frenck1998; @Haussmann2003], and that the extent of this telomere shortening is influenced by the conditions experienced early in life [@Price2013; @Monaghan2014; @Nettle2015; @Reichert2015].
-Importantly, early-life telomere dynamics are related to both short-term and late-life survival [@Heidinger2012; @Boonekamp2014], and to other parameters such as cognition [@Nettle2015a].
-However, the link between early-life conditions and late-life survival is limited to a few species, and thus poorly understood.
+@Frenck1998; @Haussmann2003], and that the extent of this telomere shortening is influenced by the conditions experienced during that period [@Price2013; @Monaghan2014; @Nettle2015; @Reichert2015].
+Importantly, early-life telomere dynamics have been associated with both short-term and late-life survival [@Heidinger2012; @Boonekamp2014], and with other parameters such as cognition [@Nettle2015a].
+However, to our knowledge no studies have simultaneously analysed how telomere dynamics, early-life conditions and late-life survival are all related in a natural setting.
 Moreover, how early-life telomere dynamics vary over spatial and temporal scales is not known.
@@
 The longitudinal study (since 1986) of the Seychelles warbler (*Acrocephalus sechellensis*) population on Cousin Island provides an excellent model system for studying senescence in the wild [reviewed in @Hammers2015].
-Due to the isolated nature of the study population and intensive field monitoring, we have excellent survival data and sampling material spanning many years (see methods, below).
-Ecological conditions and population density on Cousin are variable over space and time due to weather-induced changes in foliage cover and food availability [@VandeCrommenacker2011a].
-Seychelles warblers remain on their natal territories for at least six months, and variation in natal territory quality has been linked to changes in levels of oxidative stress [@VandeCrommenacker2011a].
-However, neither natal territory quality nor natal local density appear to be linked to early-life or adult survival [@Brouwer2006; @Hammers2013].
-Facultative cooperative breeding occurs in the Seychelles warbler [@Komdeur1994a; @Richardson2003a], and the presence of helping subordinates (but not non-helping subordinates) in the natal territory has been associated with increased survival later in life [@Brouwer2012].
+Due to the isolated nature of the study population and intensive field monitoring, we have unusually comprehensive survival data and tissue samples spanning many years (see Methods, below).
+Ecological conditions and warbler population density on Cousin vary across space and time due to weather-induced changes in foliage cover and food availability [@VandeCrommenacker2011a].
+Seychelles warblers remain on their natal territories for at least six months, and variation in the oxidative stress experienced by individuals is associated with natal  territory quality [@VandeCrommenacker2011a].
+However, neither early-life nor adult survival appear to be associated with natal territory quality or natal local density [@Brouwer2006; @Hammers2013].
+Facultative cooperative breeding occurs in the Seychelles warbler [@Komdeur1994a; @Richardson2003a], and the presence of helping subordinates (but not non-helping subordinates) in the natal territory is associated with increased survival later in life [@Brouwer2012].
 Lastly, we have an established protocol for assessing absolute telomere length in this species [@Barrett2012], and telomere length predicts survival independently of age in adult Seychelles warblers, suggesting that telomeres act as a biomarker of cost [@Barrett2013a].
-Thus, we have an excellent system in which to assess the costs of different social and environmental conditions experienced early in life, and to assess the later life consequences of early-life conditions.
+Thus, we have an excellent system in which to assess the costs of different social and environmental conditions experienced early in life, and to assess the later-life consequences of early-life conditions.
 
 In this study we test the prediction that telomeres link early-life environmental variation to late-life fitness.
-Because telomere dynamics are expected to reflect individual-level variation in the costs of early-life experiences, they may allow a more sensitive analysis of the effects of early-life environmental variation than would be possible with a direct comparison of the early-life environment and survival.
-With this in mind, we predict that: i) individuals raised in good quality conditions will have longer telomeres, and ii) having longer telomeres in early life is associated with greater survival.
+Because telomere dynamics are expected to reflect individual-level variation in the costs of early-life experiences, they may allow a more sensitive analysis of the effects of early-life environmental variation than would be possible with a direct comparison of how survival is affected by the early-life environment.
+With this in mind, we predict that: i) individuals raised in higher quality conditions will have longer telomeres and reduced telomere shortening, and ii) longer telomeres and lower rates of telomere shortening in early life are associated with greater survival.
 
@@
 115 territories lay one or, rarely, two or three eggs [@Komdeur1991].
-As a result of this low reproductive output, combined with a higher mortality rate among first-year birds [39%; @Brouwer2006], cohort sizes in the Seychelles warbler are small (<50).
+As a result of this low reproductive output, combined with higher mortality in first-year birds [39%; @Brouwer2006], cohort sizes in the Seychelles warbler are small (<50).
 
@@
 Each time a bird is caught on Cousin body mass and tarsus length are measured (to the nearest 0.1g and 0.1mm, respectively), and age is confirmed on the basis of eye colour [@Komdeur1991a] and previous captures [@Richardson2003b].
-A blood sample (ca 25 $\mu$l) is taken from each caught bird via brachial venipuncture, and stored in 1 ml of absolute ethanol in a 1.5 ml screw-cap microfuge tube at room temperature.
+A blood sample (*ca* 25 $\mu$l) is taken from each bird captured via brachial venipuncture, and stored at room temperature in 1 ml of absolute ethanol in a 1.5 ml screw-cap microfuge tube.
 
@@
 
-Telomere lengths were measured for `r nrow(FlSA)+nrow(chicks)+nrow(Loss)` samples.
+Telomere lengths were measured in `r nrow(FlSA)+nrow(chicks)+nrow(Loss)` samples.
 Of these, `r nrow(FlSA)+nrow(chicks)` were taken cross-sectionally from birds caught within their first year of life, between `r min(as.numeric(paste(juv$LayYear)))` and `r max(as.numeric(paste(juv$LayYear)))`; `r nrow(chicks)` were nestlings less than one month old (chicks), while the remaining `r nrow(FlSA)` were fledglings and subadults aged between two and twelve months (hereafter referred to as fledglings).
@@
 We performed all analyses using R version 3.0.1 [@RDevelopmentCoreTeam2011].
-Our sampling regime covers `r max(juv$LayYear)-min(juv$LayYear)` years, and temporal variation in rainfall and food availability on Cousin over this period is expected to result in variation  in a range of social and environmental variables [@Brouwer2012].
+Our sampling regime covers `r max(juv$LayYear)-min(juv$LayYear)` years, and temporal variation in rainfall and food availability on Cousin Island over this period is expected to result in variation  in a range of social and environmental variables [@Brouwer2012].
 This makes it difficult to disentangle how spatial and temporal processes differentially affect early-life telomere dynamics.
-To overcome this, we analysed how variation in early-life conditions was related to variation in early-life telomere length separately among and within breeding seasons.
+To overcome this, we analysed how variation in early-life telomere length was related to variation in early-life conditions both among and within breeding seasons.
 
-First, to analyse whether temporal variation in food availability drives population-level variation in telomere dynamics, we calculated average telomere length among all birds born in a given (major or minor) breeding season (i.e.
-a cohort), excluding cohorts where we had samples from $\leq$ 5 birds (n = `r nrow(juvseason)` cohorts).
+First, to analyse whether temporal variation in food availability drives population-level variation in telomere dynamics, we calculated mean telomere length among all birds born in a given (major or minor) breeding season (i.e.
+a cohort), excluding cohorts where the sample size $\leq$ 5 birds (n = `r nrow(juvseason)` cohorts).
 We tested for a difference in telomere length among cohorts using a one-way ANOVA, and then used linear regression to test whether variation in mean telomere length among cohorts is explained by temporal variation in insect abundance.
 
-We then conducted individual-level analyses to test how spatially varying early-life environmental and social conditions influence telomere length within cohorts, using general linear models.
-For these analyses, both response and explanatory variables were standardised by mean-centering within cohorts, thus controlling for any between-cohort temporal variation in telomere length and the social and ecological environment.
+We then conducted individual-level analyses to test how spatial variation in early-life environmental and social conditions influenced telomere length within cohorts, using general linear models.
+For these analyses, both response and explanatory variables were standardised by mean-centring within cohorts, thus controlling for any between-cohort temporal variation in telomere length and the social and ecological environment.
 As explanatory variables we included age class (nestling or fledgling), tarsus length, sex, territory quality and the number of helping and non-helping subordinate birds present in the natal territory.
@@
 We created a full model containing all of the above terms, and a top model set was then defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011].
-We report model averaged coefficients, confidence intervals, and 'relative importance', which reflects the relative weights of each predictor variable across the top model set.
+We report model-averaged coefficients, confidence intervals and 'relative importance', which reflects the relative weights of each predictor variable across the top model set.
 Model averaging was carried out using the MuMIn package (version 1.10.5) in R [@Barton2012].
 
-For individuals with longitudinal data (n = `r nrow(Loss)`) we calculated a rate of telomere loss between the first-year sample and earliest adult sample by subtracting the adult telomere length from the early-life telomere length and dividing this difference by the length of time (in days) between sampling events.
+For individuals with longitudinal data (*n* = `r nrow(Loss)`) we calculated a rate of telomere loss between the first-year sample and earliest adult sample by subtracting the adult telomere length from the early-life telomere length and dividing this difference by the length of time (in days) between sampling events.
 To account for regression to the mean effects, we applied a correction based on correlations among samples within individuals following Verhulst *et al.* [-@Verhulst2013].
@@
 Finally, we analysed the effects of telomere dynamics on survival, again by looking at variation both within and among cohorts.
-First, we tested the prediction that cohort-level telomere dynamics are associated with survival by comparing the mean telomere length and rate of loss of each cohort with the mean lifespan of all individuals in that cohort, using linear regression.
-We then used parametric survival analysis to test whether individual-level telomere length and rate of loss were related to to survival.
+First, we tested the prediction that cohort-level telomere dynamics are associated with survival by comparing the mean lifespan of all individuals within each cohort with mean telomere length and rate of loss in that cohort, using linear regression.
+We then used parametric survival analysis to test whether survival was related to individual-level telomere length and rate of loss.
 Again, we mean-centred variables within cohorts to eliminate between-season temporal variation in telomeres and survival.
-The survival time distribution was chosen by comparing the fit of a range of distributions (exponential, Weibull, log-normal, gamma, gompertz and generalized F) against the observed survival data, by visually inspecting plots and by comparing AIC values.
-As explanatory variables of survival we included early-life telomere length/loss (we ran a separate analysis for each, using the cross-sectional and longitudinal data, respectively), plus the factors identified as being related to telomere length in the previous analysis.
-With this approach we aimed to assess the effect of early-life telomere length on survival while controlling for potentially confounding variables that may be correlated with telomere dynamics.
+The survival time distribution was chosen by comparing the fit of the observed survival data against a range of distributions (exponential, Weibull, log-normal, gamma, Gompertz and generalized F) by visually inspecting plots and comparing AIC values.
+As explanatory variables of survival, we included early-life telomere length/loss (we ran a separate analysis for each, using the cross-sectional and longitudinal data, respectively), plus the factors identified as being related to telomere length in the previous analysis.
+With this approach we aimed to assess the relationship between early-life telomere length and survival, while controlling for potentially confounding variables that might be correlated with telomere dynamics.
 

### 2015-10-16 Manuscript/SW_early_life_telomeres.Rmd
Final edits before submission to Evolution

@@
 
-###Telomeres reveal silver spoon effects in a wild population
+**Telomeres reveal silver spoon effects in a wild population**
 
@@
 
-a. School of Biological Sciences, University of East Anglia, Norwich Research Park, NR4 7TJ, United Kingdom
-b. Department of Zoology, Edward Grey Institute, University of Oxford, Oxford, UK
-c. Groningen Institute for Evolutionary Life Sciences, University of Groningen, Groningen, The Netherlands
-d. Department of Animal and Plant Sciences, University of Sheffield, Sheffield, UK
-e. Nature Seychelles, Roche Caiman, Mahé, Republic of Seychelles
+1. School of Biological Sciences, University of East Anglia, Norwich Research Park, NR4 7TJ, United Kingdom
+2. Department of Zoology, Edward Grey Institute, University of Oxford, Oxford, UK
+3. Groningen Institute for Evolutionary Life Sciences, University of Groningen, Groningen, The Netherlands
+4. Department of Animal and Plant Sciences, University of Sheffield, Sheffield, UK
+5. Nature Seychelles, Roche Caiman, Mahé, Republic of Seychelles
 
-**Correspondence:** Lewis Spurgin: lewisspurgin@gmail.com; David Richardson: david.richardson@uea.ac.uk
+Lewis Spurgin: lewisspurgin@gmail.com; Kat Bebbington: K.Bebbington@uea.ac.uk Eleanor Fairfield: E.Fairfield@uea.ac.uk; Jan Komdeur: j.komdeur@rug.nl; Terry Burke: t.a.burke@sheffield.ac.uk; Hannah Dugdale: h.l.dugdale@rug.nl; David Richardson: david.richardson@uea.ac.uk
+
+**Running head:** Telomeres and silver spoons
 
 ###Abstract
-Understanding the links between early-life experiences and later-life survival (so-called "silver spoon effects") is fundamental to understanding life-history evolution.
-Telomeres – the protective caps at the ends of chromosomes – shorten  in response to stress, and telomere shortening in early life is correlated with reduced later-life survival.
-Thus, telomere dyamics may help us quantify individual variation in early-life costs and silver spoon effects.
-We tested how telomere dynamics are related to spatiotemporal variation in early-life conditions and later-life survival  using a long-term study of the Seychelles warbler (*Acrocephalus sechellensis*).
+Understanding the links between early-life experiences and later-life survival ("silver spoon effects") is fundamental to understanding life-history evolution.
+Telomeres, the protective caps at the ends of chromosomes, shorten  in response to oxidative stress, and telomere shortening is correlated with reduced survival.
+Thus, telomere dynamics may help us quantify individual variation in early-life costs, and thus enhance our understanding of how and why silver spoon effects occur.
+We tested how telomere dynamics are related to spatiotemporal variation in early-life conditions and later-life survival in the Seychelles warbler (*Acrocephalus sechellensis*), across multiple cohorts spanning 14 years.
 We show that telomere lengths and rates of shortening vary across cohorts, and that individuals born in breeding seasons with higher food availability had longer telomeres and reduced telomere shortening.
-Within cohorts, we found that telomere length in nestlings was negatively related to tarsus length (which probably  represents nestling growth), and positively related to the number of helpers present in the natal territory.
+Within cohorts, we found that telomere length in nestlings was negatively related to tarsus length (which  represents nestling growth), and positively related to the number of helpers present in the natal territory.
 Finally, we found that increased survival later in life was associated with longer telomeres and reduced telomere shortening rates in early life.
-Our results show that telomeres can both measure variation in early-costs and predict late-life survival in wild populations.
+Our results show that telomeres can both measure variation in early-life costs and predict late-life survival in wild populations.
 
-**Keywords:** Life-history; Seychelles warbler; Senescence; Survival; Telomeres
+**Keywords:** Life-history; Seychelles warbler; Senescence; Survival
+
+**Data archival location:** This manuscript was written in R Markdown (http://rmarkdown.rstudio.com/).
+All data and scripts required to reproduce the manuscript, figures and analyses will be made available on GitHub.
 
@@
 ###Introduction
-Exposure to favourable environmental conditions during development and growth can confer fitness advantages later in life (so called 'silver spoon effects') [@Grafen1988; @Monaghan2008].
+Exposure to favourable environmental conditions during development and growth can confer fitness advantages later in life [so called 'silver spoon effects'; @Grafen1988; @Monaghan2008].
 Adult fitness can be affected by a range of early-life experiences, including the quality of the habitat and available resources [@Madsen2000; @VandePol2006; @Hayward2013], population density [@Nussey2007;@Douhard2013] and natural or anthropogenic environmental disturbance [@Reid2003; @Cartwright2014].
@@
 When telomeres become critically short cells senesce [@Campisi2003], and the accumulation of these cells can result in organismal senescence and death  [@Wong2003].
-This association between senescence and telomere length has inspired a great deal of recent research into telomere ecology [reviewed in @Horn2010; @Haussmann2010; @Monaghan2014].
+This association between senescence and telomere length has inspired a great deal of recent research into telomere evolutionary ecology [reviewed in @Horn2010; @Haussmann2010; @Monaghan2014].
 While there is little direct evidence that the relationship between telomere dynamics and survival is causal [@Barrett2011; @Simons2015], there is now excellent evidence that telomeres can act as biomarkers of cost in wild populations, providing a signature of the ecological stress that has been experienced and is otherwise difficult to detect [@Monaghan2014; @Schultner2014; @Asghar2015].
@@
 Individuals are usually ringed in their first year of life, and so are of known age.
-They are then followed throughout their lives, and as they are non-migratory endemics naturally confined to the island [@Komdeur2004a], a biannual census of birds on Cousin during each breeding season gives accurate measures of local density and individual survival [@Barrett2013a].
+They are then followed throughout their lives, and as they are non-migratory endemics naturally confined to the island [@Komdeur2004a], a biannual census of birds on Cousin during each breeding season gives accurate measures of local density, social status (e.g.
+breeding male/female, helping subordinate, non-helping subordinate) and individual survival [@VandeCrommenacker2011; @Barrett2013a].
 The isolated nature of the Cousin population is a key advantage of the system for analyses involving survival, which in other systems are often confounded by emigration [see @Ergon2014 for a recent discussion].
@@
 #####Molecular methods
-Genomic DNA was extracted from a ~2 mm^2^ flake of preserved blood using the DNeasy Blood and Tissue Kit (Qiagen), following the manufacturer's protocol, with the modification of overnight lysis at 37^o^C and a final DNA elution volume of 80 $\mu$l.
+For each sample, genomic DNA was extracted from a ~2 mm^2^ flake of preserved blood using the DNeasy Blood and Tissue Kit (Qiagen), following the manufacturer's protocol, with the modification of overnight lysis at 37^o^C and a final DNA elution volume of 80 $\mu$l.
 DNA concentration and purity were quantified using a NanoDrop 8000 Spectrophotometer (ThermoScientific), and DNA integrity was validated by visualization with ethidium bromide after electrophoresis on a 1.2% agarose gel.
@@
 First, to analyse whether temporal variation in food availability drives population-level variation in telomere dynamics, we calculated mean telomere length among all birds born in a given (major or minor) breeding season (i.e.
-a cohort), excluding cohorts where the sample size $\leq$ 5 birds (n = `r nrow(juvseason)` cohorts).
+a cohort), excluding cohorts where the sample size $\leq$ 5 birds (leaving a sample size of n = `r nrow(juvseason)` cohorts).
 We tested for a difference in telomere length among cohorts using a one-way ANOVA, and then used linear regression to test whether variation in mean telomere length among cohorts is explained by temporal variation in insect abundance.
@@
 We then used parametric survival analysis to test whether survival was related to individual-level telomere length and rate of loss.
-Again, we mean-centred variables within cohorts to eliminate between-season temporal variation in telomeres and survival.
+For the latter analysis, again we mean-centred variables within cohorts to eliminate between-season temporal variation in telomeres and survival.
 The survival time distribution was chosen by comparing the fit of the observed survival data against a range of distributions (exponential, Weibull, log-normal, gamma, Gompertz and generalized F) by visually inspecting plots and comparing AIC values.
@@
 Mean ($\pm$ standard error) telomere length in first-year birds was `r round2(mean(FlSAall$TLKB))` $\pm$ `r round2(se(FlSAall$TLKB))` kb.
-This is lower than the mean for one year-old birds based on cross-sectional data (`r round2(mean(subset(dd,Age==1)$TLKB))` $\pm$ `r round2(se(subset(dd,Age==1)$TLKB))` kb).
-However, when we considered age classes within the first year of life separately, we found substantial age-related variation in telomere length (F = `r round2(anova(lm(TLKB~Fledged,data = juv))[1,4])`; *P* = `r round3(anova(lm(TLKB~Fledged,data = juv))[1,5])`).
+This was lower than the mean for one-year-old birds based on cross-sectional data (`r round2(mean(subset(dd,Age==1)$TLKB))` $\pm$ `r round2(se(subset(dd,Age==1)$TLKB))` kb).
+However, when we considered age classes within the first year of life separately, we found substantial age-related variation in telomere length (F = `r round2(anova(lm(TLKB~Fledged,data = juv))[1,4])`; P = `r round3(anova(lm(TLKB~Fledged,data = juv))[1,5])`).
 Nestlings had longer telomeres than any other age group, but there was an apparent increase in average telomere length from the fledgling to adult stages (Fig.
 1A).
-A longitudinal analysis of telomere loss showed that nestlings and fledglings both lost telomeres early in life, but that there was no difference in the rate of telomere loss among age classes (F = `r round2(anova(lm(TROC~Fledged,data = Loss))[1,4])`; *P* = `r round2(anova(lm(TROC~Fledged,data = Loss))[1,5])`; Fig.
-1B).
+However, a longitudinal analysis of telomere loss showed that nestlings and fledglings both lost telomeric DNA early in life, and that there was no difference in the rate of telomere loss between the two age classes (F = `r round2(anova(lm(TROC~Fledged,data = Loss))[1,4])`; P = `r round2(anova(lm(TROC~Fledged,data = Loss))[1,5])`; Fig.
+1B), suggesting that the cross-sectional results represent selective disappearance.
 
-#####Temporal variation in early-life telomere dynamics
-Early-life telomere length varied significantly among breeding seasons (one-way ANOVA, F = `r round2(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,4])`; *P* = `r round3(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,5])`), and this variation in average telomere length over breeding seasons was positively related to island-wide food availability (linear regression, R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,5])`; Fig.
+#####Among-cohort variation in early-life telomere dynamics
+Early-life telomere length varied significantly among breeding seasons (one-way ANOVA, F = `r round2(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,4])`; P = `r round3(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,5])`), and this variation in mean telomere length over breeding seasons was positively related to island-wide food availability (linear regression, R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,4])`; P = `r round3(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,5])`; Fig.
 2A).
-Running the regression of cohort-level telomere length and food availability separately for each age class revealed a positive, significant relationship in fledglings (R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=flseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data=flseason,na.action=na.exclude))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKBmean,data=flseason,na.action=na.exclude))[1,5])`), but no relationship in nestlings (R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=chickseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data=chickseason,na.action=na.exclude))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKBmean,data=chickseason,na.action=na.exclude))[1,5])`), indicating selective disappearance.
+Running the regression of cohort-level telomere length and food availability separately for the two age classes revealed a positive, significant relationship in fledglings (R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=flseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data=flseason,na.action=na.exclude))[1,4])`; P = `r round3(anova(lm(Insect~TLKBmean,data=flseason,na.action=na.exclude))[1,5])`), but no relationship in nestlings (R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=chickseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data=chickseason,na.action=na.exclude))[1,4])`; P = `r round3(anova(lm(Insect~TLKBmean,data=chickseason,na.action=na.exclude))[1,5])`).
 
-The rate of telomere shortening in early life (taken from the smaller subset of birds with longitudinal samples) did not vary significantly among breeding seasons (F = `r round2(anova(lm(TROC~factor(FieldPeriodID),data = Loss))[1,4])`; *P* = `r round2(anova(lm(TROC~factor(FieldPeriodID),data = Loss))[1,5])`); however, what variation there was in rates of telomere shortening could be explained by temporal variation in food availability (R^2^ = `r round2(summary(lm(Insect~TROCmean,data=loss.season,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TROCmean,data = loss.season,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Insect~TROCmean,data = loss.season,na.action=na.exclude))[1,5])`).
+The rate of telomere shortening in early life (taken from the smaller subset of birds for which we had longitudinal samples) also varied significantly among breeding seasons (F = `r round2(anova(lm(TROC~factor(LayYear),data = Loss))[1,4])`; P = `r round2(anova(lm(TROC~factor(LayYear),data = Loss))[1,5])`), and rates of telomere shortening could be explained by temporal variation in food availability (R^2^ = `r round2(summary(lm(Insect~TROCmean,data=loss.season,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TROCmean,data = loss.season,na.action=na.exclude))[1,4])`; P = `r round2(anova(lm(Insect~TROCmean,data = loss.season,na.action=na.exclude))[1,5])`).
 This pattern corresponded with the cross-sectional data: individuals born in seasons with low food availability lost telomeres, on average, at a faster rate than those born in years with high food availability (Fig.
 2B).
-Unfortunately we did not have sufficient sample sizes within cohorts to run this analysis separately among age classes.
+Unfortunately we did not have sufficient sample sizes within cohorts to run this analysis separately between age classes.
 
-#####Spatial variation in early-life telomere dynamics
+#####Within-cohort variation in early-life telomere dynamics
 The top model explaining within-season variation in early-life telomere length contained age class, tarsus length and the number of helpers present in the natal territory, as well as interactions between age class and both tarsus length and the number of helpers (Table S1).
@@
 
-For the subset of individuals with longitudinal data, we tested whether the factors previously identified as being related to early-life telomere length (i.e.
+For the subset of individuals for which we had longitudinal data, we tested whether the factors previously identified as being related to early-life telomere length (i.e.
 tarsus length and number of helpers) were also associated with differences in rates of early-life telomere loss.
 Due to a limited sample size, we considered all age classes together for this analysis.
-We found that none of these variables were associated with differences in telomere loss (all *P* > 0.05).
+We found that none of these variables were associated with differences in telomere loss (all P > 0.05).
 
-#####early-life telomere dynamics and survival
-At the population-level, there was no relationship between telomere length and lifespan (R^2^ = `r round3(summary(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))[1,5])`; Fig.
+#####Early-life telomere dynamics and survival
+At the population-level, there was no relationship between lifespan and telomere length (R^2^ = `r round3(summary(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))[1,4])`; P = `r round2(anova(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))[1,5])`; Fig.
 4A).
-Testing this relationship separately for each age class revealed that there was no age-specific relationship between cohort-level telomere length and lifespan (both *P* > 0.2).
-There was a negative relationship between population-level telomere loss and lifespan, although this was marginally non-significant (R^2^ = `r round3(summary(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))[1,5])`; Fig.
+Testing this relationship separately for each age class revealed that there was no age-specific relationship between lifespan and cohort-level telomere length (both P > 0.2).
+There was a negative relationship between population-level telomere loss and lifespan, although this was marginally non-significant (R^2^ = `r round3(summary(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))[1,4])`; P = `r round2(anova(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))[1,5])`; Fig.
 4B).
@@
 For the individual-based survival analysis, a log-normal survival model best fitted the Seychelles warbler data (AIC = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="lnorm")$AIC)`, AIC of next best distribution (Weibull) = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="weibull")$AIC)`).
-A survival model including all first year birds showed that early-life telomere length had a positive effect on survival, but this relationship was marginally non-significant (estimate = `r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`).
-Running this analysis among age classes revealed differential survival effects.
+A survival model including all first-year birds showed that early-life telomere length had a positive effect on survival, but this relationship was marginally non-significant (estimate = `r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`).
+Running this analysis between age classes revealed differential survival effects.
 Nestling telomere length had no effect on survival  (estimate = `r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`; Fig.
 5A), but telomere length at the fledgling stage was significantly related to survival (estimate = `r round3(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`).
-This effect was positive, with longer telomeres at the fledgling stage associated with increased survival later in life (Fig.
+This effect was positive, with increased survival later in life associated with longer telomeres (Fig.
 6B).
-Finally, using the longitudinal data, we found no effect of early-life telomere shortening on subsequent survival (estimate = `r round2(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,1])`, CI = `r round3(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,3], lessthan = F)`).
+Finally, using the longitudinal data, we found no effect of early-life telomere shortening on subsequent survival, although again the estimate was positive, and confidence intervals only marginally overlapped zero (estimate = `r round2(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,1])`, CI = `r round3(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,3], lessthan = F)`).
 
 ###Discussion
-Here we show that telomeres link early-life conditions to later-life survival in the Seychelles warbler.
+Here we show that telomeres are related to both early-life conditions and later-life survival in the Seychelles warbler.
 We first show that telomere length and rates of shortening in early life are subject to strong cohort effects, and that cohort-level telomere dynamics can be explained by island-wide temporal variation in food availability.
@@
 
-Our finding that food availability is related to early-life telomere length and rate of shortening adds to a growing body of evidence from humans and wild animals showing that the natal environment can have pronounced impacts on early-life telomere dynamics [reviewed in @Price2013; @Monaghan2014].
-However, very few studies have shown that temporal variation in environmental conditions affects telomere dynamics in natural populations, and the studies that have done so are limited to just two seasons [@Mizutani2013; @Watson2015].
-The long-term Seychelles warbler dataset has allowed us, using many more cohorts than has been used in other studies, to show that temporal variation in environmental conditions does indeed affect telomere dynamics, and that this effect can be seen at the population level.
-Interestingly, in the Seychelles warbler annual variation in food availability is not directly linked to survival [@Brouwer2006; @Hammers2013].
+Our finding that early-life telomere length and rate of shortening are related to food availability adds to a growing body of evidence from humans and wild animals showing that the natal environment can have pronounced impacts on early-life telomere dynamics [reviewed in @Price2013; @Monaghan2014].
+However, very few studies have shown that temporal variation in telomere dynamics occurs in natural populations, and the studies that have done so were limited to just two seasons [@Mizutani2013; @Watson2015], making it impossible to statistically evaluate whether this variation is driven by temporal variation in the environment.
+The long-term Seychelles warbler dataset has allowed us, using many more cohorts than has been used in other studies, to show that temporal variation in environmental conditions does affect telomere dynamics, and that this effect can be seen at the population level.
+Interestingly, in the Seychelles warbler survival is not directly linked to annual variation in food availability [@Brouwer2006; @Hammers2013].
 It has been suggested that telomeres act as biomarkers of cost in natural populations, as they are able to capture individual and cohort-level variability in the costs experienced as a result of environmental pressures [@Asghar2015].
-Our results confirm that telomeres can detect hidden costs in natural populations that are not detectable using life-history and environmental data alone
+Our results show that telomeres can detect hidden costs in natural populations that are not detectable using life-history and environmental data alone.
 
-We suspect the novelty of our findings of temporal, environmentally-induced variation in telomere dynamics within a population is more due to a lack of available long-term datasets with telomere screening, rather than the Seychelles warbler being unique.
+We suspect that the novelty in our finding temporal, environmentally-induced variation in telomere dynamics within a population is more due to a lack of available long-term datasets with telomere screening, rather than the Seychelles warbler being unique.
 Indeed, the environment on Cousin is benign in comparison with many regions outside the tropics, where populations undergo large fluctuations in size [e.g.
@@
 More research is now needed within and across multiple cohorts and populations to better understand the complexity of telomere evolutionary ecology.
-Secondly, our findings raise the interesting prospect of using telomeres as indicators of population health - an approach that would be useful to a broad range of conservation and animal health practitioners.
+Second, our findings raise the interesting prospect of using telomeres as indicators of population health - an approach that would be useful to a broad range of conservation and animal health practitioners.
 However, further research across populations with different environments and genetic histories is required in order to provide further insight into the usefulness of telomeres as biomarkers of population health.
 
-We found that within cohorts, tarsus length and the number of helpers were the most important predictors of telomere length in Seychelles warblers, and that these effects occurred specifically at the nestling stage.
-The tarsus effect most likely reflects the fact that in passerine birds tarsus length is correlated with age during the nestling stage [@Ricklefs1976].
+We found that, within cohorts, tarsus length and the number of helpers were the most important predictors of telomere length in Seychelles warblers, and that these effects occurred specifically at the nestling stage.
+The tarsus effect probably reflects the fact that in passerine birds tarsus length is correlated with age during the nestling stage [@Ricklefs1976].
 Telomere loss is most rapid early in life due to ongoing cell replication [@Frenck1998], and a negative correlation between telomere length and body size in early life is therefore expected.
@@
 Ours is the first study to show that the presence of helpers has a beneficial effect on offspring in terms of their telomeres.
-In the Seychelles warbler the survival benefits to a nestling of their being helpers in the natal territory is well established [@Komdeur1994;@Brouwer2012], and that this effect is detected in terms of differential telomere length, is therefore encouraging.
-However, we found no evidence that the number of helpers was related to telomere loss - this is perhaps surprising as telomere loss is expected to be a better indicator of stress than telomere length *per se* [@Boonekamp2014].
-Our (lack of) finding here is probably due to the limited nature of our longitudinal dataset, both in terms of sample size and resolution (i.e.
+In the Seychelles warbler the survival benefits to a nestling of having helpers in the natal territory is well established [@Komdeur1994;@Brouwer2012], and that the beneficial effect of helpers is detectable in terms of differential telomere length, is therefore encouraging.
+However, we found no evidence that telomere loss was related to the number of helpers - this is perhaps surprising as telomere loss is expected to be a better indicator of stress than telomere length *per se*, and effects of telomere length may occur as a result of correlated to telomere loss [@Boonekamp2014].
+However, our longitudinal dataset was limited, both in terms of sample size and resolution (i.e.
 time between sampling events).
-Seychelles warblers are rarely been sampled multiple times within their first year of life, so much of the telomere shortening that occurs in early life will be missed with our sampling regime.
-It is likely, therefore that the effects of environmental variation on telomere loss will only be detected when they are very strong using this dataset.
+Seychelles warblers are rarely sampled more than once within their first year of life, so much of the telomere shortening that occurs in early life will be missed with our sampling regime.
+It is likely, therefore that only very strong effects of environmental variation on telomere loss will  be detected in this dataset.
 Thus telomere length constitutes a better indicator of early-life stress.
 
-While the relationship between telomeres and mortality in adults has been established for some time [@Cawthon2003], only recently has the link between early-life telomere dynamics and later-life survival been studied.
-In captive zebra finches, juvenile telomere length predicts late-life survival [@Heidinger2012], and in wild bird populations early-life telomere dynamics have been linked to survival during the nestling phase [@Watson2015], and survival to adulthood [@Boonekamp2014].
-Our results are, to our knowledge, the first to demonstrate that early-life telomere length is associated with later-life survival in a wild population.
-This effect can be seen at the individual level, and (to a lesser extent) at the cohort level, although the nature of the relationship varies according to the age class at which telomeres are measured.
-At the individual level, we found that telomere length in fledglings, but not nestlings, was related to survival.
-This is unsurprising, as only by the fledgling stage is telomere length likely to reflect the accumulation of early-life stress (see also previous paragraph).
-The relationships between population-level telomere length and lifespan were in the expected direction (Fig.
+While the relationship between mortality and telomeres in adults has been established for some time [@Cawthon2003], only recently has the link between later-life survival and early-life telomere dynamics been studied.
+In captive zebra finches, juvenile telomere length predicts late-life survival [@Heidinger2012], and in wild bird populations survival to the nestling phase [@Watson2015], and survival to adulthood [@Boonekamp2014], have been linked with early-life telomere dynamics.
+Our results are, to our knowledge, the first to demonstrate that later-life survival is related to early-life telomere length in a wild population.
+This effect can be seen at the individual level, and (to a lesser extent) at the cohort level, although the nature of the relationship varies according to the age class in which telomeres are measured.
+At the individual level, we found that telomere length was related to survival in fledglings, but not nestlings.
+This is not surprising, as only by the fledgling stage is telomere length likely to reflect the accumulation of early-life stress (see also previous paragraph).
+The relationships between lifespan and population-level telomere length were in the expected direction (Fig.
 4), but not significant, which is perhaps not surprising given that even though we have a long-term dataset, we are still restricted in terms of sample size when conducting cohort-level analyses.
@@
 Here we have considered survival, but telomere length and shortening in early life may also be linked to other components of fitness [reviewed in @Monaghan2014].
-Reproductive senescence occurs in the Seychelles warbler [reviewed in @Hammers2015], making this system well suited to examining how telomere length affects lifetime reproductive success.
+Reproductive senescence occurs in the Seychelles warbler [reviewed in @Hammers2015], making this system well suited to examining how telomere length predicts lifetime reproductive success.
 A further avenue for future research will be examining, in a quantitative genetic framework, how genetic and environmental components, and their interactions, affect telomere dynamics and senescence in natural populations [@Becker2015; @Asghar2014].
@@
 
-###Data accessibility
-This manuscript was written in R Markdown (http://rmarkdown.rstudio.com/). All data and scripts required to reproduce the manuscript, figures and analyses will be made available on GitHub.
-
 ###Acknowledgements
-Nature Seychelles facilitate the long-term Seychelles warbler project, allowing us to stay on Cousin Island, and providing accommodation and facilities during field work.
+We thank Nature Seychelles for facilitating the long-term Seychelles warbler project.
 The Seychelles Bureau of Standards and Department of Environment gave permission for sampling and fieldwork.
@@
 We thank everyone who has helped in the field, and the current Seychelles warbler research group for useful discussions.
-This work was funded by two Natural Environment Research Council (NERC) grants to DSR (NE/F02083X/1 and NE/K005502/1) on which JK, TB and HLD were project partners.
+This work was funded by two Natural Environment Research Council (NERC) grants to DSR (NE/F02083X/1 and NE/K005502/1).
 HLD was funded by a NERC fellowship.

### 2015-10-16 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Final edits before submission to Evolution

@@
 opts_knit$set(root.dir = '..')
-opts_chunk$set(echo=FALSE,
-               warning=FALSE,
+opts_chunk$set(include=FALSE,
                message=FALSE,
-               fig.path='../Figures/',
-               fig.show='asis',
-               dev=c('png','pdf'))
+               fig.show='hide')
 
@@
 
-**Figure 1** Telomere length (mean $\pm$ s.e.) and age in Seychelles warblers, based on cross-sectional **(A)** and longitudinal **(B)** data.
-In **B**, telomere loss is a standardised and corrected rate (see methods for details) based on the difference in telomere length from the sample taken from birds in their first year of life, and the first available sample as an adult.
+**Figure 1** Telomere length (mean $\pm$ s.e.) in relation to age in Seychelles warblers, based on cross-sectional **(A)** and longitudinal **(B)** data.
+In **B**, telomere loss is a standardised and corrected rate (see Methods for details) based on the difference in telomere length in the sample taken from birds in their first year of life, and the first available sample as an adult.
 Values above zero indicate telomere shortening in early life.
@@
 1 and main text for details).
-Points and error bars represent mean and standard error telomere length for all birds born in each main (black) or minor (grey) breeding season, in relation to island-wide food availability.
+Points and error bars represent mean and standard error of telomere length for all birds born in each main (black) or minor (grey) breeding season, in relation to island-wide food availability in that season.
 Lines and shaded areas represent fitted values and 95% confidence limits from a linear regression.
@@
 Numbers in brackets are the relative importance of each term in the top model set (see main text for details; TQ = territory quality).
-**B** Presence/absence of helpers and **c** tarsus length in relation to raw telomere length in nestlings (black, solid lines) and fledglings (grey, dashed lines).
-Points and error bars in **B** are mean $\pm$ s.e., and the line and shaded areas from **C** represent fitted values and 95% confidence limits from a linear regression.
+**B,C** Raw telomere length Presence/absence of helpers (**B**) and tarsus length (**C**) in nestlings (black, solid lines) and fledglings (grey, dashed lines).
+Points and error bars in **B** are means $\pm$ s.e., and the line and shaded areas in **C** represent fitted values and 95% confidence limits from a linear regression.
 
-**Figure 4** Temporal, cohort-level variation in early-life telomere dynamics and lifespan in the Seychelles warbler, based on cross-sectional **(A)** and longitudinal **(B)** data (see Fig.
+**Figure 4** Temporal, cohort-level variation in lifespan in relation to early-life telomere dynamics in the Seychelles warbler, based on cross-sectional **(A)** and longitudinal **(B)** data (see Fig.
 1 and main text for details).
-Points and error bars represent means and $\pm$ s.e., respectively, of all birds born in each main (black) or minor (grey) breeding season.
+Points and error bars represent means and $\pm$ s.e., respectively, of all birds born in each main (black, solid lines) or minor (grey, dashed lines) breeding season.
 Lines and shaded areas represent fitted values and 95% confidence limits from a linear regression.
 
-**Figure 5** Kaplan-Meier curves showing the relationship between telomere length and survival for individual Seychelles warblers measured as chicks (**A**), and fledglings  (**B**).
+**Figure 5** Kaplan-Meier curves showing the relationship between survival and telomere length for individual Seychelles warblers measured as chicks (**A**), and fledglings  (**B**).
 Telomere length is binned into groups here for visualisation purposes only (long and short = greater than or less than median telomere length, respectively).
 
-PAGEBREAK
-
-**Figure 1**
-
-```{r Figure 1,fig.width=4,fig.height=8}
+```{r Figure 1}
 ######################
@@
 
+pdf('Figures/Figure 1.pdf',
+    width = 4,
+    height = 8,
+    paper = 'a4')
 multiplot(Fig1A,Fig1B)
+dev.off()
 
@@
 
-PAGEBREAK
-
-**Figure 2**
-
-```{r Figure 2,fig.width=4,fig.height=8}
+```{r Figure 2}
 
@@
   stat_smooth(method = 'lm',se = T,col = 'black') +
-  scale_color_manual(values = c('black',grey(0.4))) +
+  scale_color_manual(values = c('black',grey(0.5))) +
   annotate('text',label = 'A',x=-1.5,y=8)
@@
   stat_smooth(method = 'lm',se = T,  col = 'black') +
-  scale_color_manual(values = c('black',grey(0.4))) +
+  scale_color_manual(values = c('black',grey(0.5))) +
   annotate('text',label = 'B',x=-1.5,y=0.0075)
 
+pdf('Figures/Figure 2.pdf',
+    width = 4,
+    height = 8,
+    paper = 'a4')
 multiplot(Fig2A,Fig2B)
+dev.off()
 
@@
 
-PAGEBREAK
-
-**Figure 3**
-
-```{r Figure 3,fig.width=9,fig.height=9}
+```{r Figure 3}
 
@@
                   position = position_dodge(width = 0.2)) +
-  scale_color_manual(values = c('black',grey(0.4))) +
+  scale_color_manual(values = c('black',grey(0.5))) +
   annotate('text',label = 'B',x=0.5,y=8)
@@
   stat_smooth(method = 'lm',se = T) +
-  scale_color_manual(values = c('black',grey(0.4))) +
+  scale_color_manual(values = c('black',grey(0.5))) +
   annotate('text',label = 'C',x=19,y=15)
 
+pdf('Figures/Figure 3.pdf',
+    width = 8,
+    height = 8,
+    paper = 'a4')
 multiplot(Fig3A,Fig3B,Fig3C,layout=matrix(c(1,1,2,3),2,2,byrow=TRUE))
+dev.off()
 
 ```
-
-PAGEBREAK
-
-**Figure 4**
 
@@
         legend.position = 'none') +
-  scale_color_manual(values = c('black',grey(0.4))) +
+  scale_color_manual(values = c('black',grey(0.5))) +
+  stat_smooth(method = 'lm',se = T,col = 'black',fullrange = T) +
   annotate('text',x=2.5,y=10,label='A')
@@
         legend.position = 'none') +
-  scale_color_manual(values = c('black',grey(0.4))) +
+  scale_color_manual(values = c('black',grey(0.5))) +
   stat_smooth(method = 'lm',se = T,col = 'black',fullrange = T) +
@@
 
+pdf('Figures/Figure 4.pdf',
+    width = 4,
+    height = 8,
+    paper = 'a4')
 multiplot(Fig4A,Fig4B)
+dev.off()
 
 ```
-
-PAGEBREAK
-
-**Figure 5**
 
@@
 s1 <- survfit(Surv(time = Lifespan+1, event = Died)~TLF, data = chicks)
-Fig5A <- ggsurv(s1,plot.cens=F) + 
+Fig5A <- ggsurv(s1,plot.cens=F,lty.est=c(1,2)) + 
   theme_classic() +
@@
         axis.title.x = element_text(size = 16),
-        legend.position = c(0.8,0.8),
-        legend.text = element_text(size = 12),
-        legend.title = element_blank()) +
+        legend.position = 'none') +
   scale_colour_manual(name = 'TL',
-                      values = c('black',grey(0.4))) +
-  guides(linetype=F) +
+                      values = c('black',grey(0.5))) +
+  guides(colour=F) +
   annotate('text',x=0.5,y=1.1,label='A')
@@
 s2 <- survfit(Surv(time = Lifespan+1, event = Died)~TLF, data = FlSAall)
-Fig5B <- ggsurv(s2,plot.cens=F) + 
+Fig5B <- ggsurv(s2,plot.cens=F,lty.est=c(1,2)) + 
   theme_classic() +
@@
         axis.title.x = element_text(size = 16),
-        legend.position = 'none',
-        legend.text = element_text(size = 12),
-        legend.title = element_blank()) +
+        legend.position = 'none') +
   scale_colour_manual(name = 'TL',
-                      values = c('black',grey(0.4))) +
-  guides(linetype=F) +
+                      values = c('black',grey(0.5))) +
+  guides(colour=F) +
   annotate('text',x=0.5,y=1.1,label='B')
 
+pdf('Figures/Figure 5.pdf',
+    width = 4,
+    height = 8,
+    paper = 'a4')
 multiplot(Fig5A,Fig5B)
+dev.off()
 

### 2016-01-19 Manuscript/SW_early_life_telomeres.Rmd
Post_evolution major changes

@@
                warning=FALSE,
-               message=FALSE,
-               fig.path='../Figures/',
-               fig.width= 4,
-               fig.height= 8,
-               fig.show='asis',
-               dev=c('png','pdf'))
+               message=FALSE)
 
@@
 
-Lewis Spurgin: lewisspurgin@gmail.com; Kat Bebbington: K.Bebbington@uea.ac.uk Eleanor Fairfield: E.Fairfield@uea.ac.uk; Jan Komdeur: j.komdeur@rug.nl; Terry Burke: t.a.burke@sheffield.ac.uk; Hannah Dugdale: h.l.dugdale@rug.nl; David Richardson: david.richardson@uea.ac.uk
+**Correspondence:** Lewis Spurgin: lewisspurgin@gmail.com; David Richardson: david.richardson@uea.ac.uk
 
@@
 Importantly, early-life telomere dynamics have been associated with both short-term and late-life survival [@Heidinger2012; @Boonekamp2014], and with other parameters such as cognition [@Nettle2015a].
-However, to our knowledge no studies have simultaneously analysed how telomere dynamics, early-life conditions and late-life survival are all related in a natural setting.
-Moreover, how early-life telomere dynamics vary over spatial and temporal scales is not known.
+However, few studies have simultaneously analysed how telomere dynamics, early-life conditions and late-life survival are all related in a natural setting.
+Moreover, how early-life telomere dynamics vary over spatial and temporal scales is poorly understood.
 
-The longitudinal study (since 1986) of the Seychelles warbler (*Acrocephalus sechellensis*) population on Cousin Island provides an excellent model system for studying senescence in the wild [reviewed in @Hammers2015].
+The longitudinal study (since 1986) of the Seychelles warbler (*Acrocephalus sechellensis*) population on Cousin Island provides an excellent system for studying senescence in the wild [reviewed in @Hammers2015].
 Due to the isolated nature of the study population and intensive field monitoring, we have unusually comprehensive survival data and tissue samples spanning many years (see Methods, below).
@@
 
-In this study we test the prediction that telomeres link early-life environmental variation to late-life fitness.
+In this study we examine how telomeres link early-life environmental variation to late-life fitness in the Seychelles warbler.
 Because telomere dynamics are expected to reflect individual-level variation in the costs of early-life experiences, they may allow a more sensitive analysis of the effects of early-life environmental variation than would be possible with a direct comparison of how survival is affected by the early-life environment.
-With this in mind, we predict that: i) individuals raised in higher quality conditions will have longer telomeres and reduced telomere shortening, and ii) longer telomeres and lower rates of telomere shortening in early life are associated with greater survival.
+With this in mind, we first conduct an exploratory analysis of how the environmental and social factors experienced in early life affect telomere dynamics.
+We then test the hypothesis that longer telomeres and lower rates of telomere shortening in early life are associated with greater survival.
 
@@
 #####Early-life telomere length and age
-Mean ($\pm$ standard error) telomere length in first-year birds was `r round2(mean(FlSAall$TLKB))` $\pm$ `r round2(se(FlSAall$TLKB))` kb.
-This was lower than the mean for one-year-old birds based on cross-sectional data (`r round2(mean(subset(dd,Age==1)$TLKB))` $\pm$ `r round2(se(subset(dd,Age==1)$TLKB))` kb).
-However, when we considered age classes within the first year of life separately, we found substantial age-related variation in telomere length (F = `r round2(anova(lm(TLKB~Fledged,data = juv))[1,4])`; P = `r round3(anova(lm(TLKB~Fledged,data = juv))[1,5])`).
-Nestlings had longer telomeres than any other age group, but there was an apparent increase in average telomere length from the fledgling to adult stages (Fig.
-1A).
-However, a longitudinal analysis of telomere loss showed that nestlings and fledglings both lost telomeric DNA early in life, and that there was no difference in the rate of telomere loss between the two age classes (F = `r round2(anova(lm(TROC~Fledged,data = Loss))[1,4])`; P = `r round2(anova(lm(TROC~Fledged,data = Loss))[1,5])`; Fig.
-1B), suggesting that the cross-sectional results represent selective disappearance.
+Mean ($\pm$ standard error) telomere length in first-year birds was longer than in adults of any age (Fig.
+1A) We also found substantial age-related variation in telomere length within the first year of life, with longer telomeres in nestlings compared to fledglings (ANOVA; F = `r round2(anova(lm(TLKB~Fledged,data = juv))[1,4])`; P = `r round3(anova(lm(TLKB~Fledged,data = juv))[1,5])`; Fig.
+1B).
 
 #####Among-cohort variation in early-life telomere dynamics
-Early-life telomere length varied significantly among breeding seasons (one-way ANOVA, F = `r round2(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,4])`; P = `r round3(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,5])`), and this variation in mean telomere length over breeding seasons was positively related to island-wide food availability (linear regression, R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,4])`; P = `r round3(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,5])`; Fig.
+Early-life telomere length varied significantly among breeding seasons (one-way ANOVA, F = `r round2(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,4])`; P = `r round3(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,5])`), with a tendency toward shorter telomeres in more recent years (linear regression, R^2^ = `r round2(summary(lm(TLKBmean~Tarsus,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,4])`; P = `r round3(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,5])`; Fig.
+2).
+Variation in mean telomere length over breeding seasons was positively related to island-wide food availability (linear regression, R^2^ = `r round2(summary(lm(TLKBmean~Tarsus,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,4])`; P = `r round3(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,5])`; Fig.
 2A).
@@
 The top model explaining within-season variation in early-life telomere length contained age class, tarsus length and the number of helpers present in the natal territory, as well as interactions between age class and both tarsus length and the number of helpers (Table S1).
-This model was much better supported than the null model ($\Delta$AICc = `r round2(AICc(lm(cenTL~1,data=juvall))-AICc(lm(cenTL~Fledged*(cenTarsus+cenHelper),data=juvall)))`).
+This model was much better supported than the null model ($\Delta$AICc = `r round2(AICc(lm(cenTL~1,data=juvall))-AICc(lm(cenTL~Fledged*(cenTarsus+cenHelper),data=juvall)))`; R^2^ = `r round2(summary(lm(cenTL~Ageclass*(cenTarsus+cenHelper),data=juvall))$r.squared)`).
 Both tarsus length and the number of helpers had high relative importance in the top model set (Fig.
-3A), and the model containing these variables explained a reasonable amount of variation in telomere length (R^2^ = `r round2(summary(lm(cenTL~Ageclass*(cenTarsus+cenHelper),data=juvall))$r.squared)`).
+3A).
 Telomere length was positively related to the number of helpers in nestlings, but not in fledglings (Fig.
@@
 Due to a limited sample size, we considered all age classes together for this analysis.
-We found that none of these variables were associated with differences in telomere loss (all P > 0.05).
+The number of helpers in the natal territory was negatively associated to the rate of telomere loss ().
+We found no relationship between tarsus length and telomere loss ().
 
 #####Early-life telomere dynamics and survival
-At the population-level, there was no relationship between lifespan and telomere length (R^2^ = `r round3(summary(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))[1,4])`; P = `r round2(anova(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))[1,5])`; Fig.
-4A).
-Testing this relationship separately for each age class revealed that there was no age-specific relationship between lifespan and cohort-level telomere length (both P > 0.2).
-There was a negative relationship between population-level telomere loss and lifespan, although this was marginally non-significant (R^2^ = `r round3(summary(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))[1,4])`; P = `r round2(anova(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))[1,5])`; Fig.
-4B).
 
-For the individual-based survival analysis, a log-normal survival model best fitted the Seychelles warbler data (AIC = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="lnorm")$AIC)`, AIC of next best distribution (Weibull) = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="weibull")$AIC)`).
-A survival model including all first-year birds showed that early-life telomere length had a positive effect on survival, but this relationship was marginally non-significant (estimate = `r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`).
+library(survival)
+m1 <- coxph(with(juv,Surv(Lifespan+1,Died)~LogTL+GroupSize+Fledged+frailty(LayYear)))
+
+A log-normal survival model best fitted the Seychelles warbler data (AIC = `r round2(flexsurvreg(with(juv,Surv(Lifespan+1,Died))~ 1, dist="lnorm")$AIC)`, AIC of next best distribution (Weibull) = `r round2(flexsurvreg(with(juv,Surv(Lifespan+1,Died))~ 1, dist="weibull")$AIC)`).
+A survival model including all first-year birds showed that early-life telomere length had a positive effect on survival, but this relationship was marginally non-significant (estimate = `r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~TLKB+LayYear), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,3], lessthan = F)`).
 Running this analysis between age classes revealed differential survival effects.
-Nestling telomere length had no effect on survival  (estimate = `r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`; Fig.
-5A), but telomere length at the fledgling stage was significantly related to survival (estimate = `r round3(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`).
+Nestling telomere length had no effect on survival  (estimate = `r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,3], lessthan = F)`; Fig.
+5A), but telomere length at the fledgling stage was significantly related to survival (estimate = `r round3(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,3], lessthan = F)`).
 This effect was positive, with increased survival later in life associated with longer telomeres (Fig.
 6B).
-Finally, using the longitudinal data, we found no effect of early-life telomere shortening on subsequent survival, although again the estimate was positive, and confidence intervals only marginally overlapped zero (estimate = `r round2(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,1])`, CI = `r round3(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,3], lessthan = F)`).
+Finally, using the longitudinal data, we found no effect of early-life telomere shortening on subsequent survival (estimate = `r round2(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,1])`, CI = `r round3(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,3], lessthan = F)`).
 

### 2016-01-19 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Post_evolution major changes

@@
 ```{r Figure 1}
-######################
-#1A
-######################
 
-ddFig1A <- ddply(dd,
-                 .(Fledged),
-                 summarize,
-                 TLKBmean = mean(TLKB),
-                 TLKBse = se(TLKB))
+####################
+temp <- subset(juvseason,Season == 'Summer')
+temp <- subset(temp,FieldPeriodID != 27)
 
-Fig1A <- ggplot(ddFig1A,
-       aes(x = Fledged,
-           y = TLKBmean)) +
-  ylab('Telomere length (kb)') + 
-  xlab('') +
+Fig1 <- ggplot(temp,
+       aes(x = CatchYear,
+           y = TLKBmean,
+           col = Season)) +
+  ylab('Log telomere length (kb)') + 
+  xlab('Lay Year') +
   theme_classic() +
@@
         axis.title.x = element_text(size = 16),
-        legend.position = c(0.1,0.9),
-        legend.title = element_blank(),
-        legend.text = element_text(size = 15)) +
+        legend.position = 'none') +
   geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse)) +
-  annotate('text',label = 'A',x=0.6,y=6.2)
+  geom_line()+
+  scale_color_manual(values = c('black',grey(0.5)))
 
-######################
-#1B
-######################
-
-ddFig1B <- ddply(Loss,
-                 .(Fledged),
-                 summarize,
-                 TROCmean = mean(TROC),
-                 TROCse = se(TROC))
-
-Fig1B <- ggplot(ddFig1B,
-       aes(x = Fledged,
-           y = TROCmean)) +
-  ylab('Corrected telomere loss') + 
-  xlab('') +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 14), 
-        axis.text.y = element_text(size = 14), 
-        axis.title.y = element_text(size = 16,vjust=0.8),
-        axis.title.x = element_text(size = 16),
-        legend.position = c(0.1,0.9),
-        legend.title = element_blank(),
-        legend.text = element_text(size = 15)) +
-  geom_pointrange(aes(ymin=TROCmean-TROCse,ymax=TROCmean+TROCse)) +
-  annotate('text',label = 'B',x=0.6,y=0.0018)
+################################
 
 pdf('Figures/Figure 1.pdf',
-    width = 4,
-    height = 8,
+    width = 6,
+    height = 4,
     paper = 'a4')
-multiplot(Fig1A,Fig1B)
+Fig1
 dev.off()
@@
 ####################
-
-Fig2A <- ggplot(juvseason,
-       aes(x = Insect,
-           y = TLKBmean,
+Fig1 <- ggplot(dd,
+       aes(x = Fledged,
+           y = LogTL,
            col = Season)) +
   ylab('Telomere length (kb)') + 
-  xlab('Food availability') +
+  xlab('Lay Year') +
   theme_classic() +
@@
   geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse)) +
-  stat_smooth(method = 'lm',se = T,col = 'black') +
-  scale_color_manual(values = c('black',grey(0.5))) +
-  annotate('text',label = 'A',x=-1.5,y=8)
+  geom_line()+
+  scale_color_manual(values = c('black',grey(0.5)))
 
@@
 
-temp <- loss.season[complete.cases(loss.season),]
-
-####################################
-Fig2B <- ggplot(temp,
-       aes(x = Insect,
-           y = TROCmean,
-           col = Season)) +
-  geom_pointrange(aes(min=TROCmean-TROCse,ymax=TROCmean+TROCse)) +
-  ylab('Corrected telomere loss') + 
-  xlab('Food availability') + 
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 14), 
-        axis.text.y = element_text(size = 14), 
-        axis.title.y = element_text(size = 16,vjust=0.8),
-        axis.title.x = element_text(size = 16),
-        legend.position = 'none') +
-  stat_smooth(method = 'lm',se = T,  col = 'black') +
-  scale_color_manual(values = c('black',grey(0.5))) +
-  annotate('text',label = 'B',x=-1.5,y=0.0075)
-
-pdf('Figures/Figure 2.pdf',
-    width = 4,
-    height = 8,
+pdf('Figures/Figure 1.pdf',
+    width = 6,
+    height = 4,
     paper = 'a4')
-multiplot(Fig2A,Fig2B)
+Fig1
 dev.off()
@@
 
-######################
-#3A
-######################
+#####################
 
 #Run the model
-lm2 <- lm(cenTL ~ Fledged*(cenHelper + cenNonHelper + Sex + cenTarsus  + cenTQ),
-            data = juvall)
+lm2 <- lmer(LogTL ~ Fledged + Tarsus + Helper + GroupSize + Sex +Tarsus + BodyMass  + cenTQ  + (1|LayYear) + (1|PlateID),
+            data = juv,
+            REML = FALSE)
 
@@
 lm3 <- dredge(lm2)
-lmavg <- model.avg(lm3,subset = delta <=6)
+lm3 <- subset(lm3,delta <=6)
+lmavg <- model.avg(lm3,fit=T)
 myimp <- importance(lmavg) #Relative importance of each explanatory variable
@@
 
-Fig3A <- forestplot(table2) +
-  ylim(c(-1,1)) +
-  annotate('text',label = 'A',x=0.6,y=1)
+Fig3 <- forestplot(table2) +
+  annotate('text',label = 'A',x=0.6,y=0.1)
 
-######################
-#3B
-######################
-
-temp <- juv
-temp$HelperF <- ifelse(juv$Helper>0,'Helpers present','Helpers absent')
-
-ddFig3B <- ddply(temp,
-                 .(HelperF,Fledged),
-                 summarize,
-                 TLKBmean = mean(TLKB),
-                 TLKBse = se(TLKB))
-
-Fig3B <- ggplot(ddFig3B,
-       aes(x = HelperF,
-           y = TLKBmean,
-           col = Fledged,
-           lty = Fledged)) +
-  ylab('Telomere length (kb)') + 
-  xlab('') +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 14), 
-        axis.text.y = element_text(size = 14), 
-        axis.title.y = element_text(size = 16,vjust=0.8),
-        axis.title.x = element_text(size = 16),
-        legend.position = 'none',
-        legend.title = element_blank(),
-        legend.text = element_text(size = 15)) +
-  geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse),
-                  position = position_dodge(width = 0.2)) +
-  scale_color_manual(values = c('black',grey(0.5))) +
-  annotate('text',label = 'B',x=0.5,y=8)
-
-######################
-#3C
-######################
-
-Fig3C <- ggplot(juv,
-                aes(x = Tarsus,
-                    y = TLKB,
-                    col = Fledged,
-                    lty = Fledged)) +
-  ylab('Telomere length (kb)') + 
-  xlab('Tarsus length (mm)') +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 14), 
-        axis.text.y = element_text(size = 14), 
-        axis.title.y = element_text(size = 16,vjust=0.8),
-        axis.title.x = element_text(size = 16),
-        legend.position = 'none') +
-  geom_point(size=3) +
-  stat_smooth(method = 'lm',se = T) +
-  scale_color_manual(values = c('black',grey(0.5))) +
-  annotate('text',label = 'C',x=19,y=15)
-
-pdf('Figures/Figure 3.pdf',
+pdf('Figures/Figure 2.pdf',
     width = 8,
@@
     paper = 'a4')
-multiplot(Fig3A,Fig3B,Fig3C,layout=matrix(c(1,1,2,3),2,2,byrow=TRUE))
+Fig3
 dev.off()
@@
 
-```{r Figure 4,fig.height=8,fig.width=4}
+```{r Figure 3,fig.height=8,fig.width=4}
 
-Fig4A <- ggplot(juvseason,
-                aes(x = TLKBmean,
-                    y = Lifespan,
-                    col = Season)) +
-  geom_pointrange(aes(ymin=Lifespan-Lifespanse,ymax=Lifespan+Lifespanse)) +
-  geom_errorbarh(aes(xmin = TLKBmean-TLKBse,xmax = TLKBmean+TLKBse)) +
-  theme_classic() +
-  ylab('Remaining life') + 
-  xlab('Telomere length (kb)') + 
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        legend.position = 'none') +
-  scale_color_manual(values = c('black',grey(0.5))) +
-  stat_smooth(method = 'lm',se = T,col = 'black',fullrange = T) +
-  annotate('text',x=2.5,y=10,label='A')
-
-Fig4B <- ggplot(loss.season,
-                aes(x = TROCmean,
-                    y = Lifespan,
-                    col = Season)) +
-  geom_pointrange(aes(ymin=Lifespan-Lifespanse,ymax=Lifespan+Lifespanse)) +
-  geom_errorbarh(aes(xmin = TROCmean-TROCse,xmax = TROCmean+TROCse)) +
-  theme_classic() +
-  ylab('Remaining life') + 
-  xlab('Corrected telomere loss') + 
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 11), 
-        axis.text.y = element_text(size = 11), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        legend.position = 'none') +
-  scale_color_manual(values = c('black',grey(0.5))) +
-  stat_smooth(method = 'lm',se = T,col = 'black',fullrange = T) +
-  annotate('text',x=-0.004,y=12.5,label='B')
-
-pdf('Figures/Figure 4.pdf',
-    width = 4,
-    height = 8,
-    paper = 'a4')
-multiplot(Fig4A,Fig4B)
-dev.off()
-
-```
-
-```{r Figure 5,fig.height=8,fig.width=4}
-
-s1 <- survfit(Surv(time = Lifespan+1, event = Died)~TLF, data = chicks)
-Fig5A <- ggsurv(s1,plot.cens=F,lty.est=c(1,2)) + 
+s1 <- survfit(Surv(time = Lifespan+1, event = Died)~TLF, data = chickall)
+Fig3A <- ggsurv(s1,plot.cens=F,lty.est=c(1,2)) + 
   theme_classic() +
@@
 s2 <- survfit(Surv(time = Lifespan+1, event = Died)~TLF, data = FlSAall)
-Fig5B <- ggsurv(s2,plot.cens=F,lty.est=c(1,2)) + 
+Fig3B <- ggsurv(s2,plot.cens=F,lty.est=c(1,2)) + 
   theme_classic() +
@@
 
-pdf('Figures/Figure 5.pdf',
+pdf('Figures/Figure 3.pdf',
     width = 4,
@@
     paper = 'a4')
-multiplot(Fig5A,Fig5B)
+multiplot(Fig3A,Fig3B)
 dev.off()

### 2016-01-19 Scripts/SW juvenile telomeres IPC clean.R

@@ -1,13 +1,11 @@
 #################################################################################*
 ## Early life telomere dynamics and late-life fitness in a wild bird population
 ## CLEAN THE DATA
-#################################################################################*
-
-
+#################################################################################
 #Average repeats of blood samples
-av <- ave(dd$TL,c(dd$BloodID,dd$Status))
+av <- ave(dd$TL,c(dd$BloodID,dd$Status,dd$PlateID))
 dd$TL <- av
-dd <- unique(dd)
+dd <- dd[!(duplicated(dd$BloodID)),]
 
 
 # Weird variable names ----------------------------------------------------
@@ -27,7 +25,7 @@ dd$CatchDate <- as.Date(dd$CatchDate,"%d/%m/%Y")
 dd$Season <- ifelse(as.numeric(format(dd$CatchDate,'%m')) %in% c(4:10),
                     'Summer','Winter')
 
-head(dd)
+
 # Tarsus ------------------------------------------------------------------
 
 dd$Tarsus <- NA
@@ -48,7 +46,7 @@ dd$LeftTarsus <- NULL
 # Remove unwanted data/outliers ----------------------------------------------------
 
 
-dd <- droplevels(subset(subset(dd,TL>1000),TL<15000))
+dd <- droplevels(subset(subset(dd,TL>1000),TL<30000))
 dd <- subset(dd,BodyMass>5)
 dd <- subset(dd,Tarsus>17)
 
@@ -56,7 +54,7 @@ dd <- subset(dd,Tarsus>17)
 # Telomere data -----------------------------------------------------------
 
 dd$TLKB <- dd$TL/1000
-dd$LogTL <- log(dd$TL)
+dd$LogTL <- log10(dd$TL)
 
 
 
@@ -132,8 +130,8 @@ for(i in 1:nrow(juv))
   juv$cenTL[i] <- (juv$TLKB[i] - mean(currentdata$TLKB))/sd(currentdata$TLKB)
 }
 
-mymed <- mean(juv$cenTL,na.rm=T)
-juv$TLF <- ifelse(juv$cenTL > mymed,'Long telomeres','Short telomeres')
+mymed <- mean(juv$TLKB,na.rm=T)
+juv$TLF <- ifelse(juv$TLKB > mymed,'Long telomeres','Short telomeres')
 
 
 # Helpers and social group size -------------------------------------------
[…]

### 2016-01-19 Scripts/SW juvenile telomeres IPC load.r

@@ -18,7 +18,7 @@ library(survival)
 library(flexsurv)
 
 # Load data ---------------------------------------------------------------
-dd <- read.csv('Data/SW TL main data for analysis 2.csv') #main dataset
+dd <- read.csv('Data/TL2 output 15 Dec 2015.csv') #main dataset
 terr <- read.csv('Data/SW TL all territories.csv')
 status <- read.csv('Data/SW TL all breedstatus.csv')
 insects <- read.csv('Data/Insects.csv')

### 2016-01-20 Manuscript/SW_early_life_telomeres.Rmd
More post evolution edits

@@
 
-**Telomeres reveal silver spoon effects in a wild population**
+**Strong cohort effects on early-life telomere length in a wild population**
 
-Lewis G. Spurgin^1,2^,  Kat Bebbington^1^, Eleanor A. Fairfield^1^, Jan Komdeur^3^, Terry Burke^4^, Hannah, L. Dugdale^3,4^, and David S. Richardson^1,5,^.
+Lewis G. Spurgin^1,2^,  Kat Bebbington^1^, Eleanor A. Fairfield^1^, Martijn Hammers^2^, Jan Komdeur^3^, Terry Burke^4^, Hannah, L. Dugdale^3,4^, and David S. Richardson^1,5,^.
 
@@
 ###Abstract
-Understanding the links between early-life experiences and later-life survival ("silver spoon effects") is fundamental to understanding life-history evolution.
+Understanding the short and long term costs of individual early-life experiences is fundamental to understanding life-history evolution.
 Telomeres, the protective caps at the ends of chromosomes, shorten in response to oxidative stress, and telomere shortening is correlated with reduced survival.
-Thus, telomere dynamics may help us quantify individual variation in early-life costs, and thus enhance our understanding of how and why silver spoon effects occur.
+Thus, telomere dynamics may help us quantify individual variation in early-life costs, and enhance our understanding of how poor conditions in early life are related to later-life survival.
 We tested how telomere dynamics are related to spatiotemporal variation in early-life conditions and later-life survival in the Seychelles warbler (*Acrocephalus sechellensis*), across multiple cohorts spanning 14 years.
-We show that telomere lengths and rates of shortening vary across cohorts, and that individuals born in breeding seasons with higher food availability had longer telomeres and reduced telomere shortening.
-Within cohorts, we found that telomere length in nestlings was negatively related to tarsus length (which  represents nestling growth), and positively related to the number of helpers present in the natal territory.
-Finally, we found that increased survival later in life was associated with longer telomeres and reduced telomere shortening rates in early life.
-Our results show that telomeres can both measure variation in early-life costs and predict late-life survival in wild populations.
+we found that, in accordance with other studies, telomere length and loss are greatest in early life.
+However, we also show that juvenile telomere lengths varies markedly among cohorts.
+We found no evidence that early-life social environment (number of helpers, group size) or ecological conditions (territory quality) were related to telomere length, although we found tentative evidence that telomere length varied among summer and winter breeding season.
+Finally, we found that increased survival later in life was associated with longer telomeres and reduced telomere shortening rates in early life, but these effects could .
+Our results highlight the inmportance of cohort effects in studies of telomere length, and suggest that
 
@@
 
-Telomere lengths were measured in `r nrow(FlSA)+nrow(chicks)+nrow(Loss)` samples.
-Of these, `r nrow(FlSA)+nrow(chicks)` were taken cross-sectionally from birds caught within their first year of life, between `r min(as.numeric(paste(juv$LayYear)))` and `r max(as.numeric(paste(juv$LayYear)))`; `r nrow(chicks)` were nestlings less than one month old (chicks), while the remaining `r nrow(FlSA)` were fledglings and subadults aged between two and twelve months (hereafter referred to as fledglings).
+Telomere lengths were measured in `r nrow(juv)+nrow(Loss)` samples, .
+Of these, `r nrow(juv)` were taken cross-sectionally from birds caught within their first year of life, between `r min(as.numeric(paste(juv$LayYear)))` and `r max(as.numeric(paste(juv$LayYear)))`.
 For a subset of first-year birds (n = `r nrow(Loss)` individuals) we had longitudinal data, with an additional sample taken within two years of the original catch.
-We measured absolute telomere quantity per diploid genome for all samples using a quantitative PCR (qPCR) assay  [full details in @Barrett2012].
+
+We measured absolute telomere quantity per diploid genome for all samples using a quantitative PCR (qPCR) assay [@Barrett2012].
+Post-qPCR processing of samples was carried out as in Barrett *et al.* [-@Barrett2012], with the following amendments.
+First, averaging of technical repeats was carried out using custom made R scripts (available as supplementary material).
+Second, a change in batch of SYBR green forced us to raise the annealing temperature of the telomere reaction from 58^oC to 61^oC.
+This resulted in consistently higher, but repeatable, CQ values for the telomere reaction, so we implemented a simple correction to ensure that telomere lengths were repeatable across annealing temperatures (supplementary material).
+Finally, we excluded telomere lengths greater than 30kb, which maximised repeatability of our own data [see also @Bauch2012].
+Inter-plate repeatability of final telomere lengths was assessed using the R package rptR [@Schielzeth2011].
+
+For the subset of samples with longitudinal data we calculated a rate of telomere loss between the first-year and adult samples by subtracting adult telomere length from early-life telomere length and dividing this difference by the length of time (in days) between sampling events.
+To account for regression to the mean effects, we applied a correction based on correlations among samples within individuals following Verhulst *et al.* [-@Verhulst2013].
 
@@
 We performed all analyses using R version 3.0.1 [@RDevelopmentCoreTeam2011].
-Our sampling regime covers `r max(juv$LayYear)-min(juv$LayYear)` years, and temporal variation in rainfall and food availability on Cousin Island over this period is expected to result in variation  in a range of social and environmental variables [@Brouwer2012].
-This makes it difficult to disentangle how spatial and temporal processes differentially affect early-life telomere dynamics.
-To overcome this, we analysed how variation in early-life telomere length was related to variation in early-life conditions both among and within breeding seasons.
-
-First, to analyse whether temporal variation in food availability drives population-level variation in telomere dynamics, we calculated mean telomere length among all birds born in a given (major or minor) breeding season (i.e.
-a cohort), excluding cohorts where the sample size $\leq$ 5 birds (leaving a sample size of n = `r nrow(juvseason)` cohorts).
-We tested for a difference in telomere length among cohorts using a one-way ANOVA, and then used linear regression to test whether variation in mean telomere length among cohorts is explained by temporal variation in insect abundance.
-
-We then conducted individual-level analyses to test how spatial variation in early-life environmental and social conditions influenced telomere length within cohorts, using general linear models.
-For these analyses, both response and explanatory variables were standardised by mean-centring within cohorts, thus controlling for any between-cohort temporal variation in telomere length and the social and ecological environment.
-As explanatory variables we included age class (nestling or fledgling), tarsus length, sex, territory quality and the number of helping and non-helping subordinate birds present in the natal territory.
-We also included interactions between age class and all the other variables.
-We used model averaging to assess how these variables shape telomere length and dynamics in early life.
-We created a full model containing all of the above terms, and a top model set was then defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011].
+We used  general linear mixed models along with model averaging to explore how spatial variation in early-life environmental and social conditions influenced telomere length within cohorts.
+As explanatory variables we included age class (nestling or fledgling), tarsus length, sex, territory quality, season (summer or winter) and the number of helping and non-helping subordinate birds present in the natal territory.
+As random effects we included birth year and qPCR plate ID.
+We first created a full model containing all of the above terms, and a top model set was then defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011].
 We report model-averaged coefficients, confidence intervals and 'relative importance', which reflects the relative weights of each predictor variable across the top model set.
+For individuals with longitudinal data (*n* = `r nrow(Loss)`) we repeated the above analyses of telomere dynamics, replacing telomere length with telomere loss as the response variable.
+In the analysis of telomere loss we also included initial telomere length as an explanatory variable, as longer telomeres have been shown to decrease in length more rapidly, even after correcting for regression to the mean [@Verhulst2013].
 Model averaging was carried out using the MuMIn package (version 1.10.5) in R [@Barton2012].
 
-For individuals with longitudinal data (*n* = `r nrow(Loss)`) we calculated a rate of telomere loss between the first-year sample and earliest adult sample by subtracting the adult telomere length from the early-life telomere length and dividing this difference by the length of time (in days) between sampling events.
-To account for regression to the mean effects, we applied a correction based on correlations among samples within individuals following Verhulst *et al.* [-@Verhulst2013].
-Using this dataset we repeated the above analyses of among and within cohort telomere dynamics, replacing telomere length with telomere loss as the response variable.
-
-Finally, we analysed the effects of telomere dynamics on survival, again by looking at variation both within and among cohorts.
-First, we tested the prediction that cohort-level telomere dynamics are associated with survival by comparing the mean lifespan of all individuals within each cohort with mean telomere length and rate of loss in that cohort, using linear regression.
-We then used parametric survival analysis to test whether survival was related to individual-level telomere length and rate of loss.
-For the latter analysis, again we mean-centred variables within cohorts to eliminate between-season temporal variation in telomeres and survival.
-The survival time distribution was chosen by comparing the fit of the observed survival data against a range of distributions (exponential, Weibull, log-normal, gamma, Gompertz and generalized F) by visually inspecting plots and comparing AIC values.
-As explanatory variables of survival, we included early-life telomere length/loss (we ran a separate analysis for each, using the cross-sectional and longitudinal data, respectively), plus the factors identified as being related to telomere length in the previous analysis.
-With this approach we aimed to assess the relationship between early-life telomere length and survival, while controlling for potentially confounding variables that might be correlated with telomere dynamics.
+We used Cox regression to test whether survival was related to individual-level telomere length and rate of loss.
+We ran these models with and without lay year as a frailty term to test whether any effects on telomere length on survival could be explained by cohort effects.
 
@@
 #####Early-life telomere length and age
+Inter-plate repeatability of telomere length, based on `r length(table(dd0[duplicated(dd0$BloodID),]$BirdID))` samples measured at least twice, was 0.78 (CI = 0.73-0.82).
 Mean ($\pm$ standard error) telomere length in first-year birds was longer than in adults of any age (Fig.
-1A) We also found substantial age-related variation in telomere length within the first year of life, with longer telomeres in nestlings compared to fledglings (ANOVA; F = `r round2(anova(lm(TLKB~Fledged,data = juv))[1,4])`; P = `r round3(anova(lm(TLKB~Fledged,data = juv))[1,5])`; Fig.
+1A).
+We also found substantial age-related variation in telomere length within the first year of life, with longer telomeres in nestlings compared to fledglings (ANOVA; F = `r round2(anova(lm(TLKB~Fledged,data = juv))[1,4])`; P = `r round3(anova(lm(TLKB~Fledged,data = juv))[1,5])`; Fig.
 1B).
 
-#####Among-cohort variation in early-life telomere dynamics
-Early-life telomere length varied significantly among breeding seasons (one-way ANOVA, F = `r round2(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,4])`; P = `r round3(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,5])`), with a tendency toward shorter telomeres in more recent years (linear regression, R^2^ = `r round2(summary(lm(TLKBmean~Tarsus,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,4])`; P = `r round3(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,5])`; Fig.
+#####Cohort-level variation in early-life telomere dynamics
+Early-life telomere length varied significantly among breeding seasons (one-way ANOVA, F = `r round2(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,4])`; P = `r round3(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,5])`), with a tendency toward shorter telomeres in more recent years (linear regression of mean telomere length and lay year, R^2^ = `r round2(summary(lm(TLKBmean~LayYear,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(TLKBmean~LayYear,data = juvseason,na.action=na.exclude))[1,4])`; P = `r round3(anova(lm(TLKBmean~LayYear,data = juvseason,na.action=na.exclude))[1,5])`; Fig.
 2).
-Variation in mean telomere length over breeding seasons was positively related to island-wide food availability (linear regression, R^2^ = `r round2(summary(lm(TLKBmean~Tarsus,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,4])`; P = `r round3(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,5])`; Fig.
-2A).
-Running the regression of cohort-level telomere length and food availability separately for the two age classes revealed a positive, significant relationship in fledglings (R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=flseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data=flseason,na.action=na.exclude))[1,4])`; P = `r round3(anova(lm(Insect~TLKBmean,data=flseason,na.action=na.exclude))[1,5])`), but no relationship in nestlings (R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=chickseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data=chickseason,na.action=na.exclude))[1,4])`; P = `r round3(anova(lm(Insect~TLKBmean,data=chickseason,na.action=na.exclude))[1,5])`).
+Variation in mean telomere length over breeding seasons was not related yearly variation in island-wide food availability (linear regression, R^2^ = `r round2(summary(lm(TLKBmean~Insect,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(TLKBmean~Insect,data = juvseason,na.action=na.exclude))[1,4])`; P = `r round2(anova(lm(TLKBmean~Insect,data = juvseason,na.action=na.exclude))[1,5])`).
 
-The rate of telomere shortening in early life (taken from the smaller subset of birds for which we had longitudinal samples) also varied significantly among breeding seasons (F = `r round2(anova(lm(TROC~factor(LayYear),data = Loss))[1,4])`; P = `r round2(anova(lm(TROC~factor(LayYear),data = Loss))[1,5])`), and rates of telomere shortening could be explained by temporal variation in food availability (R^2^ = `r round2(summary(lm(Insect~TROCmean,data=loss.season,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TROCmean,data = loss.season,na.action=na.exclude))[1,4])`; P = `r round2(anova(lm(Insect~TROCmean,data = loss.season,na.action=na.exclude))[1,5])`).
-This pattern corresponded with the cross-sectional data: individuals born in seasons with low food availability lost telomeres, on average, at a faster rate than those born in years with high food availability (Fig.
-2B).
-Unfortunately we did not have sufficient sample sizes within cohorts to run this analysis separately between age classes.
+#####Individual-level variation in early-life telomere dynamics
+The top model explaining within-season variation in early-life telomere length contained age class, tarsus length and season (summer vs winter) (Table S1).
+This model was much better supported than the null model ($\Delta$AICc = `r round2(AICc(lmer(LogTL~1+(1|FieldPeriodID)+(1|PlateID),data=juv,REML = FALSE))-AICc(lmer(LogTL~Fledged+Tarsus+Season+(1|FieldPeriodID)+(1|PlateID),data=juv,REML = FALSE)))`).
+All three variables importance in the top model set, but only ageclass had confidence intervals that did not overlap zero (Fig.
+3A).
+Telomere length was slightly longer in juveniles born in winter compared to summmer seasons, and this effect appeared most pronounced in nestlings (although sample size of nestlings born in winter seasons is very small) (Fig.
+4A).
+Tarsus length was negatively related to telomere length, although much of this effect could be explained by ageclass, as fledglings have both longer tarsi and shorter telomeres compared to nestlings  (Fig.
+4B).
 
-#####Within-cohort variation in early-life telomere dynamics
-The top model explaining within-season variation in early-life telomere length contained age class, tarsus length and the number of helpers present in the natal territory, as well as interactions between age class and both tarsus length and the number of helpers (Table S1).
-This model was much better supported than the null model ($\Delta$AICc = `r round2(AICc(lm(cenTL~1,data=juvall))-AICc(lm(cenTL~Fledged*(cenTarsus+cenHelper),data=juvall)))`; R^2^ = `r round2(summary(lm(cenTL~Ageclass*(cenTarsus+cenHelper),data=juvall))$r.squared)`).
-Both tarsus length and the number of helpers had high relative importance in the top model set (Fig.
-3A).
-Telomere length was positively related to the number of helpers in nestlings, but not in fledglings (Fig.
+The longitudinal data showed the same patterns as the cross-sectional data.
+The top model explaining telomere loss again contained age class, tarsus length and season, although in this instance the top model was only a marginally better fit than the null model ($\Delta$AICc = `r round3(AICc(lmer(TROC~(1|LayYear),data=Loss,REML = FALSE))-AICc(lmer(TROC~Fledged+Tarsus+Season+(1|LayYear),data=Loss, REML = FALSE)),lessthan=F)`).
+Fledglings lost telomeres at a lower rate compared to nestlings, while birds with longer tarsi and birds born in winter seasons had higher rates of telomere shortening compared to those born in the summer (Fig.
 3B).
-Tarsus length was negatively related to telomere length in nestlings, but not in fledglings (Fig.
-3C).
-
-For the subset of individuals for which we had longitudinal data, we tested whether the factors previously identified as being related to early-life telomere length (i.e.
-tarsus length and number of helpers) were also associated with differences in rates of early-life telomere loss.
-Due to a limited sample size, we considered all age classes together for this analysis.
-The number of helpers in the natal territory was negatively associated to the rate of telomere loss ().
-We found no relationship between tarsus length and telomere loss ().
+However, it is important to note that sample size for this longitudinal analysis was small and all confidence intervals overlapped zero (Fig.
+3B).
 
@@
 
-library(survival)
-m1 <- coxph(with(juv,Surv(Lifespan+1,Died)~LogTL+GroupSize+Fledged+frailty(LayYear)))
-
-A log-normal survival model best fitted the Seychelles warbler data (AIC = `r round2(flexsurvreg(with(juv,Surv(Lifespan+1,Died))~ 1, dist="lnorm")$AIC)`, AIC of next best distribution (Weibull) = `r round2(flexsurvreg(with(juv,Surv(Lifespan+1,Died))~ 1, dist="weibull")$AIC)`).
-A survival model including all first-year birds showed that early-life telomere length had a positive effect on survival, but this relationship was marginally non-significant (estimate = `r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~TLKB+LayYear), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,3], lessthan = F)`).
-Running this analysis between age classes revealed differential survival effects.
-Nestling telomere length had no effect on survival  (estimate = `r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,3], lessthan = F)`; Fig.
-5A), but telomere length at the fledgling stage was significantly related to survival (estimate = `r round3(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,3], lessthan = F)`).
-This effect was positive, with increased survival later in life associated with longer telomeres (Fig.
-6B).
-Finally, using the longitudinal data, we found no effect of early-life telomere shortening on subsequent survival (estimate = `r round2(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,1])`, CI = `r round3(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,3], lessthan = F)`).
+A survival model including only early-life telomere length revealed a significant and positive effect on survival (estimate = `r round3(summary(coxph(with(juv,Surv(Lifespan+1,Died)~LogTL)))$coefficients[1])`, P = `r round3(summary(coxph(with(juv,Surv(Lifespan+1,Died)~LogTL)))$coefficients[5])`; Fig.
+2).
+However, when we controlled for cohort effects by including lay year as a frailty term in the model, telomere length was no longer a significant predictor of survival (estimate = `r round3(summary(coxph(with(juv,Surv(Lifespan+1,Died)~LogTL+frailty(LayYear))))$coefficients[1,1])`, P = `r round3(summary(coxph(with(juv,Surv(Lifespan+1,Died)~LogTL+frailty(LayYear))))$coefficients[1,6])`).
+The longitudinal data showed that the amount of telomere shortening experienced in early life had no effect on survival, regardless of whether we included lay year (estimate `r round3(summary(coxph(with(Loss,Surv(Lifespan+1,Died)~TROC+frailty(LayYear))))$coefficients[1,1])`, P = `r round3(summary(coxph(with(Loss,Surv(Lifespan+1,Died)~TROC+frailty(LayYear))))$coefficients[1,6])`), or excluded it (estimate `r round3(summary(coxph(with(Loss,Surv(Lifespan+1,Died)~TROC)))$coefficients[1])`, P = `r round3(summary(coxph(with(Loss,Surv(Lifespan+1,Died)~TROC)))$coefficients[5])`).
 
@@
 
-Our finding that early-life telomere length and rate of shortening are related to food availability adds to a growing body of evidence from humans and wild animals showing that the natal environment can have pronounced impacts on early-life telomere dynamics [reviewed in @Price2013; @Monaghan2014].
-However, very few studies have shown that temporal variation in telomere dynamics occurs in natural populations, and the studies that have done so were limited to just two seasons [@Mizutani2013; @Watson2015], making it impossible to statistically evaluate whether this variation is driven by temporal variation in the environment.
-The long-term Seychelles warbler dataset has allowed us, using many more cohorts than has been used in other studies, to show that temporal variation in environmental conditions does affect telomere dynamics, and that this effect can be seen at the population level.
-Interestingly, in the Seychelles warbler survival is not directly linked to annual variation in food availability [@Brouwer2006; @Hammers2013].
-It has been suggested that telomeres act as biomarkers of cost in natural populations, as they are able to capture individual and cohort-level variability in the costs experienced as a result of environmental pressures [@Asghar2015].
-Our results show that telomeres can detect hidden costs in natural populations that are not detectable using life-history and environmental data alone.
+The clearest result from our study is that telomere length varies markedly among cohorts.
+Very few studies have shown that temporal variation in telomere dynamics occurs in natural populations, and the studies that have done so were limited to just two seasons [@Mizutani2013; @Watson2015].
+The long-term Seychelles warbler dataset has allowed us, using many more cohorts than has been used in other studies, to show that temporal variation in telomere dynamics does occur at the population level.
 
@@
 
-We found that, within cohorts, tarsus length and the number of helpers were the most important predictors of telomere length in Seychelles warblers, and that these effects occurred specifically at the nestling stage.
-The tarsus effect probably reflects the fact that in passerine birds tarsus length is correlated with age during the nestling stage [@Ricklefs1976].
+Despite a large sample size and reasonable repeatability, we found little evidence that any of the social or environmental vairables we measured affect juvenile telomere length in the Seychelles warbler.
+This is surprising a poor social and ecological environment is known to be detrimental to juvenile Seychelles warblers, both in terms of oxidative stress and later life survival.
+Future research should therefore examine, in a quantitative genetic framework, how genetic and environmental components, and their interactions, affect telomere dynamics and senescence in natural populations [@Becker2015; @Asghar2014].
+
+We found tentative evidence that, in addition to age class, tarsus length and season had an effect on juvenile telomere length (Figs 3, 4).
+The tarsus effect most likely reflects the fact that in passerine birds tarsus length is correlated with age during the nestling stage [@Ricklefs1976].
 Telomere loss is most rapid early in life due to ongoing cell replication [@Frenck1998], and a negative correlation between telomere length and body size in early life is therefore expected.
+Indeed, that age class affects both tarsus length and telomere lengths can be clearly seen in our data (Fig.
+4B).
+The difference in telomere length between birds born in summer and winter seasons is more surprising, especially given the directionality of the effect.
+We find that birds born in winter seasons having longer telomeres (Fig.
+4A), while the opp.
+One possible explanation for this is that parents in good condition are more likely to breed
 
-Ours is the first study to show that the presence of helpers has a beneficial effect on offspring in terms of their telomeres.
-In the Seychelles warbler the survival benefits to a nestling of having helpers in the natal territory is well established [@Komdeur1994;@Brouwer2012], and that the beneficial effect of helpers is detectable in terms of differential telomere length, is therefore encouraging.
-However, we found no evidence that telomere loss was related to the number of helpers - this is perhaps surprising as telomere loss is expected to be a better indicator of stress than telomere length *per se*, and effects of telomere length may occur as a result of correlated to telomere loss [@Boonekamp2014].
-However, our longitudinal dataset was limited, both in terms of sample size and resolution (i.e.
+While the relationship between mortality and telomeres in adults has been established for some time [@Cawthon2003], only recently has the link between later-life survival and early-life telomere dynamics been studied.
+In captive zebra finches, juvenile telomere length predicts late-life survival [@Heidinger2012], and in wild bird populations survival to the nestling phase [@Watson2015], and survival to adulthood [@Boonekamp2014], have been linked with early-life telomere dynamics.
+
+Finally, our study highlights some of the diffculties asosciated with studying telomere dynamics in natural populations.
+
+Our longitudinal dataset was limited, both in terms of sample size and resolution (i.e.
 time between sampling events).
@@
 
-While the relationship between mortality and telomeres in adults has been established for some time [@Cawthon2003], only recently has the link between later-life survival and early-life telomere dynamics been studied.
-In captive zebra finches, juvenile telomere length predicts late-life survival [@Heidinger2012], and in wild bird populations survival to the nestling phase [@Watson2015], and survival to adulthood [@Boonekamp2014], have been linked with early-life telomere dynamics.
-Our results are, to our knowledge, the first to demonstrate that later-life survival is related to early-life telomere length in a wild population.
-This effect can be seen at the individual level, and (to a lesser extent) at the cohort level, although the nature of the relationship varies according to the age class in which telomeres are measured.
-At the individual level, we found that telomere length was related to survival in fledglings, but not nestlings.
-This is not surprising, as only by the fledgling stage is telomere length likely to reflect the accumulation of early-life stress (see also previous paragraph).
-The relationships between lifespan and population-level telomere length were in the expected direction (Fig.
-4), but not significant, which is perhaps not surprising given that even though we have a long-term dataset, we are still restricted in terms of sample size when conducting cohort-level analyses.
-Nonetheless, our data support the general conclusion that early-life telomere dynamics are important indicators of early-life stress.
-
-There are numerous promising avenues for future research into telomere evolutionary ecology in this system and others.
+There are numerous avenues for future research into telomere evolutionary ecology in this system and others.
 Here we have considered survival, but telomere length and shortening in early life may also be linked to other components of fitness [reviewed in @Monaghan2014].
 Reproductive senescence occurs in the Seychelles warbler [reviewed in @Hammers2015], making this system well suited to examining how telomere length predicts lifetime reproductive success.
-A further avenue for future research will be examining, in a quantitative genetic framework, how genetic and environmental components, and their interactions, affect telomere dynamics and senescence in natural populations [@Becker2015; @Asghar2014].
 We expect that by gaining a fuller understanding of telomere dynamics in natural populations, the fields of life-history evolution and evolutionary ecology will be greatly enhanced.

### 2016-01-20 Manuscript/SW_early_life_telomeres_SOM.Rmd
More post evolution edits

@@
 #Run the model
-lm1 <- lm(cenTL ~ Fledged*(cenHelper + cenNonHelper + Sex + cenTarsus  + cenTQ),
-            data = juvall)
+lm1 <- lmer(LogTL ~ Fledged + Tarsus + Helper + NonHelper + Sex + Tarsus + BodyMass  + cenTQ  + Season + (1|LayYear),
+            data = juv,
+            REML = FALSE)
 
@@
 nams <- colnames(t1)
-nams[nams == 'cenTarsus'] <- 'Tarsus'
-nams[nams == 'cenNonHelper'] <- 'Non helpers'
+#Sort out row/col names
+nams[nams == 'NonHelper'] <- 'Non helpers'
 nams[nams == 'SexMales'] <- 'Sex'
 nams[nams == 'cenTQ'] <- 'TQ'
-nams[nams == 'cenHelper'] <- 'Helpers'
-nams[nams == 'cenHelper:Fledged'] <- 'Ageclass x Helpers'
-nams[nams == 'cenNonHelper:Fledged'] <- 'Ageclass x Non helpers'
-nams[nams == 'cenTarsus:Fledged'] <- 'Ageclass x Tarsus'
-nams[nams == 'cenTQ:Fledged'] <- 'Ageclass x TQ'
-nams[nams == 'Fledged:SexMales'] <- 'Ageclass x Sex'
-nams[nams == 'Fledged'] <- 'Ageclass'
+nams[nams == 'Helper'] <- 'Helpers'
+nams[nams == 'BodyMass'] <- 'Body mass'
+nams[nams == 'FledgedFledglings'] <- 'Age class'
+nams[nams == 'SeasonWinter'] <- 'Season'
 
@@
 ```
+
+```{r Table S2}
+
+#Run the model
+lm2 <- lmer(TROC ~ Fledged + Tarsus + Helper + NonHelper + Sex + Tarsus + BodyMass  + cenTQ  + Season + LogTL + (1|LayYear),
+            data = Loss,
+            REML = FALSE)
+
+#Model averaging
+
+lm3 <- dredge(lm2)
+lm3 <- subset(lm3,delta <=6)
+
+#Model averaging
+t1 <- as.matrix(subset(lm3, delta<=6))
+
+#Sort out row/col names
+nams <- colnames(t1)
+
+nams[nams == 'NonHelper'] <- 'Non helpers'
+nams[nams == 'SexMales'] <- 'Sex'
+nams[nams == 'cenTQ'] <- 'TQ'
+nams[nams == 'Helper'] <- 'Helpers'
+nams[nams == 'BodyMass'] <- 'Body mass'
+nams[nams == 'FledgedFledglings'] <- 'Age class'
+nams[nams == 'SeasonWinter'] <- 'Season'
+nams[nams == 'LogTL'] <- 'Initial TL'
+
+colnames(t1) <- nams
+
+tableS1 <- data.frame(Model = apply(t1,1,getterms,c(2:12)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
+colnames(tableS1) <- c('Model','df','AICc', 'Delta AICc','Weight')
+rownames(tableS1) <- NULL
+tableS1$Model[tableS1$Model == ''] <- 'Null model'
+tableS1[,3:5] <- lapply(tableS1[,3:5],as.numeric)
+kable(tableS1,digits=2)
+
+```

### 2016-01-20 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
More post evolution edits

@@
 
+temp <- dd
+temp <- subset(temp,Age<14)
+
+ddFig1A <- ddply(temp,
+                 .(Age),
+                 summarize,
+                 TL = mean(LogTL),
+                 TLse = se(LogTL),
+                 n = length(LogTL))
+
+Fig1A <- ggplot(ddFig1A,
+       aes(x = Age,
+           y = TL)) +
+  ylab('Log telomere length (kb)') + 
+  xlab('Age (years)') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 14), 
+        axis.text.y = element_text(size = 14), 
+        axis.title.y = element_text(size = 16,vjust=0.8),
+        axis.title.x = element_text(size = 16),
+        legend.position = 'none') +
+  geom_pointrange(aes(ymax=TL+TLse,ymin=TL-TLse))+
+  geom_line()+
+  annotate('text',label = paste0('(',ddFig1A$n,')'),x = ddFig1A$Age,y = 3.3,size=3)+
+  annotate('text',label='A',x = 0,y = 3.8)
+
+####################
+ss <- table(juv$Fledged)
+
+Fig1B <- ggplot(juv,
+       aes(x = Fledged,
+           y = LogTL)) +
+  ylab('Telomere length (kb)') + 
+  xlab('') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 14), 
+        axis.text.y = element_text(size = 14), 
+        axis.title.y = element_text(size = 16,vjust=0.8),
+        axis.title.x = element_text(size = 16),
+        legend.position = 'none') +
+  geom_boxplot(notch=T)+
+  scale_color_manual(values = c('black',grey(0.5))) +
+  annotate('text',label = paste0('(',ss,')'),x = c(1:2),y = 2.85)+
+  annotate('text',label='B',x = 0.5,y = 4.5)
+
+################################
+
+pdf('Figures/Figure 1.pdf',
+    width = 9,
+    height = 3,
+    paper = 'a4r')
+multiplot(Fig1A,Fig1B,layout = matrix(c(1,1,1,1,2,2,2),nrow=1))
+dev.off()
+
+```
+
+```{r Figure 2}
+
 ####################
@@
 
-Fig1 <- ggplot(temp,
-       aes(x = CatchYear,
-           y = TLKBmean,
-           col = Season)) +
+ss <- temp$n
+
+Fig2 <- ggplot(temp,
+       aes(x = LayYear,
+           y = TLKBmean) +
   ylab('Log telomere length (kb)') + 
@@
   geom_line()+
-  scale_color_manual(values = c('black',grey(0.5)))
+  annotate('text',label=paste0('(',ss,')'),x = temp$LayYear,y = 3.4,size = 4)
 
@@
 
-pdf('Figures/Figure 1.pdf',
-    width = 6,
+pdf('Figures/Figure 2.pdf',
+    width = 8,
     height = 4,
-    paper = 'a4')
-Fig1
+    paper = 'a4r')
+Fig2
 dev.off()
 
-```
+adults2 <- subset(adults,Age>0)
 
-```{r Figure 2}
+mymean <- tapply(adults2$LogTL,adults2$CatchYear,mean)
+myse <- tapply(adults2$LogTL,adults2$CatchYear,se)
+myn <- tapply(adults2$LogTL,adults2$CatchYear,length)
 
-####################
-Fig1 <- ggplot(dd,
-       aes(x = Fledged,
-           y = LogTL,
-           col = Season)) +
-  ylab('Telomere length (kb)') + 
+ddFig2B <- data.frame(TL = mymean, TLse = myse, n = myn,LayYear = as.numeric(names(mymean)))
+
+ddFig2B <- subset(ddFig2B,n>5)
+
+ss <- ddFig2B$n
+
+Fig2 <- ggplot(ddFig2B,
+       aes(x = LayYear,
+           y = TL)) +
+  ylab('Log telomere length (kb)') + 
   xlab('Lay Year') +
@@
         legend.position = 'none') +
-  geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse)) +
+  geom_pointrange(aes(ymin=TL-TLse,ymax=TL+TLse)) +
   geom_line()+
-  scale_color_manual(values = c('black',grey(0.5)))
-
-################################
-
-pdf('Figures/Figure 1.pdf',
-    width = 6,
-    height = 4,
-    paper = 'a4')
-Fig1
-dev.off()
+  annotate('text',label=paste0('(',ss,')'),x = ddFig2B$LayYear,y = 3.3,size = 4)
 
@@
 #####################
+#3A
+#####################
 
 #Run the model
-lm2 <- lmer(LogTL ~ Fledged + Tarsus + Helper + GroupSize + Sex +Tarsus + BodyMass  + cenTQ  + (1|LayYear) + (1|PlateID),
+lm2 <- lmer(LogTL ~ Fledged + Tarsus + Helper + NonHelper + Sex + Tarsus + BodyMass  + cenTQ  + Season + (1|LayYear) + (1|PlateID) + (1|BirdID),
             data = juv,
@@
 nams <- rownames(table2)
-nams[nams == 'cenTarsus'] <- 'Tarsus'
-nams[nams == 'cenNonHelper'] <- 'Non helpers'
+nams[nams == 'NonHelper'] <- 'Non helpers'
 nams[nams == 'SexMales'] <- 'Sex'
 nams[nams == 'cenTQ'] <- 'TQ'
-nams[nams == 'cenHelper'] <- 'Helpers'
-nams[nams == 'cenHelper:FledgedFledglings'] <- 'Ageclass x Helpers'
-nams[nams == 'cenNonHelper:FledgedFledglings'] <- 'Ageclass x Non helpers'
-nams[nams == 'cenTarsus:FledgedFledglings'] <- 'Ageclass x Tarsus'
-nams[nams == 'cenTQ:FledgedFledglings'] <- 'Ageclass x TQ'
-nams[nams == 'FledgedFledglings:SexMales'] <- 'Ageclass x Sex'
-nams[nams == 'FledgedFledglings'] <- 'Ageclass'
+nams[nams == 'Helper'] <- 'Helpers'
+nams[nams == 'BodyMass'] <- 'Body mass'
+nams[nams == 'FledgedFledglings'] <- 'Age class'
+nams[nams == 'SeasonWinter'] <- 'Season'
 
@@
 
-Fig3 <- forestplot(table2) +
-  annotate('text',label = 'A',x=0.6,y=0.1)
+Fig3A <- forestplot(table2)
 
-pdf('Figures/Figure 2.pdf',
-    width = 8,
+#####################
+#3B
+#####################
+
+#Run the model
+lm2 <- lmer(TROC ~ Fledged + Tarsus + Helper + NonHelper + Sex + Tarsus + BodyMass  + cenTQ  + Season + LogTL + (1|LayYear),
+            data = Loss,
+            REML = FALSE)
+
+#Model averaging
+
+lm3 <- dredge(lm2)
+lm3 <- subset(lm3,delta <=6)
+lmavg <- model.avg(lm3,fit=T)
+myimp <- importance(lmavg) #Relative importance of each explanatory variable
+
+#Put it all into a nice table
+table2 <- imptable(lmavg,myimp)
+table2 <- signif(table2,2)
+
+#Sort out row/col names
+nams <- rownames(table2)
+nams[nams == 'NonHelper'] <- 'Non helpers'
+nams[nams == 'SexMales'] <- 'Sex'
+nams[nams == 'cenTQ'] <- 'TQ'
+nams[nams == 'Helper'] <- 'Helpers'
+nams[nams == 'BodyMass'] <- 'Body mass'
+nams[nams == 'FledgedFledglings'] <- 'Age class'
+nams[nams == 'SeasonWinter'] <- 'Season'
+nams[nams == 'LogTL'] <- 'Initial TL'
+
+rownames(table2) <- nams
+colnames(table2) <- c('Estimate','SE','2.5%','97.5%','RI')
+
+table2 <- data.frame(table2,myx=row.names(table2))
+
+Fig3B <- forestplot(table2)
+
+pdf('Figures/Figure 3.pdf',
+    width = 7,
     height = 8,
     paper = 'a4')
-Fig3
+multiplot(Fig3A,Fig3B)
 dev.off()
@@
 
-```{r Figure 3,fig.height=8,fig.width=4}
+```{r Figure 4}
 
-s1 <- survfit(Surv(time = Lifespan+1, event = Died)~TLF, data = chickall)
-Fig3A <- ggsurv(s1,plot.cens=F,lty.est=c(1,2)) + 
+ss <- c(tapply(juv$LogTL,list(juv$Season,juv$Fledged),length))
+
+Fig4A <- ggplot(juv,
+       aes(x = Fledged,
+           y = LogTL,
+           fill=Season)) +
+  ylab('Telomere length (kb)') + 
+  xlab('') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 14), 
+        axis.text.y = element_text(size = 14), 
+        axis.title.y = element_text(size = 16,vjust=0.8),
+        axis.title.x = element_text(size = 16),
+        legend.position = 'none') +
+  geom_boxplot(notch=F)+
+  scale_fill_manual(values = c('white',grey(0.5))) +
+  annotate('text',label='A',x = 0.5,y = 4.5) +
+  annotate('text',label = paste0('(',ss,')'),x=c(0.8,1.2,1.8,2.2),y = 2.8)
+
+Fig4B <- ggplot(juv,
+       aes(x = Tarsus,
+           y = LogTL,
+           col = Fledged)) +
+  ylab('Telomere length (kb)') + 
+  xlab('') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 14), 
+        axis.text.y = element_text(size = 14), 
+        axis.title.y = element_text(size = 16,vjust=0.8),
+        axis.title.x = element_text(size = 16),
+        legend.position = 'none') +
+  geom_point()+
+  scale_colour_manual(values = c('black',grey(0.5))) +
+  annotate('text',label='B',x = 17.5,y = 4.5) +
+  stat_smooth(method = 'lm')
+
+pdf('Figures/Figure 4.pdf',
+    width = 4,
+    height = 8,
+    paper = 'a4')
+multiplot(Fig4A,Fig4B)
+dev.off()
+
+```
+
+```{r Figure 5}
+s1 <- survfit(Surv(time = Lifespan+1, event = Died)~TLF, data = juv)
+Fig5 <- ggsurv(s1,plot.cens=F,lty.est=c(1,2)) + 
   theme_classic() +
@@
                       values = c('black',grey(0.5))) +
-  guides(colour=F) +
-  annotate('text',x=0.5,y=1.1,label='A')
+  guides(colour=F)
 
-###########################################
-s2 <- survfit(Surv(time = Lifespan+1, event = Died)~TLF, data = FlSAall)
-Fig3B <- ggsurv(s2,plot.cens=F,lty.est=c(1,2)) + 
-  theme_classic() +
-  ylab('Proportion survived') + 
-  xlab('Age (years)') + 
-  xlim(c(0,15)) +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 14), 
-        axis.text.y = element_text(size = 14), 
-        axis.title.y = element_text(size = 16,vjust=0.8),
-        axis.title.x = element_text(size = 16),
-        legend.position = 'none') +
-  scale_colour_manual(name = 'TL',
-                      values = c('black',grey(0.5))) +
-  guides(colour=F) +
-  annotate('text',x=0.5,y=1.1,label='B')
-
-pdf('Figures/Figure 3.pdf',
+pdf('Figures/Figure 5.pdf',
     width = 4,
-    height = 8,
+    height = 4,
     paper = 'a4')
-multiplot(Fig3A,Fig3B)
+Fig5
 dev.off()

### 2016-01-20 Scripts/SW juvenile telomeres IPC clean.R

@@ -2,10 +2,14 @@
 ## Early life telomere dynamics and late-life fitness in a wild bird population
 ## CLEAN THE DATA
 #################################################################################
+
+#Only use Ellie's samples
+dd0 <- subset(dd0,Whodunnit == 'EAF')
+
 #Average repeats of blood samples
-av <- ave(dd$TL,c(dd$BloodID,dd$Status,dd$PlateID))
-dd$TL <- av
-dd <- dd[!(duplicated(dd$BloodID)),]
+av <- ave(dd0$TL,c(dd0$BloodID,dd0$Status,dd0$PlateID))
+dd0$TL <- av
+dd <- dd0[!(duplicated(dd0$BloodID)),]
 
 
 # Weird variable names ----------------------------------------------------
@@ -84,16 +88,6 @@ dd$AgemonthF <- factor(dd$AgemonthF,levels = c('<1','1-9','9-12','>12'))
 
 dd <- subset(dd,Agemonths>0)
 
-#Delta age
-for(i in 1:nrow(dd))
-{
-  currentdata = subset(dd,BirdID == dd$BirdID[i])
-  dd$MeanAge[i] <- mean(currentdata$Age)
-  dd$DeltaAge[i] <- dd$Age[i]-dd$MeanAge[i]
-}
-
-
-
 # Survival and lifespan ---------------------------------------------------
 
 dd$RemainingLife <- dd$DeathYear-dd$CatchYear
@@ -121,9 +115,10 @@ dd$Sex <- ifelse(dd$SexEstimate == 1,'Males','Females')
 juv <- droplevels(subset(dd,Ageclass %in% c('CH','FL','OFL','SA')))
 adults <- droplevels(subset(dd,Ageclass == 'A'))
 
-juv$cenTL <- NA
+
 
 #Centre telomere length by birth year
+juv$cenTL <- NA
 for(i in 1:nrow(juv))
 {
   currentdata <- subset(juv,FieldPeriodID == FieldPeriodID[i])
@@ -191,13 +186,14 @@ for(i in 1:nrow(juv))
 
 # Subset Fledglings and subadults -----------------------------------------------------------
 
-xf <- subset(juv,Status == 'XF')
 
+
+#Get rid of NAs and cross-fostered birds
 juv <- subset(juv,!is.na(Tarsus))
 juv <- subset(juv,!is.na(BodyMass))
 juv <- subset(juv,Status!='XF')
 
-
[…]

### 2016-01-20 Scripts/SW juvenile telomeres IPC load.r

@@ -15,10 +15,9 @@ library(plyr)
 library(ggplot2)
 library(gvlma)
 library(survival)
-library(flexsurv)
 
 # Load data ---------------------------------------------------------------
-dd <- read.csv('Data/TL2 output 15 Dec 2015.csv') #main dataset
+dd0 <- read.csv('Data/TL2 output 15 Dec 2015.csv') #main dataset
 terr <- read.csv('Data/SW TL all territories.csv')
 status <- read.csv('Data/SW TL all breedstatus.csv')
 insects <- read.csv('Data/Insects.csv')

### 2016-02-03 Manuscript/SW_early_life_telomeres.Rmd
Redraft edits

@@
 
-Lewis G. Spurgin^1,2^,  Kat Bebbington^1^, Eleanor A. Fairfield^1^, Martijn Hammers^2^, Jan Komdeur^3^, Terry Burke^4^, Hannah, L. Dugdale^3,4^, and David S. Richardson^1,5,^.
+Lewis G. Spurgin^1,2^,  Kat Bebbington^1^, Eleanor A. Fairfield^1^, Martijn Hammers^3^, Jan Komdeur^3^, Terry Burke^4^, Hannah, L. Dugdale^3,4^, and David S. Richardson^1,5,^.
 
@@
 
-**Running head:** Telomeres and silver spoons
+**Running head:** Early-life telomeres and survival
 
@@
 we found that, in accordance with other studies, telomere length and loss are greatest in early life.
-However, we also show that juvenile telomere lengths varies markedly among cohorts.
-We found no evidence that early-life social environment (number of helpers, group size) or ecological conditions (territory quality) were related to telomere length, although we found tentative evidence that telomere length varied among summer and winter breeding season.
-Finally, we found that increased survival later in life was associated with longer telomeres and reduced telomere shortening rates in early life, but these effects could .
-Our results highlight the inmportance of cohort effects in studies of telomere length, and suggest that
+We then show that juvenile telomere lengths varies markedly among cohorts, with average telomere length varying by .
+We found no evidence that early-life social environment (number of helpers, group size) or ecological conditions (territory quality) were related to telomere length, although we found tentative evidence that telomere length varied among summer and winter breeding seasons.
+Finally, we found that increased survival later in life was associated with longer telomeres in early life, but this effect was at the cohort, rather than individual, level.
+Our results highlight the inmportance of cohort effects in studies of telomere length.
 
@@
 As an estimate of seasonal variation in food availability, we calculated an index of the number of insects across the entire island during each breeding season.
-This  index represents the average number of insects found per unit leaf area over all surveys carried out on the island in a breeding season.
+This index is calculated as the mean number of insects found per unit leaf area over all surveys carried out on the island in a breeding season.
 
-Each time a bird is caught on Cousin body mass and tarsus length are measured (to the nearest 0.1g and 0.1mm, respectively), and age is confirmed on the basis of eye colour [@Komdeur1991a] and previous captures [@Richardson2003b].
+Each time a bird is caught on Cousin body mass and tarsus length are measured (to the nearest 0.1g and 0.1mm, respectively).
+Using information on eye colour [@Komdeur1991a] and previous captures [@Richardson2003b], we grouped birds into three age categories: one month old (birds still in the nest), 6 months old (fledglings with light grey eyes) and 10 months old (subadults with light brown eyes).
 A blood sample (*ca* 25 $\mu$l) is taken from each bird captured via brachial venipuncture, and stored at room temperature in 1 ml of absolute ethanol in a 1.5 ml screw-cap microfuge tube.
@@
 For each sample, genomic DNA was extracted from a ~2 mm^2^ flake of preserved blood using the DNeasy Blood and Tissue Kit (Qiagen), following the manufacturer's protocol, with the modification of overnight lysis at 37^o^C and a final DNA elution volume of 80 $\mu$l.
-DNA concentration and purity were quantified using a NanoDrop 8000 Spectrophotometer (ThermoScientific), and DNA integrity was validated by visualization with ethidium bromide after electrophoresis on a 1.2% agarose gel.
-Sex was determined using the molecular method outlined by Griffiths *et al.* [-@Griffiths1998].
+Sex was determined using the PCR-based method outlined by Griffiths *et al.* [-@Griffiths1998].
+Prior to telomere analysis, DNA concentration and purity were quantified using a NanoDrop 8000 Spectrophotometer (ThermoScientific), and the following thesholds were applied before samples were included for further analysis: i) DNA concentration must be at least 15 ng $\mu$l^-1^ (based on a mean of three measurements), ii) the 260/280 ratio has to be between 1.8 and 2 and, iii) the 260/230 ratio should be higher than 1.8.
+DNA integrity was further validated by visualization with ethidium bromide after electrophoresis on a 1.2% agarose gel, and all samples with evidence of DNA degradation were re-extracted or excluded.
 
-Telomere lengths were measured in `r nrow(juv)+nrow(Loss)` samples, .
+All DNA extractions that passed the above criteria were diluted to 3.3 ng $\mu$l^-1^ before telomere measurement.
+We measured absolute telomere quantity per diploid genome for all samples using a quantitative PCR (qPCR) assay, outlined in detail by Barrett *et al.* [-@Barrett2012], with one amendment.
+A change in batch of SYBR green forced us to raise the annealing temperature of the telomere reaction from 58^o^C to 61^o^C for the majority of the samples (n = `r nrow(subset(dd,Objective %in% c(7:9)))`; see also below).
+Post-qPCR processing of samples was carried out as in Barrett *et al.* [-@Barrett2012], with the following amendments.
+First, averaging of technical repeats was carried out using custom-made R scripts (available as supplementary material) instead of GenEx.
+Second, the change in annealing temperature resulted in consistently higher, but repeatable, CQ values for the telomere reaction, so we implemented a simple correction to ensure that telomere lengths were repeatable across annealing temperatures (supplementary material).
+Finally, we excluded telomere lengths greater than 30kb, which maximised repeatability of our own data [see also @Bauch2012].
+Inter-plate repeatability of final telomere lengths was assessed using the R package rptR [@Schielzeth2011], and for all subsequent analyses we used mean telomere length per sample where we had repeats.
+
+Telomere lengths were measured using a total of `r nrow(juv)+nrow(Loss)` samples, .
 Of these, `r nrow(juv)` were taken cross-sectionally from birds caught within their first year of life, between `r min(as.numeric(paste(juv$LayYear)))` and `r max(as.numeric(paste(juv$LayYear)))`.
-For a subset of first-year birds (n = `r nrow(Loss)` individuals) we had longitudinal data, with an additional sample taken within two years of the original catch.
-
-We measured absolute telomere quantity per diploid genome for all samples using a quantitative PCR (qPCR) assay [@Barrett2012].
-Post-qPCR processing of samples was carried out as in Barrett *et al.* [-@Barrett2012], with the following amendments.
-First, averaging of technical repeats was carried out using custom made R scripts (available as supplementary material).
-Second, a change in batch of SYBR green forced us to raise the annealing temperature of the telomere reaction from 58^oC to 61^oC.
-This resulted in consistently higher, but repeatable, CQ values for the telomere reaction, so we implemented a simple correction to ensure that telomere lengths were repeatable across annealing temperatures (supplementary material).
-Finally, we excluded telomere lengths greater than 30kb, which maximised repeatability of our own data [see also @Bauch2012].
-Inter-plate repeatability of final telomere lengths was assessed using the R package rptR [@Schielzeth2011].
-
-For the subset of samples with longitudinal data we calculated a rate of telomere loss between the first-year and adult samples by subtracting adult telomere length from early-life telomere length and dividing this difference by the length of time (in days) between sampling events.
-To account for regression to the mean effects, we applied a correction based on correlations among samples within individuals following Verhulst *et al.* [-@Verhulst2013].
+For a subset of first-year birds (n = `r nrow(Loss)` individuals) we had longitudinal data, with an additional sample taken as an adult.
+For these individuals we calculated the absolute amount of telomere loss between the first-year and adult samples by subtracting adult telomere length from early-life telomere length, as well as a rate of telomere loss by dividing this difference by the length of time (in days) between sampling events.
 
@@
 We performed all analyses using R version 3.0.1 [@RDevelopmentCoreTeam2011].
-We used  general linear mixed models along with model averaging to explore how spatial variation in early-life environmental and social conditions influenced telomere length within cohorts.
-As explanatory variables we included age class (nestling or fledgling), tarsus length, sex, territory quality, season (summer or winter) and the number of helping and non-helping subordinate birds present in the natal territory.
-As random effects we included birth year and qPCR plate ID.
-We first created a full model containing all of the above terms, and a top model set was then defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011].
+Telomere length was log transfomred to fit assumptions of normality.
+We first explored how telomere length varied within the first year of life and over time using linear regression (telomere length vs age in months), and one-way ANOVA (telomere length vs cohort).
+With the longitudinal data, we tested how telomeres shorten with age by testing how telomere loss and rate of loss were related to the time interval between sampling events, using linear models.
+If telomeres are lost at a constant rate from early age, we expect the time between sampling events to be positively and linearly relaited to telomere loss no, and unrelated to rate of loss.
+If, however, telomeres are lost at a greater rate early in life we expect a non-linear or negative relationship between telomere loss and time between sampling events, and a significant (linear or non-linear) decrease in telomere rate of loss with time interval.
+
+We then used a general linear mixed model approach to explore how spatial variation in early-life environmental and social conditions influenced telomere length within cohorts.
+Model averaging was carried out using the MuMIn package (version 1.10.5) in R [@Barton2012].
+We first created a full model containing the following explanatory variables: age (in months), tarsus length, sex, territory quality, season (summer or winter) and the number of helping and non-helping subordinate birds present in the natal territory.
+We also included interaction terms between age and all the other variables as telomere dynamics are epxected to vary within the first year of life [@Heidinger2012].
+As random effects we included cohort, territory ID and qPCR plate ID.
+Model selection was then performed and a top model set defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011].
 We report model-averaged coefficients, confidence intervals and 'relative importance', which reflects the relative weights of each predictor variable across the top model set.
-For individuals with longitudinal data (*n* = `r nrow(Loss)`) we repeated the above analyses of telomere dynamics, replacing telomere length with telomere loss as the response variable.
-In the analysis of telomere loss we also included initial telomere length as an explanatory variable, as longer telomeres have been shown to decrease in length more rapidly, even after correcting for regression to the mean [@Verhulst2013].
-Model averaging was carried out using the MuMIn package (version 1.10.5) in R [@Barton2012].
+For individuals with longitudinal data (*n* = `r nrow(Loss)`) we repeated the above analyses of telomere dynamics, replacing telomere length with telomere loss as the response variable, and excluding the plate ID random effect (as each measurement of telomere loss was based on two or more measurements, and so run on multiple plates).
 
-We used Cox regression to test whether survival was related to individual-level telomere length and rate of loss.
-We ran these models with and without lay year as a frailty term to test whether any effects on telomere length on survival could be explained by cohort effects.
+We used Cox regression, implemented in the survival package in R [@Therneau2011], to test whether survival was related to individual-level telomere length and rate of loss.
+For the cross-sectional data we used lifespan (in years) as survival time, while for the longitudinal data we used remaining lifespan from the second sampling event.
+We ran these models with cohort as a frailty term to exclude cohort effects.
 
@@
 #####Early-life telomere length and age
-Inter-plate repeatability of telomere length, based on `r length(table(dd0[duplicated(dd0$BloodID),]$BirdID))` samples measured at least twice, was 0.78 (CI = 0.73-0.82).
-Mean ($\pm$ standard error) telomere length in first-year birds was longer than in adults of any age (Fig.
+Inter-plate repeatability of telomere length, based on XX samples measured at least twice, was 0.78 (CI = 0.73-0.82).
+Telomere length dereased with age within the first year of life (R^2^ = `r round2(summary(lm(LogTL~Agemonths,data=juv))$r.squared)`; F = `r round2(anova(lm(LogTL~Agemonths,data=juv))[1,4])`; P = `r round3(anova(lm(LogTL~Agemonths,data=juv))[1,5])`; Fig.
 1A).
-We also found substantial age-related variation in telomere length within the first year of life, with longer telomeres in nestlings compared to fledglings (ANOVA; F = `r round2(anova(lm(TLKB~Fledged,data = juv))[1,4])`; P = `r round3(anova(lm(TLKB~Fledged,data = juv))[1,5])`; Fig.
+Similarly, telomere shortening based on longitudinal data was significantly greater than zero in nestlings (one sample t-test; P < 0.001), but not in any other age group (all P > 0.5; Fig.
 1B).
@@
 #####Cohort-level variation in early-life telomere dynamics
-Early-life telomere length varied significantly among breeding seasons (one-way ANOVA, F = `r round2(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,4])`; P = `r round3(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,5])`), with a tendency toward shorter telomeres in more recent years (linear regression of mean telomere length and lay year, R^2^ = `r round2(summary(lm(TLKBmean~LayYear,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(TLKBmean~LayYear,data = juvseason,na.action=na.exclude))[1,4])`; P = `r round3(anova(lm(TLKBmean~LayYear,data = juvseason,na.action=na.exclude))[1,5])`; Fig.
-2).
-Variation in mean telomere length over breeding seasons was not related yearly variation in island-wide food availability (linear regression, R^2^ = `r round2(summary(lm(TLKBmean~Insect,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(TLKBmean~Insect,data = juvseason,na.action=na.exclude))[1,4])`; P = `r round2(anova(lm(TLKBmean~Insect,data = juvseason,na.action=na.exclude))[1,5])`).
+Early-life telomere length varied significantly among breeding seasons (one-way ANOVA, F = `r round2(anova(lm(LogTL~factor(FieldPeriodID),data = juv))[1,4])`; P = `r round3(anova(lm(LogTL~factor(FieldPeriodID),data = juv))[1,5])`).
+Separating temporal variaiton in telomere length among age classes revealed no significant cohort effects in chicks (F = `r round2(anova(lm(LogTL~factor(FieldPeriodID),data = subset(juv,Agemonths == 1)))[1,4])`; P = `r round2(anova(lm(LogTL~factor(FieldPeriodID),data = subset(juv,Agemonths == 1)))[1,5])`; Fig.
+2A), but significant effects in fledglings (F = `r round2(anova(lm(LogTL~factor(FieldPeriodID),data = subset(juv,Agemonths == 6)))[1,4])`; P = `r round3(anova(lm(LogTL~factor(FieldPeriodID),data = subset(juv,Agemonths == 6)))[1,5])`; Fig.
+2B) and subadults (F = `r round2(anova(lm(LogTL~factor(FieldPeriodID),data = subset(juv,Agemonths == 9)))[1,4])`; P = `r round3(anova(lm(LogTL~factor(FieldPeriodID),data = subset(juv,Agemonths == 9)))[1,5])`; Fig.
+2C).
+Variation in median telomere length over breeding seasons in fledglings and subadults was not related temporal variation in  territory quality (linear regression, R^2^ = `r round2(summary(lm(TL~TQ,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(TL~TQ,data = juvseason,na.action=na.exclude))[1,4])`; P = `r round2(anova(lm(TL~TQ,data = juvseason,na.action=na.exclude))[1,5])`), island-wide food availability (R^2^ = `r round2(summary(lm(TL~Insect,data=subset(juvseason,Insect<9),na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(TL~Insect,data = subset(juvseason,Insect<9),na.action=na.exclude))[1,4])`; P = `r round2(anova(lm(TL~Insect,data = subset(juvseason,Insect<9),na.action=na.exclude))[1,5])`) or population density (R^2^ = `r round2(summary(lm(TL~Density,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(TL~Density,data = juvseason,na.action=na.exclude))[1,4])`; P = `r round2(anova(lm(TL~Density,data = juvseason,na.action=na.exclude))[1,5])`).
 
 #####Individual-level variation in early-life telomere dynamics
-The top model explaining within-season variation in early-life telomere length contained age class, tarsus length and season (summer vs winter) (Table S1).
-This model was much better supported than the null model ($\Delta$AICc = `r round2(AICc(lmer(LogTL~1+(1|FieldPeriodID)+(1|PlateID),data=juv,REML = FALSE))-AICc(lmer(LogTL~Fledged+Tarsus+Season+(1|FieldPeriodID)+(1|PlateID),data=juv,REML = FALSE)))`).
-All three variables importance in the top model set, but only ageclass had confidence intervals that did not overlap zero (Fig.
+The top model explaining within-season variation in early-life telomere length contained age class and the number of non-helping subordinates in the natal territory (Table S1).
+This model was much better supported than the null model ($\Delta$AICc = `r round2(AICc(lmer(LogTL~1+(1|FieldPeriodID)+(1|PlateID)+(1|BirdID),data=juv,REML = FALSE))-AICc(lmer(LogTL~Fledged+NonHelper+(1|FieldPeriodID)+(1|PlateID)+(1|BirdID),data=juv,REML = FALSE)))`).
+The age class effect was very strong and significant, and was present in every model in the top model set (Fig.
 3A).
-Telomere length was slightly longer in juveniles born in winter compared to summmer seasons, and this effect appeared most pronounced in nestlings (although sample size of nestlings born in winter seasons is very small) (Fig.
-4A).
-Tarsus length was negatively related to telomere length, although much of this effect could be explained by ageclass, as fledglings have both longer tarsi and shorter telomeres compared to nestlings  (Fig.
-4B).
+The effect of non-helpers, on the other had did not have higher importance across the top model set than the other variables, and confidence intervals overlapped zero (Fig.
+3A).
+Furthermore the model including non-helpers and age class model was not a significantly better fit than a model including age class alone ($\Delta$AICc = 0.06; Table S1).
 
-The longitudinal data showed the same patterns as the cross-sectional data.
-The top model explaining telomere loss again contained age class, tarsus length and season, although in this instance the top model was only a marginally better fit than the null model ($\Delta$AICc = `r round3(AICc(lmer(TROC~(1|LayYear),data=Loss,REML = FALSE))-AICc(lmer(TROC~Fledged+Tarsus+Season+(1|LayYear),data=Loss, REML = FALSE)),lessthan=F)`).
-Fledglings lost telomeres at a lower rate compared to nestlings, while birds with longer tarsi and birds born in winter seasons had higher rates of telomere shortening compared to those born in the summer (Fig.
-3B).
-However, it is important to note that sample size for this longitudinal analysis was small and all confidence intervals overlapped zero (Fig.
+The top model explaining telomere loss contained age class and tarsus length, and agian this was a significantly better fit than the null model ($\Delta$AICc = `r round2(AICc(lmer(TROC~1+(1|FieldPeriodID),data=Loss,REML = FALSE))-AICc(lmer(TROC~Fledged+Tarsus+(1|FieldPeriodID),data=Loss,REML = FALSE)))`).
+Fledglings lost telomeres at a lower rate compared to nestlings, while birds with longer tarsi higher rates of telomere shortening (Fig.
 3B).
@@
 
-A survival model including only early-life telomere length revealed a significant and positive effect on survival (estimate = `r round3(summary(coxph(with(juv,Surv(Lifespan+1,Died)~LogTL)))$coefficients[1])`, P = `r round3(summary(coxph(with(juv,Surv(Lifespan+1,Died)~LogTL)))$coefficients[5])`; Fig.
-2).
-However, when we controlled for cohort effects by including lay year as a frailty term in the model, telomere length was no longer a significant predictor of survival (estimate = `r round3(summary(coxph(with(juv,Surv(Lifespan+1,Died)~LogTL+frailty(LayYear))))$coefficients[1,1])`, P = `r round3(summary(coxph(with(juv,Surv(Lifespan+1,Died)~LogTL+frailty(LayYear))))$coefficients[1,6])`).
-The longitudinal data showed that the amount of telomere shortening experienced in early life had no effect on survival, regardless of whether we included lay year (estimate `r round3(summary(coxph(with(Loss,Surv(Lifespan+1,Died)~TROC+frailty(LayYear))))$coefficients[1,1])`, P = `r round3(summary(coxph(with(Loss,Surv(Lifespan+1,Died)~TROC+frailty(LayYear))))$coefficients[1,6])`), or excluded it (estimate `r round3(summary(coxph(with(Loss,Surv(Lifespan+1,Died)~TROC)))$coefficients[1])`, P = `r round3(summary(coxph(with(Loss,Surv(Lifespan+1,Died)~TROC)))$coefficients[5])`).
+Telomere length in early life did not affect survival to adulthood (estimate = `r round2(summary(glm(SurvivedNext~LogTL+Agemonths,data=juv,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glm(SurvivedNext~LogTL+Agemonths,data=juv,family = 'binomial'))$coefficients[2,4])`), and there was no interaction effect between telomere length and juvenile age on survival to adulthood (estimate = `r round2(summary(glm(SurvivedNext~LogTL*Agemonths,data=juv,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glm(SurvivedNext~LogTL*Agemonths,data=juv,family = 'binomial'))$coefficients[4,4])`).
+To separate out cohort-level telomere length affected survival we calculated cohort-level telomere length (i.e.
+the mean for each cohort) and added this term to a logistic regression.
+However, we found no effect of telomere length on survival to adulthood (estimate = `r round2(summary(glm(SurvivedNext~cohortTL+Agemonths,data=juv,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glm(SurvivedNext~cohortTL+Agemonths,data=juv,family = 'binomial'))$coefficients[2,4])`), and no interaction with juvenile age (estimate = `r round2(summary(glm(SurvivedNext~cohortTL*Agemonths,data=juv,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glm(SurvivedNext~cohortTL*Agemonths,data=juv,family = 'binomial'))$coefficients[4,4])`)
+
+Cox regression also showed no effect of individual-level or cohort-level telomere length on survival later in life (individual-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~LogTL+Fledged)))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~LogTL)))$coefficients[1,5])`;  cohort-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL+Fledged+frailty(LayYear))))$coefficients[2,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL+Fledged+frailty(LayYear))))$coefficients[2,6])`).
+
+The longitudinal data showed that the amount of telomere shortening experienced in early life had no effect on survival to adulthood (estimate = `r round2(summary(glm(SurvivedNext~TROC+Agemonths,data=subset(Loss,LayYear<20013),family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glm(SurvivedNext~TROC+Agemonths,data=subset(Loss,LayYear<2013),family = 'binomial'))$coefficients[2,4])`), or in later life (estimate `r round3(summary(coxph(with(subset(Loss,LayYear<2009),Surv(RemainingLife2+1,Died)~TROC+Agemonths)))$coefficients[1,1])`, P = `r round3(summary(coxph(with(subset(Loss,LayYear<2009),Surv(RemainingLife2+1,Died)~TROC+Agemonths)))$coefficients[1,5])`).
 
 ###Discussion
-Here we show that telomeres are related to both early-life conditions and later-life survival in the Seychelles warbler.
-We first show that telomere length and rates of shortening in early life are subject to strong cohort effects, and that cohort-level telomere dynamics can be explained by island-wide temporal variation in food availability.
-We then show that within cohorts, the social environment (i.e.
-number of helpers) affects telomere length specifically at the nestling stage.
-Finally, we show that telomere length in early life predicts later-life survival.
-These findings have important implications for research on life history evolution in wild populations.
+Here we use the long-term study of a closed population of Seychelles warblers to assess the relationships between early-life conditions, telomere length and survival.
+We find that while telomere length varies
 
-The clearest result from our study is that telomere length varies markedly among cohorts.
-Very few studies have shown that temporal variation in telomere dynamics occurs in natural populations, and the studies that have done so were limited to just two seasons [@Mizutani2013; @Watson2015].
-The long-term Seychelles warbler dataset has allowed us, using many more cohorts than has been used in other studies, to show that temporal variation in telomere dynamics does occur at the population level.
+The clearest result from our study is that telomere length varies among cohorts.
+Very few studies have shown that temporal variation in telomere dynamics occurs in natural populations, and to our knowledge the studies that have done so were limited to just two seasons [@Mizutani2013; @Watson2015].
+The long-term Seychelles warbler dataset has allowed us to show that temporal variation in telomere dynamics does occur at the population level over longer time periods.
 
@@
 @Coulson2001].
-If our findings are replicated in other systems and population-level variation in early-life telomere dynamics is common in nature, this has a number of ramifications for our understanding of senescence in natural populations.
-First, our findings suggest that the telomere dynamics of a population at a given point in time represent a snapshot of a temporally varying process.
-More research is now needed within and across multiple cohorts and populations to better understand the complexity of telomere evolutionary ecology.
-Second, our findings raise the interesting prospect of using telomeres as indicators of population health - an approach that would be useful to a broad range of conservation and animal health practitioners.
-However, further research across populations with different environments and genetic histories is required in order to provide further insight into the usefulness of telomeres as biomarkers of population health.
+If our findings are replicated in other systems and population-level variation in early-life telomere dynamics is common in nature, this has potential conseaquences for our understanding of telomere dynamics in natural populations.
+In particular, our findings suggest that the telomere dynamics of a population at a given point in time represent a snapshot of a temporally varying process.
+More research is now needed within and across multiple cohorts and populations to better understand how how and why population-level telomere dynamics vary over space and time.
 
-Despite a large sample size and reasonable repeatability, we found little evidence that any of the social or environmental vairables we measured affect juvenile telomere length in the Seychelles warbler.
+In the Seychelles warbler we found only weak evidence that any of the social or environmental vairables we measured affect juvenile telomere length.
 This is surprising a poor social and ecological environment is known to be detrimental to juvenile Seychelles warblers, both in terms of oxidative stress and later life survival.
+Possible explanations for this finding include i) temporal variation in telomere length in our data is so strong that we are unable to detect spatial trends, ii) within-cohort variation in telomere length is explained by a variable that we have not included in our analyses, or iii) that effects are generally weak and levels of noise in our telomere measurments and/or ecological data preclude detection of significant effects.
 Future research should therefore examine, in a quantitative genetic framework, how genetic and environmental components, and their interactions, affect telomere dynamics and senescence in natural populations [@Becker2015; @Asghar2014].
 
-We found tentative evidence that, in addition to age class, tarsus length and season had an effect on juvenile telomere length (Figs 3, 4).
+Our data suggest that, in addition to age class, tarsus length and season had weak effects on juvenile telomere length (Figs 3, 4).
 The tarsus effect most likely reflects the fact that in passerine birds tarsus length is correlated with age during the nestling stage [@Ricklefs1976].

### 2016-02-03 Manuscript/SW_early_life_telomeres_SOM.Rmd
Redraft edits

--- 2015-10-05 Manuscript/SW_early_life_telomeres_SOM.Rmd
@@
 
-**Table S1** Model selection table for analysis of factors affecting early life telomere length in Seychelles warblers.
+**Table S1** Model selection table for analysis of factors affecting early life telomere length in juvenile Seychelles warblers.
 Only models that remained in the top model set for model averaging (AICc $\leq$ 6 compared to the top model) are included here.
@@
 #Run the model
-lm1 <- lm(cenTL ~ Fledged*(cenHelper + cenNonHelper + Sex + cenTarsus  + cenTQ),
-            data = juvall)
+lm1 <- lmer(LogTL ~ Agemonths * (Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ) + (1|LayYear)+(1|PlateID)+(1|BirdID),
+            data = juv,
+            REML = FALSE)
+lm1 <- standardize(lm1)
 
@@
 nams <- colnames(t1)
-nams[nams == 'cenTarsus'] <- 'Tarsus'
-nams[nams == 'cenNonHelper'] <- 'Non helpers'
-nams[nams == 'SexMales'] <- 'Sex'
-nams[nams == 'cenTQ'] <- 'TQ'
-nams[nams == 'cenHelper'] <- 'Helpers'
-nams[nams == 'cenHelper:Fledged'] <- 'Ageclass x Helpers'
-nams[nams == 'cenNonHelper:Fledged'] <- 'Ageclass x Non helpers'
-nams[nams == 'cenTarsus:Fledged'] <- 'Ageclass x Tarsus'
-nams[nams == 'cenTQ:Fledged'] <- 'Ageclass x TQ'
-nams[nams == 'Fledged:SexMales'] <- 'Ageclass x Sex'
-nams[nams == 'Fledged'] <- 'Ageclass'
+
+nams[nams == 'z.NonHelper'] <- 'Non helpers'
+nams[nams == 'c.Sex'] <- 'Sex'
+nams[nams == 'z.TQ'] <- 'TQ'
+nams[nams == 'c.Helper'] <- 'Helpers'
+nams[nams == 'z.BodyMass'] <- 'Body mass'
+nams[nams == 'z.Agemonths'] <- 'Age'
+nams[nams == 'z.Tarsus'] <- 'Tarsus'
+nams[nams == 'z.Agemonths:z.NonHelper'] <- 'Age x non helpers'
+nams[nams == 'c.Sex:z.Agemonths'] <- 'Age x sex'
+nams[nams == 'z.Agemonths:z.TQ'] <- 'Age x TQ'
+nams[nams == 'c.Helper:z.Agemonths'] <- 'Age x helpers'
+nams[nams == 'z.Agemonths:z.BodyMass'] <- 'Age x mass'
+nams[nams == 'z.Agemonths:z.Tarsus'] <- 'Age x tarsus'
 
@@
 
-tableS1 <- data.frame(Model = apply(t1,1,getterms,c(2:12)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
+tableS1 <- data.frame(Model = apply(t1,1,getterms,c(2:14)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
 colnames(tableS1) <- c('Model','df','AICc', 'Delta AICc','Weight')
@@
 ```
+
+**Table S2** Model selection table for analysis of factors affecting early life telomere loss in Seychelles warbler chicks, using longitudinal data.
+Only models that remained in the top model set for model averaging (AICc $\leq$ 6 compared to the top model) are included here.
+TQ = territory quality.
+
+```{r Table S2}
+temp <- subset(subset(Loss,Agemonths == 1),TimeDiff>0.3)
+
+#Run the model
+lm2 <- lmer(TROC ~ Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ + (1|LayYear),
+            data = temp,
+            REML = FALSE)
+lm2 <- standardize(lm2)
+#Model averaging
+
+lm3 <- dredge(lm2)
+lm3 <- subset(lm3,delta <=6)
+
+#Model averaging
+t1 <- as.matrix(subset(lm3, delta<=6))
+
+#Sort out row/col names
+nams <- colnames(t1)
+
+nams[nams == 'z.NonHelper'] <- 'Non helpers'
+nams[nams == 'c.Sex'] <- 'Sex'
+nams[nams == 'z.TQ'] <- 'TQ'
+nams[nams == 'c.Helper'] <- 'Helpers'
+nams[nams == 'z.BodyMass'] <- 'Body mass'
+nams[nams == 'z.Agemonths'] <- 'Age'
+nams[nams == 'z.Tarsus'] <- 'Tarsus'
+nams[nams == 'z.Agemonths:z.NonHelper'] <- 'Age x non helpers'
+nams[nams == 'c.Sex:z.Agemonths'] <- 'Age x sex'
+nams[nams == 'z.Agemonths:z.TQ'] <- 'Age x TQ'
+nams[nams == 'c.Helper:z.Agemonths'] <- 'Age x helpers'
+nams[nams == 'z.Agemonths:z.BodyMass'] <- 'Age x mass'
+nams[nams == 'z.Agemonths:z.Tarsus'] <- 'Age x tarsus'
+
+colnames(t1) <- nams
+
+tableS1 <- data.frame(Model = apply(t1,1,getterms,c(2:7)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
+colnames(tableS1) <- c('Model','df','AICc', 'Delta AICc','Weight')
+rownames(tableS1) <- NULL
+tableS1$Model[tableS1$Model == ''] <- 'Null model'
+tableS1[,3:5] <- lapply(tableS1[,3:5],as.numeric)
+kable(tableS1,digits=2)
+
+```
+
+**Figure S1**
+```{r Figure S1}
+#####################
+#4A
+#####################
+temp <- subset(subset(Loss,Agemonths == 1),TimeDiff>0.3)
+
+#Run the model
+lm2 <- lmer(TROC ~ Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ + (1|LayYear),
+            data = temp,
+            REML = FALSE)
+lm2 <- standardize(lm2)
+#Model averaging
+
+lm3 <- dredge(lm2)
+lm3 <- subset(lm3,delta <=6)
+lmavg <- model.avg(lm3,fit=T)
+myimp <- importance(lmavg) #Relative importance of each explanatory variable
+
+#Put it all into a nice table
+table2 <- imptable(lmavg,myimp)
+table2 <- signif(table2,2)
+
+#Sort out row/col names
+nams <- rownames(table2)
+nams[nams == 'z.NonHelper'] <- 'Non helpers'
+nams[nams == 'c.Sex'] <- 'Sex'
+nams[nams == 'z.TQ'] <- 'TQ'
+nams[nams == 'c.Helper'] <- 'Helpers'
+nams[nams == 'z.BodyMass'] <- 'Body mass'
+nams[nams == 'z.Agemonths'] <- 'Age'
+nams[nams == 'z.LogTL'] <- 'Initial TL'
+nams[nams == 'z.Tarsus'] <- 'Tarsus'
+nams[nams == 'z.Agemonths:z.NonHelper'] <- 'Age x non helpers'
+nams[nams == 'c.Sex:z.Agemonths'] <- 'Age x sex'
+nams[nams == 'z.Agemonths:z.TQ'] <- 'Age x TQ'
+nams[nams == 'c.Helper:z.Agemonths'] <- 'Age x helpers'
+nams[nams == 'z.Agemonths:z.BodyMass'] <- 'Age x mass'
+nams[nams == 'z.Agemonths:z.Tarsus'] <- 'Age x tarsus'
+
+rownames(table2) <- nams
+colnames(table2) <- c('Estimate','SE','2.5%','97.5%','RI')
+
+table2 <- data.frame(table2,myx=row.names(table2))
+
+FigS1 <- forestplot(table2)
+FigS1
+
+```

### 2016-02-03 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Redraft edits

--- 2016-01-19 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
@@
 
-**Figure 1** Telomere length (mean $\pm$ s.e.) in relation to age in Seychelles warblers, based on cross-sectional **(A)** and longitudinal **(B)** data.
-In **B**, telomere loss is a standardised and corrected rate (see Methods for details) based on the difference in telomere length in the sample taken from birds in their first year of life, and the first available sample as an adult.
-Values above zero indicate telomere shortening in early life.
+**Figure 1** Telomere length in relation to age in Seychelles warblers.
+**(A)** Mean  ($\pm$ s.e.) age for all birds.
+**(B)** Notched boxplot of telomere length at different age classes within the first year of life.
+Sample sizes for each group are provided in brackets.
 
-**Figure 2** Temporal variation in early-life telomere dynamics in the Seychelles warbler based on cross-sectional **(A)** and longitudinal **(B)** data (see Fig.
-1 and main text for details).
-Points and error bars represent mean and standard error of telomere length for all birds born in each main (black) or minor (grey) breeding season, in relation to island-wide food availability in that season.
-Lines and shaded areas represent fitted values and 95% confidence limits from a linear regression.
+**Figure 2** Temporal variation in early-life telomere length in the Seychelles warbler based on cross-sectional data.
+Points and error bars represent mean and standard error of telomere length for all birds born in each summer breeding season.
+Sample sizes for each year are provided in brackets.
 
 **Figure 3** Factors affecting early-life telomere length in Seychelles warblers at the individual level, within seasons.
-**A** Model averaged estimates and 95% confidence intervals for explanatory terms used in a linear model with telomere length as the response variable.
-Numbers in brackets are the relative importance of each term in the top model set (see main text for details; TQ = territory quality).
-**B,C** Raw telomere length Presence/absence of helpers (**B**) and tarsus length (**C**) in nestlings (black, solid lines) and fledglings (grey, dashed lines).
-Points and error bars in **B** are means $\pm$ s.e., and the line and shaded areas in **C** represent fitted values and 95% confidence limits from a linear regression.
+Model averaged estimates and 95% confidence intervals for explanatory terms used in a linear mixed model with telomere length (cross-sectional data, **A**) and telomere loss (longitudinal data, **B**) as the response variable.
+Numbers in brackets are the relative importance of each term in the top model set (see main text for details; TQ = territory quality, TL = telomere length).
 
-**Figure 4** Temporal, cohort-level variation in lifespan in relation to early-life telomere dynamics in the Seychelles warbler, based on cross-sectional **(A)** and longitudinal **(B)** data (see Fig.
-1 and main text for details).
-Points and error bars represent means and $\pm$ s.e., respectively, of all birds born in each main (black, solid lines) or minor (grey, dashed lines) breeding season.
-Lines and shaded areas represent fitted values and 95% confidence limits from a linear regression.
+**Figure 4** Telomere length in relation to **A** season and **B** tarsus length in juvenile Seychelles warblers.
+In *A**, white and grey boxes represent summer and winter seasons, respectively, while in **B** black and grey points represent nestlings and fledglings, respectively.
+Sample sizes for each group are provided in brackets.
 
-**Figure 5** Kaplan-Meier curves showing the relationship between survival and telomere length for individual Seychelles warblers measured as chicks (**A**), and fledglings  (**B**).
+**Figure 5** Kaplan-Meier curves showing the relationship between survival and telomere length for juvenile Seychelles warblers.
 Telomere length is binned into groups here for visualisation purposes only (long and short = greater than or less than median telomere length, respectively).
@@
 
-####################
-temp <- subset(juvseason,Season == 'Summer')
-temp <- subset(temp,FieldPeriodID != 27)
+temp <- dd
+temp$Agemonths[temp$Agemonths>10] <- '12+'
+temp$Agemonths <- factor(temp$Agemonths,levels = c('1','6','10','12+'))
 
-Fig1 <- ggplot(temp,
-       aes(x = CatchYear,
-           y = TLKBmean,
-           col = Season)) +
+Fig1A <- ggplot(temp,aes(x = Agemonths,y = LogTL))+
+  geom_jitter(col='grey') +
   ylab('Log telomere length (kb)') + 
-  xlab('Lay Year') +
+  xlab('Age (months)') +
   theme_classic() +
-  theme(axis.text.x = element_text(size = 14), 
-        axis.text.y = element_text(size = 14), 
-        axis.title.y = element_text(size = 16,vjust=0.8),
-        axis.title.x = element_text(size = 16),
-        legend.position = 'none') +
-  geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse)) +
-  geom_line()+
-  scale_color_manual(values = c('black',grey(0.5)))
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        legend.position = 'none')+
+  geom_boxplot(outlier.colour=NA)+
+  annotate('text',x = 0.6,y = 4.8,label = 'A')
 
-################################
+#
+temp <- Loss
+temp[,'Age (months)'] <- factor(temp$Agemonths)
+
+Fig1B <- ggplot(temp,aes(x = LogTL,y = LogTL1,col = `Age (months)`,shape = `Age (months)`)) +
+  ylab('Log adult telomere length (kb)') + 
+  xlab('Log juvenile telomere length (kb)') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        legend.position = 'none')+
+  geom_point() +
+  stat_smooth(method = 'lm')+
+  xlim(2,4.5)+
+  ylim(2,4.5)+
+    scale_colour_manual(values = c('darkgrey','darkgreen','darkorange'))+
+  annotate('text',x = 2,y = 4.5,label = 'B')
+
+#
+temp <- Loss
+temp[,'Age (months)'] <- factor(temp$Agemonths)
+
+Fig1C <- ggplot(temp,aes(x = TimeDiff,y = Loss,col=`Age (months)`,shape = `Age (months)`)) +
+  ylab('Log telomere loss (kb)') + 
+  xlab('Time interval (years)') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        legend.position = 'none')+
+  geom_point()+
+  geom_abline(slope = 0,intercept = 0,lty = 2)+
+  stat_smooth(method='lm')+
+  scale_colour_manual(values = c('darkgrey','darkgreen','darkorange'))+
+  annotate('text',x = 0,y = 2,label = 'C')
+
+Fig1D <- ggplot(temp,aes(x = TimeDiff,y = TROC,col=`Age (months)`,shape = `Age (months)`)) +
+  ylab('Log telomere loss rate (kb/year)') + 
+  xlab('Time interval (years)') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        legend.position = c(0.9,0.8))+
+  geom_point()+
+  geom_abline(slope = 0,intercept = 0,lty = 2)+
+  scale_colour_manual(values = c('darkgrey','darkgreen','darkorange'))+
+  annotate('text',x = 0,y = 8,label = 'D')
 
 pdf('Figures/Figure 1.pdf',
-    width = 6,
-    height = 4,
-    paper = 'a4')
-Fig1
+    width = 8,
+    height = 8,
+    paper = 'a4r')
+multiplot(Fig1A,Fig1B,Fig1C,Fig1D,layout=matrix(c(1:4),nrow = 2,byrow=T))
 dev.off()
@@
 
-####################
-Fig1 <- ggplot(dd,
-       aes(x = Fledged,
-           y = LogTL,
-           col = Season)) +
-  ylab('Telomere length (kb)') + 
-  xlab('Lay Year') +
+###############################
+temp <- juv
+temp$LayYear <- factor(temp$LayYear)
+
+Fig2A <- ggplot(subset(temp,Agemonths == 1),
+       aes(x = LayYear,
+           y = LogTL)) +
+  ylab('') + 
+  xlab('') +
   theme_classic() +
-  theme(axis.text.x = element_text(size = 14), 
+  theme(axis.text.x = element_text(angle = 60,size = 14,vjust=0.5), 
         axis.text.y = element_text(size = 14), 
@@
         legend.position = 'none') +
-  geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse)) +
-  geom_line()+
-  scale_color_manual(values = c('black',grey(0.5)))
+  geom_jitter(col = 'grey')+
+  ylim(c(2,4.5))+
+  geom_boxplot(outlier.colour = NA)+
+  scale_x_discrete(limits = factor(1998:2014),drop=F)+
+  annotate('text',x = 1,y = 4.4,label = 'A')
 
-################################
+Fig2B <- ggplot(subset(juv,Agemonths == 6),
+       aes(x = factor(LayYear),
+           y = LogTL)) +
+  ylab('Log telomere length (kb)') + 
+  xlab('') +
+  theme_classic() +
+  theme(axis.text.x = element_text(angle = 60,size = 14,vjust=0.5), 
+        axis.text.y = element_text(size = 14), 
+        axis.title.y = element_text(size = 16,vjust=0.8),
+        axis.title.x = element_text(size = 16),
+        legend.position = 'none') +
+  geom_jitter(col = 'grey')+
+  ylim(c(2,4.5))+
+  geom_boxplot(outlier.colour = NA)+
+  scale_x_discrete(limits = factor(1998:2014),drop=F)+
+  annotate('text',x = 1,y = 4.4,label = 'B')
 
-pdf('Figures/Figure 1.pdf',
+Fig2C <- ggplot(subset(juv,Agemonths == 10),
+       aes(x = factor(LayYear),
+           y = LogTL)) +
+  ylab('') + 
+  xlab('Cohort') +
+  theme_classic() +
+  theme(axis.text.x = element_text(angle = 60,size = 14,vjust=0.5), 
+        axis.text.y = element_text(size = 14), 
+        axis.title.y = element_text(size = 16,vjust=0.8),
+        axis.title.x = element_text(size = 16),
+        legend.position = 'none') +
+  geom_jitter(col = 'grey')+
+  ylim(c(2,4.5))+
+  geom_boxplot(outlier.colour = NA)+
+  scale_x_discrete(limits = factor(1998:2014),drop=F)+
+  annotate('text',x = 1,y = 4.4,label = 'C')
+
+pdf('Figures/Figure 2.pdf',
     width = 6,
-    height = 4,
-    paper = 'a4')
-Fig1
+    height = 8,
+    paper = 'a4r')
+multiplot(Fig2A,Fig2B,Fig2C)
 dev.off()
@@
 
+temp <- juv
+
+#####################
+#3A
 #####################
@@
 #Run the model
-lm2 <- lmer(LogTL ~ Fledged + Tarsus + Helper + GroupSize + Sex +Tarsus + BodyMass  + cenTQ  + (1|LayYear) + (1|PlateID),
-            data = juv,
+lm2 <- lmer(LogTL ~ Agemonths * (Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ)  + (1|LayYear) + (1|PlateID) + (1|TerritoryID),
+            data = temp,
             REML = FALSE)
-
+lm2 <- standardize(lm2)
 #Model averaging
@@
 nams <- rownames(table2)
-nams[nams == 'cenTarsus'] <- 'Tarsus'
-nams[nams == 'cenNonHelper'] <- 'Non helpers'
-nams[nams == 'SexMales'] <- 'Sex'
-nams[nams == 'cenTQ'] <- 'TQ'
-nams[nams == 'cenHelper'] <- 'Helpers'
-nams[nams == 'cenHelper:FledgedFledglings'] <- 'Ageclass x Helpers'
-nams[nams == 'cenNonHelper:FledgedFledglings'] <- 'Ageclass x Non helpers'
-nams[nams == 'cenTarsus:FledgedFledglings'] <- 'Ageclass x Tarsus'
-nams[nams == 'cenTQ:FledgedFledglings'] <- 'Ageclass x TQ'
-nams[nams == 'FledgedFledglings:SexMales'] <- 'Ageclass x Sex'
-nams[nams == 'FledgedFledglings'] <- 'Ageclass'
+nams[nams == 'z.NonHelper'] <- 'Non helpers'
+nams[nams == 'c.Sex'] <- 'Sex'
+nams[nams == 'z.TQ'] <- 'TQ'
+nams[nams == 'c.Helper'] <- 'Helpers'
+nams[nams == 'z.BodyMass'] <- 'Body mass'
+nams[nams == 'z.Agemonths'] <- 'Age'
+nams[nams == 'z.Tarsus'] <- 'Tarsus'
+nams[nams == 'z.Agemonths:z.NonHelper'] <- 'Age x non helpers'
+nams[nams == 'c.Sex:z.Agemonths'] <- 'Age x sex'
+nams[nams == 'z.Agemonths:z.TQ'] <- 'Age x TQ'
+nams[nams == 'c.Helper:z.Agemonths'] <- 'Age x helpers'
+nams[nams == 'z.Agemonths:z.BodyMass'] <- 'Age x mass'
+nams[nams == 'z.Agemonths:z.Tarsus'] <- 'Age x tarsus'
 
@@
 table2 <- data.frame(table2,myx=row.names(table2))
+table2$myx <- factor(table2$myx,levels = table2$myx[order(table2$Estimate,decreasing = T)])
 
-Fig3 <- forestplot(table2) +
-  annotate('text',label = 'A',x=0.6,y=0.1)
+Fig3 <- forestplot(table2)+
+  ylim(-0.4,0.4)
 
-pdf('Figures/Figure 2.pdf',
+pdf('Figures/Figure 3.pdf',
     width = 8,
-    height = 8,
-    paper = 'a4')
+    height = 5,
+    paper = 'a4r')
 Fig3
@@
 
-```{r Figure 3,fig.height=8,fig.width=4}
+```{r Figure 4}
+temp <- juv
 
-s1 <- survfit(Surv(time = Lifespan+1, event = Died)~TLF, data = chickall)
-Fig3A <- ggsurv(s1,plot.cens=F,lty.est=c(1,2)) + 
+temp[,'Helpers'] <- ifelse(temp$Helper>0,'Present', 'Absent')
+temp[,'Non helpers'] <- ifelse(temp$NonHelper>0,'Present', 'Absent')
+temp[,'Age (months)'] <- factor(temp$Agemonths)
+
+Fig4A <- ggplot(temp,
+       aes(x = factor(Agemonths),
+           y = LogTL,
+           fill=`Non helpers`)) +
+  ylab('Log telomere length (kb)') + 
+  xlab('Age (months)') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        legend.position = 'top') +
+  geom_jitter(aes(colour = `Non helpers`),position=position_jitterdodge(dodge.width = 0.8))+
+  geom_boxplot(outlier.colour = NA,notch=F)+
+  scale_colour_manual(values = c('grey','darkorange'))+
+  scale_fill_manual(values = c('grey','darkorange'))+
+  annotate('text',x = 0.55,y = 4.5,label = 'A')
+
+Fig4B <- ggplot(temp,
+       aes(x = factor(Agemonths),
+           y = LogTL,
+           fill=`Helpers`)) +
+  ylab('Log telomere length (kb)') + 
+  xlab('Age (months)') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        legend.position = 'top') +
+  geom_jitter(aes(colour = `Helpers`),position=position_jitterdodge(dodge.width = 0.8))+
+  geom_boxplot(outlier.colour = NA,notch=F)+
+  scale_colour_manual(values = c('grey','darkorange'))+
+  scale_fill_manual(values = c('grey','darkorange'))+
+  annotate('text',x = 0.55,y = 4.5,label = 'B')
+
+Fig4C <- ggplot(temp,
+       aes(x = TQ,
+           y = LogTL,
+           colour=`Age (months)`,
+           fill = `Age (months)`,
+           shape = `Age (months)`)) +
+  ylab('Log telomere length (kb)') + 
+  xlab('Territory quality') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        legend.position = 'top') +
+  geom_point(alpha = 0.2)+
+  stat_smooth(method = 'lm')+
+  scale_colour_manual(values = c(grey(0.2),'darkorange','darkgreen'))+
+  scale_fill_manual(values = c('darkgrey','darkorange','darkgreen'))+
+  annotate('text',x = 3,y = 4.5,label = 'C')
+
+Fig4D <- ggplot(temp,
+       aes(x = Tarsus,
+           y = LogTL,
+           colour=`Age (months)`,
+           fill = `Age (months)`,
+           shape = `Age (months)`)) +
+  ylab('Log telomere length (kb)') + 
+  xlab('Tarsus length (mm)') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        legend.position = 'top') +
+  geom_point(alpha = 0.2)+
+  stat_smooth(method = 'lm')+
+  scale_colour_manual(values = c(grey(0.2),'darkorange','darkgreen'))+
+  scale_fill_manual(values = c('darkgrey','darkorange','darkgreen'))+
+  annotate('text',x = 17,y = 4.5,label = 'D')
+
+pdf('Figures/Figure 4.pdf',
+    width = 7,
+    height = 8,
+    paper = 'a4r')
+multiplot(Fig4A,Fig4B,Fig4C,Fig4D,layout=matrix(c(1:4),byrow = TRUE,nrow = 2))
+dev.off()
+
+```
+
+```{r Figure 5}
+temp <- subset(juv,LayYear<2009)
+mymed <- median(temp$cohortTL)
+temp$coTLF <- ifelse(temp$cohortTL>mymed,'h','l')
+
+s1 <- survfit(Surv(time = Lifespan, event = Died)~coTLF, data = temp)
+Fig5A <- ggsurv(s1,plot.cens=F,lty.est=c(1:2)) + 
   theme_classic() +
@@
   scale_colour_manual(name = 'TL',
-                      values = c('black',grey(0.5))) +
-  guides(colour=F) +
-  annotate('text',x=0.5,y=1.1,label='A')
+                      values = c('grey','black')) +
+  guides(colour=F)
 
-###########################################
-s2 <- survfit(Surv(time = Lifespan+1, event = Died)~TLF, data = FlSAall)
-Fig3B <- ggsurv(s2,plot.cens=F,lty.est=c(1,2)) + 
+mymed <- median(temp$cenTL)
+temp$cenTLF <- ifelse(temp$cenTL>mymed,'h','l')
+s1 <- survfit(Surv(time = Lifespan, event = Died)~cenTLF, data = temp)
+Fig5B <- ggsurv(s1,plot.cens=F,lty.est=c(1,2)) + 
   theme_classic() +
@@
   scale_colour_manual(name = 'TL',
-                      values = c('black',grey(0.5))) +
-  guides(colour=F) +
-  annotate('text',x=0.5,y=1.1,label='B')
+                      values = c('black','grey')) +
+  guides(colour=F)
 
-pdf('Figures/Figure 3.pdf',
+pdf('Figures/Figure 5.pdf',
     width = 4,
@@
     paper = 'a4')
-multiplot(Fig3A,Fig3B)
+multiplot(Fig5A,Fig5B)
 dev.off()

### 2016-02-03 Scripts/SW juvenile telomeres IPC clean.R

@@ -50,7 +50,7 @@ dd$LeftTarsus <- NULL
 # Remove unwanted data/outliers ----------------------------------------------------
 
 
-dd <- droplevels(subset(subset(dd,TL>1000),TL<30000))
+dd <- droplevels(subset(subset(dd,TL>50),TL<30000))
 dd <- subset(dd,BodyMass>5)
 dd <- subset(dd,Tarsus>17)
 
@@ -65,8 +65,7 @@ dd$LogTL <- log10(dd$TL)
 
 # Age data ----------------------------------------------------------------
 
-dd$Age <- dd$CatchYear-dd$LayYear
-
+dd$Age <- (dd$CatchYear-dd$LayYear)+1
 
 #Sort out and order ageclass levels
 dd <- droplevels(dd[dd$Ageclass != '',])
@@ -79,24 +78,26 @@ dd$Fledged <- factor(dd$Fledged,levels = c('Nestlings','Fledglings','Adults'))
 
 
 dd$Agemonths <- ifelse(dd$Ageclass == 'CH',1,
-                       ifelse(dd$Ageclass == 'FL',3,
-                              ifelse(dd$Ageclass == 'SA',9,dd$Age*12)))
-dd$AgemonthF <- ifelse(dd$Ageclass == 'CH','<1',
-                       ifelse(dd$Ageclass == 'FL','1-9',
-                              ifelse(dd$Ageclass == 'SA','9-12','>12')))
-dd$AgemonthF <- factor(dd$AgemonthF,levels = c('<1','1-9','9-12','>12'))
+                       ifelse(dd$Ageclass == 'FL',6,
+                              ifelse(dd$Ageclass == 'SA',10,dd$Age*12)))
 
 dd <- subset(dd,Agemonths>0)
+
+dd$Age[dd$Fledged != 'Adults'] <- 1
 
 # Survival and lifespan ---------------------------------------------------
 
 dd$RemainingLife <- dd$DeathYear-dd$CatchYear
-dd$SurvivedNext <- ifelse(dd$RemainingLife>0,1,0)
-dd$Lifespan <- dd$DeathYear-dd$LayYear
+dd$SurvivedNext <- ifelse(dd$RemainingLife>1,1,0)
+dd$Lifespan <- (dd$DeathYear-dd$LayYear)+1
 dd$Died <- ifelse(dd$DeathYear<2013,1,0)
 
 
 
[…]

### 2016-02-03 Scripts/other messing about.r

@@ -1,0 +1,108 @@
+
+
+juv2 <- juv
+
+temp <- rbind(juv2[,c('BirdID','LogTL','Agemonths','Fledged','SurvivedNext','Died')],adults[,c('BirdID','LogTL','Agemonths','Fledged','SurvivedNext','Died')])
+temp$Agemonths <- factor(temp$Agemonths)
+temp <- subset(temp, Died ==1)
+
+output <- mat.or.vec(nlevels(temp$Agemonths),5)
+ss <- rep(NA,nlevels(temp$Agemonths))
+
+for(i in 1:nlevels(temp$Age))
+{
+  d1 <- subset(temp,Agemonths == levels(temp$Agemonths)[i+1])
+  d2 <- subset(temp,Agemonths == levels(temp$Agemonths)[i])
+  
+  d1 <- d1[!(duplicated(d1$BirdID)),]
+  d2 <- d2[!(duplicated(d2$BirdID)),]
+  
+  output[i,1] <- paste0(levels(temp$Agemonths)[i],' - ',levels(temp$Agemonths)[i+1])
+  output[i,2] <- mean(d1$LogTL)-mean(d2$LogTL)
+  
+  within1 <- subset(d1,BirdID %in% d2$BirdID)
+  
+  within2 <- subset(d2,BirdID %in% d1$BirdID)
+  ss[i] <- nrow(within2)
+  
+  within <- merge(within2,within1,by = 'BirdID')
+  output[i,3] <- mean((within$LogTL.y-within$LogTL.x))
+  
+  disap <- subset(d2,!(BirdID %in% d1$BirdID))
+  
+  
+  output[i,4] <- mean(d1$LogTL)-mean(disap$LogTL)
+  
+  ap <- subset(d1,!(BirdID %in% d2$BirdID))
+  output[i,5] <- mean(d1$LogTL)-mean(ap$LogTL)
+  
+}
+
+
+ss
+output <- output[complete.cases(output),]
+output <- output[1:7,]
+
+plot(output[,2],type = 'l',ylim=c(-0.5,0.3),axes=F)
+points(output[,3],type = 'l',col='red',lty=2)
+points(output[,4],type = 'l',col='darkgrey',lty=2)
+points(output[,5],type = 'l',col='blue',lty=2)
+abline(h=0,lty=3)
+
+axis(1,at=c(1:nrow(output)),labels=output[,1])
+axis(2)
+
+
+fa <- ggplot(juvseason,aes(x = Insect,y = TL))+
[…]

### 2016-02-08 Manuscript/SW_early_life_telomeres.Rmd
More edits

@@
 
-#####Early-life telomere length and age
+#####Early-life telomere dynamics
 Inter-plate repeatability of telomere length, based on XX samples measured at least twice, was 0.78 (CI = 0.73-0.82).
@@
 1A).
-Similarly, telomere shortening based on longitudinal data was significantly greater than zero in nestlings (one sample t-test; P < 0.001), but not in any other age group (all P > 0.5; Fig.
+Longitudinal data showed that within-individual telomere length between early and later life samples was significantly, albeit weakly, correlated (R^2^ = `r round2(summary(lm(LogTL1~LogTL,data=Loss))$r.squared)`; F = `r round2(anova(lm(LogTL1~LogTL,data=Loss))[1,4])`; P = `r round3(anova(lm(LogTL1~LogTL,data=Loss))[1,5])`; Fig.
 1B).
+In nestlings, the absolute reduction telomere in telomere length was not related to the length of time between sampling events (R^2^ = `r round2(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths == 1)))$r.squared)`; F = `r round2(anova(lm(Loss~TimeDiff,data=subset(Loss,Agemonths == 1)))[1,4])`; P = `r round3(anova(lm(Loss~TimeDiff,data=subset(Loss,Agemonths == 1)))[1,5])`), but was consistently greater than zero, even when the second sample was taken very shortly after the first (Fig.
+1C).
+Conversely, in both fledglings and subadults, the reduciton in telomere length was, on average, close to zero when the second sample was taken shortly after the first.
+The reduction in telomere length increased with the amount of time between sampling events in subadults (R^2^ = `r round2(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths == 10)))$r.squared)`; F = `r round2(anova(lm(Loss~TimeDiff,data=subset(Loss,Agemonths == 10)))[1,4])`; P = `r round3(anova(lm(Loss~TimeDiff,data=subset(Loss,Agemonths == 10)))[1,5])`), but not fledglings (R^2^ = `r round2(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths == 6)))$r.squared)`; F = `r round2(anova(lm(Loss~TimeDiff,data=subset(Loss,Agemonths == 6)))[1,4])`; P = `r round3(anova(lm(Loss~TimeDiff,data=subset(Loss,Agemonths == 6)))[1,5])`; Fig.
+1C).
+Conversely, the rate of telomere shortening decreased with the time between sampling events in chicks (R^2^ = `r round2(summary(lm(TROC~TimeDiff,data=subset(Loss,Agemonths == 1)))$r.squared)`; F = `r round2(anova(lm(TROC~TimeDiff,data=subset(Loss,Agemonths == 1)))[1,4])`; P = `r round3(anova(lm(TROC~TimeDiff,data=subset(Loss,Agemonths == 1)))[1,5])`), but not in fledglings (R^2^ = `r round2(summary(lm(TROC~TimeDiff,data=subset(Loss,Agemonths == 6)))$r.squared)`; F = `r round2(anova(lm(TROC~TimeDiff,data=subset(Loss,Agemonths == 6)))[1,4])`; P = `r round3(anova(lm(TROC~TimeDiff,data=subset(Loss,Agemonths == 6)))[1,5])`) or subadults (R^2^ = `r round2(summary(lm(TROC~TimeDiff,data=subset(Loss,Agemonths == 10)))$r.squared)`; F = `r round2(anova(lm(TROC~TimeDiff,data=subset(Loss,Agemonths == 10)))[1,4])`; P = `r round3(anova(lm(TROC~TimeDiff,data=subset(Loss,Agemonths == 10)))[1,5])`).
+Together, these results suggest that in our samples, telomere shortening occurs at the greatest rate shortly after the nestling stage, and comparitively little telomere shortening occurs later in life.
 
@@
 
-#####Individual-level variation in early-life telomere dynamics
-The top model explaining within-season variation in early-life telomere length contained age class and the number of non-helping subordinates in the natal territory (Table S1).
+#####Early-life environment and early-life telomere dynamics
+The top model explaining within-season variation in early-life telomere length contained age, territor the number of non-helping subordinates in the natal territory (Table S1).
 This model was much better supported than the null model ($\Delta$AICc = `r round2(AICc(lmer(LogTL~1+(1|FieldPeriodID)+(1|PlateID)+(1|BirdID),data=juv,REML = FALSE))-AICc(lmer(LogTL~Fledged+NonHelper+(1|FieldPeriodID)+(1|PlateID)+(1|BirdID),data=juv,REML = FALSE)))`).
-The age class effect was very strong and significant, and was present in every model in the top model set (Fig.
-3A).
-The effect of non-helpers, on the other had did not have higher importance across the top model set than the other variables, and confidence intervals overlapped zero (Fig.
-3A).
-Furthermore the model including non-helpers and age class model was not a significantly better fit than a model including age class alone ($\Delta$AICc = 0.06; Table S1).
 

### 2016-02-08 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
More edits

@@
   geom_abline(slope = 0,intercept = 0,lty = 2)+
+  stat_smooth(method = 'lm', formula = y~(1/exp(x)))+
   scale_colour_manual(values = c('darkgrey','darkgreen','darkorange'))+

### 2016-02-29 Manuscript/SW_early_life_telomeres.Rmd
More major edits

--- 2016-02-03 Manuscript/SW_early_life_telomeres.Rmd
@@
 We measured absolute telomere quantity per diploid genome for all samples using a quantitative PCR (qPCR) assay, outlined in detail by Barrett *et al.* [-@Barrett2012], with one amendment.
-A change in batch of SYBR green forced us to raise the annealing temperature of the telomere reaction from 58^o^C to 61^o^C for the majority of the samples (n = `r nrow(subset(dd,Objective %in% c(7:9)))`; see also below).
+A change in batch of SYBR green forced us to raise the annealing temperature of the telomere reaction from 58^o^C to 61^o^C for the majority of the samples (n = `r nrow(dd)`; see also below).
 Post-qPCR processing of samples was carried out as in Barrett *et al.* [-@Barrett2012], with the following amendments.
@@
 
-#####Early-life telomere length and age
+#####Early-life telomere dynamics
 Inter-plate repeatability of telomere length, based on XX samples measured at least twice, was 0.78 (CI = 0.73-0.82).
-Telomere length dereased with age within the first year of life (R^2^ = `r round2(summary(lm(LogTL~Agemonths,data=juv))$r.squared)`; F = `r round2(anova(lm(LogTL~Agemonths,data=juv))[1,4])`; P = `r round3(anova(lm(LogTL~Agemonths,data=juv))[1,5])`; Fig.
+Telomere length dereased with age within the first year of life (R^2^ = `r round2(summary(lm(RTL~Agemonths,data=juv))$r.squared)`; F = `r round2(anova(lm(RTL~Agemonths,data=juv))[1,4])`; P `r round3(anova(lm(RTL~Agemonths,data=juv))[1,5])`; Fig.
 1A).
-Similarly, telomere shortening based on longitudinal data was significantly greater than zero in nestlings (one sample t-test; P < 0.001), but not in any other age group (all P > 0.5; Fig.
-1B).
+Longitudinal data showed that the extent of telomere loss in early life was also greatest in the nestling phase, and decreased with age (R^2^ = `r round2(summary(lm(TROC~Agemonths,data=Loss))$r.squared)`; F = `r round2(anova(lm(TROC~Agemonths,data=Loss))[1,4])`; P `r round3(anova(lm(TROC~Agemonths,data=Loss))[1,5])`; Fig.
+1B)
 
-#####Cohort-level variation in early-life telomere dynamics
-Early-life telomere length varied significantly among breeding seasons (one-way ANOVA, F = `r round2(anova(lm(LogTL~factor(FieldPeriodID),data = juv))[1,4])`; P = `r round3(anova(lm(LogTL~factor(FieldPeriodID),data = juv))[1,5])`).
-Separating temporal variaiton in telomere length among age classes revealed no significant cohort effects in chicks (F = `r round2(anova(lm(LogTL~factor(FieldPeriodID),data = subset(juv,Agemonths == 1)))[1,4])`; P = `r round2(anova(lm(LogTL~factor(FieldPeriodID),data = subset(juv,Agemonths == 1)))[1,5])`; Fig.
-2A), but significant effects in fledglings (F = `r round2(anova(lm(LogTL~factor(FieldPeriodID),data = subset(juv,Agemonths == 6)))[1,4])`; P = `r round3(anova(lm(LogTL~factor(FieldPeriodID),data = subset(juv,Agemonths == 6)))[1,5])`; Fig.
-2B) and subadults (F = `r round2(anova(lm(LogTL~factor(FieldPeriodID),data = subset(juv,Agemonths == 9)))[1,4])`; P = `r round3(anova(lm(LogTL~factor(FieldPeriodID),data = subset(juv,Agemonths == 9)))[1,5])`; Fig.
+Both early-life telomere length and rate of loss varied significantly among breeding seasons (one-way ANOVA, telomere length: F = `r round2(anova(lm(RTL~factor(FieldPeriodID),data = juv))[1,4])`; P `r round3(anova(lm(RTL~factor(FieldPeriodID),data = juv))[1,5])`; Fig, 1C; telomere loss: F = `r round2(anova(lm(TROC~factor(FieldPeriodID),data = Loss))[1,4])`; P = `r round3(anova(lm(TROC~factor(FieldPeriodID),data = Loss))[1,5])`; Fig, 2).
+Variation in median telomere length over breeding seasons in fledglings and subadults was not related temporal variation in  territory quality (linear regression, R^2^ = `r round2(summary(lm(RTL~TQ,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(RTL~TQ,data = juvseason,na.action=na.exclude))[1,4])`; P = `r round2(anova(lm(RTL~TQ,data = juvseason,na.action=na.exclude))[1,5])`), island-wide food availability (R^2^ = `r round2(summary(lm(RTL~Insect,data=subset(juvseason,Insect<9),na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(RTL~Insect,data = subset(juvseason,Insect<9),na.action=na.exclude))[1,4])`; P = `r round2(anova(lm(RTL~Insect,data = subset(juvseason,Insect<9),na.action=na.exclude))[1,5])`) or population density (R^2^ = `r round2(summary(lm(RTL~Density,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(RTL~Density,data = juvseason,na.action=na.exclude))[1,4])`; P = `r round2(anova(lm(RTL~Density,data = juvseason,na.action=na.exclude))[1,5])`).
+
+#####Early-life environment and early-life telomere dynamics
+The top model explaining within-season variation in early-life RTL contained age, territory quality, the number of helpers in the natal territory, tarsus length, and interactions between age and the number of helpers and tarsus length (Table S1).
+All the the terms in the top model were significant (FIg.
+2A), and the top model was much better supported than a null model ($\Delta$AICc = `r round2(AICc(lmer(RTL~1+(1|LayYear)+(1|TerritoryID),data=juv,REML = FALSE))-AICc(lmer(RTL~Agemonths+Helper+TQ+Tarsus+Agemonths*Helper+Agemonths*Tarsus+(1|LayYear)+(1|TerritoryID),data=juv,REML = FALSE)))`).
+Tarsus length was negatively related to RTL in nestlings, unrelated to RTL in fledglings, and positively related to RTL in subadults (Fig.
+2B).
+The number of helpers was positvely associated with RTL in nestlings, but not in fledglings or subadults (Fig.
 2C).
-Variation in median telomere length over breeding seasons in fledglings and subadults was not related temporal variation in  territory quality (linear regression, R^2^ = `r round2(summary(lm(TL~TQ,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(TL~TQ,data = juvseason,na.action=na.exclude))[1,4])`; P = `r round2(anova(lm(TL~TQ,data = juvseason,na.action=na.exclude))[1,5])`), island-wide food availability (R^2^ = `r round2(summary(lm(TL~Insect,data=subset(juvseason,Insect<9),na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(TL~Insect,data = subset(juvseason,Insect<9),na.action=na.exclude))[1,4])`; P = `r round2(anova(lm(TL~Insect,data = subset(juvseason,Insect<9),na.action=na.exclude))[1,5])`) or population density (R^2^ = `r round2(summary(lm(TL~Density,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(TL~Density,data = juvseason,na.action=na.exclude))[1,4])`; P = `r round2(anova(lm(TL~Density,data = juvseason,na.action=na.exclude))[1,5])`).
+Finally, territory quality was negatively associated with RTL, and this effect was most pronounced in fledglings and subadults (Fig.
+2D).
 
-#####Individual-level variation in early-life telomere dynamics
-The top model explaining within-season variation in early-life telomere length contained age class and the number of non-helping subordinates in the natal territory (Table S1).
-This model was much better supported than the null model ($\Delta$AICc = `r round2(AICc(lmer(LogTL~1+(1|FieldPeriodID)+(1|PlateID)+(1|BirdID),data=juv,REML = FALSE))-AICc(lmer(LogTL~Fledged+NonHelper+(1|FieldPeriodID)+(1|PlateID)+(1|BirdID),data=juv,REML = FALSE)))`).
-The age class effect was very strong and significant, and was present in every model in the top model set (Fig.
-3A).
-The effect of non-helpers, on the other had did not have higher importance across the top model set than the other variables, and confidence intervals overlapped zero (Fig.
-3A).
-Furthermore the model including non-helpers and age class model was not a significantly better fit than a model including age class alone ($\Delta$AICc = 0.06; Table S1).
-
-The top model explaining telomere loss contained age class and tarsus length, and agian this was a significantly better fit than the null model ($\Delta$AICc = `r round2(AICc(lmer(TROC~1+(1|FieldPeriodID),data=Loss,REML = FALSE))-AICc(lmer(TROC~Fledged+Tarsus+(1|FieldPeriodID),data=Loss,REML = FALSE)))`).
-Fledglings lost telomeres at a lower rate compared to nestlings, while birds with longer tarsi higher rates of telomere shortening (Fig.
-3B).
+With the longitudinal data, the top model explaining telomere loss contained age, sex and body mass (Table S2).
+This model was a significantly better fit than the null model ($\Delta$AICc = `r round2(AICc(lmer(TROC~1+(1|LayYear),data=subset(Loss,TimeDiff>0.3),REML = FALSE))-AICc(lmer(TROC~Agemonths+Sex+BodyMass+(1|LayYear),data=subset(Loss,TimeDiff>0.3),REML = FALSE)))`), but was not a significantly better fit than a model containing only age ($\Delta$AICc = `r round2(AICc(lmer(TROC~Agemonths+(1|LayYear),data=subset(Loss,TimeDiff>0.3),REML = FALSE))-AICc(lmer(TROC~Agemonths+Sex+BodyMass+(1|LayYear),data=subset(Loss,TimeDiff>0.3),REML = FALSE)))`).
+Furthermore, the sex and body mass effects both had confidence intervals overlapping zero (Fig.
+3A), with very little signal of a pattern in the raw data (Fig.
+3B, 3C).
 
@@
 
-Telomere length in early life did not affect survival to adulthood (estimate = `r round2(summary(glm(SurvivedNext~LogTL+Agemonths,data=juv,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glm(SurvivedNext~LogTL+Agemonths,data=juv,family = 'binomial'))$coefficients[2,4])`), and there was no interaction effect between telomere length and juvenile age on survival to adulthood (estimate = `r round2(summary(glm(SurvivedNext~LogTL*Agemonths,data=juv,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glm(SurvivedNext~LogTL*Agemonths,data=juv,family = 'binomial'))$coefficients[4,4])`).
-To separate out cohort-level telomere length affected survival we calculated cohort-level telomere length (i.e.
+Telomere length in early life did not affect survival to adulthood (estimate = `r round2(summary(glm(SurvivedNext~RTL+Agemonths,data=juv,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glm(SurvivedNext~RTL+Agemonths,data=juv,family = 'binomial'))$coefficients[2,4])`), and there was no interaction effect between RTL and juvenile age on survival to adulthood (estimate = `r round2(summary(glm(SurvivedNext~RTL*Agemonths,data=juv,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glm(SurvivedNext~RTL*Agemonths,data=juv,family = 'binomial'))$coefficients[4,4])`).
+To separate out cohort-level RTL affected survival we calculated cohort-level RTL (i.e.
 the mean for each cohort) and added this term to a logistic regression.
-However, we found no effect of telomere length on survival to adulthood (estimate = `r round2(summary(glm(SurvivedNext~cohortTL+Agemonths,data=juv,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glm(SurvivedNext~cohortTL+Agemonths,data=juv,family = 'binomial'))$coefficients[2,4])`), and no interaction with juvenile age (estimate = `r round2(summary(glm(SurvivedNext~cohortTL*Agemonths,data=juv,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glm(SurvivedNext~cohortTL*Agemonths,data=juv,family = 'binomial'))$coefficients[4,4])`)
+However, we found no effect of RTL on survival to adulthood (estimate = `r round2(summary(glm(SurvivedNext~cohortTL+Agemonths,data=juv,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glm(SurvivedNext~cohortTL+Agemonths,data=juv,family = 'binomial'))$coefficients[2,4])`), and no interaction with juvenile age (estimate = `r round2(summary(glm(SurvivedNext~cohortTL*Agemonths,data=juv,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glm(SurvivedNext~cohortTL*Agemonths,data=juv,family = 'binomial'))$coefficients[4,4])`)
 
-Cox regression also showed no effect of individual-level or cohort-level telomere length on survival later in life (individual-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~LogTL+Fledged)))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~LogTL)))$coefficients[1,5])`;  cohort-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL+Fledged+frailty(LayYear))))$coefficients[2,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL+Fledged+frailty(LayYear))))$coefficients[2,6])`).
+Cox regression also showed no effect of individual-level or cohort-level RTL on survival later in life (individual-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL+Fledged)))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL)))$coefficients[1,5])`;  cohort-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL+Fledged+frailty(LayYear))))$coefficients[2,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL+Fledged+frailty(LayYear))))$coefficients[2,6])`).
 
@@
 ###Discussion
-Here we use the long-term study of a closed population of Seychelles warblers to assess the relationships between early-life conditions, telomere length and survival.
-We find that while telomere length varies
+Here we use the long-term study of a closed population of Seychelles warblers to assess the relationships between early-life conditions, telomere length and survival. We find that while RTL varies
 
-The clearest result from our study is that telomere length varies among cohorts.
+The clearest result from our study is that RTL varies among cohorts.
 Very few studies have shown that temporal variation in telomere dynamics occurs in natural populations, and to our knowledge the studies that have done so were limited to just two seasons [@Mizutani2013; @Watson2015].
@@
 
-In the Seychelles warbler we found only weak evidence that any of the social or environmental vairables we measured affect juvenile telomere length.
+In the Seychelles warbler we found only weak evidence that any of the social or environmental vairables we measured affect juvenile RTL.
 This is surprising a poor social and ecological environment is known to be detrimental to juvenile Seychelles warblers, both in terms of oxidative stress and later life survival.
-Possible explanations for this finding include i) temporal variation in telomere length in our data is so strong that we are unable to detect spatial trends, ii) within-cohort variation in telomere length is explained by a variable that we have not included in our analyses, or iii) that effects are generally weak and levels of noise in our telomere measurments and/or ecological data preclude detection of significant effects.
+Possible explanations for this finding include i) temporal variation in RTL in our data is so strong that we are unable to detect spatial trends, ii) within-cohort variation in RTL is explained by a variable that we have not included in our analyses, or iii) that effects are generally weak and levels of noise in our telomere measurments and/or ecological data preclude detection of significant effects.
 Future research should therefore examine, in a quantitative genetic framework, how genetic and environmental components, and their interactions, affect telomere dynamics and senescence in natural populations [@Becker2015; @Asghar2014].
 
-Our data suggest that, in addition to age class, tarsus length and season had weak effects on juvenile telomere length (Figs 3, 4).
+Our data suggest that, in addition to age class, tarsus length and season had weak effects on juvenile RTL (Figs 3, 4).
 The tarsus effect most likely reflects the fact that in passerine birds tarsus length is correlated with age during the nestling stage [@Ricklefs1976].
-Telomere loss is most rapid early in life due to ongoing cell replication [@Frenck1998], and a negative correlation between telomere length and body size in early life is therefore expected.
-Indeed, that age class affects both tarsus length and telomere lengths can be clearly seen in our data (Fig.
+Telomere loss is most rapid early in life due to ongoing cell replication [@Frenck1998], and a negative correlation between RTL and body size in early life is therefore expected.
+Indeed, that age class affects both tarsus length and RTL can be clearly seen in our data (Fig.
 4B).
-The difference in telomere length between birds born in summer and winter seasons is more surprising, especially given the directionality of the effect.
+The difference in RTL between birds born in summer and winter seasons is more surprising, especially given the directionality of the effect.
 We find that birds born in winter seasons having longer telomeres (Fig.

### 2016-02-29 Manuscript/SW_early_life_telomeres_SOM.Rmd
More major edits

@@
 #Run the model
-lm1 <- lmer(LogTL ~ Agemonths * (Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ) + (1|LayYear)+(1|PlateID)+(1|BirdID),
+lm1 <- lmer(RTL ~ Agemonths * (Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ) + (1|LayYear)+(1|TerritoryID),
             data = juv,
@@
 ```{r Table S2}
-temp <- subset(subset(Loss,Agemonths == 1),TimeDiff>0.3)
+temp <- subset(Loss,TimeDiff>0.3)
 
 #Run the model
-lm2 <- lmer(TROC ~ Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ + (1|LayYear),
+lm1 <- lmer(TROC ~ Agemonths * (Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ) + RTL + (1|LayYear),
             data = temp,
             REML = FALSE)
-lm2 <- standardize(lm2)
-#Model averaging
-
-lm3 <- dredge(lm2)
-lm3 <- subset(lm3,delta <=6)
+lm1 <- standardize(lm1)
 
 #Model averaging
-t1 <- as.matrix(subset(lm3, delta<=6))
+lm2 <- dredge(lm1)
+t1 <- as.matrix(subset(lm2, delta<=6))
 
@@
 
+nams[nams == 'z.RTL'] <- 'RTL'
 nams[nams == 'z.NonHelper'] <- 'Non helpers'
@@
 
-tableS1 <- data.frame(Model = apply(t1,1,getterms,c(2:7)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
-colnames(tableS1) <- c('Model','df','AICc', 'Delta AICc','Weight')
-rownames(tableS1) <- NULL
-tableS1$Model[tableS1$Model == ''] <- 'Null model'
-tableS1[,3:5] <- lapply(tableS1[,3:5],as.numeric)
-kable(tableS1,digits=2)
+tableS2 <- data.frame(Model = apply(t1,1,getterms,c(2:15)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
+colnames(tableS2) <- c('Model','df','AICc', 'Delta AICc','Weight')
+rownames(tableS2) <- NULL
+tableS2$Model[tableS2$Model == ''] <- 'Null model'
+tableS2[,3:5] <- lapply(tableS2[,3:5],as.numeric)
+kable(tableS2,digits=2)
 
@@
 #####################
-temp <- subset(subset(Loss,Agemonths == 1),TimeDiff>0.3)
+temp <- subset(Loss,TimeDiff>0.3)
+temp <- subset(temp,Agemonths == 1)
+temp <- subset(temp,TROC<2)
+temp <- subset(temp,TQI<4)
 
@@
 nams[nams == 'z.Agemonths'] <- 'Age'
-nams[nams == 'z.LogTL'] <- 'Initial TL'
+nams[nams == 'z.RTL'] <- 'Initial TL'
 nams[nams == 'z.Tarsus'] <- 'Tarsus'
@@
 ```
+
+ggplot(temp,aes(x = RTL,y = TROC,col=factor(Helper)))+
+geom_point()+
+stat_smooth(method = 'lm')

### 2016-02-29 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
More major edits

--- 2016-02-03 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
@@
 
-Fig1A <- ggplot(temp,aes(x = Agemonths,y = LogTL))+
+Fig1A <- ggplot(temp,aes(x = Agemonths,y = RTL))+
   geom_jitter(col='grey') +
-  ylab('Log telomere length (kb)') + 
+  ylab('RTL') + 
   xlab('Age (months)') +
@@
         legend.position = 'none')+
-  geom_boxplot(outlier.colour=NA)+
+  geom_boxplot(outlier.colour=NA)
   annotate('text',x = 0.6,y = 4.8,label = 'A')
 
-#
   temp <- Loss
@@
 
-Fig1B <- ggplot(temp,aes(x = LogTL,y = LogTL1,col = `Age (months)`,shape = `Age (months)`)) +
-  ylab('Log adult telomere length (kb)') + 
-  xlab('Log juvenile telomere length (kb)') +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        legend.position = 'none')+
-  geom_point() +
-  stat_smooth(method = 'lm')+
-  xlim(2,4.5)+
-  ylim(2,4.5)+
-    scale_colour_manual(values = c('darkgrey','darkgreen','darkorange'))+
-  annotate('text',x = 2,y = 4.5,label = 'B')
-
-#
-temp <- Loss
-temp[,'Age (months)'] <- factor(temp$Agemonths)
-
-Fig1C <- ggplot(temp,aes(x = TimeDiff,y = Loss,col=`Age (months)`,shape = `Age (months)`)) +
-  ylab('Log telomere loss (kb)') + 
-  xlab('Time interval (years)') +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        legend.position = 'none')+
-  geom_point()+
-  geom_abline(slope = 0,intercept = 0,lty = 2)+
-  stat_smooth(method='lm')+
-  scale_colour_manual(values = c('darkgrey','darkgreen','darkorange'))+
-  annotate('text',x = 0,y = 2,label = 'C')
-
-Fig1D <- ggplot(temp,aes(x = TimeDiff,y = TROC,col=`Age (months)`,shape = `Age (months)`)) +
-  ylab('Log telomere loss rate (kb/year)') + 
-  xlab('Time interval (years)') +
+Fig1B <- ggplot(temp,aes(x = `Age (months)`,y = TROC)) +
+    geom_jitter(col='grey') +
+  ylab('Loss in RTL') + 
+  xlab('Age (months)') +
   theme_classic() +
@@
         legend.position = c(0.9,0.8))+
-  geom_point()+
-  geom_abline(slope = 0,intercept = 0,lty = 2)+
-  scale_colour_manual(values = c('darkgrey','darkgreen','darkorange'))+
+  geom_boxplot(outlier.shape = NA)+
+  stat_smooth(method = 'lm', formula = y~(1/exp(x)))+
+  scale_colour_manual(values = c('darkgrey','darkgreen','darkorange'))
   annotate('text',x = 0,y = 8,label = 'D')
 
-pdf('Figures/Figure 1.pdf',
-    width = 8,
-    height = 8,
-    paper = 'a4r')
-multiplot(Fig1A,Fig1B,Fig1C,Fig1D,layout=matrix(c(1:4),nrow = 2,byrow=T))
-dev.off()
-
-```
-
-```{r Figure 2}
-
-###############################
 temp <- juv
@@
 
-Fig2A <- ggplot(subset(temp,Agemonths == 1),
+Fig1C <- ggplot(temp,
        aes(x = LayYear,
-           y = LogTL)) +
+           y = RTL)) +
   ylab('') + 
-  xlab('') +
+  xlab('RTL') +
   theme_classic() +
@@
   geom_jitter(col = 'grey')+
-  ylim(c(2,4.5))+
   geom_boxplot(outlier.colour = NA)+
-  scale_x_discrete(limits = factor(1998:2014),drop=F)+
-  annotate('text',x = 1,y = 4.4,label = 'A')
+  scale_x_discrete(limits = factor(1998:2014),drop=F)
+  annotate('text',x = 1,y = 4.4,label = 'C')
 
-Fig2B <- ggplot(subset(juv,Agemonths == 6),
-       aes(x = factor(LayYear),
-           y = LogTL)) +
-  ylab('Log telomere length (kb)') + 
-  xlab('') +
+temp <- Loss
+temp$LayYear <- factor(temp$LayYear)
+
+Fig1D <- ggplot(temp,
+       aes(x = LayYear,
+           y = RTL)) +
+  ylab('') + 
+  xlab('Loss in RTL') +
   theme_classic() +
@@
   geom_jitter(col = 'grey')+
-  ylim(c(2,4.5))+
   geom_boxplot(outlier.colour = NA)+
-  scale_x_discrete(limits = factor(1998:2014),drop=F)+
-  annotate('text',x = 1,y = 4.4,label = 'B')
+  scale_x_discrete(limits = factor(1998:2014),drop=F)
+  annotate('text',x = 1,y = 4.4,label = 'D')
 
-Fig2C <- ggplot(subset(juv,Agemonths == 10),
-       aes(x = factor(LayYear),
-           y = LogTL)) +
-  ylab('') + 
-  xlab('Cohort') +
-  theme_classic() +
-  theme(axis.text.x = element_text(angle = 60,size = 14,vjust=0.5), 
-        axis.text.y = element_text(size = 14), 
-        axis.title.y = element_text(size = 16,vjust=0.8),
-        axis.title.x = element_text(size = 16),
-        legend.position = 'none') +
-  geom_jitter(col = 'grey')+
-  ylim(c(2,4.5))+
-  geom_boxplot(outlier.colour = NA)+
-  scale_x_discrete(limits = factor(1998:2014),drop=F)+
-  annotate('text',x = 1,y = 4.4,label = 'C')
-
-pdf('Figures/Figure 2.pdf',
-    width = 6,
+pdf('Figures/Figure 1.pdf',
+    width = 8,
     height = 8,
     paper = 'a4r')
-multiplot(Fig2A,Fig2B,Fig2C)
+multiplot(Fig1A,Fig1B,Fig1C,Fig1D,layout=matrix(c(1,2,3,3,4,4),nrow = 3,byrow=T))
 dev.off()
@@
 
-```{r Figure 3}
+```{r Figure 2}
 
@@
 #Run the model
-lm2 <- lmer(LogTL ~ Agemonths * (Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ)  + (1|LayYear) + (1|PlateID) + (1|TerritoryID),
+lm2 <- lmer(RTL ~ Agemonths * (Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ)  + (1|LayYear) + (1|TerritoryID),
             data = temp,
@@
 
-Fig3 <- forestplot(table2)+
-  ylim(-0.4,0.4)
+Fig2A <- forestplot(table2)
 
-pdf('Figures/Figure 3.pdf',
-    width = 8,
-    height = 5,
-    paper = 'a4r')
-Fig3
-dev.off()
-
-```
-
-```{r Figure 4}
 temp <- juv
@@
 
-Fig4A <- ggplot(temp,
+Fig2B <- ggplot(temp,
        aes(x = factor(Agemonths),
-           y = LogTL,
-           fill=`Non helpers`)) +
-  ylab('Log telomere length (kb)') + 
-  xlab('Age (months)') +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        legend.position = 'top') +
-  geom_jitter(aes(colour = `Non helpers`),position=position_jitterdodge(dodge.width = 0.8))+
-  geom_boxplot(outlier.colour = NA,notch=F)+
-  scale_colour_manual(values = c('grey','darkorange'))+
-  scale_fill_manual(values = c('grey','darkorange'))+
-  annotate('text',x = 0.55,y = 4.5,label = 'A')
-
-Fig4B <- ggplot(temp,
-       aes(x = factor(Agemonths),
-           y = LogTL,
+           y = RTL,
            fill=`Helpers`)) +
-  ylab('Log telomere length (kb)') + 
+  ylab('RTL') + 
   xlab('Age (months)') +
@@
   scale_colour_manual(values = c('grey','darkorange'))+
-  scale_fill_manual(values = c('grey','darkorange'))+
+  scale_fill_manual(values = c('grey','darkorange'))
   annotate('text',x = 0.55,y = 4.5,label = 'B')
 
-Fig4C <- ggplot(temp,
+Fig2C <- ggplot(temp,
        aes(x = TQ,
-           y = LogTL,
+           y = RTL,
            colour=`Age (months)`,
@@
            shape = `Age (months)`)) +
-  ylab('Log telomere length (kb)') + 
+  ylab('RTL') + 
   xlab('Territory quality') +
@@
   scale_colour_manual(values = c(grey(0.2),'darkorange','darkgreen'))+
-  scale_fill_manual(values = c('darkgrey','darkorange','darkgreen'))+
+  scale_fill_manual(values = c('darkgrey','darkorange','darkgreen'))
   annotate('text',x = 3,y = 4.5,label = 'C')
 
-Fig4D <- ggplot(temp,
+Fig2D <- ggplot(temp,
        aes(x = Tarsus,
-           y = LogTL,
+           y = RTL,
            colour=`Age (months)`,
@@
            shape = `Age (months)`)) +
-  ylab('Log telomere length (kb)') + 
+  ylab('RTL') + 
   xlab('Tarsus length (mm)') +
@@
   scale_colour_manual(values = c(grey(0.2),'darkorange','darkgreen'))+
-  scale_fill_manual(values = c('darkgrey','darkorange','darkgreen'))+
+  scale_fill_manual(values = c('darkgrey','darkorange','darkgreen'))
   annotate('text',x = 17,y = 4.5,label = 'D')
 
-pdf('Figures/Figure 4.pdf',
-    width = 7,
-    height = 8,
+pdf('Figures/Figure 2.pdf',
+    width = 9,
+    height = 9,
     paper = 'a4r')
-multiplot(Fig4A,Fig4B,Fig4C,Fig4D,layout=matrix(c(1:4),byrow = TRUE,nrow = 2))
+multiplot(Fig2A,Fig2B,Fig2C,Fig2D,layout = matrix(c(1,1,1,2,3,4),nrow = 2,byrow = T))
 dev.off()
@@
 
-```{r Figure 5}
+```{r Figure 3}
+
+temp <- subset(Loss,TimeDiff>0.3)
+
+#####################
+#3A
+#####################
+
+#Run the model
+lm2 <- lmer(TROC ~ Agemonths * (Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ)  + RTL + (1|LayYear),
+            data = temp,
+            REML = FALSE)
+lm2 <- standardize(lm2)
+#Model averaging
+
+lm3 <- dredge(lm2)
+lm3 <- subset(lm3,delta <=6)
+lmavg <- model.avg(lm3,fit=T)
+myimp <- importance(lmavg) #Relative importance of each explanatory variable
+
+#Put it all into a nice table
+table2 <- imptable(lmavg,myimp)
+table2 <- signif(table2,2)
+
+#Sort out row/col names
+nams <- rownames(table2)
+nams[nams == 'z.RTL'] <- 'RTL'
+nams[nams == 'z.NonHelper'] <- 'Non helpers'
+nams[nams == 'c.Sex'] <- 'Sex'
+nams[nams == 'z.TQ'] <- 'TQ'
+nams[nams == 'c.Helper'] <- 'Helpers'
+nams[nams == 'z.BodyMass'] <- 'Body mass'
+nams[nams == 'z.Agemonths'] <- 'Age'
+nams[nams == 'z.Tarsus'] <- 'Tarsus'
+nams[nams == 'z.Agemonths:z.NonHelper'] <- 'Age x non helpers'
+nams[nams == 'c.Sex:z.Agemonths'] <- 'Age x sex'
+nams[nams == 'z.Agemonths:z.TQ'] <- 'Age x TQ'
+nams[nams == 'c.Helper:z.Agemonths'] <- 'Age x helpers'
+nams[nams == 'z.Agemonths:z.BodyMass'] <- 'Age x mass'
+nams[nams == 'z.Agemonths:z.Tarsus'] <- 'Age x tarsus'
+
+rownames(table2) <- nams
+colnames(table2) <- c('Estimate','SE','2.5%','97.5%','RI')
+
+table2 <- data.frame(table2,myx=row.names(table2))
+table2$myx <- factor(table2$myx,levels = table2$myx[order(table2$Estimate,decreasing = T)])
+
+Fig3A <- forestplot(table2)
+
+temp[,'Helpers'] <- ifelse(temp$Helper>0,'Present', 'Absent')
+temp[,'Non helpers'] <- ifelse(temp$NonHelper>0,'Present', 'Absent')
+temp[,'Age (months)'] <- factor(temp$Agemonths)
+
+Fig3B <- ggplot(temp,
+       aes(x = factor(Agemonths),
+           y = TROC,
+           fill=`Helpers`)) +
+  ylab('Loss in RTL') + 
+  xlab('Age (months)') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        legend.position = 'top') +
+  geom_jitter(aes(colour = `Helpers`),position=position_jitterdodge(dodge.width = 0.8))+
+  geom_boxplot(outlier.colour = NA,notch=F)+
+  scale_colour_manual(values = c('grey','darkorange'))+
+  scale_fill_manual(values = c('grey','darkorange'))
+  annotate('text',x = 0.55,y = 4.5,label = 'B')
+
+Fig3C <- ggplot(temp,
+       aes(x = factor(Agemonths),
+           y = TROC,
+           fill=Sex)) +
+  ylab('Loss in RTL') + 
+  xlab('Age (months)') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        legend.position = 'top') +
+  geom_jitter(aes(colour = Sex),position=position_jitterdodge(dodge.width = 0.8))+
+  geom_boxplot(outlier.colour = NA,notch=F)+
+  scale_colour_manual(values = c('grey','darkorange'))+
+  scale_fill_manual(values = c('grey','darkorange'))
+  annotate('text',x = 0.55,y = 4.5,label = 'B')
+
+Fig3D <- ggplot(temp,
+       aes(x = BodyMass,
+           y = TROC,
+           col=`Age (months)`)) +
+  ylab('Loss in RTL') + 
+  xlab('Body mass (g)') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        legend.position = 'top') +
+  geom_point(alpha = 0.2)+
+  stat_smooth(method = 'lm')+
+    scale_colour_manual(values = c(grey(0.2),'darkorange','darkgreen'))+
+  scale_fill_manual(values = c('darkgrey','darkorange','darkgreen'))
+  annotate('text',x = 17,y = 4.5,label = 'D')
+
+pdf('Figures/Figure 3.pdf',
+    width = 9,
+    height = 9,
+    paper = 'a4r')
+multiplot(Fig3A,Fig3B,Fig3C,Fig3D,layout = matrix(c(1,1,1,2,3,4),nrow = 2,byrow = T))
+dev.off()
+
+```
+
+```{r Figure 4}
 temp <- subset(juv,LayYear<2009)
@@
 s1 <- survfit(Surv(time = Lifespan, event = Died)~coTLF, data = temp)
-Fig5A <- ggsurv(s1,plot.cens=F,lty.est=c(1:2)) + 
+Fig4A <- ggsurv(s1,plot.cens=F,lty.est=c(1:2)) + 
   theme_classic() +
@@
 
+temp <- subset(temp,!(is.na(cenTL)))
 mymed <- median(temp$cenTL)
 temp$cenTLF <- ifelse(temp$cenTL>mymed,'h','l')
+
 s1 <- survfit(Surv(time = Lifespan, event = Died)~cenTLF, data = temp)
-Fig5B <- ggsurv(s1,plot.cens=F,lty.est=c(1,2)) + 
+Fig4B <- ggsurv(s1,plot.cens=F,lty.est=c(1,2)) + 
   theme_classic() +
@@
 
-pdf('Figures/Figure 5.pdf',
+pdf('Figures/Figure 4.pdf',
     width = 4,
@@
     paper = 'a4')
-multiplot(Fig5A,Fig5B)
+multiplot(Fig4A,Fig4B)
 dev.off()

### 2016-02-29 Scripts/SW juvenile telomeres IPC clean.R

@@ -4,11 +4,11 @@
 #################################################################################
 
 #Only use Ellie's samples
-dd0 <- subset(dd0,Whodunnit == 'EAF')
+#dd0 <- subset(dd0,Whodunnit == 'EAF')
 
 #Average repeats of blood samples
-av <- ave(dd0$TL,c(dd0$BloodID,dd0$Status,dd0$PlateID))
-dd0$TL <- av
+av <- ave(dd0$RTL,c(dd0$BloodID,dd0$Status,dd0$PlateID))
+dd0$RTL <- av
 dd <- dd0[!(duplicated(dd0$BloodID)),]
 
 
@@ -49,17 +49,10 @@ dd$LeftTarsus <- NULL
 
 # Remove unwanted data/outliers ----------------------------------------------------
 
-
-dd <- droplevels(subset(subset(dd,TL>50),TL<30000))
+dd <- subset(dd,RTL<3)
+dd <- subset(dd,RTL > 0.04)
 dd <- subset(dd,BodyMass>5)
 dd <- subset(dd,Tarsus>17)
-
-
-# Telomere data -----------------------------------------------------------
-
-dd$TLKB <- dd$TL/1000
-dd$LogTL <- log10(dd$TL)
-
 
 
 
@@ -124,8 +117,8 @@ juv$cohortTL <- NA
 for(i in 1:nrow(juv))
 {
   currentdata <- subset(juv,FieldPeriodID == FieldPeriodID[i])
-  juv$cenTL[i] <- (juv$LogTL[i] - mean(currentdata$LogTL))/sd(currentdata$LogTL)
-  juv$cohortTL[i] <- mean(currentdata$LogTL)
+  juv$cenTL[i] <- (juv$RTL[i] - mean(currentdata$RTL))/sd(currentdata$RTL)
+  juv$cohortTL[i] <- mean(currentdata$RTL)
 }
 
 mymed <- median(juv$cenTL,na.rm=T)
@@ -251,22 +244,23 @@ x4 <- x4[order(x4$BirdID),]
 
 
 #Apply Verhulst's correction for regresison to the mean
-xx1 <- x4$TLKB-mean(x4$TLKB)
-xx2 <- x2$TLKB-mean(x2$TLKB)          
-rho <- cor(x4$TLKB,x2$TLKB)
+xx1 <- x4$RTL-mean(x4$RTL)
+xx2 <- x2$RTL-mean(x2$RTL)          
+rho <- cor(x4$RTL,x2$RTL)
 
[…]

### 2016-03-01 Manuscript/SW_early_life_telomeres.Rmd
More edits (RTL)

@@
 All DNA extractions that passed the above criteria were diluted to 3.3 ng $\mu$l^-1^ before telomere measurement.
-We measured absolute telomere quantity per diploid genome for all samples using a quantitative PCR (qPCR) assay, outlined in detail by Barrett *et al.* [-@Barrett2012], with one amendment.
-A change in batch of SYBR green forced us to raise the annealing temperature of the telomere reaction from 58^o^C to 61^o^C for the majority of the samples (n = `r nrow(dd)`; see also below).
-Post-qPCR processing of samples was carried out as in Barrett *et al.* [-@Barrett2012], with the following amendments.
-First, averaging of technical repeats was carried out using custom-made R scripts (available as supplementary material) instead of GenEx.
-Second, the change in annealing temperature resulted in consistently higher, but repeatable, CQ values for the telomere reaction, so we implemented a simple correction to ensure that telomere lengths were repeatable across annealing temperatures (supplementary material).
-Finally, we excluded telomere lengths greater than 30kb, which maximised repeatability of our own data [see also @Bauch2012].
+We measured telomere length for all samples using a quantitative PCR (qPCR) assay of telomeres and a GPADH control gene, using the molecular methods outlined by Barrett *et al.* [-@Barrett2012], with one amendment.
+A change in batch of SYBR green forced us to raise the annealing temperature of the telomere reaction from 58^o^C to 61^o^C for the majority of samples.
+However, this did not affect final telomere length values (see below).
+We used the program LinRegPCR [@Ruijter2009] to correct for baseline fluorescence and calculate efficiencies and Cq values for each sample replicate.
+Averaging of technical repeats was carried out using custom-made R scripts (available as supplementary material), excluding samples with Cq values differing by >0.5.
+We then calculated relative telomere length (RTL) for each sample using equation 1 in Pfaffi *et al* [@Pfaffi2001].
+We chose to use RTL rather than continuing with the previously used method for claculating absolute telomere length, as i) using RTL enabled us to run more samples per plate (as an oligo standard is not required), ii) the RTL method was less susceptible to batch effects (Appendix 1), and iii) very few other studies have calculated absolute telomere length, and our experience suggests that cross-species comparisons are unlikely to be reliable.
+
 Inter-plate repeatability of final telomere lengths was assessed using the R package rptR [@Schielzeth2011], and for all subsequent analyses we used mean telomere length per sample where we had repeats.
@@
 For a subset of first-year birds (n = `r nrow(Loss)` individuals) we had longitudinal data, with an additional sample taken as an adult.
-For these individuals we calculated the absolute amount of telomere loss between the first-year and adult samples by subtracting adult telomere length from early-life telomere length, as well as a rate of telomere loss by dividing this difference by the length of time (in days) between sampling events.
+For these individuals we calculated the rate of telomere loss by subtracting adult RTL from early-life RTL, and dividing this difference by the length of time (to the nearest 6 months) between sampling events.
 
@@
 #####Early-life telomere dynamics
-Inter-plate repeatability of telomere length, based on XX samples measured at least twice, was 0.78 (CI = 0.73-0.82).
-Telomere length dereased with age within the first year of life (R^2^ = `r round2(summary(lm(RTL~Agemonths,data=juv))$r.squared)`; F = `r round2(anova(lm(RTL~Agemonths,data=juv))[1,4])`; P `r round3(anova(lm(RTL~Agemonths,data=juv))[1,5])`; Fig.
+Efficiecies (mean $\pm$) for our telomere and GAPDH reactions were (`r round2(mean(dd$EffTelomere))` $\pm$ `r round2(sd(dd$EffTelomere))`) and (`r round2(mean(dd$EffGAPDH))` $\pm$ `r round2(sd(dd$EffGAPDH))`) respectively.
+Inter-plate repeatability of RTL, based on XX samples measured at least twice, was 0.78 (CI = 0.73-0.82).
+
+RTL dereased with age within the first year of life (R^2^ = `r round2(summary(lm(RTL~Agemonths,data=juv))$r.squared)`; F = `r round2(anova(lm(RTL~Agemonths,data=juv))[1,4])`; P `r round3(anova(lm(RTL~Agemonths,data=juv))[1,5])`), and across the entire Seychelles warbler lifespan (R^2^ = `r round2(summary(lm(RTL~Agemonths,data=dd))$r.squared)`; F = `r round2(anova(lm(RTL~Agemonths,data=dd))[1,4])`; P `r round3(anova(lm(RTL~Agemonths,data=dd))[1,5])`; Fig.
 1A).
-Longitudinal data showed that the extent of telomere loss in early life was also greatest in the nestling phase, and decreased with age (R^2^ = `r round2(summary(lm(TROC~Agemonths,data=Loss))$r.squared)`; F = `r round2(anova(lm(TROC~Agemonths,data=Loss))[1,4])`; P `r round3(anova(lm(TROC~Agemonths,data=Loss))[1,5])`; Fig.
-1B)
+Longitudinal data showed that the rate of early-life telomere shortening also decreased with age (R^2^ = `r round2(summary(lm(TROC~Agemonths,data=Loss))$r.squared)`; F = `r round2(anova(lm(TROC~Agemonths,data=Loss))[1,4])`; P = `r round3(anova(lm(TROC~Agemonths,data=Loss))[1,5])`), suggesting that the highest rates of telomere loss occurred following the nestling stage (Fig.
+1B).
 
-Both early-life telomere length and rate of loss varied significantly among breeding seasons (one-way ANOVA, telomere length: F = `r round2(anova(lm(RTL~factor(FieldPeriodID),data = juv))[1,4])`; P `r round3(anova(lm(RTL~factor(FieldPeriodID),data = juv))[1,5])`; Fig, 1C; telomere loss: F = `r round2(anova(lm(TROC~factor(FieldPeriodID),data = Loss))[1,4])`; P = `r round3(anova(lm(TROC~factor(FieldPeriodID),data = Loss))[1,5])`; Fig, 2).
-Variation in median telomere length over breeding seasons in fledglings and subadults was not related temporal variation in  territory quality (linear regression, R^2^ = `r round2(summary(lm(RTL~TQ,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(RTL~TQ,data = juvseason,na.action=na.exclude))[1,4])`; P = `r round2(anova(lm(RTL~TQ,data = juvseason,na.action=na.exclude))[1,5])`), island-wide food availability (R^2^ = `r round2(summary(lm(RTL~Insect,data=subset(juvseason,Insect<9),na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(RTL~Insect,data = subset(juvseason,Insect<9),na.action=na.exclude))[1,4])`; P = `r round2(anova(lm(RTL~Insect,data = subset(juvseason,Insect<9),na.action=na.exclude))[1,5])`) or population density (R^2^ = `r round2(summary(lm(RTL~Density,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(RTL~Density,data = juvseason,na.action=na.exclude))[1,4])`; P = `r round2(anova(lm(RTL~Density,data = juvseason,na.action=na.exclude))[1,5])`).
+Both early-life RTL and rate of loss varied significantly among cohorts (one-way ANOVA, telomere length: F = `r round2(anova(lm(RTL~factor(FieldPeriodID),data = juv))[1,4])`; P `r round3(anova(lm(RTL~factor(FieldPeriodID),data = juv))[1,5])`; Fig, 1C; telomere loss: F = `r round2(anova(lm(TROC~factor(FieldPeriodID),data = Loss))[1,4])`; P `r round3(anova(lm(TROC~factor(FieldPeriodID),data = Loss))[1,5])`).
+Variation in median RTL over breeding seasons in fledglings and subadults was not related temporal variation in  territory quality (linear regression, R^2^ = `r round2(summary(lm(RTL~TQ,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(RTL~TQ,data = juvseason,na.action=na.exclude))[1,4])`; P = `r round2(anova(lm(RTL~TQ,data = juvseason,na.action=na.exclude))[1,5])`), island-wide food availability (R^2^ = `r round2(summary(lm(RTL~Insect,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(RTL~Insect,data = juvseason,na.action=na.exclude))[1,4])`; P = `r round2(anova(lm(RTL~Insect,data = juvseason,na.action=na.exclude))[1,5])`) or population density (R^2^ = `r round2(summary(lm(RTL~Density,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(RTL~Density,data = juvseason,na.action=na.exclude))[1,4])`; P = `r round2(anova(lm(RTL~Density,data = juvseason,na.action=na.exclude))[1,5])`).
 
 #####Early-life environment and early-life telomere dynamics
-The top model explaining within-season variation in early-life RTL contained age, territory quality, the number of helpers in the natal territory, tarsus length, and interactions between age and the number of helpers and tarsus length (Table S1).
-All the the terms in the top model were significant (FIg.
-2A), and the top model was much better supported than a null model ($\Delta$AICc = `r round2(AICc(lmer(RTL~1+(1|LayYear)+(1|TerritoryID),data=juv,REML = FALSE))-AICc(lmer(RTL~Agemonths+Helper+TQ+Tarsus+Agemonths*Helper+Agemonths*Tarsus+(1|LayYear)+(1|TerritoryID),data=juv,REML = FALSE)))`).
+The top model explaining variation in early-life RTL contained age, season (summer vs winter), tarsus length, and the interaction between age and tarsus length (Table S1).
+The top model was much better supported than a null model ($\Delta$AICc = `r round2(AICc(lmer(RTL~1+(1|LayYear)+(1|TerritoryID),data=juv,REML = FALSE))-AICc(lmer(RTL~Agemonths+Season+Tarsus+Agemonths*Tarsus+(1|LayYear)+(1|TerritoryID),data=juv,REML = FALSE)))`), and all three effects were statistically significant (Fig.
+2A).
+RTL was higher in winter compared to summer seasons (Fig.
+2B).
 Tarsus length was negatively related to RTL in nestlings, unrelated to RTL in fledglings, and positively related to RTL in subadults (Fig.
-2B).
-The number of helpers was positvely associated with RTL in nestlings, but not in fledglings or subadults (Fig.
 2C).
-Finally, territory quality was negatively associated with RTL, and this effect was most pronounced in fledglings and subadults (Fig.
-2D).
 
-With the longitudinal data, the top model explaining telomere loss contained age, sex and body mass (Table S2).
-This model was a significantly better fit than the null model ($\Delta$AICc = `r round2(AICc(lmer(TROC~1+(1|LayYear),data=subset(Loss,TimeDiff>0.3),REML = FALSE))-AICc(lmer(TROC~Agemonths+Sex+BodyMass+(1|LayYear),data=subset(Loss,TimeDiff>0.3),REML = FALSE)))`), but was not a significantly better fit than a model containing only age ($\Delta$AICc = `r round2(AICc(lmer(TROC~Agemonths+(1|LayYear),data=subset(Loss,TimeDiff>0.3),REML = FALSE))-AICc(lmer(TROC~Agemonths+Sex+BodyMass+(1|LayYear),data=subset(Loss,TimeDiff>0.3),REML = FALSE)))`).
-Furthermore, the sex and body mass effects both had confidence intervals overlapping zero (Fig.
-3A), with very little signal of a pattern in the raw data (Fig.
-3B, 3C).
+Model averaging results from the longitudinal data are shown in Figure 3A.
+The top model explaining telomere loss contained age and season (Table S2).
+This model was a significantly better fit than the null model ($\Delta$AICc = `r round2(AICc(lmer(TROC~1+(1|LayYear),data=Loss,REML = FALSE))-AICc(lmer(TROC~Agemonths+Season+(1|LayYear),data=Loss,REML = FALSE)))`).
+Juveniles born in winter seasons had higher rates of telomere shortening compared to birds born in summer seasons, but this effect was weak (Fig.
+3B), although this effect was weak and confidence intervals overlapped zero (Fig.
+3A).
+There was no evidence that tarsus length was related to telomere shortening in the same way that telomere length was (Fig.
+3C).
 
@@
 the mean for each cohort) and added this term to a logistic regression.
-However, we found no effect of RTL on survival to adulthood (estimate = `r round2(summary(glm(SurvivedNext~cohortTL+Agemonths,data=juv,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glm(SurvivedNext~cohortTL+Agemonths,data=juv,family = 'binomial'))$coefficients[2,4])`), and no interaction with juvenile age (estimate = `r round2(summary(glm(SurvivedNext~cohortTL*Agemonths,data=juv,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glm(SurvivedNext~cohortTL*Agemonths,data=juv,family = 'binomial'))$coefficients[4,4])`)
+However, we found no effect of RTL on survival to adulthood (estimate = `r round2(summary(glm(SurvivedNext~cohortTL+Agemonths,data=juv,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glm(SurvivedNext~cohortTL+Agemonths,data=juv,family = 'binomial'))$coefficients[2,4])`), and no interaction with juvenile age (estimate = `r round2(summary(glm(SurvivedNext~cohortTL*Agemonths,data=juv,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glm(SurvivedNext~cohortTL*Agemonths,data=juv,family = 'binomial'))$coefficients[4,4])`).
 
-Cox regression also showed no effect of individual-level or cohort-level RTL on survival later in life (individual-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL+Fledged)))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL)))$coefficients[1,5])`;  cohort-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL+Fledged+frailty(LayYear))))$coefficients[2,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL+Fledged+frailty(LayYear))))$coefficients[2,6])`).
-
-The longitudinal data showed that the amount of telomere shortening experienced in early life had no effect on survival to adulthood (estimate = `r round2(summary(glm(SurvivedNext~TROC+Agemonths,data=subset(Loss,LayYear<20013),family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glm(SurvivedNext~TROC+Agemonths,data=subset(Loss,LayYear<2013),family = 'binomial'))$coefficients[2,4])`), or in later life (estimate `r round3(summary(coxph(with(subset(Loss,LayYear<2009),Surv(RemainingLife2+1,Died)~TROC+Agemonths)))$coefficients[1,1])`, P = `r round3(summary(coxph(with(subset(Loss,LayYear<2009),Surv(RemainingLife2+1,Died)~TROC+Agemonths)))$coefficients[1,5])`).
+Cox regression also showed no effect of individual-level or cohort-level RTL on survival later in life (individual-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL+Fledged)))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL)))$coefficients[1,5])`;  cohort-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL+Fledged+frailty(LayYear))))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL+Fledged+frailty(LayYear))))$coefficients[1,6])`).
+Using the longitudinal data, we found that the amount of telomere shortening experienced in early life had no effect on survival (estimate `r round3(summary(coxph(with(subset(Loss,LayYear<2009),Surv(RemainingLife2+1,Died)~TROC+Agemonths)))$coefficients[1,1])`, P = `r round3(summary(coxph(with(subset(Loss,LayYear<2009),Surv(RemainingLife2+1,Died)~TROC+Agemonths)))$coefficients[1,5])`).
 
 ###Discussion
-Here we use the long-term study of a closed population of Seychelles warblers to assess the relationships between early-life conditions, telomere length and survival. We find that while RTL varies
+Here we use the long-term study of a closed population of Seychelles warblers to assess the relationships between early-life conditions, telomere length and survival.
+We find RTL is subject to strong cohort effects
 

### 2016-03-01 Manuscript/SW_early_life_telomeres_SOM.Rmd
More edits (RTL)

--- 2016-01-20 Manuscript/SW_early_life_telomeres_SOM.Rmd
@@
 #Run the model
-lm1 <- lmer(LogTL ~ Fledged + Tarsus + Helper + NonHelper + Sex + Tarsus + BodyMass  + cenTQ  + Season + (1|LayYear),
+lm1 <- lmer(RTL ~ Agemonths * (Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ + Season) + (1|LayYear)+(1|TerritoryID),
             data = juv,
             REML = FALSE)
+lm1 <- standardize(lm1)
 
@@
 nams <- colnames(t1)
-#Sort out row/col names
-nams[nams == 'NonHelper'] <- 'Non helpers'
-nams[nams == 'SexMales'] <- 'Sex'
-nams[nams == 'cenTQ'] <- 'TQ'
-nams[nams == 'Helper'] <- 'Helpers'
-nams[nams == 'BodyMass'] <- 'Body mass'
-nams[nams == 'FledgedFledglings'] <- 'Age class'
-nams[nams == 'SeasonWinter'] <- 'Season'
+
+nams[nams == 'z.NonHelper'] <- 'Non helpers'
+nams[nams == 'c.Sex'] <- 'Sex'
+nams[nams == 'z.TQ'] <- 'TQ'
+nams[nams == 'c.Helper'] <- 'Helpers'
+nams[nams == 'z.BodyMass'] <- 'Body mass'
+nams[nams == 'z.Agemonths'] <- 'Age'
+nams[nams == 'z.Tarsus'] <- 'Tarsus'
+nams[nams == 'c.Season'] <- 'Season'
+nams[nams == 'c.Season:z.Agemonths'] <- 'Age x season'
+nams[nams == 'z.Agemonths:z.NonHelper'] <- 'Age x non helpers'
+nams[nams == 'c.Sex:z.Agemonths'] <- 'Age x sex'
+nams[nams == 'z.Agemonths:z.TQ'] <- 'Age x TQ'
+nams[nams == 'c.Helper:z.Agemonths'] <- 'Age x helpers'
+nams[nams == 'z.Agemonths:z.BodyMass'] <- 'Age x mass'
+nams[nams == 'z.Agemonths:z.Tarsus'] <- 'Age x tarsus'
 
@@
 
-tableS1 <- data.frame(Model = apply(t1,1,getterms,c(2:12)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
+tableS1 <- data.frame(Model = apply(t1,1,getterms,c(2:16)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
 colnames(tableS1) <- c('Model','df','AICc', 'Delta AICc','Weight')
@@
 
+**Table S2** Model selection table for analysis of factors affecting early life telomere loss in Seychelles warbler chicks, using longitudinal data.
+Only models that remained in the top model set for model averaging (AICc $\leq$ 6 compared to the top model) are included here.
+TQ = territory quality.
+
 ```{r Table S2}
+temp <- Loss
 
 #Run the model
-lm2 <- lmer(TROC ~ Fledged + Tarsus + Helper + NonHelper + Sex + Tarsus + BodyMass  + cenTQ  + Season + LogTL + (1|LayYear),
-            data = Loss,
+lm1 <- lmer(TROC ~ Agemonths * (Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ + Season) + (1|LayYear),
+            data = temp,
             REML = FALSE)
+lm1 <- standardize(lm1)
 
 #Model averaging
-
-lm3 <- dredge(lm2)
-lm3 <- subset(lm3,delta <=6)
-
-#Model averaging
-t1 <- as.matrix(subset(lm3, delta<=6))
+lm2 <- dredge(lm1)
+t1 <- as.matrix(subset(lm2, delta<=6))
 
@@
 
-nams[nams == 'NonHelper'] <- 'Non helpers'
-nams[nams == 'SexMales'] <- 'Sex'
-nams[nams == 'cenTQ'] <- 'TQ'
-nams[nams == 'Helper'] <- 'Helpers'
-nams[nams == 'BodyMass'] <- 'Body mass'
-nams[nams == 'FledgedFledglings'] <- 'Age class'
-nams[nams == 'SeasonWinter'] <- 'Season'
-nams[nams == 'LogTL'] <- 'Initial TL'
+nams[nams == 'z.RTL'] <- 'RTL'
+nams[nams == 'z.NonHelper'] <- 'Non helpers'
+nams[nams == 'c.Sex'] <- 'Sex'
+nams[nams == 'z.TQ'] <- 'TQ'
+nams[nams == 'c.Helper'] <- 'Helpers'
+nams[nams == 'z.BodyMass'] <- 'Body mass'
+nams[nams == 'z.Agemonths'] <- 'Age'
+nams[nams == 'z.Tarsus'] <- 'Tarsus'
+nams[nams == 'c.Season'] <- 'Season'
+nams[nams == 'c.Season:z.Agemonths'] <- 'Age x season'
+nams[nams == 'z.Agemonths:z.NonHelper'] <- 'Age x non helpers'
+nams[nams == 'c.Sex:z.Agemonths'] <- 'Age x sex'
+nams[nams == 'z.Agemonths:z.TQ'] <- 'Age x TQ'
+nams[nams == 'c.Helper:z.Agemonths'] <- 'Age x helpers'
+nams[nams == 'z.Agemonths:z.BodyMass'] <- 'Age x mass'
+nams[nams == 'z.Agemonths:z.Tarsus'] <- 'Age x tarsus'
 
@@
 
-tableS1 <- data.frame(Model = apply(t1,1,getterms,c(2:12)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
-colnames(tableS1) <- c('Model','df','AICc', 'Delta AICc','Weight')
-rownames(tableS1) <- NULL
-tableS1$Model[tableS1$Model == ''] <- 'Null model'
-tableS1[,3:5] <- lapply(tableS1[,3:5],as.numeric)
-kable(tableS1,digits=2)
+tableS2 <- data.frame(Model = apply(t1,1,getterms,c(2:16)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
+colnames(tableS2) <- c('Model','df','AICc', 'Delta AICc','Weight')
+rownames(tableS2) <- NULL
+tableS2$Model[tableS2$Model == ''] <- 'Null model'
+tableS2[,3:5] <- lapply(tableS2[,3:5],as.numeric)
+kable(tableS2,digits=2)
 

### 2016-03-01 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
More edits (RTL)

--- 2016-02-03 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
@@
 
-**Figure 2** Temporal variation in early-life telomere length in the Seychelles warbler based on cross-sectional data.
-Points and error bars represent mean and standard error of telomere length for all birds born in each summer breeding season.
-Sample sizes for each year are provided in brackets.
+**Figure 2** Factors affecting early-life telomere length in Seychelles warblers.
+**A** Model averaged estimates and 95% confidence intervals for explanatory terms used in a linear mixed model with telomere length (see main text for details; TQ = territory quality, TL = telomere length).
 
-**Figure 3** Factors affecting early-life telomere length in Seychelles warblers at the individual level, within seasons.
+**Figure 3** Factors affecting early-life telomere shortening rates in Seychelles warblers, within seasons.
 Model averaged estimates and 95% confidence intervals for explanatory terms used in a linear mixed model with telomere length (cross-sectional data, **A**) and telomere loss (longitudinal data, **B**) as the response variable.
@@
 
-**Figure 4** Telomere length in relation to **A** season and **B** tarsus length in juvenile Seychelles warblers.
-In *A**, white and grey boxes represent summer and winter seasons, respectively, while in **B** black and grey points represent nestlings and fledglings, respectively.
-Sample sizes for each group are provided in brackets.
-
-**Figure 5** Kaplan-Meier curves showing the relationship between survival and telomere length for juvenile Seychelles warblers.
+**Figure 4** Kaplan-Meier curves showing the relationship between survival and telomere length for juvenile Seychelles warblers.
 Telomere length is binned into groups here for visualisation purposes only (long and short = greater than or less than median telomere length, respectively).
@@
         legend.position = 'none')+
-  geom_boxplot(outlier.colour=NA)+
-  annotate('text',x = 0.6,y = 4.8,label = 'A')
+  geom_boxplot(outlier.colour=NA,alpha = 0.2)+
+  annotate('text',x = 0.6,y = 2.1,label = 'A')
 
-#
 temp <- Loss
-temp[,'Age (months)'] <- factor(temp$Agemonths)
+temp$Agemonths <- factor(temp$Agemonths)
 
-Fig1B <- ggplot(temp,aes(x = LogTL,y = LogTL1,col = `Age (months)`,shape = `Age (months)`)) +
-  ylab('Log adult telomere length (kb)') + 
-  xlab('Log juvenile telomere length (kb)') +
+Fig1B <- ggplot(temp,aes(x = Agemonths,y = TROC))+
+  geom_jitter(col='grey') +
+  ylab('Loss in RTL') + 
+  xlab('Age (months)') +
   theme_classic() +
@@
         legend.position = 'none')+
-  geom_point() +
-  stat_smooth(method = 'lm')+
-  xlim(2,4.5)+
-  ylim(2,4.5)+
-    scale_colour_manual(values = c('darkgrey','darkgreen','darkorange'))+
-  annotate('text',x = 2,y = 4.5,label = 'B')
+  geom_boxplot(outlier.colour=NA,alpha = 0.2)+
+  annotate('text',x = 0.6,y = 2.1,label = 'B')
 
-#
-temp <- Loss
-temp[,'Age (months)'] <- factor(temp$Agemonths)
-
-Fig1C <- ggplot(temp,aes(x = TimeDiff,y = Loss,col=`Age (months)`,shape = `Age (months)`)) +
-  ylab('Log telomere loss (kb)') + 
-  xlab('Time interval (years)') +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        legend.position = 'none')+
-  geom_point()+
-  geom_abline(slope = 0,intercept = 0,lty = 2)+
-  stat_smooth(method='lm')+
-  scale_colour_manual(values = c('darkgrey','darkgreen','darkorange'))+
-  annotate('text',x = 0,y = 2,label = 'C')
-
-Fig1D <- ggplot(temp,aes(x = TimeDiff,y = TROC,col=`Age (months)`,shape = `Age (months)`)) +
-  ylab('Log telomere loss rate (kb/year)') + 
-  xlab('Time interval (years)') +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        legend.position = c(0.9,0.8))+
-  geom_point()+
-  geom_abline(slope = 0,intercept = 0,lty = 2)+
-  scale_colour_manual(values = c('darkgrey','darkgreen','darkorange'))+
-  annotate('text',x = 0,y = 8,label = 'D')
-
-pdf('Figures/Figure 1.pdf',
-    width = 8,
-    height = 8,
-    paper = 'a4r')
-multiplot(Fig1A,Fig1B,Fig1C,Fig1D,layout=matrix(c(1:4),nrow = 2,byrow=T))
-dev.off()
-
-```
-
-```{r Figure 2}
-
-###############################
 temp <- juv
@@
 
-Fig2A <- ggplot(subset(temp,Agemonths == 1),
+Fig1C <- ggplot(temp,
        aes(x = LayYear,
-           y = LogTL)) +
-  ylab('') + 
+           y = RTL)) +
+  ylab('RTL') + 
   xlab('') +
@@
   geom_jitter(col = 'grey')+
-  ylim(c(2,4.5))+
-  geom_boxplot(outlier.colour = NA)+
+  geom_boxplot(outlier.colour = NA,alpha = 0.2)+
   scale_x_discrete(limits = factor(1998:2014),drop=F)+
-  annotate('text',x = 1,y = 4.4,label = 'A')
+  annotate('text',x = 1,y = 2.1,label = 'C')
 
-Fig2B <- ggplot(subset(juv,Agemonths == 6),
-       aes(x = factor(LayYear),
-           y = LogTL)) +
-  ylab('Log telomere length (kb)') + 
-  xlab('') +
-  theme_classic() +
-  theme(axis.text.x = element_text(angle = 60,size = 14,vjust=0.5), 
-        axis.text.y = element_text(size = 14), 
-        axis.title.y = element_text(size = 16,vjust=0.8),
-        axis.title.x = element_text(size = 16),
-        legend.position = 'none') +
-  geom_jitter(col = 'grey')+
-  ylim(c(2,4.5))+
-  geom_boxplot(outlier.colour = NA)+
-  scale_x_discrete(limits = factor(1998:2014),drop=F)+
-  annotate('text',x = 1,y = 4.4,label = 'B')
-
-Fig2C <- ggplot(subset(juv,Agemonths == 10),
-       aes(x = factor(LayYear),
-           y = LogTL)) +
-  ylab('') + 
-  xlab('Cohort') +
-  theme_classic() +
-  theme(axis.text.x = element_text(angle = 60,size = 14,vjust=0.5), 
-        axis.text.y = element_text(size = 14), 
-        axis.title.y = element_text(size = 16,vjust=0.8),
-        axis.title.x = element_text(size = 16),
-        legend.position = 'none') +
-  geom_jitter(col = 'grey')+
-  ylim(c(2,4.5))+
-  geom_boxplot(outlier.colour = NA)+
-  scale_x_discrete(limits = factor(1998:2014),drop=F)+
-  annotate('text',x = 1,y = 4.4,label = 'C')
-
-pdf('Figures/Figure 2.pdf',
+pdf('Figures/Figure 1.pdf',
     width = 6,
-    height = 8,
+    height = 6,
     paper = 'a4r')
-multiplot(Fig2A,Fig2B,Fig2C)
+multiplot(Fig1A,Fig1B,Fig1C,layout=matrix(c(1,2,3,3),nrow = 2,byrow=T))
 dev.off()
@@
 #####################
-#3A
+#2A
 #####################
@@
 #Run the model
-lm2 <- lmer(LogTL ~ Agemonths * (Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ)  + (1|LayYear) + (1|PlateID) + (1|TerritoryID),
+lm2 <- lmer(RTL ~ Agemonths * (Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ + Season)  + (1|LayYear) + (1|TerritoryID),
             data = temp,
@@
 nams[nams == 'z.Tarsus'] <- 'Tarsus'
+nams[nams == 'c.Season'] <- 'Season'
+nams[nams == 'c.Season:z.Agemonths'] <- 'Age x season'
 nams[nams == 'z.Agemonths:z.NonHelper'] <- 'Age x non helpers'
@@
 
-Fig3 <- forestplot(table2)+
-  ylim(-0.4,0.4)
+Fig2A <- forestplot(table2)+
+  annotate('text',x = 1,y = 0.25,label = 'A')
 
-pdf('Figures/Figure 3.pdf',
-    width = 8,
-    height = 5,
-    paper = 'a4r')
-Fig3
-dev.off()
-
-```
-
-```{r Figure 4}
 temp <- juv
@@
 
-Fig4A <- ggplot(temp,
+Fig2B <- ggplot(temp,
        aes(x = factor(Agemonths),
-           y = LogTL,
-           fill=`Non helpers`)) +
-  ylab('Log telomere length (kb)') + 
+           y = RTL,
+           fill=Season)) +
+  ylab('RTL') + 
   xlab('Age (months)') +
@@
         legend.position = 'top') +
-  geom_jitter(aes(colour = `Non helpers`),position=position_jitterdodge(dodge.width = 0.8))+
-  geom_boxplot(outlier.colour = NA,notch=F)+
+  geom_jitter(aes(colour = Season),position=position_jitterdodge(dodge.width = 0.8))+
+  geom_boxplot(outlier.colour = NA,notch=F,alpha=0.2)+
   scale_colour_manual(values = c('grey','darkorange'))+
   scale_fill_manual(values = c('grey','darkorange'))+
-  annotate('text',x = 0.55,y = 4.5,label = 'A')
+  annotate('text',x = 0.55,y = 2.1,label = 'B')
 
-Fig4B <- ggplot(temp,
-       aes(x = factor(Agemonths),
-           y = LogTL,
-           fill=`Helpers`)) +
-  ylab('Log telomere length (kb)') + 
-  xlab('Age (months)') +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        legend.position = 'top') +
-  geom_jitter(aes(colour = `Helpers`),position=position_jitterdodge(dodge.width = 0.8))+
-  geom_boxplot(outlier.colour = NA,notch=F)+
-  scale_colour_manual(values = c('grey','darkorange'))+
-  scale_fill_manual(values = c('grey','darkorange'))+
-  annotate('text',x = 0.55,y = 4.5,label = 'B')
-
-Fig4C <- ggplot(temp,
-       aes(x = TQ,
-           y = LogTL,
+Fig2C <- ggplot(temp,
+       aes(x = Tarsus,
+           y = RTL,
            colour=`Age (months)`,
@@
            shape = `Age (months)`)) +
-  ylab('Log telomere length (kb)') + 
-  xlab('Territory quality') +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        legend.position = 'top') +
-  geom_point(alpha = 0.2)+
-  stat_smooth(method = 'lm')+
-  scale_colour_manual(values = c(grey(0.2),'darkorange','darkgreen'))+
-  scale_fill_manual(values = c('darkgrey','darkorange','darkgreen'))+
-  annotate('text',x = 3,y = 4.5,label = 'C')
-
-Fig4D <- ggplot(temp,
-       aes(x = Tarsus,
-           y = LogTL,
-           colour=`Age (months)`,
-           fill = `Age (months)`,
-           shape = `Age (months)`)) +
-  ylab('Log telomere length (kb)') + 
+  ylab('RTL') + 
   xlab('Tarsus length (mm)') +
@@
   scale_fill_manual(values = c('darkgrey','darkorange','darkgreen'))+
-  annotate('text',x = 17,y = 4.5,label = 'D')
+  annotate('text',x = 17,y = 2.1,label = 'C')
 
-pdf('Figures/Figure 4.pdf',
-    width = 7,
-    height = 8,
+pdf('Figures/Figure 2.pdf',
+    width = 8,
+    height = 9,
     paper = 'a4r')
-multiplot(Fig4A,Fig4B,Fig4C,Fig4D,layout=matrix(c(1:4),byrow = TRUE,nrow = 2))
+multiplot(Fig2A,Fig2B,Fig2C,layout = matrix(c(1,1,2,3),nrow = 2,byrow = T))
 dev.off()
@@
 
-```{r Figure 5}
-temp <- subset(juv,LayYear<2009)
-mymed <- median(temp$cohortTL)
-temp$coTLF <- ifelse(temp$cohortTL>mymed,'h','l')
+```{r Figure 3}
 
-s1 <- survfit(Surv(time = Lifespan, event = Died)~coTLF, data = temp)
-Fig5A <- ggsurv(s1,plot.cens=F,lty.est=c(1:2)) + 
+temp <- Loss
+
+#####################
+#3A
+#####################
+
+#Run the model
+lm2 <- lmer(TROC ~ Agemonths * (Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ + Season) + (1|LayYear),
+            data = temp,
+            REML = FALSE)
+lm2 <- standardize(lm2)
+#Model averaging
+
+lm3 <- dredge(lm2)
+lm3 <- subset(lm3,delta <=6)
+lmavg <- model.avg(lm3,fit=T)
+myimp <- importance(lmavg) #Relative importance of each explanatory variable
+
+#Put it all into a nice table
+table2 <- imptable(lmavg,myimp)
+table2 <- signif(table2,2)
+
+#Sort out row/col names
+nams <- rownames(table2)
+nams[nams == 'z.NonHelper'] <- 'Non helpers'
+nams[nams == 'c.Sex'] <- 'Sex'
+nams[nams == 'z.TQ'] <- 'TQ'
+nams[nams == 'c.Helper'] <- 'Helpers'
+nams[nams == 'z.BodyMass'] <- 'Body mass'
+nams[nams == 'z.Agemonths'] <- 'Age'
+nams[nams == 'z.Tarsus'] <- 'Tarsus'
+nams[nams == 'c.Season'] <- 'Season'
+nams[nams == 'c.Season:z.Agemonths'] <- 'Age x season'
+nams[nams == 'z.Agemonths:z.NonHelper'] <- 'Age x non helpers'
+nams[nams == 'c.Sex:z.Agemonths'] <- 'Age x sex'
+nams[nams == 'z.Agemonths:z.TQ'] <- 'Age x TQ'
+nams[nams == 'c.Helper:z.Agemonths'] <- 'Age x helpers'
+nams[nams == 'z.Agemonths:z.BodyMass'] <- 'Age x mass'
+nams[nams == 'z.Agemonths:z.Tarsus'] <- 'Age x tarsus'
+
+rownames(table2) <- nams
+colnames(table2) <- c('Estimate','SE','2.5%','97.5%','RI')
+
+table2 <- data.frame(table2,myx=row.names(table2))
+table2$myx <- factor(table2$myx,levels = table2$myx[order(table2$Estimate,decreasing = T)])
+
+Fig3A <- forestplot(table2)+
+  annotate('text',x = 1,y = 0.45,label = 'A')
+
+temp <- Loss
+
+temp[,'Helpers'] <- ifelse(temp$Helper>0,'Present', 'Absent')
+temp[,'Non helpers'] <- ifelse(temp$NonHelper>0,'Present', 'Absent')
+temp[,'Age (months)'] <- factor(temp$Agemonths)
+
+Fig3B <- ggplot(temp,
+       aes(x = factor(Agemonths),
+           y = TROC,
+           fill=Season)) +
+  ylab('Loss in RTL') + 
+  xlab('Age (months)') +
   theme_classic() +
-  ylab('Proportion survived') + 
-  xlab('Age (years)') + 
-  xlim(c(0,15)) +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        legend.position = 'top') +
+  geom_jitter(aes(colour = Season),position=position_jitterdodge(dodge.width = 0.8))+
+  geom_boxplot(outlier.colour = NA,notch=F,alpha=0.2)+
+  scale_colour_manual(values = c('grey','darkorange'))+
+  scale_fill_manual(values = c('grey','darkorange'))+
+  annotate('text',x = 0.55,y = 2.1,label = 'B')
+
+Fig3C <- ggplot(temp,
+       aes(x = Tarsus,
+           y = TROC,
+           col=`Age (months)`)) +
+  ylab('Loss in RTL') + 
+  xlab('Tarsus length (mm)') +
   theme_classic() +
-  theme(axis.text.x = element_text(size = 14), 
-        axis.text.y = element_text(size = 14), 
-        axis.title.y = element_text(size = 16,vjust=0.8),
-        axis.title.x = element_text(size = 16),
-        legend.position = 'none') +
-  scale_colour_manual(name = 'TL',
-                      values = c('grey','black')) +
-  guides(colour=F)
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        legend.position = 'top') +
+  geom_point(alpha = 0.2)+
+  stat_smooth(method = 'lm')+
+    scale_colour_manual(values = c(grey(0.2),'darkorange','darkgreen'))+
+  scale_fill_manual(values = c('darkgrey','darkorange','darkgreen'))+
+  annotate('text',x = 20,y = 2.1,label = 'C')
 
-mymed <- median(temp$cenTL)
-temp$cenTLF <- ifelse(temp$cenTL>mymed,'h','l')
-s1 <- survfit(Surv(time = Lifespan, event = Died)~cenTLF, data = temp)
-Fig5B <- ggsurv(s1,plot.cens=F,lty.est=c(1,2)) + 
-  theme_classic() +
-  ylab('Proportion survived') + 
-  xlab('Age (years)') + 
-  xlim(c(0,15)) +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 14), 
-        axis.text.y = element_text(size = 14), 
-        axis.title.y = element_text(size = 16,vjust=0.8),
-        axis.title.x = element_text(size = 16),
-        legend.position = 'none') +
-  scale_colour_manual(name = 'TL',
-                      values = c('black','grey')) +
-  guides(colour=F)
-
-pdf('Figures/Figure 5.pdf',
-    width = 4,
-    height = 8,
-    paper = 'a4')
-multiplot(Fig5A,Fig5B)
+pdf('Figures/Figure 3.pdf',
+    width = 8,
+    height = 9,
+    paper = 'a4r')
+multiplot(Fig3A,Fig3B,Fig3C,layout = matrix(c(1,1,2,3),nrow = 2,byrow = T))
 dev.off()

### 2016-03-01 Scripts/SW juvenile telomeres IPC clean.R

@@ -49,7 +49,7 @@ dd$LeftTarsus <- NULL
 
 # Remove unwanted data/outliers ----------------------------------------------------
 
-dd <- subset(dd,RTL<3)
+dd <- subset(dd,RTL<2)
 dd <- subset(dd,RTL > 0.04)
 dd <- subset(dd,BodyMass>5)
 dd <- subset(dd,Tarsus>17)
@@ -74,7 +74,7 @@ dd$Agemonths <- ifelse(dd$Ageclass == 'CH',1,
                        ifelse(dd$Ageclass == 'FL',6,
                               ifelse(dd$Ageclass == 'SA',10,dd$Age*12)))
 
-dd <- subset(dd,Agemonths>0)
+dd <- subset(dd,Agemonths!=12)
 
 dd$Age[dd$Fledged != 'Adults'] <- 1
 
@@ -90,7 +90,7 @@ dd$Died <- ifelse(dd$DeathYear<2013,1,0)
 # Exclude winter seasons and early years ----------------------------------
 
 dd <- subset(dd,LayYear>1997)
-dd <- subset(dd,Season == 'Summer')
+#dd <- subset(dd,Season == 'Summer')
 
 
 # Sex ---------------------------------------------------------------------
@@ -244,8 +244,8 @@ x4 <- x4[order(x4$BirdID),]
 
 
 #Apply Verhulst's correction for regresison to the mean
-xx1 <- x4$RTL-mean(x4$RTL)
-xx2 <- x2$RTL-mean(x2$RTL)          
+xx1 <- x4$RTL
+xx2 <- x2$RTL        
 rho <- cor(x4$RTL,x2$RTL)
 
 #Creat data frame with TROC
@@ -253,17 +253,17 @@ Loss <- data.frame(x4,
                    Loss = x4$RTL-x2$RTL,
                    RTL1 = x2$RTL,
                    Agemonths1 = x2$Agemonths,
-                   TimeDiff = round(as.numeric(x2$CatchDate-x4$CatchDate)/365,0),
+                   TimeDiff = round(as.numeric(x2$CatchDate-x4$CatchDate)/182.5)*0.5,
                    RemainingLife2 = x2$RemainingLife,
                    D=(rho*xx1)-xx2)
-Loss$TROC <- with(Loss,D/TimeDiff)
+Loss$TROC <- with(Loss,Loss/TimeDiff)
 
[…]

### 2016-03-01 Scripts/SW juvenile telomeres IPC functions.r

@@ -292,7 +292,7 @@ forestplot <- function(mytable)
           axis.title.y = element_text(size = 16,vjust=0.8),
           axis.title.x = element_text(size = 16)) +
     geom_abline(slope=0,intercept=0,lty=2) +
-    annotate('text',x=mytable$myx,y=min(mytable$X2.5)*1.2,label=paste('(',mytable$RI,')',sep=''))
+    annotate('text',x=mytable$myx,y=min(mytable$X2.5)*1.2,label=paste('(',mytable$RI,')',sep=''),size = 4)
     
 }
 

### 2016-03-16 Manuscript/SW_early_life_telomeres.Rmd
Discussion and fogure edits

@@
 Telomeres, the protective caps at the ends of chromosomes, shorten in response to oxidative stress, and telomere shortening is correlated with reduced survival.
-Thus, telomere dynamics may help us quantify individual variation in early-life costs, and enhance our understanding of how poor conditions in early life are related to later-life survival.
-We tested how telomere dynamics are related to spatiotemporal variation in early-life conditions and later-life survival in the Seychelles warbler (*Acrocephalus sechellensis*), across multiple cohorts spanning 14 years.
-we found that, in accordance with other studies, telomere length and loss are greatest in early life.
-We then show that juvenile telomere lengths varies markedly among cohorts, with average telomere length varying by .
-We found no evidence that early-life social environment (number of helpers, group size) or ecological conditions (territory quality) were related to telomere length, although we found tentative evidence that telomere length varied among summer and winter breeding seasons.
-Finally, we found that increased survival later in life was associated with longer telomeres in early life, but this effect was at the cohort, rather than individual, level.
-Our results highlight the inmportance of cohort effects in studies of telomere length.
+Thus, it is possible that telomere dynamics may help us quantify individual variation in early-life costs, and enhance our understanding of how poor conditions in early life are related to later-life survival.
+We tested how telomere dynamics are related to spatio-temporal variation in early-life conditions and later-life survival in the Seychelles warbler (*Acrocephalus sechellensis*), using a large dataset of `r nrow(juv)` juveniles from multiple cohorts spanning `r max(juv$LayYear)-min(juv$LayYear)` years.
+Despite cross-sectional data showing that telomere length decreases with age, and most that telomere length decreases most rapidly early in life, using longitudinal data we found that significant reductions in telomere length can only be detected within the first few months of life - after this, longitudinal changes in telomere length were not significantly different from zero.
+Juvenile telomere length varied markedly among cohorts, and this cohort variation was related to the season (birds born in winter seasons had longer telomeres), and to population density (birds bird in years with higher populaiton density had longer telomered).
+Within cohorts, we found no evidence that the early-life social environment or ecological conditions were related to telomere length or rates of shortening.
+Finally, we found no relationship between early-life and survival to adulthood.
+Our data suggest that telomeres are subject to strong population-level variation, but suggest that the resolution of telomeres as a useful biomarker of individual-level cost is likely to be age-dependent, and in many cases limited.
 
@@
 While it is clear that silver spoon effects can occur, we have little understanding of how and under what conditions early-life environments will affect adult fitness.
-However , we do expect the later-life consequences of a good or poor start to vary among individuals, populations and species because the phenotypic consequences of an adverse environment can depend on an individual's initial condition, and genetic or epigenetic makeup [@Hoffman2000; @Richards2006].
+We expect the later-life consequences of a good or poor start to vary among individuals, populations and species because the phenotypic consequences of an adverse environment can depend on an individual's initial condition, and genetic or epigenetic makeup [@Hoffman2000; @Richards2006].
 Moreover, it is not always possible to fully quantify what constitutes a good or bad environment, and any 'hidden' environmental variation may obscure relationships between the early-life conditions that are measured and the resulting adult phenotypes.
@@
 This association between senescence and telomere length has inspired a great deal of recent research into telomere evolutionary ecology [reviewed in @Horn2010; @Haussmann2010; @Monaghan2014].
-While there is little direct evidence that the relationship between telomere dynamics and survival is causal [@Barrett2011; @Simons2015], there is now excellent evidence that telomeres can act as biomarkers of cost in wild populations, providing a signature of the ecological stress that has been experienced and is otherwise difficult to detect [@Monaghan2014; @Schultner2014; @Asghar2015].
+While there is little direct evidence that the relationship between telomere dynamics and survival is causal [@Simons2015], there is mounting evidence that telomeres can act as biomarkers of cost in wild populations, providing a signature of the ecological stress that has been experienced and is otherwise difficult to detect [@Monaghan2014; @Schultner2014; @Asghar2015].
 
-There is evidence from a range of taxa that the greatest rate of telomere loss occurs in early life [e.g.
-@Frenck1998; @Haussmann2003], and that the extent of this telomere shortening is influenced by the conditions experienced during that period [@Price2013; @Monaghan2014; @Nettle2015; @Reichert2015].
-Importantly, early-life telomere dynamics have been associated with both short-term and late-life survival [@Heidinger2012; @Boonekamp2014], and with other parameters such as cognition [@Nettle2015a].
+There is evidence from a range of taxa that the greatest rate of telomere shortening occurs in early life [e.g.
+@Frenck1998; @Haussmann2003], and that the extent of this shortening is influenced by the conditions experienced during that period [@Price2013; @Monaghan2014; @Nettle2015; @Reichert2015].
+Early-life telomere dynamics have been associated with both short-term and late-life survival [@Heidinger2012; @Boonekamp2014], and with other parameters such as cognition [@Nettle2015a].
 However, few studies have simultaneously analysed how telomere dynamics, early-life conditions and late-life survival are all related in a natural setting.
@@
 
-In this study we examine how telomeres link early-life environmental variation to late-life fitness in the Seychelles warbler.
+In this study we examine how telomeres are related to early-life environmental variation and short and long-term survival in the Seychelles warbler.
 Because telomere dynamics are expected to reflect individual-level variation in the costs of early-life experiences, they may allow a more sensitive analysis of the effects of early-life environmental variation than would be possible with a direct comparison of how survival is affected by the early-life environment.
@@
 Sex was determined using the PCR-based method outlined by Griffiths *et al.* [-@Griffiths1998].
-Prior to telomere analysis, DNA concentration and purity were quantified using a NanoDrop 8000 Spectrophotometer (ThermoScientific), and the following thesholds were applied before samples were included for further analysis: i) DNA concentration must be at least 15 ng $\mu$l^-1^ (based on a mean of three measurements), ii) the 260/280 ratio has to be between 1.8 and 2 and, iii) the 260/230 ratio should be higher than 1.8.
+Prior to telomere analysis, DNA concentration and purity were quantified using a NanoDrop 8000 Spectrophotometer (ThermoScientific), and the following thresholds were applied before samples were included for further analysis: i) DNA concentration must be at least 15 ng $\mu$l^-1^ (based on a mean of three measurements), ii) the 260/280 ratio has to be between 1.8 and 2 and, iii) the 260/230 ratio should be higher than 1.8.
 DNA integrity was further validated by visualization with ethidium bromide after electrophoresis on a 1.2% agarose gel, and all samples with evidence of DNA degradation were re-extracted or excluded.
@@
 All DNA extractions that passed the above criteria were diluted to 3.3 ng $\mu$l^-1^ before telomere measurement.
-We measured telomere length for all samples using a quantitative PCR (qPCR) assay of telomeres and a GPADH control gene, using the molecular methods outlined by Barrett *et al.* [-@Barrett2012], with one amendment.
+We measured telomere length for all samples using a quantitative PCR (qPCR) assay of telomeres and a GAPDH control gene, using the molecular methods outlined by Barrett *et al.* [-@Barrett2012], with one amendment.
 A change in batch of SYBR green forced us to raise the annealing temperature of the telomere reaction from 58^o^C to 61^o^C for the majority of samples.
@@
 We then calculated relative telomere length (RTL) for each sample using equation 1 in Pfaffi *et al* [@Pfaffi2001].
-We chose to use RTL rather than continuing with the previously used method for claculating absolute telomere length, as i) using RTL enabled us to run more samples per plate (as an oligo standard is not required), ii) the RTL method was less susceptible to batch effects (Appendix 1), and iii) very few other studies have calculated absolute telomere length, and our experience suggests that cross-species comparisons are unlikely to be reliable.
-
+We chose to use RTL rather than continuing with the previously used method for calculating absolute telomere length [@Barrett2012], as i) using RTL enabled us to run more samples per plate (as an oligo standard is not required), ii) the RTL method was less susceptible to batch effects (Appendix 1), and iii) very few other studies have calculated absolute telomere length, and our experience suggests that cross-species comparisons are unlikely to be reliable.
 Inter-plate repeatability of final telomere lengths was assessed using the R package rptR [@Schielzeth2011], and for all subsequent analyses we used mean telomere length per sample where we had repeats.
 
-Telomere lengths were measured using a total of `r nrow(juv)+nrow(Loss)` samples, .
-Of these, `r nrow(juv)` were taken cross-sectionally from birds caught within their first year of life, between `r min(as.numeric(paste(juv$LayYear)))` and `r max(as.numeric(paste(juv$LayYear)))`.
-For a subset of first-year birds (n = `r nrow(Loss)` individuals) we had longitudinal data, with an additional sample taken as an adult.
-For these individuals we calculated the rate of telomere loss by subtracting adult RTL from early-life RTL, and dividing this difference by the length of time (to the nearest 6 months) between sampling events.
+For a subset of first-year birds for which we had longitudinal data, with an additional sample taken as an adult.
+For these individuals we calculated the within-individual change in RTL by subtracting adult RTL from early-life RTL (hereafter $\Delta$RTL).
+Negative values of $\Delta$RTL reflect telomere shortening, while positive values reflect telomere lengthening.
 
@@
 We performed all analyses using R version 3.0.1 [@RDevelopmentCoreTeam2011].
-Telomere length was log transfomred to fit assumptions of normality.
-We first explored how telomere length varied within the first year of life and over time using linear regression (telomere length vs age in months), and one-way ANOVA (telomere length vs cohort).
-With the longitudinal data, we tested how telomeres shorten with age by testing how telomere loss and rate of loss were related to the time interval between sampling events, using linear models.
-If telomeres are lost at a constant rate from early age, we expect the time between sampling events to be positively and linearly relaited to telomere loss no, and unrelated to rate of loss.
-If, however, telomeres are lost at a greater rate early in life we expect a non-linear or negative relationship between telomere loss and time between sampling events, and a significant (linear or non-linear) decrease in telomere rate of loss with time interval.
+Because we are using a different measure of telomere length to the previous study of adult telomere dynamics in this species [@Barrett2013], as well as including many additional samples, we first re-tested, using all available samples, whether telomere length decreases with age and $\Delta$age [a longitudinal measure based on within-subsject centring; @VandePol2009] in the Seychelles warbler, using linear mixed models, with individual ID as a random effect.
+For the analysis of $\Delta$age we we only included birds who had been sampled at least twice.
+We then explored how telomere length varied within the first year of life (telomere length vs age in months) and over time (telomere length vs cohort ID).
+For this analysis we had bvery few birds with multiple samples, so we used linear models including only the first sample for each individual.
 
-We then used a general linear mixed model approach to explore how spatial variation in early-life environmental and social conditions influenced telomere length within cohorts.
+To assess our resolution for detecting telomere shortening, we tested, for each age class, whether $\Delta$RTL was significantly lower than zero, using one-sample t-tests.
+We then tested how $\Delta$RTL was related to the time interval between sampling events, using linear models.
+A significantly negative slope indicates that telomere shortening occurs linearly over time, while the time point at which $\Delta$RTL is significantly lower than zero can be considered the time period over which we have the resolution to detect reductions in telomere length.
+
+We used general linear mixed models with model averaging to explore how spatial variation in early-life environmental and social conditions influenced telomere length within cohorts.
 Model averaging was carried out using the MuMIn package (version 1.10.5) in R [@Barton2012].
 We first created a full model containing the following explanatory variables: age (in months), tarsus length, sex, territory quality, season (summer or winter) and the number of helping and non-helping subordinate birds present in the natal territory.
-We also included interaction terms between age and all the other variables as telomere dynamics are epxected to vary within the first year of life [@Heidinger2012].
+We also included interaction terms between age and all the other variables as telomere dynamics are expected to vary within the first year of life [@Heidinger2012].
 As random effects we included cohort, territory ID and qPCR plate ID.
@@
 
-We used Cox regression, implemented in the survival package in R [@Therneau2011], to test whether survival was related to individual-level telomere length and rate of loss.
+We used generalised linear mixed models to test whether telomere length/loss were related to survival to adulthood.
+Models were set up with a binomial error structure, and included RTL/$\Delta$RTL and juvenile age as explanatory variable, along with the same random effects as for the model averaging (see above).
+We then used Cox regression, implemented in the survival package in R [@Therneau2011], to test whether long-term survival was related to individual-level telomere dynamics.
 For the cross-sectional data we used lifespan (in years) as survival time, while for the longitudinal data we used remaining lifespan from the second sampling event.
-We ran these models with cohort as a frailty term to exclude cohort effects.
+We ran these models with cohort as a frailty term to exclude cohort effects, and only included cohorts in which at least 70 percent of individuals had died (all years up to and including 2007).
 
 ###Results
+We measured telomere lengths using a total of `r nrow(dd)` unique samples from juvenile and adult Seychelles warblers.
+Of these, `r nrow(juv)` samples were taken cross-sectionally from birds caught within their first year of life, and of these juvenile birds, we had additional longitudinal samples for `r nrow(Loss)`.
+Efficiencies (mean $\pm$) for our telomere and GAPDH reactions were (`r round2(mean(dd$EffTelomere))` $\pm$ `r round2(sd(dd$EffTelomere))`) and (`r round2(mean(dd$EffGAPDH))` $\pm$ `r round2(sd(dd$EffGAPDH))`) respectively.
+Inter-plate repeatability of RTL, based on 392 samples measured at least twice, was 0.78 (CI = 0.73-0.82).
 
-#####Early-life telomere dynamics
-Efficiecies (mean $\pm$) for our telomere and GAPDH reactions were (`r round2(mean(dd$EffTelomere))` $\pm$ `r round2(sd(dd$EffTelomere))`) and (`r round2(mean(dd$EffGAPDH))` $\pm$ `r round2(sd(dd$EffGAPDH))`) respectively.
-Inter-plate repeatability of RTL, based on XX samples measured at least twice, was 0.78 (CI = 0.73-0.82).
+#####Early-life RTL
+Considering the entire Seychelles warbler lifespan, RTL decreased with age (t = `r round2(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,4])`; P `r round3(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,5])`; Fig.
+1A), and with $\Delta$age (t = `r round2(summary(lme(RTL~DeltaAge,data=subset(dd,DeltaAge!=0),random=~1|BirdID))$tTable[2,4])`; P `r round3(summary(lme(RTL~DeltaAge,data=subset(dd,DeltaAge!=0),random=~1|BirdID))$tTable[2,5])`; Fig.
+1B).
+RTL also decreased with age within the first year of life (i.e.
+excluding adult samples; R^2^ = `r round2(summary(lm(RTL~Agemonths,data=juv))$r.squared)`; t = `r round2(summary(lm(RTL~Agemonths,data=juv))$coefficients[2,3])`; P `r round3(summary(lm(RTL~Agemonths,data=juv))$coefficients[2,4])`).
+Early-life RTL varied significantly among cohorts (one-way ANOVA, telomere length: F = `r round2(anova(lm(RTL~factor(FieldPeriodID),data = juv))[1,4])`; P `r round3(anova(lm(RTL~factor(FieldPeriodID),data = juv))[1,5])`), and cohort-level variation in juvenile RTL (taken as the median for each cohort) was positively related to adult population density (R^2^ = `r round2(summary(lm(RTL~Density,data=juvseason,na.action=na.exclude))$r.squared)`; t = `r round2(summary(lm(RTL~Density,data = juvseason,na.action=na.exclude))$coefficients[2,3])`; P = `r round2(anova(lm(RTL~Density,data = juvseason,na.action=na.exclude))[1,5])`; Fig.
+1C).
+Variation in cohort-level RTL was not related temporal variation in  territory quality (linear regression, R^2^ = `r round2(summary(lm(RTL~TQ,data=juvseason,na.action=na.exclude))$r.squared)`; t = `r round2(summary(lm(RTL~TQ,data = juvseason,na.action=na.exclude))$coefficients[2,3])`; P = `r round2(anova(lm(RTL~TQ,data = juvseason,na.action=na.exclude))[1,5])`), or island-wide food availability (R^2^ = `r round2(summary(lm(RTL~Insect,data=juvseason,na.action=na.exclude))$r.squared)`; t = `r round2(summary(lm(RTL~Insect,data = juvseason,na.action=na.exclude))$coefficients[2,3])`; P = `r round2(anova(lm(RTL~Insect,data = juvseason,na.action=na.exclude))[1,5])`).
 
-RTL dereased with age within the first year of life (R^2^ = `r round2(summary(lm(RTL~Agemonths,data=juv))$r.squared)`; F = `r round2(anova(lm(RTL~Agemonths,data=juv))[1,4])`; P `r round3(anova(lm(RTL~Agemonths,data=juv))[1,5])`), and across the entire Seychelles warbler lifespan (R^2^ = `r round2(summary(lm(RTL~Agemonths,data=dd))$r.squared)`; F = `r round2(anova(lm(RTL~Agemonths,data=dd))[1,4])`; P `r round3(anova(lm(RTL~Agemonths,data=dd))[1,5])`; Fig.
-1A).
-Longitudinal data showed that the rate of early-life telomere shortening also decreased with age (R^2^ = `r round2(summary(lm(TROC~Agemonths,data=Loss))$r.squared)`; F = `r round2(anova(lm(TROC~Agemonths,data=Loss))[1,4])`; P = `r round3(anova(lm(TROC~Agemonths,data=Loss))[1,5])`), suggesting that the highest rates of telomere loss occurred following the nestling stage (Fig.
-1B).
-
-Both early-life RTL and rate of loss varied significantly among cohorts (one-way ANOVA, telomere length: F = `r round2(anova(lm(RTL~factor(FieldPeriodID),data = juv))[1,4])`; P `r round3(anova(lm(RTL~factor(FieldPeriodID),data = juv))[1,5])`; Fig, 1C; telomere loss: F = `r round2(anova(lm(TROC~factor(FieldPeriodID),data = Loss))[1,4])`; P `r round3(anova(lm(TROC~factor(FieldPeriodID),data = Loss))[1,5])`).
-Variation in median RTL over breeding seasons in fledglings and subadults was not related temporal variation in  territory quality (linear regression, R^2^ = `r round2(summary(lm(RTL~TQ,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(RTL~TQ,data = juvseason,na.action=na.exclude))[1,4])`; P = `r round2(anova(lm(RTL~TQ,data = juvseason,na.action=na.exclude))[1,5])`), island-wide food availability (R^2^ = `r round2(summary(lm(RTL~Insect,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(RTL~Insect,data = juvseason,na.action=na.exclude))[1,4])`; P = `r round2(anova(lm(RTL~Insect,data = juvseason,na.action=na.exclude))[1,5])`) or population density (R^2^ = `r round2(summary(lm(RTL~Density,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(RTL~Density,data = juvseason,na.action=na.exclude))[1,4])`; P = `r round2(anova(lm(RTL~Density,data = juvseason,na.action=na.exclude))[1,5])`).
-
-#####Early-life environment and early-life telomere dynamics
-The top model explaining variation in early-life RTL contained age, season (summer vs winter), tarsus length, and the interaction between age and tarsus length (Table S1).
-The top model was much better supported than a null model ($\Delta$AICc = `r round2(AICc(lmer(RTL~1+(1|LayYear)+(1|TerritoryID),data=juv,REML = FALSE))-AICc(lmer(RTL~Agemonths+Season+Tarsus+Agemonths*Tarsus+(1|LayYear)+(1|TerritoryID),data=juv,REML = FALSE)))`), and all three effects were statistically significant (Fig.
+The top model explaining variation in early-life RTL contained age, sex, season (summer vs winter), tarsus length, and the interaction between age and tarsus length (Table S1).
+The top model was much better supported than a null model ($\Delta$AICc = `r round2(AICc(lmer(RTL~1+(1|LayYear)+(1|TerritoryID),data=juv,REML = FALSE))-AICc(lmer(RTL~Agemonths+Sex+Season+Tarsus+Agemonths*Tarsus+(1|LayYear)+(1|TerritoryID),data=juv,REML = FALSE)))`), and with the exception of the sex and tarsus effects, model-averaged confidence intervals did not overlap zero (Fig.
 2A).
@@
 2C).
+Males had slightly longer telomeres than females, prarticularly at the fledgling and subadult stages (Fig.
+2D), although this effect was weak.
 
-Model averaging results from the longitudinal data are shown in Figure 3A.
-The top model explaining telomere loss contained age and season (Table S2).
-This model was a significantly better fit than the null model ($\Delta$AICc = `r round2(AICc(lmer(TROC~1+(1|LayYear),data=Loss,REML = FALSE))-AICc(lmer(TROC~Agemonths+Season+(1|LayYear),data=Loss,REML = FALSE)))`).
-Juveniles born in winter seasons had higher rates of telomere shortening compared to birds born in summer seasons, but this effect was weak (Fig.
-3B), although this effect was weak and confidence intervals overlapped zero (Fig.
+#####Early-life telomere shortening
+Longitudinal data showed that early-life RTL was positively, but weakly, related to adult RTL (R^2^ = `r round3(summary(lm(RTL1~RTL+Agemonths,data=Loss))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data=Loss))$coefficients[2,3])`; P = `r round3(summary(lm(RTL1~RTL+Agemonths,data = Loss))$coefficients[2,4])`; Fig.
 3A).
-There was no evidence that tarsus length was related to telomere shortening in the same way that telomere length was (Fig.
+$\Delta$RTL was significantly lower than zero in nestlings (t = `r round2(t.test(subset(Loss,Agemonths == 1)$Loss)$statistic)`; P `r round3(t.test(subset(Loss,Agemonths == 1)$Loss)$p.value)`), but not in fledglings (t = `r round2(t.test(subset(Loss,Agemonths == 6)$Loss)$statistic)`; P = `r round2(t.test(subset(Loss,Agemonths == 6)$Loss)$p.value)`) or subadults (t = `r round2(t.test(subset(Loss,Agemonths == 10)$Loss)$statistic)`; P = `r round2(t.test(subset(Loss,Agemonths == 10)$Loss)$p.value)`).
+In line with this, $\Delta$RTL increased with age in early life (R^2^ = `r round2(summary(lm(Loss~Agemonths,data=Loss))$r.squared)`; t = `r round2(summary(lm(Loss~Agemonths,data=Loss))$coefficients[2,3])`; P `r round3(summary(lm(Loss~Agemonths,data = Loss))$coefficients[2,4])`), suggesting that telomere shortening was highest shortly after the nestling phase.
+In nestlings, $\Delta$RTL  did not vary with the time between sampling events (R^2^ `r round3(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths ==1)))$r.squared)`; t = `r round2(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths ==1)))$coefficients[2,3])`; P = `r round3(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths ==1)))$coefficients[2,4])`).
+However, for fledglings and subadults $\Delta$RTL decreased with the length of time between sampling events (R^2^ = `r round2(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths !=1)))$r.squared)`; t = `r round2(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths !=1)))$coefficients[2,3])`; P = `r round3(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths !=1)))$coefficients[2,4])`).
+In flesdglings and subadults, $\Delta$RTL did not decrease below zero until ~3 years between sampling events was allowed, and confidence limits overlapped zero regardless of the amount of time between sampling events (Fig 3B).
+Together, these results suggest that the majority of observable telomere shortening in Seychelles warblers occurs in the first six months of life.
+We therefore restricted the remaining analyses of $\Delta$RTL to birds sampled as nestlings.
+
+$\Delta$RTL did not differ significantly among cohorts (F = `r round2(anova(lm(Loss~factor(LayYear),data = subset(Loss,Agemonths == 1)))[1,4])`; P `r round3(anova(lm(Loss~factor(LayYear),data = subset(Loss,Agemonths == 1)))[1,5])`), although power for this analysis was limited as longitudinal sample sizes within cohorts was low (Fig.
 3C).
+Model selection revealed that the top model explaining $\Delta$RTL was the null model (Table S2), and confidence limits for all explanatory terms overlapped zero (Fig.
+4A).
 
@@
 
-Telomere length in early life did not affect survival to adulthood (estimate = `r round2(summary(glm(SurvivedNext~RTL+Agemonths,data=juv,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glm(SurvivedNext~RTL+Agemonths,data=juv,family = 'binomial'))$coefficients[2,4])`), and there was no interaction effect between RTL and juvenile age on survival to adulthood (estimate = `r round2(summary(glm(SurvivedNext~RTL*Agemonths,data=juv,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glm(SurvivedNext~RTL*Agemonths,data=juv,family = 'binomial'))$coefficients[4,4])`).
-To separate out cohort-level RTL affected survival we calculated cohort-level RTL (i.e.
-the mean for each cohort) and added this term to a logistic regression.
-However, we found no effect of RTL on survival to adulthood (estimate = `r round2(summary(glm(SurvivedNext~cohortTL+Agemonths,data=juv,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glm(SurvivedNext~cohortTL+Agemonths,data=juv,family = 'binomial'))$coefficients[2,4])`), and no interaction with juvenile age (estimate = `r round2(summary(glm(SurvivedNext~cohortTL*Agemonths,data=juv,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glm(SurvivedNext~cohortTL*Agemonths,data=juv,family = 'binomial'))$coefficients[4,4])`).
+Early-life RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~RTL+(1|TerritoryID)+(1|LayYear),data=juv13,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL+Agemonths+Tarsus+(1|TerritoryID)+(1|LayYear),data=juv13,family = 'binomial'))$coefficients[2,4])`), with no interaction effect between RTL and juvenile age on survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|LayYear),data=juv13,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|LayYear),data=juv13,family = 'binomial'))$coefficients[4,4])`).
+To test whether cohort-level effects on RTL influenced survival we calculated cohort-level RTL (i.e.
+the median RTL for each cohort) and added this term to a GLMM, and found that cofort-level RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~cohortTL+Agemonths+(1|TerritoryID)+(1|LayYear),data=juv13,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~cohortTL+Agemonths+(1|TerritoryID)+(1|LayYear),data=juv13,family = 'binomial'))$coefficients[2,4])`), with no interaction with juvenile age (estimate = `r round2(summary(glmer(SurvivedNext~cohortTL*Agemonths+(1|TerritoryID)+(1|LayYear),data=juv13,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~cohortTL*Agemonths+(1|TerritoryID)+(1|LayYear),data=juv13,family = 'binomial'))$coefficients[4,4])`).
+Using the longitudinal data, we found that $\Delta$RTL had no effect on survival to the year after the second sampling event (estimate = `r round2(summary(glm(SurvivedNext2~Loss,data=subset(subset(Loss,Agemonths == 1),LayYear<2013),family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glm(SurvivedNext2~RTL+Agemonths,data=subset(subset(Loss,Agemonths == 1),LayYear<2013),family = 'binomial'))$coefficients[2,4])`).
 
-Cox regression also showed no effect of individual-level or cohort-level RTL on survival later in life (individual-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL+Fledged)))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL)))$coefficients[1,5])`;  cohort-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL+Fledged+frailty(LayYear))))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL+Fledged+frailty(LayYear))))$coefficients[1,6])`).
-Using the longitudinal data, we found that the amount of telomere shortening experienced in early life had no effect on survival (estimate `r round3(summary(coxph(with(subset(Loss,LayYear<2009),Surv(RemainingLife2+1,Died)~TROC+Agemonths)))$coefficients[1,1])`, P = `r round3(summary(coxph(with(subset(Loss,LayYear<2009),Surv(RemainingLife2+1,Died)~TROC+Agemonths)))$coefficients[1,5])`).
+Cox regression showed no effect of individual-level or cohort-level RTL on survival later in life (individual-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL+Agemonths+frailty(LayYear))))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL+Agemonths+frailty(LayYear))))$coefficients[1,6])`; cohort-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL+Agemonths+frailty(LayYear))))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL+Agemonths+frailty(LayYear))))$coefficients[1,6])`), and no interaction between RTL and age on survival (individual-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL*Agemonths+frailty(LayYear))))$coefficients[4,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL*Agemonths+frailty(LayYear))))$coefficients[4,6])`; cohort-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL*Agemonths+frailty(LayYear))))$coefficients[4,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL*Agemonths+frailty(LayYear))))$coefficients[4,6])`).
+$\Delta$RTL was not related to longer-term survival (estimate `r round3(summary(coxph(with(subset(subset(Loss,LayYear<2008),Agemonths == 1),Surv(RemainingLife2+1,Died)~Loss)))$coefficients[1,1])`, P = `r round3(summary(coxph(with(subset(subset(Loss,LayYear<2008),Agemonths ==1 ),Surv(RemainingLife2+1,Died)~Loss)))$coefficients[1,5])`).
 
 ###Discussion
-Here we use the long-term study of a closed population of Seychelles warblers to assess the relationships between early-life conditions, telomere length and survival.
-We find RTL is subject to strong cohort effects
+Here we use one of the largest datasets to date to assess the relationships between early-life conditions, telomere length and survival in a closed population of Seychelles warblers.
+We found that, as in other studies, telomeres are lost at the greatest rate very early in life.
+We found that telomere dynamics are subject to strong seasonal and cohort effects, but that temporal variation in RTL is not related to any of the environmental variables tested.
+Within seasons, we found no evidence that the ecological or social environment influenced telomere dynamics, and we found no evidence that early-life telomere length is related to juvenile or later-life survival.
 
-The clearest result from our study is that RTL varies among cohorts.
-Very few studies have shown that temporal variation in telomere dynamics occurs in natural populations, and to our knowledge the studies that have done so were limited to just two seasons [@Mizutani2013; @Watson2015].
-The long-term Seychelles warbler dataset has allowed us to show that temporal variation in telomere dynamics does occur at the population level over longer time periods.
+Our study adds the the growing number showing that telomere length decreases with age, and that this decrease is most rapid in early life [e.g.
+@Frenck1998; @Haussmann2003; @Heidinger2012].
+Our finding that tarsus length was related to juvenile RTL in an age-dependent manner (Fig.
+2) most likely an additional age effect - in passerine birds tarsus length is correlated with age during the nestling stage, but not after fledging [@Ricklefs1976].
+Telomere loss is most rapid early in life due to ongoing growth [@Frenck1998], and a negative correlation between RTL and body size in early life is therefore expected.
+However, despite our clear finding that population-level RTL decreases with age throughout the warbler lifespan (Fig.
+1), we found that individual-level telomere shortening can only be detected  very early in life (Fig.
+3).
+This is in contrast to other avian studies where birds were reared in laboratory [@Heidinger2012] or nestbox [@Boonekamp2014] conditions, in which individual-level telomere shortening clearly occurs throughout the juvenile period, and into adulthood.
+However, in accordance with our study, a recent study on Soay sheep (*Ovis aries*) found a weak correlation between juvenilt and adult RTL, and a highly complex pattern of telomere dynamics across the entire lifespan, with periods of early-life telomere shortening followed by selective disappearance and apparent telomere lengthening as individuals age [@Fairlie2015].
+Indeed, the complex relationship between RTL and age in our study is strikingly similar to that reported by Fairlie *et al.* [-@Fairlie2015].
+It is too soon to speculate as to what may be generating this variation among species, but one possibility is that environmental variability itself affects the resolution with which we can observe telomere shortening in natural populations [see also @Fairlie 2015].
+As early-life telomere dynamics are investigated in more systems, the causes of this variation should become clear.
 
-We suspect that the novelty in our finding temporal, environmentally-induced variation in telomere dynamics within a population is more due to a lack of available long-term datasets with telomere screening, rather than the Seychelles warbler being unique.
-Indeed, the environment on Cousin is benign in comparison with many regions outside the tropics, where populations undergo large fluctuations in size [e.g.
+Perhaps the clearest result from our study is that RTL varies among cohorts.
+Very few studies have shown that temporal variation in telomere dynamics occurs in natural populations, these have been limited in the number of seasons [@Mizutani2013; @Watson2015; but see @Fairlie2015].
+The long-term Seychelles warbler dataset has allowed us to show that temporal variation in telomere dynamics does occur at the population level, over decadal time periods, and that this variation was related to variation in the environment (Fig.
+1C).
+The difference in RTL between birds born in summer and winter seasons is more surprising, especially given the strength and directionality of the effect.
+We find that birds born in winter seasons having longer telomeres, despite the .
+One possible explanation for this is that parents in good condition are more likely to breed
+
+We suspect that the novelty in our finding temporal variation in telomere dynamics within a population is more due to a lack of available long-term datasets with telomere screening, rather than the Seychelles warbler being unique.
+Indeed, the population on Cousin is relatively stable in comparison with other populations, which can undergo large fluctuations in size [e.g.
 @Coulson2001].
-If our findings are replicated in other systems and population-level variation in early-life telomere dynamics is common in nature, this has potential conseaquences for our understanding of telomere dynamics in natural populations.
+If our findings are replicated in other systems and population-level variation in early-life telomere dynamics is common in nature, this has potential consequences for our understanding of telomere dynamics in natural populations.
 In particular, our findings suggest that the telomere dynamics of a population at a given point in time represent a snapshot of a temporally varying process.
@@
 
-In the Seychelles warbler we found only weak evidence that any of the social or environmental vairables we measured affect juvenile RTL.
-This is surprising a poor social and ecological environment is known to be detrimental to juvenile Seychelles warblers, both in terms of oxidative stress and later life survival.
-Possible explanations for this finding include i) temporal variation in RTL in our data is so strong that we are unable to detect spatial trends, ii) within-cohort variation in RTL is explained by a variable that we have not included in our analyses, or iii) that effects are generally weak and levels of noise in our telomere measurments and/or ecological data preclude detection of significant effects.
-Future research should therefore examine, in a quantitative genetic framework, how genetic and environmental components, and their interactions, affect telomere dynamics and senescence in natural populations [@Becker2015; @Asghar2014].
-
-Our data suggest that, in addition to age class, tarsus length and season had weak effects on juvenile RTL (Figs 3, 4).
-The tarsus effect most likely reflects the fact that in passerine birds tarsus length is correlated with age during the nestling stage [@Ricklefs1976].
-Telomere loss is most rapid early in life due to ongoing cell replication [@Frenck1998], and a negative correlation between RTL and body size in early life is therefore expected.
-Indeed, that age class affects both tarsus length and RTL can be clearly seen in our data (Fig.
-4B).
-The difference in RTL between birds born in summer and winter seasons is more surprising, especially given the directionality of the effect.
-We find that birds born in winter seasons having longer telomeres (Fig.
-4A), while the opp.
-One possible explanation for this is that parents in good condition are more likely to breed
+Within cohorts, we found no evidence that any of the social or environmental variables we measured affect juvenile RTL.
+This is surprising, as a poor social and ecological environment is known to be detrimental to juvenile Seychelles warblers, both in terms of oxidative stress and later life survival.
+Possible explanations for this finding include i) temporal variation in RTL in our data masks our ability to detect spatial trends, ii) within-cohort variation in RTL is explained by a variable that we have not included in our analyses, or iii) that effects are generally weak and levels of noise in our telomere measurements and/or ecological data preclude detection of significant effects.
+The first of these explanations seems the least likely, as within-cohort variation in RTL greatly exceeds among-cohort variation (Fig.
+1C).
+One thing unclear at present is to what extent early-life telomere length reflects inheritance [e.g.
+@Becker2015; @Asghar2014], and this may capture a large amount of unexplained variation in early-life RTL within our dataset.
+Future research on the Seychelles warbler will examine, in a quantitative genetic framework, how genetic and environmental components, and their interactions, affect telomere dynamics and senescence in natural populations.
 
@@
 In captive zebra finches, juvenile telomere length predicts late-life survival [@Heidinger2012], and in wild bird populations survival to the nestling phase [@Watson2015], and survival to adulthood [@Boonekamp2014], have been linked with early-life telomere dynamics.
-
-Finally, our study highlights some of the diffculties asosciated with studying telomere dynamics in natural populations.
-
-Our longitudinal dataset was limited, both in terms of sample size and resolution (i.e.
-time between sampling events).
-Seychelles warblers are rarely sampled more than once within their first year of life, so much of the telomere shortening that occurs in early life will be missed with our sampling regime.
-It is likely, therefore that only very strong effects of environmental variation on telomere loss will  be detected in this dataset.
-Thus telomere length constitutes a better indicator of early-life stress.
 
@@
 The Seychelles Bureau of Standards and Department of Environment gave permission for sampling and fieldwork.
-Emma Barrett laid the foundations for this study, generating much of the telomere data.
-We thank everyone who has helped in the field, and the current Seychelles warbler research group for useful discussions.
+Emma Barrett laid the foundations for this study, generating the original telomere qPCR protocol.
+We thank everyone who has helped in the field, and the Seychelles warbler research group for discussions.
 This work was funded by two Natural Environment Research Council (NERC) grants to DSR (NE/F02083X/1 and NE/K005502/1).
-HLD was funded by a NERC fellowship.
+LGS was also funded by an EGI Fellowship from the University of Oxford, and HLD by a NERC fellowship.
 

### 2016-03-16 Manuscript/SW_early_life_telomeres_SOM.Rmd
Discussion and fogure edits

@@
 ```{r Table S2}
-temp <- Loss
+temp <- subset(Loss,Agemonths == 1)
 
 #Run the model
-lm1 <- lmer(TROC ~ Agemonths * (Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ + Season) + (1|LayYear),
+lm1 <- lmer(Loss ~ (Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ + Season) + (1|LayYear),
             data = temp,
@@
 
-tableS2 <- data.frame(Model = apply(t1,1,getterms,c(2:16)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
+tableS2 <- data.frame(Model = apply(t1,1,getterms,c(2:8)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
 colnames(tableS2) <- c('Model','df','AICc', 'Delta AICc','Weight')

### 2016-03-16 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Discussion and fogure edits

@@
 temp <- dd
-temp$Agemonths[temp$Agemonths>10] <- '12+'
-temp$Agemonths <- factor(temp$Agemonths,levels = c('1','6','10','12+'))
 
 Fig1A <- ggplot(temp,aes(x = Agemonths,y = RTL))+
-  geom_jitter(col='grey') +
   ylab('RTL') + 
@@
         legend.position = 'none')+
-  geom_boxplot(outlier.colour=NA,alpha = 0.2)+
-  annotate('text',x = 0.6,y = 2.1,label = 'A')
+  geom_point(col='grey')+
+  stat_smooth(method = 'lm',col='darkgrey')+
+  annotate('text',x = 1,y = 2.1,label = 'A')
 
-temp <- Loss
-temp$Agemonths <- factor(temp$Agemonths)
+temp <- subset(dd,DeltaAge !=0)
 
-Fig1B <- ggplot(temp,aes(x = Agemonths,y = TROC))+
-  geom_jitter(col='grey') +
-  ylab('Loss in RTL') + 
-  xlab('Age (months)') +
+Fig1B <- ggplot(temp,aes(x = DeltaAge,y = RTL))+
+  ylab('RTL') + 
+  xlab(expression(Delta*' age (months)')) +
   theme_classic() +
@@
         legend.position = 'none')+
-  geom_boxplot(outlier.colour=NA,alpha = 0.2)+
-  annotate('text',x = 0.6,y = 2.1,label = 'B')
-
-temp <- juv
-temp$LayYear <- factor(temp$LayYear)
-
-Fig1C <- ggplot(temp,
-       aes(x = LayYear,
-           y = RTL)) +
-  ylab('RTL') + 
-  xlab('') +
-  theme_classic() +
-  theme(axis.text.x = element_text(angle = 60,size = 14,vjust=0.5), 
-        axis.text.y = element_text(size = 14), 
-        axis.title.y = element_text(size = 16,vjust=0.8),
-        axis.title.x = element_text(size = 16),
-        legend.position = 'none') +
-  geom_jitter(col = 'grey')+
-  geom_boxplot(outlier.colour = NA,alpha = 0.2)+
-  scale_x_discrete(limits = factor(1998:2014),drop=F)+
-  annotate('text',x = 1,y = 2.1,label = 'C')
+  geom_point(col='grey')+
+  stat_smooth(method = 'lm',col = 'darkgrey')+
+  annotate('text',x = -100,y = 2.1,label = 'B')
 
 pdf('Figures/Figure 1.pdf',
-    width = 6,
-    height = 6,
-    paper = 'a4r')
-multiplot(Fig1A,Fig1B,Fig1C,layout=matrix(c(1,2,3,3),nrow = 2,byrow=T))
+    width = 4,
+    height = 8,
+    paper = 'a4')
+multiplot(Fig1A,Fig1B,layout=matrix(c(1,2),nrow = 2,byrow=T))
 dev.off()
@@
 ```{r Figure 2}
+
+temp <- juv
+temp$LayYear <- factor(temp$LayYear)
+
+Fig2A <- ggplot(temp,aes(x = LayYear,y = RTL,col=Season,fill=Season))+
+  ylab('RTL') + 
+  xlab('Cohort') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12,angle = 90), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        legend.position = 'top')+
+  geom_jitter(aes(colour = Season),position=position_jitterdodge(dodge.width = 0.8))+
+  geom_boxplot(outlier.colour = NA,notch=F,alpha=0.2)+
+  scale_colour_manual(values = c('grey','darkorange'))+
+  scale_fill_manual(values = c('grey','darkorange'))+
+  annotate('text',x = 1,y = 2.1,label = 'A')
+
+temp <- juv
+x <- ifelse(temp$Season == 'Winter',0.5,0)
+temp$LayYear2 <- temp$LayYear+x
+temp <- temp[!(duplicated(temp$LayYear2)),]
+
+Fig2B <- ggplot(temp,aes(x = LayYear2,y = Density))+
+  ylab('Adult population size') + 
+  xlab('Cohort') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        legend.position = 'none')+
+  geom_line(stat = 'identity')+
+  annotate('text',x = 1998,y = 360,label = 'B')
+
+temp <- juv
+
+Fig2C <- ggplot(temp,
+       aes(x = Density,
+           y = RTL,
+           colour=Season,
+           fill = Season)) +
+  ylab('RTL') + 
+  xlab('Adult population size') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        legend.position = 'none') +
+  geom_point()+
+  stat_smooth(method = 'lm')+
+  scale_colour_manual(values = c('grey','darkorange','darkgreen'))+
+  scale_fill_manual(values = c('grey','darkorange','darkgreen'))+
+  annotate('text',x = 240,y = 2.1,label = 'C')
+
+pdf('Figures/Figure 2.pdf',
+    width = 8,
+    height = 8,
+    paper = 'a4')
+multiplot(Fig2A,Fig2B,Fig2C,layout=matrix(c(1,1,2,3),nrow = 2,byrow=T))
+dev.off()
+
+```
+
+```{r Figure 3}
 
@@
 nams[nams == 'z.NonHelper'] <- 'Non helpers'
-nams[nams == 'c.Sex'] <- 'Sex'
+nams[nams == 'c.Sex'] <- 'Sex (males)'
 nams[nams == 'z.TQ'] <- 'TQ'
@@
 nams[nams == 'z.Tarsus'] <- 'Tarsus'
-nams[nams == 'c.Season'] <- 'Season'
+nams[nams == 'c.Season'] <- 'Season (winter)'
 nams[nams == 'c.Season:z.Agemonths'] <- 'Age x season'
@@
 
-Fig2A <- forestplot(table2)+
+Fig3A <- forestplot(table2)+
   annotate('text',x = 1,y = 0.25,label = 'A')
@@
 
-Fig2B <- ggplot(temp,
-       aes(x = factor(Agemonths),
+Fig3B <- ggplot(temp,
+       aes(x = `Age (months)`,
            y = RTL,
            fill=Season)) +
-  ylab('RTL') + 
-  xlab('Age (months)') +
   theme_classic() +
@@
 
-Fig2C <- ggplot(temp,
+Fig3C <- ggplot(temp,
        aes(x = Tarsus,
@@
         legend.position = 'top') +
-  geom_point(alpha = 0.2)+
+  geom_point()+
   stat_smooth(method = 'lm')+
@@
 
-pdf('Figures/Figure 2.pdf',
-    width = 8,
+Fig3D <- ggplot(temp,
+       aes(x = `Age (months)`,
+           y = RTL,
+           fill=Sex)) +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        legend.position = 'top') +
+  geom_jitter(aes(colour = Sex),position=position_jitterdodge(dodge.width = 0.8))+
+  geom_boxplot(outlier.colour = NA,notch=F,alpha=0.2)+
+  scale_colour_manual(values = c('grey','darkorange'))+
+  scale_fill_manual(values = c('grey','darkorange'))+
+  annotate('text',x = 0.55,y = 2.1,label = 'D')
+
+pdf('Figures/Figure 3.pdf',
+    width = 9,
     height = 9,
     paper = 'a4r')
-multiplot(Fig2A,Fig2B,Fig2C,layout = matrix(c(1,1,2,3),nrow = 2,byrow = T))
+multiplot(Fig3A,Fig3B,Fig3C,Fig3D,layout = matrix(c(1,1,1,2,3,4),nrow = 2,byrow = T))
 dev.off()
@@
 
-```{r Figure 3}
+```{r Figure 4}
 
 temp <- Loss
+temp[,'Age (months)'] <- factor(temp$Agemonths)
+temp$LayYear <- factor(temp$LayYear)
+
+Fig4A <- ggplot(temp,
+       aes(x = RTL,
+           y = RTL1)) +
+  ylab('Adult RTL') + 
+  xlab('Juvenile RTL') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        legend.position = 'top') +
+  geom_point(col='grey')+
+  stat_smooth(method = 'lm',col='darkgrey')+
+  annotate('text',x = 0,y = 2.1,label = 'A')
+
+temp2 <- temp
+temp2$TimeDiff <- temp2$TimeDiff/365
+
+Fig4B <- ggplot(temp2,
+       aes(x = TimeDiff,
+           y = Loss,
+           colour=`Age (months)`,
+           fill = `Age (months)`)) +
+  ylab(expression(Delta*'RTL')) + 
+  xlab('Time (years)') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        legend.position = 'top') +
+  geom_point(alpha=0)+
+  stat_smooth(method = 'lm')+
+  scale_colour_manual(values = c(grey(0.2),'darkorange','darkgreen'))+
+  scale_fill_manual(values = c('darkgrey','darkorange','darkgreen'))+
+    geom_abline(slope = 0,intercept = 0,lty = 2)+
+  coord_cartesian(ylim = c(-0.5,0.25))+
+  annotate('text',x = 0,y = 0.5,label = 'B')
+
+pdf('Figures/Figure 4.pdf',
+    width = 4,
+    height = 8,
+    paper = 'a4')
+multiplot(Fig4A,Fig4B,layout=matrix(c(1,2),nrow=2,byrow=T))
+dev.off()
+
+```
+
+```{r Figure 5}
 
 #####################
-#3A
+#4A
 #####################
-
+temp <- subset(Loss,Agemonths == 1)
 #Run the model
-lm2 <- lmer(TROC ~ Agemonths * (Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ + Season) + (1|LayYear),
+lm2 <- lmer(Loss ~ (Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ) + (1|LayYear),
             data = temp,
@@
 lm2 <- standardize(lm2)
+
 #Model averaging
-
 lm3 <- dredge(lm2)
@@
 lmavg <- model.avg(lm3,fit=T)
-myimp <- importance(lmavg) #Relative importance of each explanatory variable
+myimp <- importance(lmavg)
 
@@
 nams[nams == 'c.Season'] <- 'Season'
-nams[nams == 'c.Season:z.Agemonths'] <- 'Age x season'
-nams[nams == 'z.Agemonths:z.NonHelper'] <- 'Age x non helpers'
-nams[nams == 'c.Sex:z.Agemonths'] <- 'Age x sex'
-nams[nams == 'z.Agemonths:z.TQ'] <- 'Age x TQ'
-nams[nams == 'c.Helper:z.Agemonths'] <- 'Age x helpers'
-nams[nams == 'z.Agemonths:z.BodyMass'] <- 'Age x mass'
-nams[nams == 'z.Agemonths:z.Tarsus'] <- 'Age x tarsus'
 
@@
 
-Fig3A <- forestplot(table2)+
-  annotate('text',x = 1,y = 0.45,label = 'A')
+Fig4 <- forestplot(table2)
 
-temp <- Loss
-
-temp[,'Helpers'] <- ifelse(temp$Helper>0,'Present', 'Absent')
-temp[,'Non helpers'] <- ifelse(temp$NonHelper>0,'Present', 'Absent')
-temp[,'Age (months)'] <- factor(temp$Agemonths)
-
-Fig3B <- ggplot(temp,
-       aes(x = factor(Agemonths),
-           y = TROC,
-           fill=Season)) +
-  ylab('Loss in RTL') + 
-  xlab('Age (months)') +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        legend.position = 'top') +
-  geom_jitter(aes(colour = Season),position=position_jitterdodge(dodge.width = 0.8))+
-  geom_boxplot(outlier.colour = NA,notch=F,alpha=0.2)+
-  scale_colour_manual(values = c('grey','darkorange'))+
-  scale_fill_manual(values = c('grey','darkorange'))+
-  annotate('text',x = 0.55,y = 2.1,label = 'B')
-
-Fig3C <- ggplot(temp,
-       aes(x = Tarsus,
-           y = TROC,
-           col=`Age (months)`)) +
-  ylab('Loss in RTL') + 
-  xlab('Tarsus length (mm)') +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        legend.position = 'top') +
-  geom_point(alpha = 0.2)+
-  stat_smooth(method = 'lm')+
-    scale_colour_manual(values = c(grey(0.2),'darkorange','darkgreen'))+
-  scale_fill_manual(values = c('darkgrey','darkorange','darkgreen'))+
-  annotate('text',x = 20,y = 2.1,label = 'C')
-
-pdf('Figures/Figure 3.pdf',
-    width = 8,
-    height = 9,
-    paper = 'a4r')
-multiplot(Fig3A,Fig3B,Fig3C,layout = matrix(c(1,1,2,3),nrow = 2,byrow = T))
+pdf('Figures/Figure 4.pdf',
+    width = 5,
+    height = 4,
+    paper = 'a4')
+Fig5
 dev.off()

### 2016-03-16 Scripts/SW juvenile telomeres IPC clean.R

@@ -4,7 +4,7 @@
 #################################################################################
 
 #Only use Ellie's samples
-#dd0 <- subset(dd0,Whodunnit == 'EAF')
+dd0 <- subset(dd0,Whodunnit == 'EAF')
 
 #Average repeats of blood samples
 av <- ave(dd0$RTL,c(dd0$BloodID,dd0$Status,dd0$PlateID))
@@ -25,35 +25,12 @@ colnames(dd)[colnames(dd) == 'MinOfFieldPeriodID'] <- 'FieldPeriodID'
 dd$CatchYear <- as.numeric(substr(dd$CatchDate,7,10))
 dd$DeathYear <- as.numeric(substr(dd$DeathDate,7,10))
 dd$CatchDate <- as.Date(dd$CatchDate,"%d/%m/%Y")
+dd$DeathDate <- as.Date(dd$DeathDate,"%d/%m/%Y")
+
+
 
 dd$Season <- ifelse(as.numeric(format(dd$CatchDate,'%m')) %in% c(4:10),
                     'Summer','Winter')
-
-
-# Tarsus ------------------------------------------------------------------
-
-dd$Tarsus <- NA
-for(i in 1:nrow(dd))
-{
-  ifelse(is.na(dd$RightTarsus[i]),
-         ifelse(is.na(dd$LeftTarsus[i]),
-                dd$Tarsus[i] <- NA,
-                dd$Tarsus[i] <- dd$LeftTarsus[i]),
-         dd$Tarsus[i] <- dd$RightTarsus[i])
-}
-
-dd$RightTarsus <- NULL
-dd$LeftTarsus <- NULL
-
-
-
-# Remove unwanted data/outliers ----------------------------------------------------
-
-dd <- subset(dd,RTL<2)
-dd <- subset(dd,RTL > 0.04)
-dd <- subset(dd,BodyMass>5)
-dd <- subset(dd,Tarsus>17)
-
 
 
 # Age data ----------------------------------------------------------------
@@ -78,9 +55,45 @@ dd <- subset(dd,Agemonths!=12)
 
 dd$Age[dd$Fledged != 'Adults'] <- 1
 
+
+
+
+# Tarsus and delta age ------------------------------------------------------------------
+
+dd$Tarsus <- NA
+dd$DeltaAge <- NA
+for(i in 1:nrow(dd))
+{
+  ifelse(is.na(dd$RightTarsus[i]),
[…]

### 2016-03-16 Scripts/SW juvenile telomeres IPC load.r

@@ -15,6 +15,7 @@ library(plyr)
 library(ggplot2)
 library(gvlma)
 library(survival)
+library(nlme)
 
 # Load data ---------------------------------------------------------------
 dd0 <- read.csv('Data/REL_TL_EB_EAF_25 Feb FULL DATA.csv') #main dataset
@@ -22,3 +23,4 @@ terr <- read.csv('Data/SW TL all territories.csv')
 status <- read.csv('Data/SW TL all breedstatus.csv')
 insects <- read.csv('Data/Insects.csv')
 dens <- read.table('Data/PsizeFP.txt')
+chickinfo <- read.csv('Data/Table chick info.csv')

### 2016-03-21 Manuscript/SW_early_life_telomeres.Rmd
More analyses

@@
 
-**Strong cohort effects on early-life telomere length in a wild population**
+**Early-life conditions, telomere dynamics and survival: insights from a long-term island bird study**
 
@@
 ###Abstract
+1.
 Understanding the short and long term costs of individual early-life experiences is fundamental to understanding life-history evolution.
@@
 Thus, it is possible that telomere dynamics may help us quantify individual variation in early-life costs, and enhance our understanding of how poor conditions in early life are related to later-life survival.
+
+2.
 We tested how telomere dynamics are related to spatio-temporal variation in early-life conditions and later-life survival in the Seychelles warbler (*Acrocephalus sechellensis*), using a large dataset of `r nrow(juv)` juveniles from multiple cohorts spanning `r max(juv$LayYear)-min(juv$LayYear)` years.
-Despite cross-sectional data showing that telomere length decreases with age, and most that telomere length decreases most rapidly early in life, using longitudinal data we found that significant reductions in telomere length can only be detected within the first few months of life - after this, longitudinal changes in telomere length were not significantly different from zero.
-Juvenile telomere length varied markedly among cohorts, and this cohort variation was related to the season (birds born in winter seasons had longer telomeres), and to population density (birds bird in years with higher populaiton density had longer telomered).
+
+3.
+We found that at the population level, telomere length decreases with cross-sectional and longitudinal measures of age, and that telomere length decreases most rapidly early in life.
+However, at the individual level we found only a very weak relationship between juvenile and adult telomere length, and that significant reductions in telomere length can only be detected within the first few months of life - after this, longitudinal changes in telomere length were not significantly different from zero.
+
+4.
+Juvenile telomere length varied markedly among cohorts, and birds born in years with higher populaiton density had longer telomeres.
 Within cohorts, we found no evidence that the early-life social environment or ecological conditions were related to telomere length or rates of shortening.
-Finally, we found no relationship between early-life and survival to adulthood.
+We also found no relationship between early-life telomere dynamics and survival later in life.
+
+5.
 Our data suggest that telomeres are subject to strong population-level variation, but suggest that the resolution of telomeres as a useful biomarker of individual-level cost is likely to be age-dependent, and in many cases limited.
@@
 All DNA extractions that passed the above criteria were diluted to 3.3 ng $\mu$l^-1^ before telomere measurement.
-We measured telomere length for all samples using a quantitative PCR (qPCR) assay of telomeres and a GAPDH control gene, using the molecular methods outlined by Barrett *et al.* [-@Barrett2012], with one amendment.
+We measured telomere length for all samples using a quantitative PCR (qPCR) assay of telomeres and a GAPDH control gene, using the molecular methods outlined by Barrett *et al.* [-@Barrett2012], with two amendments.
 A change in batch of SYBR green forced us to raise the annealing temperature of the telomere reaction from 58^o^C to 61^o^C for the majority of samples.
 However, this did not affect final telomere length values (see below).
+Second, to avoid bacth effects, we randomised samples among plates.
 We used the program LinRegPCR [@Ruijter2009] to correct for baseline fluorescence and calculate efficiencies and Cq values for each sample replicate.
@@
 For the analysis of $\Delta$age we we only included birds who had been sampled at least twice.
-We then explored how telomere length varied within the first year of life (telomere length vs age in months) and over time (telomere length vs cohort ID).
-For this analysis we had bvery few birds with multiple samples, so we used linear models including only the first sample for each individual.
+We then explored how telomere length varied within the first year of life (telomere length vs age in months) and among cohorts (telomere length vs cohort ID), again using linear mixed models with individual ID as a random effect.
 
@@
 Model averaging was carried out using the MuMIn package (version 1.10.5) in R [@Barton2012].
-We first created a full model containing the following explanatory variables: age (in months), tarsus length, sex, territory quality, season (summer or winter) and the number of helping and non-helping subordinate birds present in the natal territory.
+We first created a full model containing the following explanatory variables: age (in months), tarsus length, sex, territory quality, season (major or minor) and the number of helping and non-helping subordinate birds present in the natal territory.
 We also included interaction terms between age and all the other variables as telomere dynamics are expected to vary within the first year of life [@Heidinger2012].
@@
 #####Early-life RTL
-Considering the entire Seychelles warbler lifespan, RTL decreased with age (t = `r round2(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,4])`; P `r round3(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,5])`; Fig.
-1A), and with $\Delta$age (t = `r round2(summary(lme(RTL~DeltaAge,data=subset(dd,DeltaAge!=0),random=~1|BirdID))$tTable[2,4])`; P `r round3(summary(lme(RTL~DeltaAge,data=subset(dd,DeltaAge!=0),random=~1|BirdID))$tTable[2,5])`; Fig.
+Considering the entire Seychelles warbler lifespan, RTL decreased with age (LMM: t = `r round2(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,4])`; P `r round3(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,5])`), although there were clear increases in cross-sectional telomere length after one year of age and at several points later in life (Fig.
+1A).
+RTL also decreased with age within individuals (i.e.
+with $\Delta$age; t = `r round2(summary(lme(RTL~DeltaAge,data=subset(dd,DeltaAge!=0),random=~1|BirdID))$tTable[2,4])`; P `r round3(summary(lme(RTL~DeltaAge,data=subset(dd,DeltaAge!=0),random=~1|BirdID))$tTable[2,5])`; Fig.
 1B).
 RTL also decreased with age within the first year of life (i.e.
-excluding adult samples; R^2^ = `r round2(summary(lm(RTL~Agemonths,data=juv))$r.squared)`; t = `r round2(summary(lm(RTL~Agemonths,data=juv))$coefficients[2,3])`; P `r round3(summary(lm(RTL~Agemonths,data=juv))$coefficients[2,4])`).
-Early-life RTL varied significantly among cohorts (one-way ANOVA, telomere length: F = `r round2(anova(lm(RTL~factor(FieldPeriodID),data = juv))[1,4])`; P `r round3(anova(lm(RTL~factor(FieldPeriodID),data = juv))[1,5])`), and cohort-level variation in juvenile RTL (taken as the median for each cohort) was positively related to adult population density (R^2^ = `r round2(summary(lm(RTL~Density,data=juvseason,na.action=na.exclude))$r.squared)`; t = `r round2(summary(lm(RTL~Density,data = juvseason,na.action=na.exclude))$coefficients[2,3])`; P = `r round2(anova(lm(RTL~Density,data = juvseason,na.action=na.exclude))[1,5])`; Fig.
-1C).
-Variation in cohort-level RTL was not related temporal variation in  territory quality (linear regression, R^2^ = `r round2(summary(lm(RTL~TQ,data=juvseason,na.action=na.exclude))$r.squared)`; t = `r round2(summary(lm(RTL~TQ,data = juvseason,na.action=na.exclude))$coefficients[2,3])`; P = `r round2(anova(lm(RTL~TQ,data = juvseason,na.action=na.exclude))[1,5])`), or island-wide food availability (R^2^ = `r round2(summary(lm(RTL~Insect,data=juvseason,na.action=na.exclude))$r.squared)`; t = `r round2(summary(lm(RTL~Insect,data = juvseason,na.action=na.exclude))$coefficients[2,3])`; P = `r round2(anova(lm(RTL~Insect,data = juvseason,na.action=na.exclude))[1,5])`).
+excluding adult samples; t = `r round2(summary(lme(RTL~Agemonths,data=juv,random=~1|BirdID))$tTable[2,4])`; P `r round3(summary(lme(RTL~Agemonths,data=juv,random=~1|BirdID))$tTable[2,5])`).
 
-The top model explaining variation in early-life RTL contained age, sex, season (summer vs winter), tarsus length, and the interaction between age and tarsus length (Table S1).
-The top model was much better supported than a null model ($\Delta$AICc = `r round2(AICc(lmer(RTL~1+(1|LayYear)+(1|TerritoryID),data=juv,REML = FALSE))-AICc(lmer(RTL~Agemonths+Sex+Season+Tarsus+Agemonths*Tarsus+(1|LayYear)+(1|TerritoryID),data=juv,REML = FALSE)))`), and with the exception of the sex and tarsus effects, model-averaged confidence intervals did not overlap zero (Fig.
-2A).
-RTL was higher in winter compared to summer seasons (Fig.
-2B).
+Early-life RTL varied significantly among cohorts (Likelihood ratio test: F = `r round2(anova(lme(RTL~factor(FieldPeriodID),data=juv,random=~1|BirdID))[2,3])`; P `r round3(anova(lme(RTL~Agemonths,data=juv,random=~1|BirdID))[2,4])`).
+Cohort-level variation in juvenile RTL was positively related to adult population density (LMM controlling for age and cohort: t = `r round2(summary(lme(RTL~Agemonths+Density,data=juv,random=~1|FieldPeriodID))$tTable[3,4])`; P `r round3(summary(lme(RTL~Agemonths+Density,data=juv,random=~1|FieldPeriodID))$tTable[3,5])`; Figs 2B,2C).
+Variation in cohort-level RTL was not related temporal variation in  food availability (t = `r round2(summary(lme(RTL~Agemonths+Insect,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,4])`; P `r round3(summary(lme(RTL~Agemonths+Insect,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,5])`).
+RTL did not differ in birds born in minor compared to major seasons (t = `r round2(summary(lme(RTL~Agemonths+Season,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,4])`; P `r round3(summary(lme(RTL~Agemonths+Season,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,5])`; Fig.
+2A, 2C).
+
+Within cohorts, the top model explaining variation in early-life RTL contained age, sex, tarsus length, and the interaction between age and tarsus length (Table S1).
+The top model was much better supported than a null model ($\Delta$AICc = `r round2(AICc(lmer(RTL~1+(1|FieldPeriodID)+(1|TerritoryID),data=juv,REML = FALSE))-AICc(lmer(RTL~Agemonths+Sex+Tarsus+Agemonths*Tarsus+(1|FieldPeriodID)+(1|TerritoryID),data=juv,REML = FALSE)))`), and with the exception of the sex and tarsus effects, model-averaged confidence intervals did not overlap zero (Fig.
+3A).
 Tarsus length was negatively related to RTL in nestlings, unrelated to RTL in fledglings, and positively related to RTL in subadults (Fig.
-2C).
-Males had slightly longer telomeres than females, prarticularly at the fledgling and subadult stages (Fig.
-2D), although this effect was weak.
+3B).
+Males had slightly longer telomeres than females, particularly at the fledgling and subadult stages, although this effect was weak  (Fig.
+3C).
 
@@
 Longitudinal data showed that early-life RTL was positively, but weakly, related to adult RTL (R^2^ = `r round3(summary(lm(RTL1~RTL+Agemonths,data=Loss))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data=Loss))$coefficients[2,3])`; P = `r round3(summary(lm(RTL1~RTL+Agemonths,data = Loss))$coefficients[2,4])`; Fig.
-3A).
+4A).
 $\Delta$RTL was significantly lower than zero in nestlings (t = `r round2(t.test(subset(Loss,Agemonths == 1)$Loss)$statistic)`; P `r round3(t.test(subset(Loss,Agemonths == 1)$Loss)$p.value)`), but not in fledglings (t = `r round2(t.test(subset(Loss,Agemonths == 6)$Loss)$statistic)`; P = `r round2(t.test(subset(Loss,Agemonths == 6)$Loss)$p.value)`) or subadults (t = `r round2(t.test(subset(Loss,Agemonths == 10)$Loss)$statistic)`; P = `r round2(t.test(subset(Loss,Agemonths == 10)$Loss)$p.value)`).
@@
 In nestlings, $\Delta$RTL  did not vary with the time between sampling events (R^2^ `r round3(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths ==1)))$r.squared)`; t = `r round2(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths ==1)))$coefficients[2,3])`; P = `r round3(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths ==1)))$coefficients[2,4])`).
-However, for fledglings and subadults $\Delta$RTL decreased with the length of time between sampling events (R^2^ = `r round2(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths !=1)))$r.squared)`; t = `r round2(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths !=1)))$coefficients[2,3])`; P = `r round3(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths !=1)))$coefficients[2,4])`).
-In flesdglings and subadults, $\Delta$RTL did not decrease below zero until ~3 years between sampling events was allowed, and confidence limits overlapped zero regardless of the amount of time between sampling events (Fig 3B).
+However, in fledglings and subadults $\Delta$RTL decreased with the length of time between sampling events (R^2^ = `r round2(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths !=1)))$r.squared)`; t = `r round2(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths !=1)))$coefficients[2,3])`; P = `r round3(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths !=1)))$coefficients[2,4])`; Fig.
+4B, Fig.
+S1).
+In both fledglings and subadults, $\Delta$RTL did not decrease below zero until ~3 years between sampling events was allowed, and confidence limits overlapped zero regardless of the amount of time between sampling events (Fig 3B).
 Together, these results suggest that the majority of observable telomere shortening in Seychelles warblers occurs in the first six months of life.
-We therefore restricted the remaining analyses of $\Delta$RTL to birds sampled as nestlings.
+We therefore restricted the remaining analyses of $\Delta$RTL to birds sampled as nestlings (N = `r nrow(subset(Loss,Agemonths == 1))`.
 
-$\Delta$RTL did not differ significantly among cohorts (F = `r round2(anova(lm(Loss~factor(LayYear),data = subset(Loss,Agemonths == 1)))[1,4])`; P `r round3(anova(lm(Loss~factor(LayYear),data = subset(Loss,Agemonths == 1)))[1,5])`), although power for this analysis was limited as longitudinal sample sizes within cohorts was low (Fig.
-3C).
-Model selection revealed that the top model explaining $\Delta$RTL was the null model (Table S2), and confidence limits for all explanatory terms overlapped zero (Fig.
-4A).
+$\Delta$RTL did not differ significantly among cohorts of nestlings (F = `r round2(anova(lm(Loss~factor(FieldPeriodID),data = subset(Loss,Agemonths == 1)))[1,4])`; P `r round3(anova(lm(Loss~factor(FieldPeriodID),data = subset(Loss,Agemonths == 1)))[1,5])`), although power for this analysis was limited as longitudinal sample sizes within cohorts were low.
+Model selection revealed that the top model explaining $\Delta$RTL in nestlings contained sex and tarsus length (Table S2).
+This model was a better fit than the null model ($\Delta$AICc = `r round2(AICc(lmer(Loss~1+(1|FieldPeriodID)+(1|TerritoryID),data=subset(Loss,Agemonths == 1),REML = FALSE))-AICc(lmer(Loss~Sex+Tarsus+(1|FieldPeriodID)+(1|TerritoryID),data=subset(Loss,Agemonths == 1),REML = FALSE)))`), although confidence limits for both the sex and tarsus effects overlapped zero (Fig.
+5A).
+$\Delta$RTL was negatively related to tarsus length, and higher in females, indicating higher rates of telomere loss in smaller nestlings, and in males.
 
@@
 
-Early-life RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~RTL+(1|TerritoryID)+(1|LayYear),data=juv13,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL+Agemonths+Tarsus+(1|TerritoryID)+(1|LayYear),data=juv13,family = 'binomial'))$coefficients[2,4])`), with no interaction effect between RTL and juvenile age on survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|LayYear),data=juv13,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|LayYear),data=juv13,family = 'binomial'))$coefficients[4,4])`).
+Early-life RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,4])`), with no interaction effect between RTL and juvenile age on survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,4])`).
 To test whether cohort-level effects on RTL influenced survival we calculated cohort-level RTL (i.e.
-the median RTL for each cohort) and added this term to a GLMM, and found that cofort-level RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~cohortTL+Agemonths+(1|TerritoryID)+(1|LayYear),data=juv13,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~cohortTL+Agemonths+(1|TerritoryID)+(1|LayYear),data=juv13,family = 'binomial'))$coefficients[2,4])`), with no interaction with juvenile age (estimate = `r round2(summary(glmer(SurvivedNext~cohortTL*Agemonths+(1|TerritoryID)+(1|LayYear),data=juv13,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~cohortTL*Agemonths+(1|TerritoryID)+(1|LayYear),data=juv13,family = 'binomial'))$coefficients[4,4])`).
-Using the longitudinal data, we found that $\Delta$RTL had no effect on survival to the year after the second sampling event (estimate = `r round2(summary(glm(SurvivedNext2~Loss,data=subset(subset(Loss,Agemonths == 1),LayYear<2013),family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glm(SurvivedNext2~RTL+Agemonths,data=subset(subset(Loss,Agemonths == 1),LayYear<2013),family = 'binomial'))$coefficients[2,4])`).
+the median RTL for each cohort) and added this term to a GLMM, and found that cofort-level RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~cohortTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~cohortTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,4])`), with no interaction with juvenile age (estimate = `r round2(summary(glmer(SurvivedNext~cohortTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~cohortTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,4])`).
+Using the longitudinal data, we found that $\Delta$RTL had no effect on survival to the year after the second sampling event (estimate = `r round2(summary(glm(SurvivedNext2~Loss,data=subset(subset(Loss,Agemonths == 1),FieldPeriodID<2013),family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glm(SurvivedNext2~RTL+Agemonths,data=subset(subset(Loss,Agemonths == 1),LayYear<2013),family = 'binomial'))$coefficients[2,4])`).
 
@@
 Here we use one of the largest datasets to date to assess the relationships between early-life conditions, telomere length and survival in a closed population of Seychelles warblers.
-We found that, as in other studies, telomeres are lost at the greatest rate very early in life.
-We found that telomere dynamics are subject to strong seasonal and cohort effects, but that temporal variation in RTL is not related to any of the environmental variables tested.
-Within seasons, we found no evidence that the ecological or social environment influenced telomere dynamics, and we found no evidence that early-life telomere length is related to juvenile or later-life survival.
+We found that, as in other studies, telomeres are lost at the greatest rate  early in life.
+We found that telomere dynamics are subject to strong cohort effects, and that temporal variation in RTL is population density and season.
+Within cohorts, we found no evidence that the ecological or social environment influenced telomere dynamics, and we found no evidence that early-life telomere length is related to juvenile or later-life survival.
 
-Our study adds the the growing number showing that telomere length decreases with age, and that this decrease is most rapid in early life [e.g.
-@Frenck1998; @Haussmann2003; @Heidinger2012].
+Our study adds the the now substantial body of literature showing that telomere length decreases with age, and that this decrease is most rapid in early life [e.g.
+@Frenck1998; @Haussmann2003; @Heidinger2012; @Fairlie2015].
 Our finding that tarsus length was related to juvenile RTL in an age-dependent manner (Fig.
-2) most likely an additional age effect - in passerine birds tarsus length is correlated with age during the nestling stage, but not after fledging [@Ricklefs1976].
-Telomere loss is most rapid early in life due to ongoing growth [@Frenck1998], and a negative correlation between RTL and body size in early life is therefore expected.
+3) most likely represents an additional age effect - in passerine birds tarsus length is correlated with age during the nestling stage, but not after fledging [@Ricklefs1976].
+Telomere loss is most rapid early in life due to ongoing growth [@Frenck1998], and a negative correlation between RTL and body size in early life is therefore expected [see also @Ringsby2015].
 However, despite our clear finding that population-level RTL decreases with age throughout the warbler lifespan (Fig.
-1), we found that individual-level telomere shortening can only be detected  very early in life (Fig.
-3).
-This is in contrast to other avian studies where birds were reared in laboratory [@Heidinger2012] or nestbox [@Boonekamp2014] conditions, in which individual-level telomere shortening clearly occurs throughout the juvenile period, and into adulthood.
-However, in accordance with our study, a recent study on Soay sheep (*Ovis aries*) found a weak correlation between juvenilt and adult RTL, and a highly complex pattern of telomere dynamics across the entire lifespan, with periods of early-life telomere shortening followed by selective disappearance and apparent telomere lengthening as individuals age [@Fairlie2015].
-Indeed, the complex relationship between RTL and age in our study is strikingly similar to that reported by Fairlie *et al.* [-@Fairlie2015].
+1), we found that at the individual-level, juvenile and adult telomere length were weakly correlated, and that telomere shortening can only be detected very early in life (Fig.
+4).
+This is in contrast to other avian studies where birds were reared in laboratory [@Heidinger2012] or nestbox [@Boonekamp2014] conditions, in which juvenile and adult telomere length were highly correlated, and individual-level telomere shortening clearly occurs throughout the juvenile period, and into adulthood.
+However, in accordance with our study, a recent study on Soay sheep (*Ovis aries*) found a weak correlation between juvenile and adult RTL, and a highly complex pattern of telomere dynamics across the entire lifespan [@Fairlie2015].
+Indeed, the complex relationship between RTL and age in our study, with periods of early-life telomere shortening followed by selective disappearance and/or apparent telomere lengthening as individuals age is strikingly similar to that reported by Fairlie *et al.* [-@Fairlie2015].
 It is too soon to speculate as to what may be generating this variation among species, but one possibility is that environmental variability itself affects the resolution with which we can observe telomere shortening in natural populations [see also @Fairlie 2015].
@@
 Perhaps the clearest result from our study is that RTL varies among cohorts.
-Very few studies have shown that temporal variation in telomere dynamics occurs in natural populations, these have been limited in the number of seasons [@Mizutani2013; @Watson2015; but see @Fairlie2015].
-The long-term Seychelles warbler dataset has allowed us to show that temporal variation in telomere dynamics does occur at the population level, over decadal time periods, and that this variation was related to variation in the environment (Fig.
+Very few studies have shown that temporal variation in telomere dynamics occurs in natural populations, and these have been limited in the number of seasons [@Mizutani2013; @Watson2015; @Fairlie2015].
+The long-term Seychelles warbler dataset has allowed us to show that temporal variation in telomere dynamics can occur over substantial time periods.
+Moroevoer, becuase of the larger number of cohorts we have been able to show statistically that that temporal variation in RTL is related to variation in the environment (Fig.
 1C).
-The difference in RTL between birds born in summer and winter seasons is more surprising, especially given the strength and directionality of the effect.
-We find that birds born in winter seasons having longer telomeres, despite the .
-One possible explanation for this is that parents in good condition are more likely to breed
-
-We suspect that the novelty in our finding temporal variation in telomere dynamics within a population is more due to a lack of available long-term datasets with telomere screening, rather than the Seychelles warbler being unique.
-Indeed, the population on Cousin is relatively stable in comparison with other populations, which can undergo large fluctuations in size [e.g.
-@Coulson2001].
-If our findings are replicated in other systems and population-level variation in early-life telomere dynamics is common in nature, this has potential consequences for our understanding of telomere dynamics in natural populations.
-In particular, our findings suggest that the telomere dynamics of a population at a given point in time represent a snapshot of a temporally varying process.
-More research is now needed within and across multiple cohorts and populations to better understand how how and why population-level telomere dynamics vary over space and time.
+We may expect a negative relationship between population density and RTL, as increased competition for resources is known to effect juvenile telomere length.
+However, We suggest that population density acts as an indicator of overall island 'quality' for Seychelles warblers.
+This can clearly be seen from data from the 1980s and 1990s, when adult population size on Cousin increased as habitat quality improved [@Komdeur2005].
+Population size has fluctuated around 300 since the late 1990s (Fig.
+2B), and captures this more accurately than measures of insect abundance or territory quality.
 
@@
 1C).
-One thing unclear at present is to what extent early-life telomere length reflects inheritance [e.g.
-@Becker2015; @Asghar2014], and this may capture a large amount of unexplained variation in early-life RTL within our dataset.
-Future research on the Seychelles warbler will examine, in a quantitative genetic framework, how genetic and environmental components, and their interactions, affect telomere dynamics and senescence in natural populations.
 
@@
 There are numerous avenues for future research into telomere evolutionary ecology in this system and others.
-Here we have considered survival, but telomere length and shortening in early life may also be linked to other components of fitness [reviewed in @Monaghan2014].
+If our findings are replicated in other systems, and population-level variation in early-life telomere dynamics is common in nature, this has potential consequences for our understanding of telomere dynamics in natural populations.
+In particular, our findings suggest that the telomere dynamics of a population at a given point in time represent a snapshot of a temporally varying process.
+More research is now needed within and across multiple cohorts and populations to better understand how how and why population-level telomere dynamics vary over space and time.
+Another question to be addressed is the extent to which early-life RTL reflects inheritance [e.g.
+@Becker2015; @Asghar2014].
+This is important, as inheritance may capture a large amount of unexplained variation in early-life RTL within our dataset.
+Future research on the Seychelles warbler will examine, in a quantitative genetic framework, how genetic and environmental components, and their interactions, affect telomere dynamics and senescence in natural populations.
+FInally, here we have considered survival, but telomere length and shortening in early life may also be linked to other components of fitness [reviewed in @Monaghan2014].
 Reproductive senescence occurs in the Seychelles warbler [reviewed in @Hammers2015], making this system well suited to examining how telomere length predicts lifetime reproductive success.

### 2016-03-21 Manuscript/SW_early_life_telomeres_SOM.Rmd
More analyses

@@
 #Run the model
-lm1 <- lmer(RTL ~ Agemonths * (Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ + Season) + (1|LayYear)+(1|TerritoryID),
+lm1 <- lmer(RTL ~ Agemonths * (Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ) + (1|FieldPeriodID)+(1|TerritoryID)+(1|PlateID),
             data = juv,
@@
 
-tableS1 <- data.frame(Model = apply(t1,1,getterms,c(2:16)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
+tableS1 <- data.frame(Model = apply(t1,1,getterms,c(2:14)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
 colnames(tableS1) <- c('Model','df','AICc', 'Delta AICc','Weight')
@@
 #Run the model
-lm1 <- lmer(Loss ~ (Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ + Season) + (1|LayYear),
+lm1 <- lmer(Loss ~ (Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ) + (1|FieldPeriodID),
             data = temp,
@@
 
-tableS2 <- data.frame(Model = apply(t1,1,getterms,c(2:8)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
+tableS2 <- data.frame(Model = apply(t1,1,getterms,c(2:7)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
 colnames(tableS2) <- c('Model','df','AICc', 'Delta AICc','Weight')

### 2016-03-21 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
More analyses

@@
 
-**Figure 1** Telomere length in relation to age in Seychelles warblers.
-**(A)** Mean  ($\pm$ s.e.) age for all birds.
-**(B)** Notched boxplot of telomere length at different age classes within the first year of life.
-Sample sizes for each group are provided in brackets.
+**Figure 1** RTL in relation to **(A)** age and **(B)** $\Delta$Age in Seychelles warblers. Lines and shaded areas are fitted values and 95% confidence limits from linear regressions.
 
-**Figure 2** Factors affecting early-life telomere length in Seychelles warblers.
-**A** Model averaged estimates and 95% confidence intervals for explanatory terms used in a linear mixed model with telomere length (see main text for details; TQ = territory quality, TL = telomere length).
+**Figure 2** Cohort effects on early-life RTL in Seychelles warblers.
+**A** temporal variation RTL in major and minor cohorts, **B** tempoeral variation in adult population size, and **C** early-life RTL in relation to cohort-level variation in adult population size.
+Lines and shaded areas in **C** are fitted values and 95% confidence limits from general linear models.
 
-**Figure 3** Factors affecting early-life telomere shortening rates in Seychelles warblers, within seasons.
-Model averaged estimates and 95% confidence intervals for explanatory terms used in a linear mixed model with telomere length (cross-sectional data, **A**) and telomere loss (longitudinal data, **B**) as the response variable.
-Numbers in brackets are the relative importance of each term in the top model set (see main text for details; TQ = territory quality, TL = telomere length).
+**Figure 3** Factors affecting early-life RTL in Seychelles warblers.
+**A** Model averaged estimates and 95% confidence intervals for explanatory terms used in a linear mixed model with RTL as the response variable (see main text for details; TQ = territory quality).
+**B-D** early life RTL in relation to season (**B**),tarsus length (**C**) and sex (**D**).
+Lines and shaded areas in **D** are fitted values and 95% confidence limits from general linear models.
 
-**Figure 4** Kaplan-Meier curves showing the relationship between survival and telomere length for juvenile Seychelles warblers.
-Telomere length is binned into groups here for visualisation purposes only (long and short = greater than or less than median telomere length, respectively).
+**Figure** Longitudinal telomere dynamics in juvenile Seychelles warblers.
+**A** Juvenile telomere length in relation to adult telomere length within individuals.
+**B** $\Delta$RTL in relation to the time between sampling events.
+Lines and shaded areas represent fitted values and 95% confidence limits from general linear models.
+Raw data have been ommited from **B** for the sake of clarity, but are plotted in Figure S1.
+
+**Figure 5** Factors affecting $\Delta$RTL in juvenile Seychelles warblers.
+Model averaged estimates and 95% confidence intervals for explanatory terms used in a linear mixed model with $\Delta$RTL as the response variable (see main text for details; TQ = territory quality).
 
@@
 temp <- juv
+x <- ifelse(temp$Season == 'Minor',0.5,0)
+temp$LayYear2 <- temp$LayYear+x
 temp$LayYear <- factor(temp$LayYear)
 
-Fig2A <- ggplot(temp,aes(x = LayYear,y = RTL,col=Season,fill=Season))+
+Fig2A <- ggplot(temp,aes(x = LayYear,y = RTL,fill=Season))+
   ylab('RTL') + 
@@
 
-temp <- juv
-x <- ifelse(temp$Season == 'Winter',0.5,0)
-temp$LayYear2 <- temp$LayYear+x
 temp <- temp[!(duplicated(temp$LayYear2)),]
@@
 temp <- juv
+temp[,'Age (months)'] <- factor(temp$Agemonths)
 
@@
   scale_fill_manual(values = c('grey','darkorange','darkgreen'))+
-  annotate('text',x = 240,y = 2.1,label = 'C')
+  annotate('text',x = 250,y = 2.1,label = 'C')
 
@@
     height = 8,
-    paper = 'a4')
+    paper = 'a4r')
 multiplot(Fig2A,Fig2B,Fig2C,layout=matrix(c(1,1,2,3),nrow = 2,byrow=T))
@@
 
-#####################
-#2A
-#####################
-
 #Run the model
-lm2 <- lmer(RTL ~ Agemonths * (Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ + Season)  + (1|LayYear) + (1|TerritoryID),
+lm2 <- lmer(RTL ~ Agemonths * (Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ)  + (1|FieldPeriodID) + (1|TerritoryID)+(1|PlateID),
             data = temp,
@@
 nams[nams == 'z.Tarsus'] <- 'Tarsus'
-nams[nams == 'c.Season'] <- 'Season (winter)'
+nams[nams == 'c.Season'] <- 'Season (minor)'
 nams[nams == 'c.Season:z.Agemonths'] <- 'Age x season'
@@
 Fig3B <- ggplot(temp,
-       aes(x = `Age (months)`,
-           y = RTL,
-           fill=Season)) +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        legend.position = 'top') +
-  geom_jitter(aes(colour = Season),position=position_jitterdodge(dodge.width = 0.8))+
-  geom_boxplot(outlier.colour = NA,notch=F,alpha=0.2)+
-  scale_colour_manual(values = c('grey','darkorange'))+
-  scale_fill_manual(values = c('grey','darkorange'))+
-  annotate('text',x = 0.55,y = 2.1,label = 'B')
-
-Fig3C <- ggplot(temp,
        aes(x = Tarsus,
@@
   stat_smooth(method = 'lm')+
-  scale_colour_manual(values = c(grey(0.2),'darkorange','darkgreen'))+
-  scale_fill_manual(values = c('darkgrey','darkorange','darkgreen'))+
-  annotate('text',x = 17,y = 2.1,label = 'C')
+  scale_colour_manual(values = c('grey','darkorange','darkgreen'))+
+  scale_fill_manual(values = c('grey','darkorange','darkgreen'))+
+  annotate('text',x = 17,y = 2.1,label = 'B')
 
-Fig3D <- ggplot(temp,
+Fig3C <- ggplot(temp,
        aes(x = `Age (months)`,
@@
   scale_fill_manual(values = c('grey','darkorange'))+
-  annotate('text',x = 0.55,y = 2.1,label = 'D')
+  annotate('text',x = 0.55,y = 2.1,label = 'C')
 
 pdf('Figures/Figure 3.pdf',
-    width = 9,
-    height = 9,
+    width = 8,
+    height = 8,
     paper = 'a4r')
-multiplot(Fig3A,Fig3B,Fig3C,Fig3D,layout = matrix(c(1,1,1,2,3,4),nrow = 2,byrow = T))
+multiplot(Fig3A,Fig3B,Fig3C,layout = matrix(c(1,1,2,3),nrow = 2,byrow = T))
 dev.off()
@@
   stat_smooth(method = 'lm')+
-  scale_colour_manual(values = c(grey(0.2),'darkorange','darkgreen'))+
-  scale_fill_manual(values = c('darkgrey','darkorange','darkgreen'))+
+  scale_colour_manual(values = c('grey','darkorange','darkgreen'))+
+  scale_fill_manual(values = c('grey','darkorange','darkgreen'))+
     geom_abline(slope = 0,intercept = 0,lty = 2)+
-  coord_cartesian(ylim = c(-0.5,0.25))+
-  annotate('text',x = 0,y = 0.5,label = 'B')
+  coord_cartesian(ylim = c(-0.5,0.3))+
+  annotate('text',x = 0,y = 0.25,label = 'B')
 
@@
 
-#####################
-#4A
-#####################
 temp <- subset(Loss,Agemonths == 1)
 #Run the model
-lm2 <- lmer(Loss ~ (Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ) + (1|LayYear),
+lm2 <- lmer(Loss ~ (Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ) + (1|FieldPeriodID),
             data = temp,
@@
 nams[nams == 'z.Tarsus'] <- 'Tarsus'
-nams[nams == 'c.Season'] <- 'Season'
 
@@
 
-Fig4 <- forestplot(table2)
+Fig5A <- forestplot(table2)+
+  annotate('text',x = 0.75,y = 0.45,label = 'A')
 
-pdf('Figures/Figure 4.pdf',
-    width = 5,
-    height = 4,
+Fig5B <- ggplot(temp,
+       aes(x = Tarsus,
+           y = Loss,
+           col=Sex)) +
+  ylab(expression(Delta*'RTL')) + 
+  xlab('Tarsus length (mm)') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        legend.position = 'top') +
+  geom_point()+
+  stat_smooth(method = 'lm')+
+  scale_colour_manual(values = c('grey','darkorange','darkgreen'))+
+  scale_fill_manual(values = c('grey','darkorange','darkgreen'))+
+  annotate('text',x = 19,y = 1.1,label = 'B')
+
+pdf('Figures/Figure 5.pdf',
+    width = 4,
+    height = 8,
     paper = 'a4')
-Fig5
+multiplot(Fig5A,Fig5B)
 dev.off()

### 2016-03-21 Scripts/SW juvenile telomeres IPC clean.R

@@ -4,7 +4,7 @@
 #################################################################################
 
 #Only use Ellie's samples
-dd0 <- subset(dd0,Whodunnit == 'EAF')
+#dd0 <- subset(dd0,Whodunnit == 'EAF')
 
 #Average repeats of blood samples
 av <- ave(dd0$RTL,c(dd0$BloodID,dd0$Status,dd0$PlateID))
@@ -30,7 +30,7 @@ dd$DeathDate <- as.Date(dd$DeathDate,"%d/%m/%Y")
 
 
 dd$Season <- ifelse(as.numeric(format(dd$CatchDate,'%m')) %in% c(4:10),
-                    'Summer','Winter')
+                    'Major','Minor')
 
 
 # Age data ----------------------------------------------------------------
@@ -100,11 +100,11 @@ dd$Died <- ifelse(dd$DeathYear<2013,1,0)
 
 
 
-# Exclude winter seasons and early years ----------------------------------
+# Exclude weird seasons and early years ----------------------------------
 
 dd <- subset(dd,LayYear>1997)
-#dd <- subset(dd,Season == 'Summer')
-
+#dd <- subset(dd,Season == 'Major')
+dd <- subset(dd,FieldPeriodID != 27)
 
 # Sex ---------------------------------------------------------------------
 
@@ -226,9 +226,9 @@ for(i in 1:nrow(juv))
 
 juv$HatchDate <- as.Date(juv$HatchDate,"%d/%m/%Y")
 juv$Chickage <- as.numeric(juv$CatchDate-juv$HatchDate)
+juv$O_HatchDate <- as.POSIXlt(juv$HatchDate,format='%yyyy-%mm-%dd')$yday
+juv$O_CatchDate <- as.POSIXlt(juv$CatchDate,format='%yyyy-%mm-%dd')$yday
 juv <- subset(juv,!(is.na(Density)))
-
-# Subset Fledglings and subadults -----------------------------------------------------------
 
 
 
@@ -239,61 +239,39 @@ juv <- subset(juv,!is.na(TQ))
 juv <- subset(juv,Status!='XF')
 
 
-#Subsetting
-FlSA <- subset(juv,Ageclass!='CH')
-chicks <- subset(juv,Ageclass == 'CH')
-
-juvall <- droplevels(juv[complete.cases(juv),])
-FlSAall <- droplevels(FlSA[complete.cases(FlSA),])
[…]

### 2016-03-21 Manuscript/SW_early_life_telomeres.Rmd
Removed RTL methods and cited Kat. Other general edits before sending to Dave

@@
 
-**Running head:** Early-life telomeres and survival
+**Running head:** Early-life telomeres
 
@@
 We found that at the population level, telomere length decreases with cross-sectional and longitudinal measures of age, and that telomere length decreases most rapidly early in life.
-However, at the individual level we found only a very weak relationship between juvenile and adult telomere length, and that significant reductions in telomere length can only be detected within the first few months of life - after this, longitudinal changes in telomere length were not significantly different from zero.
+However, at the individual level we found no relationship between juvenile and adult telomere length, and that significant reductions in telomere length can only be detected within the first few months of life - after this, longitudinal changes in telomere length were not significantly different from zero.
 
 4.
-Juvenile telomere length varied markedly among cohorts, and birds born in years with higher populaiton density had longer telomeres.
+Juvenile telomere length varied markedly among cohorts, and birds born in years with higher population density had longer telomeres.
 Within cohorts, we found no evidence that the early-life social environment or ecological conditions were related to telomere length or rates of shortening.
@@
 5.
-Our data suggest that telomeres are subject to strong population-level variation, but suggest that the resolution of telomeres as a useful biomarker of individual-level cost is likely to be age-dependent, and in many cases limited.
+Our data suggest that telomeres are subject to strong population-level variation, and that telomeres at a point in time represent a snapshot of a fluid eco-evolutionary process.
+At the individual-level, however, our data suggest that the resolution of telomeres as biomarkers of cost is likely to be age-dependent, and in some cases limited.
 
-**Keywords:** Life-history; Seychelles warbler; Senescence; Survival
+**Keywords:** Life-history; Telomere; Seychelles warbler; Senescence
 
@@
 DNA integrity was further validated by visualization with ethidium bromide after electrophoresis on a 1.2% agarose gel, and all samples with evidence of DNA degradation were re-extracted or excluded.
-
 All DNA extractions that passed the above criteria were diluted to 3.3 ng $\mu$l^-1^ before telomere measurement.
-We measured telomere length for all samples using a quantitative PCR (qPCR) assay of telomeres and a GAPDH control gene, using the molecular methods outlined by Barrett *et al.* [-@Barrett2012], with two amendments.
-A change in batch of SYBR green forced us to raise the annealing temperature of the telomere reaction from 58^o^C to 61^o^C for the majority of samples.
-However, this did not affect final telomere length values (see below).
-Second, to avoid bacth effects, we randomised samples among plates.
-We used the program LinRegPCR [@Ruijter2009] to correct for baseline fluorescence and calculate efficiencies and Cq values for each sample replicate.
-Averaging of technical repeats was carried out using custom-made R scripts (available as supplementary material), excluding samples with Cq values differing by >0.5.
-We then calculated relative telomere length (RTL) for each sample using equation 1 in Pfaffi *et al* [@Pfaffi2001].
-We chose to use RTL rather than continuing with the previously used method for calculating absolute telomere length [@Barrett2012], as i) using RTL enabled us to run more samples per plate (as an oligo standard is not required), ii) the RTL method was less susceptible to batch effects (Appendix 1), and iii) very few other studies have calculated absolute telomere length, and our experience suggests that cross-species comparisons are unlikely to be reliable.
-Inter-plate repeatability of final telomere lengths was assessed using the R package rptR [@Schielzeth2011], and for all subsequent analyses we used mean telomere length per sample where we had repeats.
+We measured relative telomere length (RTL) for all samples using a quantitative PCR (qPCR) assay of telomeres and a GAPDH control gene, using the molecular methods outlined by Bebbington *et al.* (In Press).
 
 For a subset of first-year birds for which we had longitudinal data, with an additional sample taken as an adult.
-For these individuals we calculated the within-individual change in RTL by subtracting adult RTL from early-life RTL (hereafter $\Delta$RTL).
+For these individuals we calculated the within-individual change in RTL by subtracting early-life RTL from adult RTL (hereafter $\Delta$RTL).
 Negative values of $\Delta$RTL reflect telomere shortening, while positive values reflect telomere lengthening.
@@
 We performed all analyses using R version 3.0.1 [@RDevelopmentCoreTeam2011].
-Because we are using a different measure of telomere length to the previous study of adult telomere dynamics in this species [@Barrett2013], as well as including many additional samples, we first re-tested, using all available samples, whether telomere length decreases with age and $\Delta$age [a longitudinal measure based on within-subsject centring; @VandePol2009] in the Seychelles warbler, using linear mixed models, with individual ID as a random effect.
+Because we are using a different measure of telomere length to the previous study of adult telomere dynamics in this species [@Barrett2013a], as well as including many additional samples, we first re-tested, using all available samples, whether telomere length decreases with age and $\Delta$age [a longitudinal measure based on within-subject centring; @VandePol2009] in the Seychelles warbler, using linear mixed models with individual ID as a random effect.
 For the analysis of $\Delta$age we we only included birds who had been sampled at least twice.
 We then explored how telomere length varied within the first year of life (telomere length vs age in months) and among cohorts (telomere length vs cohort ID), again using linear mixed models with individual ID as a random effect.
+We tested how temporal variation in telomere length varied between birds born in main and minor breeding seasons, and with temporal variation in population density (measured as the number of adult birds on the island at the end of each breeding season) and food availability, using linear mixed models with cohort ID as a random effect.
 
@@
 Model averaging was carried out using the MuMIn package (version 1.10.5) in R [@Barton2012].
-We first created a full model containing the following explanatory variables: age (in months), tarsus length, sex, territory quality, season (major or minor) and the number of helping and non-helping subordinate birds present in the natal territory.
+We first created a full model containing the following explanatory variables: age (in months), tarsus length, sex, territory quality, season (main or minor) and the number of helping and non-helping subordinate birds present in the natal territory.
 We also included interaction terms between age and all the other variables as telomere dynamics are expected to vary within the first year of life [@Heidinger2012].
@@
 Of these, `r nrow(juv)` samples were taken cross-sectionally from birds caught within their first year of life, and of these juvenile birds, we had additional longitudinal samples for `r nrow(Loss)`.
-Efficiencies (mean $\pm$) for our telomere and GAPDH reactions were (`r round2(mean(dd$EffTelomere))` $\pm$ `r round2(sd(dd$EffTelomere))`) and (`r round2(mean(dd$EffGAPDH))` $\pm$ `r round2(sd(dd$EffGAPDH))`) respectively.
+Efficiencies (mean $\pm$ s.d.) for our telomere and GAPDH reactions were (`r round2(mean(dd$EffTelomere))` $\pm$ `r round2(sd(dd$EffTelomere))`) and (`r round2(mean(dd$EffGAPDH))` $\pm$ `r round2(sd(dd$EffGAPDH))`) respectively.
 Inter-plate repeatability of RTL, based on 392 samples measured at least twice, was 0.78 (CI = 0.73-0.82).
@@
 with $\Delta$age; t = `r round2(summary(lme(RTL~DeltaAge,data=subset(dd,DeltaAge!=0),random=~1|BirdID))$tTable[2,4])`; P `r round3(summary(lme(RTL~DeltaAge,data=subset(dd,DeltaAge!=0),random=~1|BirdID))$tTable[2,5])`; Fig.
-1B).
-RTL also decreased with age within the first year of life (i.e.
+1B), and with age within the first year of life (i.e.
 excluding adult samples; t = `r round2(summary(lme(RTL~Agemonths,data=juv,random=~1|BirdID))$tTable[2,4])`; P `r round3(summary(lme(RTL~Agemonths,data=juv,random=~1|BirdID))$tTable[2,5])`).
@@
 Cohort-level variation in juvenile RTL was positively related to adult population density (LMM controlling for age and cohort: t = `r round2(summary(lme(RTL~Agemonths+Density,data=juv,random=~1|FieldPeriodID))$tTable[3,4])`; P `r round3(summary(lme(RTL~Agemonths+Density,data=juv,random=~1|FieldPeriodID))$tTable[3,5])`; Figs 2B,2C).
-Variation in cohort-level RTL was not related temporal variation in  food availability (t = `r round2(summary(lme(RTL~Agemonths+Insect,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,4])`; P `r round3(summary(lme(RTL~Agemonths+Insect,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,5])`).
-RTL did not differ in birds born in minor compared to major seasons (t = `r round2(summary(lme(RTL~Agemonths+Season,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,4])`; P `r round3(summary(lme(RTL~Agemonths+Season,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,5])`; Fig.
+Variation in cohort-level RTL was not related temporal variation in  food availability (t = `r round2(summary(lme(RTL~Agemonths+Insect,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,4])`; P `r round3(summary(lme(RTL~Agemonths+Insect,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,5])`), and did not differ between birds born in minor compared to main breeding seasons (t = `r round2(summary(lme(RTL~Agemonths+Season,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,4])`; P `r round3(summary(lme(RTL~Agemonths+Season,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,5])`; Fig.
 2A, 2C).
@@
 #####Early-life telomere shortening
-Longitudinal data showed that early-life RTL was positively, but weakly, related to adult RTL (R^2^ = `r round3(summary(lm(RTL1~RTL+Agemonths,data=Loss))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data=Loss))$coefficients[2,3])`; P = `r round3(summary(lm(RTL1~RTL+Agemonths,data = Loss))$coefficients[2,4])`; Fig.
+Longitudinal data showed that early-life RTL was not related to adult RTL (R^2^ = `r round3(summary(lm(RTL1~RTL+Agemonths,data=Loss))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data=Loss))$coefficients[2,3])`; P = `r round3(summary(lm(RTL1~RTL+Agemonths,data = Loss))$coefficients[2,4])`; Fig.
 4A).
@@
 S1).
-In both fledglings and subadults, $\Delta$RTL did not decrease below zero until ~3 years between sampling events was allowed, and confidence limits overlapped zero regardless of the amount of time between sampling events (Fig 3B).
+In both fledglings and subadults, $\Delta$RTL did not decrease below zero until ~3 years between sampling events was allowed, and confidence limits overlapped zero regardless of the amount of time between sampling events (Fig 4B).
 Together, these results suggest that the majority of observable telomere shortening in Seychelles warblers occurs in the first six months of life.
-We therefore restricted the remaining analyses of $\Delta$RTL to birds sampled as nestlings (N = `r nrow(subset(Loss,Agemonths == 1))`.
+We therefore restricted the remaining analyses of $\Delta$RTL to birds sampled as nestlings (N = `r nrow(subset(Loss,Agemonths == 1))`).
 
 $\Delta$RTL did not differ significantly among cohorts of nestlings (F = `r round2(anova(lm(Loss~factor(FieldPeriodID),data = subset(Loss,Agemonths == 1)))[1,4])`; P `r round3(anova(lm(Loss~factor(FieldPeriodID),data = subset(Loss,Agemonths == 1)))[1,5])`), although power for this analysis was limited as longitudinal sample sizes within cohorts were low.
-Model selection revealed that the top model explaining $\Delta$RTL in nestlings contained sex and tarsus length (Table S2).
+Model selection revealed that the top model explaining $\Delta$RTL within cohorts of nestlings contained sex and tarsus length (Table S2).
 This model was a better fit than the null model ($\Delta$AICc = `r round2(AICc(lmer(Loss~1+(1|FieldPeriodID)+(1|TerritoryID),data=subset(Loss,Agemonths == 1),REML = FALSE))-AICc(lmer(Loss~Sex+Tarsus+(1|FieldPeriodID)+(1|TerritoryID),data=subset(Loss,Agemonths == 1),REML = FALSE)))`), although confidence limits for both the sex and tarsus effects overlapped zero (Fig.
 5A).
-$\Delta$RTL was negatively related to tarsus length, and higher in females, indicating higher rates of telomere loss in smaller nestlings, and in males.
+$\Delta$RTL was negatively related to tarsus length, and higher in females, indicating greater telomere shortening in smaller nestlings, and in males (Fig.
+5B).
 
@@
 
-Early-life RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,4])`), with no interaction effect between RTL and juvenile age on survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,4])`).
+Early-life RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,4])`), with no interaction effect between RTL and juvenile age (estimate = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,4])`), or sex (estimate = `r round2(summary(glmer(SurvivedNext~RTL*Sex+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[5,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL*Sex+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[5,4])`).
 To test whether cohort-level effects on RTL influenced survival we calculated cohort-level RTL (i.e.
-the median RTL for each cohort) and added this term to a GLMM, and found that cofort-level RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~cohortTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~cohortTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,4])`), with no interaction with juvenile age (estimate = `r round2(summary(glmer(SurvivedNext~cohortTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~cohortTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,4])`).
+the median RTL for each cohort) and added this term to a GLMM, and found that cohort-level RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~cohortTL+RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~cohortTL+RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,4])`).
 Using the longitudinal data, we found that $\Delta$RTL had no effect on survival to the year after the second sampling event (estimate = `r round2(summary(glm(SurvivedNext2~Loss,data=subset(subset(Loss,Agemonths == 1),FieldPeriodID<2013),family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glm(SurvivedNext2~RTL+Agemonths,data=subset(subset(Loss,Agemonths == 1),LayYear<2013),family = 'binomial'))$coefficients[2,4])`).
 
-Cox regression showed no effect of individual-level or cohort-level RTL on survival later in life (individual-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL+Agemonths+frailty(LayYear))))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL+Agemonths+frailty(LayYear))))$coefficients[1,6])`; cohort-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL+Agemonths+frailty(LayYear))))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL+Agemonths+frailty(LayYear))))$coefficients[1,6])`), and no interaction between RTL and age on survival (individual-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL*Agemonths+frailty(LayYear))))$coefficients[4,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL*Agemonths+frailty(LayYear))))$coefficients[4,6])`; cohort-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL*Agemonths+frailty(LayYear))))$coefficients[4,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL*Agemonths+frailty(LayYear))))$coefficients[4,6])`).
+Cox regression showed no effect of individual-level or cohort-level RTL on survival later in life (individual-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL+Agemonths+frailty(FieldPeriodID))))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL+Agemonths+frailty(FieldPeriodID))))$coefficients[1,6])`; cohort-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL*Agemonths+RTL+frailty(FieldPeriodID))))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL*Agemonths+RTL+frailty(FieldPeriodID))))$coefficients[1,6])`), and no interaction between RTL and age on survival (individual-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL*Agemonths+frailty(FieldPeriodID))))$coefficients[4,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL*Agemonths+frailty(FieldPeriodID))))$coefficients[4,6])`).
 $\Delta$RTL was not related to longer-term survival (estimate `r round3(summary(coxph(with(subset(subset(Loss,LayYear<2008),Agemonths == 1),Surv(RemainingLife2+1,Died)~Loss)))$coefficients[1,1])`, P = `r round3(summary(coxph(with(subset(subset(Loss,LayYear<2008),Agemonths ==1 ),Surv(RemainingLife2+1,Died)~Loss)))$coefficients[1,5])`).
@@
 Our study adds the the now substantial body of literature showing that telomere length decreases with age, and that this decrease is most rapid in early life [e.g.
-@Frenck1998; @Haussmann2003; @Heidinger2012; @Fairlie2015].
+@Frenck1998; @Haussmann2003; @Heidinger2012; @Fairlie2016].
 Our finding that tarsus length was related to juvenile RTL in an age-dependent manner (Fig.
@@
 Telomere loss is most rapid early in life due to ongoing growth [@Frenck1998], and a negative correlation between RTL and body size in early life is therefore expected [see also @Ringsby2015].
-However, despite our clear finding that population-level RTL decreases with age throughout the warbler lifespan (Fig.
-1), we found that at the individual-level, juvenile and adult telomere length were weakly correlated, and that telomere shortening can only be detected very early in life (Fig.
+We also found weak evidence for sex-specific telomere dynamics in Seychelles warblers, with males having longer-telomeres and higher loss of telomeres (Fig.
+3C, Fig.
+5B).
+We do not wish to speculate too much on this effect as confidence intervals for both sex effects overlapped zero.
+However, it is worth noting while we found no evidence that juvenile RTL has a sex-specific effect on survival, in adults the effect of telomere length on survival is strongest for males [@Barrett2013a].
+Together, these results suggest that sex effects should at least be considered when studying any aspect of telomere dynamics [see also @Barrett2011].
+
+Despite our clear finding that RTL decreases with age throughout the Seychelles warbler lifespan (Fig.
+1), we found that at the individual-level, juvenile and adult telomere were not correlated, and that telomere shortening can only be detected very early in life (Fig.
 4).
-This is in contrast to other avian studies where birds were reared in laboratory [@Heidinger2012] or nestbox [@Boonekamp2014] conditions, in which juvenile and adult telomere length were highly correlated, and individual-level telomere shortening clearly occurs throughout the juvenile period, and into adulthood.
-However, in accordance with our study, a recent study on Soay sheep (*Ovis aries*) found a weak correlation between juvenile and adult RTL, and a highly complex pattern of telomere dynamics across the entire lifespan [@Fairlie2015].
-Indeed, the complex relationship between RTL and age in our study, with periods of early-life telomere shortening followed by selective disappearance and/or apparent telomere lengthening as individuals age is strikingly similar to that reported by Fairlie *et al.* [-@Fairlie2015].
-It is too soon to speculate as to what may be generating this variation among species, but one possibility is that environmental variability itself affects the resolution with which we can observe telomere shortening in natural populations [see also @Fairlie 2015].
-As early-life telomere dynamics are investigated in more systems, the causes of this variation should become clear.
+This is in contrast to other avian studies where birds were reared in laboratory [@Heidinger2012] or nest-box [@Boonekamp2014] conditions, in which juvenile and adult telomere length were highly correlated, and individual-level telomere shortening clearly occurs throughout the juvenile period, and into adulthood.
+However, in accordance with our study, a recent study on Soay sheep (*Ovis aries*) found only a very weak correlation between juvenile and adult RTL, and a highly complex pattern of telomere dynamics across the lifespan [@Fairlie2016].
+It is too soon to speculate as to what may be generating this variation among species, but one possibility is that environmental variability itself affects the resolution with which we can observe telomere shortening in natural populations [see also @Fairlie2016].
+Regardless, our data suggest the efficacy of telomeres as individual biomarkers of cost in wild populations will depend on the age structure of the study population, as well as the sampling resolution.
 
 Perhaps the clearest result from our study is that RTL varies among cohorts.
-Very few studies have shown that temporal variation in telomere dynamics occurs in natural populations, and these have been limited in the number of seasons [@Mizutani2013; @Watson2015; @Fairlie2015].
+A few studies have now shown that temporal variation in telomere dynamics occurs in natural populations, although these have been limited in the number of seasons [@Mizutani2013; @Watson2015; @Fairlie2016].
 The long-term Seychelles warbler dataset has allowed us to show that temporal variation in telomere dynamics can occur over substantial time periods.
-Moroevoer, becuase of the larger number of cohorts we have been able to show statistically that that temporal variation in RTL is related to variation in the environment (Fig.
+Moreover, because of the large number of cohorts we have been able to show statistically that that temporal variation in RTL is related to variation in the environment (Fig.
 1C).
-We may expect a negative relationship between population density and RTL, as increased competition for resources is known to effect juvenile telomere length.
-However, We suggest that population density acts as an indicator of overall island 'quality' for Seychelles warblers.
+In some systems a negative relationship between population density and RTL may be expected, as increased competition for resources is known to effect juvenile telomere length.
+However, in the Seychelles warbler it is likely that population density acts as an indicator of overall island 'quality' for Seychelles warblers.
 This can clearly be seen from data from the 1980s and 1990s, when adult population size on Cousin increased as habitat quality improved [@Komdeur2005].
 Population size has fluctuated around 300 since the late 1990s (Fig.
-2B), and captures this more accurately than measures of insect abundance or territory quality.
+2B), and we suggest that this variation captures this island-wide conditions more accurately than measures of insect abundance or territory quality.
+If cohort-level variation in telomere dynamics continues to be found in other systems, this has potential consequences for our understanding of telomere dynamics in natural populations.
+In particular, our findings suggest that the telomere dynamics of a population at a given point in time represent a snapshot of a temporally varying process.
+More research is now needed within and across multiple cohorts and populations to better understand how how and why population-level telomere dynamics vary over space and time.
 
-Within cohorts, we found no evidence that any of the social or environmental variables we measured affect juvenile RTL.
-This is surprising, as a poor social and ecological environment is known to be detrimental to juvenile Seychelles warblers, both in terms of oxidative stress and later life survival.
-Possible explanations for this finding include i) temporal variation in RTL in our data masks our ability to detect spatial trends, ii) within-cohort variation in RTL is explained by a variable that we have not included in our analyses, or iii) that effects are generally weak and levels of noise in our telomere measurements and/or ecological data preclude detection of significant effects.
-The first of these explanations seems the least likely, as within-cohort variation in RTL greatly exceeds among-cohort variation (Fig.
+Despite substantial cohort-level variation in juvenile RTL, within cohorts we found no evidence that any of the social or environmental variables we measured affect juvenile RTL.
+This is surprising, as a poor social and ecological environment is known to be detrimental to juvenile Seychelles warblers, both in terms of oxidative stress and later life survival [@VandeCrommenacker2011a; @Brouwer2012].
+A possible explanation for this could be that temporal variation in RTL in our data masks our ability to detect spatial trends.
+However, this is unlikely we controlled for cohort ID, and within-cohort variation in RTL greatly exceeds among-cohort variation (Fig.
 1C).
+A second possibility is that effects are generally weak and levels of noise in our telomere measurements and/or ecological data preclude detection of significant effects.
+While noise in telomere measurements and problems with power are a problem in any study using qPCR to measure telomere length [@Nussey2014], our sample sizes and levels of repeatability compare favourably with other studies of natural populations that have found significant effects of the early-life environment on telomere length [e.g.
+@Fairlie2015].
+More likely is that early-life RTL in the Seychelles warbler is explained by a set of variables that we have not measured.
+A key question to be addressed is the extent to which early-life RTL reflects inheritance [e.g.
+@Becker2015; @Asghar2014].
+This is important, as inheritance may capture a large amount of unexplained variation in early-life RTL within our dataset.
+Future research on the Seychelles warbler will examine, in a quantitative genetic framework, how genetic and environmental components, and their interactions, affect telomere dynamics and senescence.
 
@@
 In captive zebra finches, juvenile telomere length predicts late-life survival [@Heidinger2012], and in wild bird populations survival to the nestling phase [@Watson2015], and survival to adulthood [@Boonekamp2014], have been linked with early-life telomere dynamics.
-
-There are numerous avenues for future research into telomere evolutionary ecology in this system and others.
-If our findings are replicated in other systems, and population-level variation in early-life telomere dynamics is common in nature, this has potential consequences for our understanding of telomere dynamics in natural populations.
-In particular, our findings suggest that the telomere dynamics of a population at a given point in time represent a snapshot of a temporally varying process.
-More research is now needed within and across multiple cohorts and populations to better understand how how and why population-level telomere dynamics vary over space and time.
-Another question to be addressed is the extent to which early-life RTL reflects inheritance [e.g.
-@Becker2015; @Asghar2014].
-This is important, as inheritance may capture a large amount of unexplained variation in early-life RTL within our dataset.
-Future research on the Seychelles warbler will examine, in a quantitative genetic framework, how genetic and environmental components, and their interactions, affect telomere dynamics and senescence in natural populations.
-FInally, here we have considered survival, but telomere length and shortening in early life may also be linked to other components of fitness [reviewed in @Monaghan2014].
+It is worth noting that here we have considered survival, but telomere length and shortening in early life may also be linked to other components of fitness [reviewed in @Monaghan2014].
 Reproductive senescence occurs in the Seychelles warbler [reviewed in @Hammers2015], making this system well suited to examining how telomere length predicts lifetime reproductive success.

### 2016-03-21 Manuscript/SW_early_life_telomeres_SOM.Rmd
Removed RTL methods and cited Kat. Other general edits before sending to Dave

@@
                fig.path='../Figures/',
-               fig.width= 4,
-               fig.height= 8,
+               fig.width= 5,
+               fig.height= 5,
                fig.show='asis',
-               dev=c('png','pdf'))
+               dev=c('png'))
 
@@
 
-**Table S2** Model selection table for analysis of factors affecting early life telomere loss in Seychelles warbler chicks, using longitudinal data.
+**Table S2** Model selection table for analysis of factors affecting $\Delta$RTL in Seychelles warbler nestlings, using longitudinal data.
 Only models that remained in the top model set for model averaging (AICc $\leq$ 6 compared to the top model) are included here.
@@
 ```
+
+**Figure S1** $\Delta$RTL in relation to the time between sampling events.
+Lines and shaded areas represent fitted values and 95% confidence limits from general linear models.
+See also Figure 4, main text.
+
+```{r Figure S1}
+
+temp <- Loss
+temp[,'Age (months)'] <- factor(temp$Agemonths)
+temp$LayYear <- factor(temp$LayYear)
+temp$TimeDiff <- temp$TimeDiff/365
+
+FigS1 <- ggplot(temp,
+       aes(x = TimeDiff,
+           y = Loss,
+           colour=`Age (months)`,
+           fill = `Age (months)`)) +
+  ylab(expression(Delta*'RTL')) + 
+  xlab('Time (years)') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        legend.position = 'top') +
+  geom_point()+
+  stat_smooth(method = 'lm')+
+  scale_colour_manual(values = c('grey','darkorange','darkgreen'))+
+  scale_fill_manual(values = c('grey','darkorange','darkgreen'))+
+    geom_abline(slope = 0,intercept = 0,lty = 2)+
+  annotate('text',x = 0,y = 1.75,label = 'B')
+
+FigS1
+
+```

### 2016-03-21 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Removed RTL methods and cited Kat. Other general edits before sending to Dave

@@
 **Figure 2** Cohort effects on early-life RTL in Seychelles warblers.
-**A** temporal variation RTL in major and minor cohorts, **B** tempoeral variation in adult population size, and **C** early-life RTL in relation to cohort-level variation in adult population size.
+**A** temporal variation RTL in major and minor cohorts, **B** temporal variation in adult population size, and **C** early-life RTL in relation to cohort-level variation in adult population size.
 Lines and shaded areas in **C** are fitted values and 95% confidence limits from general linear models.
@@
 **A** Model averaged estimates and 95% confidence intervals for explanatory terms used in a linear mixed model with RTL as the response variable (see main text for details; TQ = territory quality).
-**B-D** early life RTL in relation to season (**B**),tarsus length (**C**) and sex (**D**).
-Lines and shaded areas in **D** are fitted values and 95% confidence limits from general linear models.
+**B,C** Early-life RTL and age in relation to tarsus length (**B**) and sex (**C**).
+Lines and shaded areas in **B** are fitted values and 95% confidence limits from a general linear model.
 
-**Figure** Longitudinal telomere dynamics in juvenile Seychelles warblers.
-**A** Juvenile telomere length in relation to adult telomere length within individuals.
+**Figure 4** Longitudinal telomere dynamics in juvenile Seychelles warblers.
+**A** Juvenile telomere length in relation to adult telomere length measured in the same individual.
 **B** $\Delta$RTL in relation to the time between sampling events.
@@
 **Figure 5** Factors affecting $\Delta$RTL in juvenile Seychelles warblers.
-Model averaged estimates and 95% confidence intervals for explanatory terms used in a linear mixed model with $\Delta$RTL as the response variable (see main text for details; TQ = territory quality).
+**A** Model averaged estimates and 95% confidence intervals for explanatory terms used in a linear mixed model with $\Delta$RTL as the response variable (see main text for details; TQ = territory quality).
+**B** $\Delta$RTL in relation to tarsus length and sex.
 
@@
        aes(x = RTL,
-           y = RTL1)) +
+           y = RTL1,
+           col=Ageclass)) +
   ylab('Adult RTL') + 
@@
 
+ggplot(juvseason9,aes(x = RTL,y = Lifespan,col = Season))+
+  geom_point()+
+  stat_smooth(method = 'lm')
+
 ```

### 2016-03-21 Scripts/SW juvenile telomeres IPC load.r

@@ -16,6 +16,7 @@ library(ggplot2)
 library(gvlma)
 library(survival)
 library(nlme)
+library(magrittr)
 
 # Load data ---------------------------------------------------------------
 dd0 <- read.csv('Data/REL_TL_EB_EAF_25 Feb FULL DATA.csv') #main dataset

### 2016-04-14 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Winter excluded!

@@
 
-temp <- juv
-
 #Run the model
-lm2 <- lmer(RTL ~ Agemonths * (Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ)  + (1|FieldPeriodID) + (1|TerritoryID)+(1|PlateID),
+lm2 <- lmer(RTL ~ Agemonths*(Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ)  + (1|FieldPeriodID) + (1|TerritoryID)+(1|PlateID),
             data = temp,
@@
         legend.position = 'top') +
-  geom_point(col='grey')+
-  stat_smooth(method = 'lm',col='darkgrey')+
+  geom_point()+
+  stat_smooth(method = 'lm')+
   annotate('text',x = 0,y = 2.1,label = 'A')
@@
 Fig4B <- ggplot(temp2,
-       aes(x = TimeDiff,
-           y = Loss,
-           colour=`Age (months)`,
-           fill = `Age (months)`)) +
+       aes(x = Ageclass,
+           y = Loss)) +
   ylab(expression(Delta*'RTL')) + 
@@
         legend.position = 'top') +
-  geom_point(alpha=0)+
-  stat_smooth(method = 'lm')+
-  scale_colour_manual(values = c('grey','darkorange','darkgreen'))+
-  scale_fill_manual(values = c('grey','darkorange','darkgreen'))+
-    geom_abline(slope = 0,intercept = 0,lty = 2)+
-  coord_cartesian(ylim = c(-0.5,0.3))+
-  annotate('text',x = 0,y = 0.25,label = 'B')
+  geom_boxplot(outlier.shape = NA)+
+  geom_jitter(col = 'grey')+
+  annotate('text',x = 0.5,y = 1.5,label = 'B')+
+  geom_abline(slope = 0,intercept = 0,lty = 2)
 
@@
 
-temp <- subset(Loss,Agemonths == 1)
+temp <- Loss
 #Run the model
-lm2 <- lmer(Loss ~ (Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ) + (1|FieldPeriodID),
+lm2 <- lmer(Loss ~ Agemonths*(Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ) + (1|FieldPeriodID),
             data = temp,
@@
 
+boxplot(Loss~Helper*Ageclass,data=Loss)
+
+temp <- Loss
+
 Fig5B <- ggplot(temp,
-       aes(x = Tarsus,
+       aes(col = factor(Helper),
            y = Loss,
-           col=Sex)) +
+           x=Ageclass)) +
   ylab(expression(Delta*'RTL')) + 
@@
         legend.position = 'top') +
-  geom_point()+
-  stat_smooth(method = 'lm')+
+  geom_boxplot(notch=T)+
   scale_colour_manual(values = c('grey','darkorange','darkgreen'))+
-  scale_fill_manual(values = c('grey','darkorange','darkgreen'))+
-  annotate('text',x = 19,y = 1.1,label = 'B')
+  scale_fill_manual(values = c('grey','darkorange','darkgreen'))
 

### 2016-04-14 Scripts/SW juvenile telomeres IPC clean.R

@@ -31,6 +31,7 @@ dd$DeathDate <- as.Date(dd$DeathDate,"%d/%m/%Y")
 
 dd$Season <- ifelse(as.numeric(format(dd$CatchDate,'%m')) %in% c(4:10),
                     'Major','Minor')
+dd <- subset(dd,Season == 'Major')
 
 
 # Age data ----------------------------------------------------------------
@@ -252,7 +253,7 @@ subset(!(duplicated(BirdID)))
 
 #Get earliest subsetquent catch
 laters <- merge(aggregate(CatchDate~BirdID,subset(dd,!(BloodID %in% earlies$BloodID)),
-                          min),
+                          max),
                 subset(dd,!(BloodID %in% earlies$BloodID))) %>% 
 subset(!(duplicated(BirdID)))
 

### 2016-04-14 Manuscript/SW_early_life_telomeres.Rmd
More post-Dave edits

@@
 
-To assess our resolution for detecting telomere shortening, we tested, for each age class, whether $\Delta$RTL was significantly lower than zero, using one-sample t-tests.
-We then tested how $\Delta$RTL was related to the time interval between sampling events, using linear models.
-A significantly negative slope indicates that telomere shortening occurs linearly over time, while the time point at which $\Delta$RTL is significantly lower than zero can be considered the time period over which we have the resolution to detect reductions in telomere length.
-
 We used general linear mixed models with model averaging to explore how spatial variation in early-life environmental and social conditions influenced telomere length within cohorts.
@@
 Cohort-level variation in juvenile RTL was positively related to adult population density (LMM controlling for age and cohort: t = `r round2(summary(lme(RTL~Agemonths+Density,data=juv,random=~1|FieldPeriodID))$tTable[3,4])`; P `r round3(summary(lme(RTL~Agemonths+Density,data=juv,random=~1|FieldPeriodID))$tTable[3,5])`; Figs 2B,2C).
-Variation in cohort-level RTL was not related temporal variation in  food availability (t = `r round2(summary(lme(RTL~Agemonths+Insect,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,4])`; P `r round3(summary(lme(RTL~Agemonths+Insect,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,5])`), and did not differ between birds born in minor compared to main breeding seasons (t = `r round2(summary(lme(RTL~Agemonths+Season,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,4])`; P `r round3(summary(lme(RTL~Agemonths+Season,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,5])`; Fig.
-2A, 2C).
+Variation in cohort-level RTL was not related temporal variation in  food availability (t = `r round2(summary(lme(RTL~Agemonths+Insect,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,4])`; P `r round3(summary(lme(RTL~Agemonths+Insect,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,5])`).
 
@@
 $\Delta$RTL was significantly lower than zero in nestlings (t = `r round2(t.test(subset(Loss,Agemonths == 1)$Loss)$statistic)`; P `r round3(t.test(subset(Loss,Agemonths == 1)$Loss)$p.value)`), but not in fledglings (t = `r round2(t.test(subset(Loss,Agemonths == 6)$Loss)$statistic)`; P = `r round2(t.test(subset(Loss,Agemonths == 6)$Loss)$p.value)`) or subadults (t = `r round2(t.test(subset(Loss,Agemonths == 10)$Loss)$statistic)`; P = `r round2(t.test(subset(Loss,Agemonths == 10)$Loss)$p.value)`).
-In line with this, $\Delta$RTL increased with age in early life (R^2^ = `r round2(summary(lm(Loss~Agemonths,data=Loss))$r.squared)`; t = `r round2(summary(lm(Loss~Agemonths,data=Loss))$coefficients[2,3])`; P `r round3(summary(lm(Loss~Agemonths,data = Loss))$coefficients[2,4])`), suggesting that telomere shortening was highest shortly after the nestling phase.
-In nestlings, $\Delta$RTL  did not vary with the time between sampling events (R^2^ `r round3(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths ==1)))$r.squared)`; t = `r round2(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths ==1)))$coefficients[2,3])`; P = `r round3(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths ==1)))$coefficients[2,4])`).
-However, in fledglings and subadults $\Delta$RTL decreased with the length of time between sampling events (R^2^ = `r round2(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths !=1)))$r.squared)`; t = `r round2(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths !=1)))$coefficients[2,3])`; P = `r round3(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths !=1)))$coefficients[2,4])`; Fig.
-4B, Fig.
-S1).
-In both fledglings and subadults, $\Delta$RTL did not decrease below zero until ~3 years between sampling events was allowed, and confidence limits overlapped zero regardless of the amount of time between sampling events (Fig 4B).
-Together, these results suggest that the majority of observable telomere shortening in Seychelles warblers occurs in the first six months of life.
-We therefore restricted the remaining analyses of $\Delta$RTL to birds sampled as nestlings (N = `r nrow(subset(Loss,Agemonths == 1))`).
+In line with this, $\Delta$RTL increased with age in early life (R^2^ = `r round2(summary(lm(Loss~Agemonths,data=Loss))$r.squared)`; t = `r round2(summary(lm(Loss~Agemonths,data=Loss))$coefficients[2,3])`; P `r round3(summary(lm(Loss~Agemonths,data = Loss))$coefficients[2,4])`; Fig.
+4B), suggesting that telomere shortening was highest shortly after the nestling phase.
 
@@
 
-Early-life RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,4])`), with no interaction effect between RTL and juvenile age (estimate = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,4])`), or sex (estimate = `r round2(summary(glmer(SurvivedNext~RTL*Sex+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[5,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL*Sex+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[5,4])`).
+Early-life RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,4])`), with no interaction effect between RTL and juvenile age (estimate = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,4])`).
 To test whether cohort-level effects on RTL influenced survival we calculated cohort-level RTL (i.e.
 the median RTL for each cohort) and added this term to a GLMM, and found that cohort-level RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~cohortTL+RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~cohortTL+RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,4])`).
-Using the longitudinal data, we found that $\Delta$RTL had no effect on survival to the year after the second sampling event (estimate = `r round2(summary(glm(SurvivedNext2~Loss,data=subset(subset(Loss,Agemonths == 1),FieldPeriodID<2013),family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glm(SurvivedNext2~RTL+Agemonths,data=subset(subset(Loss,Agemonths == 1),LayYear<2013),family = 'binomial'))$coefficients[2,4])`).
+Using the longitudinal data, we found that $\Delta$RTL had no effect on survival to the year after the second sampling event (estimate = `r round2(summary(glm(SurvivedNext2~Loss+Agemonths,data = subset(Loss,FieldPeriodID<2013),family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glm(SurvivedNext2~Loss+Agemonths,data=subset(Loss,LayYear<2013),family = 'binomial'))$coefficients[2,4])`).
 

### 2016-04-14 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
More post-Dave edits

@@
 temp <- juv
-x <- ifelse(temp$Season == 'Minor',0.5,0)
-temp$LayYear2 <- temp$LayYear+x
 temp$LayYear <- factor(temp$LayYear)
 
-Fig2A <- ggplot(temp,aes(x = LayYear,y = RTL,fill=Season))+
+Fig2A <- ggplot(temp,aes(x = LayYear,y = RTL))+
   ylab('RTL') + 
@@
         legend.position = 'top')+
-  geom_jitter(aes(colour = Season),position=position_jitterdodge(dodge.width = 0.8))+
+  geom_jitter(col = 'grey')+
   geom_boxplot(outlier.colour = NA,notch=F,alpha=0.2)+
@@
 
-temp <- temp[!(duplicated(temp$LayYear2)),]
+temp$LayYear <- as.numeric(paste(temp$LayYear))
 
-Fig2B <- ggplot(temp,aes(x = LayYear2,y = Density))+
+Fig2B <- ggplot(temp,aes(x = LayYear,y = Density))+
   ylab('Adult population size') + 
@@
 
-temp <- juv
-temp[,'Age (months)'] <- factor(temp$Agemonths)
-
-Fig2C <- ggplot(temp,
-       aes(x = Density,
-           y = RTL,
-           colour=Season,
-           fill = Season)) +
-  ylab('RTL') + 
-  xlab('Adult population size') +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        legend.position = 'none') +
-  geom_point()+
-  stat_smooth(method = 'lm')+
-  scale_colour_manual(values = c('grey','darkorange','darkgreen'))+
-  scale_fill_manual(values = c('grey','darkorange','darkgreen'))+
-  annotate('text',x = 250,y = 2.1,label = 'C')
-
 pdf('Figures/Figure 2.pdf',
     width = 8,
-    height = 8,
+    height = 6,
     paper = 'a4r')
-multiplot(Fig2A,Fig2B,Fig2C,layout=matrix(c(1,1,2,3),nrow = 2,byrow=T))
+multiplot(Fig2A,Fig2B,layout=matrix(c(1,2),nrow = 2,byrow=T))
 dev.off()
@@
 ```{r Figure 3}
+
+temp <- juv
 
@@
 Fig3C <- ggplot(temp,
-       aes(x = `Age (months)`,
+       aes(col = `Age (months)`,
+           fill =  `Age (months)`,
            y = RTL,
-           fill=Sex)) +
+           x=TQ)) +
   theme_classic() +
@@
         legend.position = 'top') +
-  geom_jitter(aes(colour = Sex),position=position_jitterdodge(dodge.width = 0.8))+
-  geom_boxplot(outlier.colour = NA,notch=F,alpha=0.2)+
-  scale_colour_manual(values = c('grey','darkorange'))+
-  scale_fill_manual(values = c('grey','darkorange'))+
-  annotate('text',x = 0.55,y = 2.1,label = 'C')
+  geom_point()+
+  stat_smooth(method = 'lm')+
+  scale_colour_manual(values = c('grey','darkorange','darkgreen'))+
+  scale_fill_manual(values = c('grey','darkorange','darkgreen'))+
+  annotate('text',x = 3,y = 2.1,label = 'C')
 
@@
 temp[,'Age (months)'] <- factor(temp$Agemonths)
-temp$LayYear <- factor(temp$LayYear)
 
@@
            y = RTL1,
-           col=Ageclass)) +
+           col=`Age (months)`)) +
   ylab('Adult RTL') + 
@@
   geom_point()+
-  stat_smooth(method = 'lm')+
+  scale_colour_manual(values = c('grey','darkorange','darkgreen'))+
+  scale_fill_manual(values = c('grey','darkorange','darkgreen'))
   annotate('text',x = 0,y = 2.1,label = 'A')
 
-temp2 <- temp
-temp2$TimeDiff <- temp2$TimeDiff/365
-
-Fig4B <- ggplot(temp2,
-       aes(x = Ageclass,
+Fig4B <- ggplot(temp,
+       aes(x = `Age (months)`,
            y = Loss)) +
   ylab(expression(Delta*'RTL')) + 
-  xlab('Time (years)') +
+  xlab('Age (months)') +
   theme_classic() +
@@
 nams[nams == 'z.Tarsus'] <- 'Tarsus'
+nams[nams == 'c.Season'] <- 'Season (minor)'
+nams[nams == 'c.Season:z.Agemonths'] <- 'Age x season'
+nams[nams == 'z.Agemonths:z.NonHelper'] <- 'Age x non helpers'
+nams[nams == 'c.Sex:z.Agemonths'] <- 'Age x sex'
+nams[nams == 'z.Agemonths:z.TQ'] <- 'Age x TQ'
+nams[nams == 'c.Helper:z.Agemonths'] <- 'Age x helpers'
+nams[nams == 'z.Agemonths:z.BodyMass'] <- 'Age x mass'
+nams[nams == 'z.Agemonths:z.Tarsus'] <- 'Age x tarsus'
 
@@
 Fig5A <- forestplot(table2)+
-  annotate('text',x = 0.75,y = 0.45,label = 'A')
-
-boxplot(Loss~Helper*Ageclass,data=Loss)
+  annotate('text',x = 0.75,y = 0.6,label = 'A')
 
 temp <- Loss
+temp[,'Helpers'] <- ifelse(temp$Helper>0,'Present', 'Absent')
+temp[,'Non helpers'] <- ifelse(temp$NonHelper>0,'Present', 'Absent')
+temp[,'Age (months)'] <- factor(temp$Agemonths)
 
 Fig5B <- ggplot(temp,
-       aes(col = factor(Helper),
+       aes(x = `Age (months)`,
            y = Loss,
-           x=Ageclass)) +
-  ylab(expression(Delta*'RTL')) + 
-  xlab('Tarsus length (mm)') +
+           fill=Helpers)) +
   theme_classic() +
+  ylab(expression(Delta*'RTL'))+
   theme(axis.text.x = element_text(size = 12), 
@@
         legend.position = 'top') +
-  geom_boxplot(notch=T)+
-  scale_colour_manual(values = c('grey','darkorange','darkgreen'))+
-  scale_fill_manual(values = c('grey','darkorange','darkgreen'))
+  geom_jitter(aes(colour = Helpers),position=position_jitterdodge(dodge.width = 0.8))+
+  geom_boxplot(outlier.colour = NA,notch=F,alpha=0.2)+
+  scale_colour_manual(values = c('grey','darkorange'))+
+  scale_fill_manual(values = c('grey','darkorange'))+
+  geom_abline(slope = 0,intercept = 0,lty = 2)+
+  annotate('text',x = 0.55,y = 1.5,label = 'C')
 
 pdf('Figures/Figure 5.pdf',
-    width = 4,
+    width = 8,
     height = 8,
@@
 
-ggplot(juvseason9,aes(x = RTL,y = Lifespan,col = Season))+
-  geom_point()+
-  stat_smooth(method = 'lm')
-
 ```

### 2016-04-14 Scripts/SW juvenile telomeres IPC clean.R

@@ -251,9 +251,9 @@ juv$Condition <- lm(BodyMass~Tarsus+Age+Sex,data=juv)$residuals
 earlies <- merge(juv,aggregate(CatchDate~BirdID,juv,min)) %>% 
 subset(!(duplicated(BirdID)))
 
-#Get earliest subsetquent catch
+#Get earliest subsequent catch
 laters <- merge(aggregate(CatchDate~BirdID,subset(dd,!(BloodID %in% earlies$BloodID)),
-                          max),
+                          min),
                 subset(dd,!(BloodID %in% earlies$BloodID))) %>% 
 subset(!(duplicated(BirdID)))
 
@@ -277,6 +277,7 @@ Loss$TROC <- with(Loss,Loss/TimeDiff)
 
 #
 Loss <- subset(Loss,TimeDiff>0)
+Loss <- subset(Loss,TimeDiff<(365*2))
 
 # Get rid of stuff not to be used -----------------------------------------
 

### 2016-04-18 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
DSR edits

@@
 
-Fig2A <- ggplot(temp,aes(x = LayYear,y = RTL))+
+Fig2 <- ggplot(temp,aes(x = LayYear,y = RTL))+
   ylab('RTL') + 
@@
   scale_colour_manual(values = c('grey','darkorange'))+
-  scale_fill_manual(values = c('grey','darkorange'))+
-  annotate('text',x = 1,y = 2.1,label = 'A')
-
-temp$LayYear <- as.numeric(paste(temp$LayYear))
-
-Fig2B <- ggplot(temp,aes(x = LayYear,y = Density))+
-  ylab('Adult population size') + 
-  xlab('Cohort') +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        legend.position = 'none')+
-  geom_line(stat = 'identity')+
-  annotate('text',x = 1998,y = 360,label = 'B')
+  scale_fill_manual(values = c('grey','darkorange'))
 
@@
     width = 8,
-    height = 6,
+    height = 4,
     paper = 'a4r')
-multiplot(Fig2A,Fig2B,layout=matrix(c(1,2),nrow = 2,byrow=T))
+Fig2
 dev.off()
@@
 #Run the model
-lm2 <- lmer(RTL ~ Agemonths*(Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ)  + (1|FieldPeriodID) + (1|TerritoryID)+(1|PlateID),
+lm2 <- lmer(RTL ~ Agemonths*(Tarsus) + Helper + NonHelper + Sex + BodyMass  + TQ  + (1|FieldPeriodID) + (1|TerritoryID)+(1|PlateID),
             data = temp,
@@
 Fig3A <- forestplot(table2)+
-  annotate('text',x = 1,y = 0.25,label = 'A')
+  annotate('text',x = 0.8,y = 0.25,label = 'A')
 
@@
 #Run the model
-lm2 <- lmer(Loss ~ Agemonths*(Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ) + (1|FieldPeriodID),
+lm2 <- lmer(Loss ~ TimeDiff+Agemonths*Helper+Tarsus + NonHelper + Sex + BodyMass  + TQ + (1|FieldPeriodID),
             data = temp,
@@
 table2 <- data.frame(table2,myx=row.names(table2))
+table2 <- table2[rownames(table2) != 'z.TimeDiff',]
 table2$myx <- factor(table2$myx,levels = table2$myx[order(table2$Estimate,decreasing = T)])
@@
   geom_abline(slope = 0,intercept = 0,lty = 2)+
-  annotate('text',x = 0.55,y = 1.5,label = 'C')
+  annotate('text',x = 0.55,y = 1.5,label = 'B')
+
+Fig5C <- ggplot(temp,
+       aes(col = `Age (months)`,
+           fill =  `Age (months)`,
+           y = Loss,
+           x=TQ)) +
+    ylab(expression(Delta*'RTL'))+
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        legend.position = 'top') +
+  geom_point()+
+  stat_smooth(method = 'lm')+
+  scale_colour_manual(values = c('grey','darkorange','darkgreen'))+
+  scale_fill_manual(values = c('grey','darkorange','darkgreen'))+
+  annotate('text',x = 3,y = 1.3,label = 'C')
 
@@
     paper = 'a4')
-multiplot(Fig5A,Fig5B)
+multiplot(Fig5A,Fig5B,Fig5C,layout = matrix(c(1,1,2,3),nrow=2,byrow = T))
 dev.off()

### 2016-04-18 Scripts/SW juvenile telomeres IPC clean.R

@@ -276,8 +276,9 @@ Loss <- data.frame(earlies,
 Loss$TROC <- with(Loss,Loss/TimeDiff)
 
 #
-Loss <- subset(Loss,TimeDiff>0)
-Loss <- subset(Loss,TimeDiff<(365*2))
+Loss <- subset(Loss,TimeDiff>(365/2))
+Loss <- subset(Loss,TimeDiff<(365*5))
+Loss$TimeDiff <- Loss$TimeDiff/365
 
 # Get rid of stuff not to be used -----------------------------------------
 

### 2016-04-18 Manuscript/SW_early_life_telomeres.Rmd
DSR edits

@@
 Facultative cooperative breeding occurs in the Seychelles warbler [@Komdeur1994a; @Richardson2003a], and the presence of helping subordinates (but not non-helping subordinates) in the natal territory is associated with increased survival later in life [@Brouwer2012].
-Lastly, we have an established protocol for assessing absolute telomere length in this species [@Barrett2012], and telomere length predicts survival independently of age in adult Seychelles warblers, suggesting that telomeres act as a biomarker of cost [@Barrett2013a].
+Lastly, we have an established protocol for assessing telomere length in this species [@Barrett2012], and telomere length predicts survival independently of age in adult Seychelles warblers, suggesting that telomeres act as a biomarker of cost [@Barrett2013a].
 Thus, we have an excellent system in which to assess the costs of different social and environmental conditions experienced early in life, and to assess the later-life consequences of early-life conditions.

### 2016-04-18 Manuscript/SW_early_life_telomeres.Rmd
Home edits

@@
 3.
-We found that at the population level, telomere length decreases with cross-sectional and longitudinal measures of age, and that telomere length decreases most rapidly early in life.
-However, at the individual level we found no relationship between juvenile and adult telomere length, and that significant reductions in telomere length can only be detected within the first few months of life - after this, longitudinal changes in telomere length were not significantly different from zero.
+We found that telomere length decreases with cross-sectional and longitudinal measures of age, and that telomere length decreases most rapidly early in life.
+However, at the individual level we found a large amount of variation in telomere length and rates of change, with no relationship between within-individual juvenile and adult telomere length.
 
 4.
-Juvenile telomere length varied markedly among cohorts, and birds born in years with higher population density had longer telomeres.
-Within cohorts, we found no evidence that the early-life social environment or ecological conditions were related to telomere length or rates of shortening.
-We also found no relationship between early-life telomere dynamics and survival later in life.
+Juvenile telomere length varied markedly among cohorts, but we found no evidence that this variation was related to temporal variation in food availability or population density.
+Within cohorts, we found some evidence that telomere dynamics were related to territory quality and the presence of helpers in the natal territory, but the majority of variation in juvenile telomere length was unexplained.
+We found no relationship between early-life telomere dynamics and survival to adulthood or later in life.
 
 5.
-Our data suggest that telomeres are subject to strong population-level variation, and that telomeres at a point in time represent a snapshot of a fluid eco-evolutionary process.
-At the individual-level, however, our data suggest that the resolution of telomeres as biomarkers of cost is likely to be age-dependent, and in some cases limited.
+Our data suggest that although telomeres shorten with age, in natual popuations telomere dynamics vary enormously over both space and time.
+Ascertaining what explains this variation will be a difficult but worhtwhile task, requiring comprehensive, long-term ecological and quantitative genetic data.
 
@@
 320 birds on Cousin Island (04'20'S, 55'40'E) has been intensively studied since 1986 [@Richardson2003b; @Spurgin2014a].
-This species has two breeding seasons, running from June-August (main breeding season) and December-February (minor breeding season), when the breeding females on each of the *ca*.
+This species has a main breeding season, running from June-August, when the breeding females on each of the *ca*.
 115 territories lay one or, rarely, two or three eggs [@Komdeur1991].
@@
 Seychelles warblers are highly territorial and all territories are mapped during the breeding seasons using detailed observational data of foraging and territorial defence behaviour, and surveyed for territory quality [@Richardson2003b].
-Territory quality is calculated based on territory size, foliage cover and insect abundance [@Komdeur1992], and territory quality estimates obtained across years are averaged to obtain a single value for each territory [@Hammers2013].
+Territory quality is calculated based on territory size, foliage cover and insect abundance [@Komdeur1992], and where territory quality estimates were not available for a specific year we used the average value for that territory across years [@Hammers2013].
 Cousin is subject to intra- and inter-annual variation in rainfall and food availability, and such island-wide temporal variation may override the effects of individual territory quality.
@@
 We then explored how telomere length varied within the first year of life (telomere length vs age in months) and among cohorts (telomere length vs cohort ID), again using linear mixed models with individual ID as a random effect.
-We tested how temporal variation in telomere length varied between birds born in main and minor breeding seasons, and with temporal variation in population density (measured as the number of adult birds on the island at the end of each breeding season) and food availability, using linear mixed models with cohort ID as a random effect.
+We tested how temporal variation in telomere length varied with temporal variation in population density (measured as the number of adult birds on the island at the end of each breeding season) and food availability, using linear mixed models with cohort ID as a random effect.
 
-We used general linear mixed models with model averaging to explore how spatial variation in early-life environmental and social conditions influenced telomere length within cohorts.
+We used linear mixed models with model averaging to explore how spatial variation in early-life environmental and social conditions influenced telomere length within cohorts.
 Model averaging was carried out using the MuMIn package (version 1.10.5) in R [@Barton2012].
-We first created a full model containing the following explanatory variables: age (in months), tarsus length, sex, territory quality, season (main or minor) and the number of helping and non-helping subordinate birds present in the natal territory.
+We first created a full model containing the following explanatory variables: age (in months), tarsus length, sex, territory quality and the number of helping and non-helping subordinate birds present in the natal territory.
 We also included interaction terms between age and all the other variables as telomere dynamics are expected to vary within the first year of life [@Heidinger2012].
@@
 For individuals with longitudinal data (*n* = `r nrow(Loss)`) we repeated the above analyses of telomere dynamics, replacing telomere length with telomere loss as the response variable, and excluding the plate ID random effect (as each measurement of telomere loss was based on two or more measurements, and so run on multiple plates).
+For both sets of models, we first ran the model averaging procedure including all interaction terms, then re-ran the pipeline excluding non-significant interactions.
 
@@
 We found that, as in other studies, telomeres are lost at the greatest rate  early in life.
-We found that telomere dynamics are subject to strong cohort effects, and that temporal variation in RTL is population density and season.
+We found that telomere dynamics are subject to strong cohort effects.
 Within cohorts, we found no evidence that the ecological or social environment influenced telomere dynamics, and we found no evidence that early-life telomere length is related to juvenile or later-life survival.

### 2016-04-19 Manuscript/SW_early_life_telomeres.Rmd
DSR edits

@@
 For individuals with longitudinal data (*n* = `r nrow(Loss)`) we repeated the above analyses of telomere dynamics, replacing telomere length with telomere loss as the response variable, and excluding the plate ID random effect (as each measurement of telomere loss was based on two or more measurements, and so run on multiple plates).
-For both sets of models, we first ran the model averaging procedure including all interaction terms, then re-ran the pipeline excluding non-significant interactions.
+For both sets of models, we first ran the model averaging procedure including all interaction terms, then re-ran the pipeline excluding non-significant interactions in order to obtain more accurate model averaged parameter estimates and confidence intervals.
 
@@
 
-$\Delta$RTL did not differ significantly among cohorts of nestlings (F = `r round2(anova(lm(Loss~factor(FieldPeriodID),data = subset(Loss,Agemonths == 1)))[1,4])`; P `r round3(anova(lm(Loss~factor(FieldPeriodID),data = subset(Loss,Agemonths == 1)))[1,5])`), although power for this analysis was limited as longitudinal sample sizes within cohorts were low.
-Model selection revealed that the top model explaining $\Delta$RTL within cohorts of nestlings contained sex and tarsus length (Table S2).
-This model was a better fit than the null model ($\Delta$AICc = `r round2(AICc(lmer(Loss~1+(1|FieldPeriodID)+(1|TerritoryID),data=subset(Loss,Agemonths == 1),REML = FALSE))-AICc(lmer(Loss~Sex+Tarsus+(1|FieldPeriodID)+(1|TerritoryID),data=subset(Loss,Agemonths == 1),REML = FALSE)))`), although confidence limits for both the sex and tarsus effects overlapped zero (Fig.
-5A).
-$\Delta$RTL was negatively related to tarsus length, and higher in females, indicating greater telomere shortening in smaller nestlings, and in males (Fig.
-5B).
+$\Delta$RTL did not differ significantly among cohorts (F = `r round2(anova(lm(Loss~factor(FieldPeriodID),data = Loss))[1,4])`; P `r round3(anova(lm(Loss~factor(FieldPeriodID),data = Loss))[1,5])`), although power for this analysis was limited as longitudinal sample sizes within cohorts were low.
+Model selection revealed that the top model explaining $\Delta$RTL within cohorts contained only age (Table S2).
+This model was a better fit than the null model ($\Delta$AICc = `r round2(AICc(lmer(Loss~1+(1|FieldPeriodID)+(1|TerritoryID),data=Loss,REML = FALSE))-AICc(lmer(Loss~Agemonths+(1|FieldPeriodID)+(1|TerritoryID),data=Loss,REML = FALSE)))`), and only the age effect was significant (Fig.
+4C).
 
@@
 
-Early-life RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,4])`), with no interaction effect between RTL and juvenile age (estimate = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,4])`).
+Early-life RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~RTL+(1|FieldPeriodID),data=subset(juv13,Agemonths ==1),family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,4])`), with no interaction effect between RTL and juvenile age (estimate = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,4])`).
 To test whether cohort-level effects on RTL influenced survival we calculated cohort-level RTL (i.e.
@@
 This work was funded by two Natural Environment Research Council (NERC) grants to DSR (NE/F02083X/1 and NE/K005502/1).
-LGS was also funded by an EGI Fellowship from the University of Oxford, and HLD by a NERC fellowship.
+LGS was also funded by fellowships from the University of Oxford and the BBSRC, and HLD by a NERC fellowship.
 

### 2016-04-19 Manuscript/SW_early_life_telomeres_SOM.Rmd
DSR edits

@@
 ```{r Table S2}
-temp <- subset(Loss,Agemonths == 1)
+temp <- Loss
 
 #Run the model
-lm1 <- lmer(Loss ~ (Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ) + (1|FieldPeriodID),
+lm1 <- lmer(Loss ~ Agemonths*(Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ) + (1|FieldPeriodID),
             data = temp,
@@
 
-nams[nams == 'z.RTL'] <- 'RTL'
 nams[nams == 'z.NonHelper'] <- 'Non helpers'
@@
 
-tableS2 <- data.frame(Model = apply(t1,1,getterms,c(2:7)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
-colnames(tableS2) <- c('Model','df','AICc', 'Delta AICc','Weight')
-rownames(tableS2) <- NULL
-tableS2$Model[tableS2$Model == ''] <- 'Null model'
-tableS2[,3:5] <- lapply(tableS2[,3:5],as.numeric)
-kable(tableS2,digits=2)
+tableS1 <- data.frame(Model = apply(t1,1,getterms,c(2:14)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
+colnames(tableS1) <- c('Model','df','AICc', 'Delta AICc','Weight')
+rownames(tableS1) <- NULL
+tableS1$Model[tableS1$Model == ''] <- 'Null model'
+tableS1[,3:5] <- lapply(tableS1[,3:5],as.numeric)
+kable(tableS1,digits=2)
 

### 2016-04-19 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
DSR edits

@@
         axis.title.x = element_text(size = 14),
-        legend.position = 'top') +
+        legend.position = c(0.9,0.9),
+        legend.background = element_rect(fill = grey(0.9))) +
   geom_point()+
   scale_colour_manual(values = c('grey','darkorange','darkgreen'))+
-  scale_fill_manual(values = c('grey','darkorange','darkgreen'))
+  scale_fill_manual(values = c('grey','darkorange','darkgreen'))+
   annotate('text',x = 0,y = 2.1,label = 'A')
@@
 
-pdf('Figures/Figure 4.pdf',
-    width = 4,
-    height = 8,
-    paper = 'a4')
-multiplot(Fig4A,Fig4B,layout=matrix(c(1,2),nrow=2,byrow=T))
-dev.off()
-
-```
-
-```{r Figure 5}
-
 temp <- Loss
 #Run the model
-lm2 <- lmer(Loss ~ TimeDiff+Agemonths*Helper+Tarsus + NonHelper + Sex + BodyMass  + TQ + (1|FieldPeriodID),
+lm2 <- lmer(Loss ~ Agemonths+Helper+Tarsus + NonHelper + Sex + BodyMass  + TQ + (1|FieldPeriodID),
             data = temp,
@@
 
-Fig5A <- forestplot(table2)+
-  annotate('text',x = 0.75,y = 0.6,label = 'A')
+Fig4C <- forestplot(table2)+
+  annotate('text',x = 0.75,y = 0.6,label = 'C')
 
-temp <- Loss
-temp[,'Helpers'] <- ifelse(temp$Helper>0,'Present', 'Absent')
-temp[,'Non helpers'] <- ifelse(temp$NonHelper>0,'Present', 'Absent')
-temp[,'Age (months)'] <- factor(temp$Agemonths)
+pdf('Figures/Figure 4.pdf',
+    width = 8,
+    height = 8,
+    paper = 'a4')
+multiplot(Fig4A,Fig4B,Fig4C,layout=matrix(c(1,2,3,3),nrow=2,byrow=T))
+dev.off()
 
-Fig5B <- ggplot(temp,
-       aes(x = `Age (months)`,
-           y = Loss,
-           fill=Helpers)) +
-  theme_classic() +
-  ylab(expression(Delta*'RTL'))+
-  theme(axis.text.x = element_text(size = 12), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        legend.position = 'top') +
-  geom_jitter(aes(colour = Helpers),position=position_jitterdodge(dodge.width = 0.8))+
-  geom_boxplot(outlier.colour = NA,notch=F,alpha=0.2)+
-  scale_colour_manual(values = c('grey','darkorange'))+
-  scale_fill_manual(values = c('grey','darkorange'))+
-  geom_abline(slope = 0,intercept = 0,lty = 2)+
-  annotate('text',x = 0.55,y = 1.5,label = 'B')
+```
 
-Fig5C <- ggplot(temp,
-       aes(col = `Age (months)`,
-           fill =  `Age (months)`,
-           y = Loss,
-           x=TQ)) +
-    ylab(expression(Delta*'RTL'))+
+```{r Figure 5}
+temp <- subset(juv13,Agemonths != 1)
+temp$LayYear <- factor(temp$LayYear)
+temp[,'Survived to following year?'] <- ifelse(temp$SurvivedNext == 1,'Y','N')
+
+Fig4B <- ggplot(temp,
+       aes(x = LayYear,
+           y = RTL,
+           fill = `Survived to following year?`)) +
+  xlab('Cohort') +
   theme_classic() +
@@
         legend.position = 'top') +
-  geom_point()+
-  stat_smooth(method = 'lm')+
-  scale_colour_manual(values = c('grey','darkorange','darkgreen'))+
-  scale_fill_manual(values = c('grey','darkorange','darkgreen'))+
-  annotate('text',x = 3,y = 1.3,label = 'C')
+  geom_boxplot(outlier.shape = NA,alpha = 0.2)+
+  geom_jitter(aes(col = `Survived to following year?`),position=position_jitterdodge(jitter.width = 0.2))+
+  annotate('text',x = 0.5,y = 1.5,label = 'B')
 
-pdf('Figures/Figure 5.pdf',
-    width = 8,
-    height = 8,
-    paper = 'a4')
-multiplot(Fig5A,Fig5B,Fig5C,layout = matrix(c(1,1,2,3),nrow=2,byrow = T))
-dev.off()
+Fig4B
 

### 2016-04-19 Scripts/SW juvenile telomeres IPC clean.R

@@ -4,7 +4,7 @@
 #################################################################################
 
 #Only use Ellie's samples
-#dd0 <- subset(dd0,Whodunnit == 'EAF')
+dd0 <- subset(dd0,Whodunnit == 'EAF')
 
 #Average repeats of blood samples
 av <- ave(dd0$RTL,c(dd0$BloodID,dd0$Status,dd0$PlateID))

### 2016-04-19 Manuscript/SW_early_life_telomeres.Rmd
DSR edits

@@
 Cox regression showed no effect of individual-level or cohort-level RTL on survival later in life (individual-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL+Agemonths+frailty(FieldPeriodID))))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL+Agemonths+frailty(FieldPeriodID))))$coefficients[1,6])`; cohort-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL*Agemonths+RTL+frailty(FieldPeriodID))))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL*Agemonths+RTL+frailty(FieldPeriodID))))$coefficients[1,6])`), and no interaction between RTL and age on survival (individual-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL*Agemonths+frailty(FieldPeriodID))))$coefficients[4,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL*Agemonths+frailty(FieldPeriodID))))$coefficients[4,6])`).
-$\Delta$RTL was not related to longer-term survival (estimate `r round3(summary(coxph(with(subset(subset(Loss,LayYear<2008),Agemonths == 1),Surv(RemainingLife2+1,Died)~Loss)))$coefficients[1,1])`, P = `r round3(summary(coxph(with(subset(subset(Loss,LayYear<2008),Agemonths ==1 ),Surv(RemainingLife2+1,Died)~Loss)))$coefficients[1,5])`).
+$\Delta$RTL was not related to longer-term survival (estimate `r round3(summary(coxph(with(subset(Loss,LayYear<2008),Surv(RemainingLife2+1,Died)~Loss+Agemonths)))$coefficients[1,1])`, P = `r round2(summary(coxph(with(subset(Loss,LayYear<2008),Surv(RemainingLife2+1,Died)~Loss)))$coefficients[1,5])`).
 

### 2016-04-19 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
DSR edits

@@
 ```{r Figure 5}
-temp <- subset(juv13,Agemonths != 1)
+temp <- juv
 temp$LayYear <- factor(temp$LayYear)
 temp[,'Survived to following year?'] <- ifelse(temp$SurvivedNext == 1,'Y','N')
+temp[,'Age (months)'] <- factor(temp$Agemonths)
 
-Fig4B <- ggplot(temp,
-       aes(x = LayYear,
-           y = RTL,
-           fill = `Survived to following year?`)) +
-  xlab('Cohort') +
+Fig5 <- ggplot(temp,
+       aes(x = RTL,
+           y = SurvivedNext,
+           fill = `Age (months)`,
+           col = `Age (months)`)) +
+  xlab('RTL') +
+  ylab('Proportion survived')+
   theme_classic() +
@@
         legend.position = 'top') +
-  geom_boxplot(outlier.shape = NA,alpha = 0.2)+
-  geom_jitter(aes(col = `Survived to following year?`),position=position_jitterdodge(jitter.width = 0.2))+
-  annotate('text',x = 0.5,y = 1.5,label = 'B')
+  geom_point(alpha=0.2)+
+  stat_smooth(method = 'glm',family = 'binomial')+
+    scale_colour_manual(values = c('grey','darkorange','darkgreen'))+
+  scale_fill_manual(values = c('grey','darkorange','darkgreen'))
 
-Fig4B
+pdf('Figures/Figure 5.pdf',
+    width = 4,
+    height = 4,
+    paper = 'a4')
+Fig5
+dev.off()
 

### 2016-04-19 Manuscript/SW_early_life_telomeres.Rmd
DSR edits

@@
 the median RTL for each cohort) and added this term to a GLMM, and found that cohort-level RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~cohortTL+RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~cohortTL+RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,4])`).
-Using the longitudinal data, we found that $\Delta$RTL had no effect on survival to the year after the second sampling event (estimate = `r round2(summary(glm(SurvivedNext2~Loss+Agemonths,data = subset(Loss,FieldPeriodID<2013),family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glm(SurvivedNext2~Loss+Agemonths,data=subset(Loss,LayYear<2013),family = 'binomial'))$coefficients[2,4])`).
+Using the longitudinal data, we found that $\Delta$RTL had no effect on survival to the year after the second sampling event (estimate = `r round2(summary(glm(SurvivedNext2~Loss+Agemonths,data = subset(Loss,LayYear<2013),family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glm(SurvivedNext2~Loss+Agemonths,data=subset(Loss,LayYear<2013),family = 'binomial'))$coefficients[2,4])`).
 

### 2016-04-20 Manuscript/SW_early_life_telomeres.Rmd
Email edit
test

@@
 
-**Correspondence:** Lewis Spurgin: lewisspurgin@gmail.com; David Richardson: david.richardson@uea.ac.uk
+**Correspondence:** Lewis Spurgin: l.spurgin@uea.ac.uk; David Richardson: david.richardson@uea.ac.uk
 
@@
 1.
-Understanding the short and long term costs of individual early-life experiences is fundamental to understanding life-history evolution.
+Understanding the costs of individual early-life experiences is fundamental to understanding life-history evolution.
 Telomeres, the protective caps at the ends of chromosomes, shorten in response to oxidative stress, and telomere shortening is correlated with reduced survival.

### 2016-04-22 Manuscript/SW_early_life_telomeres.Rmd
DSR edtis

@@
 1.
-Understanding the costs of individual early-life experiences is fundamental to understanding life-history evolution.
+Understanding the costs of individual early-life experiences is fundamental to understanding life-history evolution and population dynmamics.
 Telomeres, the protective caps at the ends of chromosomes, shorten in response to oxidative stress, and telomere shortening is correlated with reduced survival.
-Thus, it is possible that telomere dynamics may help us quantify individual variation in early-life costs, and enhance our understanding of how poor conditions in early life are related to later-life survival.
+Thus, telomere dynamics may help us quantify individual variation in early-life costs, and enhance our understanding of how poor conditions in early life are related to later-life survival.
 
@@
 We found that telomere length decreases with cross-sectional and longitudinal measures of age, and that telomere length decreases most rapidly early in life.
-However, at the individual level we found a large amount of variation in telomere length and rates of change, with no relationship between within-individual juvenile and adult telomere length.
+Juvenile telomere length varied markedly among cohorts, but we found no evidence that this variation was related to temporal variation in food availability or population density.
 
 4.
-Juvenile telomere length varied markedly among cohorts, but we found no evidence that this variation was related to temporal variation in food availability or population density.
-Within cohorts, we found some evidence that telomere dynamics were related to territory quality and the presence of helpers in the natal territory, but the majority of variation in juvenile telomere length was unexplained.
+At the individual level we found a large amount of variation in telomere length and rates of change, with no relationship between within-individual juvenile and adult telomere length.
+We found weak evidence that telomere dnegatively related to telomere length, but the majority of variation in juvenile telomere length and dynamics was unexplained.
 We found no relationship between early-life telomere dynamics and survival to adulthood or later in life.
@@
 5.
-Our data suggest that although telomeres shorten with age, in natual popuations telomere dynamics vary enormously over both space and time.
-Ascertaining what explains this variation will be a difficult but worhtwhile task, requiring comprehensive, long-term ecological and quantitative genetic data.
+Our comprehensive, long-term data suggest that although telomeres shorten with age, in natual popuations telomere dynamics vary enormously over both space and time.
+Ascertaining what explains this variation will be a difficult but worhtwhile task, which will require combining long-term ecological and quantitative genetic data.
 
@@
 320 birds on Cousin Island (04'20'S, 55'40'E) has been intensively studied since 1986 [@Richardson2003b; @Spurgin2014a].
-This species has a main breeding season, running from June-August, when the breeding females on each of the *ca*.
+This species' main breeding season runs from June-August, when the breeding females on each of the *ca*.
 115 territories lay one or, rarely, two or three eggs [@Komdeur1991].
@@
 
-For a subset of first-year birds for which we had longitudinal data, with an additional sample taken as an adult.
+For a subset of first-year birds we had longitudinal data, with an additional sample taken as an adult.
 For these individuals we calculated the within-individual change in RTL by subtracting early-life RTL from adult RTL (hereafter $\Delta$RTL).
@@
 We performed all analyses using R version 3.0.1 [@RDevelopmentCoreTeam2011].
-Because we are using a different measure of telomere length to the previous study of adult telomere dynamics in this species [@Barrett2013a], as well as including many additional samples, we first re-tested, using all available samples, whether telomere length decreases with age and $\Delta$age [a longitudinal measure based on within-subject centring; @VandePol2009] in the Seychelles warbler, using linear mixed models with individual ID as a random effect.
+Our dataset differs from slightly from that of Barrett *et al.* [-@Barrett2013a] in that we are using RTL calculated driectly from the qPCR data rather than converting into an absolute measure , and we are including many additional samples.
+We therefore first re-tested, using all available samples, whether telomere length decreases with age and $\Delta$age [a longitudinal measure based on within-subject centring; @VandePol2009] in the Seychelles warbler, using linear mixed models with individual ID as a random effect.
 For the analysis of $\Delta$age we we only included birds who had been sampled at least twice.
-We then explored how telomere length varied within the first year of life (telomere length vs age in months) and among cohorts (telomere length vs cohort ID), again using linear mixed models with individual ID as a random effect.
+We then, using the juvenile samples, explored how telomere length varied within the first year of life (telomere length vs age in months) and among cohorts (telomere length vs cohort ID), again using linear mixed models with individual ID as a random effect.
 We tested how temporal variation in telomere length varied with temporal variation in population density (measured as the number of adult birds on the island at the end of each breeding season) and food availability, using linear mixed models with cohort ID as a random effect.
@@
 We used generalised linear mixed models to test whether telomere length/loss were related to survival to adulthood.
-Models were set up with a binomial error structure, and included RTL/$\Delta$RTL and juvenile age as explanatory variable, along with the same random effects as for the model averaging (see above).
+Models were set up with a binomial error structure, and included RTL or $\Delta$RTL and juvenile age as explanatory variables, along with the same random effects as for the model averaging (see above).
 We then used Cox regression, implemented in the survival package in R [@Therneau2011], to test whether long-term survival was related to individual-level telomere dynamics.
@@
 #####Early-life RTL
-Considering the entire Seychelles warbler lifespan, RTL decreased with age (LMM: t = `r round2(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,4])`; P `r round3(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,5])`), although there were clear increases in cross-sectional telomere length after one year of age and at several points later in life (Fig.
+Using the cross-sectional data RTL decreased with age across the entire lifespan of the Seychelles warbler (LMM: t = `r round2(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,4])`; P `r round3(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,5])`), although there were clear increases in (cross-sectional) telomere length after one year of age and at several points later in life (Fig.
 1A).
-RTL also decreased with age within individuals (i.e.
-with $\Delta$age; t = `r round2(summary(lme(RTL~DeltaAge,data=subset(dd,DeltaAge!=0),random=~1|BirdID))$tTable[2,4])`; P `r round3(summary(lme(RTL~DeltaAge,data=subset(dd,DeltaAge!=0),random=~1|BirdID))$tTable[2,5])`; Fig.
-1B), and with age within the first year of life (i.e.
+RTL also decreased with age within the first year of life (i.e.
 excluding adult samples; t = `r round2(summary(lme(RTL~Agemonths,data=juv,random=~1|BirdID))$tTable[2,4])`; P `r round3(summary(lme(RTL~Agemonths,data=juv,random=~1|BirdID))$tTable[2,5])`).
+In the longitudinal analysis RTL decreased with age within individuals (i.e.
+with $\Delta$age; t = `r round2(summary(lme(RTL~DeltaAge,data=subset(dd,DeltaAge!=0),random=~1|BirdID))$tTable[2,4])`; P = `r round3(summary(lme(RTL~DeltaAge,data=subset(dd,DeltaAge!=0),random=~1|BirdID))$tTable[2,5])`; Fig.
+1B).
 
 Early-life RTL varied significantly among cohorts (Likelihood ratio test: F = `r round2(anova(lme(RTL~factor(FieldPeriodID),data=juv,random=~1|BirdID))[2,3])`; P `r round3(anova(lme(RTL~Agemonths,data=juv,random=~1|BirdID))[2,4])`).
-Cohort-level variation in juvenile RTL was positively related to adult population density (LMM controlling for age and cohort: t = `r round2(summary(lme(RTL~Agemonths+Density,data=juv,random=~1|FieldPeriodID))$tTable[3,4])`; P `r round3(summary(lme(RTL~Agemonths+Density,data=juv,random=~1|FieldPeriodID))$tTable[3,5])`; Figs 2B,2C).
-Variation in cohort-level RTL was not related temporal variation in  food availability (t = `r round2(summary(lme(RTL~Agemonths+Insect,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,4])`; P `r round3(summary(lme(RTL~Agemonths+Insect,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,5])`).
+Cohort-level variation in juvenile RTL was not related to adult population density (LMM controlling for age and cohort: t = `r round2(summary(lme(RTL~Agemonths+Density,data=juv,random=~1|FieldPeriodID))$tTable[3,4])`; P = `r round3(summary(lme(RTL~Agemonths+Density,data=juv,random=~1|FieldPeriodID))$tTable[3,5])`), or to temporal variation in  food availability (t = `r round2(summary(lme(RTL~Agemonths+Insect,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,4])`; P = `r round3(summary(lme(RTL~Agemonths+Insect,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,5])`).
 
-Within cohorts, the top model explaining variation in early-life RTL contained age, sex, tarsus length, and the interaction between age and tarsus length (Table S1).
-The top model was much better supported than a null model ($\Delta$AICc = `r round2(AICc(lmer(RTL~1+(1|FieldPeriodID)+(1|TerritoryID),data=juv,REML = FALSE))-AICc(lmer(RTL~Agemonths+Sex+Tarsus+Agemonths*Tarsus+(1|FieldPeriodID)+(1|TerritoryID),data=juv,REML = FALSE)))`), and with the exception of the sex and tarsus effects, model-averaged confidence intervals did not overlap zero (Fig.
+Within cohorts, the top model explaining variation in early-life RTL contained age, territory quality, tarsus length, and the interaction between age and tarsus length (Table S1).
+The top model was much better supported than a null model ($\Delta$AICc = `r round2(AICc(lmer(RTL~1+(1|FieldPeriodID)+(1|TerritoryID),data=juv,REML = FALSE))-AICc(lmer(RTL~Agemonths+TQ+Tarsus+Agemonths*Tarsus+(1|FieldPeriodID)+(1|TerritoryID),data=juv,REML = FALSE)))`).
+Model-averaged confidence intervals of the age and tarsus x age effects did not overlap zero, but the tarsus and territory qusality effects were not significant (Fig.
 3A).
@@
 3B).
-Males had slightly longer telomeres than females, particularly at the fledgling and subadult stages, although this effect was weak  (Fig.
+Territory quality was negatively related to RTL, particularly in nestlings, although this effect was very weak (Fig.
 3C).
@@
 #####Early-life telomere shortening
-Longitudinal data showed that early-life RTL was not related to adult RTL (R^2^ = `r round3(summary(lm(RTL1~RTL+Agemonths,data=Loss))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data=Loss))$coefficients[2,3])`; P = `r round3(summary(lm(RTL1~RTL+Agemonths,data = Loss))$coefficients[2,4])`; Fig.
+Longitudinal data showed that early-life RTL was not related to adult RTL (R^2^ = `r round2(summary(lm(RTL1~RTL+Agemonths,data=Loss))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data=Loss))$coefficients[2,3])`; P = `r round2(summary(lm(RTL1~RTL+Agemonths,data = Loss))$coefficients[2,4])`; Fig.
 4A).
+Running this analysis separetly for juvenile age classes revealed that this relationship was not significant for nestlings (R^2^ = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(Loss,Agemonths == 1)))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(Loss,Agemonths == 1)))$coefficients[2,3])`; P = `r round2(summary(lm(RTL1~RTL+Agemonths,data = subset(Loss,Agemonths == 1)))$coefficients[2,4])`), fledglings (R^2^ = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(Loss,Agemonths == 6)))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(Loss,Agemonths == 6)))$coefficients[2,3])`; P = `r round2(summary(lm(RTL1~RTL+Agemonths,data = subset(Loss,Agemonths == 6)))$coefficients[2,4])`) or subadults (R^2^ = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(Loss,Agemonths == 10)))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(Loss,Agemonths == 10)))$coefficients[2,3])`; P = `r round2(summary(lm(RTL1~RTL+Agemonths,data = subset(Loss,Agemonths == 10)))$coefficients[2,4])`).
 $\Delta$RTL was significantly lower than zero in nestlings (t = `r round2(t.test(subset(Loss,Agemonths == 1)$Loss)$statistic)`; P `r round3(t.test(subset(Loss,Agemonths == 1)$Loss)$p.value)`), but not in fledglings (t = `r round2(t.test(subset(Loss,Agemonths == 6)$Loss)$statistic)`; P = `r round2(t.test(subset(Loss,Agemonths == 6)$Loss)$p.value)`) or subadults (t = `r round2(t.test(subset(Loss,Agemonths == 10)$Loss)$statistic)`; P = `r round2(t.test(subset(Loss,Agemonths == 10)$Loss)$p.value)`).
@@
 Model selection revealed that the top model explaining $\Delta$RTL within cohorts contained only age (Table S2).
-This model was a better fit than the null model ($\Delta$AICc = `r round2(AICc(lmer(Loss~1+(1|FieldPeriodID)+(1|TerritoryID),data=Loss,REML = FALSE))-AICc(lmer(Loss~Agemonths+(1|FieldPeriodID)+(1|TerritoryID),data=Loss,REML = FALSE)))`), and only the age effect was significant (Fig.
+This model was a better fit than the null model ($\Delta$AICc = `r round2(AICc(lmer(Loss~1+(1|FieldPeriodID)+(1|TerritoryID),data=Loss,REML = FALSE))-AICc(lmer(Loss~Agemonths+(1|FieldPeriodID)+(1|TerritoryID),data=Loss,REML = FALSE)))`), and only the age effect was significant over the top model set (Fig.
 4C).
@@
 
-Early-life RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~RTL+(1|FieldPeriodID),data=subset(juv13,Agemonths ==1),family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,4])`), with no interaction effect between RTL and juvenile age (estimate = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,4])`).
+Early-life RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~cenTL+Agemonths+(1|PlateID)+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~cenTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,4])`), with no interaction effect between RTL and juvenile age (estimate = `r round2(summary(glmer(SurvivedNext~cenTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~cenTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,4])`), suggesting that there were no age-specific effects of juvenile RTL on survival to adulthood.
 To test whether cohort-level effects on RTL influenced survival we calculated cohort-level RTL (i.e.
@@
 
-Cox regression showed no effect of individual-level or cohort-level RTL on survival later in life (individual-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL+Agemonths+frailty(FieldPeriodID))))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL+Agemonths+frailty(FieldPeriodID))))$coefficients[1,6])`; cohort-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL*Agemonths+RTL+frailty(FieldPeriodID))))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL*Agemonths+RTL+frailty(FieldPeriodID))))$coefficients[1,6])`), and no interaction between RTL and age on survival (individual-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL*Agemonths+frailty(FieldPeriodID))))$coefficients[4,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL*Agemonths+frailty(FieldPeriodID))))$coefficients[4,6])`).
+Cox regression showed no effect of individual-level or cohort-level RTL on survival later in life (individual-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cenTL+Agemonths+frailty(FieldPeriodID))))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cenTL+Agemonths+frailty(FieldPeriodID))))$coefficients[1,6])`; cohort-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL*Agemonths+RTL+frailty(FieldPeriodID))))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL*Agemonths+RTL+frailty(FieldPeriodID))))$coefficients[1,6])`), and no interaction between RTL and juvenile age on survival (individual-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cenTL*Agemonths+frailty(FieldPeriodID))))$coefficients[4,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cenTL*Agemonths+frailty(FieldPeriodID))))$coefficients[4,6])`).
 $\Delta$RTL was not related to longer-term survival (estimate `r round3(summary(coxph(with(subset(Loss,LayYear<2008),Surv(RemainingLife2+1,Died)~Loss+Agemonths)))$coefficients[1,1])`, P = `r round2(summary(coxph(with(subset(Loss,LayYear<2008),Surv(RemainingLife2+1,Died)~Loss)))$coefficients[1,5])`).
@@
 We found that telomere dynamics are subject to strong cohort effects.
-Within cohorts, we found no evidence that the ecological or social environment influenced telomere dynamics, and we found no evidence that early-life telomere length is related to juvenile or later-life survival.
+Within cohorts, we found little evidence that the ecological or social environment influenced telomere dynamics, and we found no evidence that early-life telomere length is related to juvenile or later-life survival.
 
@@
 
-Despite our clear finding that RTL decreases with age throughout the Seychelles warbler lifespan (Fig.
-1), we found that at the individual-level, juvenile and adult telomere were not correlated, and that telomere shortening can only be detected very early in life (Fig.
+ We found that at the individual-level, juvenile and adult telomere were not correlated, and that consistent telomere shortening only occurs  very early in life (Fig.
 4).

### 2016-04-22 Manuscript/SW_early_life_telomeres_SOM.Rmd
DSR edtis

@@
 #Run the model
-lm1 <- lmer(Loss ~ Agemonths*(Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ) + (1|FieldPeriodID),
+lm1 <- lmer(Loss ~ Agemonths*(Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ) + (1|FieldPeriodID) + (1|TerritoryID),
             data = temp,

### 2016-04-22 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
DSR edtis

--- 2016-04-19 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
@@
         axis.title.x = element_text(size = 14),
+                axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'),
         legend.position = 'none')+
@@
         axis.title.x = element_text(size = 14),
+        axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'),
         legend.position = 'none')+
@@
         axis.title.x = element_text(size = 14),
+                axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'),
         legend.position = 'top')+
   geom_jitter(col = 'grey')+
-  geom_boxplot(outlier.colour = NA,notch=F,alpha=0.2)+
-  scale_colour_manual(values = c('grey','darkorange'))+
-  scale_fill_manual(values = c('grey','darkorange'))
+  geom_boxplot(outlier.colour = NA,notch=F,alpha=0)
 
@@
 Fig3A <- forestplot(table2)+
-  annotate('text',x = 0.8,y = 0.25,label = 'A')
+  annotate('text',x = 0.6,y = 0.25,label = 'A')+
+  theme(axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'))
 
@@
         axis.title.x = element_text(size = 14),
+                axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'),
         legend.position = 'top') +
@@
        aes(x = RTL,
-           y = RTL1,
-           col=`Age (months)`)) +
+           y = RTL1)) +
   ylab('Adult RTL') + 
@@
         axis.title.x = element_text(size = 14),
-        legend.position = c(0.9,0.9),
+        legend.position = c(0.82,0.88),
+        legend.title = element_text(size = 12),
+        axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'),
         legend.background = element_rect(fill = grey(0.9))) +
+  geom_point(col='grey')+
+  stat_smooth(method = 'lm',col='darkgrey')+
+  annotate('text',x = 0,y = 1.8,label = 'A')
+
+temp <- subset(dd,DeltaAge !=0)
+reps <- tapply(temp$RTL,temp$BirdID,length)
+keep <- names(reps[reps>3])
+temp <- subset(temp,BirdID %in% keep)
+
+mins <- tapply(temp$Agemonths,temp$BirdID,min)
+keep2 <- names(mins[mins <5])
+temp <- subset(temp,BirdID %in% keep2)
+temp <- subset(temp,!duplicated(paste0(BirdID,Agemonths)))
+
+Fig4B <- ggplot(temp,aes(x = Agemonths,y = RTL,group = BirdID,col=factor(BirdID)))+
 geom_point()+
-  scale_colour_manual(values = c('grey','darkorange','darkgreen'))+
-  scale_fill_manual(values = c('grey','darkorange','darkgreen'))+
-  annotate('text',x = 0,y = 2.1,label = 'A')
-
-Fig4B <- ggplot(temp,
-       aes(x = `Age (months)`,
-           y = Loss)) +
-  ylab(expression(Delta*'RTL')) + 
+geom_line()+
   xlab('Age (months)') +
@@
         axis.title.x = element_text(size = 14),
-        legend.position = 'top') +
-  geom_boxplot(outlier.shape = NA)+
-  geom_jitter(col = 'grey')+
-  annotate('text',x = 0.5,y = 1.5,label = 'B')+
-  geom_abline(slope = 0,intercept = 0,lty = 2)
+        axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'),
+        legend.position = 'none')+
+  annotate('text',x = -5,y = 2,label = 'B')
 
@@
 #Run the model
-lm2 <- lmer(Loss ~ Agemonths+Helper+Tarsus + NonHelper + Sex + BodyMass  + TQ + (1|FieldPeriodID),
+lm2 <- lmer(Loss ~ Agemonths+Helper+Tarsus + NonHelper + Sex + BodyMass  + TQ + (1|FieldPeriodID) + (1|TerritoryID),
             data = temp,
@@
 Fig4C <- forestplot(table2)+
-  annotate('text',x = 0.75,y = 0.6,label = 'C')
+  annotate('text',x = 0.55,y = 0.6,label = 'C')+
+  theme(        axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'))
 
@@
 
-```{r Figure 5}
-temp <- subset(juv13,Agemonths != 1)
-temp$LayYear <- factor(temp$LayYear)
-temp[,'Survived to following year?'] <- ifelse(temp$SurvivedNext == 1,'Y','N')
+temp <- subset(dd,DeltaAge !=0)
+reps <- tapply(temp$RTL,temp$BirdID,length)
+keep <- names(reps[reps>3])
+temp <- subset(temp,BirdID %in% keep)
 
-Fig4B <- ggplot(temp,
-       aes(x = LayYear,
-           y = RTL,
-           fill = `Survived to following year?`)) +
-  xlab('Cohort') +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        legend.position = 'top') +
-  geom_boxplot(outlier.shape = NA,alpha = 0.2)+
-  geom_jitter(aes(col = `Survived to following year?`),position=position_jitterdodge(jitter.width = 0.2))+
-  annotate('text',x = 0.5,y = 1.5,label = 'B')
+mins <- tapply(temp$Agemonths,temp$BirdID,min)
+keep2 <- names(mins[mins <5])
+temp <- subset(temp,BirdID %in% keep2)
+temp <- subset(temp,!duplicated(paste0(BirdID,Agemonths)))
 
-Fig4B
-
-```
+ggplot(temp,aes(x = Agemonths,y = RTL,group = BirdID,col=factor(BirdID)))+
+geom_point()+
+geom_line()+
+theme(legend.position='none')

### 2016-04-22 Scripts/SW juvenile telomeres IPC clean.R

@@ -4,7 +4,7 @@
 #################################################################################
 
 #Only use Ellie's samples
-dd0 <- subset(dd0,Whodunnit == 'EAF')
+#dd0 <- subset(dd0,Whodunnit == 'EAF')
 
 #Average repeats of blood samples
 av <- ave(dd0$RTL,c(dd0$BloodID,dd0$Status,dd0$PlateID))
@@ -63,6 +63,11 @@ dd$Age[dd$Fledged != 'Adults'] <- 1
 
 dd$Tarsus <- NA
 dd$DeltaAge <- NA
+dd$Malaria <- NA
+
+malaria <- subset(malaria,!duplicated(BloodID))
+malaria <- subset(malaria,Consensus <2)
+
 for(i in 1:nrow(dd))
 {
   ifelse(is.na(dd$RightTarsus[i]),
@@ -72,7 +77,10 @@ for(i in 1:nrow(dd))
          dd$Tarsus[i] <- dd$RightTarsus[i])
   
   dd$DeltaAge[i] <- dd$Agemonths[i] - mean(subset(dd,BirdID == dd$BirdID[i])$Agemonths)
-  
+  if(dd$BloodID[i] %in% malaria$BloodID)
+     {
+       dd$Malaria[i] <- subset(malaria,BloodID == dd$BloodID[i])$Consensus
+     }
 }
 
 dd$RightTarsus <- NULL
@@ -197,7 +205,7 @@ for(i in 1:nrow(juv))
       juv$TQ[i] <- log10(subset(cd,FieldPeriodID == juv$FieldPeriodID[i])$TQcorrected)
     } else
     {
-      juv$TQ[i] <- log10(mean(cd$TQcorrected))
+      log10(mean(cd$TQcorrected))
     }
     juv$TQI[i] <- juv$TQ[i]/juv$GroupSize[i]
     
@@ -253,7 +261,7 @@ subset(!(duplicated(BirdID)))
 
 #Get earliest subsequent catch
 laters <- merge(aggregate(CatchDate~BirdID,subset(dd,!(BloodID %in% earlies$BloodID)),
-                          min),
+                          mean),
                 subset(dd,!(BloodID %in% earlies$BloodID))) %>% 
 subset(!(duplicated(BirdID)))
 
@@ -277,7 +285,6 @@ Loss$TROC <- with(Loss,Loss/TimeDiff)
 
 #
 Loss <- subset(Loss,TimeDiff>(365/2))
[…]

### 2016-04-22 Scripts/SW juvenile telomeres IPC load.r

@@ -25,3 +25,4 @@ status <- read.csv('Data/SW TL all breedstatus.csv')
 insects <- read.csv('Data/Insects.csv')
 dens <- read.table('Data/PsizeFP.txt')
 chickinfo <- read.csv('Data/Table chick info.csv')
+malaria <- read.csv('Data/Malaria.csv')

### 2016-05-06 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
More edits

@@
 
-temp <- Loss
-#Run the model
-lm2 <- lmer(Loss ~ Agemonths+Helper+Tarsus + NonHelper + Sex + BodyMass  + TQ + (1|FieldPeriodID) + (1|TerritoryID),
-            data = temp,
-            REML = FALSE)
-lm2 <- standardize(lm2)
+#First calculate Delta RTL 
+temp <- subset(dd,!duplicated(paste0(BirdID,Agemonths)))
+counts <- with(temp,tapply(RTL,BirdID,length))
+c3 <- names(counts[counts >= 2])
 
-#Model averaging
-lm3 <- dredge(lm2)
-lm3 <- subset(lm3,delta <=6)
-lmavg <- model.avg(lm3,fit=T)
-myimp <- importance(lmavg)
+dd3 <- subset(temp,BirdID %in% c3)
 
-#Put it all into a nice table
-table2 <- imptable(lmavg,myimp)
-table2 <- signif(table2,2)
+ggplot(dd3,aes(x = Agemonths,y = RTL, group = BirdID))+
+  geom_line(alpha = 0.5)
 
-#Sort out row/col names
-nams <- rownames(table2)
-nams[nams == 'z.NonHelper'] <- 'Non helpers'
-nams[nams == 'c.Sex'] <- 'Sex (males)'
-nams[nams == 'z.TQ'] <- 'TQ'
-nams[nams == 'c.Helper'] <- 'Helpers'
-nams[nams == 'z.BodyMass'] <- 'Body mass'
-nams[nams == 'z.Agemonths'] <- 'Age'
-nams[nams == 'z.Tarsus'] <- 'Tarsus'
-nams[nams == 'c.Season'] <- 'Season (minor)'
-nams[nams == 'c.Season:z.Agemonths'] <- 'Age x season'
-nams[nams == 'z.Agemonths:z.NonHelper'] <- 'Age x non helpers'
-nams[nams == 'c.Sex:z.Agemonths'] <- 'Age x sex'
-nams[nams == 'z.Agemonths:z.TQ'] <- 'Age x TQ'
-nams[nams == 'c.Helper:z.Agemonths'] <- 'Age x helpers'
-nams[nams == 'z.Agemonths:z.BodyMass'] <- 'Age x mass'
-nams[nams == 'z.Agemonths:z.Tarsus'] <- 'Age x tarsus'
+dd3$DeltaRTL <- NA
 
-rownames(table2) <- nams
-colnames(table2) <- c('Estimate','SE','2.5%','97.5%','RI')
+for(i in 1:nrow(dd3))
+{
+  cd <- subset(dd3,BirdID == dd3$BirdID[i])
+  cd <- cd[order(cd$Agemonths),]
 
-table2 <- data.frame(table2,myx=row.names(table2))
-table2 <- table2[rownames(table2) != 'z.TimeDiff',]
-table2$myx <- factor(table2$myx,levels = table2$myx[order(table2$Estimate,decreasing = T)])
+  birdpos <- which(cd$BloodID == dd3$BloodID[i])
 
-Fig4C <- forestplot(table2)+
-  annotate('text',x = 0.55,y = 0.6,label = 'C')+
-  theme(        axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black'))
+  if(birdpos == 1)
+  {
+    dd3$DeltaRTL[i] <- 0
+  } else
+  {
+      prevbird <- cd[birdpos-1,'RTL']
+  dd3$DeltaRTL[i] <- dd3$RTL[i]-prevbird
+  dd3$MidAge[i] <- mean(c(dd3$Agemonths[i],cd[birdpos-1,'Agemonths']))
+  dd3$MinAge[i] <- cd[birdpos-1,'Agemonths']
+  dd3$TimeDiff[i] <- dd3$Agemonths[i] - cd[birdpos-1,'Agemonths']
+  }
+
+}
+
+temp <- subset(dd3,DeltaRTL != 0)
+
+boxplot(DeltaRTL~SurvivedNext,data=subset(temp,MinAge>1))
+summary(lm(DeltaRTL~RemainingLife,data=subset(temp,MinAge>1)))
+
+plot(DeltaRTL~LayYear,data=subset(temp,MinAge>1 ))
+
+temp <- subset(dd0,!duplicated(paste0(BloodID,RTL)))
+
+temp <- subset(temp,RTL<2)
+temp <- subset(temp,RTL > 0.04)
+
+counts <- with(temp,tapply(RTL,BloodID,length))
+c3 <- names(counts[counts %in% c(2:5)])
+
+dd3_2 <- subset(temp,BloodID %in% c3)
+dd3_2 <- dd3_2[order(dd3_2$BloodID),]
+counts <- with(dd3_2,tapply(RTL,BloodID,length))
+
+dd3_2$birdpos <- unlist(sapply(counts,function(x) c(1:x)))
+
+dd3_2$DeltaRTL <- NA
+
+for(i in 1:nrow(dd3_2))
+{
+  cd <- subset(dd3_2,BloodID == dd3_2$BloodID[i])
+
+  if(dd3_2$birdpos[i] == 1)
+  {
+    dd3_2$DeltaRTL[i] <- 0
+  } else
+  {
+      prevbird <- subset(cd,birdpos == dd3_2$birdpos[i]-1)$RTL
+  dd3_2$DeltaRTL[i] <- dd3_2$RTL[i]-prevbird
+  }
+
+}
+
+dd3 <- subset(dd3,DeltaRTL !=0)
+dd3_2 <- subset(dd3_2,DeltaRTL!=0)
+
+dd3_2 <- dd3_2[1:nrow(dd3),]
+
+temp <- data.frame(DeltaRTL = c(dd3$DeltaRTL,dd3_2$DeltaRTL),Group = rep(c('Among samples','Within samples'),c(nrow(dd3),nrow(dd3_2))))
+
+Fig4C <- ggplot(temp,aes(col=Group,fill = Group,x = DeltaRTL, y = ..scaled..))+
+    geom_density(alpha = 0.2)+
+  theme_classic()+
+    theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        legend.position = c(0.22,0.78),
+        legend.title = element_blank(),
+        axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black')) +
+  scale_colour_manual(values = c('darkgrey','gold'))+
+  scale_fill_manual(values = c('darkgrey','gold'))+
+  geom_vline(xintercept = 0,lty = 2)+
+  annotate('text',x = -5,y = 2,label = 'C')
 
 pdf('Figures/Figure 4.pdf',
-    width = 8,
-    height = 8,
+    width = 4,
+    height = 9,
     paper = 'a4')
-multiplot(Fig4A,Fig4B,Fig4C,layout=matrix(c(1,2,3,3),nrow=2,byrow=T))
+multiplot(Fig4A,Fig4B,Fig4C,layout=matrix(c(1,2,3),nrow=3,byrow=T))
 dev.off()
@@
 
-temp <- subset(dd,DeltaAge !=0)
-reps <- tapply(temp$RTL,temp$BirdID,length)
-keep <- names(reps[reps>3])
-temp <- subset(temp,BirdID %in% keep)
+#First calculate Delta RTL 
+temp <- subset(dd,!duplicated(paste0(BirdID,Agemonths)))
+counts <- with(temp,tapply(RTL,BirdID,length))
+c3 <- names(counts[counts >= 2])
 
-mins <- tapply(temp$Agemonths,temp$BirdID,min)
-keep2 <- names(mins[mins <5])
-temp <- subset(temp,BirdID %in% keep2)
-temp <- subset(temp,!duplicated(paste0(BirdID,Agemonths)))
+dd3 <- subset(temp,BirdID %in% c3)
 
-ggplot(temp,aes(x = Agemonths,y = RTL,group = BirdID,col=factor(BirdID)))+
-geom_point()+
-geom_line()+
-theme(legend.position='none')
+ggplot(dd3,aes(x = Agemonths,y = RTL, group = BirdID))+
+  geom_line(alpha = 0.5)
+
+dd3$DeltaRTL <- NA
+
+for(i in 1:nrow(dd3))
+{
+  cd <- subset(dd3,BirdID == dd3$BirdID[i])
+  cd <- cd[order(cd$Agemonths),]
+
+  birdpos <- which(cd$BloodID == dd3$BloodID[i])
+
+  if(birdpos == 1)
+  {
+    dd3$DeltaRTL[i] <- 0
+  } else
+  {
+      prevbird <- cd[birdpos-1,'RTL']
+  dd3$DeltaRTL[i] <- dd3$RTL[i]-prevbird
+  dd3$MidAge[i] <- mean(c(dd3$Agemonths[i],cd[birdpos-1,'Agemonths']))
+  dd3$MinAge[i] <- cd[birdpos-1,'Agemonths']
+  dd3$TimeDiff[i] <- dd3$Agemonths[i] - cd[birdpos-1,'Agemonths']
+  }
+
+}
+
+temp <- subset(dd3,DeltaRTL != 0)
+
+boxplot(DeltaRTL~SurvivedNext,data=subset(temp,MinAge>1))
+summary(lm(DeltaRTL~RemainingLife,data=subset(temp,MinAge>1)))
+
+plot(DeltaRTL~LayYear,data=subset(temp,MinAge>1 ))
+
+temp <- subset(dd0,!duplicated(paste0(BloodID,RTL)))
+
+temp <- subset(temp,RTL<2)
+temp <- subset(temp,RTL > 0.04)
+
+counts <- with(temp,tapply(RTL,BloodID,length))
+c3 <- names(counts[counts %in% c(2:5)])
+
+dd3_2 <- subset(temp,BloodID %in% c3)
+dd3_2 <- dd3_2[order(dd3_2$BloodID),]
+counts <- with(dd3_2,tapply(RTL,BloodID,length))
+
+dd3_2$birdpos <- unlist(sapply(counts,function(x) c(1:x)))
+
+dd3_2$DeltaRTL <- NA
+
+for(i in 1:nrow(dd3_2))
+{
+  cd <- subset(dd3_2,BloodID == dd3_2$BloodID[i])
+
+  if(dd3_2$birdpos[i] == 1)
+  {
+    dd3_2$DeltaRTL[i] <- 0
+  } else
+  {
+      prevbird <- subset(cd,birdpos == dd3_2$birdpos[i]-1)$RTL
+  dd3_2$DeltaRTL[i] <- dd3_2$RTL[i]-prevbird
+  }
+
+}
+
+dd3 <- subset(dd3,DeltaRTL !=0)
+dd3_2 <- subset(dd3_2,DeltaRTL!=0)
+
+dd3_2 <- dd3_2[1:nrow(dd3),]
+
+temp <- data.frame(DeltaRTL = c(dd3$DeltaRTL,dd3_2$DeltaRTL),Group = rep(c('Among samples','Within samples'),c(nrow(dd3),nrow(dd3_2))))
+
+ggplot(temp,aes(col=Group,fill = Group,x = DeltaRTL))+
+    geom_histogram()+
+  theme_classic()+
+  scale_colour_manual(values = c('darkgrey','gold'))+
+  scale_fill_manual(values = c('darkgrey','gold'))+
+  geom_vline(xintercept = 0,lty = 2)

### 2016-05-06 Scripts/SW juvenile telomeres IPC clean.R

@@ -4,13 +4,14 @@
 #################################################################################
 
 #Only use Ellie's samples
-#dd0 <- subset(dd0,Whodunnit == 'EAF')
+dd0 <- subset(dd0,Whodunnit == 'EAF')
 
 #Average repeats of blood samples
 av <- ave(dd0$RTL,c(dd0$BloodID,dd0$Status,dd0$PlateID))
-dd0$RTL <- av
+dd0$RTL2 <- av
 dd <- dd0[!(duplicated(dd0$BloodID)),]
-
+dd$RTL <- dd$RTL2
+dd$RTL2 <- NULL
 
 # Weird variable names ----------------------------------------------------
 
@@ -185,7 +186,6 @@ terr <- terr[complete.cases(terr),] #Get rid of blank rows
 insects <- subset(insects,FieldPeriodID != 26)
 dens$SPsize <- with(dens,(PsizeFP-mean(PsizeFP))/sd(PsizeFP))
 
-terrmeans 
 
 juv$TQ <- NA
 juv$TQI <- NA

### 2016-05-09 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
More edits

@@
         legend.background = element_rect(fill = grey(0.9))) +
+  geom_abline(slope = 1,intercept = 0,lty = 2)+
   geom_point(col='grey')+
@@
 c3 <- names(counts[counts >= 2])
-
 dd3 <- subset(temp,BirdID %in% c3)
-
-ggplot(dd3,aes(x = Agemonths,y = RTL, group = BirdID))+
-  geom_line(alpha = 0.5)
-
-dd3$DeltaRTL <- NA
+DeltaRTL <- NA
 
@@
 
-temp <- subset(dd3,DeltaRTL != 0)
-
-boxplot(DeltaRTL~SurvivedNext,data=subset(temp,MinAge>1))
-summary(lm(DeltaRTL~RemainingLife,data=subset(temp,MinAge>1)))
-
-plot(DeltaRTL~LayYear,data=subset(temp,MinAge>1 ))
-
 temp <- subset(dd0,!duplicated(paste0(BloodID,RTL)))
-
 temp <- subset(temp,RTL<2)
@@
 dd3_2 <- subset(dd3_2,DeltaRTL!=0)
-
 dd3_2 <- dd3_2[1:nrow(dd3),]
@@
   theme_classic()+
+  ylab ('Scaled density')+
+  xlab(expression(Delta*'RTL'))+
     theme(axis.text.x = element_text(size = 12), 
@@
   geom_vline(xintercept = 0,lty = 2)+
-  annotate('text',x = -5,y = 2,label = 'C')
+  annotate('text',x = -2,y = 1,label = 'C')
 
@@
 ```
-
-#First calculate Delta RTL 
-temp <- subset(dd,!duplicated(paste0(BirdID,Agemonths)))
-counts <- with(temp,tapply(RTL,BirdID,length))
-c3 <- names(counts[counts >= 2])
-
-dd3 <- subset(temp,BirdID %in% c3)
-
-ggplot(dd3,aes(x = Agemonths,y = RTL, group = BirdID))+
-  geom_line(alpha = 0.5)
-
-dd3$DeltaRTL <- NA
-
-for(i in 1:nrow(dd3))
-{
-  cd <- subset(dd3,BirdID == dd3$BirdID[i])
-  cd <- cd[order(cd$Agemonths),]
-
-  birdpos <- which(cd$BloodID == dd3$BloodID[i])
-
-  if(birdpos == 1)
-  {
-    dd3$DeltaRTL[i] <- 0
-  } else
-  {
-      prevbird <- cd[birdpos-1,'RTL']
-  dd3$DeltaRTL[i] <- dd3$RTL[i]-prevbird
-  dd3$MidAge[i] <- mean(c(dd3$Agemonths[i],cd[birdpos-1,'Agemonths']))
-  dd3$MinAge[i] <- cd[birdpos-1,'Agemonths']
-  dd3$TimeDiff[i] <- dd3$Agemonths[i] - cd[birdpos-1,'Agemonths']
-  }
-
-}
-
-temp <- subset(dd3,DeltaRTL != 0)
-
-boxplot(DeltaRTL~SurvivedNext,data=subset(temp,MinAge>1))
-summary(lm(DeltaRTL~RemainingLife,data=subset(temp,MinAge>1)))
-
-plot(DeltaRTL~LayYear,data=subset(temp,MinAge>1 ))
-
-temp <- subset(dd0,!duplicated(paste0(BloodID,RTL)))
-
-temp <- subset(temp,RTL<2)
-temp <- subset(temp,RTL > 0.04)
-
-counts <- with(temp,tapply(RTL,BloodID,length))
-c3 <- names(counts[counts %in% c(2:5)])
-
-dd3_2 <- subset(temp,BloodID %in% c3)
-dd3_2 <- dd3_2[order(dd3_2$BloodID),]
-counts <- with(dd3_2,tapply(RTL,BloodID,length))
-
-dd3_2$birdpos <- unlist(sapply(counts,function(x) c(1:x)))
-
-dd3_2$DeltaRTL <- NA
-
-for(i in 1:nrow(dd3_2))
-{
-  cd <- subset(dd3_2,BloodID == dd3_2$BloodID[i])
-
-  if(dd3_2$birdpos[i] == 1)
-  {
-    dd3_2$DeltaRTL[i] <- 0
-  } else
-  {
-      prevbird <- subset(cd,birdpos == dd3_2$birdpos[i]-1)$RTL
-  dd3_2$DeltaRTL[i] <- dd3_2$RTL[i]-prevbird
-  }
-
-}
-
-dd3 <- subset(dd3,DeltaRTL !=0)
-dd3_2 <- subset(dd3_2,DeltaRTL!=0)
-
-dd3_2 <- dd3_2[1:nrow(dd3),]
-
-temp <- data.frame(DeltaRTL = c(dd3$DeltaRTL,dd3_2$DeltaRTL),Group = rep(c('Among samples','Within samples'),c(nrow(dd3),nrow(dd3_2))))
-
-ggplot(temp,aes(col=Group,fill = Group,x = DeltaRTL))+
-    geom_histogram()+
-  theme_classic()+
-  scale_colour_manual(values = c('darkgrey','gold'))+
-  scale_fill_manual(values = c('darkgrey','gold'))+
-  geom_vline(xintercept = 0,lty = 2)

### 2016-05-09 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
More home edits

@@
 
-#First calculate Delta RTL 
-temp <- subset(dd,!duplicated(paste0(BirdID,Agemonths)))
-counts <- with(temp,tapply(RTL,BirdID,length))
-c3 <- names(counts[counts >= 2])
-dd3 <- subset(temp,BirdID %in% c3)
-DeltaRTL <- NA
-
-for(i in 1:nrow(dd3))
-{
-  cd <- subset(dd3,BirdID == dd3$BirdID[i])
-  cd <- cd[order(cd$Agemonths),]
-
-  birdpos <- which(cd$BloodID == dd3$BloodID[i])
-
-  if(birdpos == 1)
-  {
-    dd3$DeltaRTL[i] <- 0
-  } else
-  {
-      prevbird <- cd[birdpos-1,'RTL']
-  dd3$DeltaRTL[i] <- dd3$RTL[i]-prevbird
-  dd3$MidAge[i] <- mean(c(dd3$Agemonths[i],cd[birdpos-1,'Agemonths']))
-  dd3$MinAge[i] <- cd[birdpos-1,'Agemonths']
-  dd3$TimeDiff[i] <- dd3$Agemonths[i] - cd[birdpos-1,'Agemonths']
-  }
-
-}
-
-temp <- subset(dd0,!duplicated(paste0(BloodID,RTL)))
-temp <- subset(temp,RTL<2)
-temp <- subset(temp,RTL > 0.04)
-
-counts <- with(temp,tapply(RTL,BloodID,length))
-c3 <- names(counts[counts %in% c(2:5)])
-
-dd3_2 <- subset(temp,BloodID %in% c3)
-dd3_2 <- dd3_2[order(dd3_2$BloodID),]
-counts <- with(dd3_2,tapply(RTL,BloodID,length))
-
-dd3_2$birdpos <- unlist(sapply(counts,function(x) c(1:x)))
-
-dd3_2$DeltaRTL <- NA
-
-for(i in 1:nrow(dd3_2))
-{
-  cd <- subset(dd3_2,BloodID == dd3_2$BloodID[i])
-
-  if(dd3_2$birdpos[i] == 1)
-  {
-    dd3_2$DeltaRTL[i] <- 0
-  } else
-  {
-      prevbird <- subset(cd,birdpos == dd3_2$birdpos[i]-1)$RTL
-  dd3_2$DeltaRTL[i] <- dd3_2$RTL[i]-prevbird
-  }
-
-}
-
-dd3 <- subset(dd3,DeltaRTL !=0)
-dd3_2 <- subset(dd3_2,DeltaRTL!=0)
-dd3_2 <- dd3_2[1:nrow(dd3),]
-
-temp <- data.frame(DeltaRTL = c(dd3$DeltaRTL,dd3_2$DeltaRTL),Group = rep(c('Among samples','Within samples'),c(nrow(dd3),nrow(dd3_2))))
-
-Fig4C <- ggplot(temp,aes(col=Group,fill = Group,x = DeltaRTL, y = ..scaled..))+
+Fig4C <- ggplot(ddL,aes(col=Group,fill = Group,x = DeltaRTL, y = ..scaled..))+
     geom_density(alpha = 0.2)+
@@
 
+ddL <- ddL[complete.cases(ddL),]
+
 ```

### 2016-05-09 Manuscript/SW_early_life_telomeres.Rmd
More home edits

@@
 $\Delta$RTL was significantly lower than zero in nestlings (t = `r round2(t.test(subset(Loss,Agemonths == 1)$Loss)$statistic)`; P `r round3(t.test(subset(Loss,Agemonths == 1)$Loss)$p.value)`), but not in fledglings (t = `r round2(t.test(subset(Loss,Agemonths == 6)$Loss)$statistic)`; P = `r round2(t.test(subset(Loss,Agemonths == 6)$Loss)$p.value)`) or subadults (t = `r round2(t.test(subset(Loss,Agemonths == 10)$Loss)$statistic)`; P = `r round2(t.test(subset(Loss,Agemonths == 10)$Loss)$p.value)`).
-In line with this, $\Delta$RTL increased with age in early life (R^2^ = `r round2(summary(lm(Loss~Agemonths,data=Loss))$r.squared)`; t = `r round2(summary(lm(Loss~Agemonths,data=Loss))$coefficients[2,3])`; P `r round3(summary(lm(Loss~Agemonths,data = Loss))$coefficients[2,4])`; Fig.
-4B), suggesting that telomere shortening was highest shortly after the nestling phase.
+In line with this, $\Delta$RTL increased with age in early life (R^2^ = `r round2(summary(lm(Loss~Agemonths,data=Loss))$r.squared)`; t = `r round2(summary(lm(Loss~Agemonths,data=Loss))$coefficients[2,3])`; P `r round3(summary(lm(Loss~Agemonths,data = Loss))$coefficients[2,4])`), suggesting that telomere shortening was highest shortly after the nestling phase, after which both telomere shortening and telomere lenghtening were observed (Fig.
+4B).
+
+To test whether the apparent telomere lengthening in our dataset could be explained by measurement error, we used the entire Seychelles warbler telomere dataset to compare variance in telomere length among repeat measurements of samples (N = XX samples) to the variance observed among different samples of the same individual (see supplementary material XX for details).
+We found significantly higher variance among individuals compared to among samples (Levene's test XXX).
+Importantly, splitting the longitudinal data revealed that not only did we observe significantly more shortening within individuals compared to within samples (XXX), but also significantly more lengthening (XXX).
 

### 2016-05-10 Manuscript/SW_early_life_telomeres.Rmd
More edits

--- 2016-04-22 Manuscript/SW_early_life_telomeres.Rmd
@@
 
+Because we observed apparent telomere lenghtening in our data, we used the entire Seychelles warbler dataset to test whether this lengthening could be explained by measurement error.
+We used all birds (adults and juveniles) with at least two telomere measurements (N = `r nrow(subset(ddL,Group == 'Among samples'))` samples from `r length(unique(subset(ddL,Group == 'Among samples')$BirdID))` birds).
+We then calculated $\Delta$RTL for each pair of samples (for example, for a bird with three logitudinal measurements throughout their life we calculated the change in RTL between the first and second sample and between the second and third sample).
+We then compared these longitudinal changes in telomere length to sampling error, using all samples where we had at least two repeat telomere measurements for the same sample (N = `r nrow(subset(ddL,Group == 'Within samples'))` measurements from `r length(unique(subset(ddL,Group == 'Within samples')$BirdID))` samples).
+
 We used linear mixed models with model averaging to explore how spatial variation in early-life environmental and social conditions influenced telomere length within cohorts.
@@
 Running this analysis separetly for juvenile age classes revealed that this relationship was not significant for nestlings (R^2^ = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(Loss,Agemonths == 1)))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(Loss,Agemonths == 1)))$coefficients[2,3])`; P = `r round2(summary(lm(RTL1~RTL+Agemonths,data = subset(Loss,Agemonths == 1)))$coefficients[2,4])`), fledglings (R^2^ = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(Loss,Agemonths == 6)))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(Loss,Agemonths == 6)))$coefficients[2,3])`; P = `r round2(summary(lm(RTL1~RTL+Agemonths,data = subset(Loss,Agemonths == 6)))$coefficients[2,4])`) or subadults (R^2^ = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(Loss,Agemonths == 10)))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(Loss,Agemonths == 10)))$coefficients[2,3])`; P = `r round2(summary(lm(RTL1~RTL+Agemonths,data = subset(Loss,Agemonths == 10)))$coefficients[2,4])`).
-$\Delta$RTL was significantly lower than zero in nestlings (t = `r round2(t.test(subset(Loss,Agemonths == 1)$Loss)$statistic)`; P `r round3(t.test(subset(Loss,Agemonths == 1)$Loss)$p.value)`), but not in fledglings (t = `r round2(t.test(subset(Loss,Agemonths == 6)$Loss)$statistic)`; P = `r round2(t.test(subset(Loss,Agemonths == 6)$Loss)$p.value)`) or subadults (t = `r round2(t.test(subset(Loss,Agemonths == 10)$Loss)$statistic)`; P = `r round2(t.test(subset(Loss,Agemonths == 10)$Loss)$p.value)`).
-In line with this, $\Delta$RTL increased with age in early life (R^2^ = `r round2(summary(lm(Loss~Agemonths,data=Loss))$r.squared)`; t = `r round2(summary(lm(Loss~Agemonths,data=Loss))$coefficients[2,3])`; P `r round3(summary(lm(Loss~Agemonths,data = Loss))$coefficients[2,4])`; Fig.
-4B), suggesting that telomere shortening was highest shortly after the nestling phase.
+$\Delta$RTL was significantly lower than zero in nestlings (t = `r round2(t.test(subset(Loss,Agemonths == 1)$Loss)$statistic)`; P `r round3(t.test(subset(Loss,Agemonths == 1)$Loss)$p.value)`), but not in fledglings (t = `r round2(t.test(subset(Loss,Agemonths == 6)$Loss)$statistic)`; P = `r round2(t.test(subset(Loss,Agemonths == 6)$Loss)$p.value)`) or subadults (t = `r round2(t.test(subset(Loss,Agemonths == 10)$Loss)$statistic)`; P = `r round2(t.test(subset(Loss,Agemonths == 10)$Loss)$p.value)`), suggesting that telomere shortening was highest shortly after the nestling phase, after which both telomere shortening and telomere lenghtening were observed (Fig.
+4B).
+
+To test whether the apparent telomere lengthening in our dataset could be explained by measurement error, we used the entire Seychelles warbler telomere dataset to compare variance in telomere length among repeat measurements of samples to the variance observed among different samples of the same individual (see methods for details).
+We found significantly higher variance among individuals compared to among samples (Levene's test XXX).
+Importantly, splitting the longitudinal data revealed that not only did we observe significantly more shortening within individuals compared to within samples (XXX), but also significantly more lengthening (XXX).
 
 $\Delta$RTL did not differ significantly among cohorts (F = `r round2(anova(lm(Loss~factor(FieldPeriodID),data = Loss))[1,4])`; P `r round3(anova(lm(Loss~factor(FieldPeriodID),data = Loss))[1,5])`), although power for this analysis was limited as longitudinal sample sizes within cohorts were low.
-Model selection revealed that the top model explaining $\Delta$RTL within cohorts contained only age (Table S2).
-This model was a better fit than the null model ($\Delta$AICc = `r round2(AICc(lmer(Loss~1+(1|FieldPeriodID)+(1|TerritoryID),data=Loss,REML = FALSE))-AICc(lmer(Loss~Agemonths+(1|FieldPeriodID)+(1|TerritoryID),data=Loss,REML = FALSE)))`), and only the age effect was significant over the top model set (Fig.
-4C).
+Model selection revealed that the top model explaining $\Delta$RTL within cohorts contained only the number of non-helping subordinates (Table S2), although model was not a significantly better fit than the null model ($\Delta$AICc = `r round2(AICc(lmer(Loss~1+(1|FieldPeriodID)+(1|TerritoryID),data=Loss,REML = FALSE))-AICc(lmer(Loss~NonHelper+(1|FieldPeriodID)+(1|TerritoryID),data=Loss,REML = FALSE)))`).
 

### 2016-05-10 Manuscript/SW_early_life_telomeres_SOM.Rmd
More edits

@@
 #Run the model
-lm1 <- lmer(RTL ~ Agemonths * (Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ) + (1|FieldPeriodID)+(1|TerritoryID)+(1|PlateID),
+lm1 <- lmer(cenTL ~ Agemonths * (Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ) + (1|FieldPeriodID)+(1|TerritoryID)+(1|PlateID),
             data = juv,

### 2016-05-10 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
More edits

@@
 #Run the model
-lm2 <- lmer(RTL ~ Agemonths*(Tarsus) + Helper + NonHelper + Sex + BodyMass  + TQ  + (1|FieldPeriodID) + (1|TerritoryID)+(1|PlateID),
+lm2 <- lmer(RTL ~ Agemonths * (Tarsus) + Helper + NonHelper + Sex + BodyMass  + TQ  + (1|FieldPeriodID) + (1|TerritoryID)+(1|PlateID),
             data = temp,
@@
 
-Fig4B <- ggplot(temp,aes(x = Agemonths,y = RTL,group = BirdID,col=factor(BirdID)))+
+Fig4B <- ggplot(temp,aes(x = Agemonths,y = CqGAPDH,group = BirdID,col=factor(BirdID)))+
 geom_point()+
@@
         axis.line.y = element_line(colour = 'black'),
-        legend.position = 'none')+
+        legend.position = 'none')
   annotate('text',x = -5,y = 2,label = 'B')
 
-Fig4C <- ggplot(ddL,aes(col=Group,fill = Group,x = DeltaRTL, y = ..scaled..))+
+Fig4C <- ggplot(ddL,aes(col=Group,fill = Group,y = ..scaled..,x = DeltaRTL))+
     geom_density(alpha = 0.2)+
@@
         axis.title.x = element_text(size = 14),
-        legend.position = c(0.22,0.78),
+        legend.position = 'none',
         legend.title = element_blank(),
@@
 
-ddL <- ddL[complete.cases(ddL),]
+```
 
-```
+ggplot(ddL,aes(y = DeltaTL,x = DeltaGAP, col = Group))+
+geom_point()+
+stat_smooth(method = 'lm')
+
+plot(DeltaTL~Agemonths,data=dd3)
+plot(DeltaGAP~Agemonths,data=dd3)
+
+plot(DeltaGAP~TimeDiff,data=dd3)
+
+hist(dd$CqGAPDH)
+
+plot(CqTelomere~CqGAPDH,data=dd)
+
+boxplot(CqTelomere~Ageclass,data=dd)
+boxplot(CqGAPDH~Ageclass,data=dd)
+
+ggplot(dd,aes(y = CqTelomere, x = CqGAPDH,col = Ageclass))+
+geom_point()+
+stat_smooth(method = 'lm')

### 2016-05-10 Scripts/SW juvenile telomeres IPC clean.R

@@ -7,8 +7,7 @@
 dd0 <- subset(dd0,Whodunnit == 'EAF')
 
 #Average repeats of blood samples
-av <- ave(dd0$RTL,c(dd0$BloodID,dd0$Status,dd0$PlateID))
-dd0$RTL2 <- av
+dd0$RTL2 <- ave(dd0$RTL,c(dd0$BloodID,dd0$Status,dd0$PlateID))
 dd <- dd0[!(duplicated(dd0$BloodID)),]
 dd$RTL <- dd$RTL2
 dd$RTL2 <- NULL
@@ -64,10 +63,6 @@ dd$Age[dd$Fledged != 'Adults'] <- 1
 
 dd$Tarsus <- NA
 dd$DeltaAge <- NA
-dd$Malaria <- NA
-
-malaria <- subset(malaria,!duplicated(BloodID))
-malaria <- subset(malaria,Consensus <2)
 
 for(i in 1:nrow(dd))
 {
@@ -78,10 +73,6 @@ for(i in 1:nrow(dd))
          dd$Tarsus[i] <- dd$RightTarsus[i])
   
   dd$DeltaAge[i] <- dd$Agemonths[i] - mean(subset(dd,BirdID == dd$BirdID[i])$Agemonths)
-  if(dd$BloodID[i] %in% malaria$BloodID)
-     {
-       dd$Malaria[i] <- subset(malaria,BloodID == dd$BloodID[i])$Consensus
-     }
 }
 
 dd$RightTarsus <- NULL
@@ -92,7 +83,10 @@ dd$LeftTarsus <- NULL
 # Remove unwanted data/outliers ----------------------------------------------------
 
 dd <- subset(dd,RTL<2)
-dd <- subset(dd,RTL > 0.04)
+dd <- subset(dd,RTL > 0.05)
+#dd <- subset(dd,CqTelomere <22)
+dd <- subset(dd,CqGAPDH < 25)
+
 dd <- subset(dd,BodyMass>5)
 dd <- subset(dd,Tarsus>17)
 
@@ -206,7 +200,7 @@ for(i in 1:nrow(juv))
       juv$TQ[i] <- log10(subset(cd,FieldPeriodID == juv$FieldPeriodID[i])$TQcorrected)
     } else
     {
-      log10(mean(cd$TQcorrected))
[…]

### 2016-05-10 Scripts/SW juvenile telomeres IPC load.r

@@ -26,4 +26,3 @@ status <- read.csv('Data/SW TL all breedstatus.csv')
 insects <- read.csv('Data/Insects.csv')
 dens <- read.table('Data/PsizeFP.txt')
 chickinfo <- read.csv('Data/Table chick info.csv')
-malaria <- read.csv('Data/Malaria.csv')

### 2016-05-12 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
More edits

@@
 
-ggplot(dd,aes(y = CqTelomere, x = CqGAPDH,col = Ageclass))+
+ggplot(dd,aes(y = CqGAPDH, x = Agemonths,col=Sex))+
 geom_point()+
 stat_smooth(method = 'lm')
+
+summary(lm(CqGAPDH~Agemonths,data=dd))
+summary(lm(CqTelomere~Agemonths,data=dd))
+summary(lm(RTL~Agemonths,data=dd))

### 2016-05-15 Manuscript/SW_early_life_telomeres.Rmd
Home edits

@@
 We then compared these longitudinal changes in telomere length to sampling error, using all samples where we had at least two repeat telomere measurements for the same sample (N = `r nrow(subset(ddL,Group == 'Within samples'))` measurements from `r length(unique(subset(ddL,Group == 'Within samples')$BirdID))` samples).
+We calculated $\Delta$RTL between pairs repeat measurements within samples in exactly the same way as for across samples, except that repeat measurements were ordered by the date at which samples were measured rather thjan by age.
 
@@
 To test whether the apparent telomere lengthening in our dataset could be explained by measurement error, we used the entire Seychelles warbler telomere dataset to compare variance in telomere length among repeat measurements of samples to the variance observed among different samples of the same individual (see methods for details).
-We found significantly higher variance among individuals compared to among samples (Levene's test XXX).
-Importantly, splitting the longitudinal data revealed that not only did we observe significantly more shortening within individuals compared to within samples (XXX), but also significantly more lengthening (XXX).
+We found significantly higher variance among individuals compared to among samples (Levene's test: F = `r round2(leveneTest(DeltaRTL~Group,data=ddL)['F value'][1,1])`; P `r round3(leveneTest(DeltaRTL~Group,data=ddL)['Pr(>F)'][1,1])`; Fig.
+4C).
+Importantly, splitting the longitudinal data into instances of shortening ($\Delta$RTL < 0) and lenghtning ($\Delta$RTL > 0) revealed that not only did we observe significantly more shortening within individuals compared to within samples (Wilcoxon test: W = `r round(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$statistic,0)`; P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$p.value)`), but also significantly more lengthening (W = `r as.character(round(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$statistic,0))`; P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$p.value)`).
 

### 2016-05-15 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Home edits

--- 2016-05-10 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
@@
 
-Fig4B <- ggplot(temp,aes(x = Agemonths,y = CqGAPDH,group = BirdID,col=factor(BirdID)))+
+Fig4B <- ggplot(temp,aes(x = Agemonths,y = RTL,group = BirdID,col=factor(BirdID)))+
 geom_point()+
@@
 ```
-
-ggplot(ddL,aes(y = DeltaTL,x = DeltaGAP, col = Group))+
-geom_point()+
-stat_smooth(method = 'lm')
-
-plot(DeltaTL~Agemonths,data=dd3)
-plot(DeltaGAP~Agemonths,data=dd3)
-
-plot(DeltaGAP~TimeDiff,data=dd3)
-
-hist(dd$CqGAPDH)
-
-plot(CqTelomere~CqGAPDH,data=dd)
-
-boxplot(CqTelomere~Ageclass,data=dd)
-boxplot(CqGAPDH~Ageclass,data=dd)
-
-ggplot(dd,aes(y = CqTelomere, x = CqGAPDH,col = Ageclass))+
-geom_point()+
-stat_smooth(method = 'lm')

### 2016-05-15 Manuscript/SW_early_life_telomeres.Rmd
Home edits

@@
 1.
-Understanding the costs of individual early-life experiences is fundamental to understanding life-history evolution and population dynmamics.
+Understanding the costs of individual early-life experiences is fundamental to understanding life-history evolution and population dynamics.
 Telomeres, the protective caps at the ends of chromosomes, shorten in response to oxidative stress, and telomere shortening is correlated with reduced survival.
@@
 We found that telomere length decreases with cross-sectional and longitudinal measures of age, and that telomere length decreases most rapidly early in life.
-Juvenile telomere length varied markedly among cohorts, but we found no evidence that this variation was related to temporal variation in food availability or population density.
+However, both cross-sectional and longitudinal data suggested that telomere lenghtening occurs in some individuals, and using a large number of repeated measurements we show statistically that this lengthening is unlikely to be explained by measurement error.
 
 4.
-At the individual level we found a large amount of variation in telomere length and rates of change, with no relationship between within-individual juvenile and adult telomere length.
-We found weak evidence that telomere dnegatively related to telomere length, but the majority of variation in juvenile telomere length and dynamics was unexplained.
+Juvenile telomere length varied markedly among cohorts, but we found no evidence that this variation was related to temporal variation in food availability or population density.
+At the individual level we found
+no relationship between within-individual juvenile and adult telomere length.
+We found weak evidence that territory quality was negatively related to telomere length, but the majority of variation in juvenile telomere length and dynamics was unexplained.
 We found no relationship between early-life telomere dynamics and survival to adulthood or later in life.
@@
 5.
-Our comprehensive, long-term data suggest that although telomeres shorten with age, in natual popuations telomere dynamics vary enormously over both space and time.
-Ascertaining what explains this variation will be a difficult but worhtwhile task, which will require combining long-term ecological and quantitative genetic data.
+Our comprehensive, long-term data suggest that although telomeres shorten with age, in natural populations telomere dynamics vary enormously over both space and time.
+Ascertaining what explains this variation will be a difficult but worthwhile task, which will require combining long-term ecological and quantitative genetic data.
 
@@
 We performed all analyses using R version 3.0.1 [@RDevelopmentCoreTeam2011].
-Our dataset differs from slightly from that of Barrett *et al.* [-@Barrett2013a] in that we are using RTL calculated driectly from the qPCR data rather than converting into an absolute measure , and we are including many additional samples.
+Our dataset differs from slightly from that of Barrett *et al.* [-@Barrett2013a] in that we are using RTL calculated directly from the qPCR data rather than converting into an absolute measure , and we are including many additional samples.
 We therefore first re-tested, using all available samples, whether telomere length decreases with age and $\Delta$age [a longitudinal measure based on within-subject centring; @VandePol2009] in the Seychelles warbler, using linear mixed models with individual ID as a random effect.
@@
 
-Because we observed apparent telomere lenghtening in our data, we used the entire Seychelles warbler dataset to test whether this lengthening could be explained by measurement error.
+Because we observed apparent telomere lengthening in our data, we used the entire Seychelles warbler dataset to test whether this lengthening could be explained by measurement error.
 We used all birds (adults and juveniles) with at least two telomere measurements (N = `r nrow(subset(ddL,Group == 'Among samples'))` samples from `r length(unique(subset(ddL,Group == 'Among samples')$BirdID))` birds).
-We then calculated $\Delta$RTL for each pair of samples (for example, for a bird with three logitudinal measurements throughout their life we calculated the change in RTL between the first and second sample and between the second and third sample).
+We then calculated $\Delta$RTL for each pair of samples (for example, for a bird with three longitudinal measurements throughout their life we calculated the change in RTL between the first and second sample and between the second and third sample).
 We then compared these longitudinal changes in telomere length to sampling error, using all samples where we had at least two repeat telomere measurements for the same sample (N = `r nrow(subset(ddL,Group == 'Within samples'))` measurements from `r length(unique(subset(ddL,Group == 'Within samples')$BirdID))` samples).
-We calculated $\Delta$RTL between pairs repeat measurements within samples in exactly the same way as for across samples, except that repeat measurements were ordered by the date at which samples were measured rather thjan by age.
+We calculated $\Delta$RTL between pairs repeat measurements within samples in exactly the same way as for across samples, except that repeat measurements were ordered by the date at which samples were measured rather than by age.
 
@@
 The top model was much better supported than a null model ($\Delta$AICc = `r round2(AICc(lmer(RTL~1+(1|FieldPeriodID)+(1|TerritoryID),data=juv,REML = FALSE))-AICc(lmer(RTL~Agemonths+TQ+Tarsus+Agemonths*Tarsus+(1|FieldPeriodID)+(1|TerritoryID),data=juv,REML = FALSE)))`).
-Model-averaged confidence intervals of the age and tarsus x age effects did not overlap zero, but the tarsus and territory qusality effects were not significant (Fig.
+Model-averaged confidence intervals of the age and tarsus x age effects did not overlap zero, but the tarsus and territory quality effects were not significant (Fig.
 3A).
@@
 4A).
-Running this analysis separetly for juvenile age classes revealed that this relationship was not significant for nestlings (R^2^ = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(Loss,Agemonths == 1)))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(Loss,Agemonths == 1)))$coefficients[2,3])`; P = `r round2(summary(lm(RTL1~RTL+Agemonths,data = subset(Loss,Agemonths == 1)))$coefficients[2,4])`), fledglings (R^2^ = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(Loss,Agemonths == 6)))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(Loss,Agemonths == 6)))$coefficients[2,3])`; P = `r round2(summary(lm(RTL1~RTL+Agemonths,data = subset(Loss,Agemonths == 6)))$coefficients[2,4])`) or subadults (R^2^ = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(Loss,Agemonths == 10)))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(Loss,Agemonths == 10)))$coefficients[2,3])`; P = `r round2(summary(lm(RTL1~RTL+Agemonths,data = subset(Loss,Agemonths == 10)))$coefficients[2,4])`).
-$\Delta$RTL was significantly lower than zero in nestlings (t = `r round2(t.test(subset(Loss,Agemonths == 1)$Loss)$statistic)`; P `r round3(t.test(subset(Loss,Agemonths == 1)$Loss)$p.value)`), but not in fledglings (t = `r round2(t.test(subset(Loss,Agemonths == 6)$Loss)$statistic)`; P = `r round2(t.test(subset(Loss,Agemonths == 6)$Loss)$p.value)`) or subadults (t = `r round2(t.test(subset(Loss,Agemonths == 10)$Loss)$statistic)`; P = `r round2(t.test(subset(Loss,Agemonths == 10)$Loss)$p.value)`), suggesting that telomere shortening was highest shortly after the nestling phase, after which both telomere shortening and telomere lenghtening were observed (Fig.
+Running this analysis separately for juvenile age classes revealed that this relationship was not significant for nestlings (R^2^ = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(Loss,Agemonths == 1)))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(Loss,Agemonths == 1)))$coefficients[2,3])`; P = `r round2(summary(lm(RTL1~RTL+Agemonths,data = subset(Loss,Agemonths == 1)))$coefficients[2,4])`), fledglings (R^2^ = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(Loss,Agemonths == 6)))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(Loss,Agemonths == 6)))$coefficients[2,3])`; P = `r round2(summary(lm(RTL1~RTL+Agemonths,data = subset(Loss,Agemonths == 6)))$coefficients[2,4])`) or subadults (R^2^ = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(Loss,Agemonths == 10)))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(Loss,Agemonths == 10)))$coefficients[2,3])`; P = `r round2(summary(lm(RTL1~RTL+Agemonths,data = subset(Loss,Agemonths == 10)))$coefficients[2,4])`).
+$\Delta$RTL was significantly lower than zero in nestlings (t = `r round2(t.test(subset(Loss,Agemonths == 1)$Loss)$statistic)`; P `r round3(t.test(subset(Loss,Agemonths == 1)$Loss)$p.value)`), but not in fledglings (t = `r round2(t.test(subset(Loss,Agemonths == 6)$Loss)$statistic)`; P = `r round2(t.test(subset(Loss,Agemonths == 6)$Loss)$p.value)`) or subadults (t = `r round2(t.test(subset(Loss,Agemonths == 10)$Loss)$statistic)`; P = `r round2(t.test(subset(Loss,Agemonths == 10)$Loss)$p.value)`), suggesting that telomere shortening was highest shortly after the nestling phase, after which both telomere shortening and telomere lengthening were observed (Fig.
 4B).

### 2016-05-16 Manuscript/SW_early_life_telomeres.Rmd
More edits

@@
 We then calculated $\Delta$RTL for each pair of samples (for example, for a bird with three longitudinal measurements throughout their life we calculated the change in RTL between the first and second sample and between the second and third sample).
-We then compared these longitudinal changes in telomere length to sampling error, using all samples where we had at least two repeat telomere measurements for the same sample (N = `r nrow(subset(ddL,Group == 'Within samples'))` measurements from `r length(unique(subset(ddL,Group == 'Within samples')$BirdID))` samples).
-We calculated $\Delta$RTL between pairs repeat measurements within samples in exactly the same way as for across samples, except that repeat measurements were ordered by the date at which samples were measured rather than by age.
+We then compared these longitudinal changes in telomere length to sampling error, using all samples where we had at least two repeat telomere measurements for the same sample (i.e.
+completely seperate reactions run on separate plates: N = `r nrow(subset(ddL,Group == 'Within samples'))` measurements from `r length(unique(subset(ddL,Group == 'Within samples')$BirdID))` samples).
+We calculated $\Delta$RTL between pairs of repeat measurements within samples in exactly the same way as for across samples, except that repeat measurements were ordered by the date at the qPCR was run rather than by age.
 
@@
 To test whether the apparent telomere lengthening in our dataset could be explained by measurement error, we used the entire Seychelles warbler telomere dataset to compare variance in telomere length among repeat measurements of samples to the variance observed among different samples of the same individual (see methods for details).
-We found significantly higher variance among individuals compared to among samples (Levene's test: F = `r round2(leveneTest(DeltaRTL~Group,data=ddL)['F value'][1,1])`; P `r round3(leveneTest(DeltaRTL~Group,data=ddL)['Pr(>F)'][1,1])`; Fig.
+We found significantly higher variance in telomere length over individual lifetimes compared to among sample replicates (Levene's test: F = `r round2(leveneTest(DeltaRTL~Group,data=ddL)['F value'][1,1])`; P `r round3(leveneTest(DeltaRTL~Group,data=ddL)['Pr(>F)'][1,1])`; Fig.
 4C).

### 2016-05-16 Manuscript/SW_early_life_telomeres_SOM.Rmd
Home edits
More edits

--- 2016-04-19 Manuscript/SW_early_life_telomeres_SOM.Rmd
@@
 #Run the model
-lm1 <- lmer(Loss ~ Agemonths*(Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ) + (1|FieldPeriodID),
-            data = temp,
+lm1 <- glmer(L ~ Agemonths*(Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ) + (1|FieldPeriodID) + (1|TerritoryID),
+            data = subset(temp,Loss<0),
             REML = FALSE)
@@
 ```
+
+ggplot(d5,aes(x = RTL,col = factor(LayYear)))+
+geom_density()
+
+with(juv,tapply(RTL,LayYear,length))
+
+d5 <- subset(juv,LayYear == 2002)
+plot(RTL~BodyMass,data=d5)
+
+lm1 <- lmer(RTL ~ Agemonths*(Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ) + (1|TerritoryID),
+            data = d5,
+            REML = FALSE)
+lm1 <- standardize(lm1)
+
+#Model averaging
+lm2 <- dredge(lm1)
+
+head(lm2)
+
+plot(RTL~BodyMass,data=d5)

### 2016-05-16 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Home edits

@@
 
+temp <- Loss
+temp$LossL <- ifelse(temp$Loss>0,'Lengthening','Shortening')
+
+Fig4D <- ggplot(temp,aes(col=LossL,fill = LossL,y = Loss,x = TimeDiff))+
+    geom_point()+
+  stat_smooth(method = 'lm')+
+  theme_classic()+
+  xlab ('Follow up time (years)')+
+  ylab(expression(Delta*'RTL'))+
+    theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        legend.position = c(0.8,0.9),
+        legend.title = element_blank(),
+        axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black')) +
+  scale_colour_manual(values = c('darkgrey','gold'))+
+  scale_fill_manual(values = c('darkgrey','gold'))+
+    geom_abline(slope = 0,intercept = 0,lty = 2)+
+  annotate('text',x = 0,y = 1,label = 'D')
+
 pdf('Figures/Figure 4.pdf',
-    width = 4,
-    height = 9,
+    width = 8,
+    height = 7,
     paper = 'a4')
-multiplot(Fig4A,Fig4B,Fig4C,layout=matrix(c(1,2,3),nrow=3,byrow=T))
+multiplot(Fig4A,Fig4B,Fig4C,Fig4D,layout=matrix(c(1,2,3,4),nrow=2,byrow=T))
 dev.off()

### 2016-05-16 Scripts/SW juvenile telomeres IPC clean.R

@@ -343,6 +343,7 @@ dd3$TimeDiff <- NA
 dd3$DeltaRTL <- NA
 dd3$DeltaTL <- NA
 dd3$DeltaGAP <- NA
+dd3$BloodID1 <- NA
 
 for(i in 1:nrow(dd3))
 {
@@ -361,6 +362,7 @@ for(i in 1:nrow(dd3))
     dd3$DeltaGAP[i] <- nextbird$CqGAPDH - dd3$CqGAPDH[i]
     dd3$DeltaTL[i] <- nextbird$CqTelomere - dd3$CqTelomere[i]
     dd3$TimeDiff[i] <- cd[birdpos+1,'Agemonths'] - dd3$Agemonths[i]
+    dd3$BloodID1[i] <- nextbird$BloodID
   }
   
   
@@ -387,6 +389,7 @@ dd3_2$birdpos <- unlist(sapply(counts,function(x) c(1:x)))
 dd3_2$DeltaRTL <- NA
 dd3_2$DeltaTL <- NA
 dd3_2$DeltaGAP <- NA
+dd3_2$BloodID1 <- NA
 
 for(i in 1:nrow(dd3_2))
 {
@@ -401,6 +404,7 @@ for(i in 1:nrow(dd3_2))
     dd3_2$DeltaRTL[i] <- nextbird$RTL - dd3_2$RTL[i]
     dd3_2$DeltaGAP[i] <- nextbird$CqGAPDH - dd3_2$CqGAPDH[i]
     dd3_2$DeltaTL[i] <- nextbird$CqTelomere - dd3_2$CqTelomere[i]
+    dd3_2$BloodID1[i] <- nextbird$BloodID
   }
   
   
@@ -411,7 +415,9 @@ dd3 <- subset(dd3,!is.na(DeltaRTL))
 dd3_2 <- subset(dd3_2,!is.na(DeltaRTL))
 
 
-ddL <- data.frame(BirdID = c(dd3$BirdID,dd3_2$BloodID),
+ddL <- data.frame(BirdID = c(dd3$BirdID,dd3_2$BirdID),
+                  BloodID = c(dd3$BloodID,dd3_2$BloodID),
+                  BloodID1 = c(dd3$BloodID1,dd3_2$BloodID1),
                   DeltaRTL = c(dd3$DeltaRTL,dd3_2$DeltaRTL),
[…]

### 2016-05-17 Manuscript/SW_early_life_telomeres.Rmd
More edits

@@
 We found that telomere length decreases with cross-sectional and longitudinal measures of age, and that telomere length decreases most rapidly early in life.
-However, both cross-sectional and longitudinal data suggested that telomere lenghtening occurs in some individuals, and using a large number of repeated measurements we show statistically that this lengthening is unlikely to be explained by measurement error.
+However, both cross-sectional and longitudinal data suggested that telomere length increases occurs in some individuals, and using a large number of repeated measurements we show statistically that this increase is unlikely to be explained by qPCR measurement error.
 
 4.
-Juvenile telomere length varied markedly among cohorts, but we found no evidence that this variation was related to temporal variation in food availability or population density.
-At the individual level we found
-no relationship between within-individual juvenile and adult telomere length.
-We found weak evidence that territory quality was negatively related to telomere length, but the majority of variation in juvenile telomere length and dynamics was unexplained.
+Juvenile telomere length varied markedly among cohorts, but we found no evidence that this variation was related to temporal variation in the environment.
+At the individual found weak evidence that territory quality was negatively related to telomere length, but the majority of variation in juvenile telomere length and dynamics was unexplained.
 We found no relationship between early-life telomere dynamics and survival to adulthood or later in life.
@@
 Our comprehensive, long-term data suggest that although telomeres shorten with age, in natural populations telomere dynamics vary enormously over both space and time.
-Ascertaining what explains this variation will be a difficult but worthwhile task, which will require combining long-term ecological and quantitative genetic data.
+Furthermore, Ascertaining what explains this variation will be a difficult but worthwhile task, which will require combining long-term ecological and quantitative genetic data.
 
@@
 Efficiencies (mean $\pm$ s.d.) for our telomere and GAPDH reactions were (`r round2(mean(dd$EffTelomere))` $\pm$ `r round2(sd(dd$EffTelomere))`) and (`r round2(mean(dd$EffGAPDH))` $\pm$ `r round2(sd(dd$EffGAPDH))`) respectively.
-Inter-plate repeatability of RTL, based on 392 samples measured at least twice, was 0.78 (CI = 0.73-0.82).
+Inter-plate repeatability of RTL, based on `r nrow(subset(ddL,Group == 'Within samples'))` samples measured at least twice, was 0.78 (CI = 0.73-0.82).
 
@@
 
-Early-life RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~cenTL+Agemonths+(1|PlateID)+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~cenTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,4])`), with no interaction effect between RTL and juvenile age (estimate = `r round2(summary(glmer(SurvivedNext~cenTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~cenTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,4])`), suggesting that there were no age-specific effects of juvenile RTL on survival to adulthood.
+Early-life RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~RTL+Agemonths+(1|PlateID)+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,4])`), with no interaction effect between RTL and juvenile age (estimate = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,4])`), suggesting that there were no age-specific effects of juvenile RTL on survival to adulthood.
 To test whether cohort-level effects on RTL influenced survival we calculated cohort-level RTL (i.e.
@@
 
-Cox regression showed no effect of individual-level or cohort-level RTL on survival later in life (individual-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cenTL+Agemonths+frailty(FieldPeriodID))))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cenTL+Agemonths+frailty(FieldPeriodID))))$coefficients[1,6])`; cohort-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL*Agemonths+RTL+frailty(FieldPeriodID))))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL*Agemonths+RTL+frailty(FieldPeriodID))))$coefficients[1,6])`), and no interaction between RTL and juvenile age on survival (individual-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cenTL*Agemonths+frailty(FieldPeriodID))))$coefficients[4,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cenTL*Agemonths+frailty(FieldPeriodID))))$coefficients[4,6])`).
+Cox regression showed no effect of individual-level or cohort-level RTL on survival later in life (individual-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL+Agemonths+frailty(FieldPeriodID))))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL+Agemonths+frailty(FieldPeriodID))))$coefficients[1,6])`; cohort-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL*Agemonths+RTL+frailty(FieldPeriodID))))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL*Agemonths+RTL+frailty(FieldPeriodID))))$coefficients[1,6])`), and no interaction between RTL and juvenile age on survival (individual-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL*Agemonths+frailty(FieldPeriodID))))$coefficients[4,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL*Agemonths+frailty(FieldPeriodID))))$coefficients[4,6])`).
 $\Delta$RTL was not related to longer-term survival (estimate `r round3(summary(coxph(with(subset(Loss,LayYear<2008),Surv(RemainingLife2+1,Died)~Loss+Agemonths)))$coefficients[1,1])`, P = `r round2(summary(coxph(with(subset(Loss,LayYear<2008),Surv(RemainingLife2+1,Died)~Loss)))$coefficients[1,5])`).

### 2016-05-17 Manuscript/SW_early_life_telomeres_SOM.Rmd
More edits

--- 2016-04-22 Manuscript/SW_early_life_telomeres_SOM.Rmd
@@
 lm1 <- lmer(Loss ~ Agemonths*(Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ) + (1|FieldPeriodID) + (1|TerritoryID),
-            data = temp,
+            data = Loss,
             REML = FALSE)
@@
 ```
+
+ggplot(dd,aes(x = RTL, col = Ageclass))+
+geom_density()

### 2016-05-17 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
More edits

--- 2016-05-15 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
@@
        aes(x = RTL,
-           y = RTL1)) +
+           y = RTL1,
+           col = LossF)) +
   ylab('Adult RTL') + 
   xlab('Juvenile RTL') +
+    xlim(c(0,3))+
+  ylim(c(0,3))+
   theme_classic() +
@@
         axis.title.x = element_text(size = 14),
-        legend.position = c(0.82,0.88),
+        legend.position = c('none'),
         legend.title = element_text(size = 12),
         axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black'),
-        legend.background = element_rect(fill = grey(0.9))) +
+        axis.line.y = element_line(colour = 'black')) +
   geom_abline(slope = 1,intercept = 0,lty = 2)+
-  geom_point(col='grey')+
-  stat_smooth(method = 'lm',col='darkgrey')+
-  annotate('text',x = 0,y = 1.8,label = 'A')
+  geom_point()+
+  scale_colour_manual(values = c('darkgrey','red'))+
+  annotate('text',x = 0,y = 3,label = 'A')
 
@@
 
+temp$Agemonths <- temp$Agemonths/12
+
+mycols <- topo.colors(length(keep2))
+
 Fig4B <- ggplot(temp,aes(x = Agemonths,y = RTL,group = BirdID,col=factor(BirdID)))+
@@
 geom_line()+
-  xlab('Age (months)') +
+  xlab('Age (years)') +
   theme_classic() +
@@
         axis.line.y = element_line(colour = 'black'),
-        legend.position = 'none')
-  annotate('text',x = -5,y = 2,label = 'B')
+        legend.position = 'none')+
+  scale_colour_brewer(palette = 'Set1')+
+  annotate('text',x = 0,y = 2,label = 'B')
 
@@
 
+temp <- Loss
+
+Fig4D <- ggplot(temp,aes(col=LossF,fill = LossF,y = Loss,x = TimeDiff))+
+    geom_point()+
+  stat_smooth(method = 'lm')+
+  theme_classic()+
+  xlab ('Follow up time (years)')+
+  ylab(expression(Delta*'RTL'))+
+    theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        legend.position = c('none'),
+        legend.title = element_blank(),
+        axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black')) +
+  scale_colour_manual(values = c('darkgrey','red'))+
+  scale_fill_manual(values = c('darkgrey','red'))+
+    geom_abline(slope = 0,intercept = 0,lty = 2)+
+  annotate('text',x = 0,y = 1,label = 'D')
+
 pdf('Figures/Figure 4.pdf',
-    width = 4,
-    height = 9,
+    width = 8,
+    height = 7,
     paper = 'a4')
-multiplot(Fig4A,Fig4B,Fig4C,layout=matrix(c(1,2,3),nrow=3,byrow=T))
+multiplot(Fig4A,Fig4B,Fig4C,Fig4D,layout=matrix(c(1,2,3,4),nrow=2,byrow=T))
 dev.off()

### 2016-05-17 Scripts/SW juvenile telomeres IPC clean.R

@@ -82,9 +82,9 @@ dd$LeftTarsus <- NULL
 
 # Remove unwanted data/outliers ----------------------------------------------------
 
-dd <- subset(dd,RTL<2)
+
 dd <- subset(dd,RTL > 0.05)
-#dd <- subset(dd,CqTelomere <22)
+dd <- subset(dd,CqTelomere <22)
 dd <- subset(dd,CqGAPDH < 25)
 
 dd <- subset(dd,BodyMass>5)
@@ -281,6 +281,7 @@ Loss$TROC <- with(Loss,Loss/TimeDiff)
 #
 Loss <- subset(Loss,TimeDiff>(365/2))
 Loss$TimeDiff <- Loss$TimeDiff/365
+Loss$LossF <- ifelse(Loss$Loss>0,'Lengthening','Shortening')
 
 # Get rid of stuff not to be used -----------------------------------------
 

### 2016-05-17 Manuscript/SW_early_life_telomeres.Rmd
More edits

@@
 We calculated $\Delta$RTL between pairs of repeat measurements within samples in exactly the same way as for across samples, except that repeat measurements were ordered by the date at the qPCR was run rather than by age.
+Secondly, if telomere lengthening is an artifact of measurement error, it is likely to occur more frequently when there is a short follow up time between samples [@Steenstrup2013].
+We therefore tested whether telomere lengthening was more likely to occur at short sampling intervals by relating the sample follow up time to both the probability of telomere lengthening (GLM with a binomial error structure, telomere lengthening yes/no as response), and to the dgree of lengthening (LM using only samples where RTL increased over time, $\Delta$RTL as response).
 
@@
 For the cross-sectional data we used lifespan (in years) as survival time, while for the longitudinal data we used remaining lifespan from the second sampling event.
-We ran these models with cohort as a frailty term to exclude cohort effects, and only included cohorts in which at least 70 percent of individuals had died (all years up to and including 2007).
+We ran these models with cohort as a frailty term to exclude cohort effects, and only included cohorts in which at least 70 percent of individuals had died (all years up to and including 2008).
 
@@
 3B).
-Territory quality was negatively related to RTL, particularly in nestlings, although this effect was very weak (Fig.
+Territory quality was negatively related to RTL, particularly in nestlings, although this effect was weak (Fig.
 3C).
@@
 Importantly, splitting the longitudinal data into instances of shortening ($\Delta$RTL < 0) and lenghtning ($\Delta$RTL > 0) revealed that not only did we observe significantly more shortening within individuals compared to within samples (Wilcoxon test: W = `r round(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$statistic,0)`; P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$p.value)`), but also significantly more lengthening (W = `r as.character(round(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$statistic,0))`; P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$p.value)`).
+We found no evidence that the likelihood of telomere lengthening was related to the follow up time between sampling events (GLM: t = `r round2(summary(glm(DeltaRTLF~TimeDiff,data=dd3,family = 'binomial'))$coefficients[2,3])`; P = `r round2(summary(glm(DeltaRTLF~TimeDiff,data=dd3,family = 'binomial'))$coefficients[2,4])`), and considering those samples that did show increases in $\Delta$RTL, the degree of increase was not related to follow up time (LM: t = `r round2(summary(lm(DeltaRTL~TimeDiff,data=subset(dd3,DeltaRTL>0)))$coefficients[2,3])`; P = `r round2(summary(lm(DeltaRTL~TimeDiff,data=subset(dd3,DeltaRTL>0)))$coefficients[2,4])`).
 
-$\Delta$RTL did not differ significantly among cohorts (F = `r round2(anova(lm(Loss~factor(FieldPeriodID),data = Loss))[1,4])`; P `r round3(anova(lm(Loss~factor(FieldPeriodID),data = Loss))[1,5])`), although power for this analysis was limited as longitudinal sample sizes within cohorts were low.
-Model selection revealed that the top model explaining $\Delta$RTL within cohorts contained only the number of non-helping subordinates (Table S2), although model was not a significantly better fit than the null model ($\Delta$AICc = `r round2(AICc(lmer(Loss~1+(1|FieldPeriodID)+(1|TerritoryID),data=Loss,REML = FALSE))-AICc(lmer(Loss~NonHelper+(1|FieldPeriodID)+(1|TerritoryID),data=Loss,REML = FALSE)))`).
+$\Delta$RTL did not differ significantly among cohorts (F = `r round2(anova(lm(Loss~factor(FieldPeriodID),data = Loss))[1,4])`; P = `r round(anova(lm(Loss~factor(FieldPeriodID),data = Loss))[1,5])`), although power for this analysis was limited as longitudinal sample sizes within cohorts were low.
+Model selection revealed that the top model explaining $\Delta$RTL was the null model (Table S2).
 
@@
 
-Cox regression showed no effect of individual-level or cohort-level RTL on survival later in life (individual-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL+Agemonths+frailty(FieldPeriodID))))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL+Agemonths+frailty(FieldPeriodID))))$coefficients[1,6])`; cohort-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL*Agemonths+RTL+frailty(FieldPeriodID))))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL*Agemonths+RTL+frailty(FieldPeriodID))))$coefficients[1,6])`), and no interaction between RTL and juvenile age on survival (individual-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL*Agemonths+frailty(FieldPeriodID))))$coefficients[4,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL*Agemonths+frailty(FieldPeriodID))))$coefficients[4,6])`).
+Cox regression showed no effect of individual-level or cohort-level RTL on survival later in life (individual-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL+Agemonths+frailty(FieldPeriodID))))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL+Agemonths+frailty(FieldPeriodID))))$coefficients[1,6])`; cohort-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL*Agemonths+RTL+frailty(FieldPeriodID))))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL*Agemonths+RTL+frailty(FieldPeriodID))))$coefficients[1,6])`), and no interaction between RTL and juvenile age on survival (individual-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL*Agemonths+frailty(FieldPeriodID))))$coefficients[4,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL*Agemonths+frailty(FieldPeriodID))))$coefficients[4,6])`; cohort-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL*Agemonths+frailty(FieldPeriodID))))$coefficients[4,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL*Agemonths+frailty(FieldPeriodID))))$coefficients[4,6])`).
 $\Delta$RTL was not related to longer-term survival (estimate `r round3(summary(coxph(with(subset(Loss,LayYear<2008),Surv(RemainingLife2+1,Died)~Loss+Agemonths)))$coefficients[1,1])`, P = `r round2(summary(coxph(with(subset(Loss,LayYear<2008),Surv(RemainingLife2+1,Died)~Loss)))$coefficients[1,5])`).
@@
 ###Discussion
-Here we use one of the largest datasets to date to assess the relationships between early-life conditions, telomere length and survival in a closed population of Seychelles warblers.
+Here we use one of the largest datasets to date to assess the relationships between early-life conditions, telomere dynamics and survival in a closed population of Seychelles warblers.
 We found that, as in other studies, telomeres are lost at the greatest rate early in life.
@@
 While noise in telomere measurements and problems with power are a problem in any study using qPCR to measure telomere length [@Nussey2014], our sample sizes and levels of repeatability compare favourably with other studies of natural populations that have found significant effects of the early-life environment on telomere length [e.g.
-@Fairlie2015].
+@Fairlie2016].
 More likely is that early-life RTL in the Seychelles warbler is explained by a set of variables that we have not measured.

### 2016-05-17 Manuscript/SW_early_life_telomeres_SOM.Rmd
More edits

@@
 #Run the model
-lm1 <- lmer(Loss ~ Agemonths*(Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ) + (1|FieldPeriodID) + (1|TerritoryID),
-            data = Loss,
+lm1 <- lmer(Loss ~ RTL+ Agemonths*(Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ) + (1|FieldPeriodID) + (1|TerritoryID),
+            data = temp,
             REML = FALSE)
@@
 ```
-
-ggplot(dd,aes(x = RTL, col = Ageclass))+
-geom_density()

### 2016-05-17 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
More edits

@@
   geom_point()+
-  scale_colour_manual(values = c('darkgrey','red'))+
+  scale_colour_manual(values = c('navy','red'))+
   annotate('text',x = 0,y = 3,label = 'A')
@@
 
-Fig4D <- ggplot(temp,aes(col=LossF,fill = LossF,y = Loss,x = TimeDiff))+
+Fig4D <- ggplot(temp,aes(col=LossF,y = Loss,x = TimeDiff))+
     geom_point()+
-  stat_smooth(method = 'lm')+
   theme_classic()+
@@
   ylab(expression(Delta*'RTL'))+
+  ylim(c(-2,2))+
     theme(axis.text.x = element_text(size = 12), 
@@
         axis.line.y = element_line(colour = 'black')) +
-  scale_colour_manual(values = c('darkgrey','red'))+
-  scale_fill_manual(values = c('darkgrey','red'))+
+  scale_colour_manual(values = c('navy','red'))+
     geom_abline(slope = 0,intercept = 0,lty = 2)+
-  annotate('text',x = 0,y = 1,label = 'D')
+  annotate('text',x = 0,y = 2,label = 'D')
 

### 2016-05-17 Scripts/SW juvenile telomeres IPC clean.R

@@ -83,9 +83,9 @@ dd$LeftTarsus <- NULL
 # Remove unwanted data/outliers ----------------------------------------------------
 
 
-dd <- subset(dd,RTL > 0.05)
-dd <- subset(dd,CqTelomere <22)
-dd <- subset(dd,CqGAPDH < 25)
+#dd <- subset(dd,RTL > 0.05)
+dd <- subset(dd,CqTelomere <28)
+dd <- subset(dd,CqGAPDH < 28)
 
 dd <- subset(dd,BodyMass>5)
 dd <- subset(dd,Tarsus>17)
@@ -100,7 +100,7 @@ dd$SurvivedNext <- ifelse(dd$RemainingLife<1,0,1)
 
 dd$SurvivedNext <- ifelse(dd$RemainingLife>1,1,0)
 dd$Lifespan <- (dd$DeathYear-dd$LayYear)+1
-dd$Died <- ifelse(dd$DeathYear<2013,1,0)
+dd$Died <- ifelse(dd$DeathYear<2014,1,0)
 
 
 
@@ -307,7 +307,7 @@ juvseason <- ddply(juv,
 juvseason <- subset(juvseason,n>4)
 
 with(juv,tapply(Died,LayYear,mean))
-juv9 <- subset(juv,LayYear<2008)
+juv9 <- subset(juv,LayYear<2009)
 juv13 <- subset(juv,LayYear<2013)
 
 
@@ -415,6 +415,8 @@ for(i in 1:nrow(dd3_2))
 dd3 <- subset(dd3,!is.na(DeltaRTL))
 dd3_2 <- subset(dd3_2,!is.na(DeltaRTL))
 
+dd3$DeltaRTLF <- ifelse(dd3$DeltaRTL>0,1,0)
+
 
 ddL <- data.frame(BirdID = c(dd3$BirdID,dd3_2$BirdID),
                   BloodID = c(dd3$BloodID,dd3_2$BloodID),
@@ -423,7 +425,6 @@ ddL <- data.frame(BirdID = c(dd3$BirdID,dd3_2$BirdID),
                   DeltaTL = c(dd3$DeltaTL,dd3_2$DeltaTL),
                   DeltaGAP = c(dd3$DeltaGAP,dd3_2$DeltaGAP),
                   Group = rep(c('Among samples','Within samples'),c(nrow(dd3),nrow(dd3_2))))
-
 
 rm(temp,dd3_2)
 

### 2016-06-09 Manuscript/SW_early_life_telomeres.Rmd
More edits

@@
 
-Lewis G. Spurgin^1,2^,  Kat Bebbington^1^, Eleanor A. Fairfield^1^, Martijn Hammers^3^, Jan Komdeur^3^, Terry Burke^4^, Hannah, L. Dugdale^3,4^, and David S. Richardson^1,5,^.
+Lewis G. Spurgin^1^,  Kat Bebbington^1^, Eleanor A. Fairfield^1^, Martijn Hammers^2^, Jan Komdeur^2^, Terry Burke^3^, Hannah, L. Dugdale^2,4^, and David S. Richardson^1,5,^.
 
 1. School of Biological Sciences, University of East Anglia, Norwich Research Park, NR4 7TJ, United Kingdom
-2. Department of Zoology, Edward Grey Institute, University of Oxford, Oxford, UK
-3. Groningen Institute for Evolutionary Life Sciences, University of Groningen, Groningen, The Netherlands
-4. Department of Animal and Plant Sciences, University of Sheffield, Sheffield, UK
+2. Groningen Institute for Evolutionary Life Sciences, University of Groningen, Groningen, The Netherlands
+3. Department of Animal and Plant Sciences, University of Sheffield, Sheffield, UK
+4. School of Biology, The Faculty of Biological Sciences, University of Leeds, Leeds LS2 9JT, UK
 5. Nature Seychelles, Roche Caiman, Mahé, Republic of Seychelles
@@
 
-Early-life RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~RTL+Agemonths+(1|PlateID)+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,4])`), with no interaction effect between RTL and juvenile age (estimate = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,4])`), suggesting that there were no age-specific effects of juvenile RTL on survival to adulthood.
+Early-life RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~cenTL+Agemonths+(1|PlateID)+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~cenTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,4])`), with no interaction effect between RTL and juvenile age (estimate = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,4])`), suggesting that there were no age-specific effects of juvenile RTL on survival to adulthood.
 To test whether cohort-level effects on RTL influenced survival we calculated cohort-level RTL (i.e.

### 2016-06-09 Scripts/SW juvenile telomeres IPC clean.R

@@ -83,7 +83,7 @@ dd$LeftTarsus <- NULL
 # Remove unwanted data/outliers ----------------------------------------------------
 
 
-#dd <- subset(dd,RTL > 0.05)
+dd <- subset(dd,RTL > 0.05)
 dd <- subset(dd,CqTelomere <28)
 dd <- subset(dd,CqGAPDH < 28)
 

### 2016-06-09 Scripts/SW juvenile telomeres IPC load.r

@@ -20,7 +20,7 @@ library(magrittr)
 library(car)
 
 # Load data ---------------------------------------------------------------
-dd0 <- read.csv('Data/REL_TL_EB_EAF_25 Feb FULL DATA.csv') #main dataset
+  dd0 <- read.csv('Data/REL_TL_EB_EAF_24 May aft TL queries.csv') #main dataset
 terr <- read.csv('Data/SW TL all territories.csv')
 status <- read.csv('Data/SW TL all breedstatus.csv')
 insects <- read.csv('Data/Insects.csv')

### 2016-06-14 Manuscript/SW_early_life_telomeres_SOM.Rmd
More edits

@@
 ```{r Table S2}
-temp <- Loss
+temp <- dd3
 
 #Run the model
-lm1 <- lmer(Loss ~ RTL+ Agemonths*(Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ) + (1|FieldPeriodID) + (1|TerritoryID),
+lm1 <- glmer(DeltaRTLF ~ Agemonths+(Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ) + (1|FieldPeriodID) + (1|TerritoryID),
             data = temp,
+            family = 'binomial',
             REML = FALSE)

### 2016-06-14 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
More edits

@@
 
-**Figure 4** Longitudinal telomere dynamics in juvenile Seychelles warblers.
-**A** Juvenile telomere length in relation to adult telomere length measured in the same individual.
+**Figure 4** Longitudinal telomere dynamics in subset(dd,Agemonths<12)enile Seychelles warblers.
+**A** subset(dd,Agemonths<12)enile telomere length in relation to adult telomere length measured in the same individual.
 **B** $\Delta$RTL in relation to the time between sampling events.
@@
 
-**Figure 5** Factors affecting $\Delta$RTL in juvenile Seychelles warblers.
+**Figure 5** Factors affecting $\Delta$RTL in subset(dd,Agemonths<12)enile Seychelles warblers.
 **A** Model averaged estimates and 95% confidence intervals for explanatory terms used in a linear mixed model with $\Delta$RTL as the response variable (see main text for details; TQ = territory quality).
@@
 
-temp <- juv
+temp <- subset(dd,Agemonths<12)
 temp$LayYear <- factor(temp$LayYear)
@@
 
-temp <- juv
+temp <- dd
 
@@
 
-temp <- juv
+temp <- subset(dd,Agemonths<12)
 
@@
   ylab('Adult RTL') + 
-  xlab('Juvenile RTL') +
+  xlab('subset(dd,Agemonths<12)enile RTL') +
     xlim(c(0,3))+
@@
 
-temp <- Loss
+temp <- dd3
 
-Fig4D <- ggplot(temp,aes(col=LossF,y = Loss,x = TimeDiff))+
+Fig4D <- ggplot(temp,aes(col=DeltaRTLF,y = DeltaRTL,x = TimeDiff))+
     geom_point()+
@@
 ```
+
+temp <- with(dd3,tapply(DeltaRTL,Agemonths,mean))
+tempSE <- with(dd3,tapply(DeltaRTL,Agemonths,se))*1.96
+
+plot(temp~names(temp),type = 'l',ylim = c(-0.8,0.8))
+points(temp+tempSE~names(temp),type = 'l',lty = 2)
+points(temp-tempSE~names(temp),type = 'l',lty = 2)
+abline(0,0,col = 'red',lwd = 2)
+
+plot(DeltaRTL~TimeDiff,data = dd3)

### 2016-06-14 Scripts/SW juvenile telomeres IPC clean.R

@@ -129,46 +129,46 @@ adults <- droplevels(subset(dd,Ageclass == 'A'))
 
 
 #Centre telomere length by birth year
-juv$cenTL <- NA
-juv$cohortTL <- NA
-for(i in 1:nrow(juv))
+dd$cenTL <- NA
+dd$cohortTL <- NA
+for(i in 1:nrow(dd))
 {
-  currentdata <- subset(juv,FieldPeriodID == FieldPeriodID[i])
+  currentdata <- subset(dd,FieldPeriodID == FieldPeriodID[i])
   if(nrow(currentdata>4))
   {
-    juv$cenTL[i] <- (juv$RTL[i] - mean(currentdata$RTL))/sd(currentdata$RTL)
-    juv$cohortTL[i] <- median(currentdata$RTL)  
+    dd$cenTL[i] <- (dd$RTL[i] - mean(currentdata$RTL))/sd(currentdata$RTL)
+    dd$cohortTL[i] <- median(currentdata$RTL)  
   }
 
 }
 
-mymed <- median(juv$cenTL,na.rm=T)
-juv$cenTLF <- ifelse(juv$cenTL < mymed,'Short telomeres','Long telomeres')
+mymed <- median(dd$cenTL,na.rm=T)
+dd$cenTLF <- ifelse(dd$cenTL < mymed,'Short telomeres','Long telomeres')
 
-mymed <- median(juv$cohortTL,na.rm=T)
-juv$coTLF <- ifelse(juv$cohortTL < mymed,'Short telomeres','Long telomeres')
+mymed <- median(dd$cohortTL,na.rm=T)
+dd$coTLF <- ifelse(dd$cohortTL < mymed,'Short telomeres','Long telomeres')
 
 
 rm(mymed)
 
-juv <- subset(juv,!(is.na(cenTL)))
+dd <- subset(dd,!(is.na(cenTL)))
 
 # Helpers and social group size -------------------------------------------
 
-status <- subset(status,BreedGroupID %in% juv$BreedGroupID)
+status <- subset(status,BreedGroupID %in% dd$BreedGroupID)
 
-for(i in 1:nrow(juv))
+for(i in 1:nrow(dd))
 {
-  currentBG <- juv$BreedGroupID[i]
-  currentdata <- subset(subset(status,BreedGroupID == currentBG),BirdID !=juv$BirdID[i])
-  juv$GroupSize[i] <- nrow(currentdata)
-  juv$Helper[i] <- nrow(subset(currentdata,Status == 'H'))
[…]

### 2016-06-16 Manuscript/SW_early_life_telomeres_SOM.Rmd
More edits - lifelong dynamics!

@@
 
+temp <- subset(juv,!is.na(cenTQ))
+
 #Run the model
-lm1 <- lmer(RTL ~ Agemonths * (Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ) + (1|FieldPeriodID)+(1|TerritoryID)+(1|PlateID),
-            data = juv,
+lm1 <- lmer(RTL ~ Agemonths * (Tarsus + Helper + NonHelper + Sex + BodyMass  + cenTQ) + (1|BirdID)+(1|FieldPeriodID)+(1|TerritoryID),
+            data = temp,
             REML = FALSE)
@@
 #Run the model
-lm1 <- glmer(DeltaRTLF ~ Agemonths+(Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ) + (1|FieldPeriodID) + (1|TerritoryID),
+lm1 <- lmer(DeltaRTL ~ Agemonths*(Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ) + (1|BirdID) + (1|FieldPeriodID) + (1|TerritoryID),
             data = temp,
-            family = 'binomial',
             REML = FALSE)

### 2016-06-16 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
More edits - lifelong dynamics!

@@
 
-**Figure 4** Longitudinal telomere dynamics in subset(dd,Agemonths<12)enile Seychelles warblers.
-**A** subset(dd,Agemonths<12)enile telomere length in relation to adult telomere length measured in the same individual.
+**Figure 4** Longitudinal telomere dynamics in juvenile Seychelles warblers.
+**A** juvenile telomere length in relation to adult telomere length measured in the same individual.
 **B** $\Delta$RTL in relation to the time between sampling events.
@@
 
-**Figure 5** Factors affecting $\Delta$RTL in subset(dd,Agemonths<12)enile Seychelles warblers.
+**Figure 5** Factors affecting $\Delta$RTL in juvenile Seychelles warblers.
 **A** Model averaged estimates and 95% confidence intervals for explanatory terms used in a linear mixed model with $\Delta$RTL as the response variable (see main text for details; TQ = territory quality).
@@
 
-temp <- dd
+temp <- subset(dd,!is.na(cenTQ))
+temp$LogAge <- log(temp$Agemonths)
 
 #Run the model
-lm2 <- lmer(RTL ~ Agemonths * (Tarsus) + Helper + NonHelper + Sex + BodyMass  + TQ  + (1|FieldPeriodID) + (1|TerritoryID)+(1|PlateID),
+lm2 <- lmer(RTL ~  LogAge* (Tarsus + NonHelper + Sex + BodyMass  + cenTQ)  + (1|FieldPeriodID) + (1|BirdID) + (1|TerritoryID),
             data = temp,
@@
 
-temp <- subset(dd,Agemonths<12)
-
 temp[,'Helpers'] <- ifelse(temp$Helper>0,'Present', 'Absent')
@@
        aes(x = Tarsus,
-           y = RTL,
-           colour=`Age (months)`,
-           fill = `Age (months)`,
-           shape = `Age (months)`)) +
+           y = RTL)) +
   ylab('RTL') + 
@@
         legend.position = 'top') +
-  geom_point()+
-  stat_smooth(method = 'lm')+
-  scale_colour_manual(values = c('grey','darkorange','darkgreen'))+
-  scale_fill_manual(values = c('grey','darkorange','darkgreen'))+
+  geom_point(col = 'grey')+
+  stat_smooth(method = 'lm',col = 'darkgrey')+
   annotate('text',x = 17,y = 2.1,label = 'B')
@@
 Fig3C <- ggplot(temp,
-       aes(col = `Age (months)`,
-           fill =  `Age (months)`,
+       aes(
            y = RTL,
-           x=TQ)) +
+           x=Agemonths,
+       col = factor(Helper))) +
   theme_classic() +
@@
   geom_point()+
-  stat_smooth(method = 'lm')+
+  stat_smooth(method = 'lm',formula = y~log(x))+
   scale_colour_manual(values = c('grey','darkorange','darkgreen'))+
@@
 
-temp <- Loss
+temp <- dd3
 temp[,'Age (months)'] <- factor(temp$Agemonths)
@@
   ylab('Adult RTL') + 
-  xlab('subset(dd,Agemonths<12)enile RTL') +
+  xlab('juvenile RTL') +
     xlim(c(0,3))+

### 2016-06-16 Scripts/SW juvenile telomeres IPC clean.R

@@ -107,8 +107,8 @@ dd$Died <- ifelse(dd$DeathYear<2014,1,0)
 # Exclude weird seasons and early years ----------------------------------
 
 dd <- subset(dd,LayYear>1997)
-#dd <- subset(dd,Season == 'Major')
-dd <- subset(dd,FieldPeriodID != 27)
+dd <- subset(dd,Season == 'Major')
+#dd <- subset(dd,FieldPeriodID != 27)
 
 # Sex ---------------------------------------------------------------------
 
@@ -116,15 +116,6 @@ dd$Sex <- ifelse(dd$SexEstimate == 1,'Males','Females')
 
 
 
-
-
-
-
-# Subset juveniles --------------------------------------------------------------
-
-
-juv <- droplevels(subset(dd,Ageclass %in% c('CH','FL','OFL','SA')))
-adults <- droplevels(subset(dd,Ageclass == 'A'))
 
 
 
@@ -151,8 +142,6 @@ dd$coTLF <- ifelse(dd$cohortTL < mymed,'Short telomeres','Long telomeres')
 
 rm(mymed)
 
-dd <- subset(dd,!(is.na(cenTL)))
-
 # Helpers and social group size -------------------------------------------
 
 status <- subset(status,BreedGroupID %in% dd$BreedGroupID)
@@ -163,7 +152,6 @@ for(i in 1:nrow(dd))
   currentdata <- subset(subset(status,BreedGroupID == currentBG),BirdID !=dd$BirdID[i])
   dd$GroupSize[i] <- nrow(currentdata)
   dd$Helper[i] <- nrow(subset(currentdata,Status == 'H'))
-  dd$Otherdds[i] <- nrow(subset(currentdata,Status %in% c('CH','FL','OFL')))
   dd$NonHelper[i] <- nrow(subset(currentdata,Status %in% c('AB','ABX')))
 }
 
@@ -179,16 +167,13 @@ dd$Helper[dd$Helper>1] <- 1
 terr <- terr[complete.cases(terr),] #Get rid of blank rows
 insects <- subset(insects,FieldPeriodID != 26)
 dens$SPsize <- with(dens,(PsizeFP-mean(PsizeFP))/sd(PsizeFP))
-
[…]

### 2016-06-16 Manuscript/SW_early_life_telomeres.Rmd
More edits

@@
 We measured telomere lengths using a total of `r nrow(dd)` unique samples from juvenile and adult Seychelles warblers.
-Of these, `r nrow(juv)` samples were taken cross-sectionally from birds caught within their first year of life, and of these juvenile birds, we had additional longitudinal samples for `r nrow(Loss)`.
+Of these, `r nrow(juv)` samples were taken cross-sectionally from birds caught within their first year of life, and of these juvenile birds, we had additional longitudinal samples for `r length(Loss)`.
 Efficiencies (mean $\pm$ s.d.) for our telomere and GAPDH reactions were (`r round2(mean(dd$EffTelomere))` $\pm$ `r round2(sd(dd$EffTelomere))`) and (`r round2(mean(dd$EffGAPDH))` $\pm$ `r round2(sd(dd$EffGAPDH))`) respectively.

### 2016-06-16 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
More edits

@@
            y = RTL1,
-           col = LossF)) +
+           col = factor(DeltaRTLF))) +
   ylab('Adult RTL') + 
@@
 
-Fig4D <- ggplot(temp,aes(col=DeltaRTLF,y = DeltaRTL,x = TimeDiff))+
+Fig4D <- ggplot(temp,aes(col=factor(DeltaRTLF),y = DeltaRTL,x = TimeDiff))+
     geom_point()+

### 2016-06-16 Scripts/SW juvenile telomeres IPC clean.R

@@ -253,6 +253,7 @@ for(i in 1:nrow(dd3))
   {
     nextbird <- cd[birdpos+1,]
     dd3$DeltaRTL[i] <- nextbird$RTL - dd3$RTL[i]
+    dd3$RTL1[i] <- nextbird$RTL
     dd3$DeltaGAP[i] <- nextbird$CqGAPDH - dd3$CqGAPDH[i]
     dd3$DeltaTL[i] <- nextbird$CqTelomere - dd3$CqTelomere[i]
     dd3$TimeDiff[i] <- cd[birdpos+1,'Agemonths'] - dd3$Agemonths[i]

### 2016-06-17 Manuscript/SW_early_life_telomeres.Rmd
More edits

--- 2016-06-09 Manuscript/SW_early_life_telomeres.Rmd
@@
 We measured telomere lengths using a total of `r nrow(dd)` unique samples from juvenile and adult Seychelles warblers.
-Of these, `r nrow(juv)` samples were taken cross-sectionally from birds caught within their first year of life, and of these juvenile birds, we had additional longitudinal samples for `r nrow(Loss)`.
+Of these, `r length(unique(subset(dd,Agemonths<12)$BirdID))` samples were taken cross-sectionally from birds caught within their first year of life.
 Efficiencies (mean $\pm$ s.d.) for our telomere and GAPDH reactions were (`r round2(mean(dd$EffTelomere))` $\pm$ `r round2(sd(dd$EffTelomere))`) and (`r round2(mean(dd$EffGAPDH))` $\pm$ `r round2(sd(dd$EffGAPDH))`) respectively.

### 2016-06-17 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
More edits

@@
 
-temp <- subset(dd,!is.na(cenTQ))
+temp <- juv_r
 temp$LogAge <- log(temp$Agemonths)
@@
 #Run the model
-lm2 <- lmer(RTL ~  LogAge* (Tarsus + NonHelper + Sex + BodyMass  + cenTQ)  + (1|FieldPeriodID) + (1|BirdID) + (1|TerritoryID),
+lm2 <- lmer(RTL ~  LogAge* (Helper)+Tarsus + NonHelper + Sex + BodyMass  +yday*LogAge + (1|FieldPeriodID) + (1|BirdID) + (1|TerritoryID),
             data = temp,
@@
 ```
-
-temp <- with(dd3,tapply(DeltaRTL,Agemonths,mean))
-tempSE <- with(dd3,tapply(DeltaRTL,Agemonths,se))*1.96
-
-plot(temp~names(temp),type = 'l',ylim = c(-0.8,0.8))
-points(temp+tempSE~names(temp),type = 'l',lty = 2)
-points(temp-tempSE~names(temp),type = 'l',lty = 2)
-abline(0,0,col = 'red',lwd = 2)
-
-plot(DeltaRTL~TimeDiff,data = dd3)

### 2016-06-17 Scripts/SW juvenile telomeres IPC clean.R

@@ -26,13 +26,14 @@ dd$CatchYear <- as.numeric(substr(dd$CatchDate,7,10))
 dd$DeathYear <- as.numeric(substr(dd$DeathDate,7,10))
 dd$CatchDate <- as.Date(dd$CatchDate,"%d/%m/%Y")
 dd$DeathDate <- as.Date(dd$DeathDate,"%d/%m/%Y")
-
+dd$CatchMonth <- as.numeric(substr(dd$CatchDate,6,7))
 
 
 dd$Season <- ifelse(as.numeric(format(dd$CatchDate,'%m')) %in% c(4:10),
                     'Major','Minor')
 dd <- subset(dd,Season == 'Major')
-
+dd <- subset(dd,CatchMonth %in% c(6:9))
+dd$yday <- as.POSIXlt(dd$CatchDate)$yday - min(as.POSIXlt(dd$CatchDate)$yday)
 
 # Age data ----------------------------------------------------------------
 
@@ -85,12 +86,14 @@ dd$LeftTarsus <- NULL
 
 dd <- subset(dd,RTL > 0.05)
 dd <- subset(dd,CqTelomere <28)
-dd <- subset(dd,CqGAPDH < 28)
+dd <- subset(dd,CqGAPDH < 27)
+dd <- subset(dd,RTL<3)
 
 dd <- subset(dd,BodyMass>5)
 dd <- subset(dd,Tarsus>17)
 
-
+hist(dd$CqTelomere)
+hist(dd$CqGAPDH)
 
 
 # Survival and lifespan ---------------------------------------------------
@@ -342,5 +345,5 @@ for(i in 1:nrow(addata))
   addata[i,toreplace] <- cd
 }
 
-juv <- rbind(juvdata,addata)
-
+juv_r <- rbind(juvdata,addata)
+juv <- juvdata

### 2016-06-27 Manuscript/SW_early_life_telomeres.Rmd
More edits

@@
 We measured telomere lengths using a total of `r nrow(dd)` unique samples from juvenile and adult Seychelles warblers.
-Of these, `r length(unique(subset(dd,Agemonths<12)$BirdID))` samples were taken cross-sectionally from birds caught within their first year of life.
 Efficiencies (mean $\pm$ s.d.) for our telomere and GAPDH reactions were (`r round2(mean(dd$EffTelomere))` $\pm$ `r round2(sd(dd$EffTelomere))`) and (`r round2(mean(dd$EffGAPDH))` $\pm$ `r round2(sd(dd$EffGAPDH))`) respectively.
@@
 
-#####Early-life RTL
-Using the cross-sectional data RTL decreased with age across the entire lifespan of the Seychelles warbler (LMM: t = `r round2(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,4])`; P `r round3(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,5])`), although there were clear increases in (cross-sectional) telomere length after one year of age and at several points later in life (Fig.
+#####Telomere dynamics and age
+Using the cross-sectional data RTL decreased with age across the entire lifespan of the Seychelles warbler (LMM: t = `r round2(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,4])`; P `r round3(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,5])`).
+We found that modelling a non-linear effect age explained almost double the variation in telomere length (R^2^ of linear regression of telomere length and log age = `r round3(summary(lm(RTL~log(Agemonths),data=dd))$r.squared)` than a linear model `r round3(summary(lm(RTL~Agemonths,data=dd))$r.squared)`.
+Also worth noting is that while the overall trend was for decreasing delomre length, there were clear increases in cross-sectional telomere length after one year of age and at several points later in life (Fig.
 1A).
@@
 
-Early-life RTL varied significantly among cohorts (Likelihood ratio test: F = `r round2(anova(lme(RTL~factor(FieldPeriodID),data=juv,random=~1|BirdID))[2,3])`; P `r round3(anova(lme(RTL~Agemonths,data=juv,random=~1|BirdID))[2,4])`).
-Cohort-level variation in juvenile RTL was not related to adult population density (LMM controlling for age and cohort: t = `r round2(summary(lme(RTL~Agemonths+Density,data=juv,random=~1|FieldPeriodID))$tTable[3,4])`; P = `r round3(summary(lme(RTL~Agemonths+Density,data=juv,random=~1|FieldPeriodID))$tTable[3,5])`), or to temporal variation in  food availability (t = `r round2(summary(lme(RTL~Agemonths+Insect,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,4])`; P = `r round3(summary(lme(RTL~Agemonths+Insect,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,5])`).
-
-Within cohorts, the top model explaining variation in early-life RTL contained age, territory quality, tarsus length, and the interaction between age and tarsus length (Table S1).
-The top model was much better supported than a null model ($\Delta$AICc = `r round2(AICc(lmer(RTL~1+(1|FieldPeriodID)+(1|TerritoryID),data=juv,REML = FALSE))-AICc(lmer(RTL~Agemonths+TQ+Tarsus+Agemonths*Tarsus+(1|FieldPeriodID)+(1|TerritoryID),data=juv,REML = FALSE)))`).
-Model-averaged confidence intervals of the age and tarsus x age effects did not overlap zero, but the tarsus and territory quality effects were not significant (Fig.
-3A).
-Tarsus length was negatively related to RTL in nestlings, unrelated to RTL in fledglings, and positively related to RTL in subadults (Fig.
-3B).
-Territory quality was negatively related to RTL, particularly in nestlings, although this effect was weak (Fig.
-3C).
-
-#####Early-life telomere shortening
 Longitudinal data showed that early-life RTL was not related to adult RTL (R^2^ = `r round2(summary(lm(RTL1~RTL+Agemonths,data=Loss))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data=Loss))$coefficients[2,3])`; P = `r round2(summary(lm(RTL1~RTL+Agemonths,data = Loss))$coefficients[2,4])`; Fig.
@@
 
+#####Early life environment and telomere dynamics
+Early-life RTL varied significantly among cohorts (Likelihood ratio test: F = `r round2(anova(lme(RTL~factor(FieldPeriodID),data=juv,random=~1|BirdID))[2,3])`; P `r round3(anova(lme(RTL~Agemonths,data=juv,random=~1|BirdID))[2,4])`).
+Cohort-level variation in juvenile RTL was not related to adult population density (LMM controlling for age and cohort: t = `r round2(summary(lme(RTL~Agemonths+Density,data=juv,random=~1|FieldPeriodID))$tTable[3,4])`; P = `r round3(summary(lme(RTL~Agemonths+Density,data=juv,random=~1|FieldPeriodID))$tTable[3,5])`), or to temporal variation in  food availability (t = `r round2(summary(lme(RTL~Agemonths+Insect,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,4])`; P = `r round3(summary(lme(RTL~Agemonths+Insect,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,5])`).
+
 $\Delta$RTL did not differ significantly among cohorts (F = `r round2(anova(lm(Loss~factor(FieldPeriodID),data = Loss))[1,4])`; P = `r round(anova(lm(Loss~factor(FieldPeriodID),data = Loss))[1,5])`), although power for this analysis was limited as longitudinal sample sizes within cohorts were low.
@@
 
+Within cohorts, the top model explaining variation in early-life RTL contained age, territory quality, tarsus length, and the interaction between age and tarsus length (Table S1).
+The top model was much better supported than a null model ($\Delta$AICc = `r round2(AICc(lmer(RTL~1+(1|FieldPeriodID)+(1|TerritoryID),data=juv,REML = FALSE))-AICc(lmer(RTL~Agemonths+TQ+Tarsus+Agemonths*Tarsus+(1|FieldPeriodID)+(1|TerritoryID),data=juv,REML = FALSE)))`).
+Model-averaged confidence intervals of the age and tarsus x age effects did not overlap zero, but the tarsus and territory quality effects were not significant (Fig.
+3A).
+Tarsus length was negatively related to RTL in nestlings, unrelated to RTL in fledglings, and positively related to RTL in subadults (Fig.
+3B).
+Territory quality was negatively related to RTL, particularly in nestlings, although this effect was weak (Fig.
+3C).
+
 #####Early-life telomere dynamics and survival
-
-Early-life RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~cenTL+Agemonths+(1|PlateID)+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~cenTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,4])`), with no interaction effect between RTL and juvenile age (estimate = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,4])`), suggesting that there were no age-specific effects of juvenile RTL on survival to adulthood.
+Early-life RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~cenTL+Agemonths+(1|PlateID)+(1|TerritoryID)+(1|FieldPeriodID),data=juv14,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~cenTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv14,family = 'binomial'))$coefficients[2,4])`), with no interaction effect between RTL and juvenile age (estimate = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv14,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv14,family = 'binomial'))$coefficients[4,4])`), suggesting that there were no age-specific effects of juvenile RTL on survival to adulthood.
 To test whether cohort-level effects on RTL influenced survival we calculated cohort-level RTL (i.e.
-the median RTL for each cohort) and added this term to a GLMM, and found that cohort-level RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~cohortTL+RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~cohortTL+RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,4])`).
+the median RTL for each cohort) and added this term to a GLMM, and found that cohort-level RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~cohortTL+RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv14,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~cohortTL+RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv14,family = 'binomial'))$coefficients[2,4])`).
 Using the longitudinal data, we found that $\Delta$RTL had no effect on survival to the year after the second sampling event (estimate = `r round2(summary(glm(SurvivedNext2~Loss+Agemonths,data = subset(Loss,LayYear<2013),family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glm(SurvivedNext2~Loss+Agemonths,data=subset(Loss,LayYear<2013),family = 'binomial'))$coefficients[2,4])`).
 
-Cox regression showed no effect of individual-level or cohort-level RTL on survival later in life (individual-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL+Agemonths+frailty(FieldPeriodID))))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL+Agemonths+frailty(FieldPeriodID))))$coefficients[1,6])`; cohort-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL*Agemonths+RTL+frailty(FieldPeriodID))))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL*Agemonths+RTL+frailty(FieldPeriodID))))$coefficients[1,6])`), and no interaction between RTL and juvenile age on survival (individual-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL*Agemonths+frailty(FieldPeriodID))))$coefficients[4,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL*Agemonths+frailty(FieldPeriodID))))$coefficients[4,6])`; cohort-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL*Agemonths+frailty(FieldPeriodID))))$coefficients[4,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL*Agemonths+frailty(FieldPeriodID))))$coefficients[4,6])`).
+summary(glmer(SurvivedNext~RTL+(1|BirdID),data=subset(dd,LayYear<2000), family = 'binomial'))
+
+Cox regression showed no effect of individual-level or cohort-level RTL on survival later in life (individual-level: estimate = `r round2(summary(coxph(with(juv12,Surv(Lifespan,Died)~RTL+Agemonths+frailty(FieldPeriodID))))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv12,Surv(Lifespan,Died)~RTL+Agemonths+frailty(FieldPeriodID))))$coefficients[1,6])`; cohort-level: estimate = `r round2(summary(coxph(with(juv12,Surv(Lifespan,Died)~cohortTL*Agemonths+RTL+frailty(FieldPeriodID))))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv12,Surv(Lifespan,Died)~cohortTL*Agemonths+RTL+frailty(FieldPeriodID))))$coefficients[1,6])`), and no interaction between RTL and juvenile age on survival (individual-level: estimate = `r round2(summary(coxph(with(juv12,Surv(Lifespan,Died)~RTL*Agemonths+frailty(FieldPeriodID))))$coefficients[4,1])`, P = `r round2(summary(coxph(with(juv12,Surv(Lifespan,Died)~RTL*Agemonths+frailty(FieldPeriodID))))$coefficients[4,6])`; cohort-level: estimate = `r round2(summary(coxph(with(juv12,Surv(Lifespan,Died)~cohortTL*Agemonths+frailty(FieldPeriodID))))$coefficients[4,1])`, P = `r round2(summary(coxph(with(juv12,Surv(Lifespan,Died)~cohortTL*Agemonths+frailty(FieldPeriodID))))$coefficients[4,6])`).
 $\Delta$RTL was not related to longer-term survival (estimate `r round3(summary(coxph(with(subset(Loss,LayYear<2008),Surv(RemainingLife2+1,Died)~Loss+Agemonths)))$coefficients[1,1])`, P = `r round2(summary(coxph(with(subset(Loss,LayYear<2008),Surv(RemainingLife2+1,Died)~Loss)))$coefficients[1,5])`).
@@
 ###References
+
+temp <- dd
+temp$y <- ifelse(temp$CqTelomere<15,0,1)
+
+summary(glm(y~Ageclass+Sex+Helper,data=temp,family = 'binomial'))
+
+ggplot(temp,aes(y = RTL,x = Ageclass))+
+geom_boxplot()

### 2016-06-27 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
More edits

@@
         legend.position = 'none')+
-  geom_point(col='grey')+
-  stat_smooth(method = 'lm',col='darkgrey')+
-  annotate('text',x = 1,y = 2.1,label = 'A')
+  geom_point()+
+    stat_smooth(col = 'darkgrey',lty = 3,se = F)+
+  stat_smooth(method = 'lm',col='darkgrey',formula = y~log(x),se = F)
+  annotate('text',x = 1,y = 2.8,label = 'A')
 
@@
 
-temp <- juv_r
+temp <- subset(juv_r,Season == 'Major')
 temp$LogAge <- log(temp$Agemonths)
@@
 #Run the model
-lm2 <- lmer(RTL ~  LogAge* (Helper)+Tarsus + NonHelper + Sex + BodyMass  +yday*LogAge + (1|FieldPeriodID) + (1|BirdID) + (1|TerritoryID),
+lm2 <- lmer(CqTelomere ~  LogAge* (Helper)+Tarsus + NonHelper + Sex + BodyMass  +yday*LogAge + (1|FieldPeriodID) + (1|BirdID) + (1|TerritoryID),
             data = temp,
@@
 temp[,'Age (months)'] <- factor(temp$Agemonths) 
+temp$Age2 <- ifelse(temp$Ageclass == 'A','Adults','Juveniles')
 
@@
        aes(x = Tarsus,
-           y = RTL)) +
+           y = RTL,
+           col = Age2)) +
   ylab('RTL') + 
@@
         legend.position = 'top') +
-  geom_point(col = 'grey')+
-  stat_smooth(method = 'lm',col = 'darkgrey')+
+  geom_point()+
+  stat_smooth(method = 'lm')+
   annotate('text',x = 17,y = 2.1,label = 'B')
@@
            y = RTL,
-           x=Agemonths,
-       col = factor(Helper))) +
+           x=yday,
+       col = (Ageclass))) +
   theme_classic() +
@@
   geom_point()+
-  stat_smooth(method = 'lm',formula = y~log(x))+
-  scale_colour_manual(values = c('grey','darkorange','darkgreen'))+
-  scale_fill_manual(values = c('grey','darkorange','darkgreen'))+
+   stat_smooth()+
+  # scale_colour_manual(values = c('grey','darkorange','darkgreen'))+
+  # scale_fill_manual(values = c('grey','darkorange','darkgreen'))+
   annotate('text',x = 3,y = 2.1,label = 'C')
@@
 temp <- subset(temp,BirdID %in% keep)
+ temp <- subset(temp,!duplicated(paste0(BirdID,Agemonths)))
 
@@
  temp <- subset(temp,BirdID %in% keep2)
-temp <- subset(temp,!duplicated(paste0(BirdID,Agemonths)))
 
@@
 Fig4B <- ggplot(temp,aes(x = Agemonths,y = RTL,group = BirdID,col=factor(BirdID)))+
-geom_point()+
-geom_line()+
+geom_line(col = 'darkgrey')+
   xlab('Age (years)') +
@@
         legend.position = 'none')+
-  scale_colour_brewer(palette = 'Set1')+
+  #scale_colour_brewer(palette = 'Set1')+
   annotate('text',x = 0,y = 2,label = 'B')

### 2016-06-27 Scripts/SW juvenile telomeres IPC clean.R

@@ -37,7 +37,7 @@ dd$yday <- as.POSIXlt(dd$CatchDate)$yday - min(as.POSIXlt(dd$CatchDate)$yday)
 
 # Age data ----------------------------------------------------------------
 
-dd$Age <- (dd$CatchYear-dd$LayYear)+1
+dd$Age <- (dd$CatchYear-dd$LayYear)
 
 #Sort out and order ageclass levels
 dd <- droplevels(dd[dd$Ageclass != '',])
@@ -51,11 +51,10 @@ dd$Fledged <- factor(dd$Fledged,levels = c('Nestlings','Fledglings','Adults'))
 
 dd$Agemonths <- ifelse(dd$Ageclass == 'CH',1,
                        ifelse(dd$Ageclass == 'FL',6,
-                              ifelse(dd$Ageclass == 'SA',10,dd$Age*12)))
+                              ifelse(dd$Ageclass == 'SA',10,dd$Age*12+6)))
 
-dd <- subset(dd,Agemonths!=12)
 
-dd$Age[dd$Fledged != 'Adults'] <- 1
+  dd$Age[dd$Fledged != 'Adults'] <- 1
 
 
 
@@ -92,8 +91,7 @@ dd <- subset(dd,RTL<3)
 dd <- subset(dd,BodyMass>5)
 dd <- subset(dd,Tarsus>17)
 
-hist(dd$CqTelomere)
-hist(dd$CqGAPDH)
+
 
 
 # Survival and lifespan ---------------------------------------------------
@@ -101,7 +99,6 @@ hist(dd$CqGAPDH)
 dd$RemainingLife <- round(as.numeric(dd$DeathDate-dd$CatchDate)/365,0)
 dd$SurvivedNext <- ifelse(dd$RemainingLife<1,0,1)
 
-dd$SurvivedNext <- ifelse(dd$RemainingLife>1,1,0)
 dd$Lifespan <- (dd$DeathYear-dd$LayYear)+1
 dd$Died <- ifelse(dd$DeathYear<2014,1,0)
 
@@ -110,7 +107,7 @@ dd$Died <- ifelse(dd$DeathYear<2014,1,0)
 # Exclude weird seasons and early years ----------------------------------
 
 dd <- subset(dd,LayYear>1997)
-dd <- subset(dd,Season == 'Major')
+#dd <- subset(dd,Season == 'Major')
 #dd <- subset(dd,FieldPeriodID != 27)
 
[…]

### 2016-09-05 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
More edits

@@
 #Run the model
-lm2 <- lmer(CqTelomere ~  LogAge* (Helper)+Tarsus + NonHelper + Sex + BodyMass  +yday*LogAge + (1|FieldPeriodID) + (1|BirdID) + (1|TerritoryID),
+lm2 <- lmer(CqTelomere ~  LogAge+Tarsus + NonHelper + Sex + BodyMass  +yday + (1|FieldPeriodID) + (1|BirdID) + (1|TerritoryID),
             data = temp,
@@
 
+temp <- subset(juv_r,Season == 'Major')
 Fig3C <- ggplot(temp,
@@
   geom_point()+
-   stat_smooth()+
+   stat_smooth(method = 'lm',formula = y~poly(x,2))+
   # scale_colour_manual(values = c('grey','darkorange','darkgreen'))+

### 2016-09-05 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
More

@@
 #Run the model
-lm2 <- lmer(CqTelomere ~  LogAge+Tarsus + NonHelper + Sex + BodyMass  +yday + (1|FieldPeriodID) + (1|BirdID) + (1|TerritoryID),
+lm2 <- lmer(RTL ~  LogAge+Tarsus + Helper + NonHelper + Sex + BodyMass  +yday + (1|FieldPeriodID) + (1|BirdID) + (1|TerritoryID),
             data = temp,

### 2016-09-20 Manuscript/SW_early_life_telomeres.Rmd
Home edits

@@
 We report model-averaged coefficients, confidence intervals and 'relative importance', which reflects the relative weights of each predictor variable across the top model set.
-For individuals with longitudinal data (*n* = `r nrow(Loss)`) we repeated the above analyses of telomere dynamics, replacing telomere length with telomere loss as the response variable, and excluding the plate ID random effect (as each measurement of telomere loss was based on two or more measurements, and so run on multiple plates).
+For individuals with longitudinal data (*n* = `r nrow(dd3)`) we repeated the above analyses of telomere dynamics, replacing telomere length with telomere loss as the response variable, and excluding the plate ID random effect (as each measurement of telomere loss was based on two or more measurements, and so run on multiple plates).
 For both sets of models, we first ran the model averaging procedure including all interaction terms, then re-ran the pipeline excluding non-significant interactions in order to obtain more accurate model averaged parameter estimates and confidence intervals.
@@
 
-Longitudinal data showed that early-life RTL was not related to adult RTL (R^2^ = `r round2(summary(lm(RTL1~RTL+Agemonths,data=Loss))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data=Loss))$coefficients[2,3])`; P = `r round2(summary(lm(RTL1~RTL+Agemonths,data = Loss))$coefficients[2,4])`; Fig.
+Longitudinal data showed that early-life RTL was not related to adult RTL (R^2^ = `r round2(summary(lm(RTL1~RTL+Agemonths,data = dd3))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data = dd3))$coefficients[2,3])`; P = `r round2(summary(lm(RTL1~RTL+Agemonths,data = dd3))$coefficients[2,4])`; Fig.
 4A).
-Running this analysis separately for juvenile age classes revealed that this relationship was not significant for nestlings (R^2^ = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(Loss,Agemonths == 1)))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(Loss,Agemonths == 1)))$coefficients[2,3])`; P = `r round2(summary(lm(RTL1~RTL+Agemonths,data = subset(Loss,Agemonths == 1)))$coefficients[2,4])`), fledglings (R^2^ = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(Loss,Agemonths == 6)))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(Loss,Agemonths == 6)))$coefficients[2,3])`; P = `r round2(summary(lm(RTL1~RTL+Agemonths,data = subset(Loss,Agemonths == 6)))$coefficients[2,4])`) or subadults (R^2^ = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(Loss,Agemonths == 10)))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(Loss,Agemonths == 10)))$coefficients[2,3])`; P = `r round2(summary(lm(RTL1~RTL+Agemonths,data = subset(Loss,Agemonths == 10)))$coefficients[2,4])`).
-$\Delta$RTL was significantly lower than zero in nestlings (t = `r round2(t.test(subset(Loss,Agemonths == 1)$Loss)$statistic)`; P `r round3(t.test(subset(Loss,Agemonths == 1)$Loss)$p.value)`), but not in fledglings (t = `r round2(t.test(subset(Loss,Agemonths == 6)$Loss)$statistic)`; P = `r round2(t.test(subset(Loss,Agemonths == 6)$Loss)$p.value)`) or subadults (t = `r round2(t.test(subset(Loss,Agemonths == 10)$Loss)$statistic)`; P = `r round2(t.test(subset(Loss,Agemonths == 10)$Loss)$p.value)`), suggesting that telomere shortening was highest shortly after the nestling phase, after which both telomere shortening and telomere lengthening were observed (Fig.
+Running this analysis separately for juvenile age classes revealed that this relationship was not significant for nestlings (R^2^ = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(dd3,Agemonths == 1)))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(dd3,Agemonths == 1)))$coefficients[2,3])`; P = `r round2(summary(lm(RTL1~RTL+Agemonths,data = subset(dd3,Agemonths == 1)))$coefficients[2,4])`), fledglings (R^2^ = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(dd3,Agemonths == 6)))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(dd3,Agemonths == 6)))$coefficients[2,3])`; P = `r round2(summary(lm(RTL1~RTL+Agemonths,data = subset(dd3,Agemonths == 6)))$coefficients[2,4])`) or subadults (R^2^ = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(dd3,Agemonths == 10)))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(dd3,Agemonths == 10)))$coefficients[2,3])`; P = `r round2(summary(lm(RTL1~RTL+Agemonths,data = subset(dd3,Agemonths == 10)))$coefficients[2,4])`).
+$\Delta$RTL was significantly lower than zero in nestlings (t = `r round2(t.test(subset(dd3,Agemonths == 1)$DeltaRTL)$statistic)`; P `r round3(t.test(subset(dd3,Agemonths == 1)$DeltaRTL)$p.value)`), but not in fledglings (t = `r round2(t.test(subset(dd3,Agemonths == 6)$DeltaRTL)$statistic)`; P = `r round2(t.test(subset(dd3,Agemonths == 6)$DeltaRTL)$p.value)`) or subadults (t = `r round2(t.test(subset(dd3,Agemonths == 10)$DeltaRTL)$statistic)`; P = `r round2(t.test(subset(dd3,Agemonths == 10)$DeltaRTL)$p.value)`), suggesting that telomere shortening was highest shortly after the nestling phase, after which both telomere shortening and telomere lengthening were observed (Fig.
 4B).
@@
 Early-life RTL varied significantly among cohorts (Likelihood ratio test: F = `r round2(anova(lme(RTL~factor(FieldPeriodID),data=juv,random=~1|BirdID))[2,3])`; P `r round3(anova(lme(RTL~Agemonths,data=juv,random=~1|BirdID))[2,4])`).
-Cohort-level variation in juvenile RTL was not related to adult population density (LMM controlling for age and cohort: t = `r round2(summary(lme(RTL~Agemonths+Density,data=juv,random=~1|FieldPeriodID))$tTable[3,4])`; P = `r round3(summary(lme(RTL~Agemonths+Density,data=juv,random=~1|FieldPeriodID))$tTable[3,5])`), or to temporal variation in  food availability (t = `r round2(summary(lme(RTL~Agemonths+Insect,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,4])`; P = `r round3(summary(lme(RTL~Agemonths+Insect,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,5])`).
+Cohort-level variation in juvenile RTL was not related to adult population density (LMM controlling for age and cohort: t = `r round2(summary(lme(RTL~Agemonths+Density,data=subset(juv,!is.na(Density)),random=~1|FieldPeriodID))$tTable[3,4])`; P = `r round3(summary(lme(RTL~Agemonths+Density,data=subset(juv,!is.na(Density)),random=~1|FieldPeriodID))$tTable[3,5])`), or to temporal variation in  food availability (t = `r round2(summary(lme(RTL~Agemonths+Insect,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,4])`; P = `r round3(summary(lme(RTL~Agemonths+Insect,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,5])`).
 
-$\Delta$RTL did not differ significantly among cohorts (F = `r round2(anova(lm(Loss~factor(FieldPeriodID),data = Loss))[1,4])`; P = `r round(anova(lm(Loss~factor(FieldPeriodID),data = Loss))[1,5])`), although power for this analysis was limited as longitudinal sample sizes within cohorts were low.
+$\Delta$RTL did not differ significantly among cohorts (F = `r round2(anova(lm(DeltaRTL~factor(FieldPeriodID),data = dd3))[1,4])`; P = `r round(anova(lm(DeltaRTL~factor(FieldPeriodID),data = dd3))[1,5])`), although power for this analysis was limited as longitudinal sample sizes within cohorts were low.
 Model selection revealed that the top model explaining $\Delta$RTL was the null model (Table S2).
@@
 Within cohorts, the top model explaining variation in early-life RTL contained age, territory quality, tarsus length, and the interaction between age and tarsus length (Table S1).
-The top model was much better supported than a null model ($\Delta$AICc = `r round2(AICc(lmer(RTL~1+(1|FieldPeriodID)+(1|TerritoryID),data=juv,REML = FALSE))-AICc(lmer(RTL~Agemonths+TQ+Tarsus+Agemonths*Tarsus+(1|FieldPeriodID)+(1|TerritoryID),data=juv,REML = FALSE)))`).
 Model-averaged confidence intervals of the age and tarsus x age effects did not overlap zero, but the tarsus and territory quality effects were not significant (Fig.
@@
 the median RTL for each cohort) and added this term to a GLMM, and found that cohort-level RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~cohortTL+RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv14,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~cohortTL+RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv14,family = 'binomial'))$coefficients[2,4])`).
-Using the longitudinal data, we found that $\Delta$RTL had no effect on survival to the year after the second sampling event (estimate = `r round2(summary(glm(SurvivedNext2~Loss+Agemonths,data = subset(Loss,LayYear<2013),family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glm(SurvivedNext2~Loss+Agemonths,data=subset(Loss,LayYear<2013),family = 'binomial'))$coefficients[2,4])`).
-
-summary(glmer(SurvivedNext~RTL+(1|BirdID),data=subset(dd,LayYear<2000), family = 'binomial'))
-
-Cox regression showed no effect of individual-level or cohort-level RTL on survival later in life (individual-level: estimate = `r round2(summary(coxph(with(juv12,Surv(Lifespan,Died)~RTL+Agemonths+frailty(FieldPeriodID))))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv12,Surv(Lifespan,Died)~RTL+Agemonths+frailty(FieldPeriodID))))$coefficients[1,6])`; cohort-level: estimate = `r round2(summary(coxph(with(juv12,Surv(Lifespan,Died)~cohortTL*Agemonths+RTL+frailty(FieldPeriodID))))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv12,Surv(Lifespan,Died)~cohortTL*Agemonths+RTL+frailty(FieldPeriodID))))$coefficients[1,6])`), and no interaction between RTL and juvenile age on survival (individual-level: estimate = `r round2(summary(coxph(with(juv12,Surv(Lifespan,Died)~RTL*Agemonths+frailty(FieldPeriodID))))$coefficients[4,1])`, P = `r round2(summary(coxph(with(juv12,Surv(Lifespan,Died)~RTL*Agemonths+frailty(FieldPeriodID))))$coefficients[4,6])`; cohort-level: estimate = `r round2(summary(coxph(with(juv12,Surv(Lifespan,Died)~cohortTL*Agemonths+frailty(FieldPeriodID))))$coefficients[4,1])`, P = `r round2(summary(coxph(with(juv12,Surv(Lifespan,Died)~cohortTL*Agemonths+frailty(FieldPeriodID))))$coefficients[4,6])`).
-$\Delta$RTL was not related to longer-term survival (estimate `r round3(summary(coxph(with(subset(Loss,LayYear<2008),Surv(RemainingLife2+1,Died)~Loss+Agemonths)))$coefficients[1,1])`, P = `r round2(summary(coxph(with(subset(Loss,LayYear<2008),Surv(RemainingLife2+1,Died)~Loss)))$coefficients[1,5])`).
 

### 2016-09-20 Scripts/SW juvenile telomeres IPC clean.R

@@ -345,6 +345,6 @@ for(i in 1:nrow(addata))
 juv_r <- rbind(juvdata,addata)
 juv <- juvdata
 
-juv14 <- subset(juv,LayYear<2014)
-juv12 <- subset(juv,LayYear<2012)
+juv14 <- subset(juv_r,LayYear<2014)
+juv12 <- subset(juv_r,LayYear<2012)
 

### 2016-09-21 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Work edits

--- 2016-09-05 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
@@
 #Run the model
-lm2 <- lmer(CqTelomere ~  LogAge+Tarsus + NonHelper + Sex + BodyMass  +yday + (1|FieldPeriodID) + (1|BirdID) + (1|TerritoryID),
+lm2 <- lmer(RTL ~  LogAge+Tarsus + Helper + NonHelper + Sex + BodyMass  +yday + (1|FieldPeriodID) + (1|BirdID) + (1|TerritoryID),
+
             data = temp,

### 2016-09-29 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Edits

--- 2016-09-05 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
@@
         legend.position = 'none')+
-  geom_point()+
-    stat_smooth(col = 'darkgrey',lty = 3,se = F)+
-  stat_smooth(method = 'lm',col='darkgrey',formula = y~log(x),se = F)
-  annotate('text',x = 1,y = 2.8,label = 'A')
+  geom_point(col = 'darkgrey')+
+  stat_smooth(method = 'lm',col='darkblue',fill = 'darkblue',formula = y~log(x),se = T)+
+  annotate('text',x = 1,y = 2.9,label = 'A')
 
@@
   geom_point(col='grey')+
-  stat_smooth(method = 'lm',col = 'darkgrey')+
-  annotate('text',x = -100,y = 2.1,label = 'B')
+  stat_smooth(method = 'lm',col = 'darkblue',fill = 'darkblue')+
+  annotate('text',x = -100,y = 2.9,label = 'B')
 
@@
 
-temp <- subset(juv_r,Season == 'Major')
-temp$LogAge <- log(temp$Agemonths)
+temp <- dd
+temp$LogAge <- log(temp$Agemonths+1)
+
+temp2 <- subset(temp,CatchYear<2014)
+
+temp2$cenTL <- (temp2$RTL-tapply(temp2$RTL,temp2$Agemonths,mean)[as.character(temp2$Agemonths)])/tapply(temp2$RTL,temp2$Agemonths,sd)[as.character(temp2$Agemonths)]
+
+#temp2 <- subset(temp2,Whodunnit != 'EAF')
+temp2 <- subset(temp2,!is.na(Agemonths))
+temp2 <- subset(temp2,Agemonths>10)
+
+temp2$obs <- c(1:nrow(temp2))
+
+summary(glmer(SurvivedNext~RTL+Agemonths+(1|BirdID)+(1|obs),data=temp2,family = 'binomial',control = glmerControl(optimizer = "bobyqa")))
+summary(glm(SurvivedNext~RTL+Agemonths,data=temp2,family = 'binomial'))
+
+ggplot(temp2,aes(y = RTL,x = Agemonths,col = factor(SurvivedNext)))+
+  geom_point()+
+  stat_smooth()
+x <- tapply(temp2$RTL,list(temp2$SurvivedNext,factor(temp2$Agemonths)),mean)[,1:10]
+plot(x[1,],type = 'l')
+points(x[2,],type = 'l',lty = 2)
 
@@
 #Model averaging
+
+summary(lm2)
+confint.merMod(lm2)
 
@@
 ```
+
+```{r Figure X}
+
+temp <- dd
+#temp$SurvivedNext <- factor(temp$SurvivedNext)
+temp$SN2 <- with(temp,paste(Sex,SurvivedNext))
+
+FigX <- ggplot(temp,aes(x = LayYear,y = RTL,col = SN2))+
+  ylab('RTL') + 
+  xlab('Cohort') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12,angle = 90), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+                axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'),
+        legend.position = 'top')+
+  geom_point()+
+  stat_smooth()
+
+FigX
+
+FigX <- ggplot(subset(temp,Agemonths),aes(x = SN2,y = RTL))+
+  ylab('RTL') + 
+  xlab('Cohort') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12,angle = 90), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+                axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'),
+        legend.position = 'top')+
+  geom_boxplot()
+
+FigX
+
+```

### 2016-09-29 Scripts/SW juvenile telomeres IPC clean.R

@@ -4,7 +4,7 @@
 #################################################################################
 
 #Only use Ellie's samples
-dd0 <- subset(dd0,Whodunnit == 'EAF')
+#dd0 <- subset(dd0,Whodunnit == 'EAF')
 
 #Average repeats of blood samples
 dd0$RTL2 <- ave(dd0$RTL,c(dd0$BloodID,dd0$Status,dd0$PlateID))
@@ -54,7 +54,7 @@ dd$Agemonths <- ifelse(dd$Ageclass == 'CH',1,
                               ifelse(dd$Ageclass == 'SA',10,dd$Age*12+6)))
 
 
-  dd$Age[dd$Fledged != 'Adults'] <- 1
+  #dd$Age[dd$Fledged != 'Adults'] <- 1
 
 
 
@@ -102,13 +102,6 @@ dd$SurvivedNext <- ifelse(dd$RemainingLife<1,0,1)
 dd$Lifespan <- (dd$DeathYear-dd$LayYear)+1
 dd$Died <- ifelse(dd$DeathYear<2014,1,0)
 
-
-
-# Exclude weird seasons and early years ----------------------------------
-
-dd <- subset(dd,LayYear>1997)
-#dd <- subset(dd,Season == 'Major')
-#dd <- subset(dd,FieldPeriodID != 27)
 
 # Sex ---------------------------------------------------------------------
 

### 2016-10-03 Manuscript/SW_early_life_telomeres_SOM.Rmd
Edits

@@
 temp <- dd3
+temp <- subset(temp,!is.na(TQ))
 
 #Run the model
-lm1 <- lmer(DeltaRTL ~ Agemonths*(Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ) + (1|BirdID) + (1|FieldPeriodID) + (1|TerritoryID),
+lm1 <- lmer(DeltaRTL ~ Agemonths+(DeltaCondition + Sex + TQ) + (1|BirdID) + (1|FieldPeriodID) + (1|TerritoryID),
             data = temp,
             REML = FALSE)
+
+lm1 <- glmer(DeltaRTLF ~ Agemonths+(Tarsus + Helper + Sex + BodyMass +TQ) + (1|TerritoryID) + (1|BirdID),
+            data = temp,
+            family = 'binomial',
+            control = glmerControl(optimizer = 'bobyqa'))
+
 lm1 <- standardize(lm1)
+summary(lm1)
+confint.merMod(lm1)
 
@@
 
+ggplot(dd3,aes(x = DeltaMass,y = (DeltaRTL)))+
+geom_point()+
+stat_smooth(method = 'lm')
+
 **Figure S1** $\Delta$RTL in relation to the time between sampling events.

### 2016-10-03 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Edits

@@
 
-temp <- dd3
+temp <- subset(dd3,Agemonths<12)
 temp[,'Age (months)'] <- factor(temp$Agemonths)
@@
        aes(x = RTL,
-           y = RTL1,
-           col = factor(DeltaRTLF))) +
+           y = RTL1)) +
   ylab('Adult RTL') + 
@@
 
-temp <- subset(dd,DeltaAge !=0)
+temp <- subset(dd,Age<15)
 reps <- tapply(temp$RTL,temp$BirdID,length)
-keep <- names(reps[reps>3])
+keep <- names(reps[reps>0])
 temp <- subset(temp,BirdID %in% keep)
@@
  mins <- tapply(temp$Agemonths,temp$BirdID,min)
- keep2 <- names(mins[mins <5])
+ keep2 <- names(mins[mins <400])
  temp <- subset(temp,BirdID %in% keep2)
@@
 Fig4B <- ggplot(temp,aes(x = Agemonths,y = RTL,group = BirdID,col=factor(BirdID)))+
-geom_line(col = 'darkgrey')+
+  geom_point(alpha = 1,col='grey')+
+geom_line(col = 'darkgrey',alpha = 0.2)+
   xlab('Age (years)') +
@@
   #scale_colour_brewer(palette = 'Set1')+
-  annotate('text',x = 0,y = 2,label = 'B')
+  annotate('text',x = 0,y = 3,label = 'B')
 
@@
 
+temp <- subset(dd,Age<15)
+reps <- tapply(temp$RTL,temp$BirdID,length)
+keep <- names(reps[reps>2])
+temp <- subset(temp,BirdID %in% keep)
+ temp <- subset(temp,!duplicated(paste0(BirdID,Agemonths)))
+
+ggplot(temp,aes(x = CatchYear,y = RTL,group = BirdID,col=factor(BirdID)))+
+  geom_point(alpha = 1,col='grey')+
+geom_line(col = 'darkgrey',alpha = 0.2)+
+  xlab('Age (years)') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'),
+        legend.position = 'none')+
+  #scale_colour_brewer(palette = 'Set1')+
+  annotate('text',x = 1997,y = 2,label = 'B')
+
+ggplot(ddL,aes(x = DeltaRTL,y = RTL,col = Group))+
+geom_point()+
+stat_smooth(method = 'lm')
+
 ```{r Figure X}

### 2016-10-03 Scripts/SW juvenile telomeres IPC clean.R

@@ -4,7 +4,7 @@
 #################################################################################
 
 #Only use Ellie's samples
-#dd0 <- subset(dd0,Whodunnit == 'EAF')
+dd0 <- subset(dd0,Whodunnit == 'EAF')
 
 #Average repeats of blood samples
 dd0$RTL2 <- ave(dd0$RTL,c(dd0$BloodID,dd0$Status,dd0$PlateID))
@@ -211,6 +211,7 @@ for(i in 1:nrow(dd))
 dd <- subset(dd,!is.na(Tarsus))
 dd <- subset(dd,!is.na(BodyMass))
 dd <- subset(dd,Status!='XF')
+dd$Condition <- lm(BodyMass~Tarsus,data = dd)$residuals 
 
 
 # Get rid of stuff not to be used -----------------------------------------
@@ -231,6 +232,7 @@ dd3$DeltaRTL <- NA
 dd3$DeltaTL <- NA
 dd3$DeltaGAP <- NA
 dd3$BloodID1 <- NA
+dd3$SurvivedNext2 <- NA
 
 for(i in 1:nrow(dd3))
 {
@@ -251,6 +253,9 @@ for(i in 1:nrow(dd3))
     dd3$DeltaTL[i] <- nextbird$CqTelomere - dd3$CqTelomere[i]
     dd3$TimeDiff[i] <- cd[birdpos+1,'Agemonths'] - dd3$Agemonths[i]
     dd3$BloodID1[i] <- nextbird$BloodID
+    dd3$SurvivedNext2[i] <- nextbird$SurvivedNext
+    dd3$DeltaCondition[i] <- nextbird$Condition - dd3$Condition[i]
+    dd3$DeltaAge[i] <- nextbird$Agemonths - dd3$Agemonths
   }
   
   
@@ -290,6 +295,7 @@ for(i in 1:nrow(dd3_2))
   {
     nextbird <- subset(cd,birdpos == dd3_2$birdpos[i]+1)
     dd3_2$DeltaRTL[i] <- nextbird$RTL - dd3_2$RTL[i]
+    dd3_2$RTL1[i] <- nextbird$RTL
     dd3_2$BloodID1[i] <- nextbird$BloodID
   }
   
@@ -306,6 +312,8 @@ dd3$DeltaRTLF <- ifelse(dd3$DeltaRTL>0,1,0)
 ddL <- data.frame(BirdID = c(dd3$BirdID,dd3_2$BirdID),
[…]

### 2016-10-05 Manuscript/SW_early_life_telomeres.Rmd
Edits

@@
 Early-life RTL varied significantly among cohorts (Likelihood ratio test: F = `r round2(anova(lme(RTL~factor(FieldPeriodID),data=juv,random=~1|BirdID))[2,3])`; P `r round3(anova(lme(RTL~Agemonths,data=juv,random=~1|BirdID))[2,4])`).
-Cohort-level variation in juvenile RTL was not related to adult population density (LMM controlling for age and cohort: t = `r round2(summary(lme(RTL~Agemonths+Density,data=subset(juv,!is.na(Density)),random=~1|FieldPeriodID))$tTable[3,4])`; P = `r round3(summary(lme(RTL~Agemonths+Density,data=subset(juv,!is.na(Density)),random=~1|FieldPeriodID))$tTable[3,5])`), or to temporal variation in  food availability (t = `r round2(summary(lme(RTL~Agemonths+Insect,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,4])`; P = `r round3(summary(lme(RTL~Agemonths+Insect,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,5])`).
+Cohort-level variation in juvenile RTL was not related to adult population density (LMM controlling for age and cohort: t = `r round2(summary(lme(RTL~Agemonths+Density,data=subset(juv,!is.na(Density)),random=~1|FieldPeriodID))$tTable[3,4])`; P = `r round3(summary(lme(RTL~Agemonths+Density,data=subset(dd,!is.na(Density)),random=~1|BirdID))$tTable[3,5])`), or to temporal variation in  food availability (t = `r round2(summary(lme(RTL~Agemonths+Insect,data=dd,random=~1|BirdID/FieldPeriodID,na.action=na.exclude))$tTable[3,4])`; P = `r round3(summary(lme(RTL~Agemonths+Insect,data=juv,random=~1|BirdID,na.action=na.exclude))$tTable[3,5])`).
 
@@
 
-temp <- dd
-temp$y <- ifelse(temp$CqTelomere<15,0,1)
+temp <- subset(dd,Ageclass !='CH')
 
-summary(glm(y~Ageclass+Sex+Helper,data=temp,family = 'binomial'))
-
-ggplot(temp,aes(y = RTL,x = Ageclass))+
-geom_boxplot()
+ggplot(temp,aes(x = Insect,y = RTL))+
+geom_boxplot(aes(group = Insect))+
+geom_smooth()+
+theme_bw()

### 2016-10-06 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Eds

@@
 
-temp <- dd
+temp <- subset(dd,LayYear %in% c(2000:2008))
 
-Fig1A <- ggplot(temp,aes(x = Agemonths,y = RTL))+
+Fig1A <- ggplot(temp,aes(x = Agemonths,y = RTL,col=factor(LayYear)))+
   ylab('RTL') + 
@@
         legend.position = 'none')+
-  geom_point(col = 'darkgrey')+
-  stat_smooth(method = 'lm',col='darkblue',fill = 'darkblue',formula = y~log(x),se = T)+
+  stat_smooth(method = 'lm',formula = y~log(x),se=F)
   annotate('text',x = 1,y = 2.9,label = 'A')
@@
 
-temp <- subset(dd,Agemonths<12)
-temp$LayYear <- factor(temp$LayYear)
+temp <- subset(dd,LayYear>1995)
 
@@
         legend.position = 'top')+
-  geom_jitter(col = 'grey')+
-  geom_boxplot(outlier.colour = NA,notch=F,alpha=0)
+  geom_boxplot(aes(group = LayYear),outlier.colour = 'grey',notch=F,alpha=0)+
+  geom_smooth()
+
+Fig2x <- ggplot(temp,aes(x = CatchYear,y = RTL))+
+  ylab('RTL') + 
+  xlab('Cohort') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12,angle = 90), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+                axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'),
+        legend.position = 'top')+
+  geom_boxplot(aes(group = CatchYear),outlier.colour = 'grey',notch=F,alpha=0)+
+  geom_smooth()
 
@@
 ```
+
+temp <- dd
+temp$LogAge <- log(temp$Agemonths)
+
+m1 <- lm(RTL~LogAge,data=temp)
+m2 <- lmer(RTL~LogAge+(1|CatchYear),data=temp)
+m3 <- lmer(RTL~LogAge+(LogAge|CatchYear),data=temp)
+m4 <- lmer(RTL~LogAge+(1|LayYear),data=temp)
+m5 <- lmer(RTL~LogAge+(LogAge|LayYear),data=temp)
+
+anova(m2,m3,m4,m5)
+summary(m2)
+summary(m3)

### 2016-10-06 Manuscript/SW_early_life_telomeres.Rmd
Home edits

@@
 
-Longitudinal data showed that early-life RTL was not related to adult RTL (R^2^ = `r round2(summary(lm(RTL1~RTL+Agemonths,data = dd3))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data = dd3))$coefficients[2,3])`; P = `r round2(summary(lm(RTL1~RTL+Agemonths,data = dd3))$coefficients[2,4])`; Fig.
+Longitudinal data showed that early-life RTL was not related to adult RTL (R^2^ = `r round2(summary(lm(RTL1~RTL+TimeDiff,data = dd3))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data = dd3))$coefficients[2,3])`; P = `r round2(summary(lm(RTL1~RTL+Agemonths,data = dd3))$coefficients[2,4])`; Fig.
 4A).
-Running this analysis separately for juvenile age classes revealed that this relationship was not significant for nestlings (R^2^ = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(dd3,Agemonths == 1)))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(dd3,Agemonths == 1)))$coefficients[2,3])`; P = `r round2(summary(lm(RTL1~RTL+Agemonths,data = subset(dd3,Agemonths == 1)))$coefficients[2,4])`), fledglings (R^2^ = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(dd3,Agemonths == 6)))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(dd3,Agemonths == 6)))$coefficients[2,3])`; P = `r round2(summary(lm(RTL1~RTL+Agemonths,data = subset(dd3,Agemonths == 6)))$coefficients[2,4])`) or subadults (R^2^ = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(dd3,Agemonths == 10)))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data=subset(dd3,Agemonths == 10)))$coefficients[2,3])`; P = `r round2(summary(lm(RTL1~RTL+Agemonths,data = subset(dd3,Agemonths == 10)))$coefficients[2,4])`).
+Running this analysis separately for juvenile age classes revealed that this relationship was not significant for nestlings (R^2^ = `r round2(summary(lm(RTL1~RTL,data=subset(dd3,Agemonths == 1)))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL,data=subset(dd3,Agemonths == 1)))$coefficients[2,3])`; P = `r round2(summary(lm(RTL1~RTL,data = subset(dd3,Agemonths == 1)))$coefficients[2,4])`), fledglings (R^2^ = `r round2(summary(lm(RTL1~RTL,data=subset(dd3,Agemonths == 6)))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL,data=subset(dd3,Agemonths == 6)))$coefficients[2,3])`; P = `r round2(summary(lm(RTL1~RTL,data = subset(dd3,Agemonths == 6)))$coefficients[2,4])`) or subadults (R^2^ = `r round2(summary(lm(RTL1~RTL,data=subset(dd3,Agemonths == 10)))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL,data=subset(dd3,Agemonths == 10)))$coefficients[2,3])`; P = `r round2(summary(lm(RTL1~RTL,data = subset(dd3,Agemonths == 10)))$coefficients[2,4])`).
 $\Delta$RTL was significantly lower than zero in nestlings (t = `r round2(t.test(subset(dd3,Agemonths == 1)$DeltaRTL)$statistic)`; P `r round3(t.test(subset(dd3,Agemonths == 1)$DeltaRTL)$p.value)`), but not in fledglings (t = `r round2(t.test(subset(dd3,Agemonths == 6)$DeltaRTL)$statistic)`; P = `r round2(t.test(subset(dd3,Agemonths == 6)$DeltaRTL)$p.value)`) or subadults (t = `r round2(t.test(subset(dd3,Agemonths == 10)$DeltaRTL)$statistic)`; P = `r round2(t.test(subset(dd3,Agemonths == 10)$DeltaRTL)$p.value)`), suggesting that telomere shortening was highest shortly after the nestling phase, after which both telomere shortening and telomere lengthening were observed (Fig.

### 2016-10-06 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Home edits

@@
 temp <- subset(dd,LayYear>1995)
+temp$JA <- ifelse(temp$Ageclass == 'A','Adults','Juveniles')
 
@@
 
-Fig2x <- ggplot(temp,aes(x = CatchYear,y = RTL))+
-  ylab('RTL') + 
-  xlab('Cohort') +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12,angle = 90), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-                axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black'),
-        legend.position = 'top')+
-  geom_boxplot(aes(group = CatchYear),outlier.colour = 'grey',notch=F,alpha=0)+
-  geom_smooth()
-
 pdf('Figures/Figure 2.pdf',
@@
 
-temp <- dd
+temp <- subset(dd,!is.na(Insect))
 temp$LogAge <- log(temp$Agemonths+1)
 
-temp2 <- subset(temp,CatchYear<2014)
-
-temp2$cenTL <- (temp2$RTL-tapply(temp2$RTL,temp2$Agemonths,mean)[as.character(temp2$Agemonths)])/tapply(temp2$RTL,temp2$Agemonths,sd)[as.character(temp2$Agemonths)]
-
-#temp2 <- subset(temp2,Whodunnit != 'EAF')
-temp2 <- subset(temp2,!is.na(Agemonths))
-temp2 <- subset(temp2,Agemonths>10)
-
-temp2$obs <- c(1:nrow(temp2))
-
-summary(glmer(SurvivedNext~RTL+Agemonths+(1|BirdID)+(1|obs),data=temp2,family = 'binomial',control = glmerControl(optimizer = "bobyqa")))
-summary(glm(SurvivedNext~RTL+Agemonths,data=temp2,family = 'binomial'))
-
-ggplot(temp2,aes(y = RTL,x = Agemonths,col = factor(SurvivedNext)))+
-  geom_point()+
-  stat_smooth()
-x <- tapply(temp2$RTL,list(temp2$SurvivedNext,factor(temp2$Agemonths)),mean)[,1:10]
-plot(x[1,],type = 'l')
-points(x[2,],type = 'l',lty = 2)
-
 #Run the model
-lm2 <- lmer(RTL ~  LogAge+Tarsus + Helper + NonHelper + Sex + BodyMass  +yday + (1|FieldPeriodID) + (1|BirdID) + (1|TerritoryID),
+lm2 <- lmer(RTL ~  LogAge+Tarsus + Helper + NonHelper + Sex + BodyMass  +yday + Insect + Density + (1|LayYear) + (1|BirdID) + (1|TerritoryID),
             data = temp,
@@
 lm2 <- standardize(lm2)
+
 #Model averaging
-
-summary(lm2)
-confint.merMod(lm2)
-
 lm3 <- dredge(lm2)
@@
 
-temp <- subset(dd3,Agemonths<12)
+temp <- dd3
 temp[,'Age (months)'] <- factor(temp$Agemonths)

### 2016-10-07 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Edits

@@
 
-temp <- subset(dd,LayYear %in% c(2000:2008))
+temp <- subset(dd,Age<15)
+reps <- tapply(temp$RTL,temp$BirdID,length)
+keep <- names(reps[reps>0])
+temp <- subset(temp,BirdID %in% keep)
+temp <- subset(temp,!duplicated(paste0(BirdID,Agemonths)))
 
-Fig1A <- ggplot(temp,aes(x = Agemonths,y = RTL,col=factor(LayYear)))+
-  ylab('RTL') + 
-  xlab('Age (months)') +
+ mins <- tapply(temp$Agemonths,temp$BirdID,min)
+ keep2 <- names(mins[mins <400])
+ temp <- subset(temp,BirdID %in% keep2)
+
+temp$Agemonths <- temp$Agemonths/12
+
+Fig1A <- ggplot(temp,aes(x = Agemonths,y = RTL,group = BirdID,col=factor(BirdID)))+
+  geom_point(alpha = 1,col='grey')+
+geom_line(col = 'darkgrey',alpha = 0.2)+
+  xlab('Age (years)') +
   theme_classic() +
@@
         legend.position = 'none')+
-  stat_smooth(method = 'lm',formula = y~log(x),se=F)
-  annotate('text',x = 1,y = 2.9,label = 'A')
+  geom_smooth(col = 'black',fill = 'black',lwd = 0.5,aes(group = NULL),method = 'lm',formula = y~log(x))+
+  annotate('text',x = 0,y = 3,label = 'A')
 
-temp <- subset(dd,DeltaAge !=0)
+temp <- dd3
 
-Fig1B <- ggplot(temp,aes(x = DeltaAge,y = RTL))+
-  ylab('RTL') + 
-  xlab(expression(Delta*' age (months)')) +
+Fig1B <- ggplot(temp,
+       aes(x = RTL,
+           y = RTL1)) +
+  ylab('Adult RTL') + 
+  xlab('juvenile RTL') +
+    xlim(c(0,3))+
+  ylim(c(0,3))+
   theme_classic() +
@@
         axis.title.x = element_text(size = 14),
+        legend.position = c('none'),
+        legend.title = element_text(size = 12),
         axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black'),
-        legend.position = 'none')+
-  geom_point(col='grey')+
-  stat_smooth(method = 'lm',col = 'darkblue',fill = 'darkblue')+
-  annotate('text',x = -100,y = 2.9,label = 'B')
+        axis.line.y = element_line(colour = 'black')) +
+    geom_point(col = 'grey')+
+  geom_abline(slope = 1,intercept = 0,lty = 2)+
+  annotate('text',x = 0,y = 3,label = 'B')
+
+Fig1C <- ggplot(ddL,aes(col=Group,fill = Group,y = ..scaled..,x = DeltaRTL))+
+    geom_density(alpha = 0.2)+
+  theme_classic()+
+  ylab ('Scaled density')+
+  xlab(expression(Delta*'RTL'))+
+    theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        legend.position = 'none',
+        legend.title = element_blank(),
+        axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black')) +
+  scale_colour_manual(values = c('darkgrey','gold'))+
+  scale_fill_manual(values = c('darkgrey','gold'))+
+  geom_vline(xintercept = 0,lty = 2)+
+  annotate('text',x = -2,y = 1,label = 'C')
+
+temp <- dd3
+
+Fig1D <- ggplot(temp,aes(y = DeltaRTL,x = TimeDiff))+
+    geom_point(col = 'grey')+
+  theme_classic()+ 
+  xlab ('Follow up time (years)')+
+  ylab(expression(Delta*'RTL'))+
+  ylim(c(-2,2))+
+    theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        legend.position = c('none'),
+        legend.title = element_blank(),
+        axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black')) +
+    geom_abline(slope = 0,intercept = 0,lty = 2)+
+  annotate('text',x = 0,y = 2,label = 'D')
 
 pdf('Figures/Figure 1.pdf',
-    width = 4,
-    height = 8,
+    width = 8,
+    height = 7,
     paper = 'a4')
-multiplot(Fig1A,Fig1B,layout=matrix(c(1,2),nrow = 2,byrow=T))
+multiplot(Fig1A,Fig1B,Fig1C,Fig1D,layout=matrix(c(1,2,3,4),nrow=2,byrow=T))
 dev.off()
@@
 #Run the model
-lm2 <- lmer(RTL ~  LogAge+Tarsus + Helper + NonHelper + Sex + BodyMass  +yday + Insect + Density + (1|LayYear) + (1|BirdID) + (1|TerritoryID),
+lm2 <- lmer(RTL ~  LogAge+Tarsus + Helper + NonHelper + Sex + BodyMass  +yday + Insect + Density + (1|LayYear) + (1|BirdID),
             data = temp,
@@
 
-```{r Figure 4}
-
-temp <- dd3
-temp[,'Age (months)'] <- factor(temp$Agemonths)
-
-Fig4A <- ggplot(temp,
-       aes(x = RTL,
-           y = RTL1)) +
-  ylab('Adult RTL') + 
-  xlab('juvenile RTL') +
-    xlim(c(0,3))+
-  ylim(c(0,3))+
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        legend.position = c('none'),
-        legend.title = element_text(size = 12),
-        axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black')) +
-  geom_abline(slope = 1,intercept = 0,lty = 2)+
-  geom_point()+
-  scale_colour_manual(values = c('navy','red'))+
-  annotate('text',x = 0,y = 3,label = 'A')
-
-temp <- subset(dd,Age<15)
-reps <- tapply(temp$RTL,temp$BirdID,length)
-keep <- names(reps[reps>0])
-temp <- subset(temp,BirdID %in% keep)
- temp <- subset(temp,!duplicated(paste0(BirdID,Agemonths)))
-
- mins <- tapply(temp$Agemonths,temp$BirdID,min)
- keep2 <- names(mins[mins <400])
- temp <- subset(temp,BirdID %in% keep2)
-
-temp$Agemonths <- temp$Agemonths/12
-
-mycols <- topo.colors(length(keep2))
-
-Fig4B <- ggplot(temp,aes(x = Agemonths,y = RTL,group = BirdID,col=factor(BirdID)))+
-  geom_point(alpha = 1,col='grey')+
-geom_line(col = 'darkgrey',alpha = 0.2)+
-  xlab('Age (years)') +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black'),
-        legend.position = 'none')+
-  #scale_colour_brewer(palette = 'Set1')+
-  annotate('text',x = 0,y = 3,label = 'B')
-
-Fig4C <- ggplot(ddL,aes(col=Group,fill = Group,y = ..scaled..,x = DeltaRTL))+
-    geom_density(alpha = 0.2)+
-  theme_classic()+
-  ylab ('Scaled density')+
-  xlab(expression(Delta*'RTL'))+
-    theme(axis.text.x = element_text(size = 12), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        legend.position = 'none',
-        legend.title = element_blank(),
-        axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black')) +
-  scale_colour_manual(values = c('darkgrey','gold'))+
-  scale_fill_manual(values = c('darkgrey','gold'))+
-  geom_vline(xintercept = 0,lty = 2)+
-  annotate('text',x = -2,y = 1,label = 'C')
-
-temp <- dd3
-
-Fig4D <- ggplot(temp,aes(col=factor(DeltaRTLF),y = DeltaRTL,x = TimeDiff))+
-    geom_point()+
-  theme_classic()+ 
-  xlab ('Follow up time (years)')+
-  ylab(expression(Delta*'RTL'))+
-  ylim(c(-2,2))+
-    theme(axis.text.x = element_text(size = 12), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        legend.position = c('none'),
-        legend.title = element_blank(),
-        axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black')) +
-  scale_colour_manual(values = c('navy','red'))+
-    geom_abline(slope = 0,intercept = 0,lty = 2)+
-  annotate('text',x = 0,y = 2,label = 'D')
-
-pdf('Figures/Figure 4.pdf',
-    width = 8,
-    height = 7,
-    paper = 'a4')
-multiplot(Fig4A,Fig4B,Fig4C,Fig4D,layout=matrix(c(1,2,3,4),nrow=2,byrow=T))
-dev.off()
-
-```
-
 temp <- subset(dd,Age<15)
@@
 
-anova(m2,m3,m4,m5)
+m1 <- lmer(RTL~(Agemonths)+(1|BirdID)+(1|LayYear),data=dd)
+m2 <- lmer(RTL~Ageclass+(1|BirdID)+(1|LayYear),data=dd)
+m3 <- lmer(RTL~LogAge+(1|BirdID)+(1|LayYear),data=dd)
+
+anova(m1,m2,m3)
 summary(m2)
 summary(m3)
+
+dd3$LogAge <- log(dd3$Agemonths)
+m1 <- lmer(RTL1~RTL+(1|LayYear),data=dd3)
+summary(m1)

### 2016-10-10 Manuscript/SW_early_life_telomeres.Rmd
Edits
Home edits

@@
 1.
-Understanding the costs of individual early-life experiences is fundamental to understanding life-history evolution and population dynamics.
+Understanding the costs individuals and populations face throughout their lifetimes is fundamental to understanding life-history evolution and population dynamics.
 Telomeres, the protective caps at the ends of chromosomes, shorten in response to oxidative stress, and telomere shortening is correlated with reduced survival.
-Thus, telomere dynamics may help us quantify individual variation in early-life costs, and enhance our understanding of how poor conditions in early life are related to later-life survival.
+Thus, telomere dynamics may help us quantify individual variation in costs, and enhance our understanding individual variation in natural populations.
 
 2.
-We tested how telomere dynamics are related to spatio-temporal variation in early-life conditions and later-life survival in the Seychelles warbler (*Acrocephalus sechellensis*), using a large dataset of `r nrow(juv)` juveniles from multiple cohorts spanning `r max(juv$LayYear)-min(juv$LayYear)` years.
+We tested how spatiotemporal in longitudinal telomere dynamics are related to ecology and survival in the Seychelles warbler (*Acrocephalus sechellensis*), using a large dataset of `r nrow(dd)` samples from multiple cohorts spanning `r max(juv$LayYear)-min(juv$LayYear)` years.
 
@@
 4.
-Juvenile telomere length varied markedly among cohorts, but we found no evidence that this variation was related to temporal variation in the environment.
-At the individual found weak evidence that territory quality was negatively related to telomere length, but the majority of variation in juvenile telomere length and dynamics was unexplained.
-We found no relationship between early-life telomere dynamics and survival to adulthood or later in life.
+Temporal length and rates of shortening varied markedly both within and among cohorts.
+Aside from age-related measurements, the only significant ecological predictor of telomer length  was temporal variation in food availability.
+We also found that cohort and age effects led to spurious associations between telomere dynamics and survival.
 
 5.
-Our comprehensive, long-term data suggest that although telomeres shorten with age, in natural populations telomere dynamics vary enormously over both space and time.
-Furthermore, Ascertaining what explains this variation will be a difficult but worthwhile task, which will require combining long-term ecological and quantitative genetic data.
+Our comprehensive, long-term data suggest that although telomeres shorten with age, in natural populations telomere dynamics vary enormously over both space and time, and that this variation cannot be attributed entirely to measurment error.
+Ascertaining what explains this variation will be a difficult but worthwhile task, and will require combining long-term ecological and quantitative genetic data.
 
@@
 theme_bw()
+
+dd$LogAge <- log(dd$Agemonths)
+m1 <- lmer(RTL~LogAge+(1|BirdID),data=dd)
+m2 <- lmer(RTL~LogAge+(1|BirdID)+(1|TerritoryID),data=dd)
+m3 <- lmer(RTL~LogAge+(1|BirdID)+(1|LayYear),data=dd)
+m4 <- lmer(RTL~LogAge+(1|BirdID)+(1|CatchYear)+(1|LayYear),data=dd)
+
+anova(m1,m2,m3,m4)

### 2016-10-11 Manuscript/SW_early_life_telomeres.Rmd
Edits

@@
 2.
-We tested how spatiotemporal in longitudinal telomere dynamics are related to ecology and survival in the Seychelles warbler (*Acrocephalus sechellensis*), using a large dataset of `r nrow(dd)` samples from multiple cohorts spanning `r max(juv$LayYear)-min(juv$LayYear)` years.
+We provide a detailed analysis of how telomere dynamics vary over individual lifespans, and with spatiotemporal variation in the social and ecological environment, in the Seychelles warbler (*Acrocephalus sechellensis*).
+We use one of the largest longitudinal datasets to date, consisting of `r nrow(dd)` samples from multiple cohorts spanning `r max(juv$LayYear)-min(juv$LayYear)` years.
 
@@
 anova(m1,m2,m3,m4)
+
+head(dd3)
+fit <- coxph(Surv(RemainingLife, RemainingLife1, Died1) ~ Agemonths1 + RTL1,
+              data=dd3)
+
+ dd3 <- subset(dd3,RemainingLife-RemainingLife1 >0)
+
+which(is.na(dd3$RemainingLife1))
+which(is.na(dd3$RTL1))

### 2016-10-11 Scripts/SW juvenile telomeres IPC clean.R

@@ -253,9 +253,12 @@ for(i in 1:nrow(dd3))
     dd3$DeltaTL[i] <- nextbird$CqTelomere - dd3$CqTelomere[i]
     dd3$TimeDiff[i] <- cd[birdpos+1,'Agemonths'] - dd3$Agemonths[i]
     dd3$BloodID1[i] <- nextbird$BloodID
-    dd3$SurvivedNext2[i] <- nextbird$SurvivedNext
+    dd3$SurvivedNext1[i] <- nextbird$SurvivedNext
+    dd3$RemainingLife1[i] <- nextbird$RemainingLife
+    dd3$Died1[i] <- nextbird$Died
     dd3$DeltaCondition[i] <- nextbird$Condition - dd3$Condition[i]
     dd3$DeltaAge[i] <- nextbird$Agemonths - dd3$Agemonths
+    dd3$Agemonths1[i] <- nextbird$Agemonths
   }
   
   

### 2016-10-12 Manuscript/SW_early_life_telomeres.Rmd
Edits

@@
 2.
-We provide a detailed analysis of how telomere dynamics vary over individual lifespans, and with spatiotemporal variation in the social and ecological environment, in the Seychelles warbler (*Acrocephalus sechellensis*).
-We use one of the largest longitudinal datasets to date, consisting of `r nrow(dd)` samples from multiple cohorts spanning `r max(juv$LayYear)-min(juv$LayYear)` years.
+The Seychelles warbler (*Acrocephalus sechellensis*) is an excellent model system to the causes and consequences of individual variation in telomere dynamics.
+Here we combine this long-term ecological data with one the largest longitudinal telomere datasets to date, consisting of `r nrow(dd)` samples from multiple cohorts spanning `r max(juv$LayYear)-min(juv$LayYear)` years.
+We provide a detailed analysis of how telomere dynamics vary over individual lifespans, and with spatiotemporal variation in the social and ecological environment.
 
@@
 4.
-Temporal length and rates of shortening varied markedly both within and among cohorts.
-Aside from age-related measurements, the only significant ecological predictor of telomer length  was temporal variation in food availability.
+Telomere length and rates of shortening varied markedly both within and among cohorts.
+Telomere length  was positively associated with island-wide temporal variation in food availability.
 We also found that cohort and age effects led to spurious associations between telomere dynamics and survival.
@@
 which(is.na(dd3$RTL1))
+
+plot(DeltaRTL~DeltaAge,data=dd3)
+
+summary(lm(DeltaRTL~DeltaInsect,data=subset(dd3,Ageclass !='A'),na.action=na.exclude))
+summary(lm(RTL~Insect,data=juv,na.action=na.exclude))
+
+ggplot(dd3,aes(x = DeltaInsect,y = DeltaRTL,col = Ageclass))+
+geom_point()+
+stat_smooth(method = 'lm')

### 2016-10-12 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Edits

@@
 
-temp <- subset(dd,!is.na(Insect))
+temp <- subset(subset(dd,!is.na(Insect)),!is.na(Malaria))
 temp$LogAge <- log(temp$Agemonths+1)
@@
 #Run the model
-lm2 <- lmer(RTL ~  LogAge+Tarsus + Helper + NonHelper + Sex + BodyMass  +yday + Insect + Density + (1|LayYear) + (1|BirdID),
+lm2 <- lmer(RTL ~  LogAge+Tarsus + Helper + NonHelper + Sex + BodyMass  +yday + Insect + Density + Malaria + (1|LayYear) + (1|BirdID),
             data = temp,

### 2016-10-12 Scripts/SW juvenile telomeres IPC load.r

@@ -26,3 +26,4 @@ status <- read.csv('Data/SW TL all breedstatus.csv')
 insects <- read.csv('Data/Insects.csv')
 dens <- read.table('Data/PsizeFP.txt')
 chickinfo <- read.csv('Data/Table chick info.csv')
+mal <- read.csv('Data/Malaria.csv')

### 2016-10-19 Manuscript/SW_early_life_telomeres.Rmd
Edits

@@
 2.
-The Seychelles warbler (*Acrocephalus sechellensis*) is an excellent model system to the causes and consequences of individual variation in telomere dynamics.
-Here we combine this long-term ecological data with one the largest longitudinal telomere datasets to date, consisting of `r nrow(dd)` samples from multiple cohorts spanning `r max(juv$LayYear)-min(juv$LayYear)` years.
+The Seychelles warbler (*Acrocephalus sechellensis*) is an excellent model system for disentangling the causes and consequences of individual variation in telomere dynamics.
+Here we study spatiotemporal variation in lifelong telomere dynamics in the Seychelles warbler.
+We combine long-term ecological data with one the largest longitudinal telomere datasets to date, consisting of `r nrow(dd)` samples from multiple cohorts spanning `r max(dd$CatchYear)-min(dd$CatchYear)` years.
 We provide a detailed analysis of how telomere dynamics vary over individual lifespans, and with spatiotemporal variation in the social and ecological environment.
@@
 5.
-Our comprehensive, long-term data suggest that although telomeres shorten with age, in natural populations telomere dynamics vary enormously over both space and time, and that this variation cannot be attributed entirely to measurment error.
+Our comprehensive, long-term data show that in natural populations telomere dynamics vary enormously over both space and time, and that this variation cannot be attributed entirely to measurment error.
 Ascertaining what explains this variation will be a difficult but worthwhile task, and will require combining long-term ecological and quantitative genetic data.
@@
 ###Introduction
-Exposure to favourable environmental conditions during development and growth can confer fitness advantages later in life [so called 'silver spoon effects'; @Grafen1988; @Monaghan2008].
-Adult fitness can be affected by a range of early-life experiences, including the quality of the habitat and available resources [@Madsen2000; @VandePol2006; @Hayward2013], population density [@Nussey2007;@Douhard2013] and natural or anthropogenic environmental disturbance [@Reid2003; @Cartwright2014].
-Understanding the causes and consequences of these early-life experiences is key to understanding many ecological and evolutionary  processes, including  patterns of natural and sexual selection,  population growth rates and even local extinction [@Coulson2001; @Roach2014].
-Understanding silver spoon effects is therefore of central interest to ecologists, evolutionary biologists and conservationists.
-
-While it is clear that silver spoon effects can occur, we have little understanding of how and under what conditions early-life environments will affect adult fitness.
-We expect the later-life consequences of a good or poor start to vary among individuals, populations and species because the phenotypic consequences of an adverse environment can depend on an individual's initial condition, and genetic or epigenetic makeup [@Hoffman2000; @Richards2006].
-Moreover, it is not always possible to fully quantify what constitutes a good or bad environment, and any 'hidden' environmental variation may obscure relationships between the early-life conditions that are measured and the resulting adult phenotypes.
-Indeed, the pervasiveness of silver spoon effects varies between species [@Drummond2011], cohorts [@Reid2003] and sexes [@Wilkin2009].
-We therefore need to understand, or at least be able to measure, how the environment affects individuals differentially within a population in order to elucidate the later-life consequences of early-life experience.
-
 Telomeres - the protective caps on the ends of chromosomes - may provide a solution to this problem.
@@
 We found that modelling a non-linear effect age explained almost double the variation in telomere length (R^2^ of linear regression of telomere length and log age = `r round3(summary(lm(RTL~log(Agemonths),data=dd))$r.squared)` than a linear model `r round3(summary(lm(RTL~Agemonths,data=dd))$r.squared)`.
-Also worth noting is that while the overall trend was for decreasing delomre length, there were clear increases in cross-sectional telomere length after one year of age and at several points later in life (Fig.
+Also worth noting is that while the overall trend was for decreasing telomere length, there were clear increases in cross-sectional telomere length after one year of age and at several points later in life (Fig.
 1A).
-RTL also decreased with age within the first year of life (i.e.
-excluding adult samples; t = `r round2(summary(lme(RTL~Agemonths,data=juv,random=~1|BirdID))$tTable[2,4])`; P `r round3(summary(lme(RTL~Agemonths,data=juv,random=~1|BirdID))$tTable[2,5])`).
-In the longitudinal analysis RTL decreased with age within individuals (i.e.
-with $\Delta$age; t = `r round2(summary(lme(RTL~DeltaAge,data=subset(dd,DeltaAge!=0),random=~1|BirdID))$tTable[2,4])`; P = `r round3(summary(lme(RTL~DeltaAge,data=subset(dd,DeltaAge!=0),random=~1|BirdID))$tTable[2,5])`; Fig.
-1B).
 
@@
 
-To test whether the apparent telomere lengthening in our dataset could be explained by measurement error, we used the entire Seychelles warbler telomere dataset to compare variance in telomere length among repeat measurements of samples to the variance observed among different samples of the same individual (see methods for details).
+To test whether the apparent telomere lengthening in our dataset could be explained by measurement error, we compared variance in telomere length among repeat measurements of samples to the variance observed among different samples of the same individual (see methods for details).
 We found significantly higher variance in telomere length over individual lifetimes compared to among sample replicates (Levene's test: F = `r round2(leveneTest(DeltaRTL~Group,data=ddL)['F value'][1,1])`; P `r round3(leveneTest(DeltaRTL~Group,data=ddL)['Pr(>F)'][1,1])`; Fig.

### 2016-10-19 Manuscript/SW_early_life_telomeres.Rmd
Home edits

@@
 
-temp <- subset(dd,Ageclass !='CH')
-
-ggplot(temp,aes(x = Insect,y = RTL))+
-geom_boxplot(aes(group = Insect))+
-geom_smooth()+
-theme_bw()
-
-dd$LogAge <- log(dd$Agemonths)
-m1 <- lmer(RTL~LogAge+(1|BirdID),data=dd)
-m2 <- lmer(RTL~LogAge+(1|BirdID)+(1|TerritoryID),data=dd)
-m3 <- lmer(RTL~LogAge+(1|BirdID)+(1|LayYear),data=dd)
-m4 <- lmer(RTL~LogAge+(1|BirdID)+(1|CatchYear)+(1|LayYear),data=dd)
-
-anova(m1,m2,m3,m4)
-
-head(dd3)
-fit <- coxph(Surv(RemainingLife, RemainingLife1, Died1) ~ Agemonths1 + RTL1,
-              data=dd3)
-
- dd3 <- subset(dd3,RemainingLife-RemainingLife1 >0)
-
-which(is.na(dd3$RemainingLife1))
-which(is.na(dd3$RTL1))
-
-plot(DeltaRTL~DeltaAge,data=dd3)
-
-summary(lm(DeltaRTL~DeltaInsect,data=subset(dd3,Ageclass !='A'),na.action=na.exclude))
-summary(lm(RTL~Insect,data=juv,na.action=na.exclude))
-
-ggplot(dd3,aes(x = DeltaInsect,y = DeltaRTL,col = Ageclass))+
-geom_point()+
-stat_smooth(method = 'lm')
+temp <- subset(dd3,BirdID %in% juv$BirdID)
+nrow(temp)

### 2016-10-19 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Home edits

@@
 
-temp <- subset(juv_r,Season == 'Major')
+temp <- dd
 Fig3C <- ggplot(temp,
@@
            y = RTL,
-           x=yday,
-       col = (Ageclass))) +
+           x=Insect)) +
   theme_classic() +
@@
   geom_point()+
-   stat_smooth(method = 'lm',formula = y~poly(x,2))+
+   stat_smooth(method = 'lm')+
   # scale_colour_manual(values = c('grey','darkorange','darkgreen'))+

### 2016-10-19 Scripts/SW juvenile telomeres IPC clean.R

@@ -213,7 +213,6 @@ for(i in 1:nrow(dd))
   }
 }
 
-
 #Get rid of NAs and cross-fostered birds
 dd <- subset(dd,!is.na(Tarsus))
 dd <- subset(dd,!is.na(BodyMass))

### 2016-10-20 Manuscript/SW_early_life_telomeres.Rmd
Edits

@@
 We performed all analyses using R version 3.0.1 [@RDevelopmentCoreTeam2011].
-Our dataset differs from slightly from that of Barrett *et al.* [-@Barrett2013a] in that we are using RTL calculated directly from the qPCR data rather than converting into an absolute measure , and we are including many additional samples.
+Our dataset differs from slightly from that of Barrett *et al.* [-@Barrett2013a] in that we are using RTL calculated directly from the qPCR data rather than converting into an absolute measure, and we are including many additional samples.
 We therefore first re-tested, using all available samples, whether telomere length decreases with age and $\Delta$age [a longitudinal measure based on within-subject centring; @VandePol2009] in the Seychelles warbler, using linear mixed models with individual ID as a random effect.

### 2016-10-20 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Edits

@@
 
-temp <- subset(subset(dd,!is.na(Insect)),!is.na(Malaria))
-temp$LogAge <- log(temp$Agemonths+1)
+temp <- subset(dd,!is.na(Insect))
+temp$LogAge <- log(temp$Agemonths)
 
 #Run the model
-lm2 <- lmer(RTL ~  LogAge+Tarsus + Helper + NonHelper + Sex + BodyMass  +yday + Insect + Density + Malaria + (1|LayYear) + (1|BirdID),
+lm2 <- lmer(RTL ~  LogAge + Tarsus + Helper + NonHelper + Sex + BodyMass  +CatchMonth + Insect + Density + (1|LayYear) + (1|BirdID) + (1|PlateID),
             data = temp,
@@
 nams[nams == 'c.Sex'] <- 'Sex (males)'
-nams[nams == 'z.TQ'] <- 'TQ'
 nams[nams == 'c.Helper'] <- 'Helpers'
 nams[nams == 'z.BodyMass'] <- 'Body mass'
-nams[nams == 'z.Agemonths'] <- 'Age'
+nams[nams == 'z.LogAge'] <- 'Log age'
 nams[nams == 'z.Tarsus'] <- 'Tarsus'
-nams[nams == 'c.Season'] <- 'Season (minor)'
-nams[nams == 'c.Season:z.Agemonths'] <- 'Age x season'
-nams[nams == 'z.Agemonths:z.NonHelper'] <- 'Age x non helpers'
-nams[nams == 'c.Sex:z.Agemonths'] <- 'Age x sex'
-nams[nams == 'z.Agemonths:z.TQ'] <- 'Age x TQ'
-nams[nams == 'c.Helper:z.Agemonths'] <- 'Age x helpers'
-nams[nams == 'z.Agemonths:z.BodyMass'] <- 'Age x mass'
-nams[nams == 'z.Agemonths:z.Tarsus'] <- 'Age x tarsus'
+nams[nams == 'z.Insect'] <- 'Insect abundance'
+nams[nams == 'z.CatchMonth'] <- 'Month'
 
@@
 
-temp[,'Helpers'] <- ifelse(temp$Helper>0,'Present', 'Absent')
-temp[,'Non helpers'] <- ifelse(temp$NonHelper>0,'Present', 'Absent')
-temp[,'Age (months)'] <- factor(temp$Agemonths) 
-temp$Age2 <- ifelse(temp$Ageclass == 'A','Adults','Juveniles')
+temp$Age2 <- ifelse(temp$Ageclass == 'A',
+                    ifelse(temp$Sex == 'Males',
+                           'Adult males',
+                           'Adult females'),
+                    ifelse(temp$Sex == 'Males',
+                           'Juvenule males',
+                           'Juvenile females'))
 
@@
         legend.position = 'top') +
-  geom_point()+
-  stat_smooth(method = 'lm')+
+  stat_smooth(method = 'lm')
   annotate('text',x = 17,y = 2.1,label = 'B')
 
-temp <- dd
+temp$InsectF <- ifelse(temp$Insect>4,'High','Low')
 Fig3C <- ggplot(temp,
@@
            y = RTL,
-           x=Insect)) +
+           x=Agemonths,
+           col = InsectF)) +
   theme_classic() +
@@
   geom_point()+
-   stat_smooth(method = 'lm')+
+   stat_smooth(method = 'lm',formula = y~log(x))
   # scale_colour_manual(values = c('grey','darkorange','darkgreen'))+
@@
 ```
+
+anova(lm(RTL~Sex+Agemonths+factor(CatchYear)+Sex*factor(LayYear),data=dd))
+
+ggplot(dd,aes(x = factor(LayYear),y = RTL,col = Sex))+
+geom_boxplot(notch=F)
 

### 2016-10-24 Manuscript/SW_early_life_telomeres.Rmd
More edits

@@
 Our comprehensive, long-term data show that in natural populations telomere dynamics vary enormously over both space and time, and that this variation cannot be attributed entirely to measurment error.
-Ascertaining what explains this variation will be a difficult but worthwhile task, and will require combining long-term ecological and quantitative genetic data.
+Ascertaining what explains this variation will require combining long-term ecological and quantitative genetic data.
 
@@
 #####Early-life telomere dynamics and survival
-Early-life RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~cenTL+Agemonths+(1|PlateID)+(1|TerritoryID)+(1|FieldPeriodID),data=juv14,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~cenTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv14,family = 'binomial'))$coefficients[2,4])`), with no interaction effect between RTL and juvenile age (estimate = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv14,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv14,family = 'binomial'))$coefficients[4,4])`), suggesting that there were no age-specific effects of juvenile RTL on survival to adulthood.
+Early-life RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~cenTL+Agemonths+(1|PlateID)+(1|TerritoryID)+(1|LayYear),data=juv14,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~cenTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv14,family = 'binomial'))$coefficients[2,4])`), with no interaction effect between RTL and juvenile age (estimate = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv14,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv14,family = 'binomial'))$coefficients[4,4])`), suggesting that there were no age-specific effects of juvenile RTL on survival to adulthood.
 To test whether cohort-level effects on RTL influenced survival we calculated cohort-level RTL (i.e.
@@
 
-temp <- subset(dd3,BirdID %in% juv$BirdID)
-nrow(temp)
+m1 <- lmer(RTL~LogAge+(1|BirdID),data=dd)
+m2 <- lmer(RTL~LogAge+(1|BirdID) + (1|LayYear),data=dd)
+m3 <- lmer(RTL~LogAge+(1|BirdID) + (LogAge|LayYear),data=dd)
+
+anova(m1,m2,m3)
+
+dd14 <- subset(dd,DeathYear<2015)
+dd14$RowID <- c(1:nrow(dd14))
+
+summary(glmer(SurvivedNext~RTL * LogAge + Tarsus + Sex + (1|BirdID)+(1|LayYear),data=dd14,family = 'binomial'))
+
+ggplot(dd14,aes(x = Agemonths, col = factor(SurvivedNext),y = RTL))+
+geom_smooth(method = 'lm',formula = y~log(x))

### 2016-10-24 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
More edits

@@
 nams[nams == 'z.CatchMonth'] <- 'Month'
+nams[nams == 'z.Density'] <- 'Density'
 
@@
 Fig3B <- ggplot(temp,
-       aes(x = Tarsus,
+       aes(x = Agemonths,
            y = RTL,
-           col = Age2)) +
+           col = Sex)) +
   ylab('RTL') + 
-  xlab('Tarsus length (mm)') +
+  xlab('') +
   theme_classic() +
@@
         legend.position = 'top') +
-  stat_smooth(method = 'lm')
+  stat_smooth(method = 'lm',formula = y~log(x))
   annotate('text',x = 17,y = 2.1,label = 'B')
 
-temp$InsectF <- ifelse(temp$Insect>4,'High','Low')
 Fig3C <- ggplot(temp,
@@
            y = RTL,
-           x=Agemonths,
-           col = InsectF)) +
+           x=Insect,
+           col = Sex)) +
   theme_classic() +
@@
 
-anova(lm(RTL~Sex+Agemonths+factor(CatchYear)+Sex*factor(LayYear),data=dd))
-
-ggplot(dd,aes(x = factor(LayYear),y = RTL,col = Sex))+
-geom_boxplot(notch=F)
-
-temp <- subset(dd,Age<15)
-reps <- tapply(temp$RTL,temp$BirdID,length)
-keep <- names(reps[reps>2])
-temp <- subset(temp,BirdID %in% keep)
- temp <- subset(temp,!duplicated(paste0(BirdID,Agemonths)))
-
-ggplot(temp,aes(x = CatchYear,y = RTL,group = BirdID,col=factor(BirdID)))+
-  geom_point(alpha = 1,col='grey')+
-geom_line(col = 'darkgrey',alpha = 0.2)+
-  xlab('Age (years)') +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black'),
-        legend.position = 'none')+
-  #scale_colour_brewer(palette = 'Set1')+
-  annotate('text',x = 1997,y = 2,label = 'B')
-
-ggplot(ddL,aes(x = DeltaRTL,y = RTL,col = Group))+
-geom_point()+
-stat_smooth(method = 'lm')
-
-```{r Figure X}
-
-temp <- dd
-#temp$SurvivedNext <- factor(temp$SurvivedNext)
-temp$SN2 <- with(temp,paste(Sex,SurvivedNext))
-
-FigX <- ggplot(temp,aes(x = LayYear,y = RTL,col = SN2))+
-  ylab('RTL') + 
-  xlab('Cohort') +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12,angle = 90), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-                axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black'),
-        legend.position = 'top')+
-  geom_point()+
-  stat_smooth()
-
-FigX
-
-FigX <- ggplot(subset(temp,Agemonths),aes(x = SN2,y = RTL))+
-  ylab('RTL') + 
-  xlab('Cohort') +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12,angle = 90), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-                axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black'),
-        legend.position = 'top')+
-  geom_boxplot()
-
-FigX
-
-```
-
-temp <- dd
-temp$LogAge <- log(temp$Agemonths)
-
-m1 <- lm(RTL~LogAge,data=temp)
-m2 <- lmer(RTL~LogAge+(1|CatchYear),data=temp)
-m3 <- lmer(RTL~LogAge+(LogAge|CatchYear),data=temp)
-m4 <- lmer(RTL~LogAge+(1|LayYear),data=temp)
-m5 <- lmer(RTL~LogAge+(LogAge|LayYear),data=temp)
-
-m1 <- lmer(RTL~(Agemonths)+(1|BirdID)+(1|LayYear),data=dd)
-m2 <- lmer(RTL~Ageclass+(1|BirdID)+(1|LayYear),data=dd)
-m3 <- lmer(RTL~LogAge+(1|BirdID)+(1|LayYear),data=dd)
-
-anova(m1,m2,m3)
-summary(m2)
-summary(m3)
-
-dd3$LogAge <- log(dd3$Agemonths)
-m1 <- lmer(RTL1~RTL+(1|LayYear),data=dd3)
-summary(m1)
+plot(tapply(dd3$DeltaRTLF,dd3$TimeDiff,mean))

### 2016-10-24 Scripts/SW juvenile telomeres IPC clean.R

@@ -4,7 +4,7 @@
 #################################################################################
 
 #Only use Ellie's samples
-dd0 <- subset(dd0,Whodunnit == 'EAF')
+# dd0 <- subset(dd0,Whodunnit == 'EAF')
 
 #Average repeats of blood samples
 dd0$RTL2 <- ave(dd0$RTL,c(dd0$BloodID,dd0$Status,dd0$PlateID))
@@ -55,8 +55,8 @@ dd$Agemonths <- ifelse(dd$Ageclass == 'CH',1,
 
 
   #dd$Age[dd$Fledged != 'Adults'] <- 1
-
-
+dd$LogAge <- log10(dd$Agemonths)
+dd <- subset(dd,!is.na(LogAge))
 
 
 # Tarsus and delta age ------------------------------------------------------------------

### 2016-10-25 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Edits

@@
 
-temp <- subset(dd,LayYear>1995)
-temp$JA <- ifelse(temp$Ageclass == 'A','Adults','Juveniles')
+ys <- names(which(table(subset(dd,Ageclass == 'CH')$LayYear)>10))
+temp <- subset(dd,LayYear %in% ys)
+temp$Cohort <- factor(temp$LayYear)
 
-Fig2 <- ggplot(temp,aes(x = LayYear,y = RTL))+
-  ylab('RTL') + 
-  xlab('Cohort') +
+Fig2A <- ggplot(temp,aes(x = Agemonths,y = RTL,col = Cohort))+
+  xlab('Age (months)') +
   theme_classic() +
-  theme(axis.text.x = element_text(size = 12,angle = 90), 
+  theme(axis.text.x = element_text(size = 12), 
         axis.text.y = element_text(size = 12), 
@@
         axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black'),
-        legend.position = 'top')+
-  geom_boxplot(aes(group = LayYear),outlier.colour = 'grey',notch=F,alpha=0)+
-  geom_smooth()
+        axis.line.y = element_line(colour = 'black'))+
+  geom_smooth(method = 'lm',formula = y~log(x),se = F)+
+  annotate('text',x = 0,y = 1.3,label = 'A')
+
+Fig2B <- ggplot(temp,aes(x = Agemonths,y = RTL,col = Cohort))+
+  xlab('Age (months)') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'))+
+  geom_smooth(se = F)+
+  annotate('text',x = 0,y = 1.4,label = 'B')
 
 pdf('Figures/Figure 2.pdf',
-    width = 8,
-    height = 4,
-    paper = 'a4r')
-Fig2
+    width = 5,
+    height = 8,
+    paper = 'a4')
+multiplot(Fig2A,Fig2B)
 dev.off()

### 2016-10-26 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Edits

@@
 ```{r Figure 1}
-
+subset
 temp <- subset(dd,Age<15)
@@
   xlab('Age (months)') +
+  geom_point()+
   theme_classic() +
@@
 #Run the model
-lm2 <- lmer(RTL ~  LogAge + Tarsus + Helper + NonHelper + Sex + BodyMass  +CatchMonth + Insect + Density + (1|LayYear) + (1|BirdID) + (1|PlateID),
+lm2 <- lmer(RTL ~  LogAge + Tarsus + Helper + NonHelper + Sex + BodyMass + Insect + Density + (1|LayYear) + (1|BirdID) + (1|PlateID),
             data = temp,
@@
 
-temp$Age2 <- ifelse(temp$Ageclass == 'A',
-                    ifelse(temp$Sex == 'Males',
-                           'Adult males',
-                           'Adult females'),
-                    ifelse(temp$Sex == 'Males',
-                           'Juvenule males',
-                           'Juvenile females'))
-
   Fig3B <- ggplot(temp,
@@
            col = Sex)) +
+  geom_point()+
   ylab('RTL') + 
-  xlab('') +
+  xlab('Age (months)') +
   theme_classic() +
@@
         axis.line.y = element_line(colour = 'black'),
-        legend.position = 'top') +
-  stat_smooth(method = 'lm',formula = y~log(x))
-  annotate('text',x = 17,y = 2.1,label = 'B')
+        legend.position = 'none') +
+  stat_smooth(method = 'lm',formula = y~log(x))+
+  annotate('text',x = 1,y = 2.5,label = 'B')
 
 Fig3C <- ggplot(temp,
-       aes(
+       aes(x = Tarsus,
            y = RTL,
-           x=Insect,
            col = Sex)) +
+  geom_point()+
+  ylab('RTL') + 
+  xlab('Tarsus length (mm)') +
   theme_classic() +
@@
         axis.line.y = element_line(colour = 'black'),
-        legend.position = 'top') +
-  geom_point()+
-   stat_smooth(method = 'lm',formula = y~log(x))
-  # scale_colour_manual(values = c('grey','darkorange','darkgreen'))+
-  # scale_fill_manual(values = c('grey','darkorange','darkgreen'))+
-  annotate('text',x = 3,y = 2.1,label = 'C')
+        legend.position = 'none') +
+  stat_smooth(method = 'lm')+
+  annotate('text',x = 17,y = 2.5,label = 'C')
+
+Fig3D <- ggplot(temp,
+       aes(
+           y = RTL,
+           x=Insect)) +
+    ylab('RTL') + 
+  xlab('Insect abundance') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+                axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'),
+        legend.position = 'none') +
+  geom_boxplot(aes(group = Insect),outlier.shape = NA)+
+   stat_smooth(method = 'lm',col = 'darkgrey')+
+  annotate('text',x = 2.5,y = 2.5,label = 'D')
+
+Fig3E <- ggplot(temp,
+       aes(
+           y = RTL,
+           x=Density)) +
+    ylab('RTL') + 
+  xlab('Density') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+                axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'),
+        legend.position = 'none') +
+  geom_boxplot(aes(group = Insect),outlier.shape = NA)+
+   stat_smooth(method = 'lm',col = 'darkgrey')+
+  annotate('text',x = 270,y = 2.5,label = 'E')
 
@@
     width = 8,
-    height = 8,
-    paper = 'a4r')
-multiplot(Fig3A,Fig3B,Fig3C,layout = matrix(c(1,1,2,3),nrow = 2,byrow = T))
+    height = 12,
+    paper = 'a4')
+multiplot(Fig3A,Fig3B,Fig3C,Fig3D,Fig3E,layout = matrix(c(1,1,2,3,4,5),nrow = 3,byrow = T))
 dev.off()
@@
 plot(tapply(dd3$DeltaRTLF,dd3$TimeDiff,mean))
+
+ggplot(dd,aes(col = Whodunnit,y = RTL,x = factor(LayYear)))+
+geom_boxplot()

### 2016-10-31 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Edits

@@
            y = RTL1)) +
-  ylab('Adult RTL') + 
-  xlab('juvenile RTL') +
+  ylab(expression('RTL ('*italic(t)*' + 1)')) + 
+  xlab(expression('RTL ('*italic(t)*')')) +
     xlim(c(0,3))+
@@
 temp$LogAge <- log(temp$Agemonths)
+temp$RTL <- log10(temp$RTL)
 
@@
 
-plot(tapply(dd3$DeltaRTLF,dd3$TimeDiff,mean))
-
-ggplot(dd,aes(col = Whodunnit,y = RTL,x = factor(LayYear)))+
-geom_boxplot()
+library(qqman)
+?manhattan

### 2016-10-31 Manuscript/SW_early_life_telomeres.Rmd
Edits
Home edits

@@
 
+m1 <- lmer(RTL~Agemonths+(1|BirdID) + (1|LayYear),data=dd)
+m2 <- lmer(RTL~LogAge+(1|BirdID) + (1|LayYear),data=dd)
+m3 <- lmer(RTL~factor(Agemonths)+(1|BirdID) + (1|LayYear),data=dd)
+
+anova(m1,m2,m3)
+
 dd14 <- subset(dd,DeathYear<2015)
-dd14$RowID <- c(1:nrow(dd14))
 
-summary(glmer(SurvivedNext~RTL * LogAge + Tarsus + Sex + (1|BirdID)+(1|LayYear),data=dd14,family = 'binomial'))
+summary(glmer(SurvivedNext~RTL*LogAge + (1|LayYear) + (1|BirdID),data=dd14,family = 'binomial'))
 
-ggplot(dd14,aes(x = Agemonths, col = factor(SurvivedNext),y = RTL))+
-geom_smooth(method = 'lm',formula = y~log(x))
+?coxph
+summary(coxph(Surv(RemainingLife,Died)~RTL*LogAge,data=dd14))

### 2016-10-31 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Home edits

@@
 temp <- dd3
+dd3
+props <- tapply(dd3$DeltaRTLF,dd3$TimeDiff,mean)
+propn <- tapply(dd3$DeltaRTLF,dd3$TimeDiff,length)
+temp <- data.frame(props,propn,TimeDiff = as.numeric(names(props))/12) %>%
+  subset(propn > 10)
 
-Fig1D <- ggplot(temp,aes(y = DeltaRTL,x = TimeDiff))+
+Fig1D <- ggplot(temp,aes(y = props,x = TimeDiff))+
     geom_point(col = 'grey')+
@@
   xlab ('Follow up time (years)')+
-  ylab(expression(Delta*'RTL'))+
-  ylim(c(-2,2))+
+  ylab('Proportion gainers')+
     theme(axis.text.x = element_text(size = 12), 
@@
         axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black')) +
-    geom_abline(slope = 0,intercept = 0,lty = 2)+
-  annotate('text',x = 0,y = 2,label = 'D')
+        axis.line.y = element_line(colour = 'black'))
 

### 2016-11-02 Manuscript/SW_early_life_telomeres.Rmd
Eds

--- 2016-10-20 Manuscript/SW_early_life_telomeres.Rmd
@@
 
-#####Early-life telomere dynamics and survival
-Early-life RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~cenTL+Agemonths+(1|PlateID)+(1|TerritoryID)+(1|FieldPeriodID),data=juv14,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~cenTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv14,family = 'binomial'))$coefficients[2,4])`), with no interaction effect between RTL and juvenile age (estimate = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv14,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv14,family = 'binomial'))$coefficients[4,4])`), suggesting that there were no age-specific effects of juvenile RTL on survival to adulthood.
-To test whether cohort-level effects on RTL influenced survival we calculated cohort-level RTL (i.e.
-the median RTL for each cohort) and added this term to a GLMM, and found that cohort-level RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~cohortTL+RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv14,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~cohortTL+RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv14,family = 'binomial'))$coefficients[2,4])`).
+#####Telomere dynamics and survival
+Early-life RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~RTL+LogAge+(1|PlateID)+(1|TerritoryID)+(1|LayYear),data=dd14,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL+LogAge+(1|TerritoryID)+(1|LayYear),data=dd14,family = 'binomial'))$coefficients[2,4])`), with no interaction effect between RTL and  age (estimate = `r round2(summary(glmer(SurvivedNext~RTL*LogAge+(1|TerritoryID)+(1|LayYear),data=dd14,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL*LogAge+(1|TerritoryID)+(1|LayYear),data=dd14,family = 'binomial'))$coefficients[4,4])`), suggesting that there were no age-specific effects of juvenile RTL on survival to adulthood.
+
+summary(glmer(SurvivedNext1~DeltaRTL+LogAge+(1|LayYear),data=dd3,family = 'binomial'))
 
@@
 ###References
-
-temp <- subset(dd3,BirdID %in% juv$BirdID)
-nrow(temp)

### 2016-11-03 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Eds

--- 2016-10-26 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
@@
         legend.position = 'none')+
-  geom_smooth(col = 'black',fill = 'black',lwd = 0.5,aes(group = NULL),method = 'lm',formula = y~log(x))+
+  geom_smooth(col = 'black',fill = 'black',lwd = 0.5,aes(group = NULL),method = 'lm',formula = y~log(x))
   annotate('text',x = 0,y = 3,label = 'A')
@@
            y = RTL1)) +
-  ylab('Adult RTL') + 
-  xlab('juvenile RTL') +
-    xlim(c(0,3))+
-  ylim(c(0,3))+
+  ylab(expression('RTL ('*italic(t)*' + 1)')) + 
+  xlab(expression('RTL ('*italic(t)*')')) +
   theme_classic() +
@@
     geom_point(col = 'grey')+
-  geom_abline(slope = 1,intercept = 0,lty = 2)+
+  geom_abline(slope = 1,intercept = 0,lty = 2)
   annotate('text',x = 0,y = 3,label = 'B')
@@
   scale_fill_manual(values = c('darkgrey','gold'))+
-  geom_vline(xintercept = 0,lty = 2)+
+  geom_vline(xintercept = 0,lty = 2)
   annotate('text',x = -2,y = 1,label = 'C')
@@
   ylab(expression(Delta*'RTL'))+
-  ylim(c(-2,2))+
     theme(axis.text.x = element_text(size = 12), 
@@
         axis.line.y = element_line(colour = 'black')) +
-    geom_abline(slope = 0,intercept = 0,lty = 2)+
+    geom_abline(slope = 0,intercept = 0,lty = 2)
   annotate('text',x = 0,y = 2,label = 'D')
@@
         axis.line.y = element_line(colour = 'black'))+
-  geom_smooth(method = 'lm',formula = y~log(x),se = F)+
+  geom_smooth(method = 'lm',formula = y~log(x),se = F)
   annotate('text',x = 0,y = 1.3,label = 'A')
 
-Fig2B <- ggplot(temp,aes(x = Agemonths,y = RTL,col = Cohort))+
+temp2 <- ddply(temp,
+               .(Agemonths,Cohort),
+               summarise,
+               MedianRTL = mean(RTL))  %>%
+  subset(Agemonths < 20)
+
+Fig2B <- ggplot(temp2,aes(x = Agemonths,y = MedianRTL,col = Cohort))+
   xlab('Age (months)') +
-  geom_point()+
+  geom_line()+
   theme_classic() +
@@
         axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black'))+
-  geom_smooth(se = F)+
+        axis.line.y = element_line(colour = 'black'))
   annotate('text',x = 0,y = 1.4,label = 'B')
@@
 Fig3A <- forestplot(table2)+
-  annotate('text',x = 0.6,y = 0.25,label = 'A')+
+  annotate('text',x = 0.6,y = 0.05,label = 'A')+
   theme(axis.line.x = element_line(colour = 'black'),
@@
         legend.position = 'none') +
-  stat_smooth(method = 'lm',formula = y~log(x))+
+  stat_smooth(method = 'lm',formula = y~log(x))
   annotate('text',x = 1,y = 2.5,label = 'B')
@@
         legend.position = 'none') +
-  stat_smooth(method = 'lm')+
+  stat_smooth(method = 'lm')
   annotate('text',x = 17,y = 2.5,label = 'C')
@@
   geom_boxplot(aes(group = Insect),outlier.shape = NA)+
-   stat_smooth(method = 'lm',col = 'darkgrey')+
+   stat_smooth(method = 'lm',col = 'darkgrey')
   annotate('text',x = 2.5,y = 2.5,label = 'D')
@@
   geom_boxplot(aes(group = Insect),outlier.shape = NA)+
-   stat_smooth(method = 'lm',col = 'darkgrey')+
+   stat_smooth(method = 'lm',col = 'darkgrey')
   annotate('text',x = 270,y = 2.5,label = 'E')
@@
 ```
-
-plot(tapply(dd3$DeltaRTLF,dd3$TimeDiff,mean))
-
-ggplot(dd,aes(col = Whodunnit,y = RTL,x = factor(LayYear)))+
-geom_boxplot()

### 2016-11-03 Scripts/SW juvenile telomeres IPC clean.R

@@ -7,17 +7,19 @@
 # dd0 <- subset(dd0,Whodunnit == 'EAF')
 
 dd0 <- subset(dd0,RTL > 0.05) %>%
-  subset(CqTelomere <28) %>%
+  subset(CqTelomere < 28) %>%
   subset(CqGAPDH < 26) %>%
   subset(CqGAPDH > 21) %>%
-  subset(RTL < 2.5)
+  subset(RTL < 3)
 
-
+dd0$RTL <- sqrt(dd0$RTL)
 #Average repeats of blood samples
 dd0$RTL2 <- ave(dd0$RTL,c(dd0$BloodID,dd0$Status,dd0$PlateID))
+
 dd <- dd0[!(duplicated(dd0$BloodID)),]
 dd$RTL <- dd$RTL2
 dd$RTL2 <- NULL
+
 
 # Weird variable names ----------------------------------------------------
 

### 2016-11-03 Scripts/SW juvenile telomeres IPC load.r

@@ -18,6 +18,7 @@ library(survival)
 library(nlme)
 library(magrittr)
 library(car)
+library(rptR)
 
 # Load data ---------------------------------------------------------------
   dd0 <- read.csv('Data/REL_TL_EB_EAF_24 May aft TL queries.csv') #main dataset

### 2016-11-07 Scripts/SW juvenile telomeres IPC clean.R
Edits

@@ -6,15 +6,27 @@
 #Only use Ellie's samples
 # dd0 <- subset(dd0,Whodunnit == 'EAF')
 
-dd0 <- subset(dd0,RTL > 0.05) %>%
-  subset(CqTelomere < 28) %>%
+dd0 <- subset(dd0,CqTelomere < 100) %>%
   subset(CqGAPDH < 26) %>%
-  subset(CqGAPDH > 21) %>%
-  subset(RTL < 3)
+subset(DiffGAPDH < 0.5) %>%
+  subset(DiffTelomere < 0.5) %>%
+  subset(EffGAPDH < 2) %>%
+  subset(RTL < 3) %>%
+  subset(BloodID>7) %>%
+  subset(!duplicated(paste0(CqTelomere,CqGAPDH)))
+
 
 dd0$RTL <- sqrt(dd0$RTL)
+
+
+
 #Average repeats of blood samples
 dd0$RTL2 <- ave(dd0$RTL,c(dd0$BloodID,dd0$Status,dd0$PlateID))
+
+# rpt(RTL~(1|BloodID),grname = 'BloodID',data = dd0,nboot = 50)
+# rpt(RTL~(1|BloodID),grname = 'BloodID',data = subset(dd0,Whodunnit == "EB"),nboot = 50)
+# rpt(RTL~(1|BloodID),grname = 'BloodID',data = subset(dd0,Whodunnit == "EAF"),nboot = 50)
+
 
 dd <- dd0[!(duplicated(dd0$BloodID)),]
 dd$RTL <- dd$RTL2

### 2016-11-07 Scripts/SW juvenile telomeres IPC load.r

@@ -19,9 +19,10 @@ library(nlme)
 library(magrittr)
 library(car)
 library(rptR)
+library(Rmisc)
 
 # Load data ---------------------------------------------------------------
-  dd0 <- read.csv('Data/REL_TL_EB_EAF_24 May aft TL queries.csv') #main dataset
+dd0 <- read.csv('Data/REL_TL_EB_EAF_24 May aft TL queries.csv') #main dataset
 terr <- read.csv('Data/SW TL all territories.csv')
 status <- read.csv('Data/SW TL all breedstatus.csv')
 insects <- read.csv('Data/Insects.csv')

### 2016-11-07 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Edits
Eds

@@
 #Run the model
-lm2 <- lmer(RTL ~  LogAge + Tarsus + Helper + NonHelper + Sex + BodyMass + Insect + Density + (1|LayYear) + (1|BirdID) + (1|PlateID),
+lm2 <- lmer(RTL ~  LogAge + Tarsus + Helper + GroupSize + Sex + BodyMass + Insect + Density + (1|LayYear) + (1|BirdID) + (1|PlateID),
             data = temp,
@@
 nams <- rownames(table2)
-nams[nams == 'z.NonHelper'] <- 'Non helpers'
+nams[nams == 'z.GroupSize'] <- 'Group Size'
 nams[nams == 'c.Sex'] <- 'Sex (males)'
@@
 Fig3B <- ggplot(temp,
-       aes(x = Agemonths,
-           y = RTL,
-           col = Sex)) +
-  geom_point()+
-  ylab('RTL') + 
-  xlab('Age (months)') +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-                axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black'),
-        legend.position = 'none') +
-  stat_smooth(method = 'lm',formula = y~log(x))
-  annotate('text',x = 1,y = 2.5,label = 'B')
-
-Fig3C <- ggplot(temp,
        aes(x = Tarsus,
@@
 
-Fig3D <- ggplot(temp,
+Fig3C <- ggplot(temp,
        aes(
@@
 
-Fig3E <- ggplot(temp,
-       aes(
-           y = RTL,
-           x=Density)) +
-    ylab('RTL') + 
-  xlab('Density') +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-                axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black'),
-        legend.position = 'none') +
-  geom_boxplot(aes(group = Insect),outlier.shape = NA)+
-   stat_smooth(method = 'lm',col = 'darkgrey')
-  annotate('text',x = 270,y = 2.5,label = 'E')
-
 pdf('Figures/Figure 3.pdf',
     width = 8,
-    height = 12,
+    height = 8,
     paper = 'a4')
-multiplot(Fig3A,Fig3B,Fig3C,Fig3D,Fig3E,layout = matrix(c(1,1,2,3,4,5),nrow = 3,byrow = T))
+multiplot(Fig3A,Fig3B,Fig3C,layout = matrix(c(1,1,2,3),nrow = 2,byrow = T))
 dev.off()
@@
 ```
+
+x <- subset(subset(dd,Ageclass 'A'))
+
+summary(lm(RTL~Tarsus,data = x))
+
+dd$Sex2 <- paste(dd$Fledged,dd$Sex)
+
+ggplot(dd,aes(x = Tarsus, y = RTL,col = Sex2))+
+geom_smooth(method = 'lm')

### 2016-11-07 Manuscript/SW_early_life_telomeres.Rmd
Home eds

@@
 
-Longitudinal data showed that early-life RTL was not related to adult RTL (R^2^ = `r round2(summary(lm(RTL1~RTL+TimeDiff,data = dd3))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data = dd3))$coefficients[2,3])`; P = `r round2(summary(lm(RTL1~RTL+Agemonths,data = dd3))$coefficients[2,4])`; Fig.
+Longitudinal data showed that early-life RTL was weakly, but significantly, related to adult RTL (R^2^ = `r round3(summary(lm(RTL1~RTL,data = dd3))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL,data = dd3))$coefficients[2,3])`; P = `r round3(summary(lm(RTL1~RTL,data = dd3))$coefficients[2,4])`; Fig.
 4A).
-Running this analysis separately for juvenile age classes revealed that this relationship was not significant for nestlings (R^2^ = `r round2(summary(lm(RTL1~RTL,data=subset(dd3,Agemonths == 1)))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL,data=subset(dd3,Agemonths == 1)))$coefficients[2,3])`; P = `r round2(summary(lm(RTL1~RTL,data = subset(dd3,Agemonths == 1)))$coefficients[2,4])`), fledglings (R^2^ = `r round2(summary(lm(RTL1~RTL,data=subset(dd3,Agemonths == 6)))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL,data=subset(dd3,Agemonths == 6)))$coefficients[2,3])`; P = `r round2(summary(lm(RTL1~RTL,data = subset(dd3,Agemonths == 6)))$coefficients[2,4])`) or subadults (R^2^ = `r round2(summary(lm(RTL1~RTL,data=subset(dd3,Agemonths == 10)))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL,data=subset(dd3,Agemonths == 10)))$coefficients[2,3])`; P = `r round2(summary(lm(RTL1~RTL,data = subset(dd3,Agemonths == 10)))$coefficients[2,4])`).
 $\Delta$RTL was significantly lower than zero in nestlings (t = `r round2(t.test(subset(dd3,Agemonths == 1)$DeltaRTL)$statistic)`; P `r round3(t.test(subset(dd3,Agemonths == 1)$DeltaRTL)$p.value)`), but not in fledglings (t = `r round2(t.test(subset(dd3,Agemonths == 6)$DeltaRTL)$statistic)`; P = `r round2(t.test(subset(dd3,Agemonths == 6)$DeltaRTL)$p.value)`) or subadults (t = `r round2(t.test(subset(dd3,Agemonths == 10)$DeltaRTL)$statistic)`; P = `r round2(t.test(subset(dd3,Agemonths == 10)$DeltaRTL)$p.value)`), suggesting that telomere shortening was highest shortly after the nestling phase, after which both telomere shortening and telomere lengthening were observed (Fig.

### 2016-11-07 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Home eds

@@
 
-x <- subset(subset(dd,Ageclass 'A'))
+```{r Figure 3}
 
-summary(lm(RTL~Tarsus,data = x))
+temp <- dd14
 
-dd$Sex2 <- paste(dd$Fledged,dd$Sex)
+Fig4A <- ggplot(temp,
+       aes(x = Agemonths,
+           col = factor(SurvivedNext),
+           y = RTL)) +
+  ylab('RTL') + 
+  xlab('Survival probability') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+                axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'),
+        legend.position = 'none') +
+  geom_smooth(method = 'lm')
+  annotate('text',x = 17,y = 2.5,label = 'C')
 
-ggplot(dd,aes(x = Tarsus, y = RTL,col = Sex2))+
-geom_smooth(method = 'lm')
+temp <- ddply(juv14,
+              .(LayYear,Fledged),
+              summarise,
+              MeanTL = median(RTL),
+              Survived = mean(SurvivedNext),
+              n = length(RTL)) %>%
+  subset(n > 9)
+
+ggplot(temp,aes(x = MeanTL,y = Survived,col = Fledged))+
+  geom_point()+
+  stat_smooth(method = 'lm')
+
+Fig4B <- ??
+
+pdf('Figures/Figure 4.pdf',
+    width = 8,
+    height = 8,
+    paper = 'a4')
+multiplot(Fig4A,Fig4B,layout = matrix(c(1,2),nrow = 2,byrow = T))
+dev.off()
+
+```

### 2016-11-07 Scripts/SW juvenile telomeres IPC clean.R

@@ -6,11 +6,12 @@
 #Only use Ellie's samples
 # dd0 <- subset(dd0,Whodunnit == 'EAF')
 
-dd0 <- subset(dd0,CqTelomere < 100) %>%
+
+dd0 <- subset(dd0,CqTelomere < 25) %>%
   subset(CqGAPDH < 26) %>%
-subset(DiffGAPDH < 0.5) %>%
+  subset(CqGAPDH > 19) %>%
+  subset(DiffGAPDH < 0.5) %>%
   subset(DiffTelomere < 0.5) %>%
-  subset(EffGAPDH < 2) %>%
   subset(RTL < 3) %>%
   subset(BloodID>7) %>%
   subset(!duplicated(paste0(CqTelomere,CqGAPDH)))
@@ -26,6 +27,9 @@ dd0$RTL2 <- ave(dd0$RTL,c(dd0$BloodID,dd0$Status,dd0$PlateID))
 # rpt(RTL~(1|BloodID),grname = 'BloodID',data = dd0,nboot = 50)
 # rpt(RTL~(1|BloodID),grname = 'BloodID',data = subset(dd0,Whodunnit == "EB"),nboot = 50)
 # rpt(RTL~(1|BloodID),grname = 'BloodID',data = subset(dd0,Whodunnit == "EAF"),nboot = 50)
+# 
+# hist(dd0$CqTelomere)
+# hist(dd0$RTL)
 
 
 dd <- dd0[!(duplicated(dd0$BloodID)),]

### 2016-11-08 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Eds

@@
 ```{r Figure 1}
-subset
-temp <- subset(dd,Age<15)
-reps <- tapply(temp$RTL,temp$BirdID,length)
-keep <- names(reps[reps>0])
-temp <- subset(temp,BirdID %in% keep)
-temp <- subset(temp,!duplicated(paste0(BirdID,Agemonths)))
 
- mins <- tapply(temp$Agemonths,temp$BirdID,min)
- keep2 <- names(mins[mins <400])
- temp <- subset(temp,BirdID %in% keep2)
-
-temp$Agemonths <- temp$Agemonths/12
-
-Fig1A <- ggplot(temp,aes(x = Agemonths,y = RTL,group = BirdID,col=factor(BirdID)))+
+Fig1A <- subset(dd,Age<15) %>%
+  ddply(.(BirdID),
+        mutate,
+        nreps = length(RTL)) %>%
+  subset(nreps>0) %>%
+  subset(!duplicated(paste0(BirdID,Agemonths))) %>%
+  transform(Agemonthsdec = Agemonths/12) %>%
+  ggplot(aes(x = Agemonthsdec,y = RTL,group = BirdID,col=factor(BirdID)))+
   geom_point(alpha = 1,col='grey')+
@@
 
-temp <- dd3
-
-Fig1B <- ggplot(temp,
+Fig1B <- ggplot(dd3,
        aes(x = RTL,
@@
 
-temp <- dd3
-dd3
-props <- tapply(dd3$DeltaRTLF,dd3$TimeDiff,mean)
-propn <- tapply(dd3$DeltaRTLF,dd3$TimeDiff,length)
-temp <- data.frame(props,propn,TimeDiff = as.numeric(names(props))/12) %>%
-  subset(propn > 10)
-
-Fig1D <- ggplot(temp,aes(y = props,x = TimeDiff))+
-    geom_point(col = 'grey')+
-  theme_classic()+ 
-  xlab ('Follow up time (years)')+
-  ylab(expression(Delta*'RTL'))+
-    theme(axis.text.x = element_text(size = 12), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        legend.position = c('none'),
-        legend.title = element_blank(),
-        axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black')) +
-    geom_abline(slope = 0,intercept = 0,lty = 2)
-  annotate('text',x = 0,y = 2,label = 'D')
-
 pdf('Figures/Figure 1.pdf',
     width = 8,
-    height = 7,
+    height = 8,
     paper = 'a4')
-multiplot(Fig1A,Fig1B,Fig1C,Fig1D,layout=matrix(c(1,2,3,4),nrow=2,byrow=T))
+multiplot(Fig1A,Fig1B,Fig1C,layout=matrix(c(1,1,1,1,2,3),nrow=3,byrow=T))
 dev.off()
@@
 
-ys <- names(which(table(subset(dd,Ageclass == 'CH')$LayYear)>10))
-temp <- subset(dd,LayYear %in% ys)
-temp$Cohort <- factor(temp$LayYear)
-
-Fig2A <- ggplot(temp,aes(x = Agemonths,y = RTL,col = Cohort))+
+Fig2 <- ddply(dd,
+               .(Agemonths,LayYear),
+               summarise,
+               meanRTL = mean(RTL),
+               n = length(RTL))  %>%
+  ddply(.(LayYear),
+        mutate,
+        nAges = length(meanRTL),
+        minAge = min(Agemonths)) %>%
+  subset(minAge == 1) %>%
+  subset(n > 5) %>%
+  transform(Cohort = factor(LayYear)) %>%
+  ggplot(aes(x = Agemonths,y = meanRTL,col = Cohort))+
   xlab('Age (months)') +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black'))+
-  geom_smooth(method = 'lm',formula = y~log(x),se = F)
-  annotate('text',x = 0,y = 1.3,label = 'A')
-
-temp2 <- ddply(temp,
-               .(Agemonths,Cohort),
-               summarise,
-               MedianRTL = mean(RTL))  %>%
-  subset(Agemonths < 20)
-
-Fig2B <- ggplot(temp2,aes(x = Agemonths,y = MedianRTL,col = Cohort))+
-  xlab('Age (months)') +
+  geom_point()+
   geom_line()+
@@
 
-```{r Figure 3}
-
-temp <- dd14
+```{r Figure 4}
+dd14$r <- sample(1:nrow(dd14),replace = F)
+temp <- dd14[order(dd14$BirdID,dd14$r,decreasing = F),] %>%
+  subset(!duplicated(BirdID)) %>%
+  subset(Died == 1)
 
 Fig4A <- ggplot(temp,
-       aes(x = Agemonths,
+       aes(x = Sex,
            col = factor(SurvivedNext),
@@
   ylab('RTL') + 
-  xlab('Survival probability') +
+  xlab('Age (months)') +
   theme_classic() +
@@
         axis.line.y = element_line(colour = 'black'),
-        legend.position = 'none') +
-  geom_smooth(method = 'lm')
+        legend.position = 'top') +
+  geom_boxplot(notch=T)
   annotate('text',x = 17,y = 2.5,label = 'C')
@@
 ```
+
+x <- subset(dd14,Fledged == 'Fledglings')
+
+table(x$Sex,x$LayYear)
+
+dd14$SexAge <- paste(dd14$Fledged,dd14$Sex)
+
+ggplot(dd14,aes(x = Tarsus,y = RTL,col = SexAge))+
+geom_smooth(method = 'lm')

### 2016-11-08 Scripts/SW juvenile telomeres IPC clean.R

@@ -380,6 +380,13 @@ juv <- juvdata
 juv14 <- subset(juv_r,LayYear<2014)
 juv12 <- subset(juv_r,LayYear<2012)
 
-dd14 <- subset(dd,LayYear<2014)
-dd12 <- subset(dd,LayYear<2012)
+dd14 <- subset(dd,LayYear<2014) %>%
+  transform(r = sample(order(RTL))) %>%
+  arrange(BirdID,r) %>%
+  subset(!duplicated(BirdID))
+
+dd8 <- subset(dd,LayYear<2008) %>%
+  transform(r = sample(order(RTL))) %>%
+  arrange(BirdID,CatchDate) %>%
+  subset(!duplicated(BirdID))
 

### 2016-11-08 Manuscript/SW_early_life_telomeres.Rmd
Eds

@@
 Longitudinal data showed that early-life RTL was weakly, but significantly, related to adult RTL (R^2^ = `r round3(summary(lm(RTL1~RTL,data = dd3))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL,data = dd3))$coefficients[2,3])`; P = `r round3(summary(lm(RTL1~RTL,data = dd3))$coefficients[2,4])`; Fig.
-4A).
-$\Delta$RTL was significantly lower than zero in nestlings (t = `r round2(t.test(subset(dd3,Agemonths == 1)$DeltaRTL)$statistic)`; P `r round3(t.test(subset(dd3,Agemonths == 1)$DeltaRTL)$p.value)`), but not in fledglings (t = `r round2(t.test(subset(dd3,Agemonths == 6)$DeltaRTL)$statistic)`; P = `r round2(t.test(subset(dd3,Agemonths == 6)$DeltaRTL)$p.value)`) or subadults (t = `r round2(t.test(subset(dd3,Agemonths == 10)$DeltaRTL)$statistic)`; P = `r round2(t.test(subset(dd3,Agemonths == 10)$DeltaRTL)$p.value)`), suggesting that telomere shortening was highest shortly after the nestling phase, after which both telomere shortening and telomere lengthening were observed (Fig.
-4B).
+1B).
+As with cross-sectional RTL, $\Delta$RTL decreased with age (R^2^ = `r round3(summary(lm(DeltaRTL~Agemonths,data = dd3))$r.squared)`; t = `r round2(summary(lm(DeltaRTL~Agemonths,data = dd3))$coefficients[2,3])`; P = `r round3(summary(lm(DeltaRTL~Agemonths,data = dd3))$coefficients[2,4])`), suggesting that telomere shortening was highest shortly after the nestling phase, after which both telomere shortening and telomere lengthening were observed.
 
@@
 We found significantly higher variance in telomere length over individual lifetimes compared to among sample replicates (Levene's test: F = `r round2(leveneTest(DeltaRTL~Group,data=ddL)['F value'][1,1])`; P `r round3(leveneTest(DeltaRTL~Group,data=ddL)['Pr(>F)'][1,1])`; Fig.
-4C).
+1C).
 Importantly, splitting the longitudinal data into instances of shortening ($\Delta$RTL < 0) and lenghtning ($\Delta$RTL > 0) revealed that not only did we observe significantly more shortening within individuals compared to within samples (Wilcoxon test: W = `r round(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$statistic,0)`; P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$p.value)`), but also significantly more lengthening (W = `r as.character(round(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$statistic,0))`; P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$p.value)`).
@@
 
-summary(glmer(SurvivedNext1~DeltaRTL+LogAge+(1|LayYear),data=dd3,family = 'binomial'))
+summary(glmer(SurvivedNext~RTL+LogAge+Sex+Tarsus+(1|LayYear),data=subset(dd14,Agemonths>30),family = 'binomial'))
+
+summary(coxph(Surv(RemainingLife,Died)~RTL+LogAge,data=dd8))
 

### 2016-11-08 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Home eds

@@
 
-Fig2 <- ddply(dd,
+Fig2A <- transform(juv,Cohort = factor(LayYear)) %>%
+  subset(Agemonths>1) %>%
+  ggplot(aes(y = RTL,x = Cohort))+
+  xlab('Cohort') +
+  geom_boxplot() +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12,angle = 90), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'))
+  annotate('text',x = 0,y = 1.4,label = 'B')
+
+Fig2B <- ddply(dd,
                .(Agemonths,LayYear),
@@
         minAge = min(Agemonths)) %>%
-  subset(minAge == 1) %>%
-  subset(n > 5) %>%
+  subset(LayYear %in% c(2001:2009)) %>%
+  subset(n>4) %>%
   transform(Cohort = factor(LayYear)) %>%
@@
   xlab('Age (months)') +
+  ylab('Mean RTL') +
   geom_point()+
@@
         axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black'))
+        axis.line.y = element_line(colour = 'black'),
+        legend.position = 'none')
   annotate('text',x = 0,y = 1.4,label = 'B')
@@
 ```{r Figure 4}
-dd14$r <- sample(1:nrow(dd14),replace = F)
-temp <- dd14[order(dd14$BirdID,dd14$r,decreasing = F),] %>%
-  subset(!duplicated(BirdID)) %>%
-  subset(Died == 1)
-
-Fig4A <- ggplot(temp,
-       aes(x = Sex,
+Fig4A <- ggplot(dd14,
+       aes(x = Agemonths,
            col = factor(SurvivedNext),
@@
         legend.position = 'top') +
-  geom_boxplot(notch=T)
+  geom_point()+
+  geom_smooth(method = 'lm',formula = y~log(x))
   annotate('text',x = 17,y = 2.5,label = 'C')
 
-temp <- ddply(juv14,
-              .(LayYear,Fledged),
-              summarise,
-              MeanTL = median(RTL),
-              Survived = mean(SurvivedNext),
-              n = length(RTL)) %>%
-  subset(n > 9)
-
-ggplot(temp,aes(x = MeanTL,y = Survived,col = Fledged))+
+Fig4B <- ggplot(dd8,
+       aes(x = RTL,
+           y = RemainingLife,
+           col = Fledged)) +
+  ylab('RTL') + 
+  xlab('Age (months)') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+                axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'),
+        legend.position = 'top') +
   geom_point()+
-  stat_smooth(method = 'lm')
-
-Fig4B <- ??
+  geom_smooth(method = 'lm')
+  annotate('text',x = 17,y = 2.5,label = 'C')
 
@@
 ```
-
-x <- subset(dd14,Fledged == 'Fledglings')
-
-table(x$Sex,x$LayYear)
-
-dd14$SexAge <- paste(dd14$Fledged,dd14$Sex)
-
-ggplot(dd14,aes(x = Tarsus,y = RTL,col = SexAge))+
-geom_smooth(method = 'lm')

### 2016-11-08 Scripts/SW juvenile telomeres IPC clean.R

@@ -382,11 +382,11 @@ juv12 <- subset(juv_r,LayYear<2012)
 
 dd14 <- subset(dd,LayYear<2014) %>%
   transform(r = sample(order(RTL))) %>%
-  arrange(BirdID,r) %>%
+  arrange(BirdID,desc(CatchDate)) %>%
   subset(!duplicated(BirdID))
 
 dd8 <- subset(dd,LayYear<2008) %>%
   transform(r = sample(order(RTL))) %>%
-  arrange(BirdID,CatchDate) %>%
+  arrange(BirdID,desc(CatchDate)) %>%
   subset(!duplicated(BirdID))
 

### 2016-11-09 Manuscript/SW_early_life_telomeres.Rmd
eds

@@
 Efficiencies (mean $\pm$ s.d.) for our telomere and GAPDH reactions were (`r round2(mean(dd$EffTelomere))` $\pm$ `r round2(sd(dd$EffTelomere))`) and (`r round2(mean(dd$EffGAPDH))` $\pm$ `r round2(sd(dd$EffGAPDH))`) respectively.
-Inter-plate repeatability of RTL, based on `r nrow(subset(ddL,Group == 'Within samples'))` samples measured at least twice, was 0.78 (CI = 0.73-0.82).
+Inter-plate repeatability of RTL, based on `r nrow(subset(ddL,Group == 'Within samples'))` samples measured at least twice, was `r round2(rptall$R)` (CI = `r round2(rptall$CI_emp[1])`-`r round2(rptall$CI_emp[2])`)
 
 #####Telomere dynamics and age
-Using the cross-sectional data RTL decreased with age across the entire lifespan of the Seychelles warbler (LMM: t = `r round2(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,4])`; P `r round3(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,5])`).
-We found that modelling a non-linear effect age explained almost double the variation in telomere length (R^2^ of linear regression of telomere length and log age = `r round3(summary(lm(RTL~log(Agemonths),data=dd))$r.squared)` than a linear model `r round3(summary(lm(RTL~Agemonths,data=dd))$r.squared)`.
-Also worth noting is that while the overall trend was for decreasing telomere length, there were clear increases in cross-sectional telomere length after one year of age and at several points later in life (Fig.
+Using the cross-sectional data, we found that RTL decreased with age in the Seychelles warbler (LMM: estimate $\pm$ s.e.
+= `r round3(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID/LayYear))$tTable[2,1])` $\pm$ `r round3(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID/LayYear))$tTable[2,2])`; t = `r round2(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID/LayYear))$tTable[2,4])`; P `r round3(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID/LayYear))$tTable[2,5])`).
+We found that modelling a non-linear effect age explained almost double the variation in RTL (R^2^ of linear regression of telomere length and log age = `r round3(summary(lm(RTL~log(Agemonths),data=dd))$r.squared)` than a linear model `r round3(summary(lm(RTL~Agemonths,data=dd))$r.squared)`, suggesting that telomere length decreased at the greatest rate in early life.
+Accordingly, RTL decreased with age at a greater rate in juveniles (estimate $\pm$ s.e.
+= `r round3(summary(lme(RTL~Agemonths,data=juv,random=~1|BirdID/LayYear))$tTable[2,1],lessthan = F)` $\pm$ `r round3(summary(lme(RTL~Agemonths,data=juv,random=~1|BirdID/LayYear))$tTable[2,2])`) compared to adults (estimate $\pm$ s.e.
+= `r round3(summary(lme(RTL~Agemonths,data=subset(dd,Ageclass == 'A'),random=~1|BirdID/LayYear))$tTable[2,1],lessthan = F)` $\pm$ `r round3(summary(lme(RTL~Agemonths,data=subset(dd,Ageclass == 'A'),random=~1|BirdID/LayYear))$tTable[2,2])`).
+Also worth noting is that while the overall trend was for decreasing RTL, there were appoarent increases in cross-sectional telomere length after one year of age and at several points later in life (Fig.
 1A).
@@
 1B).
-As with cross-sectional RTL, $\Delta$RTL decreased with age (R^2^ = `r round3(summary(lm(DeltaRTL~Agemonths,data = dd3))$r.squared)`; t = `r round2(summary(lm(DeltaRTL~Agemonths,data = dd3))$coefficients[2,3])`; P = `r round3(summary(lm(DeltaRTL~Agemonths,data = dd3))$coefficients[2,4])`), suggesting that telomere shortening was highest shortly after the nestling phase, after which both telomere shortening and telomere lengthening were observed.
-
+As with cross-sectional RTL, $\Delta$RTL decreased with age (R^2^ = `r round3(summary(lm(DeltaRTL~LogAge,data = dd3))$r.squared)`; t = `r round2(summary(lm(DeltaRTL~LogAge,data = dd3))$coefficients[2,3])`; P = `r round3(summary(lm(DeltaRTL~LogAge,data = dd3))$coefficients[2,4])`), suggesting that telomere shortening was highest in early life.
+Although both or cross-sectional and longitudinal data indicated a general trend of telomere shortening, we found that `r round((nrow(subset(dd3,DeltaRTL>0))/nrow(dd3))*100,0)% of our $\Delta$RTL measurments were positive, and thus indicative of telomere lengthening.
 To test whether the apparent telomere lengthening in our dataset could be explained by measurement error, we compared variance in telomere length among repeat measurements of samples to the variance observed among different samples of the same individual (see methods for details).
@@
 Importantly, splitting the longitudinal data into instances of shortening ($\Delta$RTL < 0) and lenghtning ($\Delta$RTL > 0) revealed that not only did we observe significantly more shortening within individuals compared to within samples (Wilcoxon test: W = `r round(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$statistic,0)`; P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$p.value)`), but also significantly more lengthening (W = `r as.character(round(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$statistic,0))`; P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$p.value)`).
-We found no evidence that the likelihood of telomere lengthening was related to the follow up time between sampling events (GLM: t = `r round2(summary(glm(DeltaRTLF~TimeDiff,data=dd3,family = 'binomial'))$coefficients[2,3])`; P = `r round2(summary(glm(DeltaRTLF~TimeDiff,data=dd3,family = 'binomial'))$coefficients[2,4])`), and considering those samples that did show increases in $\Delta$RTL, the degree of increase was not related to follow up time (LM: t = `r round2(summary(lm(DeltaRTL~TimeDiff,data=subset(dd3,DeltaRTL>0)))$coefficients[2,3])`; P = `r round2(summary(lm(DeltaRTL~TimeDiff,data=subset(dd3,DeltaRTL>0)))$coefficients[2,4])`).
 
-#####Early life environment and telomere dynamics
-Early-life RTL varied significantly among cohorts (Likelihood ratio test: F = `r round2(anova(lme(RTL~factor(FieldPeriodID),data=juv,random=~1|BirdID))[2,3])`; P `r round3(anova(lme(RTL~Agemonths,data=juv,random=~1|BirdID))[2,4])`).
-Cohort-level variation in juvenile RTL was not related to adult population density (LMM controlling for age and cohort: t = `r round2(summary(lme(RTL~Agemonths+Density,data=subset(juv,!is.na(Density)),random=~1|FieldPeriodID))$tTable[3,4])`; P = `r round3(summary(lme(RTL~Agemonths+Density,data=subset(dd,!is.na(Density)),random=~1|BirdID))$tTable[3,5])`), or to temporal variation in  food availability (t = `r round2(summary(lme(RTL~Agemonths+Insect,data=dd,random=~1|BirdID/FieldPeriodID,na.action=na.exclude))$tTable[3,4])`; P = `r round3(summary(lme(RTL~Agemonths+Insect,data=juv,random=~1|BirdID,na.action=na.exclude))$tTable[3,5])`).
+#####Spatiotemporal variation in telomere dynamics and the environment
+There was significant variation in RTL among cohorts (), although we found no evidence that the relationshio between telomere length and age varied among cohorts.
+Rather, cohort-level differences in RTL were consistent over the lifespan of Seychelles warblers.
+
+A linear mixed model of RTL in relation to social and ecological environmental variables revealed age, tarsus length, sex and annual variation in food availability were significantly related to RTL (Table 1).
+Model selection revealed that sex only appeared in models where tarsus length was also present (Table S1), and sex was no longer significant when tarsus length was removed from the full model (P = ).
+RTL increased with tarsus length in males, but decreased in females.
+In general, the model averaging approach yielded qualitatively identical results to the full LMM, with the same explanatory variables 'significant' in terms of being retained in top models, and having model-average confidence intervals not overlapping zero (Table S1; Fig.
+S1).
 
@@
 Model selection revealed that the top model explaining $\Delta$RTL was the null model (Table S2).
-
-Within cohorts, the top model explaining variation in early-life RTL contained age, territory quality, tarsus length, and the interaction between age and tarsus length (Table S1).
-Model-averaged confidence intervals of the age and tarsus x age effects did not overlap zero, but the tarsus and territory quality effects were not significant (Fig.
-3A).
-Tarsus length was negatively related to RTL in nestlings, unrelated to RTL in fledglings, and positively related to RTL in subadults (Fig.
-3B).
-Territory quality was negatively related to RTL, particularly in nestlings, although this effect was weak (Fig.
-3C).
-
-#####Telomere dynamics and survival
-Early-life RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~RTL+LogAge+(1|PlateID)+(1|TerritoryID)+(1|LayYear),data=dd14,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL+LogAge+(1|TerritoryID)+(1|LayYear),data=dd14,family = 'binomial'))$coefficients[2,4])`), with no interaction effect between RTL and  age (estimate = `r round2(summary(glmer(SurvivedNext~RTL*LogAge+(1|TerritoryID)+(1|LayYear),data=dd14,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL*LogAge+(1|TerritoryID)+(1|LayYear),data=dd14,family = 'binomial'))$coefficients[4,4])`), suggesting that there were no age-specific effects of juvenile RTL on survival to adulthood.
-
-summary(glmer(SurvivedNext~RTL+LogAge+Sex+Tarsus+(1|LayYear),data=subset(dd14,Agemonths>30),family = 'binomial'))
-
-summary(coxph(Surv(RemainingLife,Died)~RTL+LogAge,data=dd8))
 

### 2016-11-09 Manuscript/SW_early_life_telomeres_SOM.Rmd
eds

--- 2016-06-14 Manuscript/SW_early_life_telomeres_SOM.Rmd
@@
 #Run the model
-lm1 <- lmer(RTL ~ Agemonths * (Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ) + (1|FieldPeriodID)+(1|TerritoryID)+(1|PlateID),
-            data = juv,
+temp <- subset(dd,!is.na(Insect))
+
+#Run the model
+lm2 <- lmer(RTL ~  LogAge + Tarsus + Helper + GroupSize + Sex + BodyMass + Insect + Density + (1|LayYear) + (1|BirdID) + (1|PlateID),
+            data = temp,
             REML = FALSE)
-lm1 <- standardize(lm1)
+lm2 <- standardize(lm2)
 
 #Model averaging
-lm2 <- dredge(lm1)
-t1 <- as.matrix(subset(lm2, delta<=6))
+lm3 <- dredge(lm2)
+t1 <- as.matrix(subset(lm3, delta<=6))
 
@@
 
-nams[nams == 'z.NonHelper'] <- 'Non helpers'
-nams[nams == 'c.Sex'] <- 'Sex'
-nams[nams == 'z.TQ'] <- 'TQ'
+nams[nams == 'z.GroupSize'] <- 'Group Size'
+nams[nams == 'c.Sex'] <- 'Sex (males)'
 nams[nams == 'c.Helper'] <- 'Helpers'
 nams[nams == 'z.BodyMass'] <- 'Body mass'
-nams[nams == 'z.Agemonths'] <- 'Age'
+nams[nams == 'z.LogAge'] <- 'Log age'
 nams[nams == 'z.Tarsus'] <- 'Tarsus'
-nams[nams == 'c.Season'] <- 'Season'
-nams[nams == 'c.Season:z.Agemonths'] <- 'Age x season'
-nams[nams == 'z.Agemonths:z.NonHelper'] <- 'Age x non helpers'
-nams[nams == 'c.Sex:z.Agemonths'] <- 'Age x sex'
-nams[nams == 'z.Agemonths:z.TQ'] <- 'Age x TQ'
-nams[nams == 'c.Helper:z.Agemonths'] <- 'Age x helpers'
-nams[nams == 'z.Agemonths:z.BodyMass'] <- 'Age x mass'
-nams[nams == 'z.Agemonths:z.Tarsus'] <- 'Age x tarsus'
+nams[nams == 'z.Insect'] <- 'Insect abundance'
+nams[nams == 'z.CatchMonth'] <- 'Month'
+nams[nams == 'z.Density'] <- 'Density'
 
@@
 ```{r Table S2}
-temp <- dd3
+#Run the model
+temp <- subset(dd3,!is.na(Insect))
 
 #Run the model
-lm1 <- glmer(DeltaRTLF ~ Agemonths+(Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ) + (1|FieldPeriodID) + (1|TerritoryID),
+lm2 <- lmer(DeltaRTL ~  LogAge + Tarsus + Helper + GroupSize + Sex + BodyMass + Insect + Density + (1|LayYear) + (1|PlateID),
             data = temp,
-            family = 'binomial',
             REML = FALSE)
-lm1 <- standardize(lm1)
+lm2 <- standardize(lm2)
 
 #Model averaging
-lm2 <- dredge(lm1)
-t1 <- as.matrix(subset(lm2, delta<=6))
+lm3 <- dredge(lm2)
+t1 <- as.matrix(subset(lm3, delta<=6))
 
@@
 ```{r Figure S1}
+temp <- subset(dd,!is.na(Insect))
+temp$LogAge <- log(temp$Agemonths)
 
-temp <- Loss
-temp[,'Age (months)'] <- factor(temp$Agemonths)
-temp$LayYear <- factor(temp$LayYear)
-temp$TimeDiff <- temp$TimeDiff/365
+#Run the model
+lm2 <- lmer(RTL ~  LogAge + Tarsus + Helper + GroupSize + Sex + BodyMass + Insect + Density + (1|LayYear) + (1|BirdID) + (1|PlateID),
+            data = temp,
+            REML = FALSE)
+lm2 <- standardize(lm2)
 
-FigS1 <- ggplot(temp,
-       aes(x = TimeDiff,
-           y = Loss,
-           colour=`Age (months)`,
-           fill = `Age (months)`)) +
-  ylab(expression(Delta*'RTL')) + 
-  xlab('Time (years)') +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        legend.position = 'top') +
-  geom_point()+
-  stat_smooth(method = 'lm')+
-  scale_colour_manual(values = c('grey','darkorange','darkgreen'))+
-  scale_fill_manual(values = c('grey','darkorange','darkgreen'))+
-    geom_abline(slope = 0,intercept = 0,lty = 2)+
-  annotate('text',x = 0,y = 1.75,label = 'B')
+#Model averaging
+lm3 <- dredge(lm2)
+lm3 <- subset(lm3,delta <=6)
+lmavg <- model.avg(lm3,fit=T)
+myimp <- importance(lmavg) #Relative importance of each explanatory variable
+
+#Put it all into a nice table
+table2 <- imptable(lmavg,myimp)
+table2 <- signif(table2,2)
+
+#Sort out row/col names
+nams <- rownames(table2)
+nams[nams == 'z.GroupSize'] <- 'Group Size'
+nams[nams == 'c.Sex'] <- 'Sex (males)'
+nams[nams == 'c.Helper'] <- 'Helpers'
+nams[nams == 'z.BodyMass'] <- 'Body mass'
+nams[nams == 'z.LogAge'] <- 'Log age'
+nams[nams == 'z.Tarsus'] <- 'Tarsus'
+nams[nams == 'z.Insect'] <- 'Insect abundance'
+nams[nams == 'z.CatchMonth'] <- 'Month'
+nams[nams == 'z.Density'] <- 'Density'
+
+rownames(table2) <- nams
+colnames(table2) <- c('Estimate','SE','2.5%','97.5%','RI')
+
+table2 <- data.frame(table2,myx=row.names(table2))
+table2$myx <- factor(table2$myx,levels = table2$myx[order(table2$Estimate,decreasing = T)])
+
+FigS1 <- forestplot(table2)+
+  annotate('text',x = 0.6,y = 0.05,label = 'A')+
+  theme(axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'))
 
@@
 ```
+
+```{r Figure S2}
+
+#Run the model
+temp <- subset(dd3,!is.na(Insect))
+
+#Run the model
+lm2 <- lmer(DeltaRTL ~  LogAge + Tarsus + Helper + GroupSize + Sex + BodyMass + Insect + Density + (1|LayYear) + (1|PlateID),
+            data = temp,
+            REML = FALSE)
+lm2 <- standardize(lm2)
+
+#Model averaging
+lm3 <- dredge(lm2)
+lm3 <- subset(lm3,delta <=6)
+lmavg <- model.avg(lm3,fit=T)
+myimp <- importance(lmavg) #Relative importance of each explanatory variable
+
+#Put it all into a nice table
+table2 <- imptable(lmavg,myimp)
+table2 <- signif(table2,2)
+
+#Sort out row/col names
+nams <- rownames(table2)
+nams[nams == 'z.GroupSize'] <- 'Group Size'
+nams[nams == 'c.Sex'] <- 'Sex (males)'
+nams[nams == 'c.Helper'] <- 'Helpers'
+nams[nams == 'z.BodyMass'] <- 'Body mass'
+nams[nams == 'z.LogAge'] <- 'Log age'
+nams[nams == 'z.Tarsus'] <- 'Tarsus'
+nams[nams == 'z.Insect'] <- 'Insect abundance'
+nams[nams == 'z.CatchMonth'] <- 'Month'
+nams[nams == 'z.Density'] <- 'Density'
+
+rownames(table2) <- nams
+colnames(table2) <- c('Estimate','SE','2.5%','97.5%','RI')
+
+table2 <- data.frame(table2,myx=row.names(table2))
+table2$myx <- factor(table2$myx,levels = table2$myx[order(table2$Estimate,decreasing = T)])
+
+FigS2 <- forestplot(table2)+
+  annotate('text',x = 0.6,y = 0.05,label = 'A')+
+  theme(axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'))
+
+FigS2
+
+```

### 2016-11-09 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
eds

--- 2016-11-08 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
@@
 source('Scripts/SW juvenile telomeres IPC clean.r')
+
+```
+
+**Table1**
+```{r Table 1}
+temp <- subset(dd,!is.na(Insect))
+
+#Run the model
+lm2 <- lmer(RTL ~  LogAge + Tarsus + Helper + GroupSize + Sex + BodyMass + Insect + Density + (1|LayYear) + (1|BirdID) + (1|PlateID),
+            data = temp,
+            REML = FALSE)
+CIs <- confint.merMod(lm2)
+t1 <- data.frame(Estimate = fixef(lm2), s.e. = se.fixef(lm2), LCI = CIs[5:nrow(CIs),1], UCI = CIs[5:nrow(CIs),2])
+kable(t1)
 
@@
   subset(!duplicated(paste0(BirdID,Agemonths))) %>%
-  transform(Agemonthsdec = Agemonths/12) %>%
-  ggplot(aes(x = Agemonthsdec,y = RTL,group = BirdID,col=factor(BirdID)))+
+  ggplot(aes(x = Agemonths,y = RTL,group = BirdID,col=factor(BirdID)))+
   geom_point(alpha = 1,col='grey')+
@@
 
-Fig2 <- ddply(dd,
+Fig2A <- transform(juv,Cohort = factor(LayYear)) %>%
+  subset(Agemonths>(1/12)) %>%
+  ggplot(aes(y = RTL,x = Cohort))+
+  xlab('Cohort') +
+  geom_boxplot() +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12,angle = 90), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'))
+  annotate('text',x = 0,y = 1.4,label = 'B')
+
+Fig2B <- ddply(dd,
                .(Agemonths,LayYear),
@@
         minAge = min(Agemonths)) %>%
-  subset(minAge == 1) %>%
-  subset(n > 5) %>%
+  subset(LayYear %in% c(2001:2009)) %>%
+  subset(n>4) %>%
   transform(Cohort = factor(LayYear)) %>%
   ggplot(aes(x = Agemonths,y = meanRTL,col = Cohort))+
-  xlab('Age (months)') +
+  xlab('Age (years)') +
+  ylab('Mean RTL') +
   geom_point()+
@@
 
-temp <- subset(dd,!is.na(Insect))
-temp$LogAge <- log(temp$Agemonths)
-
-#Run the model
-lm2 <- lmer(RTL ~  LogAge + Tarsus + Helper + GroupSize + Sex + BodyMass + Insect + Density + (1|LayYear) + (1|BirdID) + (1|PlateID),
-            data = temp,
-            REML = FALSE)
-lm2 <- standardize(lm2)
-
-#Model averaging
-lm3 <- dredge(lm2)
-lm3 <- subset(lm3,delta <=6)
-lmavg <- model.avg(lm3,fit=T)
-myimp <- importance(lmavg) #Relative importance of each explanatory variable
-
-#Put it all into a nice table
-table2 <- imptable(lmavg,myimp)
-table2 <- signif(table2,2)
-
-#Sort out row/col names
-nams <- rownames(table2)
-nams[nams == 'z.GroupSize'] <- 'Group Size'
-nams[nams == 'c.Sex'] <- 'Sex (males)'
-nams[nams == 'c.Helper'] <- 'Helpers'
-nams[nams == 'z.BodyMass'] <- 'Body mass'
-nams[nams == 'z.LogAge'] <- 'Log age'
-nams[nams == 'z.Tarsus'] <- 'Tarsus'
-nams[nams == 'z.Insect'] <- 'Insect abundance'
-nams[nams == 'z.CatchMonth'] <- 'Month'
-nams[nams == 'z.Density'] <- 'Density'
-
-rownames(table2) <- nams
-colnames(table2) <- c('Estimate','SE','2.5%','97.5%','RI')
-
-table2 <- data.frame(table2,myx=row.names(table2))
-table2$myx <- factor(table2$myx,levels = table2$myx[order(table2$Estimate,decreasing = T)])
-
-Fig3A <- forestplot(table2)+
-  annotate('text',x = 0.6,y = 0.05,label = 'A')+
-  theme(axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black'))
-
-Fig3B <- ggplot(temp,
+Fig3A <- ggplot(temp,
        aes(x = Tarsus,
@@
 
-Fig3C <- ggplot(temp,
+Fig3B <- ggplot(temp,
        aes(
@@
 pdf('Figures/Figure 3.pdf',
-    width = 8,
+    width = 5,
     height = 8,
     paper = 'a4')
-multiplot(Fig3A,Fig3B,Fig3C,layout = matrix(c(1,1,2,3),nrow = 2,byrow = T))
+multiplot(Fig3A,Fig3B)
 dev.off()
@@
 ```
-
-```{r Figure 4}
-dd14$r <- sample(1:nrow(dd14),replace = F)
-temp <- dd14[order(dd14$BirdID,dd14$r,decreasing = F),] %>%
-  subset(!duplicated(BirdID)) %>%
-  subset(Died == 1)
-
-Fig4A <- ggplot(temp,
-       aes(x = Sex,
-           col = factor(SurvivedNext),
-           y = RTL)) +
-  ylab('RTL') + 
-  xlab('Age (months)') +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-                axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black'),
-        legend.position = 'top') +
-  geom_boxplot(notch=T)
-  annotate('text',x = 17,y = 2.5,label = 'C')
-
-temp <- ddply(juv14,
-              .(LayYear,Fledged),
-              summarise,
-              MeanTL = median(RTL),
-              Survived = mean(SurvivedNext),
-              n = length(RTL)) %>%
-  subset(n > 9)
-
-ggplot(temp,aes(x = MeanTL,y = Survived,col = Fledged))+
-  geom_point()+
-  stat_smooth(method = 'lm')
-
-Fig4B <- ??
-
-pdf('Figures/Figure 4.pdf',
-    width = 8,
-    height = 8,
-    paper = 'a4')
-multiplot(Fig4A,Fig4B,layout = matrix(c(1,2),nrow = 2,byrow = T))
-dev.off()
-
-```
-
-x <- subset(dd14,Fledged == 'Fledglings')
-
-table(x$Sex,x$LayYear)
-
-dd14$SexAge <- paste(dd14$Fledged,dd14$Sex)
-
-ggplot(dd14,aes(x = Tarsus,y = RTL,col = SexAge))+
-geom_smooth(method = 'lm')

### 2016-11-09 Scripts/SW juvenile telomeres IPC clean.R

@@ -13,18 +13,17 @@ dd0 <- subset(dd0,CqTelomere < 25) %>%
   subset(DiffGAPDH < 0.5) %>%
   subset(DiffTelomere < 0.5) %>%
   subset(RTL < 3) %>%
-  subset(BloodID>7) %>%
-  subset(!duplicated(paste0(CqTelomere,CqGAPDH)))
-
+  subset(BloodID>0)
 
 dd0$RTL <- sqrt(dd0$RTL)
+
 
 
 
 #Average repeats of blood samples
 dd0$RTL2 <- ave(dd0$RTL,c(dd0$BloodID,dd0$Status,dd0$PlateID))
 
-# rpt(RTL~(1|BloodID),grname = 'BloodID',data = dd0,nboot = 50)
+rptall <- rpt(RTL~(1|BloodID),grname = 'BloodID',data = dd0,nboot = 50)
 # rpt(RTL~(1|BloodID),grname = 'BloodID',data = subset(dd0,Whodunnit == "EB"),nboot = 50)
 # rpt(RTL~(1|BloodID),grname = 'BloodID',data = subset(dd0,Whodunnit == "EAF"),nboot = 50)
 # 
@@ -76,7 +75,7 @@ dd$Fledged <- factor(dd$Fledged,levels = c('Nestlings','Fledglings','Adults'))
 
 dd$Agemonths <- ifelse(dd$Ageclass == 'CH',1,
                        ifelse(dd$Ageclass == 'FL',6,
-                              ifelse(dd$Ageclass == 'SA',10,dd$Age*12+6)))
+                              ifelse(dd$Ageclass == 'SA',10,dd$Age*12+6)))/12
 
 
   #dd$Age[dd$Fledged != 'Adults'] <- 1

### 2016-11-09 Scripts/SW juvenile telomeres IPC load.r

@@ -10,16 +10,16 @@ rm(list=ls())
 
 # Load relevant libraries -------------------------------------------------
 library(MuMIn)
-library(arm)
 library(plyr)
 library(ggplot2)
-library(gvlma)
 library(survival)
 library(nlme)
 library(magrittr)
-library(car)
 library(rptR)
 library(Rmisc)
+library(sjPlot)
+library(lme4)
+library(arm)
 
 # Load data ---------------------------------------------------------------
 dd0 <- read.csv('Data/REL_TL_EB_EAF_24 May aft TL queries.csv') #main dataset

### 2016-11-10 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Home eds

@@
 
-**Figure 1** RTL in relation to **(A)** age and **(B)** $\Delta$Age in Seychelles warblers. Lines and shaded areas are fitted values and 95% confidence limits from linear regressions.
+**Figure 1** Telomere length in relation to age in Seychelles warblers.
+Points and  thin grey lines represent individual samples and birds, respectively.
+The thick line and shaded area represent the fitted values and 95% confidence limits of a linea regression of RTL and log-transformed age.
 
-**Figure 2** Cohort effects on early-life RTL in Seychelles warblers.
-**A** temporal variation RTL in major and minor cohorts, **B** temporal variation in adult population size, and **C** early-life RTL in relation to cohort-level variation in adult population size.
-Lines and shaded areas in **C** are fitted values and 95% confidence limits from general linear models.
+**Figure 2** Longitudinal telomere dynamics in the Seychelles warbler.
+**A** Variation in RTL within individuals sampled at different time points.
+The dotted line represents parity, and thus points above and below the line represent increases and decreases in RTL, respectively.
+**B** Scaled density plots of repeated RTL measurements among individual samples, and among different samples taken from the same individual.
 
-**Figure 3** Factors affecting early-life RTL in Seychelles warblers.
-**A** Model averaged estimates and 95% confidence intervals for explanatory terms used in a linear mixed model with RTL as the response variable (see main text for details; TQ = territory quality).
-**B,C** Early-life RTL and age in relation to tarsus length (**B**) and sex (**C**).
-Lines and shaded areas in **B** are fitted values and 95% confidence limits from a general linear model.
+**Figure 3** Cohort variaiton in telomere length in the Seychelles warbler.
+**A** Bocplot of RTL among all Seychelles warbler cohorts.
+**B** Relationship between RTL and age in individual cohorts.
+Only cohorts with samples of at least 10 nestlings were included.
+Lines represent represents the fitted values from a linear regression of RTL and log-transformed age.
 
-**Figure 4** Longitudinal telomere dynamics in juvenile Seychelles warblers.
-**A** juvenile telomere length in relation to adult telomere length measured in the same individual.
-**B** $\Delta$RTL in relation to the time between sampling events.
-Lines and shaded areas represent fitted values and 95% confidence limits from general linear models.
-Raw data have been ommited from **B** for the sake of clarity, but are plotted in Figure S1.
-
-**Figure 5** Factors affecting $\Delta$RTL in juvenile Seychelles warblers.
-**A** Model averaged estimates and 95% confidence intervals for explanatory terms used in a linear mixed model with $\Delta$RTL as the response variable (see main text for details; TQ = territory quality).
-**B** $\Delta$RTL in relation to tarsus length and sex.
+**Figure 4** Individual variation in telomere length in Seychelles warbler.
+**A** RTL in relation to tarsus length and sex.
+**B** RTL in relation to variation in annual food availability.
+Lines and shaded areas represent the fitted values and 95% confidence limits from linear regressions.
 
@@
 
-Fig1A <- subset(dd,Age<15) %>%
-  ddply(.(BirdID),
+Fig1 <- ddply(dd,
+               .(BirdID),
         mutate,
@@
 
-Fig1B <- ggplot(dd3,
+pdf('Figures/Figure 1.pdf',
+    width = 4,
+    height = 4,
+    paper = 'a4')
+Fig1
+dev.off()
+
+```
+
+```{r Figure 2}
+
+Fig2A <- ggplot(dd3,
        aes(x = RTL,
@@
 
-Fig1C <- ggplot(ddL,aes(col=Group,fill = Group,y = ..scaled..,x = DeltaRTL))+
+Fig2B <- ggplot(ddL,aes(col=Group,fill = Group,y = ..scaled..,x = DeltaRTL))+
     geom_density(alpha = 0.2)+
@@
 
-pdf('Figures/Figure 1.pdf',
-    width = 8,
+pdf('Figures/Figure 2.pdf',
+    width = 5,
     height = 8,
     paper = 'a4')
-multiplot(Fig1A,Fig1B,Fig1C,layout=matrix(c(1,1,1,1,2,3),nrow=3,byrow=T))
+multiplot(Fig2A,Fig2B)
 dev.off()
@@
 
-```{r Figure 2}
+```{r Figure 3}
 
-Fig2A <- transform(juv,Cohort = factor(LayYear)) %>%
+Fig3A <- transform(juv,Cohort = factor(LayYear)) %>%
   subset(Agemonths>(1/12)) %>%
@@
 
-Fig2B <- ddply(dd,
-               .(Agemonths,LayYear),
-               summarise,
-               meanRTL = mean(RTL),
-               n = length(RTL))  %>%
-  ddply(.(LayYear),
+Fig3B <- transform(dd,Cohort = factor(LayYear)) %>%
+  ddply(.(Cohort),
         mutate,
-        nAges = length(meanRTL),
-        minAge = min(Agemonths)) %>%
-  subset(LayYear %in% c(2001:2009)) %>%
-  subset(n>4) %>%
-  transform(Cohort = factor(LayYear)) %>%
-  ggplot(aes(x = Agemonths,y = meanRTL,col = Cohort))+
+        nc = length(which(Agemonths ==1/12))) %>%
+  subset(nc > 11) %>%
+  ggplot(aes(y = RTL,x = Agemonths, col = Cohort))+
   xlab('Age (years)') +
-  ylab('Mean RTL') +
-  geom_point()+
-  geom_line()+
+  geom_smooth(method = 'lm', formula = y~log(x),se = F) +
   theme_classic() +
@@
         axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black'),
-        legend.position = 'none')
+        axis.line.y = element_line(colour = 'black'))
   annotate('text',x = 0,y = 1.4,label = 'B')
 
-pdf('Figures/Figure 2.pdf',
+pdf('Figures/Figure 3.pdf',
     width = 5,
@@
     paper = 'a4')
-multiplot(Fig2A,Fig2B)
+multiplot(Fig3A,Fig3B)
 dev.off()
@@
 
-```{r Figure 3}
+```{r Figure 4}
 
-Fig3A <- ggplot(temp,
+Fig4A <- ggplot(temp,
        aes(x = Tarsus,
@@
 
-Fig3B <- ggplot(temp,
+Fig4B <- ggplot(temp,
        aes(
@@
 
-pdf('Figures/Figure 3.pdf',
-    width = 5,
+pdf('Figures/Figure 4.pdf',
+    width = 4,
     height = 8,
     paper = 'a4')
-multiplot(Fig3A,Fig3B)
+multiplot(Fig4A,Fig4B)
 dev.off()

### 2016-11-11 Manuscript/SW_early_life_telomeres.Rmd
Home edits

@@
 4.
-Telomere length and rates of shortening varied markedly both within and among cohorts.
-Telomere length  was positively associated with island-wide temporal variation in food availability.
-We also found that cohort and age effects led to spurious associations between telomere dynamics and survival.
+Telomere length and rates of shortening varied markedly both within and among cohorts, and cohort-level differences in telomere length persisted throughout individual lifespan.
+Temporal variation in telomere length was positively associated with island-wide temporal variation in food availability.
+The other significant predictors of telomere length were sex, with males having longer telomeres than females, and tarus length, which affected telomere length in a sex-specific manner.
 
 5.
-Our comprehensive, long-term data show that in natural populations telomere dynamics vary enormously over both space and time, and that this variation cannot be attributed entirely to measurment error.
-Ascertaining what explains this variation will require combining long-term ecological and quantitative genetic data.
+Our comprehensive, long-term data show that in natural populations, telomere dynamics vary enormously over both space and time.
+Ascertaining what explains this variation requires combining large sample sizes spanning multiple cohorts, and long-term ecological data.
 
@@
 Facultative cooperative breeding occurs in the Seychelles warbler [@Komdeur1994a; @Richardson2003a], and the presence of helping subordinates (but not non-helping subordinates) in the natal territory is associated with increased survival later in life [@Brouwer2012].
-Lastly, we have an established protocol for assessing telomere length in this species [@Barrett2012], and telomere length predicts survival independently of age in adult Seychelles warblers, suggesting that telomeres act as a biomarker of cost [@Barrett2013a].
+Lastly, we have an established protocol for assessing telomere length in this species [@Barrett2012; @Bebbington2016], and telomere length predicts survival independently of age in adult Seychelles warblers, suggesting that telomeres act as a biomarker of cost in this species [@Barrett2013a].
 Thus, we have an excellent system in which to assess the costs of different social and environmental conditions experienced early in life, and to assess the later-life consequences of early-life conditions.
 
-In this study we examine how telomeres are related to early-life environmental variation and short and long-term survival in the Seychelles warbler.
-Because telomere dynamics are expected to reflect individual-level variation in the costs of early-life experiences, they may allow a more sensitive analysis of the effects of early-life environmental variation than would be possible with a direct comparison of how survival is affected by the early-life environment.
-With this in mind, we first conduct an exploratory analysis of how the environmental and social factors experienced in early life affect telomere dynamics.
-We then test the hypothesis that longer telomeres and lower rates of telomere shortening in early life are associated with greater survival.
+In this study we examine how telomeres are related to early-life environmental variation over the entire lifespans across multiple cohorts of the Seychelles warbler.
 
@@
 All DNA extractions that passed the above criteria were diluted to 3.3 ng $\mu$l^-1^ before telomere measurement.
-We measured relative telomere length (RTL) for all samples using a quantitative PCR (qPCR) assay of telomeres and a GAPDH control gene, using the molecular methods outlined by Bebbington *et al.* (In Press).
+We measured relative telomere length (RTL) for all samples using a quantitative PCR (qPCR) assay of telomeres and a GAPDH control gene, using the molecular methods outlined by Bebbington *et al.* [-@Bebbington2016].
 
@@
 We performed all analyses using R version 3.0.1 [@RDevelopmentCoreTeam2011].
-Our dataset differs from slightly from that of Barrett *et al.* [-@Barrett2013a] in that we are using RTL calculated directly from the qPCR data rather than converting into an absolute measure, and we are including many additional samples.
-We therefore first re-tested, using all available samples, whether telomere length decreases with age and $\Delta$age [a longitudinal measure based on within-subject centring; @VandePol2009] in the Seychelles warbler, using linear mixed models with individual ID as a random effect.
+We tested the relationship between telomere length age and $\Delta$age [a longitudinal measure based on within-subject centring; @VandePol2009] in the Seychelles warbler, using linear mixed models with individual ID as a random effect.
 For the analysis of $\Delta$age we we only included birds who had been sampled at least twice.

### 2016-11-11 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Home edits

@@
 lm2 <- lmer(RTL ~  LogAge + Tarsus + Helper + GroupSize + Sex + BodyMass + Insect + Density + (1|LayYear) + (1|BirdID) + (1|PlateID),
+            data = temp,
+            REML = FALSE)
+CIs <- confint.merMod(lm2)
+t1 <- data.frame(Estimate = fixef(lm2), s.e. = se.fixef(lm2), LCI = CIs[5:nrow(CIs),1], UCI = CIs[5:nrow(CIs),2])
+kable(t1)
+
+```
+
+**Table 2**
+
+```{r Table 2}
+temp <- subset(dd3,!is.na(Insect))
+
+#Run the model
+lm2 <- lmer(DeltaRTL ~  LogAge + Tarsus + Helper + GroupSize + Sex + BodyMass + Insect + Density + (1|LayYear) + (1|BirdID) + (1|PlateID),
             data = temp,

### 2016-11-11 Manuscript/SW_early_life_telomeres.Rmd
Home eds

@@
 #####Telomere dynamics and age
-Using the cross-sectional data, we found that RTL decreased with age in the Seychelles warbler (LMM: estimate $\pm$ s.e.
-= `r round3(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID/LayYear))$tTable[2,1])` $\pm$ `r round3(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID/LayYear))$tTable[2,2])`; t = `r round2(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID/LayYear))$tTable[2,4])`; P `r round3(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID/LayYear))$tTable[2,5])`).
-We found that modelling a non-linear effect age explained almost double the variation in RTL (R^2^ of linear regression of telomere length and log age = `r round3(summary(lm(RTL~log(Agemonths),data=dd))$r.squared)` than a linear model `r round3(summary(lm(RTL~Agemonths,data=dd))$r.squared)`, suggesting that telomere length decreased at the greatest rate in early life.
+Using the cross-sectional data, we found that RTL decreased with age in the Seychelles warbler (LMM with individual ID as random effect: estimate $\pm$ s.e.
+= `r round3(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,1])` $\pm$ `r round3(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,2])`; t = `r round2(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,4])`; P `r round3(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,5])`).
+We found that modelling a non-linear effect age explained almost double the variation in RTL (R^2^ of linear regression of telomere length and log age = `r round3(summary(lm(RTL~log(Agemonths),data=dd))$r.squared)` than fitting a linear relationship `r round3(summary(lm(RTL~Agemonths,data=dd))$r.squared)`, suggesting that telomere length decreased at the greatest rate in early life.
 Accordingly, RTL decreased with age at a greater rate in juveniles (estimate $\pm$ s.e.
@@
 Also worth noting is that while the overall trend was for decreasing RTL, there were appoarent increases in cross-sectional telomere length after one year of age and at several points later in life (Fig.
-1A).
+1).
+
+Including birth year as a random effect in a LMM of RTL and (log) age resulted in a significantly improved model to one with only individual ID as a random effect (), and was a significantly better model than one including catch year as a random effect.
+This is indicative of cohort effects on telomere length across the Seychelles warbler lifespan.
+Furthermore, we found no evidence that the relationship between telomere length and age varied among cohorts ().
+Rather, cohort-level differences in RTL were consistent over the lifespan of Seychelles warblers.
 
@@
 #####Spatiotemporal variation in telomere dynamics and the environment
-There was significant variation in RTL among cohorts (), although we found no evidence that the relationshio between telomere length and age varied among cohorts.
-Rather, cohort-level differences in RTL were consistent over the lifespan of Seychelles warblers.
-
 A linear mixed model of RTL in relation to social and ecological environmental variables revealed age, tarsus length, sex and annual variation in food availability were significantly related to RTL (Table 1).

### 2016-11-13 Manuscript/SW_early_life_telomeres.Rmd
Home eds

@@
 source('Scripts/SW juvenile telomeres IPC clean.r')
+source('Scripts/SW juvenile telomeres IPC models.r')
 
@@
 We measured relative telomere length (RTL) for all samples using a quantitative PCR (qPCR) assay of telomeres and a GAPDH control gene, using the molecular methods outlined by Bebbington *et al.* [-@Bebbington2016].
+We assessed repeatability of RTL using the rptR package in R [@RDevelopmentCoreTeam2011].
 
-For a subset of first-year birds we had longitudinal data, with an additional sample taken as an adult.
-For these individuals we calculated the within-individual change in RTL by subtracting early-life RTL from adult RTL (hereafter $\Delta$RTL).
+For a subset of birds we had longitudinal data, with an additional samples taken at different ages.
+For these individuals we calculated the within-individual change in RTL by subtracting RTL at time point *t* from RTL at timepoint *t* + 1 (hereafter $\Delta$RTL).
 Negative values of $\Delta$RTL reflect telomere shortening, while positive values reflect telomere lengthening.
@@
 #####Statistical analyses
-We performed all analyses using R version 3.0.1 [@RDevelopmentCoreTeam2011].
-We tested the relationship between telomere length age and $\Delta$age [a longitudinal measure based on within-subject centring; @VandePol2009] in the Seychelles warbler, using linear mixed models with individual ID as a random effect.
-For the analysis of $\Delta$age we we only included birds who had been sampled at least twice.
-We then, using the juvenile samples, explored how telomere length varied within the first year of life (telomere length vs age in months) and among cohorts (telomere length vs cohort ID), again using linear mixed models with individual ID as a random effect.
-We tested how temporal variation in telomere length varied with temporal variation in population density (measured as the number of adult birds on the island at the end of each breeding season) and food availability, using linear mixed models with cohort ID as a random effect.
+We performed all statistical analyses using R.
+We first tested the relationship between RTL and age using linear mixed models (LMMs) with individual ID as a random effect, carried out in the lme4 package [@Bates2014].
+To test whether the decrease in telomere length with age is greatest in early life, we compared LMMs with age (in months) and log-transformed age, using a likelihood ratio test.
+We also tested whether telomores shorten with age in individuals with samples from multiple timepoints, using a LMM of RTL and $\Delta$age [a longitudinal measure based on within-subject centring; @VandePol2009].
+To test how telomere length, and its relationship with age, varied among cohorts, we compared the above LMMs with models that included birth year as an additional random effect term.
+We also compared a model in which only the intercept (i.e.
+RTL) was allowed to vary among among cohorts, to one in which the slope of the relationship between RTL and age was also allowed to vary.
 
 Because we observed apparent telomere lengthening in our data, we used the entire Seychelles warbler dataset to test whether this lengthening could be explained by measurement error.
-We used all birds (adults and juveniles) with at least two telomere measurements (N = `r nrow(subset(ddL,Group == 'Among samples'))` samples from `r length(unique(subset(ddL,Group == 'Among samples')$BirdID))` birds).
-We then calculated $\Delta$RTL for each pair of samples (for example, for a bird with three longitudinal measurements throughout their life we calculated the change in RTL between the first and second sample and between the second and third sample).
+Using all birds (adults and juveniles) with at least two telomere measurement, we calculated $\Delta$RTL for each pair of samples (for example, for a bird with three longitudinal measurements throughout their life we calculated the change in RTL between the first and second sample and between the second and third sample).
 We then compared these longitudinal changes in telomere length to sampling error, using all samples where we had at least two repeat telomere measurements for the same sample (i.e.
@@
 We calculated $\Delta$RTL between pairs of repeat measurements within samples in exactly the same way as for across samples, except that repeat measurements were ordered by the date at the qPCR was run rather than by age.
-Secondly, if telomere lengthening is an artifact of measurement error, it is likely to occur more frequently when there is a short follow up time between samples [@Steenstrup2013].
-We therefore tested whether telomere lengthening was more likely to occur at short sampling intervals by relating the sample follow up time to both the probability of telomere lengthening (GLM with a binomial error structure, telomere lengthening yes/no as response), and to the dgree of lengthening (LM using only samples where RTL increased over time, $\Delta$RTL as response).
 
-We used linear mixed models with model averaging to explore how spatial variation in early-life environmental and social conditions influenced telomere length within cohorts.
-Model averaging was carried out using the MuMIn package (version 1.10.5) in R [@Barton2012].
-We first created a full model containing the following explanatory variables: age (in months), tarsus length, sex, territory quality and the number of helping and non-helping subordinate birds present in the natal territory.
-We also included interaction terms between age and all the other variables as telomere dynamics are expected to vary within the first year of life [@Heidinger2012].
-As random effects we included cohort, territory ID and qPCR plate ID.
-Model selection was then performed and a top model set defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011].
+We used LMMs to explore how variation in early-life environmental and social conditions influenced telomere length within cohorts.
+We created a full model with RTL as a response variable, alongside the following explanatory variables: age, tarsus length, body mass, sex, annual food availability, population density (an annual measure estimated from the summer breeding census) and the number of helping and non-helping subordinate birds present in the natal territory.
+As random effects we included birth year and individual identity.
+We report model estimates and confidence intervals for all effects included in the full model.
+As our analysis is exploratory, and because multicollinearity between variables in our full model may have obscured real effects, we also performed model averaging, using the MuMIn package in R [@Barton2012].
+Model selection was performed using the full model described above and a top model set defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011].
 We report model-averaged coefficients, confidence intervals and 'relative importance', which reflects the relative weights of each predictor variable across the top model set.
-For individuals with longitudinal data (*n* = `r nrow(dd3)`) we repeated the above analyses of telomere dynamics, replacing telomere length with telomere loss as the response variable, and excluding the plate ID random effect (as each measurement of telomere loss was based on two or more measurements, and so run on multiple plates).
-For both sets of models, we first ran the model averaging procedure including all interaction terms, then re-ran the pipeline excluding non-significant interactions in order to obtain more accurate model averaged parameter estimates and confidence intervals.
-
-We used generalised linear mixed models to test whether telomere length/loss were related to survival to adulthood.
-Models were set up with a binomial error structure, and included RTL or $\Delta$RTL and juvenile age as explanatory variables, along with the same random effects as for the model averaging (see above).
-We then used Cox regression, implemented in the survival package in R [@Therneau2011], to test whether long-term survival was related to individual-level telomere dynamics.
-For the cross-sectional data we used lifespan (in years) as survival time, while for the longitudinal data we used remaining lifespan from the second sampling event.
-We ran these models with cohort as a frailty term to exclude cohort effects, and only included cohorts in which at least 70 percent of individuals had died (all years up to and including 2008).
+For individuals with longitudinal data (*n* = `r nrow(dd3)`) we repeated the above analyses of telomere dynamics, replacing telomere length with $\Delta$RTL as the response variable, and excluding the plate ID random effect (as each measurement of $\Delta$RTL was based on two or more measurements, and so run on multiple plates).
 
@@
 #####Telomere dynamics and age
-Using the cross-sectional data, we found that RTL decreased with age in the Seychelles warbler (LMM with individual ID as random effect: estimate $\pm$ s.e.
-= `r round3(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,1])` $\pm$ `r round3(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,2])`; t = `r round2(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,4])`; P `r round3(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,5])`).
-We found that modelling a non-linear effect age explained almost double the variation in RTL (R^2^ of linear regression of telomere length and log age = `r round3(summary(lm(RTL~log(Agemonths),data=dd))$r.squared)` than fitting a linear relationship `r round3(summary(lm(RTL~Agemonths,data=dd))$r.squared)`, suggesting that telomere length decreased at the greatest rate in early life.
-Accordingly, RTL decreased with age at a greater rate in juveniles (estimate $\pm$ s.e.
-= `r round3(summary(lme(RTL~Agemonths,data=juv,random=~1|BirdID/LayYear))$tTable[2,1],lessthan = F)` $\pm$ `r round3(summary(lme(RTL~Agemonths,data=juv,random=~1|BirdID/LayYear))$tTable[2,2])`) compared to adults (estimate $\pm$ s.e.
-= `r round3(summary(lme(RTL~Agemonths,data=subset(dd,Ageclass == 'A'),random=~1|BirdID/LayYear))$tTable[2,1],lessthan = F)` $\pm$ `r round3(summary(lme(RTL~Agemonths,data=subset(dd,Ageclass == 'A'),random=~1|BirdID/LayYear))$tTable[2,2])`).
+Using the cross-sectional data, we found that RTL decreased with age in the Seychelles warbler (LMM: estimate = `r round3(fixef(RTL_Age)['Agemonths'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Age)['Agemonths',],lessthan = F),collapse = ', ')`).
+RTL also decreased with log-transformed age (LMM: estimate = `r round2(fixef(RTL_LogAge)['LogAge'])`, CIs = `r paste(round2(CIwald(RTL_LogAge)['LogAge',]),collapse = ', ')`), and this model was a significantly better fit than one modelling a linear relationship between RTL and age (P = `r round3(anova(RTL_Age,RTL_LogAge)$"Pr(>Chisq)"[2])`), suggesting that telomere length decreased at the greatest rate in early life - somehting that was evident from the raw data (Fig.
+1).
+All further models therefore used log-transformed age as a covariate.
 Also worth noting is that while the overall trend was for decreasing RTL, there were appoarent increases in cross-sectional telomere length after one year of age and at several points later in life (Fig.
@@
 
-Including birth year as a random effect in a LMM of RTL and (log) age resulted in a significantly improved model to one with only individual ID as a random effect (), and was a significantly better model than one including catch year as a random effect.
-This is indicative of cohort effects on telomere length across the Seychelles warbler lifespan.
-Furthermore, we found no evidence that the relationship between telomere length and age varied among cohorts ().
-Rather, cohort-level differences in RTL were consistent over the lifespan of Seychelles warblers.
+Including birth year as a random effect in a LMM of RTL and (log) age resulted in a significantly improved model to one with only individual ID as a random effect (P = `r round3(anova(RTL_LogAge,RTL_LogAge_LY)$"Pr(>Chisq)"[2])`).
+Allowing the slope of the age effect to vary among birth years resulted in a slightly significant improvement on one in which only the intercept was allowed to vary (P = `r round3(anova(RTL_LogAge_LY,RTL_LogAge_AgeLY)$"Pr(>Chisq)"[2])`).
+However, a model including catch year and individual ID as random effects was significantly better than a model indluding birth year and individual ID (P = `r round3(anova(RTL_LogAge_LY,RTL_LogAge_CY)$"Pr(>Chisq)"[2])`), and a model that allowed the slope of the age effect to vary among catch years was better still (P = `r round3(anova(RTL_LogAge_CY,RTL_LogAge_AgeCY)$"Pr(>Chisq)"[2])`).
 
@@
 As with cross-sectional RTL, $\Delta$RTL decreased with age (R^2^ = `r round3(summary(lm(DeltaRTL~LogAge,data = dd3))$r.squared)`; t = `r round2(summary(lm(DeltaRTL~LogAge,data = dd3))$coefficients[2,3])`; P = `r round3(summary(lm(DeltaRTL~LogAge,data = dd3))$coefficients[2,4])`), suggesting that telomere shortening was highest in early life.
-Although both or cross-sectional and longitudinal data indicated a general trend of telomere shortening, we found that `r round((nrow(subset(dd3,DeltaRTL>0))/nrow(dd3))*100,0)% of our $\Delta$RTL measurments were positive, and thus indicative of telomere lengthening.
+Although both or cross-sectional and longitudinal data indicated a general trend of telomere shortening, we found that `r round((nrow(subset(dd3,DeltaRTL>0))/nrow(dd3))*100,0)`% of our $\Delta$RTL measurments were positive, and thus indicative of telomere lengthening.
 To test whether the apparent telomere lengthening in our dataset could be explained by measurement error, we compared variance in telomere length among repeat measurements of samples to the variance observed among different samples of the same individual (see methods for details).

### 2016-11-13 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Home eds

@@
 
-Fig3A <- transform(juv,Cohort = factor(LayYear)) %>%
+Fig3A <- transform(juv,Cohort = factor(CatchYear)) %>%
   subset(Agemonths>(1/12)) %>%
@@
 
-Fig3B <- transform(dd,Cohort = factor(LayYear)) %>%
+Fig3B <- transform(dd,Cohort = factor(CatchYear)) %>%
   ddply(.(Cohort),

### 2016-11-13 Scripts/SW juvenile telomeres IPC functions.r

@@ -457,3 +457,7 @@ give.n <- function(x)
   return(c(y = median(x)*0.96, label = length(x)))
   }
 
+CIwald <- function(model)
+{
+  confint.merMod(model,method = 'Wald')
+}

### 2016-11-14 Manuscript/SW_early_life_telomeres.Rmd
Home edits

@@
 
-Including birth year as a random effect in a LMM of RTL and (log) age resulted in a significantly improved model to one with only individual ID as a random effect (P = `r round3(anova(RTL_LogAge,RTL_LogAge_LY)$"Pr(>Chisq)"[2])`).
-Allowing the slope of the age effect to vary among birth years resulted in a slightly significant improvement on one in which only the intercept was allowed to vary (P = `r round3(anova(RTL_LogAge_LY,RTL_LogAge_AgeLY)$"Pr(>Chisq)"[2])`).
-However, a model including catch year and individual ID as random effects was significantly better than a model indluding birth year and individual ID (P = `r round3(anova(RTL_LogAge_LY,RTL_LogAge_CY)$"Pr(>Chisq)"[2])`), and a model that allowed the slope of the age effect to vary among catch years was better still (P = `r round3(anova(RTL_LogAge_CY,RTL_LogAge_AgeCY)$"Pr(>Chisq)"[2])`).
+Including birth year as a random effect in a LMM of RTL and (log) age resulted in a significantly improved model to one with only individual ID as a random effect (P `r round3(anova(RTL_LogAge,RTL_LogAge_LY)$"Pr(>Chisq)"[2])`; Fig.
+2A).
+Allowing the slope of the age effect to vary among birth years resulted in a slightly significant improvement on one in which only the intercept was allowed to vary (P = `r round3(anova(RTL_LogAge_LY,RTL_LogAge_AgeLY)$"Pr(>Chisq)"[2])`; Fig.
+2B).
+However, a model including catch year and individual ID as random effects was significantly better than a model indluding birth year and individual ID (P `r round3(anova(RTL_LogAge_LY,RTL_LogAge_CY)$"Pr(>Chisq)"[2])`) suggesting that variation among catch years was greater than among cohorts (Fig.
+2C), and a model that allowed the slope of the age effect to vary among catch years was better still (P = `r round3(anova(RTL_LogAge_CY,RTL_LogAge_AgeCY)$"Pr(>Chisq)"[2])`; Fig.
+2D).
+Using longitudinal data and $\Delta$Age as an explanatory variable we found a similar pattern - a model including catch year as a random effect was better than that did not allow for temporal variation in RTL (P = `r round3(anova(RTL_DeltaAge,RTL_DeltaAge_CY)$"Pr(>Chisq)"[2])`), and better than one with birth year as a random effect (P `r round3(anova(RTL_DeltaAge_LY,RTL_DeltaAge_CY)$"Pr(>Chisq)"[2])`).
+However a model allowing the slope of the $\Delta$Age effect was not better than one in which only the intercept varied (P = `r round3(anova(RTL_DeltaAge_CY,RTL_DeltaAge_AgeCY)$"Pr(>Chisq)"[2])`).
 
-Longitudinal data showed that early-life RTL was weakly, but significantly, related to adult RTL (R^2^ = `r round3(summary(lm(RTL1~RTL,data = dd3))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL,data = dd3))$coefficients[2,3])`; P = `r round3(summary(lm(RTL1~RTL,data = dd3))$coefficients[2,4])`; Fig.
+Within individuals, longitudinal data showed that RTL measurements taken from different timepoints were weakly, but significantly, positvely correlated (R^2^ = `r round3(summary(lm(RTL1~RTL,data = dd3))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL,data = dd3))$coefficients[2,3])`; P = `r round3(summary(lm(RTL1~RTL,data = dd3))$coefficients[2,4])`; Fig.
 1B).
-As with cross-sectional RTL, $\Delta$RTL decreased with age (R^2^ = `r round3(summary(lm(DeltaRTL~LogAge,data = dd3))$r.squared)`; t = `r round2(summary(lm(DeltaRTL~LogAge,data = dd3))$coefficients[2,3])`; P = `r round3(summary(lm(DeltaRTL~LogAge,data = dd3))$coefficients[2,4])`), suggesting that telomere shortening was highest in early life.
-Although both or cross-sectional and longitudinal data indicated a general trend of telomere shortening, we found that `r round((nrow(subset(dd3,DeltaRTL>0))/nrow(dd3))*100,0)`% of our $\Delta$RTL measurments were positive, and thus indicative of telomere lengthening.
+Although both or cross-sectional and longitudinal data indicated a general trend of telomere shortening with age, we found that `r round((nrow(subset(dd3,DeltaRTL>0))/nrow(dd3))*100,0)`% of our $\Delta$RTL measurments were positive, and thus indicative of telomere lengthening.
 To test whether the apparent telomere lengthening in our dataset could be explained by measurement error, we compared variance in telomere length among repeat measurements of samples to the variance observed among different samples of the same individual (see methods for details).
@@
 A linear mixed model of RTL in relation to social and ecological environmental variables revealed age, tarsus length, sex and annual variation in food availability were significantly related to RTL (Table 1).
-Model selection revealed that sex only appeared in models where tarsus length was also present (Table S1), and sex was no longer significant when tarsus length was removed from the full model (P = ).
+Model selection revealed that sex only appeared in models where tarsus length was also present (Table S1), and sex was no longer significant when tarsus length was removed from the full model (estimate = `r round3(fixef(RTL_Full_NoTarsus)['SexMales'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Full_NoTarsus)['SexMales',],lessthan = F),collapse = ', ')`).
 RTL increased with tarsus length in males, but decreased in females.
-In general, the model averaging approach yielded qualitatively identical results to the full LMM, with the same explanatory variables 'significant' in terms of being retained in top models, and having model-average confidence intervals not overlapping zero (Table S1; Fig.
+The model averaging approach yielded qualitatively identical results to the full LMM, with the same explanatory variables 'significant' in terms of being retained in top models, and having model-averaged confidence intervals not overlapping zero (Table S1; Fig.
 S1).
 
-$\Delta$RTL did not differ significantly among cohorts (F = `r round2(anova(lm(DeltaRTL~factor(FieldPeriodID),data = dd3))[1,4])`; P = `r round(anova(lm(DeltaRTL~factor(FieldPeriodID),data = dd3))[1,5])`), although power for this analysis was limited as longitudinal sample sizes within cohorts were low.
-Model selection revealed that the top model explaining $\Delta$RTL was the null model (Table S2).
+Model selection revealed that the top model explaining $\Delta$RTL was...
 

### 2016-11-14 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Home edits

@@
 ```{r Table 1}
-temp <- subset(dd,!is.na(Insect))
 
-#Run the model
-lm2 <- lmer(RTL ~  LogAge + Tarsus + Helper + GroupSize + Sex + BodyMass + Insect + Density + (1|LayYear) + (1|BirdID) + (1|PlateID),
-            data = temp,
-            REML = FALSE)
-CIs <- confint.merMod(lm2)
-t1 <- data.frame(Estimate = fixef(lm2), s.e. = se.fixef(lm2), LCI = CIs[5:nrow(CIs),1], UCI = CIs[5:nrow(CIs),2])
+CIs <- CIwald(RTL_Full)
+t1 <- data.frame(Estimate = fixef(RTL_Full), LCI = CIs[6:nrow(CIs),1], UCI = CIs[6:nrow(CIs),2])
 kable(t1)
@@
 ```{r Table 2}
+
 temp <- subset(dd3,!is.na(Insect))
@@
 #Run the model
-lm2 <- lmer(DeltaRTL ~  LogAge + Tarsus + Helper + GroupSize + Sex + BodyMass + Insect + Density + (1|LayYear) + (1|BirdID) + (1|PlateID),
-            data = temp,
-            REML = FALSE)
-CIs <- confint.merMod(lm2)
-t1 <- data.frame(Estimate = fixef(lm2), s.e. = se.fixef(lm2), LCI = CIs[5:nrow(CIs),1], UCI = CIs[5:nrow(CIs),2])
+CIs <- CIwald(DeltaRTL_Full)
+t1 <- data.frame(Estimate = fixef(DeltaRTL_Full), LCI = CIs[4:nrow(CIs),1], UCI = CIs[4:nrow(CIs),2])
 kable(t1)

### 2016-11-14 Scripts/SW juvenile telomeres IPC models.r

@@ -1,7 +1,34 @@
+#################################################################################*
+## Telomere dynamics in a wild bird population
+## LOAD MODELS
+#################################################################################*
+
+
+# Telomere length, age and cohort -----------------------------------------
 
 RTL_Age <- lmer(RTL~Agemonths + (1|BirdID),data = dd)
 RTL_LogAge <- lmer(RTL~LogAge + (1|BirdID),data = dd)
 RTL_LogAge_LY <- lmer(RTL~LogAge + (1|BirdID) + (1|LayYear),data = dd)
-RTL_LogAge_AgeLY <- lmer(RTL~LogAge + (1|BirdID) + (Agemonths|LayYear),data = dd)
 RTL_LogAge_CY <- lmer(RTL~LogAge + (1|BirdID) + (1|CatchYear),data = dd)
+RTL_LogAge_AgeLY <- lmer(RTL~LogAge + (1|BirdID) + (Agemonths|LayYear),data = dd)
 RTL_LogAge_AgeCY <- lmer(RTL~LogAge + (1|BirdID) + (Agemonths|CatchYear),data = dd)
+
+RTL_DeltaAge <- lmer(RTL~DeltaAge + (1|BirdID),data = dd3)
+RTL_DeltaAge_LY <- lmer(RTL~DeltaAge + (1|BirdID) + (1|LayYear),data = dd3)
+RTL_DeltaAge_CY <- lmer(RTL~DeltaAge + (1|BirdID) + (1|CatchYear),data = dd3)
+RTL_DeltaAge_AgeLY <- lmer(RTL~DeltaAge + (1|BirdID) + (DeltaAge|LayYear),data = dd3)
+RTL_DeltaAge_AgeCY <- lmer(RTL~DeltaAge + (1|BirdID) + (DeltaAge|CatchYear),data = dd3)
+
+
+# RTL and ecology ---------------------------------------------------------
+
+RTL_Full <- lmer(RTL ~  LogAge + Tarsus + Helper + GroupSize + Sex + BodyMass + Insect + Density + (Agemonths|CatchYear) + (1|BirdID),
+                 data = subset(dd,!is.na(Insect)),
+                 REML = FALSE)
+RTL_Full_NoTarsus <- lmer(RTL ~  LogAge + Helper + GroupSize + Sex + BodyMass + Insect + Density + (Agemonths|CatchYear) + (1|BirdID),
+                          data = subset(dd,!is.na(Insect)),
+                          REML = FALSE)
+
[…]

### 2016-11-16 Manuscript/SW_early_life_telomeres.Rmd
Eds

@@
 #####Telomere dynamics and age
-Using the cross-sectional data, we found that RTL decreased with age in the Seychelles warbler (LMM: estimate = `r round3(fixef(RTL_Age)['Agemonths'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Age)['Agemonths',],lessthan = F),collapse = ', ')`).
+Using cross-sectional data we found that RTL decreased with age in the Seychelles warbler (LMM: estimate = `r round3(fixef(RTL_Age)['Agemonths'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Age)['Agemonths',],lessthan = F),collapse = ', ')`).
 RTL also decreased with log-transformed age (LMM: estimate = `r round2(fixef(RTL_LogAge)['LogAge'])`, CIs = `r paste(round2(CIwald(RTL_LogAge)['LogAge',]),collapse = ', ')`), and this model was a significantly better fit than one modelling a linear relationship between RTL and age (P = `r round3(anova(RTL_Age,RTL_LogAge)$"Pr(>Chisq)"[2])`), suggesting that telomere length decreased at the greatest rate in early life - somehting that was evident from the raw data (Fig.
@@
 A linear mixed model of RTL in relation to social and ecological environmental variables revealed age, tarsus length, sex and annual variation in food availability were significantly related to RTL (Table 1).
-Model selection revealed that sex only appeared in models where tarsus length was also present (Table S1), and sex was no longer significant when tarsus length was removed from the full model (estimate = `r round3(fixef(RTL_Full_NoTarsus)['SexMales'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Full_NoTarsus)['SexMales',],lessthan = F),collapse = ', ')`).
-RTL increased with tarsus length in males, but decreased in females.
 The model averaging approach yielded qualitatively identical results to the full LMM, with the same explanatory variables 'significant' in terms of being retained in top models, and having model-averaged confidence intervals not overlapping zero (Table S1; Fig.
 S1).
+One interesting finding from the model selection was that sex only appeared in top models where tarsus length was also present (Table S1).
+In accordance with this, when tarsus length was removed from the full model sex was no longer significant (estimate = `r round3(fixef(RTL_Full_NoTarsus)['SexMales'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Full_NoTarsus)['SexMales',],lessthan = F),collapse = ', ')`), and a sex x tarsus interaction was significant when included (estimate = `r round3(fixef(RTL_Full_SexTarsus)['SexMales:Tarsus'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Full_SexTarsus)['SexMales:Tarsus',],lessthan = F),collapse = ', ')`); RTL increased with tarsus length in males, but decreased in females.
 
-Model selection revealed that the top model explaining $\Delta$RTL was...
+No social or ecological environmental variables were significant in a model of longitudinal data with $\Delta$RTL as the response variable (Table 2).
+The model averaging approach suggested that $\Delta$RTL was positively related to age (i.e.
+telomere shortening increased with age), and negatively associated with annual density (i.e.
+more shortening in years with higher population density; Table S2; Fig.
+S2).
 
 ###Discussion
-Here we use one of the largest datasets to date to assess the relationships between early-life conditions, telomere dynamics and survival in a closed population of Seychelles warblers.
-We found that, as in other studies, telomeres are lost at the greatest rate early in life.
-We found that telomere dynamics are subject to strong cohort effects.
-Within cohorts, we found little evidence that the ecological or social environment influenced telomere dynamics, and we found no evidence that early-life telomere length is related to juvenile or later-life survival.
+Here we use a large, long-term, multi-cohort dataset to assess the relationships between the spatiotemporal variation in the ecological environment and lifelong telomere dynamics in a closed population of Seychelles warblers.
+We found that, as in other studies, telomeres shorten over time, and that telomere loss is greatest in early life.
+However, bouts of apparent telomere lengthening occur, and the extent of this lengthening cannot be explained solely by measurement error.
+We then found that telomere dynamics vary among sexes, and over time, with this temporal variation is related to fluctuations in food availability.
 
-Our study adds the the now substantial body of literature showing that telomere length decreases with age, and that this decrease is most rapid in early life [e.g.
+Our study adds the the now substantial body of literature showing that telomere length decreases with age in both humans and wild animals, and that this decrease is most rapid in early life [e.g.
 @Frenck1998; @Haussmann2003; @Heidinger2012; @Fairlie2016].
-Our finding that tarsus length was related to juvenile RTL in an age-dependent manner (Fig.
-3) most likely represents an additional age effect - in passerine birds tarsus length is correlated with age during the nestling stage, but not after fledging [@Ricklefs1976].
-Telomere loss is most rapid early in life due to ongoing growth [@Frenck1998], and a negative correlation between RTL and body size in early life is therefore expected [see also @Ringsby2015].
-We also found weak evidence for sex-specific telomere dynamics in Seychelles warblers, with males having longer-telomeres and higher loss of telomeres (Fig.
-3C, Fig.
-5B).
-We do not wish to speculate too much on this effect as confidence intervals for both sex effects overlapped zero.
-However, it is worth noting while we found no evidence that juvenile RTL has a sex-specific effect on survival, in adults the effect of telomere length on survival is strongest for males [@Barrett2013a].
-Together, these results suggest that sex effects should at least be considered when studying any aspect of telomere dynamics [see also @Barrett2011].
+However, using longitudinal data we found that within-individual telomere measurments were only weakly correlated, and that consistent telomere shortening only occurs early in life, after which both shortening and lengthening occurs.
+This is in contrast to other avian studies where birds were reared in laboratory [@Heidinger2012] or nest-box [@Boonekamp2014] conditions, in which juvenile and adult telomere length were highly correlated, and individual-level telomere shortening clearly occurs throughout the juvenile period, and into adulthood.
+However, in accordance with our study, a recent study on a wild population of Soay sheep (*Ovis aries*) found only a very weak correlation between juvenile and adult RTL, and a highly complex pattern of telomere dynamics across the lifespan [@Fairlie2016].
 
- We found that at the individual-level, juvenile and adult telomere were not correlated, and that consistent telomere shortening only occurs  very early in life (Fig.
-4).
-This is in contrast to other avian studies where birds were reared in laboratory [@Heidinger2012] or nest-box [@Boonekamp2014] conditions, in which juvenile and adult telomere length were highly correlated, and individual-level telomere shortening clearly occurs throughout the juvenile period, and into adulthood.
-However, in accordance with our study, a recent study on Soay sheep (*Ovis aries*) found only a very weak correlation between juvenile and adult RTL, and a highly complex pattern of telomere dynamics across the lifespan [@Fairlie2016].
-It is too soon to speculate as to what may be generating this variation among species, but one possibility is that environmental variability itself affects the resolution with which we can observe telomere shortening in natural populations [see also @Fairlie2016].
-Regardless, our data suggest the efficacy of telomeres as individual biomarkers of cost in wild populations will depend on the age structure of the study population, as well as the sampling resolution.
+We also found evidence for sex-specific telomere dynamics in Seychelles warblers, with males having longer-telomeres than females, in and that this sex-difference interacts with tarsus length.
+In Seychelles warblers the effect of telomere length on survival is strongest for males [@Barrett2013a].
+Together, these results suggest that sex effects should be considered when studying any aspect of telomere dynamics [see also @Barrett2011].
 

### 2016-11-16 Manuscript/SW_early_life_telomeres_SOM.Rmd
Eds

@@
 FigS1 <- forestplot(table2)+
-  annotate('text',x = 0.6,y = 0.05,label = 'A')+
   theme(axis.line.x = element_line(colour = 'black'),
@@
 FigS2 <- forestplot(table2)+
-  annotate('text',x = 0.6,y = 0.05,label = 'A')+
   theme(axis.line.x = element_line(colour = 'black'),

### 2016-11-16 Scripts/SW juvenile telomeres IPC models.r

@@ -28,7 +28,9 @@ RTL_Full <- lmer(RTL ~  LogAge + Tarsus + Helper + GroupSize + Sex + BodyMass + 
 RTL_Full_NoTarsus <- lmer(RTL ~  LogAge + Helper + GroupSize + Sex + BodyMass + Insect + Density + (Agemonths|CatchYear) + (1|BirdID),
                           data = subset(dd,!is.na(Insect)),
                           REML = FALSE)
-
+RTL_Full_SexTarsus <- lmer(RTL ~  LogAge + Helper + GroupSize + Sex + BodyMass + Insect + Density + Sex*Tarsus + (Agemonths|CatchYear) + (1|BirdID),
+                          data = subset(dd,!is.na(Insect)),
+                          REML = FALSE)
 DeltaRTL_Full <- lmer(DeltaRTL ~  LogAge + Tarsus + Helper + GroupSize + Sex + BodyMass + Insect + Density + (1|CatchYear) + (1|BirdID),
                              data = subset(dd3,!is.na(Insect)),
                              REML = FALSE)

### 2016-11-18 Manuscript/SW_early_life_telomeres.Rmd
Eds

@@
 
-**Early-life conditions, telomere dynamics and survival: insights from a long-term island bird study**
+**Spatiotemporal variation in lifelong telomere dynamics in the Seychelles warbler**
 
@@
 We found that telomere length decreases with cross-sectional and longitudinal measures of age, and that telomere length decreases most rapidly early in life.
-However, both cross-sectional and longitudinal data suggested that telomere length increases occurs in some individuals, and using a large number of repeated measurements we show statistically that this increase is unlikely to be explained by qPCR measurement error.
+However, both cross-sectional and longitudinal data suggested that telomere length increases occurs in some individuals for short periods, and using a large number of repeated measurements we show statistically that this increase cannot be explained solely by qPCR measurement error.
 
@@
 115 territories lay one or, rarely, two or three eggs [@Komdeur1991].
-As a result of this low reproductive output, combined with higher mortality in first-year birds [39%; @Brouwer2006], cohort sizes in the Seychelles warbler are small (<50).
+As a result of this low reproductive output, combined with higher mortality in first-year birds [39%; @Brouwer2006], cohort sizes in the Seychelles warbler are small (< 50).
 
@@
 We calculated $\Delta$RTL between pairs of repeat measurements within samples in exactly the same way as for across samples, except that repeat measurements were ordered by the date at the qPCR was run rather than by age.
+We first compared the variance in $\Delta$RTL within and among individuals using a Levene's test.
+Then, to separately test whether the extent of telomere shortening and lenghtening within individuals was greater than expected by measurement error, we split $\Delta$RTL measurements into groups in which shortening ($\Delta$RTL < 0) and lengthening $\Delta$RTL ($\Delta$RTL > 0) occurred, and tested whether $\Delta$RTL among samples was significantly different between within-sample and across-sample measurements, using Wilcoxon tests.
 
-We used LMMs to explore how variation in early-life environmental and social conditions influenced telomere length within cohorts.
+We used LMMs to test how variation in early-life environmental and social conditions influenced telomere length within cohorts.
 We created a full model with RTL as a response variable, alongside the following explanatory variables: age, tarsus length, body mass, sex, annual food availability, population density (an annual measure estimated from the summer breeding census) and the number of helping and non-helping subordinate birds present in the natal territory.
@@
 
-Including birth year as a random effect in a LMM of RTL and (log) age resulted in a significantly improved model to one with only individual ID as a random effect (P `r round3(anova(RTL_LogAge,RTL_LogAge_LY)$"Pr(>Chisq)"[2])`; Fig.
+Including birth year as a random effect in a LMM of RTL and (log) age resulted in a significantly improved model to one with only individual ID ans plate ID as random effects (P `r round3(anova(RTL_LogAge,RTL_LogAge_LY)$"Pr(>Chisq)"[2])`; Fig.
 2A).
@@
 2B).
-However, a model including catch year and individual ID as random effects was significantly better than a model indluding birth year and individual ID (P `r round3(anova(RTL_LogAge_LY,RTL_LogAge_CY)$"Pr(>Chisq)"[2])`) suggesting that variation among catch years was greater than among cohorts (Fig.
+However, a model including catch year and individual ID as random effects was significantly better than a model indluding birth year and individual ID (P `r round3(anova(RTL_LogAge_LY,RTL_LogAge_CY)$"Pr(>Chisq)"[2])`; Fig 2C) suggesting that variation among catch years was greater than among cohorts (Fig.
 2C), and a model that allowed the slope of the age effect to vary among catch years was better still (P = `r round3(anova(RTL_LogAge_CY,RTL_LogAge_AgeCY)$"Pr(>Chisq)"[2])`; Fig.
@@
 Within individuals, longitudinal data showed that RTL measurements taken from different timepoints were weakly, but significantly, positvely correlated (R^2^ = `r round3(summary(lm(RTL1~RTL,data = dd3))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL,data = dd3))$coefficients[2,3])`; P = `r round3(summary(lm(RTL1~RTL,data = dd3))$coefficients[2,4])`; Fig.
-1B).
+3A).
 Although both or cross-sectional and longitudinal data indicated a general trend of telomere shortening with age, we found that `r round((nrow(subset(dd3,DeltaRTL>0))/nrow(dd3))*100,0)`% of our $\Delta$RTL measurments were positive, and thus indicative of telomere lengthening.
@@
 We found significantly higher variance in telomere length over individual lifetimes compared to among sample replicates (Levene's test: F = `r round2(leveneTest(DeltaRTL~Group,data=ddL)['F value'][1,1])`; P `r round3(leveneTest(DeltaRTL~Group,data=ddL)['Pr(>F)'][1,1])`; Fig.
-1C).
-Importantly, splitting the longitudinal data into instances of shortening ($\Delta$RTL < 0) and lenghtning ($\Delta$RTL > 0) revealed that not only did we observe significantly more shortening within individuals compared to within samples (Wilcoxon test: W = `r round(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$statistic,0)`; P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$p.value)`), but also significantly more lengthening (W = `r as.character(round(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$statistic,0))`; P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$p.value)`).
+3B).
+Importantly, splitting the longitudinal data into instances of shortening ($\Delta$RTL < 0) and lenghtning ($\Delta$RTL > 0) revealed that not only did we observe significantly more shortening within individuals compared to within samples (Wilcoxon test: W = `r round(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$statistic,0)`; P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$p.value)`), but also significantly more lengthening (W = `r as.character(round(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$statistic,0))`; P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$p.value)`; Fig.
+3B).
 
 #####Spatiotemporal variation in telomere dynamics and the environment
-A linear mixed model of RTL in relation to social and ecological environmental variables revealed age, tarsus length, sex and annual variation in food availability were significantly related to RTL (Table 1).
+A LMM revealed that age, tarsus length, sex and annual variation in food availability were significantly related to RTL (Table 1).
+Tarsus length was negatively related to RTL and males had longer telomeres than females (Table 1; Fig.
+4A), while insect availability was positively related to RTL (Table 1; Fig.
+4B).
 The model averaging approach yielded qualitatively identical results to the full LMM, with the same explanatory variables 'significant' in terms of being retained in top models, and having model-averaged confidence intervals not overlapping zero (Table S1; Fig.
@@
 One interesting finding from the model selection was that sex only appeared in top models where tarsus length was also present (Table S1).
-In accordance with this, when tarsus length was removed from the full model sex was no longer significant (estimate = `r round3(fixef(RTL_Full_NoTarsus)['SexMales'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Full_NoTarsus)['SexMales',],lessthan = F),collapse = ', ')`), and a sex x tarsus interaction was significant when included (estimate = `r round3(fixef(RTL_Full_SexTarsus)['SexMales:Tarsus'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Full_SexTarsus)['SexMales:Tarsus',],lessthan = F),collapse = ', ')`); RTL increased with tarsus length in males, but decreased in females.
+In accordance with this, when tarsus length was removed from the full model sex was no longer significant (estimate = `r round3(fixef(RTL_Full_NoTarsus)['SexMales'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Full_NoTarsus)['SexMales',],lessthan = F),collapse = ', ')`), and a sex x tarsus interaction was significant when included (estimate = `r round3(fixef(RTL_Full_SexTarsus)['SexMales:Tarsus'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Full_SexTarsus)['SexMales:Tarsus',],lessthan = F),collapse = ', ')`); RTL increased with tarsus length in males, but decreased in females (Fig.
+4A).
 
@@
 telomere shortening increased with age), and negatively associated with annual density (i.e.
-more shortening in years with higher population density; Table S2; Fig.
-S2).
+more shortening in years with higher population density; Table S2; Figs S2; S3).
 
 ###Discussion
-Here we use a large, long-term, multi-cohort dataset to assess the relationships between the spatiotemporal variation in the ecological environment and lifelong telomere dynamics in a closed population of Seychelles warblers.
-We found that, as in other studies, telomeres shorten over time, and that telomere loss is greatest in early life.
+Here we use a long-term, multi-cohort dataset to assess the relationships between the spatiotemporal variation in the ecological environment and lifelong telomere dynamics in a closed population of Seychelles warblers.
+We found that telomeres shorten over time, and that telomere loss is greatest in early life.
 However, bouts of apparent telomere lengthening occur, and the extent of this lengthening cannot be explained solely by measurement error.
@@
 This work was funded by two Natural Environment Research Council (NERC) grants to DSR (NE/F02083X/1 and NE/K005502/1).
-LGS was also funded by fellowships from the University of Oxford and the BBSRC, and HLD by a NERC fellowship.
+LGS was also funded by a fellowship from the BBSRC, and HLD by a NERC fellowship.
 

### 2016-11-18 Manuscript/SW_early_life_telomeres_SOM.Rmd
Eds

@@
 
-#Run the model
-temp <- subset(dd,!is.na(Insect))
+#Model averaging
 
-#Run the model
-lm2 <- lmer(RTL ~  LogAge + Tarsus + Helper + GroupSize + Sex + BodyMass + Insect + Density + (1|LayYear) + (1|BirdID) + (1|PlateID),
-            data = temp,
-            REML = FALSE)
-lm2 <- standardize(lm2)
-
-#Model averaging
-lm3 <- dredge(lm2)
 t1 <- as.matrix(subset(lm3, delta<=6))
@@
 ```{r Table S2}
-#Run the model
-temp <- subset(dd3,!is.na(Insect))
-
-#Run the model
-lm2 <- lmer(DeltaRTL ~  LogAge + Tarsus + Helper + GroupSize + Sex + BodyMass + Insect + Density + (1|LayYear) + (1|PlateID),
-            data = temp,
-            REML = FALSE)
-lm2 <- standardize(lm2)
 
 #Model averaging
-lm3 <- dredge(lm2)
+lm3 <- dredge(DeltaRTL_Full)
 t1 <- as.matrix(subset(lm3, delta<=6))

### 2016-11-18 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Eds

@@
 
-Fig2A <- ggplot(dd3,
+Fig2A <- transform(juv,Cohort = factor(CatchYear)) %>%
+  subset(Agemonths>(1/12)) %>%
+  ggplot(aes(y = RTL,x = Cohort))+
+  xlab('Cohort') +
+  geom_boxplot() +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12,angle = 90), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'))
+  annotate('text',x = 0,y = 1.4,label = 'B')
+
+Fig2B <- transform(dd,Cohort = factor(CatchYear)) %>%
+  ddply(.(Cohort),
+        mutate,
+        nc = length(which(Agemonths ==1/12))) %>%
+  subset(nc > 11) %>%
+  ggplot(aes(y = RTL,x = Agemonths, col = Cohort))+
+  xlab('Age (years)') +
+  geom_smooth(method = 'lm', formula = y~log(x),se = F) +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'))
+  annotate('text',x = 0,y = 1.4,label = 'B')
+
+pdf('Figures/Figure 2.pdf',
+    width = 5,
+    height = 8,
+    paper = 'a4')
+multiplot(Fig2A,Fig2B)
+dev.off()
+
+```
+
+```{r Figure 3}
+
+Fig3A <- ggplot(dd3,
        aes(x = RTL,
@@
 
-Fig2B <- ggplot(ddL,aes(col=Group,fill = Group,y = ..scaled..,x = DeltaRTL))+
+Fig3B <- ggplot(ddL,aes(col=Group,fill = Group,y = ..scaled..,x = DeltaRTL))+
     geom_density(alpha = 0.2)+
@@
   annotate('text',x = -2,y = 1,label = 'C')
-
-pdf('Figures/Figure 2.pdf',
-    width = 5,
-    height = 8,
-    paper = 'a4')
-multiplot(Fig2A,Fig2B)
-dev.off()
-
-```
-
-```{r Figure 3}
-
-Fig3A <- transform(juv,Cohort = factor(CatchYear)) %>%
-  subset(Agemonths>(1/12)) %>%
-  ggplot(aes(y = RTL,x = Cohort))+
-  xlab('Cohort') +
-  geom_boxplot() +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12,angle = 90), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black'))
-  annotate('text',x = 0,y = 1.4,label = 'B')
-
-Fig3B <- transform(dd,Cohort = factor(CatchYear)) %>%
-  ddply(.(Cohort),
-        mutate,
-        nc = length(which(Agemonths ==1/12))) %>%
-  subset(nc > 11) %>%
-  ggplot(aes(y = RTL,x = Agemonths, col = Cohort))+
-  xlab('Age (years)') +
-  geom_smooth(method = 'lm', formula = y~log(x),se = F) +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black'))
-  annotate('text',x = 0,y = 1.4,label = 'B')
 

### 2016-11-18 Scripts/SW juvenile telomeres IPC clean.R

@@ -9,11 +9,11 @@
 
 dd0 <- subset(dd0,CqTelomere < 25) %>%
   subset(CqGAPDH < 26) %>%
-  subset(CqGAPDH > 19) %>%
+  subset(CqGAPDH > 21) %>%
   subset(DiffGAPDH < 0.5) %>%
   subset(DiffTelomere < 0.5) %>%
   subset(RTL < 3) %>%
-  subset(BloodID>0)
+  subset(BloodID>6)
 
 dd0$RTL <- sqrt(dd0$RTL)
 

### 2016-11-18 Scripts/SW juvenile telomeres IPC models.r

@@ -6,12 +6,12 @@
 
 # Telomere length, age and cohort -----------------------------------------
 
-RTL_Age <- lmer(RTL~Agemonths + (1|BirdID),data = dd)
-RTL_LogAge <- lmer(RTL~LogAge + (1|BirdID),data = dd)
-RTL_LogAge_LY <- lmer(RTL~LogAge + (1|BirdID) + (1|LayYear),data = dd)
-RTL_LogAge_CY <- lmer(RTL~LogAge + (1|BirdID) + (1|CatchYear),data = dd)
-RTL_LogAge_AgeLY <- lmer(RTL~LogAge + (1|BirdID) + (Agemonths|LayYear),data = dd)
-RTL_LogAge_AgeCY <- lmer(RTL~LogAge + (1|BirdID) + (Agemonths|CatchYear),data = dd)
+RTL_Age <- lmer(RTL~Agemonths + (1|BirdID) + (1|PlateID),data = dd)
+RTL_LogAge <- lmer(RTL~LogAge + (1|BirdID) + (1|PlateID),data = dd)
+RTL_LogAge_LY <- lmer(RTL~LogAge + (1|BirdID) + (1|PlateID) + (1|LayYear),data = dd)
+RTL_LogAge_CY <- lmer(RTL~LogAge + (1|BirdID) + (1|PlateID) + (1|CatchYear),data = dd)
+RTL_LogAge_AgeLY <- lmer(RTL~LogAge + (1|BirdID) + (1|PlateID) + (Agemonths|LayYear),data = dd)
+RTL_LogAge_AgeCY <- lmer(RTL~LogAge + (1|BirdID) + (1|PlateID) + (Agemonths|CatchYear),data = dd)
 
 RTL_DeltaAge <- lmer(RTL~DeltaAge + (1|BirdID),data = dd3)
 RTL_DeltaAge_LY <- lmer(RTL~DeltaAge + (1|BirdID) + (1|LayYear),data = dd3)
@@ -22,13 +22,13 @@ RTL_DeltaAge_AgeCY <- lmer(RTL~DeltaAge + (1|BirdID) + (DeltaAge|CatchYear),data
 
 # RTL and ecology ---------------------------------------------------------
 
-RTL_Full <- lmer(RTL ~  LogAge + Tarsus + Helper + GroupSize + Sex + BodyMass + Insect + Density + (Agemonths|CatchYear) + (1|BirdID),
[…]

### 2016-11-20 Manuscript/SW_early_life_telomeres.Rmd
Home eds

@@
 We found that telomeres shorten over time, and that telomere loss is greatest in early life.
-However, bouts of apparent telomere lengthening occur, and the extent of this lengthening cannot be explained solely by measurement error.
-We then found that telomere dynamics vary among sexes, and over time, with this temporal variation is related to fluctuations in food availability.
+However, bouts of apparent telomere lengthening occur, and the extent of this lengthening cannot be explained solely by qPCR measurement error.
+We also found that telomere dynamics vary among sexes and over time, with temporal variation in telomere dynamics related to fluctuations in food availability.
 
-Our study adds the the now substantial body of literature showing that telomere length decreases with age in both humans and wild animals, and that this decrease is most rapid in early life [e.g.
+Our study adds the the now substantial body of literature from humans and wild animals, showing that telomere length decreases with age, and that this decrease is most rapid in early life [e.g.
 @Frenck1998; @Haussmann2003; @Heidinger2012; @Fairlie2016].
@@
 We also found evidence for sex-specific telomere dynamics in Seychelles warblers, with males having longer-telomeres than females, in and that this sex-difference interacts with tarsus length.
-In Seychelles warblers the effect of telomere length on survival is strongest for males [@Barrett2013a].
+This is likely to be due to differential growth costs between sexes.
+Furthermore, in Seychelles warblers the effect of telomere length on survival is strongest for males [@Barrett2013a].
 Together, these results suggest that sex effects should be considered when studying any aspect of telomere dynamics [see also @Barrett2011].

### 2016-11-21 Manuscript/SW_early_life_telomeres.Rmd
eds

--- 2016-11-18 Manuscript/SW_early_life_telomeres.Rmd
@@
 ###Introduction
-Telomeres - the protective caps on the ends of chromosomes - may provide a solution to this problem.
 Telomeres shorten with age [@Monaghan2006], and in response to oxidative stress, which can be elevated  due to environmental factors [@VonZglinicki2002].
@@
 We also tested whether telomores shorten with age in individuals with samples from multiple timepoints, using a LMM of RTL and $\Delta$age [a longitudinal measure based on within-subject centring; @VandePol2009].
-To test how telomere length, and its relationship with age, varied among cohorts, we compared the above LMMs with models that included birth year as an additional random effect term.
+To test how telomere length, and its relationship with age, varied among cohorts, we compared the above LMMs with models that included birth year and catch year as additional random effects.
 We also compared a model in which only the intercept (i.e.
-RTL) was allowed to vary among among cohorts, to one in which the slope of the relationship between RTL and age was also allowed to vary.
+RTL) was allowed to vary among among cohorts or years, to one in which the slope of the relationship between RTL and age was also allowed to vary.
 
@@
 We used LMMs to test how variation in early-life environmental and social conditions influenced telomere length within cohorts.
-We created a full model with RTL as a response variable, alongside the following explanatory variables: age, tarsus length, body mass, sex, annual food availability, population density (an annual measure estimated from the summer breeding census) and the number of helping and non-helping subordinate birds present in the natal territory.
-As random effects we included birth year and individual identity.
+We created a full model with RTL as a response variable, alongside the following explanatory variables: age, tarsus length, body mass, sex, annual food availability, territory quality, population density (an annual measure estimated from the summer breeding census) and the number of birds and the number of helping  subordinate birds present in the territory.
+As random effects we included catch year and individual identity.
 We report model estimates and confidence intervals for all effects included in the full model.
-As our analysis is exploratory, and because multicollinearity between variables in our full model may have obscured real effects, we also performed model averaging, using the MuMIn package in R [@Barton2012].
+To obtain an estimate of explanatory power, we calculated marginal R^2 (just incoporatng fixed effects) and conditional R^2 (incorporating fixed and random effects) following Nakagawa & Schielzeth [-@Nakagawa2013] and Johnson [-@Johnson2014], respectively As a complementary approach, we also performed model averaging, using the MuMIn package in R [@Barton2012].
 Model selection was performed using the full model described above and a top model set defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011].
@@
 Using cross-sectional data we found that RTL decreased with age in the Seychelles warbler (LMM: estimate = `r round3(fixef(RTL_Age)['Agemonths'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Age)['Agemonths',],lessthan = F),collapse = ', ')`).
-RTL also decreased with log-transformed age (LMM: estimate = `r round2(fixef(RTL_LogAge)['LogAge'])`, CIs = `r paste(round2(CIwald(RTL_LogAge)['LogAge',]),collapse = ', ')`), and this model was a significantly better fit than one modelling a linear relationship between RTL and age (P = `r round3(anova(RTL_Age,RTL_LogAge)$"Pr(>Chisq)"[2])`), suggesting that telomere length decreased at the greatest rate in early life - somehting that was evident from the raw data (Fig.
-1).
-All further models therefore used log-transformed age as a covariate.
+RTL also decreased with log-transformed age (LMM: estimate = `r round2(fixef(RTL_LogAge)['LogAge'])`, CIs = `r paste(round2(CIwald(RTL_LogAge)['LogAge',]),collapse = ', ')`), and this model was a significantly better fit than one modelling a linear relationship between RTL and age (P = `r round3(anova(RTL_Age,RTL_LogAge)$"Pr(>Chisq)"[2])`), suggesting that telomere length decreased at the greatest rate in early life - something that was evident from the raw data (Fig.
+1A).
 Also worth noting is that while the overall trend was for decreasing RTL, there were appoarent increases in cross-sectional telomere length after one year of age and at several points later in life (Fig.
-1).
+1A).
+Using longitudinal data, we found that RTL decreased with $\Delta$Age (LMM: estimate = `r round3(fixef(RTL_DeltaAge)['DeltaAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_DeltaAge)['DeltaAge',],lessthan = F),collapse = ', ')`), showing that telomere length decreases with age within individuals (Fig.
+1B).
 
-Including birth year as a random effect in a LMM of RTL and (log) age resulted in a significantly improved model to one with only individual ID ans plate ID as random effects (P `r round3(anova(RTL_LogAge,RTL_LogAge_LY)$"Pr(>Chisq)"[2])`; Fig.
-2A).
-Allowing the slope of the age effect to vary among birth years resulted in a slightly significant improvement on one in which only the intercept was allowed to vary (P = `r round3(anova(RTL_LogAge_LY,RTL_LogAge_AgeLY)$"Pr(>Chisq)"[2])`; Fig.
+Including birth year as a random effect in a LMM of RTL and (log) age resulted in a significantly improved model to one with only individual ID and plate ID as random effects (likelihood ratio test, P `r round3(anova(RTL_LogAge,RTL_LogAge_LY)$"Pr(>Chisq)"[2])`).
+Furthermore, allowing the slope of the age effect to vary among birth years resulted in a significant improvement on one in which only the intercept was allowed to vary (conditional R^2^ = `r round2(r.squaredGLMM(RTL_LogAge_AgeLY)['R2c'])`; P = `r round3(anova(RTL_LogAge_LY,RTL_LogAge_AgeLY)$"Pr(>Chisq)"[2])`; Fig.
 2B).
-However, a model including catch year and individual ID as random effects was significantly better than a model indluding birth year and individual ID (P `r round3(anova(RTL_LogAge_LY,RTL_LogAge_CY)$"Pr(>Chisq)"[2])`; Fig 2C) suggesting that variation among catch years was greater than among cohorts (Fig.
-2C), and a model that allowed the slope of the age effect to vary among catch years was better still (P = `r round3(anova(RTL_LogAge_CY,RTL_LogAge_AgeCY)$"Pr(>Chisq)"[2])`; Fig.
-2D).
-Using longitudinal data and $\Delta$Age as an explanatory variable we found a similar pattern - a model including catch year as a random effect was better than that did not allow for temporal variation in RTL (P = `r round3(anova(RTL_DeltaAge,RTL_DeltaAge_CY)$"Pr(>Chisq)"[2])`), and better than one with birth year as a random effect (P `r round3(anova(RTL_DeltaAge_LY,RTL_DeltaAge_CY)$"Pr(>Chisq)"[2])`).
-However a model allowing the slope of the $\Delta$Age effect was not better than one in which only the intercept varied (P = `r round3(anova(RTL_DeltaAge_CY,RTL_DeltaAge_AgeCY)$"Pr(>Chisq)"[2])`).
+Temporal dynamics of longitudinal telomere data were similar to the cross-sectional data.
+A model of RTL and $\Delta$Age including catch year as a random effect was not better than one that did not allow for temporal variation in RTL (P = `r round3(anova(RTL_DeltaAge,RTL_DeltaAge_LY)$"Pr(>Chisq)"[2])`).
+However a model allowing the slope of the $\Delta$Age effect was not better than one in which only the intercept varied (P = `r round3(anova(RTL_DeltaAge,RTL_DeltaAge_AgeLY)$"Pr(>Chisq)"[2])`).
 
@@
 4B).
-The model averaging approach yielded qualitatively identical results to the full LMM, with the same explanatory variables 'significant' in terms of being retained in top models, and having model-averaged confidence intervals not overlapping zero (Table S1; Fig.
+The full model was weak in terms of explanatory power of fixed effects (marginal R^2^ = `r round2(r.squaredGLMM(RTL_Full)['R2m'])`), although including the random effect terms increased this substantially (conditional R^2^ = `r round2(r.squaredGLMM(RTL_Full)['R2c'])`) The model averaging approach yielded qualitatively identical results to the full LMM, with the same explanatory variables 'significant' in terms of being retained in top models, and having model-averaged confidence intervals not overlapping zero (Table S1; Fig.
 S1).
@@
 4A).
-
-No social or ecological environmental variables were significant in a model of longitudinal data with $\Delta$RTL as the response variable (Table 2).
-The model averaging approach suggested that $\Delta$RTL was positively related to age (i.e.
-telomere shortening increased with age), and negatively associated with annual density (i.e.
-more shortening in years with higher population density; Table S2; Figs S2; S3).
+No social or ecological environmental variables were significant predictors of $\Delta$RTL, regardless of whether we interpreted the full model  (Table 2), or used model averaging (Table S2; Fig.
+S2).
 
@@
 
-Perhaps the clearest result from our study is that RTL varies among cohorts.
+Perhaps the clearest result from our study is that RTL, among cohorts, and among sampling years.
 A few studies have now shown that temporal variation in telomere dynamics occurs in natural populations, although these have been limited in the number of seasons [@Mizutani2013; @Watson2015; @Fairlie2016].
 The long-term Seychelles warbler dataset has allowed us to show that temporal variation in telomere dynamics can occur over substantial time periods.
-Moreover, because of the large number of cohorts we have been able to show statistically that that temporal variation in RTL is related to variation in the environment (Fig.
-1C).
-In some systems a negative relationship between population density and RTL may be expected, as increased competition for resources is known to effect juvenile telomere length.
-However, in the Seychelles warbler it is likely that population density acts as an indicator of overall island 'quality' for Seychelles warblers.
-This can clearly be seen from data from the 1980s and 1990s, when adult population size on Cousin increased as habitat quality improved [@Komdeur2005].
-Population size has fluctuated around 300 since the late 1990s (Fig.
-2B), and we suggest that this variation captures this island-wide conditions more accurately than measures of insect abundance or territory quality.
+Moreover, because of the large number of cohorts we have been able to show statistically that that temporal variation in RTL is related to variation in food availability (Fig.
+4B).
 If cohort-level variation in telomere dynamics continues to be found in other systems, this has potential consequences for our understanding of telomere dynamics in natural populations.
@@
 
-Despite substantial cohort-level variation in juvenile RTL, within cohorts we found no evidence that any of the social or environmental variables we measured affect juvenile RTL.
+Despite substantial cohort-level variation in RTL, within cohorts we found no evidence that any of the social or environmental variables we measured affect juvenile RTL.
 This is surprising, as a poor social and ecological environment is known to be detrimental to juvenile Seychelles warblers, both in terms of oxidative stress and later life survival [@VandeCrommenacker2011a; @Brouwer2012].
-A possible explanation for this could be that temporal variation in RTL in our data masks our ability to detect spatial trends.
-However, this is unlikely we controlled for cohort ID, and within-cohort variation in RTL greatly exceeds among-cohort variation (Fig.
+An explanation for this could be that temporal variation in RTL in our data masks our ability to detect spatial trends.
+However, this is unlikely as we controlled for cohort ID, and within-cohort variation in RTL greatly exceeds among-cohort variation (Fig.
 1C).

### 2016-11-21 Manuscript/SW_early_life_telomeres_SOM.Rmd
eds

@@
 source('Scripts/SW juvenile telomeres IPC models.r')
+source('Scripts/SW juvenile telomeres IPC model averaging.r')
 
@@
 Only models that remained in the top model set for model averaging (AICc $\leq$ 6 compared to the top model) are included here.
-TQ = territory quality.
 
@@
 
-t1 <- as.matrix(subset(lm3, delta<=6))
+t1 <- as.matrix(RTL_Full_Dredge)
 
@@
 nams[nams == 'z.Density'] <- 'Density'
+nams[nams == 'z.TQ'] <- 'Territory quality'
 
@@
 
-tableS1 <- data.frame(Model = apply(t1,1,getterms,c(2:14)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
+tableS1 <- data.frame(Model = apply(t1,1,getterms,c(2:10)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
 colnames(tableS1) <- c('Model','df','AICc', 'Delta AICc','Weight')
@@
 #Model averaging
-lm3 <- dredge(DeltaRTL_Full)
-t1 <- as.matrix(subset(lm3, delta<=6))
+t1 <- as.matrix(DeltaRTL_Full_Dredge)
 
@@
 
-tableS1 <- data.frame(Model = apply(t1,1,getterms,c(2:14)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
-colnames(tableS1) <- c('Model','df','AICc', 'Delta AICc','Weight')
-rownames(tableS1) <- NULL
-tableS1$Model[tableS1$Model == ''] <- 'Null model'
-tableS1[,3:5] <- lapply(tableS1[,3:5],as.numeric)
-kable(tableS1,digits=2)
+tableS2 <- data.frame(Model = apply(t1,1,getterms,c(2:10)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
+colnames(tableS2) <- c('Model','df','AICc', 'Delta AICc','Weight')
+rownames(tableS2) <- NULL
+tableS2$Model[tableS2$Model == ''] <- 'Null model'
+tableS2[,3:5] <- lapply(tableS2[,3:5],as.numeric)
+kable(tableS2,digits=2)
 
@@
 
-**Figure S1** $\Delta$RTL in relation to the time between sampling events.
-Lines and shaded areas represent fitted values and 95% confidence limits from general linear models.
-See also Figure 4, main text.
+**Figure S1** Model averaged estimates and 95% cofidence intervals for social and ecological environmental variables in relation to RTL in Seychelles warblers.
 
 ```{r Figure S1}
-temp <- subset(dd,!is.na(Insect))
-temp$LogAge <- log(temp$Agemonths)
-
-#Run the model
-lm2 <- lmer(RTL ~  LogAge + Tarsus + Helper + GroupSize + Sex + BodyMass + Insect + Density + (1|LayYear) + (1|BirdID) + (1|PlateID),
-            data = temp,
-            REML = FALSE)
-lm2 <- standardize(lm2)
 
 #Model averaging
-lm3 <- dredge(lm2)
-lm3 <- subset(lm3,delta <=6)
-lmavg <- model.avg(lm3,fit=T)
+lmavg <- model.avg(RTL_Full_Dredge,fit=T)
 myimp <- importance(lmavg) #Relative importance of each explanatory variable
@@
 
+**Figure S2** Model averaged estimates and 95% cofidence intervals for social and ecological environmental variables in relation to $\Delta$RTL in Seychelles warblers, using longitudinal data.
+
 ```{r Figure S2}
 
-#Run the model
-temp <- subset(dd3,!is.na(Insect))
-
-#Run the model
-lm2 <- lmer(DeltaRTL ~  LogAge + Tarsus + Helper + GroupSize + Sex + BodyMass + Insect + Density + (1|LayYear) + (1|PlateID),
-            data = temp,
-            REML = FALSE)
-lm2 <- standardize(lm2)
-
 #Model averaging
-lm3 <- dredge(lm2)
-lm3 <- subset(lm3,delta <=6)
-lmavg <- model.avg(lm3,fit=T)
+lmavg <- model.avg(DeltaRTL_Full_Dredge,fit=T)
 myimp <- importance(lmavg) #Relative importance of each explanatory variable
 
-#Put it all into a nice table
+#Put it all into a table
 table2 <- imptable(lmavg,myimp)

### 2016-11-21 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
eds

@@
 
-**Table1**
+**Table 1**
 ```{r Table 1}
 
+AIC(RTL_LogAge)
+AIC(RTL_LogAge_LY)
+
+```
+
+**Table 2**
+```{r Table 2}
+
 CIs <- CIwald(RTL_Full)
-t1 <- data.frame(Estimate = fixef(RTL_Full), LCI = CIs[6:nrow(CIs),1], UCI = CIs[6:nrow(CIs),2])
+t1 <- data.frame(Estimate = fixef(RTL_Full), LCI = CIs[7:nrow(CIs),1], UCI = CIs[7:nrow(CIs),2])
+
+nams <- row.names(t1)
+nams[nams == 'LogAge'] <- 'Age'
+nams[nams == 'Tarsus'] <- 'Tarsus length'
+nams[nams == 'Helper'] <- 'Number helpers'
+nams[nams == 'GroupSize'] <- 'Group size'
+nams[nams == 'SexMales'] <- 'Sex (males)'
+nams[nams == 'BodyMass'] <- 'Body mass'
+nams[nams == 'Insect'] <- 'Annual food availability'
+nams[nams == 'TQ'] <- 'Territory quality'
+
+row.names(t1) <- nams
+
 kable(t1)
@@
 
-temp <- subset(dd3,!is.na(Insect))
-
 #Run the model
@@
 t1 <- data.frame(Estimate = fixef(DeltaRTL_Full), LCI = CIs[4:nrow(CIs),1], UCI = CIs[4:nrow(CIs),2])
+
+nams <- row.names(t1)
+nams[nams == 'LogAge'] <- 'Age'
+nams[nams == 'Tarsus'] <- 'Tarsus length'
+nams[nams == 'Helper'] <- 'Number helpers'
+nams[nams == 'GroupSize'] <- 'Group size'
+nams[nams == 'SexMales'] <- 'Sex (males)'
+nams[nams == 'BodyMass'] <- 'Body mass'
+nams[nams == 'Insect'] <- 'Annual food availability'
+nams[nams == 'TQ'] <- 'Territory quality'
+
+row.names(t1) <- nams
+
 kable(t1)
@@
 
-Fig1 <- ddply(dd,
+Fig1A <- ddply(dd,
                .(BirdID),
@@
         legend.position = 'none')+
-  geom_smooth(col = 'black',fill = 'black',lwd = 0.5,aes(group = NULL),method = 'lm',formula = y~log(x))
-  annotate('text',x = 0,y = 3,label = 'A')
+  geom_smooth(col = 'black',fill = 'black',lwd = 0.5,aes(group = NULL),method = 'lm',formula = y~log(x))+
+  annotate('text',x = 0,y = 1.7,label = 'A')
+
+  Fig1B <- ggplot(dd3, aes(x = DeltaAge,y = RTL))+
+  geom_point(alpha = 1,col='grey')+
+  xlab(expression(Delta*'Age (years)')) +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'),
+        legend.position = 'none')+
+  geom_smooth(col = 'black',method = 'lm')+
+  annotate('text',x = -7.5,y = 1.7,label = 'B')
 
@@
     width = 4,
-    height = 4,
+    height = 8,
     paper = 'a4')
-Fig1
+multiplot(Fig1A,Fig1B)
 dev.off()
@@
 ```{r Figure 2}
+c15 <- c("dodgerblue2","#E31A1C", # red
+                "green4",
+                "#FF7F00", # orange
+                "gold1",
+                "skyblue2", # lt pink
+                "palegreen2",
+                "#FDBF6F", # lt orange
+                "gray70", "khaki2","blue1","steelblue4",
+                "darkturquoise",
+                "darkorange4","brown")
 
-Fig2A <- transform(juv,Cohort = factor(CatchYear)) %>%
-  subset(Agemonths>(1/12)) %>%
-  ggplot(aes(y = RTL,x = Cohort))+
-  xlab('Cohort') +
-  geom_boxplot() +
+Fig2A <- transform(dd,Cohort = factor(LayYear)) %>%
+  subset(Cohort %in% c(2002:2011)) %>%
+  ggplot(aes(y = RTL,x = Cohort,fill = Cohort))+
+  xlab('Birth year') +
+  geom_boxplot(notch=T) +
   theme_classic() +
@@
         axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black'))
-  annotate('text',x = 0,y = 1.4,label = 'B')
+        axis.line.y = element_line(colour = 'black'),
+        legend.position = 'none')+
+  scale_fill_manual(values = c15)+
+  annotate('text',x = 1,y = 1.7,label = 'A')
 
-Fig2B <- transform(dd,Cohort = factor(CatchYear)) %>%
-  ddply(.(Cohort),
-        mutate,
-        nc = length(which(Agemonths ==1/12))) %>%
-  subset(nc > 11) %>%
+Fig2B <- transform(dd,Cohort = factor(LayYear)) %>%
+  subset(Cohort %in% c(2002:2011)) %>%
   ggplot(aes(y = RTL,x = Agemonths, col = Cohort))+
@@
         axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black'))
-  annotate('text',x = 0,y = 1.4,label = 'B')
+        axis.line.y = element_line(colour = 'black'),
+        legend.position = 'none') +
+    scale_color_manual(values = c15)+
+  annotate('text',x = 0,y = 1.15,label = 'B')
+
+Fig2C <- transform(dd,Cohort = factor(CatchYear)) %>%
+  subset(Cohort %in% c(2002:2011)) %>%
+  ggplot(aes(y = RTL,x = Cohort,fill = Cohort))+
+  xlab('Sampling year') +
+  geom_boxplot(notch=T) +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12,angle = 90), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'),
+        legend.position = 'none') + 
+    scale_fill_manual(values = c15)+
+  annotate('text',x = 1,y = 1.7,label = 'C')
+
+Fig2D <- transform(dd,Cohort = factor(CatchYear)) %>%
+  subset(Cohort %in% c(2002:2011)) %>%
+  ggplot(aes(y = RTL,x = Agemonths, col = Cohort))+
+  xlab('Age (years)') +
+  geom_smooth(method = 'lm', formula = y~log(x),se = F) +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'),
+        legend.position = 'none') +
+    scale_color_manual(values = c15)+
+  annotate('text',x = 0,y = 1.15,label = 'D')
 
 pdf('Figures/Figure 2.pdf',
-    width = 5,
+    width = 8,
     height = 8,
     paper = 'a4')
-multiplot(Fig2A,Fig2B)
+multiplot(Fig2A,Fig2B,Fig2C,Fig2D,layout = matrix(1:4, 2, 2,byrow = T))
+dev.off()
+
+```
+
+```{r Figure 2}
+
+c15 <- c("dodgerblue2","#E31A1C", # red
+                "green4",
+                "#FF7F00", # orange
+                "gold1",
+                "skyblue2", # lt pink
+                "palegreen2",
+                "#FDBF6F", # lt orange
+                "gray70", "khaki2","blue1","steelblue4",
+                "darkturquoise",
+                "darkorange4","brown")
+
+Fig2A <- transform(dd,Cohort = factor(LayYear)) %>%
+  subset(Cohort %in% c(2002:2011)) %>%
+  ggplot(aes(y = RTL,x = Cohort,fill = Cohort))+
+  xlab('Birth year') +
+  geom_boxplot(notch=T) +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12,angle = 90), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'),
+        legend.position = 'none')+
+  scale_fill_manual(values = c15)+
+  annotate('text',x = 1,y = 1.7,label = 'A')
+
+Fig2B <- transform(dd3,Cohort = factor(LayYear)) %>%
+  subset(Cohort %in% c(2002:2011)) %>%
+  ggplot(aes(y = RTL,x = DeltaAge, col = Cohort))+
+  xlab('Age (years)') +
+  geom_smooth(method = 'lm', formula = y~(x),se = F) +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'),
+        legend.position = 'none') +
+    scale_color_manual(values = c15)+
+  annotate('text',x = 0,y = 1.15,label = 'B')
+
+Fig2C <- transform(dd,Cohort = factor(CatchYear)) %>%
+  subset(Cohort %in% c(2002:2011)) %>%
+  ggplot(aes(y = RTL,x = Cohort,fill = Cohort))+
+  xlab('Sampling year') +
+  geom_boxplot(notch=T) +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12,angle = 90), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'),
+        legend.position = 'none') + 
+    scale_fill_manual(values = c15)+
+  annotate('text',x = 1,y = 1.7,label = 'C')
+
+Fig2D <- transform(dd,Cohort = factor(CatchYear)) %>%
+  subset(Cohort %in% c(2002:2011)) %>%
+  ggplot(aes(y = RTL,x = DeltaAge, col = Cohort))+
+  xlab('Age (years)') +
+  geom_smooth(method = 'lm', formula = y~(x),se = F) +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'),
+        legend.position = 'none') +
+    scale_color_manual(values = c15)+
+  annotate('text',x = 0,y = 1.15,label = 'D')
+
+pdf('Figures/Figure 2.pdf',
+    width = 8,
+    height = 8,
+    paper = 'a4')
+multiplot(Fig2A,Fig2B,Fig2C,Fig2D,layout = matrix(1:4, 2, 2,byrow = T))
 dev.off()
@@
     geom_point(col = 'grey')+
-  geom_abline(slope = 1,intercept = 0,lty = 2)
-  annotate('text',x = 0,y = 3,label = 'B')
+  geom_abline(slope = 1,intercept = 0,lty = 2) +
+  annotate('text',x = 0.3,y = 1.7,label = 'A')
 
@@
   scale_fill_manual(values = c('darkgrey','gold'))+
-  geom_vline(xintercept = 0,lty = 2)
-  annotate('text',x = -2,y = 1,label = 'C')
+  geom_vline(xintercept = 0,lty = 2) +
+  annotate('text',x = -1,y = 1,label = 'B')
 
 pdf('Figures/Figure 3.pdf',
-    width = 5,
+    width = 4,
     height = 8,
@@
 
-Fig4A <- ggplot(temp,
+Fig4A <- ggplot(dd,
        aes(x = Tarsus,
@@
         legend.position = 'none') +
-  stat_smooth(method = 'lm')
-  annotate('text',x = 17,y = 2.5,label = 'C')
+  stat_smooth(method = 'lm') +
+  scale_color_manual(values = c('gold2','red'))+
+  annotate('text',x = 17,y = 1.7,label = 'A')
 
-Fig4B <- ggplot(temp,
-       aes(
-           y = RTL,
+Fig4B <- ggplot(dd,
+       aes(y = RTL,
            x=Insect)) +
@@
   geom_boxplot(aes(group = Insect),outlier.shape = NA)+
-   stat_smooth(method = 'lm',col = 'darkgrey')
-  annotate('text',x = 2.5,y = 2.5,label = 'D')
+   stat_smooth(method = 'lm',col = 'darkgrey') +
+  annotate('text',x = 2.5,y = 1.6,label = 'B')
 

### 2016-11-21 Scripts/SW juvenile telomeres IPC clean.R

@@ -283,7 +283,6 @@ for(i in 1:nrow(dd3))
     dd3$RemainingLife1[i] <- nextbird$RemainingLife
     dd3$Died1[i] <- nextbird$Died
     dd3$DeltaCondition[i] <- nextbird$Condition - dd3$Condition[i]
-    dd3$DeltaAge[i] <- nextbird$Agemonths - dd3$Agemonths[i]
     dd3$DeltaInsect[i] <- nextbird$Insect - dd3$Insect[i]
     dd3$Agemonths1[i] <- nextbird$Agemonths
   }

### 2016-11-21 Scripts/SW juvenile telomeres IPC model averaging.R

@@ -1,0 +1,10 @@
+#################################################################################*
+## Telomere dynamics in a wild bird population
+## PERFORM MODEL AVERAGING
+#################################################################################*
+
+RTL_Full <- standardize(RTL_Full)
+RTL_Full_Dredge <- subset(dredge(RTL_Full),delta <= 6)
+
+DeltaRTL_Full <- standardize(DeltaRTL_Full)
+DeltaRTL_Full_Dredge <- subset(dredge(DeltaRTL_Full),delta <= 6)

### 2016-11-21 Scripts/SW juvenile telomeres IPC models.r

@@ -22,15 +22,20 @@ RTL_DeltaAge_AgeCY <- lmer(RTL~DeltaAge + (1|BirdID) + (DeltaAge|CatchYear),data
 
 # RTL and ecology ---------------------------------------------------------
 
-RTL_Full <- lmer(RTL ~  LogAge + Tarsus + Helper + GroupSize + Sex + BodyMass + Insect + Density + (Agemonths|CatchYear) + (1|BirdID) + (1|PlateID),
-                 data = subset(dd,!is.na(Insect)),
+ddNA <- subset(subset(dd,!is.na(Insect)),!is.na(TQ))
+dd3NA <- subset(subset(dd3,!is.na(Insect)),!is.na(TQ))
+
+RTL_Full <- lmer(RTL ~  LogAge + Tarsus + Helper + GroupSize + Sex + BodyMass + Insect + Density + TQ + (Agemonths|CatchYear) + (1|BirdID) + (1|PlateID),
+                 data = ddNA,
                  REML = FALSE)
-RTL_Full_NoTarsus <- lmer(RTL ~  LogAge + Helper + GroupSize + Sex + BodyMass + Insect + Density + (Agemonths|CatchYear) + (1|BirdID) + (1|PlateID),
-                          data = subset(dd,!is.na(Insect)),
+RTL_Full_NoTarsus <- lmer(RTL ~  LogAge + Helper + GroupSize + Sex + BodyMass + Insect + Density + TQ + (Agemonths|CatchYear) + (1|BirdID) + (1|PlateID),
+                          data = ddNA,
                           REML = FALSE)
-RTL_Full_SexTarsus <- lmer(RTL ~  LogAge + Helper + GroupSize + Sex + BodyMass + Insect + Density + Sex*Tarsus + (Agemonths|CatchYear) + (1|BirdID) + (1|PlateID),
-                          data = subset(dd,!is.na(Insect)),
+RTL_Full_SexTarsus <- lmer(RTL ~  LogAge + Helper + GroupSize + Sex + BodyMass + Insect + Density + Sex*Tarsus + TQ + (Agemonths|CatchYear) + (1|BirdID) + (1|PlateID),
+                          data = ddNA,
                           REML = FALSE)
-DeltaRTL_Full <- lmer(DeltaRTL ~  LogAge + Tarsus + Helper + GroupSize + Sex + BodyMass + Insect + Density + (1|CatchYear) + (1|BirdID),
[…]

### 2016-11-22 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Eds

@@
   xlab('Birth year') +
-  geom_boxplot(notch=T) +
+  geom_boxplot(notch=T,aes(outlier.colour = Cohort)) +
   theme_classic() +
@@
   subset(Cohort %in% c(2002:2011)) %>%
-  ggplot(aes(y = RTL,x = DeltaAge, col = Cohort))+
+  ggplot(aes(y = RTL,x = Agemonths, col = Cohort))+
   xlab('Age (years)') +
-  geom_smooth(method = 'lm', formula = y~(x),se = F) +
+  geom_smooth(method = 'lm', formula = y~log(x),se = F) +
   theme_classic() +
@@
 
-Fig2C <- transform(dd,Cohort = factor(CatchYear)) %>%
-  subset(Cohort %in% c(2002:2011)) %>%
-  ggplot(aes(y = RTL,x = Cohort,fill = Cohort))+
-  xlab('Sampling year') +
-  geom_boxplot(notch=T) +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12,angle = 90), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black'),
-        legend.position = 'none') + 
-    scale_fill_manual(values = c15)+
-  annotate('text',x = 1,y = 1.7,label = 'C')
-
-Fig2D <- transform(dd,Cohort = factor(CatchYear)) %>%
-  subset(Cohort %in% c(2002:2011)) %>%
-  ggplot(aes(y = RTL,x = DeltaAge, col = Cohort))+
-  xlab('Age (years)') +
-  geom_smooth(method = 'lm', formula = y~(x),se = F) +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black'),
-        legend.position = 'none') +
-    scale_color_manual(values = c15)+
-  annotate('text',x = 0,y = 1.15,label = 'D')
-
 pdf('Figures/Figure 2.pdf',
-    width = 8,
+    width = 4,
     height = 8,
     paper = 'a4')
-multiplot(Fig2A,Fig2B,Fig2C,Fig2D,layout = matrix(1:4, 2, 2,byrow = T))
+multiplot(Fig2A,Fig2B)
 dev.off()
@@
 ```
+
+dd$FAge <- factor(dd$Agemonths)
+dd$Cohort <- factor(dd$LayYear)
+
+lm1 <- lmer(RTL ~ Cohort*(Agemonths + LogAge + FAge) + (1|BirdID) + (1|CatchYear) + (1|PlateID),data=dd,REML = FALSE)
+
+lm2 <- dredge(lm1,subset = !(LogAge & FAge) & !(LogAge & Agemonths ) & !(Agemonths & FAge))
+
+lm2

### 2016-11-22 Scripts/SW juvenile telomeres IPC clean.R

@@ -78,7 +78,6 @@ dd$Agemonths <- ifelse(dd$Ageclass == 'CH',1,
                               ifelse(dd$Ageclass == 'SA',10,dd$Age*12+6)))/12
 
 
-  #dd$Age[dd$Fledged != 'Adults'] <- 1
 dd$LogAge <- log10(dd$Agemonths)
 dd <- subset(dd,!is.na(LogAge))
 
@@ -129,31 +128,6 @@ dd$Sex <- ifelse(dd$SexEstimate == 1,'Males','Females')
 
 
 
-
-
-
-#Centre telomere length by birth year
-dd$cenTL <- NA
-dd$cohortTL <- NA
-for(i in 1:nrow(dd))
-{
-  currentdata <- subset(dd,FieldPeriodID == FieldPeriodID[i])
-  if(nrow(currentdata>4))
-  {
-    dd$cenTL[i] <- (dd$RTL[i] - mean(currentdata$RTL))/sd(currentdata$RTL)
-    dd$cohortTL[i] <- median(currentdata$RTL)  
-  }
-
-}
-
-mymed <- median(dd$cenTL,na.rm=T)
-dd$cenTLF <- ifelse(dd$cenTL < mymed,'Short telomeres','Long telomeres')
-
-mymed <- median(dd$cohortTL,na.rm=T)
-dd$coTLF <- ifelse(dd$cohortTL < mymed,'Short telomeres','Long telomeres')
-
-
-rm(mymed)
 
 # Helpers and social group size -------------------------------------------
 

### 2016-11-22 Scripts/SW juvenile telomeres IPC models.r

@@ -9,15 +9,11 @@
 RTL_Age <- lmer(RTL~Agemonths + (1|BirdID) + (1|PlateID),data = dd)
 RTL_LogAge <- lmer(RTL~LogAge + (1|BirdID) + (1|PlateID),data = dd)
 RTL_LogAge_LY <- lmer(RTL~LogAge + (1|BirdID) + (1|PlateID) + (1|LayYear),data = dd)
-RTL_LogAge_CY <- lmer(RTL~LogAge + (1|BirdID) + (1|PlateID) + (1|CatchYear),data = dd)
 RTL_LogAge_AgeLY <- lmer(RTL~LogAge + (1|BirdID) + (1|PlateID) + (Agemonths|LayYear),data = dd)
-RTL_LogAge_AgeCY <- lmer(RTL~LogAge + (1|BirdID) + (1|PlateID) + (Agemonths|CatchYear),data = dd)
 
 RTL_DeltaAge <- lmer(RTL~DeltaAge + (1|BirdID),data = dd3)
 RTL_DeltaAge_LY <- lmer(RTL~DeltaAge + (1|BirdID) + (1|LayYear),data = dd3)
-RTL_DeltaAge_CY <- lmer(RTL~DeltaAge + (1|BirdID) + (1|CatchYear),data = dd3)
 RTL_DeltaAge_AgeLY <- lmer(RTL~DeltaAge + (1|BirdID) + (DeltaAge|LayYear),data = dd3)
-RTL_DeltaAge_AgeCY <- lmer(RTL~DeltaAge + (1|BirdID) + (DeltaAge|CatchYear),data = dd3)
 
 
 # RTL and ecology ---------------------------------------------------------

### 2016-11-23 Manuscript/SW_early_life_telomeres.Rmd
EDS

@@
 #####Telomere dynamics and age
-Using cross-sectional data we found that RTL decreased with age in the Seychelles warbler (LMM: estimate = `r round3(fixef(RTL_Age)['Agemonths'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Age)['Agemonths',],lessthan = F),collapse = ', ')`).
-RTL also decreased with log-transformed age (LMM: estimate = `r round2(fixef(RTL_LogAge)['LogAge'])`, CIs = `r paste(round2(CIwald(RTL_LogAge)['LogAge',]),collapse = ', ')`), and this model was a significantly better fit than one modelling a linear relationship between RTL and age (P = `r round3(anova(RTL_Age,RTL_LogAge)$"Pr(>Chisq)"[2])`), suggesting that telomere length decreased at the greatest rate in early life - something that was evident from the raw data (Fig.
-1A).
-Also worth noting is that while the overall trend was for decreasing RTL, there were appoarent increases in cross-sectional telomere length after one year of age and at several points later in life (Fig.
-1A).
-Using longitudinal data, we found that RTL decreased with $\Delta$Age (LMM: estimate = `r round3(fixef(RTL_DeltaAge)['DeltaAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_DeltaAge)['DeltaAge',],lessthan = F),collapse = ', ')`), showing that telomere length decreases with age within individuals (Fig.
-1B).
-
-Including birth year as a random effect in a LMM of RTL and (log) age resulted in a significantly improved model to one with only individual ID and plate ID as random effects (likelihood ratio test, P `r round3(anova(RTL_LogAge,RTL_LogAge_LY)$"Pr(>Chisq)"[2])`).
-Furthermore, allowing the slope of the age effect to vary among birth years resulted in a significant improvement on one in which only the intercept was allowed to vary (conditional R^2^ = `r round2(r.squaredGLMM(RTL_LogAge_AgeLY)['R2c'])`; P = `r round3(anova(RTL_LogAge_LY,RTL_LogAge_AgeLY)$"Pr(>Chisq)"[2])`; Fig.
-2B).
-Temporal dynamics of longitudinal telomere data were similar to the cross-sectional data.
-A model of RTL and $\Delta$Age including catch year as a random effect was not better than one that did not allow for temporal variation in RTL (P = `r round3(anova(RTL_DeltaAge,RTL_DeltaAge_LY)$"Pr(>Chisq)"[2])`).
-However a model allowing the slope of the $\Delta$Age effect was not better than one in which only the intercept varied (P = `r round3(anova(RTL_DeltaAge,RTL_DeltaAge_AgeLY)$"Pr(>Chisq)"[2])`).
+Model comparison
 

### 2016-11-23 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
EDS

@@
 
-AIC(RTL_LogAge)
-AIC(RTL_LogAge_LY)
+t1 <- as.matrix(RTL_Age_Dredge)
+
+#Sort out row/col names
+nams <- colnames(t1)
+
+nams[nams == 'Agemonths'] <- 'Age (continuous)'
+nams[nams == 'LogAge'] <- 'Log age'
+nams[nams == 'FAge'] <- 'Age (factor)'
+nams[nams == 'Cohort:FAge'] <- 'Age (factor)*cohort'
+nams[nams == 'Cohort:Agemonths'] <- 'Age (continuous)*cohort'
+nams[nams == 'Cohort:LogAge'] <- 'Log age*cohort'
+
+colnames(t1) <- nams
+
+tableS1 <- data.frame(Model = apply(t1,1,getterms,c(2:8)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
+colnames(tableS1) <- c('Model','df','AICc', 'Delta AICc','Weight')
+rownames(tableS1) <- NULL
+tableS1$Model[tableS1$Model == ''] <- 'Null model'
+tableS1[,3:5] <- lapply(tableS1[,3:5],as.numeric)
+kable(tableS1,digits=3)
 
@@
 ```
-
-dd$FAge <- factor(dd$Agemonths)
-dd$Cohort <- factor(dd$LayYear)
-
-lm1 <- lmer(RTL ~ Cohort*(Agemonths + LogAge + FAge) + (1|BirdID) + (1|CatchYear) + (1|PlateID),data=dd,REML = FALSE)
-
-lm2 <- dredge(lm1,subset = !(LogAge & FAge) & !(LogAge & Agemonths ) & !(Agemonths & FAge))
-
-lm2

### 2016-11-23 Scripts/SW juvenile telomeres IPC clean.R

@@ -326,7 +326,8 @@ rm(temp,dd3_2)
 
 
 
-
+dd$Cohort <- factor(dd$LayYear)
+dd$FAge <- factor(dd$Age)
 
 # Subset birds with juvenile samples --------------------------------------
 

### 2016-11-23 Scripts/SW juvenile telomeres IPC model averaging.R

@@ -3,6 +3,9 @@
 ## PERFORM MODEL AVERAGING
 #################################################################################*
 
+RTL_Age_Dredge <- dredge(RTL_Age,
+                         subset = !(LogAge & FAge) & !(LogAge & Agemonths ) & !(Agemonths & FAge))
+
 RTL_Full <- standardize(RTL_Full)
 RTL_Full_Dredge <- subset(dredge(RTL_Full),delta <= 6)
 

### 2016-11-23 Scripts/SW juvenile telomeres IPC models.r

@@ -6,15 +6,9 @@
 
 # Telomere length, age and cohort -----------------------------------------
 
-RTL_Age <- lmer(RTL~Agemonths + (1|BirdID) + (1|PlateID),data = dd)
-RTL_LogAge <- lmer(RTL~LogAge + (1|BirdID) + (1|PlateID),data = dd)
-RTL_LogAge_LY <- lmer(RTL~LogAge + (1|BirdID) + (1|PlateID) + (1|LayYear),data = dd)
-RTL_LogAge_AgeLY <- lmer(RTL~LogAge + (1|BirdID) + (1|PlateID) + (Agemonths|LayYear),data = dd)
-
-RTL_DeltaAge <- lmer(RTL~DeltaAge + (1|BirdID),data = dd3)
-RTL_DeltaAge_LY <- lmer(RTL~DeltaAge + (1|BirdID) + (1|LayYear),data = dd3)
-RTL_DeltaAge_AgeLY <- lmer(RTL~DeltaAge + (1|BirdID) + (DeltaAge|LayYear),data = dd3)
-
+RTL_Age <- lmer(RTL ~ Cohort*(Agemonths + LogAge + FAge) + (1|BirdID) + (1|CatchYear) + (1|PlateID),
+                data=dd,
+                REML = FALSE)
 
 # RTL and ecology ---------------------------------------------------------
 

### 2016-11-23 Manuscript/SW_early_life_telomeres.Rmd
Home eds

@@
 We performed all statistical analyses using R.
-We first tested the relationship between RTL and age using linear mixed models (LMMs) with individual ID as a random effect, carried out in the lme4 package [@Bates2014].
-To test whether the decrease in telomere length with age is greatest in early life, we compared LMMs with age (in months) and log-transformed age, using a likelihood ratio test.
-We also tested whether telomores shorten with age in individuals with samples from multiple timepoints, using a LMM of RTL and $\Delta$age [a longitudinal measure based on within-subject centring; @VandePol2009].
-To test how telomere length, and its relationship with age, varied among cohorts, we compared the above LMMs with models that included birth year and catch year as additional random effects.
-We also compared a model in which only the intercept (i.e.
-RTL) was allowed to vary among among cohorts or years, to one in which the slope of the relationship between RTL and age was also allowed to vary.
+We first explored the relationship between RTL and age among cohorts using linear mixed models (LMMs) carried out in the lme4 package [@Bates2014].
+Following a similar approach to Fairlie *et al.* [-@Fairlie2016], we compared a selection of models fitting different relationships between RTL and age.
+We created models where age was fitted as a linear covariate, where age was log-transformed, and where age was fitted as a factor.
+We also created models including birth year (cohort) as a factor, and ones which included interactions between cohort and each measure of age.
+All fitted models are included in Table 1.
+As random effects we included individual ID, catch year and qPCR plate ID.
+Models were compared using AIC.
 
+Using the longitudinal data, we tested whether telomores shorten with age in individuals, using a LMM of RTL and $\Delta$age [a longitudinal measure based on within-subject centring; @VandePol2009].
 Because we observed apparent telomere lengthening in our data, we used the entire Seychelles warbler dataset to test whether this lengthening could be explained by measurement error.

### 2016-11-23 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Home eds

--- 2016-11-22 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
@@
 ```{r Table 1}
+RTL_Age_Dredge <- dredge(RTL_Age,
+                         subset = !(LogAge & FAge) & 
+                           !(LogAge & Agemonths) &
+                           !(LogAge & Age2) &
+                           !(Agemonths & FAge) & 
+                           !(Agemonths & Age2) &
+                           !(FAge & Age2))
 
-AIC(RTL_LogAge)
-AIC(RTL_LogAge_LY)
+t1 <- as.matrix(RTL_Age_Dredge)
 
+#Sort out row/col names
+nams <- colnames(t1)
+
+nams[nams == 'Agemonths'] <- 'Age (linear)'
+nams[nams == 'LogAge'] <- 'Age (log)'
+nams[nams == 'FAge'] <- 'Age (factor)'
+nams[nams == 'Cohort:FAge'] <- 'Age (factor)*cohort'
+nams[nams == 'Agemonths:Cohort'] <- 'Age (continuous)*cohort'
+nams[nams == 'Cohort:LogAge'] <- 'Age (log)*cohort'
+nams[nams == 'Age2:Cohort'] <- 'Age (quadratic)*cohort'
+nams[nams == 'Age2'] <- 'Age (quadratic)'
+
+colnames(t1) <- nams
+
+tableS1A <- data.frame(Model = apply(t1,1,getterms,c(2:10)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
+colnames(tableS1A) <- c('Model','df','AICc', 'Delta AICc','Weight')
+rownames(tableS1A) <- NULL
+tableS1A$Model[tableS1A$Model == ''] <- 'Null model'
+tableS1A[,3:5] <- lapply(tableS1A[,3:5],as.numeric)
+
+RTL_DeltaAge_Dredge <- dredge(RTL_DeltaAge,
+                         subset = !(DeltaLogAge & DeltaAge) &
+                           !(DeltaLogAge & DeltaAge2) &
+                           !(DeltaAge & DeltaAge2))
+
+t1 <- as.matrix(RTL_DeltaAge_Dredge)
+
+#Sort out row/col names
+nams <- colnames(t1)
+
+nams[nams == 'DeltaAge'] <- 'Delta age (linear)'
+nams[nams == 'DeltaLogAge'] <- 'Delta age (log)'
+nams[nams == 'Cohort:DeltaAge'] <- 'Delta age (continuous)*cohort'
+nams[nams == 'Cohort:DeltaLogAge'] <- 'Delta age (log)*cohort'
+nams[nams == 'Cohort:DeltaAge2'] <- 'Delta age (quadratic)*cohort'
+nams[nams == 'DeltaAge2'] <- 'Delta age (quadratic)'
+
+colnames(t1) <- nams
+
+tableS1B <- data.frame(Model = apply(t1,1,getterms,c(2:8)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
+colnames(tableS1B) <- c('Model','df','AICc', 'Delta AICc','Weight')
+rownames(tableS1B) <- NULL
+tableS1B$Model[tableS1B$Model == ''] <- 'Null model'
+tableS1B[,3:5] <- lapply(tableS1B[,3:5],as.numeric)
+
+tableS1 <- rbind(c('A',rep(' ',4)),tableS1A,c('--',rep(' ',4)),c('B',rep(' ',4)),tableS1B)
+
+kable(tableS1,digits = 3)
 ```
@@
 CIs <- CIwald(RTL_Full)
-t1 <- data.frame(Estimate = fixef(RTL_Full), LCI = CIs[7:nrow(CIs),1], UCI = CIs[7:nrow(CIs),2])
+t1a <- data.frame(Variable = row.names(CIs)[7:nrow(CIs)],
+                  Estimate = fixef(RTL_Full),
+                  LCI = CIs[7:nrow(CIs),1],
+                  UCI = CIs[7:nrow(CIs),2],
+                  row.names = NULL,
+                  stringsAsFactors = F)
 
-nams <- row.names(t1)
+nams <- t1a$Variable
 nams[nams == 'LogAge'] <- 'Age'
@@
 
-row.names(t1) <- nams
-
-kable(t1)
-
-```
-
-**Table 2**
-
-```{r Table 2}
+t1a$Variable <- nams
 
@@
 CIs <- CIwald(DeltaRTL_Full)
-t1 <- data.frame(Estimate = fixef(DeltaRTL_Full), LCI = CIs[4:nrow(CIs),1], UCI = CIs[4:nrow(CIs),2])
+t1b <- data.frame(Variable = row.names(CIs)[4:nrow(CIs)],
+                  Estimate = fixef(DeltaRTL_Full),
+                  LCI = CIs[4:nrow(CIs),1],
+                  UCI = CIs[4:nrow(CIs),2],
+                  row.names = NULL,
+                  stringsAsFactors = F)
 
-nams <- row.names(t1)
+nams <- t1b$Variable
 nams[nams == 'LogAge'] <- 'Age'
@@
 
-row.names(t1) <- nams
+t1b$Variable <- nams
 
-kable(t1)
+tableS2 <- rbind(c('A',rep(' ',4)),t1a,c('--',rep(' ',4)),c('B',rep(' ',4)),t1b)
+
+kable(tableS2)
 

### 2016-11-23 Scripts/SW juvenile telomeres IPC clean.R

@@ -80,7 +80,9 @@ dd$Agemonths <- ifelse(dd$Ageclass == 'CH',1,
 
 dd$LogAge <- log10(dd$Agemonths)
 dd <- subset(dd,!is.na(LogAge))
-
+dd$Cohort <- factor(dd$LayYear)
+dd$FAge <- factor(dd$Age)
+dd$Age2 <- poly(dd$Agemonths,2)[,2]
 
 # Tarsus and delta age ------------------------------------------------------------------
 
@@ -96,7 +98,11 @@ for(i in 1:nrow(dd))
          dd$Tarsus[i] <- dd$RightTarsus[i])
   
   dd$DeltaAge[i] <- dd$Agemonths[i] - mean(subset(dd,BirdID == dd$BirdID[i])$Agemonths)
+  dd$DeltaLogAge[i] <- dd$LogAge[i] - mean(subset(dd,BirdID == dd$BirdID[i])$LogAge)
+  dd$DeltaAge2[i] <- dd$Age2[i] - mean(subset(dd,BirdID == dd$BirdID[i])$Age2)
 }
+
+dd$FDeltaAge <- factor(dd$DeltaAge)
 
 dd$RightTarsus <- NULL
 dd$LeftTarsus <- NULL
@@ -324,10 +330,6 @@ ddL <- data.frame(BirdID = c(dd3$BirdID,dd3_2$BirdID),
 
 rm(temp,dd3_2)
 
-
-
-dd$Cohort <- factor(dd$LayYear)
-dd$FAge <- factor(dd$Age)
 
 # Subset birds with juvenile samples --------------------------------------
 

### 2016-11-23 Scripts/SW juvenile telomeres IPC model averaging.R

@@ -3,8 +3,6 @@
 ## PERFORM MODEL AVERAGING
 #################################################################################*
 
-RTL_Age_Dredge <- dredge(RTL_Age,
-                         subset = !(LogAge & FAge) & !(LogAge & Agemonths ) & !(Agemonths & FAge))
 
 RTL_Full <- standardize(RTL_Full)
 RTL_Full_Dredge <- subset(dredge(RTL_Full),delta <= 6)

### 2016-11-24 Manuscript/SW_early_life_telomeres.Rmd
Eds

@@
 
-**Spatiotemporal variation in lifelong telomere dynamics in the Seychelles warbler**
+**Spatiotemporal variation in lifelong telomere dynamics in a long-term ecological study**
 
@@
 Here we study spatiotemporal variation in lifelong telomere dynamics in the Seychelles warbler.
-We combine long-term ecological data with one the largest longitudinal telomere datasets to date, consisting of `r nrow(dd)` samples from multiple cohorts spanning `r max(dd$CatchYear)-min(dd$CatchYear)` years.
+We combine long-term ecological data with one the largest longitudinal telomere datasets to date, consisting of `r nrow(dd)` samples from `r length(table(dd$Cohort))` cohorts born between `r min(dd$LayYear)` and `r max(dd$LayYear)`.
 We provide a detailed analysis of how telomere dynamics vary over individual lifespans, and with spatiotemporal variation in the social and ecological environment.
@@
 As random effects we included individual ID, catch year and qPCR plate ID.
-Models were compared using AIC.
+Models were compared using AIC with correction for finite sample size (AICc).
+Using the longitudinal data, we tested whether telomores shorten with age in individuals, using LMMs of RTL and $\Delta$age [a longitudinal measure based on within-subject centring; @VandePol2009].
+We calculated $\Delta$age using log and polynomial transformed age data, and carried out model selection as above, with the exception that we did not model $\Delta$age as a factor (due to a lack of discreet groupings).
 
-Using the longitudinal data, we tested whether telomores shorten with age in individuals, using a LMM of RTL and $\Delta$age [a longitudinal measure based on within-subject centring; @VandePol2009].
-Because we observed apparent telomere lengthening in our data, we used the entire Seychelles warbler dataset to test whether this lengthening could be explained by measurement error.
+Because, when examining the distribution of $\Delta$RTL, we observed apparent telomere lengthening in our data, we used the entire Seychelles warbler dataset to test whether this lengthening could be explained by measurement error.
 Using all birds (adults and juveniles) with at least two telomere measurement, we calculated $\Delta$RTL for each pair of samples (for example, for a bird with three longitudinal measurements throughout their life we calculated the change in RTL between the first and second sample and between the second and third sample).
@@
 We created a full model with RTL as a response variable, alongside the following explanatory variables: age, tarsus length, body mass, sex, annual food availability, territory quality, population density (an annual measure estimated from the summer breeding census) and the number of birds and the number of helping  subordinate birds present in the territory.
-As random effects we included catch year and individual identity.
+Random effects structure was informed by the analysis of telomere dynamics and age (see results): we included individual ID, cohort ID, and allowed the slope of the age effect to vary among cohorts.
 We report model estimates and confidence intervals for all effects included in the full model.
@@
 We report model-averaged coefficients, confidence intervals and 'relative importance', which reflects the relative weights of each predictor variable across the top model set.
-For individuals with longitudinal data (*n* = `r nrow(dd3)`) we repeated the above analyses of telomere dynamics, replacing telomere length with $\Delta$RTL as the response variable, and excluding the plate ID random effect (as each measurement of $\Delta$RTL was based on two or more measurements, and so run on multiple plates).
+For individuals with longitudinal data (*n* = `r nrow(dd3)`) we repeated the above analyses of telomere dynamics, replacing telomere length with $\Delta$RTL as the response variable, and excluding the random effects for plate ID (as each measurement of $\Delta$RTL was based on two or more measurements, and so run on multiple plates) and cohort ID (as longitudinal telomere dynamics did not differ among cohorts; see results).
 
 ###Results
-We measured telomere lengths using a total of `r nrow(dd)` unique samples from juvenile and adult Seychelles warblers.
+We measured telomere lengths using a total of `r nrow(dd)` unique samples from juvenile and adult Seychelles warblers from `r length(table(dd$Cohort))` cohorts born between `r min(dd$LayYear)` and `r max(dd$LayYear)`.
 Efficiencies (mean $\pm$ s.d.) for our telomere and GAPDH reactions were (`r round2(mean(dd$EffTelomere))` $\pm$ `r round2(sd(dd$EffTelomere))`) and (`r round2(mean(dd$EffGAPDH))` $\pm$ `r round2(sd(dd$EffGAPDH))`) respectively.
-Inter-plate repeatability of RTL, based on `r nrow(subset(ddL,Group == 'Within samples'))` samples measured at least twice, was `r round2(rptall$R)` (CI = `r round2(rptall$CI_emp[1])`-`r round2(rptall$CI_emp[2])`)
+Inter-plate repeatability of RTL, based on `r nrow(subset(ddL,Group == 'Within samples'))` samples measured at least twice, was `r round2(rptall$R)` (CI = `r round2(rptall$CI_emp[1])`-`r round2(rptall$CI_emp[2])`).
 
-#####Telomere dynamics and age
-Model comparison
+#####Telomere dynamics and age amnong cohorts
+We first tested how RTL was related to age among cohorts using a model selection approach.
+We found that the top model, by some distance, contained a log age, cohort ID, and a log age x cohort interaction (Table 1A).
+The second best model contained log age and cohort ID, with no interaction term, although this was substantially poorer in terms of model fit ($\Delta$AICc > 10), and all other models were much poorer still ($\Delta$AICc > 35; Table 1).
+The log-linear relationship between RTL and age couls be clearly seen in the raw data; RTL decreased with age (estimate = `r round3(fixef(RTL_LogAge)['LogAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_LogAge)['LogAge',],lessthan = F),collapse = ', ')`), with the greatest decrease occurring early in life (Fig.
+1A).
+There was also substantial variation in RTL among cohorts, with no obvious trend over time (Fig.
+1B).
+There was a negative relationship between RTL and age in almost all `r length(table(dd$Cohort))` cohorts, but the slope of the relationship between log age and RTL varied substantially among cohorts (Fig.1C).
 
-Within individuals, longitudinal data showed that RTL measurements taken from different timepoints were weakly, but significantly, positvely correlated (R^2^ = `r round3(summary(lm(RTL1~RTL,data = dd3))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL,data = dd3))$coefficients[2,3])`; P = `r round3(summary(lm(RTL1~RTL,data = dd3))$coefficients[2,4])`; Fig.
-3A).
-Although both or cross-sectional and longitudinal data indicated a general trend of telomere shortening with age, we found that `r round((nrow(subset(dd3,DeltaRTL>0))/nrow(dd3))*100,0)`% of our $\Delta$RTL measurments were positive, and thus indicative of telomere lengthening.
+A within-individual analysis of RTL and age revealed that the top model explaining RTL contained $\Delta$log age, which reflects within-individual changes in log-transformed age (Table 1B).
+Models including cohort ID and cohort x age interactions were substantially poorer than models only containing age (Table 1B).
+RTL dereased with $\Delta$log age (estimate = `r round3(fixef(RTL_DeltaLogAge)['DeltaLogAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_DeltaLogAge)['DeltaLogAge',],lessthan = F),collapse = ', ')`), confirming that within-individual telomere shortening occurs across the Seychelles warbler dataset.
+
+There was a hint of a positive correlation between telomere measurements taken from different timepoints (Fig.
+2A), but this was very weak (marginal R^2^ = `r round2(r.squaredGLMM(RTL_RTL1)['R2m'])`), and not quite significant (estimate = `r round3(fixef(RTL_RTL1)['RTL1'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_RTL1)['RTL',],lessthan = F),collapse = ', ')`).
+Although both or cross-sectional and longitudinal data indicated a general trend of telomere shortening with age, we found that RTL within individuals increaseed in `r round((nrow(subset(dd3,DeltaRTL>0))/nrow(dd3))*100,0)`% of cases (Fig.
+2A), which indicative of telomere lengthening.
 To test whether the apparent telomere lengthening in our dataset could be explained by measurement error, we compared variance in telomere length among repeat measurements of samples to the variance observed among different samples of the same individual (see methods for details).
 We found significantly higher variance in telomere length over individual lifetimes compared to among sample replicates (Levene's test: F = `r round2(leveneTest(DeltaRTL~Group,data=ddL)['F value'][1,1])`; P `r round3(leveneTest(DeltaRTL~Group,data=ddL)['Pr(>F)'][1,1])`; Fig.
-3B).
-Importantly, splitting the longitudinal data into instances of shortening ($\Delta$RTL < 0) and lenghtning ($\Delta$RTL > 0) revealed that not only did we observe significantly more shortening within individuals compared to within samples (Wilcoxon test: W = `r round(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$statistic,0)`; P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$p.value)`), but also significantly more lengthening (W = `r as.character(round(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$statistic,0))`; P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$p.value)`; Fig.
-3B).
+2B).
+Importantly, splitting the longitudinal data into instances of shortening (i.e.
+$\Delta$RTL < 0) and lenghtning (i.e.
+$\Delta$RTL > 0) revealed that not only did we observe significantly more shortening within individuals compared to within samples (Wilcoxon test: P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$p.value)`), but also significantly more lengthening (P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$p.value)`; Fig.
+2B).
 
-#####Spatiotemporal variation in telomere dynamics and the environment
-A LMM revealed that age, tarsus length, sex and annual variation in food availability were significantly related to RTL (Table 1).
-Tarsus length was negatively related to RTL and males had longer telomeres than females (Table 1; Fig.
-4A), while insect availability was positively related to RTL (Table 1; Fig.
-4B).
+#####Telomere dynamics and the environment
+A LMM revealed that age, tarsus length, sex and insect abundance were significantly related to RTL (Fig.
+3A).
+Tarsus length was negatively related to RTL and males had longer telomeres than females (Fig.
+3B), while insect abundance was positively related to RTL (Fig.
+3C).
 The full model was weak in terms of explanatory power of fixed effects (marginal R^2^ = `r round2(r.squaredGLMM(RTL_Full)['R2m'])`), although including the random effect terms increased this substantially (conditional R^2^ = `r round2(r.squaredGLMM(RTL_Full)['R2c'])`) The model averaging approach yielded qualitatively identical results to the full LMM, with the same explanatory variables 'significant' in terms of being retained in top models, and having model-averaged confidence intervals not overlapping zero (Table S1; Fig.
@@
 4A).
-No social or ecological environmental variables were significant predictors of $\Delta$RTL, regardless of whether we interpreted the full model  (Table 2), or used model averaging (Table S2; Fig.
+No social or ecological environmental variables were significant predictors of $\Delta$RTL using the full model  approach (Table S2).
+Using model selection, we found that the top model explaining $\Delta$RTL contained age and population density (Table S3).
+$\Delta$RTL was positively related to age, consistent with telomere shortening being highest in early life, and negatively related to population density; however, in both instances model averaged confidence intervals overlapped zero (Fig.
 S2).
@@
 Here we use a long-term, multi-cohort dataset to assess the relationships between the spatiotemporal variation in the ecological environment and lifelong telomere dynamics in a closed population of Seychelles warblers.
-We found that telomeres shorten over time, and that telomere loss is greatest in early life.
-However, bouts of apparent telomere lengthening occur, and the extent of this lengthening cannot be explained solely by qPCR measurement error.
-We also found that telomere dynamics vary among sexes and over time, with temporal variation in telomere dynamics related to fluctuations in food availability.
+We found that telomere length decreases with age, and that this decrease is greatest in early life.
+Telomere length decreased with age in almost all `r length(table(dd$Cohort)` cohorts studied, but telomere length, and the rate of decrease of telomere length with age, varied substantially among cohorts.
+Despite an overall trend for telomere shortening in the Seychelles warbler, we found that bouts of apparent telomere lengthening occur within individuals, and the extent of this lengthening cannot be explained solely by qPCR measurement error.
+Finally, we found that telomeres are related to tarsus length in a sex-specific manner, and that telomer length is positively associated with temporal fluctuations in food availability.
 
-Our study adds the the now substantial body of literature from humans and wild animals, showing that telomere length decreases with age, and that this decrease is most rapid in early life [e.g.
+Our study adds the the now substantial body of literature from humans and wild animals showing that telomere length decreases with age, and that this decrease is most rapid in early life [e.g.
 @Frenck1998; @Haussmann2003; @Heidinger2012; @Fairlie2016].
-However, using longitudinal data we found that within-individual telomere measurments were only weakly correlated, and that consistent telomere shortening only occurs early in life, after which both shortening and lengthening occurs.
-This is in contrast to other avian studies where birds were reared in laboratory [@Heidinger2012] or nest-box [@Boonekamp2014] conditions, in which juvenile and adult telomere length were highly correlated, and individual-level telomere shortening clearly occurs throughout the juvenile period, and into adulthood.
-However, in accordance with our study, a recent study on a wild population of Soay sheep (*Ovis aries*) found only a very weak correlation between juvenile and adult RTL, and a highly complex pattern of telomere dynamics across the lifespan [@Fairlie2016].
+However, we found that telomere shortening was only consistently observed early in life, after which both shortening and lengthening occurred.
+Importantly, these increases were observed in longitudinal as well as cross-sectional data, suggesting that selective disappearance does not explain this pattern entirely.
+Longitudinal increases in measured telomere length have been observed in humans and [] wild animals [], but whether these observations reflect true elongation or measurement error has been debated.
+
+We are aware of two previously described methods that statistically identify whether telomere elongation occurs within a population.
+One approach uses individuals with at least three measurements to differentiate between hypothesis of consistent lenghtening and a combination of shortening and lengthening [@Simons2014].
+Another study found that the likelihood of detectening increases in telomere length was related to the follow-up time between sampling events, which they argued is inconsistent with telomere elongation [@Steenstrup2013].
+However both of these approaches assume the telomere elongation within individuals are consistent over time.
+We suggest that telomere elongation could occur at specific, short periods throughout individual lifespans, against a backdrop of overall telomere shortening.
+
+Consistent with a dynamic pattern of telomere shortening with age, we found that within-individual telomere measurments were only weakly correlated.
+This is in contrast to other avian studies where birds were reared in laboratory [@Heidinger2012] or nest-box [@Boonekamp2014] conditions, in which within-individual telomere length measurements were highly correlated, and individual-level telomere shortening  occured throughout the juvenile period and into adulthood.
+However, a recent study on a wild population of Soay sheep (*Ovis aries*) found only a very weak correlation between within-individual telomere measurements [@Fairlie2016], and we anticipate that a similar pattern may be found in many other wild populations.
+
+Perhaps the clearest result from our study is that RTL varied among cohorts.
+A few studies have now shown that temporal variation in telomere dynamics occurs in natural populations, although these have been limited in the number of seasons [@Mizutani2013; @Watson2015; @Fairlie2016].
+The long-term Seychelles warbler dataset has allowed us to show that temporal variation in telomere dynamics can occur over substantial time periods.
+Moreover, because of the large number of cohorts we have been able to show statistically that that temporal variation in RTL is related to variation in food availability.
+If cohort-level variation in telomere dynamics continues to be found in other systems, this has potential consequences for our understanding of telomere dynamics in natural populations.
+In particular, our findings suggest that the telomere dynamics of a population at a given point in time represent a snapshot of a temporally varying process.
+More research is now needed within and across multiple cohorts and populations to better understand how how and why population-level telomere dynamics vary over space and time.
 
@@
 
-Perhaps the clearest result from our study is that RTL, among cohorts, and among sampling years.
-A few studies have now shown that temporal variation in telomere dynamics occurs in natural populations, although these have been limited in the number of seasons [@Mizutani2013; @Watson2015; @Fairlie2016].
-The long-term Seychelles warbler dataset has allowed us to show that temporal variation in telomere dynamics can occur over substantial time periods.
-Moreover, because of the large number of cohorts we have been able to show statistically that that temporal variation in RTL is related to variation in food availability (Fig.
-4B).
-If cohort-level variation in telomere dynamics continues to be found in other systems, this has potential consequences for our understanding of telomere dynamics in natural populations.
-In particular, our findings suggest that the telomere dynamics of a population at a given point in time represent a snapshot of a temporally varying process.
-More research is now needed within and across multiple cohorts and populations to better understand how how and why population-level telomere dynamics vary over space and time.
-
-Despite substantial cohort-level variation in RTL, within cohorts we found no evidence that any of the social or environmental variables we measured affect juvenile RTL.
-This is surprising, as a poor social and ecological environment is known to be detrimental to juvenile Seychelles warblers, both in terms of oxidative stress and later life survival [@VandeCrommenacker2011a; @Brouwer2012].
+Despite finding associations between the environment and telomere dynamics, we should bear in mind that our social and ecological variables explained a small proportion of the variance in RTL.
+A poor social and ecological environment is known to be detrimental to juvenile Seychelles warblers, both in terms of oxidative stress and later life survival [@VandeCrommenacker2011a; @Brouwer2012].
 An explanation for this could be that temporal variation in RTL in our data masks our ability to detect spatial trends.
@@
 
-While the relationship between mortality and telomeres in adults has been established for some time [@Cawthon2003], only recently has the link between later-life survival and early-life telomere dynamics been studied.
-In captive zebra finches, juvenile telomere length predicts late-life survival [@Heidinger2012], and in wild bird populations survival to the nestling phase [@Watson2015], and survival to adulthood [@Boonekamp2014], have been linked with early-life telomere dynamics.
-It is worth noting that here we have considered survival, but telomere length and shortening in early life may also be linked to other components of fitness [reviewed in @Monaghan2014].
-Reproductive senescence occurs in the Seychelles warbler [reviewed in @Hammers2015], making this system well suited to examining how telomere length predicts lifetime reproductive success.
-We expect that by gaining a fuller understanding of telomere dynamics in natural populations, the fields of life-history evolution and evolutionary ecology will be greatly enhanced.
-
 ###Acknowledgements

### 2016-11-24 Manuscript/SW_early_life_telomeres_SOM.Rmd
EDS
Eds

@@
 
-###Supplementary information for Spurgin *et al.* "Telomeres reveal silver spoon effects in a wild population"
+###Supplementary information for Spurgin *et al.* "Spatiotemporal variation in lifelong telomere dynamics in a long-term ecological study"
 
@@
 
-#Model averaging
-
 t1 <- as.matrix(RTL_Full_Dredge)
@@
 
-**Table S2** Model selection table for analysis of factors affecting $\Delta$RTL in Seychelles warbler nestlings, using longitudinal data.
+**Table S2** Full model of factors affecting $\Delta$RTL in Seychelles warblers, using longitudinal data.
+```{r Table S2}
+
+#Run the model
+CIs <- CIwald(DeltaRTL_Full)
+t1b <- data.frame(Variable = row.names(CIs)[3:nrow(CIs)],
+                  Estimate = fixef(DeltaRTL_Full),
+                  LCI = CIs[3:nrow(CIs),1],
+                  UCI = CIs[3:nrow(CIs),2],
+                  row.names = NULL,
+                  stringsAsFactors = F)
+
+nams <- t1b$Variable
+nams[nams == 'z.LogAge'] <- 'Age'
+nams[nams == 'z.Tarsus'] <- 'Tarsus length'
+nams[nams == 'c.Helper'] <- 'Number helpers'
+nams[nams == 'z.GroupSize'] <- 'Group size'
+nams[nams == 'c.Sex'] <- 'Sex (males)'
+nams[nams == 'z.BodyMass'] <- 'Body mass'
+nams[nams == 'z.Insect'] <- 'Insect abundance'
+nams[nams == 'z.TQ'] <- 'Territory quality'
+nams[nams == 'z.Density'] <- 'Density'
+
+t1b$Variable <- nams
+
+kable(t1b,digits = 3)
+
+```
+
+**Table S3** Model selection table for analysis of factors affecting $\Delta$RTL in Seychelles warblers, using longitudinal data.
 Only models that remained in the top model set for model averaging (AICc $\leq$ 6 compared to the top model) are included here.
@@
 
-```{r Table S2}
+```{r Table S3}
 
@@
 
-nams[nams == 'z.GroupSize'] <- 'Group Size'
+nams[nams == 'z.GroupSize'] <- 'Group size'
 nams[nams == 'c.Sex'] <- 'Sex (males)'
@@
 
-tableS2 <- data.frame(Model = apply(t1,1,getterms,c(2:10)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
-colnames(tableS2) <- c('Model','df','AICc', 'Delta AICc','Weight')
-rownames(tableS2) <- NULL
-tableS2$Model[tableS2$Model == ''] <- 'Null model'
-tableS2[,3:5] <- lapply(tableS2[,3:5],as.numeric)
-kable(tableS2,digits=2)
+tableS3 <- data.frame(Model = apply(t1,1,getterms,c(2:10)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
+colnames(tableS3) <- c('Model','df','AICc', 'Delta AICc','Weight')
+rownames(tableS3) <- NULL
+tableS3$Model[tableS3$Model == ''] <- 'Null model'
+tableS3[,3:5] <- lapply(tableS3[,3:5],as.numeric)
+kable(tableS3,digits=2)
 
@@
 nams <- rownames(table2)
-nams[nams == 'z.GroupSize'] <- 'Group Size'
+nams[nams == 'z.GroupSize'] <- 'Group size'
 nams[nams == 'c.Sex'] <- 'Sex (males)'

### 2016-11-24 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Eds

--- 2016-11-22 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
@@
 
-**Table 1**
-```{r Table 1}
+**Table 1** Telomere dynamics and age in Seychelles warbler cohorts.
+Linear mixed models were created with RTL as the response variable, and different measures of age, cohort ID, and cohort x age interactions as explanatory variables (see methods for details).
+Models are ranked by AICc, with best models at the top of the table.
 
-AIC(RTL_LogAge)
-AIC(RTL_LogAge_LY)
+```{r Table 1,include = T,echo = F,warning = F}
+RTL_Age_Dredge <- dredge(RTL_Age,
+                         subset = !(LogAge & FAge) & 
+                           !(LogAge & Agemonths) &
+                           !(LogAge & Age2) &
+                           !(Agemonths & FAge) & 
+                           !(Agemonths & Age2) &
+                           !(FAge & Age2))
 
-```
+t1 <- as.matrix(RTL_Age_Dredge)
 
-**Table 2**
-```{r Table 2}
+#Sort out row/col names
+nams <- colnames(t1)
 
-CIs <- CIwald(RTL_Full)
-t1 <- data.frame(Estimate = fixef(RTL_Full), LCI = CIs[7:nrow(CIs),1], UCI = CIs[7:nrow(CIs),2])
+nams[nams == 'Agemonths'] <- 'Age (linear)'
+nams[nams == 'LogAge'] <- 'Age (log)'
+nams[nams == 'FAge'] <- 'Age (factor)'
+nams[nams == 'Cohort:FAge'] <- 'Age (factor)*cohort'
+nams[nams == 'Agemonths:Cohort'] <- 'Age (continuous)*cohort'
+nams[nams == 'Cohort:LogAge'] <- 'Age (log)*cohort'
+nams[nams == 'Age2:Cohort'] <- 'Age (quadratic)*cohort'
+nams[nams == 'Age2'] <- 'Age (quadratic)'
 
-nams <- row.names(t1)
-nams[nams == 'LogAge'] <- 'Age'
-nams[nams == 'Tarsus'] <- 'Tarsus length'
-nams[nams == 'Helper'] <- 'Number helpers'
-nams[nams == 'GroupSize'] <- 'Group size'
-nams[nams == 'SexMales'] <- 'Sex (males)'
-nams[nams == 'BodyMass'] <- 'Body mass'
-nams[nams == 'Insect'] <- 'Annual food availability'
-nams[nams == 'TQ'] <- 'Territory quality'
+colnames(t1) <- nams
 
-row.names(t1) <- nams
+table1A <- data.frame(Model = apply(t1,1,getterms,c(2:10)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
+colnames(table1A) <- c('Model','df','AICc', 'Delta AICc','Weight')
+rownames(table1A) <- NULL
+table1A$Model[table1A$Model == ''] <- 'Null model'
+table1A[,3:5] <- lapply(table1A[,3:5],as.numeric)
 
-kable(t1)
+RTL_DeltaAge_Dredge <- dredge(RTL_DeltaAge,
+                         subset = !(DeltaLogAge & DeltaAge) &
+                           !(DeltaLogAge & DeltaAge2) &
+                           !(DeltaAge & DeltaAge2))
 
-```
+t1 <- as.matrix(RTL_DeltaAge_Dredge)
 
-**Table 2**
+#Sort out row/col names
+nams <- colnames(t1)
 
-```{r Table 2}
+nams[nams == 'DeltaAge'] <- 'Delta age (linear)'
+nams[nams == 'DeltaLogAge'] <- 'Delta age (log)'
+nams[nams == 'Cohort:DeltaAge'] <- 'Delta age (continuous)*cohort'
+nams[nams == 'Cohort:DeltaLogAge'] <- 'Delta age (log)*cohort'
+nams[nams == 'Cohort:DeltaAge2'] <- 'Delta age (quadratic)*cohort'
+nams[nams == 'DeltaAge2'] <- 'Delta age (quadratic)'
 
-#Run the model
-CIs <- CIwald(DeltaRTL_Full)
-t1 <- data.frame(Estimate = fixef(DeltaRTL_Full), LCI = CIs[4:nrow(CIs),1], UCI = CIs[4:nrow(CIs),2])
+colnames(t1) <- nams
 
-nams <- row.names(t1)
-nams[nams == 'LogAge'] <- 'Age'
-nams[nams == 'Tarsus'] <- 'Tarsus length'
-nams[nams == 'Helper'] <- 'Number helpers'
-nams[nams == 'GroupSize'] <- 'Group size'
-nams[nams == 'SexMales'] <- 'Sex (males)'
-nams[nams == 'BodyMass'] <- 'Body mass'
-nams[nams == 'Insect'] <- 'Annual food availability'
-nams[nams == 'TQ'] <- 'Territory quality'
+table1B <- data.frame(Model = apply(t1,1,getterms,c(2:8)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
+colnames(table1B) <- c('Model','df','AICc', 'Delta AICc','Weight')
+rownames(table1B) <- NULL
+table1B$Model[table1B$Model == ''] <- 'Null model'
+table1B[,3:5] <- lapply(table1B[,3:5],as.numeric)
 
-row.names(t1) <- nams
+#Sort out rounding and exlcude NAs from blank spaces. Clunky but works
+table1 <- rbind(c('A',rep(' ',4)),table1A,c('--',rep(' ',4)),c('B',rep(' ',4)),table1B)
+table1[,3:5] <- lapply(table1[,3:5],as.numeric)
+table1[,3:5] <- round(table1[,3:5],3)
+table1[,3:5] <- lapply(table1[,3:5],as.character)
+table1[is.na(table1)] <- " "
 
-kable(t1)
-
+kable(table1,digits = 3)
 ```
@@
 
-**Figure 1** Telomere length in relation to age in Seychelles warblers.
+**Figure 1** Telomere dynamics in relation to age in Seychelles warbler cohorts.
+**A** RTL and age across all individuals.
 Points and  thin grey lines represent individual samples and birds, respectively.
-The thick line and shaded area represent the fitted values and 95% confidence limits of a linea regression of RTL and log-transformed age.
+The thick line and shaded area represent the fitted values and 95% confidence limits of a linear regression of RTL and log-transformed age.**B** Boxplot of variation in RTL among cohorts.
+For visualisation purposes a selection of cohorts with large sample sizes across a wide age range was chosen, but all cohorts were included in models.
+**C** RTL and age among cohorts.
+Lines represent fitted values from a linear regression and log-transformed age, and colours correspond to **B**.
+**D** RTL in relation to and $\Delta$Log age (i.e.
+within indiviual variation in log age).
 
@@
 
-**Figure 3** Cohort variaiton in telomere length in the Seychelles warbler.
-**A** Bocplot of RTL among all Seychelles warbler cohorts.
-**B** Relationship between RTL and age in individual cohorts.
-Only cohorts with samples of at least 10 nestlings were included.
-Lines represent represents the fitted values from a linear regression of RTL and log-transformed age.
-
-**Figure 4** Individual variation in telomere length in Seychelles warbler.
-**A** RTL in relation to tarsus length and sex.
-**B** RTL in relation to variation in annual food availability.
+**Figure 3** Telomere length in relation to the social and ecological environment in the Seychelles warbler.
+**A** Estimates and 95% confidence intervals for all explanatory variables fitted in a linear mixed model (see methods for details).
+**B** RTL in relation to tarsus length and sex.
+**C** RTL in relation to variation in annual food availability.
 Lines and shaded areas represent the fitted values and 95% confidence limits from linear regressions.
@@
 
-  Fig1B <- ggplot(dd3, aes(x = DeltaAge,y = RTL))+
-  geom_point(alpha = 1,col='grey')+
-  xlab(expression(Delta*'Age (years)')) +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black'),
-        legend.position = 'none')+
-  geom_smooth(col = 'black',method = 'lm')+
-  annotate('text',x = -7.5,y = 1.7,label = 'B')
-
-pdf('Figures/Figure 1.pdf',
-    width = 4,
-    height = 8,
-    paper = 'a4')
-multiplot(Fig1A,Fig1B)
-dev.off()
-
-```
-
-```{r Figure 2}
 c15 <- c("dodgerblue2","#E31A1C", # red
@@
                 "darkturquoise",
-                "darkorange4","brown")
+                "darkorange4","brown","navy",
+         'yellow','green','khaki3','blue','pink','orange')
 
-Fig2A <- transform(dd,Cohort = factor(LayYear)) %>%
-  subset(Cohort %in% c(2002:2011)) %>%
-  ggplot(aes(y = RTL,x = Cohort,fill = Cohort))+
+Fig1B <- transform(dd,Cohort = factor(LayYear)) %>%
+  ggplot(aes(y = RTL,x = Cohort, fill = Cohort))+
   xlab('Birth year') +
-  geom_boxplot(notch=T) +
+  geom_boxplot(outlier.shape = NA) +
   theme_classic() +
@@
         axis.line.y = element_line(colour = 'black'),
-        legend.position = 'none')+
+        legend.position = 'none') +
         scale_fill_manual(values = c15)+
-  annotate('text',x = 1,y = 1.7,label = 'A')
+  annotate('text',x = 1,y = 1.6,label = 'B')
 
-Fig2B <- transform(dd,Cohort = factor(LayYear)) %>%
-  subset(Cohort %in% c(2002:2011)) %>%
-  ggplot(aes(y = RTL,x = Agemonths, col = Cohort))+
+Fig1C <- transform(dd,Cohort = factor(LayYear)) %>%
+  ggplot(aes(y = RTL,x = Agemonths, col = Cohort,fill = Cohort))+
   xlab('Age (years)') +
@@
     scale_color_manual(values = c15)+
-  annotate('text',x = 0,y = 1.15,label = 'B')
+        scale_fill_manual(values = c15)+
+  annotate('text',x = 0,y = 1.3,label = 'C')
 
-Fig2C <- transform(dd,Cohort = factor(CatchYear)) %>%
-  subset(Cohort %in% c(2002:2011)) %>%
-  ggplot(aes(y = RTL,x = Cohort,fill = Cohort))+
-  xlab('Sampling year') +
-  geom_boxplot(notch=T) +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12,angle = 90), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black'),
-        legend.position = 'none') + 
-    scale_fill_manual(values = c15)+
-  annotate('text',x = 1,y = 1.7,label = 'C')
-
-Fig2D <- transform(dd,Cohort = factor(CatchYear)) %>%
-  subset(Cohort %in% c(2002:2011)) %>%
-  ggplot(aes(y = RTL,x = Agemonths, col = Cohort))+
-  xlab('Age (years)') +
-  geom_smooth(method = 'lm', formula = y~log(x),se = F) +
+  Fig1D <- ggplot(dd3, aes(x = DeltaLogAge,y = RTL))+
+  geom_point(alpha = 1,col='grey')+
+  xlab(expression(Delta*'Log age (years)')) +
   theme_classic() +
@@
         axis.line.y = element_line(colour = 'black'),
-        legend.position = 'none') +
-    scale_color_manual(values = c15)+
-  annotate('text',x = 0,y = 1.15,label = 'D')
+        legend.position = 'none')+
+  geom_smooth(col = 'black',method = 'lm')+
+  annotate('text',x = -1.5,y = 1.7,label = 'D')
 
-pdf('Figures/Figure 2.pdf',
+pdf('Figures/Figure 1.pdf',
     width = 8,
@@
     paper = 'a4')
-multiplot(Fig2A,Fig2B,Fig2C,Fig2D,layout = matrix(1:4, 2, 2,byrow = T))
+multiplot(Fig1A,Fig1B,Fig1C,Fig1D,layout = matrix(1:4,2,2,byrow = T))
 dev.off()
@@
 
-c15 <- c("dodgerblue2","#E31A1C", # red
-                "green4",
-                "#FF7F00", # orange
-                "gold1",
-                "skyblue2", # lt pink
-                "palegreen2",
-                "#FDBF6F", # lt orange
-                "gray70", "khaki2","blue1","steelblue4",
-                "darkturquoise",
-                "darkorange4","brown")
-
-Fig2A <- transform(dd,Cohort = factor(LayYear)) %>%
-  subset(Cohort %in% c(2002:2011)) %>%
-  ggplot(aes(y = RTL,x = Cohort,fill = Cohort))+
-  xlab('Birth year') +
-  geom_boxplot(notch=T,aes(outlier.colour = Cohort)) +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12,angle = 90), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black'),
-        legend.position = 'none')+
-  scale_fill_manual(values = c15)+
-  annotate('text',x = 1,y = 1.7,label = 'A')
-
-Fig2B <- transform(dd3,Cohort = factor(LayYear)) %>%
-  subset(Cohort %in% c(2002:2011)) %>%
-  ggplot(aes(y = RTL,x = Agemonths, col = Cohort))+
-  xlab('Age (years)') +
-  geom_smooth(method = 'lm', formula = y~log(x),se = F) +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black'),
-        legend.position = 'none') +
-    scale_color_manual(values = c15)+
-  annotate('text',x = 0,y = 1.15,label = 'B')
-
-pdf('Figures/Figure 2.pdf',
-    width = 4,
-    height = 8,
-    paper = 'a4')
-multiplot(Fig2A,Fig2B)
-dev.off()
-
-```
-
-```{r Figure 3}
-
-Fig3A <- ggplot(dd3,
+Fig2A <- ggplot(dd3,
        aes(x = RTL,
@@
 
-Fig3B <- ggplot(ddL,aes(col=Group,fill = Group,y = ..scaled..,x = DeltaRTL))+
+Fig2B <- ggplot(ddL,aes(col=Group,fill = Group,y = ..scaled..,x = DeltaRTL))+
     geom_density(alpha = 0.2)+
@@
 
-pdf('Figures/Figure 3.pdf',
+pdf('Figures/Figure 2.pdf',
     width = 4,
@@
     paper = 'a4')
-multiplot(Fig3A,Fig3B)
+multiplot(Fig2A,Fig2B)
 dev.off()
@@
 
-```{r Figure 4}
+```{r Figure 3}
 
-Fig4A <- ggplot(dd,
+CIs <- CIwald(RTL_Full2)
+t1a <- data.frame(myx = row.names(CIs)[7:nrow(CIs)],
+                  Estimate = fixef(RTL_Full2),
+                  X2.5. = CIs[7:nrow(CIs),1],
+                  X97.5. = CIs[7:nrow(CIs),2],
+                  row.names = NULL,
+                  stringsAsFactors = F)
+
+nams <- t1a$myx
+nams[nams == 'z.LogAge'] <- 'Age'
+nams[nams == 'z.Tarsus'] <- 'Tarsus length'
+nams[nams == 'c.Helper'] <- 'Number helpers'
+nams[nams == 'z.GroupSize'] <- 'Group size'
+nams[nams == 'c.Sex'] <- 'Sex (males)'
+nams[nams == 'z.BodyMass'] <- 'Body mass'
+nams[nams == 'z.Insect'] <- 'Insect abundance'
+nams[nams == 'z.TQ'] <- 'Territory quality'
+nams[nams == 'z.Density'] <- 'Density'
+
+t1a$myx <- nams
+
+t1a$myx <- factor(t1a$myx,levels = t1a$myx[order(t1a$Estimate,decreasing = T)])
+t1a <- subset(t1a,myx != "(Intercept)")
+
+Fig3A <- forestplot(t1a,RI = F)+
+  annotate('text',y = 0.1,x = 0.6,label = 'A')
+
+Fig3B <- ggplot(dd,
        aes(x = Tarsus,
@@
   scale_color_manual(values = c('gold2','red'))+
-  annotate('text',x = 17,y = 1.7,label = 'A')
+  annotate('text',x = 17,y = 1.7,label = 'B')
 
-Fig4B <- ggplot(dd,
+Fig3C <- ggplot(dd,
        aes(y = RTL,
@@
    stat_smooth(method = 'lm',col = 'darkgrey') +
-  annotate('text',x = 2.5,y = 1.6,label = 'B')
+  annotate('text',x = 2.5,y = 1.6,label = 'C')
 
-pdf('Figures/Figure 4.pdf',
-    width = 4,
+pdf('Figures/Figure 3.pdf',
+    width = 8,
     height = 8,
     paper = 'a4')
-multiplot(Fig4A,Fig4B)
+multiplot(Fig3A,Fig3B,Fig3C,layout = matrix(c(1,1,2,3),byrow = T,nrow = 2))
 dev.off()

### 2016-11-24 Scripts/SW juvenile telomeres IPC clean.R

@@ -59,6 +59,8 @@ dd <- subset(dd,Season == 'Major')
 dd <- subset(dd,CatchMonth %in% c(6:9))
 dd$yday <- as.POSIXlt(dd$CatchDate)$yday - min(as.POSIXlt(dd$CatchDate)$yday)
 
+dd <- subset(dd,LayYear > 1992)
+
 # Age data ----------------------------------------------------------------
 
 dd$Age <- (dd$CatchYear-dd$LayYear)

### 2016-11-24 Scripts/SW juvenile telomeres IPC functions.r

@@ -276,10 +276,10 @@ propmales <- function(x)
 
 #Forestplot
 
-forestplot <- function(mytable)
+forestplot <- function(mytable,RI = T)
 {
   
-  ggplot(mytable,
+  x <- ggplot(mytable,
          aes(x = myx,
              y=Estimate)) +
     geom_point() + 
@@ -290,9 +290,19 @@ forestplot <- function(mytable)
     theme(axis.text.x = element_text(angle=90,size = 14,hjust=1,vjust=0.5), 
           axis.text.y = element_text(size = 14), 
           axis.title.y = element_text(size = 16,vjust=0.8),
-          axis.title.x = element_text(size = 16)) +
-    geom_abline(slope=0,intercept=0,lty=2) +
-    annotate('text',x=mytable$myx,y=min(mytable$X2.5)*1.2,label=paste('(',mytable$RI,')',sep=''),size = 4)
+          axis.title.x = element_text(size = 16),
+          axis.line.x = element_line(colour = 'black'),
+          axis.line.y = element_line(colour = 'black')) +
+    geom_abline(slope=0,intercept=0,lty=2)
+  
+  if(RI == T)
+  {
+    x + annotate('text',x=mytable$myx,y=min(mytable$X2.5)*1.2,label=paste('(',mytable$RI,')',sep=''),size = 4)
+  } else
+  {
+    x
+  }
+
     
 }
 
@@ -456,6 +466,11 @@ give.n <- function(x)
   {
   return(c(y = median(x)*0.96, label = length(x)))
   }
+
+
+
+
+# CIWald - get Wald confidence intervals ----------------------------------
 
 CIwald <- function(model)
 {

### 2016-11-24 Scripts/SW juvenile telomeres IPC models.r

@@ -9,26 +9,34 @@
 RTL_Age <- lmer(RTL ~ Cohort*(Agemonths + LogAge + FAge + Age2) + (1|BirdID) + (1|CatchYear) + (1|PlateID),
                 data=dd,
                 REML = FALSE)
-
+RTL_LogAge <- lmer(RTL ~ LogAge + (1|BirdID) + (1|CatchYear) + (1|PlateID),
+                data=dd,
+                REML = FALSE)
 RTL_DeltaAge <- lmer(RTL ~ Cohort*(DeltaAge + DeltaLogAge + DeltaAge2) + (1|BirdID) + (1|CatchYear) + (1|PlateID),
                 data=dd3,
                 REML = FALSE)
+RTL_DeltaLogAge <- lmer(RTL ~ DeltaLogAge + (1|BirdID) + (1|CatchYear) + (1|PlateID),
+                   data=dd3,
+                   REML = FALSE)
+RTL_RTL1 <- lmer(RTL1 ~ RTL + (1|BirdID),
+                        data=dd3,
+                        REML = FALSE)
 
 # RTL and ecology ---------------------------------------------------------
 
 ddNA <- subset(subset(dd,!is.na(Insect)),!is.na(TQ))
 dd3NA <- subset(subset(dd3,!is.na(Insect)),!is.na(TQ))
 
-RTL_Full <- lmer(RTL ~  LogAge + Tarsus + Helper + GroupSize + Sex + BodyMass + Insect + Density + TQ + (Agemonths|CatchYear) + (1|BirdID) + (1|PlateID),
+RTL_Full <- lmer(RTL ~  LogAge + Tarsus + Helper + GroupSize + Sex + BodyMass + Insect + Density + TQ + (Agemonths|LayYear) + (1|BirdID) + (1|PlateID),
                  data = ddNA,
                  REML = FALSE)
-RTL_Full_NoTarsus <- lmer(RTL ~  LogAge + Helper + GroupSize + Sex + BodyMass + Insect + Density + TQ + (Agemonths|CatchYear) + (1|BirdID) + (1|PlateID),
+RTL_Full_NoTarsus <- lmer(RTL ~  LogAge + Helper + GroupSize + Sex + BodyMass + Insect + Density + TQ + (Agemonths|LayYear) + (1|BirdID) + (1|PlateID),
                           data = ddNA,
                           REML = FALSE)
[…]

### 2016-11-25 Manuscript/SW_early_life_telomeres.Rmd
Eds

@@
 Telomere length and rates of shortening varied markedly both within and among cohorts, and cohort-level differences in telomere length persisted throughout individual lifespan.
-Temporal variation in telomere length was positively associated with island-wide temporal variation in food availability.
-The other significant predictors of telomere length were sex, with males having longer telomeres than females, and tarus length, which affected telomere length in a sex-specific manner.
+Temporal variation in telomere length was positively associated with island-wide temporal variation in insect abundance, suggesting that temporal variation in food availability imposes a cost in Seychelles warblers.
+We also found evidence for sex-specific relationships between telomeres and tarsus length, potentially reflecting differential costs of growth.
 
 5.
-Our comprehensive, long-term data show that in natural populations, telomere dynamics vary enormously over both space and time.
-Ascertaining what explains this variation requires combining large sample sizes spanning multiple cohorts, and long-term ecological data.
+Our long-term data show that in natural populations, telomere dynamics vary in a highly complex manner over individual lifespans, and across space and time.
+Variance in telomere dynamics among individuals reflect wide array of genetic, parental and evironmental factors.
+Explaining this variation more fully will require the integration of comprehensive long-term ecological and genetic data from multiple populations and species.
 
@@
 ###Introduction
-Telomeres shorten with age [@Monaghan2006], and in response to oxidative stress, which can be elevated  due to environmental factors [@VonZglinicki2002].
+A major aim of ecologists is to be able to quantify and understand why individuals vary in their response to different environments.
+Identifying the costs imposed on individuals within populations is central to understanding variation in fitness [@Lindstrom1999], and thus for understanding population and community dynamics [@Bolnick2011].
+However, fully quantifying individual-level variation in costs is impossible in wild systems, and thus effective biomarkers that reflect individual-level experiences are required.
+
+Telomeres are repetitive DNA sequences found on the ends of linear chromosomes that protect against DNA damage.
+Telomeres shorten with age [@Monaghan2006], and there is evidence from a range of taxa that the greatest rate of telomere shortening occurs in early life [e.g.
+@Frenck1998; @Haussmann2003].
+Telomere shortening also occurs in response to oxidative stress, which can be elevated  due to environmental factors [@VonZglinicki2002].
+There is evidence from both humans and wild animal populations that telomere shortening is influenced by the conditions experienced during both early life and adulthood [@Price2013; @Monaghan2014; @Nettle2015; @Reichert2015].
+Importantly, the extent of telomere shortening that occurs has a direct link to senescence and survival.
 When telomeres become critically short cells senesce [@Campisi2003], and the accumulation of these cells can result in organismal senescence and death [@Wong2003].
-This association between senescence and telomere length has inspired a great deal of recent research into telomere evolutionary ecology [reviewed in @Horn2010; @Haussmann2010; @Monaghan2014].
-While there is little direct evidence that the relationship between telomere dynamics and survival is causal [@Simons2015], there is mounting evidence that telomeres can act as biomarkers of cost in wild populations, providing a signature of the ecological stress that has been experienced and is otherwise difficult to detect [@Monaghan2014; @Schultner2014; @Asghar2015].
+This association between senescence and telomere length has inspired a great deal of recent research into telomere evolutionary ecology, and relationships between telomere dynamics and survival have been documented in several wild organisms [reviewed in @Horn2010; @Haussmann2010; @Simons2015].
+While there is as yet little direct evidence that the relationship between telomere dynamics and survival is causal [@Simons2015], there is mounting evidence that telomeres can act as biomarkers of cost in wild populations, providing a signature of the ecological stress that has been experienced and is otherwise difficult to detect [e.g.
+@Schultner2014; @Asghar2015].
 
-There is evidence from a range of taxa that the greatest rate of telomere shortening occurs in early life [e.g.
-@Frenck1998; @Haussmann2003], and that the extent of this shortening is influenced by the conditions experienced during that period [@Price2013; @Monaghan2014; @Nettle2015; @Reichert2015].
-Early-life telomere dynamics have been associated with both short-term and late-life survival [@Heidinger2012; @Boonekamp2014], and with other parameters such as cognition [@Nettle2015a].
-However, few studies have simultaneously analysed how telomere dynamics, early-life conditions and late-life survival are all related in a natural setting.
-Moreover, how early-life telomere dynamics vary over spatial and temporal scales is poorly understood.
+Although there has now been a considerable amount of effort put into studying telomere dynamics in natural populations, our understanding of the forces responsible for explaining variation in telomere length is still limited.
+This is important, as before we can use telomeres as a measure of the costs experienced by individuals and populations, we need to how different developmental, genetic and ecological variables interact to affect telomeres.
+Telomere length and rates of shortening can vary according to parental characteristics [Njajour2007; @Heidinger2016], among sexes [@Barrett2011], and with a whole host of environmental conditions experienced at different life-history stages [@Monaghan2014].
+There is also some evidence that telomeres can increase, as well as decrease, in length [@Simons2014], which has important ramifications for our understanding of how telomeres reflect costs; however, telomere lengthening is often attributed to measurement error [@Steenstrup2015].
+Recent evidence suggests that telomere dynamics are indeed highly variable over individual lifespans, and that even the relationship between telomeres and age can vary markedly among cohorts [@Fairlie2016].
+To better understand what explains variation in telomere dynamics, more studies are required that incorporate telomere variation over entire lifepans with comprehensive, long-term ecological data.
 
 The longitudinal study (since 1986) of the Seychelles warbler (*Acrocephalus sechellensis*) population on Cousin Island provides an excellent system for studying senescence in the wild [reviewed in @Hammers2015].
-Due to the isolated nature of the study population and intensive field monitoring, we have unusually comprehensive survival data and tissue samples spanning many years (see Methods, below).
-Ecological conditions and warbler population density on Cousin vary across space and time due to weather-induced changes in foliage cover and food availability [@VandeCrommenacker2011a].
-Seychelles warblers remain on their natal territories for at least six months, and variation in the oxidative stress experienced by individuals is associated with natal territory quality [@VandeCrommenacker2011a].
-However, neither early-life nor adult survival appear to be associated with natal territory quality or natal local density [@Brouwer2006; @Hammers2013].
+Due to the isolated nature of the study population and intensive field monitoring, we have unusually comprehensive ecological and survival data spanning many years (see Methods, below).
+Conditions and warbler population density on Cousin vary across space and time due to weather-induced changes in foliage cover and food availability [@VandeCrommenacker2011a].
+Variation in the oxidative stress experienced by individuals is associated with territory quality [@VandeCrommenacker2011a].
+However, neither early-life nor adult survival appear to be associated with territory quality or local density [@Brouwer2006; @Hammers2013].
 Facultative cooperative breeding occurs in the Seychelles warbler [@Komdeur1994a; @Richardson2003a], and the presence of helping subordinates (but not non-helping subordinates) in the natal territory is associated with increased survival later in life [@Brouwer2012].
-Lastly, we have an established protocol for assessing telomere length in this species [@Barrett2012; @Bebbington2016], and telomere length predicts survival independently of age in adult Seychelles warblers, suggesting that telomeres act as a biomarker of cost in this species [@Barrett2013a].
-Thus, we have an excellent system in which to assess the costs of different social and environmental conditions experienced early in life, and to assess the later-life consequences of early-life conditions.
+Lastly, we have an established protocol for assessing telomere length in this species [@Barrett2012; @Bebbington2016], and telomere length predicts survival independently of age, suggesting that telomeres act as a biomarker of cost in this species [@Barrett2013a].
+Thus, we have an excellent system in which to assess the costs of different social and environmental conditions experienced by individuals, and to assess how these costs vary over space and time.
 
-In this study we examine how telomeres are related to early-life environmental variation over the entire lifespans across multiple cohorts of the Seychelles warbler.
+In this study we study how lifelong telomere dynamics are related to environmental variation across `r length(table(dd$Cohort))` Seychelles warbler cohorts.
+We first study how telomere length and rates of shortening are related to age, and how this relationship varies among cohorts, in order to gain an in-depth understanding of the temporal dynamics of telomere changes.
+We then examine, within individuals, how telomere length changes with age, and statistically test whether increases in telomere length are biologically meaningful.
+Finally, we test how a wide range of social and environmental variables are related to telomere length and shortening in order gain a fuller understanding of the forces driving telomere dynamics in natural populations.
 
@@
 
-Because, when examining the distribution of $\Delta$RTL, we observed apparent telomere lengthening in our data, we used the entire Seychelles warbler dataset to test whether this lengthening could be explained by measurement error.
-Using all birds (adults and juveniles) with at least two telomere measurement, we calculated $\Delta$RTL for each pair of samples (for example, for a bird with three longitudinal measurements throughout their life we calculated the change in RTL between the first and second sample and between the second and third sample).
-We then compared these longitudinal changes in telomere length to sampling error, using all samples where we had at least two repeat telomere measurements for the same sample (i.e.
+We used our longitudinal data to determine individual-level consistency in RTL.
+We constructed a LMM with RTL at time *t* + 1 as the response variable, RTL at time *t* and age as fixed effects, and individual ID and cohort as random effects.
+Because, when examining the distribution longitudinal telomere changes, we observed apparent telomere lengthening in our data, we used the entire Seychelles warbler dataset to test whether this lengthening could be explained by measurement error.
+We compared longitudinal changes in telomere length among different samples from the same birds to samples where we had at least two repeat telomere measurements for the same sample (i.e.
 completely seperate reactions run on separate plates: N = `r nrow(subset(ddL,Group == 'Within samples'))` measurements from `r length(unique(subset(ddL,Group == 'Within samples')$BirdID))` samples).
@@
 
-There was a hint of a positive correlation between telomere measurements taken from different timepoints (Fig.
-2A), but this was very weak (marginal R^2^ = `r round2(r.squaredGLMM(RTL_RTL1)['R2m'])`), and not quite significant (estimate = `r round3(fixef(RTL_RTL1)['RTL1'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_RTL1)['RTL',],lessthan = F),collapse = ', ')`).
-Although both or cross-sectional and longitudinal data indicated a general trend of telomere shortening with age, we found that RTL within individuals increaseed in `r round((nrow(subset(dd3,DeltaRTL>0))/nrow(dd3))*100,0)`% of cases (Fig.
-2A), which indicative of telomere lengthening.
-To test whether the apparent telomere lengthening in our dataset could be explained by measurement error, we compared variance in telomere length among repeat measurements of samples to the variance observed among different samples of the same individual (see methods for details).
+There was positive correlation between telomere measurements taken from different timepoints (Fig.
+2A), but this was very weak (marginal R^2^ = `r round2(r.squaredGLMM(RTL_RTL1)['R2m'])`), and not quite significant (estimate = `r round3(fixef(RTL_RTL1)['RTL'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_RTL1)['RTL',],lessthan = F),collapse = ', ')`).
+Although both cross-sectional and longitudinal data indicated a general trend of telomere shortening with age, we found that RTL within individuals increased in `r round((nrow(subset(dd3,DeltaRTL>0))/nrow(dd3))*100,0)`% of our `r nrow(dd3)` $\Delta$RTL measurements (Fig.
+2A), which is indicative of telomere lengthening.
+To test whether this apparent telomere lengthening in our dataset could be explained by measurement error, we compared variance in telomere length among repeat measurements of samples to the variance observed among different samples of the same individual.
 We found significantly higher variance in telomere length over individual lifetimes compared to among sample replicates (Levene's test: F = `r round2(leveneTest(DeltaRTL~Group,data=ddL)['F value'][1,1])`; P `r round3(leveneTest(DeltaRTL~Group,data=ddL)['Pr(>F)'][1,1])`; Fig.
 2B).
-Importantly, splitting the longitudinal data into instances of shortening (i.e.
+Splitting the longitudinal data into instances of shortening (i.e.
 $\Delta$RTL < 0) and lenghtning (i.e.
@@
 #####Telomere dynamics and the environment
-A LMM revealed that age, tarsus length, sex and insect abundance were significantly related to RTL (Fig.
+In addition to age, tarsus length, sex and insect abundance were significantly associated with RTL (Fig.
 3A).
@@
 Our study adds the the now substantial body of literature from humans and wild animals showing that telomere length decreases with age, and that this decrease is most rapid in early life [e.g.
-@Frenck1998; @Haussmann2003; @Heidinger2012; @Fairlie2016].
+@Frenck1998; @Haussmann2003; @Heidinger2012].
 However, we found that telomere shortening was only consistently observed early in life, after which both shortening and lengthening occurred.
 Importantly, these increases were observed in longitudinal as well as cross-sectional data, suggesting that selective disappearance does not explain this pattern entirely.
-Longitudinal increases in measured telomere length have been observed in humans and [] wild animals [], but whether these observations reflect true elongation or measurement error has been debated.
+Longitudinal increases in measured telomere length have been observed in humans and [reviewed in Steenstrup2015] wild animals [@Kotrschal2007; @Fairlie2016].
+The most commonly invoked explanation for increases in telomere length is measurement error, which can be a particular problem in qPCR-based telomere studies [@Nussey2014; @Steenstrup2015; @Verhulst2015].
+To our knowledge, ours is the first study to explicitly compare intra-individual variation among samples to variation among sample replicates, on a large scale.
+Our results suggest that qPCR measurement error alone cannot explain observed increases in RTL observed within individuals.
 
-We are aware of two previously described methods that statistically identify whether telomere elongation occurs within a population.
-One approach uses individuals with at least three measurements to differentiate between hypothesis of consistent lenghtening and a combination of shortening and lengthening [@Simons2014].
-Another study found that the likelihood of detectening increases in telomere length was related to the follow-up time between sampling events, which they argued is inconsistent with telomere elongation [@Steenstrup2013].
-However both of these approaches assume the telomere elongation within individuals are consistent over time.
-We suggest that telomere elongation could occur at specific, short periods throughout individual lifespans, against a backdrop of overall telomere shortening.
+Our results (Fig.
+1A) suggest that increases in telomere length occur at specific, short periods throughout individual lifespans, against a backdrop of overall telomere shortening.
+Consistent with a dynamic pattern of telomere shortening with age, we found that within-individual telomere measurments were only weakly correlated.
+These findings are in contrast to other avian studies where birds were reared in laboratory [@Heidinger2012] or nest-box [@Boonekamp2014] conditions, in which within-individual telomere length measurements were highly consistent, and individual-level telomere shortening occured throughout the juvenile period and into adulthood.
+However, the lifelong telomere dynamics found in Seychelles warblers are strikingly similar to those found in Soay sheep [@Fairlie2016], and we anticipate that a similar pattern may be found in other wild populations.
+That increases in telomere length may be sporadic and interspersed by bouts of shortening is a crucial point, because previously described approaches to distinguish telomere elongation from measurement error based on assumptions about follow-up time between measurements [@Steenstrup2015], or based on measuring variance among measurements [@Simons2014] assume the telomere elongation within individuals is consistent over time.
+Such sporadic changes in telomere length over lifespans could occur due to a change in the composition of cell types within individual samples, or due to the actual elongation of telomeres.
+Determining the mechanism of these changes is essential for how we view telomeres as biomarkers of costs.
+If telomeres can be lengthened in response to improvements in environmental conditions, for example, this would suggest the they reflect short-to medium-term costs, rather than the  the cumulative costs that an individual has faced over its lifespan [@Bateson2016].
+New statistical and technical approaches are therefore now required to determine the mechanisms behind increases in telomere length within individuals.
 
-Consistent with a dynamic pattern of telomere shortening with age, we found that within-individual telomere measurments were only weakly correlated.
-This is in contrast to other avian studies where birds were reared in laboratory [@Heidinger2012] or nest-box [@Boonekamp2014] conditions, in which within-individual telomere length measurements were highly correlated, and individual-level telomere shortening  occured throughout the juvenile period and into adulthood.
-However, a recent study on a wild population of Soay sheep (*Ovis aries*) found only a very weak correlation between within-individual telomere measurements [@Fairlie2016], and we anticipate that a similar pattern may be found in many other wild populations.
+A few studies have shown that temporal variation in telomere dynamics occurs in natural populations, although these have been limited in the number of seasons [@Mizutani2013; @Watson2015; @Fairlie2016].
+The long-term Seychelles warbler dataset has allowed us to show that temporal variation in telomere dynamics can occur over substantial time periods.
+Our data suggest that birth year is a highly important factor in shaping telomere dynamics, but also that age-related declines in telomere length vary among cohorts.
+This is consistent with telomere length being controlled by both genetic and environmental factors - something that is becoming apparent from quantitative genetic studies of telomeres.
+Moreover, our findings suggest that the telomere dynamics of a population at a given point in time represent a snapshot of a temporally varying process.
+Research of telomere dynamics within and across multiple cohorts and populations will enable us to better understand how how and why population-level telomere dynamics vary over space and time.
 
-Perhaps the clearest result from our study is that RTL varied among cohorts.
-A few studies have now shown that temporal variation in telomere dynamics occurs in natural populations, although these have been limited in the number of seasons [@Mizutani2013; @Watson2015; @Fairlie2016].
-The long-term Seychelles warbler dataset has allowed us to show that temporal variation in telomere dynamics can occur over substantial time periods.
-Moreover, because of the large number of cohorts we have been able to show statistically that that temporal variation in RTL is related to variation in food availability.
-If cohort-level variation in telomere dynamics continues to be found in other systems, this has potential consequences for our understanding of telomere dynamics in natural populations.
-In particular, our findings suggest that the telomere dynamics of a population at a given point in time represent a snapshot of a temporally varying process.
-More research is now needed within and across multiple cohorts and populations to better understand how how and why population-level telomere dynamics vary over space and time.
+The long-term nature of the Seychelles warbler study enabled us to show that temporal variation in food availability was positively related to telomere length.
+This is consistent with strong cohort effects, and suggests that temporal variation in conditions may be a key driver of costs in the Seychelles warbler.
+Although Cousin Island is relatively benign in comparison to other island systems [e.g.
+@Coulson2001], substantial annual variation in rainfall does occur, with associated changes in insect abundance, and it appears that this confers a cost to Seychelles warblers.
+We also found evidence for sex-specific telomere dynamics, with males having longer-telomeres than females, and that this sex-difference interacts with tarsus length.
+If the sex-dependent relationship between telomere and tarsus lengths was due to differential growth alone we would expect the opposite pattern to that observed, as male Seychelles warblers are larger than females (Fig.
+3B).
+One possibility is that the environment imposes differential costs on males and females: a recent study in captive zebra finches found that manipulation of dietary nutrients had sex-sependent affects on telomere dynamics [@Noguera2015].
+Also worth noting is that the effect of telomere length on survival in strongest in male Seychelles warblers [@Barrett2013a], although the nature of the relationship between sex, telomeres and survival is not yet clear [@Barrett2011]
 
-We also found evidence for sex-specific telomere dynamics in Seychelles warblers, with males having longer-telomeres than females, in and that this sex-difference interacts with tarsus length.
-This is likely to be due to differential growth costs between sexes.
-Furthermore, in Seychelles warblers the effect of telomere length on survival is strongest for males [@Barrett2013a].
-Together, these results suggest that sex effects should be considered when studying any aspect of telomere dynamics [see also @Barrett2011].
+Although we found clear associations between the environment and telomere dynamics, we should bear in mind that our social and ecological variables explained a small proportion of the variance in RTL.
+A poor social and ecological environment is known to be detrimental to Seychelles warblers, both in terms of oxidative stress and survival [@VandeCrommenacker2011a; @Brouwer2012], and it is therefore perhaps surprising that these variables do not explain more variance in RTL.
+While measurement error clearly accounts for some of this noise, it is also likely that early-life RTL in the Seychelles warbler is explained by a set of environmental and genetic variables not considered here [e.g.
+@Bebbington2016].
+A key question to be addressed is the extent to which RTL, especially in early life, reflects inheritance and parental effects [e.g.
+@Becker2015; @Asghar2014; @Heidinger2016].
+Future research on the Seychelles warbler will examine, in a quantitative genetic framework, how genetic and environmental components, and their interactions, affect telomere dynamics and senescence.
 
-Despite finding associations between the environment and telomere dynamics, we should bear in mind that our social and ecological variables explained a small proportion of the variance in RTL.
-A poor social and ecological environment is known to be detrimental to juvenile Seychelles warblers, both in terms of oxidative stress and later life survival [@VandeCrommenacker2011a; @Brouwer2012].
-An explanation for this could be that temporal variation in RTL in our data masks our ability to detect spatial trends.
-However, this is unlikely as we controlled for cohort ID, and within-cohort variation in RTL greatly exceeds among-cohort variation (Fig.
-1C).
-A second possibility is that effects are generally weak and levels of noise in our telomere measurements and/or ecological data preclude detection of significant effects.
-While noise in telomere measurements and problems with power are a problem in any study using qPCR to measure telomere length [@Nussey2014], our sample sizes and levels of repeatability compare favourably with other studies of natural populations that have found significant effects of the early-life environment on telomere length [e.g.
-@Fairlie2016].
-More likely is that early-life RTL in the Seychelles warbler is explained by a set of variables that we have not measured.
-A key question to be addressed is the extent to which early-life RTL reflects inheritance [e.g.
-@Becker2015; @Asghar2014].
-This is important, as inheritance may capture a large amount of unexplained variation in early-life RTL within our dataset.
-Future research on the Seychelles warbler will examine, in a quantitative genetic framework, how genetic and environmental components, and their interactions, affect telomere dynamics and senescence.
+Finally, we emphasise once more that many of the findings here were dependent on long-term ecological data.
+Long-term ecological study systems are uniquely suited to addressing a wide range of problems in ecology [@Clutton-Brock2010], and understanding variation in telomere dynamics is one such problem.
+To gain a full understanding of telomere dynamics in natural systems, long-term studies combining ecological and genetic data will be required from a range of species.
 

### 2016-11-25 Scripts/SW juvenile telomeres IPC clean.R

@@ -3,10 +3,7 @@
 ## CLEAN THE DATA
 #################################################################################
 
-#Only use Ellie's samples
-# dd0 <- subset(dd0,Whodunnit == 'EAF')
-
-
+#Subset bad samples
 dd0 <- subset(dd0,CqTelomere < 25) %>%
   subset(CqGAPDH < 26) %>%
   subset(CqGAPDH > 21) %>%

### 2016-11-25 Scripts/SW juvenile telomeres IPC models.r

@@ -18,7 +18,7 @@ RTL_DeltaAge <- lmer(RTL ~ Cohort*(DeltaAge + DeltaLogAge + DeltaAge2) + (1|Bird
 RTL_DeltaLogAge <- lmer(RTL ~ DeltaLogAge + (1|BirdID) + (1|CatchYear) + (1|PlateID),
                    data=dd3,
                    REML = FALSE)
-RTL_RTL1 <- lmer(RTL1 ~ RTL + (1|BirdID),
+RTL_RTL1 <- lmer(RTL1 ~ RTL + LogAge + (1|BirdID) + (1|LayYear),
                         data=dd3,
                         REML = FALSE)
 

### 2016-11-25 Manuscript/SW_early_life_telomeres.Rmd
EDITS BEFORE DRAFT TO DAVE

@@
 1.
-Understanding the costs individuals and populations face throughout their lifetimes is fundamental to understanding life-history evolution and population dynamics.
+Understanding the costs that individuals and populations face throughout their lifetimes is fundamental to understanding life-history evolution and population dynamics.
 Telomeres, the protective caps at the ends of chromosomes, shorten in response to oxidative stress, and telomere shortening is correlated with reduced survival.
-Thus, telomere dynamics may help us quantify individual variation in costs, and enhance our understanding individual variation in natural populations.
+Thus, telomere dynamics may help us quantify individual variation in costs, and enhance our understanding of individual variation in natural populations.
 
 2.
-The Seychelles warbler (*Acrocephalus sechellensis*) is an excellent model system for disentangling the causes and consequences of individual variation in telomere dynamics.
-Here we study spatiotemporal variation in lifelong telomere dynamics in the Seychelles warbler.
+Here we study spatio-temporal variation in lifelong telomere dynamics in the Seychelles warbler (*Acrocephalus sechellensis*).
 We combine long-term ecological data with one the largest longitudinal telomere datasets to date, consisting of `r nrow(dd)` samples from `r length(table(dd$Cohort))` cohorts born between `r min(dd$LayYear)` and `r max(dd$LayYear)`.
-We provide a detailed analysis of how telomere dynamics vary over individual lifespans, and with spatiotemporal variation in the social and ecological environment.
+We provide a detailed analysis of how telomere dynamics vary over individual lifespans, and with spatio-temporal variation in the social and ecological environment.
 
@@
 We found that telomere length decreases with cross-sectional and longitudinal measures of age, and that telomere length decreases most rapidly early in life.
-However, both cross-sectional and longitudinal data suggested that telomere length increases occurs in some individuals for short periods, and using a large number of repeated measurements we show statistically that this increase cannot be explained solely by qPCR measurement error.
+However, both cross-sectional and longitudinal data suggested that against this overall pattern of shortening, bouts of telomere length increases occur in some individuals.
+Using a large number of repeated measurements we show statistically that these increases cannot be explained solely by qPCR measurement error.
 
 4.
-Telomere length and rates of shortening varied markedly both within and among cohorts, and cohort-level differences in telomere length persisted throughout individual lifespan.
-Temporal variation in telomere length was positively associated with island-wide temporal variation in insect abundance, suggesting that temporal variation in food availability imposes a cost in Seychelles warblers.
+Telomere length decreased with age in almost all cohorts studied, but telomere length and the slope of the relation ship with age varied markedly among cohorts.
+Variation in telomere length was positively associated with island-wide temporal variation in insect abundance, suggesting that temporal variation in food availability imposes a cost in Seychelles warblers.
 We also found evidence for sex-specific relationships between telomeres and tarsus length, potentially reflecting differential costs of growth.
@@
 Our long-term data show that in natural populations, telomere dynamics vary in a highly complex manner over individual lifespans, and across space and time.
-Variance in telomere dynamics among individuals reflect wide array of genetic, parental and evironmental factors.
+Variance in telomere dynamics among individuals is the product of wide array of genetic, parental and environmental factors.
 Explaining this variation more fully will require the integration of comprehensive long-term ecological and genetic data from multiple populations and species.
@@
 
-PAGEBREAK
-
 ###Introduction
@@
 Identifying the costs imposed on individuals within populations is central to understanding variation in fitness [@Lindstrom1999], and thus for understanding population and community dynamics [@Bolnick2011].
-However, fully quantifying individual-level variation in costs is impossible in wild systems, and thus effective biomarkers that reflect individual-level experiences are required.
+However, fully quantifying individual-level variation in costs is impossible in wild systems, and thus effective biomarkers that reflect the physiological consequences of individual-level experiences are required.
 
@@
 Although there has now been a considerable amount of effort put into studying telomere dynamics in natural populations, our understanding of the forces responsible for explaining variation in telomere length is still limited.
-This is important, as before we can use telomeres as a measure of the costs experienced by individuals and populations, we need to how different developmental, genetic and ecological variables interact to affect telomeres.
+This is important, as before we can use telomeres as a measure of the costs experienced by individuals and populations, we need to know how different developmental, genetic and ecological variables interact to affect telomeres.
 Telomere length and rates of shortening can vary according to parental characteristics [Njajour2007; @Heidinger2016], among sexes [@Barrett2011], and with a whole host of environmental conditions experienced at different life-history stages [@Monaghan2014].
-There is also some evidence that telomeres can increase, as well as decrease, in length [@Simons2014], which has important ramifications for our understanding of how telomeres reflect costs; however, telomere lengthening is often attributed to measurement error [@Steenstrup2015].
+There is also some evidence that telomeres can increase, as well as decrease, in length [@Simons2014], which has important ramifications for our understanding of how telomeres reflect costs.
+However, increases in telomere length are often attributed to measurement error [@Steenstrup2013], and as such its ecological significance is unknown.
 Recent evidence suggests that telomere dynamics are indeed highly variable over individual lifespans, and that even the relationship between telomeres and age can vary markedly among cohorts [@Fairlie2016].
-To better understand what explains variation in telomere dynamics, more studies are required that incorporate telomere variation over entire lifepans with comprehensive, long-term ecological data.
+To better understand what explains variation in telomere dynamics, more studies are required that incorporate telomere variation over entire lifespans with comprehensive, long-term ecological data.
 
@@
 Facultative cooperative breeding occurs in the Seychelles warbler [@Komdeur1994a; @Richardson2003a], and the presence of helping subordinates (but not non-helping subordinates) in the natal territory is associated with increased survival later in life [@Brouwer2012].
-Lastly, we have an established protocol for assessing telomere length in this species [@Barrett2012; @Bebbington2016], and telomere length predicts survival independently of age, suggesting that telomeres act as a biomarker of cost in this species [@Barrett2013a].
+Lastly, we have an established protocol for assessing telomere length in this species [@Barrett2012; @Bebbington2016], and telomere length predicts survival independently of age, suggesting that telomeres act as a biomarker of cost in this species [@Barrett2013a; @Hammers2015; @Bebbington2016].
 Thus, we have an excellent system in which to assess the costs of different social and environmental conditions experienced by individuals, and to assess how these costs vary over space and time.
@@
 115 territories lay one or, rarely, two or three eggs [@Komdeur1991].
-As a result of this low reproductive output, combined with higher mortality in first-year birds [39%; @Brouwer2006], cohort sizes in the Seychelles warbler are small (< 50).
+As a result of this low reproductive output, combined with higher mortality in first-year birds [39%; @Brouwer2006], cohort sizes in the Seychelles warbler are typically small (< 50).
 
@@
 breeding male/female, helping subordinate, non-helping subordinate) and individual survival [@VandeCrommenacker2011; @Barrett2013a].
-The isolated nature of the Cousin population is a key advantage of the system for analyses involving survival, which in other systems are often confounded by emigration [see @Ergon2014 for a recent discussion].
-Full details of catching and monitoring methods can be found in Brouwer *et al.* [-@Brouwer2012].
+Full details of monitoring methods can be found in Brouwer *et al.* [-@Brouwer2012].
 
@@
 Cousin is subject to intra- and inter-annual variation in rainfall and food availability, and such island-wide temporal variation may override the effects of individual territory quality.
-As an estimate of seasonal variation in food availability, we calculated an index of the number of insects across the entire island during each breeding season.
+As an estimate of seasonal variation in food availability, we calculated an index of the number of insects across the entire island during each breeding season (referred to hereafter as 'insect abundance').
 This index is calculated as the mean number of insects found per unit leaf area over all surveys carried out on the island in a breeding season.
@@
 Each time a bird is caught on Cousin body mass and tarsus length are measured (to the nearest 0.1g and 0.1mm, respectively).
-Using information on eye colour [@Komdeur1991a] and previous captures [@Richardson2003b], we grouped birds into three age categories: one month old (birds still in the nest), 6 months old (fledglings with light grey eyes) and 10 months old (subadults with light brown eyes).
+We aged all birds using information on eye colour [@Komdeur1991a] and previous captures [@Richardson2003b].
 A blood sample (*ca* 25 $\mu$l) is taken from each bird captured via brachial venipuncture, and stored at room temperature in 1 ml of absolute ethanol in a 1.5 ml screw-cap microfuge tube.
@@
 
-For a subset of birds we had longitudinal data, with an additional samples taken at different ages.
-For these individuals we calculated the within-individual change in RTL by subtracting RTL at time point *t* from RTL at timepoint *t* + 1 (hereafter $\Delta$RTL).
-Negative values of $\Delta$RTL reflect telomere shortening, while positive values reflect telomere lengthening.
+For a subset of birds we had longitudinal data, with one or more additional samples taken at different ages (n = `r length(unique(c(dd3$BloodID,dd3$BloodID1)))` measurements from `r length(unique(dd3$BirdID))` birds).
+For these individuals we calculated the within-individual change in RTL by subtracting RTL at time point *t* from RTL at time point *t* + 1 (hereafter $\Delta$RTL, n = `r nrow(dd3)` measurements).
+Negative values of $\Delta$RTL reflect decreases in telomere length with age, while positive values increases.
 
 #####Statistical analyses
-We performed all statistical analyses using R.
+We performed all statistical analyses using R [@RDevelopmentCoreTeam2011].
+RTL was square root transformed to improve linear model fits.
 We first explored the relationship between RTL and age among cohorts using linear mixed models (LMMs) carried out in the lme4 package [@Bates2014].
 Following a similar approach to Fairlie *et al.* [-@Fairlie2016], we compared a selection of models fitting different relationships between RTL and age.
-We created models where age was fitted as a linear covariate, where age was log-transformed, and where age was fitted as a factor.
-We also created models including birth year (cohort) as a factor, and ones which included interactions between cohort and each measure of age.
+We created models where the relationship between RTL and age was linear, quadratic, log-linear, and where age was fitted as a factor.
+For each age term, we fitted additional models including birth year (cohort) as a factor, and an interactions between cohort age.
 All fitted models are included in Table 1.
@@
 Models were compared using AIC with correction for finite sample size (AICc).
-Using the longitudinal data, we tested whether telomores shorten with age in individuals, using LMMs of RTL and $\Delta$age [a longitudinal measure based on within-subject centring; @VandePol2009].
+Using the longitudinal data, we tested how telomeres change with age in individuals, using LMMs of RTL and $\Delta$age [a longitudinal measure based on within-subject centring; @VandePol2009].
 We calculated $\Delta$age using log and polynomial transformed age data, and carried out model selection as above, with the exception that we did not model $\Delta$age as a factor (due to a lack of discreet groupings).
 
-We used our longitudinal data to determine individual-level consistency in RTL.
+We then used longitudinal data to determine individual-level consistency in RTL.
 We constructed a LMM with RTL at time *t* + 1 as the response variable, RTL at time *t* and age as fixed effects, and individual ID and cohort as random effects.
-Because, when examining the distribution longitudinal telomere changes, we observed apparent telomere lengthening in our data, we used the entire Seychelles warbler dataset to test whether this lengthening could be explained by measurement error.
-We compared longitudinal changes in telomere length among different samples from the same birds to samples where we had at least two repeat telomere measurements for the same sample (i.e.
-completely seperate reactions run on separate plates: N = `r nrow(subset(ddL,Group == 'Within samples'))` measurements from `r length(unique(subset(ddL,Group == 'Within samples')$BirdID))` samples).
-We calculated $\Delta$RTL between pairs of repeat measurements within samples in exactly the same way as for across samples, except that repeat measurements were ordered by the date at the qPCR was run rather than by age.
-We first compared the variance in $\Delta$RTL within and among individuals using a Levene's test.
-Then, to separately test whether the extent of telomere shortening and lenghtening within individuals was greater than expected by measurement error, we split $\Delta$RTL measurements into groups in which shortening ($\Delta$RTL < 0) and lengthening $\Delta$RTL ($\Delta$RTL > 0) occurred, and tested whether $\Delta$RTL among samples was significantly different between within-sample and across-sample measurements, using Wilcoxon tests.
+To obtain an estimate of explanatory power, we calculated marginal R^2^ (just incorporating fixed effects) following Nakagawa & Schielzeth [-@Nakagawa2013].
+When examining the distribution longitudinal telomere changes we observed increases in telomere length with age, so we used repeat measurements to test whether these increases could be explained by measurement error.
+We calculated $\Delta$RTL between pairs of repeat measurements within samples in exactly the same way as for across samples (i.e.
+completely separate reactions run on separate plates: N = `r nrow(subset(ddL,Group == 'Within samples'))` measurements from `r length(unique(subset(ddL,Group == 'Within samples')$BirdID))` samples), and compared the variance in $\Delta$RTL within and among individuals using a Levene's test.
+Then, to separately test whether the extent of telomere increases and decreases within individuals was greater than expected by measurement error, we split $\Delta$RTL measurements into groups in which RTL decreased ($\Delta$RTL < 0) and increased $\Delta$RTL ($\Delta$RTL > 0), and tested whether $\Delta$RTL among samples was significantly different between within-sample and across-sample measurements, using Wilcoxon tests.
 
 We used LMMs to test how variation in early-life environmental and social conditions influenced telomere length within cohorts.
-We created a full model with RTL as a response variable, alongside the following explanatory variables: age, tarsus length, body mass, sex, annual food availability, territory quality, population density (an annual measure estimated from the summer breeding census) and the number of birds and the number of helping  subordinate birds present in the territory.
-Random effects structure was informed by the analysis of telomere dynamics and age (see results): we included individual ID, cohort ID, and allowed the slope of the age effect to vary among cohorts.
+We created a full model with RTL as a response variable, alongside the following explanatory variables: age, tarsus length, body mass, sex, insect abundance, territory quality, island-wide population density (an annual measure estimated from the summer breeding census), territory group size, and the number of helping subordinate birds present in the territory.
+The random effects structure was informed by the analysis of telomere dynamics and age (see results): we included individual ID, cohort ID, and a random slope of age among cohorts.
 We report model estimates and confidence intervals for all effects included in the full model.
-To obtain an estimate of explanatory power, we calculated marginal R^2 (just incoporatng fixed effects) and conditional R^2 (incorporating fixed and random effects) following Nakagawa & Schielzeth [-@Nakagawa2013] and Johnson [-@Johnson2014], respectively As a complementary approach, we also performed model averaging, using the MuMIn package in R [@Barton2012].
+We also calculated marginal R^2^ and conditional R^2^ [incorporating fixed and random effects; @Johnson2014] to assess the explanatory power of these models.
+As a complementary approach, we also performed model averaging, using the MuMIn package in R [@Barton2012].
 Model selection was performed using the full model described above and a top model set defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011].
 We report model-averaged coefficients, confidence intervals and 'relative importance', which reflects the relative weights of each predictor variable across the top model set.
-For individuals with longitudinal data (*n* = `r nrow(dd3)`) we repeated the above analyses of telomere dynamics, replacing telomere length with $\Delta$RTL as the response variable, and excluding the random effects for plate ID (as each measurement of $\Delta$RTL was based on two or more measurements, and so run on multiple plates) and cohort ID (as longitudinal telomere dynamics did not differ among cohorts; see results).
+For individuals with longitudinal data we repeated the above analyses of telomere dynamics, replacing telomere length with $\Delta$RTL as the response variable, and excluding the random effects for plate ID (as each measurement of $\Delta$RTL was based on two or more measurements, and so run on multiple plates) and cohort ID (as longitudinal telomere dynamics did not differ among cohorts; see results).
 
 ###Results
-We measured telomere lengths using a total of `r nrow(dd)` unique samples from juvenile and adult Seychelles warblers from `r length(table(dd$Cohort))` cohorts born between `r min(dd$LayYear)` and `r max(dd$LayYear)`.
+We measured telomere lengths using a total of `r length(unique(dd$BloodID))` unique samples from juvenile and adult Seychelles warblers from `r length(table(dd$Cohort))` cohorts born between `r min(dd$LayYear)` and `r max(dd$LayYear)`.
 Efficiencies (mean $\pm$ s.d.) for our telomere and GAPDH reactions were (`r round2(mean(dd$EffTelomere))` $\pm$ `r round2(sd(dd$EffTelomere))`) and (`r round2(mean(dd$EffGAPDH))` $\pm$ `r round2(sd(dd$EffGAPDH))`) respectively.
@@
 We first tested how RTL was related to age among cohorts using a model selection approach.
-We found that the top model, by some distance, contained a log age, cohort ID, and a log age x cohort interaction (Table 1A).
+We found that the top model, by some distance, contained a loglinear relationship between RTL and age, as well as a log age x cohort interaction (Table 1A).
 The second best model contained log age and cohort ID, with no interaction term, although this was substantially poorer in terms of model fit ($\Delta$AICc > 10), and all other models were much poorer still ($\Delta$AICc > 35; Table 1).
-The log-linear relationship between RTL and age couls be clearly seen in the raw data; RTL decreased with age (estimate = `r round3(fixef(RTL_LogAge)['LogAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_LogAge)['LogAge',],lessthan = F),collapse = ', ')`), with the greatest decrease occurring early in life (Fig.
+The log-linear relationship between RTL and age could be seen clearly in the raw data; RTL decreased with age (estimate = `r round3(fixef(RTL_LogAge)['LogAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_LogAge)['LogAge',],lessthan = F),collapse = ', ')`), with the greatest decrease occurring early in life (Fig.
 1A).
@@
 A within-individual analysis of RTL and age revealed that the top model explaining RTL contained $\Delta$log age, which reflects within-individual changes in log-transformed age (Table 1B).
-Models including cohort ID and cohort x age interactions were substantially poorer than models only containing age (Table 1B).
-RTL dereased with $\Delta$log age (estimate = `r round3(fixef(RTL_DeltaLogAge)['DeltaLogAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_DeltaLogAge)['DeltaLogAge',],lessthan = F),collapse = ', ')`), confirming that within-individual telomere shortening occurs across the Seychelles warbler dataset.
+Models including cohort ID and cohort x age interactions were substantially poorer fits than models only containing age (Table 1B).
+RTL decreased with $\Delta$log age (estimate = `r round3(fixef(RTL_DeltaLogAge)['DeltaLogAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_DeltaLogAge)['DeltaLogAge',],lessthan = F),collapse = ', ')`), confirming that within-individual telomere shortening occurs across the Seychelles warbler dataset.
 
-There was positive correlation between telomere measurements taken from different timepoints (Fig.
+There was positive correlation between telomere measurements taken from different time points (Fig.
 2A), but this was very weak (marginal R^2^ = `r round2(r.squaredGLMM(RTL_RTL1)['R2m'])`), and not quite significant (estimate = `r round3(fixef(RTL_RTL1)['RTL'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_RTL1)['RTL',],lessthan = F),collapse = ', ')`).
 Although both cross-sectional and longitudinal data indicated a general trend of telomere shortening with age, we found that RTL within individuals increased in `r round((nrow(subset(dd3,DeltaRTL>0))/nrow(dd3))*100,0)`% of our `r nrow(dd3)` $\Delta$RTL measurements (Fig.
-2A), which is indicative of telomere lengthening.
-To test whether this apparent telomere lengthening in our dataset could be explained by measurement error, we compared variance in telomere length among repeat measurements of samples to the variance observed among different samples of the same individual.
+2A).
+To test whether increases in telomere length in our dataset could be explained by measurement error, we compared variance in telomere length among repeat measurements of samples to the variance observed among different samples of the same individual.
 We found significantly higher variance in telomere length over individual lifetimes compared to among sample replicates (Levene's test: F = `r round2(leveneTest(DeltaRTL~Group,data=ddL)['F value'][1,1])`; P `r round3(leveneTest(DeltaRTL~Group,data=ddL)['Pr(>F)'][1,1])`; Fig.
 2B).
-Splitting the longitudinal data into instances of shortening (i.e.
-$\Delta$RTL < 0) and lenghtning (i.e.
-$\Delta$RTL > 0) revealed that not only did we observe significantly more shortening within individuals compared to within samples (Wilcoxon test: P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$p.value)`), but also significantly more lengthening (P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$p.value)`; Fig.
+Splitting the longitudinal data into instances of decreasing (i.e.
+$\Delta$RTL < 0) and increasing (i.e.
+$\Delta$RTL > 0) telomere length revealed that not only did we observe significantly greater decrease in RTL within individuals compared to within samples (Wilcoxon test: P `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$p.value)`), but also a significantly greater increase (P `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$p.value)`; Fig.
 2B).
@@
 #####Telomere dynamics and the environment
-In addition to age, tarsus length, sex and insect abundance were significantly associated with RTL (Fig.
+In addition to age, RTL was associated with tarsus length, sex and insect abundance (Fig.
 3A).
@@
 3C).
-The full model was weak in terms of explanatory power of fixed effects (marginal R^2^ = `r round2(r.squaredGLMM(RTL_Full)['R2m'])`), although including the random effect terms increased this substantially (conditional R^2^ = `r round2(r.squaredGLMM(RTL_Full)['R2c'])`) The model averaging approach yielded qualitatively identical results to the full LMM, with the same explanatory variables 'significant' in terms of being retained in top models, and having model-averaged confidence intervals not overlapping zero (Table S1; Fig.
+The full model was fairly weak in terms of explanatory power of fixed effects (marginal R^2^ = `r round2(r.squaredGLMM(RTL_Full)['R2m'])`), although including the random effect terms increased this substantially (conditional R^2^ = `r round2(r.squaredGLMM(RTL_Full)['R2c'])`) The model averaging approach yielded qualitatively identical results to the full LMM, with the same explanatory variables 'significant' in terms of being retained in top models, and having model-averaged confidence intervals not overlapping zero (Table S1; Fig.
 S1).
@@
 ###Discussion
-Here we use a long-term, multi-cohort dataset to assess the relationships between the spatiotemporal variation in the ecological environment and lifelong telomere dynamics in a closed population of Seychelles warblers.
+Here we use a long-term, multi-cohort dataset to assess the relationships between spatio-temporal variation in the ecological environment and lifelong telomere dynamics in a closed population of Seychelles warblers.
 We found that telomere length decreases with age, and that this decrease is greatest in early life.
-Telomere length decreased with age in almost all `r length(table(dd$Cohort)` cohorts studied, but telomere length, and the rate of decrease of telomere length with age, varied substantially among cohorts.
-Despite an overall trend for telomere shortening in the Seychelles warbler, we found that bouts of apparent telomere lengthening occur within individuals, and the extent of this lengthening cannot be explained solely by qPCR measurement error.
-Finally, we found that telomeres are related to tarsus length in a sex-specific manner, and that telomer length is positively associated with temporal fluctuations in food availability.
+Telomere length decreased with age in almost all `r length(table(dd$Cohort))` cohorts studied, but telomere length, and the rate of decrease of telomere length with age, varied substantially among cohorts.
+Despite an overall trend for telomere shortening in the Seychelles warbler, we found that telomere length increased within some individuals at particular life stages, and that the extent of these increases cannot be explained solely by qPCR measurement error.
+Finally, we found that telomeres are related to tarsus length in a sex-specific manner, and that telomere length is positively associated with temporal fluctuations in food availability.
 
@@
 @Frenck1998; @Haussmann2003; @Heidinger2012].
-However, we found that telomere shortening was only consistently observed early in life, after which both shortening and lengthening occurred.
+However, we also found that despite an overall trend for shortening, telomere length both increased and decreased, especially after the juvenile period.
 Importantly, these increases were observed in longitudinal as well as cross-sectional data, suggesting that selective disappearance does not explain this pattern entirely.
-Longitudinal increases in measured telomere length have been observed in humans and [reviewed in Steenstrup2015] wild animals [@Kotrschal2007; @Fairlie2016].
-The most commonly invoked explanation for increases in telomere length is measurement error, which can be a particular problem in qPCR-based telomere studies [@Nussey2014; @Steenstrup2015; @Verhulst2015].
+Longitudinal increases in measured telomere length have been observed in humans and [reviewed in @Steenstrup2013] wild animals [@Kotrschal2007; @Fairlie2016].
+The most commonly invoked explanation for increases in telomere length is measurement error, which can be a particular problem in qPCR-based telomere studies [@Nussey2014; @Steenstrup2013; @Verhulst2015].
 To our knowledge, ours is the first study to explicitly compare intra-individual variation among samples to variation among sample replicates, on a large scale.
@@
 
-Our results (Fig.
-1A) suggest that increases in telomere length occur at specific, short periods throughout individual lifespans, against a backdrop of overall telomere shortening.
-Consistent with a dynamic pattern of telomere shortening with age, we found that within-individual telomere measurments were only weakly correlated.
-These findings are in contrast to other avian studies where birds were reared in laboratory [@Heidinger2012] or nest-box [@Boonekamp2014] conditions, in which within-individual telomere length measurements were highly consistent, and individual-level telomere shortening occured throughout the juvenile period and into adulthood.
+We found that increases in telomere length are not consistent over individual lifespans, but that increases occur at specific, short periods, against a backdrop of overall telomere shortening.
+Consistent with a pattern of sporadic changes in telomere length with age, we found that within-individual telomere measurements were only weakly correlated.
+These findings are in contrast to other avian studies where birds were reared in laboratory [@Heidinger2012] or nest-box [@Boonekamp2014] conditions, in which within-individual telomere length measurements were highly consistent, and individual-level telomere shortening occurred throughout the juvenile period and into adulthood.
 However, the lifelong telomere dynamics found in Seychelles warblers are strikingly similar to those found in Soay sheep [@Fairlie2016], and we anticipate that a similar pattern may be found in other wild populations.
-That increases in telomere length may be sporadic and interspersed by bouts of shortening is a crucial point, because previously described approaches to distinguish telomere elongation from measurement error based on assumptions about follow-up time between measurements [@Steenstrup2015], or based on measuring variance among measurements [@Simons2014] assume the telomere elongation within individuals is consistent over time.
-Such sporadic changes in telomere length over lifespans could occur due to a change in the composition of cell types within individual samples, or due to the actual elongation of telomeres.
+That increases in telomere length may be sporadic and overlaid onto an overall pattern of shortening is a crucial point, because previously described approaches to distinguish telomere elongation from measurement error, based on assumptions about follow-up time between measurements [@Steenstrup2013], or based on measuring variance among measurements [@Simons2014], assume the telomere elongation within individuals is consistent over time.
+Such sporadic changes in telomere length over lifespans could occur due to changes in the composition of cell types within individual samples, or due to the actual elongation of telomeres.
 Determining the mechanism of these changes is essential for how we view telomeres as biomarkers of costs.
-If telomeres can be lengthened in response to improvements in environmental conditions, for example, this would suggest the they reflect short-to medium-term costs, rather than the  the cumulative costs that an individual has faced over its lifespan [@Bateson2016].
-New statistical and technical approaches are therefore now required to determine the mechanisms behind increases in telomere length within individuals.
+For example, if telomeres can be lengthened in response to improvements in environmental conditions, this would suggest that they reflect short- to medium-term costs, rather than the cumulative costs that an individual has faced over its lifespan [@Bateson2016].
+New statistical and technical approaches are therefore now required to determine the mechanisms behind increases in telomere length within individuals, so that biologically informed hypotheses about the ecological causes and consequences of these increases can be generated and tested.
 
@@
 
-The long-term nature of the Seychelles warbler study enabled us to show that temporal variation in food availability was positively related to telomere length.
+We found that temporal variation in food availability was positively related to telomere length.
 This is consistent with strong cohort effects, and suggests that temporal variation in conditions may be a key driver of costs in the Seychelles warbler.
@@
 3B).
-One possibility is that the environment imposes differential costs on males and females: a recent study in captive zebra finches found that manipulation of dietary nutrients had sex-sependent affects on telomere dynamics [@Noguera2015].
+One possibility is that the environment imposes differential costs on males and females: a recent study in captive zebra finches found that manipulation of dietary nutrients had sex-dependent affects on telomere dynamics [@Noguera2015].
 Also worth noting is that the effect of telomere length on survival in strongest in male Seychelles warblers [@Barrett2013a], although the nature of the relationship between sex, telomeres and survival is not yet clear [@Barrett2011]
 
-Although we found clear associations between the environment and telomere dynamics, we should bear in mind that our social and ecological variables explained a small proportion of the variance in RTL.
+Although we found clear associations between the environment and telomere dynamics, we should bear in mind that our social and ecological variables explained a modest proportion of the variance in RTL.
 A poor social and ecological environment is known to be detrimental to Seychelles warblers, both in terms of oxidative stress and survival [@VandeCrommenacker2011a; @Brouwer2012], and it is therefore perhaps surprising that these variables do not explain more variance in RTL.

### 2016-11-25 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Eds
EDITS BEFORE DRAFT TO DAVE

@@
 table1A[,3:5] <- lapply(table1A[,3:5],as.numeric)
+table1A[,3:5] <- round(table1A[,3:5],3)
 
@@
 table1B[,3:5] <- lapply(table1B[,3:5],as.numeric)
+table1B[,3:5] <- round(table1B[,3:5],3)
 
 #Sort out rounding and exlcude NAs from blank spaces. Clunky but works
-table1 <- rbind(c('A',rep(' ',4)),table1A,c('--',rep(' ',4)),c('B',rep(' ',4)),table1B)
-table1[,3:5] <- lapply(table1[,3:5],as.numeric)
-table1[,3:5] <- round(table1[,3:5],3)
-table1[,3:5] <- lapply(table1[,3:5],as.character)
-table1[is.na(table1)] <- " "
+table1 <- rbind(c('A',rep('-',4)),table1A,c('B',rep('-',4)),table1B)
 
@@
 
+RTL_Full2 <- standardize(RTL_Full)
+
 CIs <- CIwald(RTL_Full2)

### 2016-11-25 Scripts/SW juvenile telomeres IPC models.r

@@ -39,5 +39,3 @@ RTL_Full_SexTarsus <- lmer(RTL ~  LogAge + Helper + GroupSize + Sex + BodyMass +
 DeltaRTL_Full <- lmer(DeltaRTL ~  LogAge + Tarsus + Helper + GroupSize + Sex + BodyMass + Insect + Density + TQ + (1|BirdID),
                       data = dd3NA,
                              REML = FALSE)
-
-

### 2016-12-06 Manuscript/Figs for Dave.rmd
Figs for Dave

--- 2016-11-25 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
@@
 ---
+
 ```{r setup, include = F}
@@
 
-Fig1A <- ddply(dd,
-               .(BirdID),
-        mutate,
-        nreps = length(RTL)) %>%
-  subset(nreps>0) %>%
-  subset(!duplicated(paste0(BirdID,Agemonths))) %>%
-  ggplot(aes(x = Agemonths,y = RTL,group = BirdID,col=factor(BirdID)))+
-  geom_point(alpha = 1,col='grey')+
-geom_line(col = 'darkgrey',alpha = 0.2)+
-  xlab('Age (years)') +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black'),
-        legend.position = 'none')+
-  geom_smooth(col = 'black',fill = 'black',lwd = 0.5,aes(group = NULL),method = 'lm',formula = y~log(x))+
-  annotate('text',x = 0,y = 1.7,label = 'A')
+pdf('Figures/Figures for Dave NERC grant.pdf',
+    width = 5,
+    height = 5,
+    paper = 'a4')
 
@@
 
-Fig1B <- transform(dd,Cohort = factor(LayYear)) %>%
-  ggplot(aes(y = RTL,x = Cohort, fill = Cohort))+
+transform(dd,Cohort = factor(LayYear)) %>%
+  ddply(.(LayYear),
+        summarise,
+        meanRTL = mean(RTL),
+        seRTL = se(RTL)) %>%
+  ggplot(aes(y = meanRTL,x = LayYear))+
   xlab('Birth year') +
-  geom_boxplot(outlier.shape = NA) +
+  ylab('RTL')+
+  geom_point()+
+  geom_line()+
+  geom_errorbar(aes(ymax = meanRTL+seRTL,ymin=meanRTL-seRTL),width =0.2)+
   theme_classic() +
@@
         legend.position = 'none') +
-        scale_fill_manual(values = c15)+
-  annotate('text',x = 1,y = 1.6,label = 'B')
+  scale_fill_manual(values = c15)
 
-Fig1C <- transform(dd,Cohort = factor(LayYear)) %>%
+transform(dd,Cohort = factor(LayYear)) %>%
+  subset(Cohort %in% c(2001,2005,2010)) %>%
   ggplot(aes(y = RTL,x = Agemonths, col = Cohort,fill = Cohort))+
   xlab('Age (years)') +
-  geom_smooth(method = 'lm', formula = y~log(x),se = F) +
+  xlim(c(0,6))+
+  geom_smooth(method = 'lm', formula = y~log(x),se = T,fullrange=T,alpha = 0.1) +
   theme_classic() +
@@
         axis.line.y = element_line(colour = 'black'),
-        legend.position = 'none') +
+        legend.position = 'top') +
   scale_color_manual(values = c15)+
-        scale_fill_manual(values = c15)+
-  annotate('text',x = 0,y = 1.3,label = 'C')
+  scale_fill_manual(values = c15)
 
-  Fig1D <- ggplot(dd3, aes(x = DeltaLogAge,y = RTL))+
-  geom_point(alpha = 1,col='grey')+
-  xlab(expression(Delta*'Log age (years)')) +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black'),
-        legend.position = 'none')+
-  geom_smooth(col = 'black',method = 'lm')+
-  annotate('text',x = -1.5,y = 1.7,label = 'D')
-
-pdf('Figures/Figure 1.pdf',
-    width = 8,
-    height = 8,
-    paper = 'a4')
-multiplot(Fig1A,Fig1B,Fig1C,Fig1D,layout = matrix(1:4,2,2,byrow = T))
-dev.off()
-
-```
-
-```{r Figure 2}
-
-Fig2A <- ggplot(dd3,
-       aes(x = RTL,
-           y = RTL1)) +
-  ylab(expression('RTL ('*italic(t)*' + 1)')) + 
-  xlab(expression('RTL ('*italic(t)*')')) +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        legend.position = c('none'),
-        legend.title = element_text(size = 12),
-        axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black')) +
-    geom_point(col = 'grey')+
-  geom_abline(slope = 1,intercept = 0,lty = 2) +
-  annotate('text',x = 0.3,y = 1.7,label = 'A')
-
-Fig2B <- ggplot(ddL,aes(col=Group,fill = Group,y = ..scaled..,x = DeltaRTL))+
-    geom_density(alpha = 0.2)+
-  theme_classic()+
-  ylab ('Scaled density')+
-  xlab(expression(Delta*'RTL'))+
-    theme(axis.text.x = element_text(size = 12), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-        legend.position = 'none',
-        legend.title = element_blank(),
-        axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black')) +
-  scale_colour_manual(values = c('darkgrey','gold'))+
-  scale_fill_manual(values = c('darkgrey','gold'))+
-  geom_vline(xintercept = 0,lty = 2) +
-  annotate('text',x = -1,y = 1,label = 'B')
-
-pdf('Figures/Figure 2.pdf',
-    width = 4,
-    height = 8,
-    paper = 'a4')
-multiplot(Fig2A,Fig2B)
-dev.off()
-
-```
-
-```{r Figure 3}
-
-RTL_Full2 <- standardize(RTL_Full)
-
-CIs <- CIwald(RTL_Full2)
-t1a <- data.frame(myx = row.names(CIs)[7:nrow(CIs)],
-                  Estimate = fixef(RTL_Full2),
-                  X2.5. = CIs[7:nrow(CIs),1],
-                  X97.5. = CIs[7:nrow(CIs),2],
-                  row.names = NULL,
-                  stringsAsFactors = F)
-
-nams <- t1a$myx
-nams[nams == 'z.LogAge'] <- 'Age'
-nams[nams == 'z.Tarsus'] <- 'Tarsus length'
-nams[nams == 'c.Helper'] <- 'Number helpers'
-nams[nams == 'z.GroupSize'] <- 'Group size'
-nams[nams == 'c.Sex'] <- 'Sex (males)'
-nams[nams == 'z.BodyMass'] <- 'Body mass'
-nams[nams == 'z.Insect'] <- 'Insect abundance'
-nams[nams == 'z.TQ'] <- 'Territory quality'
-nams[nams == 'z.Density'] <- 'Density'
-
-t1a$myx <- nams
-
-t1a$myx <- factor(t1a$myx,levels = t1a$myx[order(t1a$Estimate,decreasing = T)])
-t1a <- subset(t1a,myx != "(Intercept)")
-
-Fig3A <- forestplot(t1a,RI = F)+
-  annotate('text',y = 0.1,x = 0.6,label = 'A')
-
-Fig3B <- ggplot(dd,
-       aes(x = Tarsus,
-           y = RTL,
-           col = Sex)) +
-  geom_point()+
-  ylab('RTL') + 
-  xlab('Tarsus length (mm)') +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12), 
-        axis.text.y = element_text(size = 12), 
-        axis.title.y = element_text(size = 14,vjust=0.8),
-        axis.title.x = element_text(size = 14),
-                axis.line.x = element_line(colour = 'black'),
-        axis.line.y = element_line(colour = 'black'),
-        legend.position = 'none') +
-  stat_smooth(method = 'lm') +
-  scale_color_manual(values = c('gold2','red'))+
-  annotate('text',x = 17,y = 1.7,label = 'B')
-
-Fig3C <- ggplot(dd,
-       aes(y = RTL,
+ddply(dd,
+      .(Insect),
+      summarise,
+      meanRTL = mean(RTL),
+      seRTL = se(RTL)) %>%
+ggplot(aes(y = meanRTL,
               x=Insect)) +
@@
         legend.position = 'none') +
-  geom_boxplot(aes(group = Insect),outlier.shape = NA)+
-   stat_smooth(method = 'lm',col = 'darkgrey') +
-  annotate('text',x = 2.5,y = 1.6,label = 'C')
+  geom_point()+
+  geom_errorbar(aes(ymin=meanRTL-seRTL,ymax=meanRTL+seRTL))+
+  stat_smooth(method = 'lm',col = 'darkgrey')
 
-pdf('Figures/Figure 3.pdf',
-    width = 8,
-    height = 8,
-    paper = 'a4')
-multiplot(Fig3A,Fig3B,Fig3C,layout = matrix(c(1,1,2,3),byrow = T,nrow = 2))
 dev.off()

### 2017-01-12 Manuscript/SW_early_life_telomeres.Rmd
More eds

@@
 1.
-Understanding the costs that individuals and populations face throughout their lifetimes is fundamental to understanding life-history evolution and population dynamics.
+Understanding the costs that individuals face throughout their lifetimes is fundamental to understanding life-history evolution and population dynamics.
 Telomeres, the protective caps at the ends of chromosomes, shorten in response to oxidative stress, and telomere shortening is correlated with reduced survival.
-Thus, telomere dynamics may help us quantify individual variation in costs, and enhance our understanding of individual variation in natural populations.
+Investigating telomere dynamics may help us quantify individual variation in the costs of the social and ecological environment, and enhance our understanding of the dynamics of natural populations.
 

### 2017-01-12 Manuscript/SW_early_life_telomeres.Rmd
Home edits

@@
 Here we study spatio-temporal variation in lifelong telomere dynamics in the Seychelles warbler (*Acrocephalus sechellensis*).
-We combine long-term ecological data with one the largest longitudinal telomere datasets to date, consisting of `r nrow(dd)` samples from `r length(table(dd$Cohort))` cohorts born between `r min(dd$LayYear)` and `r max(dd$LayYear)`.
-We provide a detailed analysis of how telomere dynamics vary over individual lifespans, and with spatio-temporal variation in the social and ecological environment.
+We combine long-term life history and ecological data with one the largest longitudinal telomere datasets to date, consisting of `r nrow(dd)` samples from `r length(table(dd$Cohort))` cohorts born between `r min(dd$LayYear)` and `r max(dd$LayYear)`.
+We provide a detailed analysis of how telomere dynamics vary over individual lifespans and cohorts, and with spatio-temporal variation in the social and ecological environment.
 
@@
 However, both cross-sectional and longitudinal data suggested that against this overall pattern of shortening, bouts of telomere length increases occur in some individuals.
-Using a large number of repeated measurements we show statistically that these increases cannot be explained solely by qPCR measurement error.
+Using a large number of repeated measurements we show statistically that these increases are unlikely to be explained solely by qPCR measurement error.
 
@@
 Telomere length decreased with age in almost all cohorts studied, but telomere length and the slope of the relation ship with age varied markedly among cohorts.
-Variation in telomere length was positively associated with island-wide temporal variation in insect abundance, suggesting that temporal variation in food availability imposes a cost in Seychelles warblers.
+Telomere length was positively associated with island-wide temporal insect abundance, suggesting that the costs associated with living in a harsher environment can be studied by investigating telomere dynamics.
 We also found evidence for sex-specific relationships between telomeres and tarsus length, potentially reflecting differential costs of growth.
@@
 5.
-Our long-term data show that in natural populations, telomere dynamics vary in a highly complex manner over individual lifespans, and across space and time.
+Our long-term data show that in a natural population, telomere dynamics vary in a highly complex manner over individual lifespans, and across space and time.
 Variance in telomere dynamics among individuals is the product of wide array of genetic, parental and environmental factors.
@@
 
-**Keywords:** Life-history; Telomere; Seychelles warbler; Senescence
+**Keywords:** Biomarkers; Intra- and inter-individual variation; Life-history; Telomere; Seychelles warbler; Senescence
 
@@
 ###Introduction
-A major aim of ecologists is to be able to quantify and understand why individuals vary in their response to different environments.
-Identifying the costs imposed on individuals within populations is central to understanding variation in fitness [@Lindstrom1999], and thus for understanding population and community dynamics [@Bolnick2011].
+A major aim of ecologists is to quantify and understand why individuals vary in their response to different environments.
+Identifying the costs imposed on populations is central to understanding variation in fitness [@Lindstrom1999], and thus for understanding population and community dynamics [@Bolnick2011].
 However, fully quantifying individual-level variation in costs is impossible in wild systems, and thus effective biomarkers that reflect the physiological consequences of individual-level experiences are required.
 
-Telomeres are repetitive DNA sequences found on the ends of linear chromosomes that protect against DNA damage.
-Telomeres shorten with age [@Monaghan2006], and there is evidence from a range of taxa that the greatest rate of telomere shortening occurs in early life [e.g.
+Telomeres have been proposed to be a potential biomarker of such costs [@Monaghan2014].
+Telomeres are repetitive DNA sequences on the ends of linear chromosomes that protect against DNA damage.
+Telomeres generally shorten with age [@Monaghan2006], and there is evidence from a range of taxa that the greatest rate of telomere shortening occurs in early life [e.g.
 @Frenck1998; @Haussmann2003].
-Telomere shortening also occurs in response to oxidative stress, which can be elevated due to environmental factors [@VonZglinicki2002].
+Telomere shortening occurs in response to oxidative stress, which can be elevated due to environmental factors [@VonZglinicki2002].
 There is evidence from both humans and wild animal populations that telomere shortening is influenced by the conditions experienced during both early life and adulthood [@Price2013; @Monaghan2014; @Nettle2015; @Reichert2015].
-Importantly, the extent of telomere shortening that occurs has a direct link to senescence and survival.
+Importantly, the extent of telomere shortening is directly linked to senescence and survival.
 When telomeres become critically short cells senesce [@Campisi2003], and the accumulation of these cells can result in organismal senescence and death [@Wong2003].
-This association between senescence and telomere length has inspired a great deal of recent research into telomere evolutionary ecology, and relationships between telomere dynamics and survival have been documented in several wild organisms [reviewed in @Horn2010; @Haussmann2010; @Simons2015].
+
+The association between senescence and telomere length has inspired a great deal of recent research into telomere evolutionary ecology, and relationships between telomere dynamics and survival have been documented in several wild organisms [reviewed in @Horn2010; @Haussmann2010; @Simons2015].
 While there is as yet little direct evidence that the relationship between telomere dynamics and survival is causal [@Simons2015], there is mounting evidence that telomeres can act as biomarkers of cost in wild populations, providing a signature of the ecological stress that has been experienced and is otherwise difficult to detect [e.g.
 @Schultner2014; @Asghar2015].
+There is also some evidence that telomeres can increase, as well as decrease, in length [@Simons2014; @Bateson2016], which has important ramifications for our understanding of how telomeres reflect costs.
+However, increases in telomere length are often attributed to measurement error [@Steenstrup2013], and as such its ecological significance is unknown.
 
-Although there has now been a considerable amount of effort put into studying telomere dynamics in natural populations, our understanding of the forces responsible for explaining variation in telomere length is still limited.
-This is important, as before we can use telomeres as a measure of the costs experienced by individuals and populations, we need to know how different developmental, genetic and ecological variables interact to affect telomeres.
+Although a considerable amount of effort has been put into studying telomere dynamics in natural populations, our understanding of the forces responsible for explaining variation in telomere length is still limited.
+Understanding how different factors shape telomere length variation is important, as before we can use telomeres as a measure of the costs experienced by individuals, we need to know how different developmental, genetic and ecological variables interact to affect telomeres.
 Telomere length and rates of shortening can vary according to parental characteristics [Njajour2007; @Heidinger2016], among sexes [@Barrett2011], and with a whole host of environmental conditions experienced at different life-history stages [@Monaghan2014].
-There is also some evidence that telomeres can increase, as well as decrease, in length [@Simons2014], which has important ramifications for our understanding of how telomeres reflect costs.
-However, increases in telomere length are often attributed to measurement error [@Steenstrup2013], and as such its ecological significance is unknown.
 Recent evidence suggests that telomere dynamics are indeed highly variable over individual lifespans, and that even the relationship between telomeres and age can vary markedly among cohorts [@Fairlie2016].
-To better understand what explains variation in telomere dynamics, more studies are required that incorporate telomere variation over entire lifespans with comprehensive, long-term ecological data.
+To understand which factors best explain variation in telomere dynamics, more studies are required that incorporate telomere variation over entire lifespans with comprehensive, long-term ecological data.
 
-The longitudinal study (since 1986) of the Seychelles warbler (*Acrocephalus sechellensis*) population on Cousin Island provides an excellent system for studying senescence in the wild [reviewed in @Hammers2015].
-Due to the isolated nature of the study population and intensive field monitoring, we have unusually comprehensive ecological and survival data spanning many years (see Methods, below).
-Conditions and warbler population density on Cousin vary across space and time due to weather-induced changes in foliage cover and food availability [@VandeCrommenacker2011a].
-Variation in the oxidative stress experienced by individuals is associated with territory quality [@VandeCrommenacker2011a].
+The longitudinal study (since 1986) of the Seychelles warbler (*Acrocephalus sechellensis*) population on Cousin Island provides an excellent system for studying telomere dynamics and senescence patterns in the wild [reviewed in @Hammers2015].
+Due to the isolated nature of the study population and intensive field monitoring, we have comprehensive ecological and survival data spanning many years (see Methods, below).
+Environmental conditions and population density on Cousin Island vary across space and time due to weather-induced changes in foliage cover and food availability [@VandeCrommenacker2011a].
+Variation in oxidative stress experienced by individuals is associated with territory quality [@VandeCrommenacker2011a].
 However, neither early-life nor adult survival appear to be associated with territory quality or local density [@Brouwer2006; @Hammers2013].
-Facultative cooperative breeding occurs in the Seychelles warbler [@Komdeur1994a; @Richardson2003a], and the presence of helping subordinates (but not non-helping subordinates) in the natal territory is associated with increased survival later in life [@Brouwer2012].
-Lastly, we have an established protocol for assessing telomere length in this species [@Barrett2012; @Bebbington2016], and telomere length predicts survival independently of age, suggesting that telomeres act as a biomarker of cost in this species [@Barrett2013a; @Hammers2015; @Bebbington2016].
+
+ As well as territory quality variation, there is also variation in the social environment.
+Facultative cooperative breeding occurs in the Seychelles warbler [@Komdeur1994a; @Richardson2003a], and the presence of helpers (but not non-helpers) in the natal territory is associated with increased survival of offspring later in life [@Brouwer2012].
+Lastly, we have an established protocol for assessing telomere length in this species [@Barrett2012; @Bebbington2016], and telomere length predicts survival independently of age, suggesting that individual variation in telomere length is ecologically relevant in this species [@Barrett2013a; @Hammers2015; @Bebbington2016].
 Thus, we have an excellent system in which to assess the costs of different social and environmental conditions experienced by individuals, and to assess how these costs vary over space and time.
 
-In this study we study how lifelong telomere dynamics are related to environmental variation across `r length(table(dd$Cohort))` Seychelles warbler cohorts.
+In this study, we test how lifelong telomere dynamics are related to environmental variation across `r length(table(dd$Cohort))` Seychelles warbler cohorts.
 We first study how telomere length and rates of shortening are related to age, and how this relationship varies among cohorts, in order to gain an in-depth understanding of the temporal dynamics of telomere changes.
-We then examine, within individuals, how telomere length changes with age, and statistically test whether increases in telomere length are biologically meaningful.
+We then examine, within individuals, how telomere length changes with age, and statistically test whether observed increases in telomere length are larger than expected based on measurement error.
 Finally, we test how a wide range of social and environmental variables are related to telomere length and shortening in order gain a fuller understanding of the forces driving telomere dynamics in natural populations.
@@
 320 birds on Cousin Island (04'20'S, 55'40'E) has been intensively studied since 1986 [@Richardson2003b; @Spurgin2014a].
-This species' main breeding season runs from June-August, when the breeding females on each of the *ca*.
+This species' main breeding season runs from June-September, when the breeding females on each of the *ca*.
 115 territories lay one or, rarely, two or three eggs [@Komdeur1991].
-As a result of this low reproductive output, combined with higher mortality in first-year birds [39%; @Brouwer2006], cohort sizes in the Seychelles warbler are typically small (< 50).
+As a result of this low reproductive output, combined with higher mortality in first-year birds [39% in first-year birds versus 16% in adults; @Brouwer2006], cohort sizes in the Seychelles warbler are typically small (< 50).
 
-Individuals are usually ringed in their first year of life, and so are of known age.
-They are then followed throughout their lives, and as they are non-migratory endemics naturally confined to the island [@Komdeur2004a], a biannual census of birds on Cousin during each breeding season gives accurate measures of local density, social status (e.g.
-breeding male/female, helping subordinate, non-helping subordinate) and individual survival [@VandeCrommenacker2011; @Barrett2013a].
+Individuals are usually ringed (with an individually numbered metal ring and unique combination of colour rings) in their first year of life, and so are of known age.
+We aged all birds using information on eye colour [@Komdeur1991a] and previous captures [@Richardson2003b].
+As Seychelles warblers are non-migratory endemics naturally confined to the island [@Komdeur2004a], a biannual census of birds on Cousin during each breeding season gives accurate measures of local density, social status (e.g.
+breedier, helper, non-helper) and individual survival [@VandeCrommenacker2011; @Barrett2013a].
 Full details of monitoring methods can be found in Brouwer *et al.* [-@Brouwer2012].
 
-Seychelles warblers are highly territorial and all territories are mapped during the breeding seasons using detailed observational data of foraging and territorial defence behaviour, and surveyed for territory quality [@Richardson2003b].
-Territory quality is calculated based on territory size, foliage cover and insect abundance [@Komdeur1992], and where territory quality estimates were not available for a specific year we used the average value for that territory across years [@Hammers2013].
+Seychelles warblers are highly territorial and all territories were mapped during each main breeding season using detailed observational data of foraging and territorial defence behaviour, and surveyed for territory quality [@Richardson2003b].
+Territory quality is calculated based on territory size, foliage cover and insect abundance [@Komdeur1992].
+Where territory quality estimates were not available for a specific year we used the average value for that territory across years [@Hammers2013].
 Cousin is subject to intra- and inter-annual variation in rainfall and food availability, and such island-wide temporal variation may override the effects of individual territory quality.
-As an estimate of seasonal variation in food availability, we calculated an index of the number of insects across the entire island during each breeding season (referred to hereafter as 'insect abundance').
-This index is calculated as the mean number of insects found per unit leaf area over all surveys carried out on the island in a breeding season.
+As an estimate of seasonal variation in food availability, we calculated an index of the number of insects across the entire island during each main breeding season (referred to hereafter as 'insect abundance').
+This index is calculated as the mean number of insects found per unit leaf area over all surveys carried out on the island in a main breeding season.
 
 Each time a bird is caught on Cousin body mass and tarsus length are measured (to the nearest 0.1g and 0.1mm, respectively).
-We aged all birds using information on eye colour [@Komdeur1991a] and previous captures [@Richardson2003b].
 A blood sample (*ca* 25 $\mu$l) is taken from each bird captured via brachial venipuncture, and stored at room temperature in 1 ml of absolute ethanol in a 1.5 ml screw-cap microfuge tube.

### 2017-01-16 Manuscript/SW_early_life_telomeres.Rmd
Home edits

@@
 Sex was determined using the PCR-based method outlined by Griffiths *et al.* [-@Griffiths1998].
-Prior to telomere analysis, DNA concentration and purity were quantified using a NanoDrop 8000 Spectrophotometer (ThermoScientific), and the following thresholds were applied before samples were included for further analysis: i) DNA concentration must be at least 15 ng $\mu$l^-1^ (based on a mean of three measurements), ii) the 260/280 ratio has to be between 1.8 and 2 and, iii) the 260/230 ratio should be higher than 1.8.
+Prior to telomere analysis, DNA concentration and purity were quantified using a NanoDrop 8000 Spectrophotometer (ThermoScientific), and the following thresholds were applied before samples were included for further analysis: i) DNA concentration must be at least 15 ng $\mu$l^-1^ (based on a mean of three measurements), ii) the 260/280 absorbance ratio has to be between 1.8 and 2 for acceptable DNA purity and, iii) the 260/230 absorbance ratio must be higher than 1.8.
 DNA integrity was further validated by visualization with ethidium bromide after electrophoresis on a 1.2% agarose gel, and all samples with evidence of DNA degradation were re-extracted or excluded.
 All DNA extractions that passed the above criteria were diluted to 3.3 ng $\mu$l^-1^ before telomere measurement.
-We measured relative telomere length (RTL) for all samples using a quantitative PCR (qPCR) assay of telomeres and a GAPDH control gene, using the molecular methods outlined by Bebbington *et al.* [-@Bebbington2016].
-We assessed repeatability of RTL using the rptR package in R [@RDevelopmentCoreTeam2011].
+We measured relative telomere length (RTL) for all samples using a quantitative PCR (qPCR) assay of telomeres and a GAPDH control gene, following Bebbington *et al.* [-@Bebbington2016].
 
@@
 #####Statistical analyses
-We performed all statistical analyses using R [@RDevelopmentCoreTeam2011].
-RTL was square root transformed to improve linear model fits.
-We first explored the relationship between RTL and age among cohorts using linear mixed models (LMMs) carried out in the lme4 package [@Bates2014].
+We performed all statistical analyses using R version 3.2.2 [@RDevelopmentCoreTeam2011].
+RTL was square root transformed to improve linear model fits, and we assessed repeatability of RTL using the rptR package.
+We explored the cross-sectional relationship between RTL and age among cohorts using linear mixed models (LMMs) carried out in the lme4 package [@Bates2014].
 Following a similar approach to Fairlie *et al.* [-@Fairlie2016], we compared a selection of models fitting different relationships between RTL and age.
@@
 All fitted models are included in Table 1.
-As random effects we included individual ID, catch year and qPCR plate ID.
+For random effects we included individual ID, and catch year.
 Models were compared using AIC with correction for finite sample size (AICc).
-Using the longitudinal data, we tested how telomeres change with age in individuals, using LMMs of RTL and $\Delta$age [a longitudinal measure based on within-subject centring; @VandePol2009].
-We calculated $\Delta$age using log and polynomial transformed age data, and carried out model selection as above, with the exception that we did not model $\Delta$age as a factor (due to a lack of discreet groupings).
 
-We then used longitudinal data to determine individual-level consistency in RTL.
-We constructed a LMM with RTL at time *t* + 1 as the response variable, RTL at time *t* and age as fixed effects, and individual ID and cohort as random effects.
-To obtain an estimate of explanatory power, we calculated marginal R^2^ (just incorporating fixed effects) following Nakagawa & Schielzeth [-@Nakagawa2013].
-When examining the distribution longitudinal telomere changes we observed increases in telomere length with age, so we used repeat measurements to test whether these increases could be explained by measurement error.
-We calculated $\Delta$RTL between pairs of repeat measurements within samples in exactly the same way as for across samples (i.e.
-completely separate reactions run on separate plates: N = `r nrow(subset(ddL,Group == 'Within samples'))` measurements from `r length(unique(subset(ddL,Group == 'Within samples')$BirdID))` samples), and compared the variance in $\Delta$RTL within and among individuals using a Levene's test.
-Then, to separately test whether the extent of telomere increases and decreases within individuals was greater than expected by measurement error, we split $\Delta$RTL measurements into groups in which RTL decreased ($\Delta$RTL < 0) and increased $\Delta$RTL ($\Delta$RTL > 0), and tested whether $\Delta$RTL among samples was significantly different between within-sample and across-sample measurements, using Wilcoxon tests.
+Using the longitudinal data, we then tested how telomeres change with age in individuals, using LMMs of RTL as a response and $\Delta$age [a longitudinal measure based on within-subject centring; @VandePol2009] as an explanatory variable.
+We calculated $\Delta$age using log and polynomial transformed age data, and carried out model selection as above, with the exception that we did not model $\Delta$age as a factor (due to a lack of discreet groupings), and mean age was also included in models to partition within-individual vs cross-sectional effects [@VandePol2009].
 
-We used LMMs to test how variation in early-life environmental and social conditions influenced telomere length within cohorts.
-We created a full model with RTL as a response variable, alongside the following explanatory variables: age, tarsus length, body mass, sex, insect abundance, territory quality, island-wide population density (an annual measure estimated from the summer breeding census), territory group size, and the number of helping subordinate birds present in the territory.
-The random effects structure was informed by the analysis of telomere dynamics and age (see results): we included individual ID, cohort ID, and a random slope of age among cohorts.
+We also used longitudinal data to determine individual-level consistency in RTL.
+We constructed a LMM with RTL at time *t* + 1 as the response variable, RTL at time *t* and age at time *t* as fixed effects, and individual ID and cohort as random effects.
+To obtain an estimate of explanatory power of the fixed effects, we calculated marginal R^2^ following Nakagawa & Schielzeth [-@Nakagawa2013].
+
+When examining the distribution of longitudinal telomere changes we observed increases in telomere length with age, so we used repeat measurements to test whether these increases could be explained by measurement error.
+We calculated the change in RTL between pairs of repeat measurements within samples (hereafter $\Delta$RTL[sample]; N = `r nrow(subset(ddL,Group == 'Within samples'))` measurements from `r length(unique(subset(ddL,Group == 'Within samples')$BirdID))` samples) in exactly the same way as for across samples (hereafter $\Delta$RTL[individual]), using completely separate reactions run on separate plates.
+To test whether greater changes in RTL were observed among individuals compared to among repeat samples, we compared the variance in $\Delta$RTL[sample] and $\Delta$RTL[individual] using a Levene's test.
+Then, to separately test whether the extent of telomere increases and decreases within individuals was greater than expected by measurement error, we split $\Delta$RTL measurements into groups in which RTL decreased ($\Delta$RTL < 0) and increased $\Delta$RTL ($\Delta$RTL > 0), and tested whether $\Delta$RTL[inidividual] values were significantly different from $\Delta$RTL[sample] values, using Wilcoxon tests.
+
+We then used LMMs to test how variation in environmental and social conditions influenced telomere length within cohorts.
+We created a full model with RTL as a response variable, alongside the following explanatory variables: log age (based on the RTL and age analysis; see results), tarsus length, body mass, sex, insect abundance, territory quality, island-wide population density (an annual measure estimated from the summer breeding census), territory group size, and the number of helping subordinate birds present in the territory.
+The random effects structure was informed by the analysis of telomere dynamics and age (see results): we included individual ID, cohort ID, and a random slope of log age among cohorts.
 We report model estimates and confidence intervals for all effects included in the full model.
-We also calculated marginal R^2^ and conditional R^2^ [incorporating fixed and random effects; @Johnson2014] to assess the explanatory power of these models.
+We also calculated marginal R^2^ and conditional R^2^ [incorporating fixed and random effects, respectively; @Johnson2014] to assess the explanatory power of these models.
 As a complementary approach, we also performed model averaging, using the MuMIn package in R [@Barton2012].
@@
 We report model-averaged coefficients, confidence intervals and 'relative importance', which reflects the relative weights of each predictor variable across the top model set.
-For individuals with longitudinal data we repeated the above analyses of telomere dynamics, replacing telomere length with $\Delta$RTL as the response variable, and excluding the random effects for plate ID (as each measurement of $\Delta$RTL was based on two or more measurements, and so run on multiple plates) and cohort ID (as longitudinal telomere dynamics did not differ among cohorts; see results).
+
+For individuals with longitudinal data we repeated the above analyses of telomere dynamics, replacing telomere length with $\Delta$RTL[sample] as the response variable, and including the environmental/social explanatory variables from the first of the two sampling points.
+We excluded the cohort ID random effect from this analysis, as longitudinal telomere dynamics did not differ among cohorts; see results.
 
@@
 Efficiencies (mean $\pm$ s.d.) for our telomere and GAPDH reactions were (`r round2(mean(dd$EffTelomere))` $\pm$ `r round2(sd(dd$EffTelomere))`) and (`r round2(mean(dd$EffGAPDH))` $\pm$ `r round2(sd(dd$EffGAPDH))`) respectively.
-Inter-plate repeatability of RTL, based on `r nrow(subset(ddL,Group == 'Within samples'))` samples measured at least twice, was `r round2(rptall$R)` (CI = `r round2(rptall$CI_emp[1])`-`r round2(rptall$CI_emp[2])`).
+Inter-plate repeatability of RTL, based on `r nrow(subset(ddL,Group == 'Within samples'))` samples measured at least twice, was `r round2(rptall$R)` (CI = `r round2(rptall$CI_emp[1])`, `r round2(rptall$CI_emp[2])`).
 
@@
 We first tested how RTL was related to age among cohorts using a model selection approach.
-We found that the top model, by some distance, contained a loglinear relationship between RTL and age, as well as a log age x cohort interaction (Table 1A).
-The second best model contained log age and cohort ID, with no interaction term, although this was substantially poorer in terms of model fit ($\Delta$AICc > 10), and all other models were much poorer still ($\Delta$AICc > 35; Table 1).
-The log-linear relationship between RTL and age could be seen clearly in the raw data; RTL decreased with age (estimate = `r round3(fixef(RTL_LogAge)['LogAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_LogAge)['LogAge',],lessthan = F),collapse = ', ')`), with the greatest decrease occurring early in life (Fig.
+The top model contained a loglinear relationship between RTL and age, as well as a log age x cohort interaction (Table 1A).
+The second-best model contained log age and cohort ID, with no interaction term, although this model was substantially poorer in terms of model fit ($\Delta$AICc > 10).
+All other models fitted the data much less well ($\Delta$AICc > 35; Table 1).
+The log-linear relationship between RTL and age could be seen clearly in the raw data; RTL decreased with age (estimate = `r round3(fixef(RTL_LogAge)['LogAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_LogAge)['LogAge',],lessthan = F),collapse = ', ')`), with the greatest decrease occurring in the first year of life (Fig.
 1A).
-There was also substantial variation in RTL among cohorts, with no obvious trend over time (Fig.
+There was substantial variation in RTL among cohorts, with no obvious trend over time (Fig.
 1B).
-There was a negative relationship between RTL and age in almost all `r length(table(dd$Cohort))` cohorts, but the slope of the relationship between log age and RTL varied substantially among cohorts (Fig.1C).
+There was a negative relationship between RTL and age in 21 of the `r length(table(dd$Cohort))` cohorts, but the slope of the relationship between log age and RTL varied substantially among cohorts (Fig.1C).
+In the one year in which RTL increased with age (2013), 17 of the 18 birds were fledglings or subadults, suggesting that the observed pattern was an artifact of the sampling in this season rather than a real relationship.
 
@@
 
-There was positive correlation between telomere measurements taken from different time points (Fig.
-2A), but this was very weak (marginal R^2^ = `r round2(r.squaredGLMM(RTL_RTL1)['R2m'])`), and not quite significant (estimate = `r round3(fixef(RTL_RTL1)['RTL'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_RTL1)['RTL',],lessthan = F),collapse = ', ')`).
-Although both cross-sectional and longitudinal data indicated a general trend of telomere shortening with age, we found that RTL within individuals increased in `r round((nrow(subset(dd3,DeltaRTL>0))/nrow(dd3))*100,0)`% of our `r nrow(dd3)` $\Delta$RTL measurements (Fig.
+There was positive correlation between telomere measurements taken at different time points (Fig.
+2A), but this was very weak (marginal R^2^ = `r round2(r.squaredGLMM(RTL_RTL1)['R2m'])`), and not significant (estimate = `r round3(fixef(RTL_RTL1)['RTL'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_RTL1)['RTL',],lessthan = F),collapse = ', ')`).
+Although both cross-sectional and longitudinal data indicated a general trend of telomere shortening with age, we found that RTL within individuals increased in `r round((nrow(subset(dd3,DeltaRTL>0))/nrow(dd3))*100,0)`% of our `r nrow(dd3)` $\Delta$RTL[individual] measurements (Fig.
 2A).
@@
 3C).
-The full model was fairly weak in terms of explanatory power of fixed effects (marginal R^2^ = `r round2(r.squaredGLMM(RTL_Full)['R2m'])`), although including the random effect terms increased this substantially (conditional R^2^ = `r round2(r.squaredGLMM(RTL_Full)['R2c'])`) The model averaging approach yielded qualitatively identical results to the full LMM, with the same explanatory variables 'significant' in terms of being retained in top models, and having model-averaged confidence intervals not overlapping zero (Table S1; Fig.
+The full model was weak in terms of explanatory power of fixed effects (marginal R^2^ = `r round2(r.squaredGLMM(RTL_Full)['R2m'])`), although including the random effect terms increased this substantially (conditional R^2^ = `r round2(r.squaredGLMM(RTL_Full)['R2c'])`).
+The model averaging approach yielded qualitatively identical results to the full LMM, with the same explanatory variables 'significant' in terms of being retained in top models, and having model-averaged confidence intervals not overlapping zero (Table S1; Fig.
 S1).
 One interesting finding from the model selection was that sex only appeared in top models where tarsus length was also present (Table S1).
-In accordance with this, when tarsus length was removed from the full model sex was no longer significant (estimate = `r round3(fixef(RTL_Full_NoTarsus)['SexMales'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Full_NoTarsus)['SexMales',],lessthan = F),collapse = ', ')`), and a sex x tarsus interaction was significant when included (estimate = `r round3(fixef(RTL_Full_SexTarsus)['SexMales:Tarsus'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Full_SexTarsus)['SexMales:Tarsus',],lessthan = F),collapse = ', ')`); RTL increased with tarsus length in males, but decreased in females (Fig.
-4A).
+In accordance with this, when tarsus length was removed from the full model sex was no longer significant (estimate = `r round3(fixef(RTL_Full_NoTarsus)['SexMales'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Full_NoTarsus)['SexMales',],lessthan = F),collapse = ', ')`), and a sex x tarsus interaction was significant when included (estimate = `r round3(fixef(RTL_Full_SexTarsus)['SexMales:Tarsus'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Full_SexTarsus)['SexMales:Tarsus',],lessthan = F),collapse = ', ')`); RTL decreased with tarsus length in both sexes, but this decrease was stronger in females (Fig.
+3B).
 No social or ecological environmental variables were significant predictors of $\Delta$RTL using the full model approach (Table S2).
@@
 ###Discussion
-Here we use a long-term, multi-cohort dataset to assess the relationships between spatio-temporal variation in the ecological environment and lifelong telomere dynamics in a closed population of Seychelles warblers.
+Here we use a long-term, multi-cohort dataset to assess the relationships between spatio-temporal variation in the ecological environment and lifelong telomere dynamics in a contained population of Seychelles warblers.
 We found that telomere length decreases with age, and that this decrease is greatest in early life.
-Telomere length decreased with age in almost all `r length(table(dd$Cohort))` cohorts studied, but telomere length, and the rate of decrease of telomere length with age, varied substantially among cohorts.
+Telomere length decreased with age in almost all of the `r length(table(dd$Cohort))` cohorts studied, but telomere length, and the rate of decrease of telomere length with age, varied substantially among cohorts.
 Despite an overall trend for telomere shortening in the Seychelles warbler, we found that telomere length increased within some individuals at particular life stages, and that the extent of these increases cannot be explained solely by qPCR measurement error.
@@
 
-Our study adds the the now substantial body of literature from humans and wild animals showing that telomere length decreases with age, and that this decrease is most rapid in early life [e.g.
+Our study adds to the substantial body of literature from humans and wild animals showing that telomere length decreases with age, and that this decrease is most rapid in early life [e.g.
 @Frenck1998; @Haussmann2003; @Heidinger2012].
 However, we also found that despite an overall trend for shortening, telomere length both increased and decreased, especially after the juvenile period.
-Importantly, these increases were observed in longitudinal as well as cross-sectional data, suggesting that selective disappearance does not explain this pattern entirely.
+Importantly, these increases were observed in longitudinal as well as cross-sectional data, suggesting that selective disappearance of individuals with shorter telomeres does not explain this pattern entirely.
 Longitudinal increases in measured telomere length have been observed in humans and [reviewed in @Steenstrup2013] wild animals [@Kotrschal2007; @Fairlie2016].
@@
 
-We found that increases in telomere length are not consistent over individual lifespans, but that increases occur at specific, short periods, against a backdrop of overall telomere shortening.
+Increases in telomere length were not consistent over individual lifespans, but increases occured at specific, short periods, against a backdrop of overall telomere shortening.
 Consistent with a pattern of sporadic changes in telomere length with age, we found that within-individual telomere measurements were only weakly correlated.
@@
 However, the lifelong telomere dynamics found in Seychelles warblers are strikingly similar to those found in Soay sheep [@Fairlie2016], and we anticipate that a similar pattern may be found in other wild populations.
-That increases in telomere length may be sporadic and overlaid onto an overall pattern of shortening is a crucial point, because previously described approaches to distinguish telomere elongation from measurement error, based on assumptions about follow-up time between measurements [@Steenstrup2013], or based on measuring variance among measurements [@Simons2014], assume the telomere elongation within individuals is consistent over time.
+
+The finding that increases in telomere length may be sporadic and overlaid onto an overall pattern of shortening is an important point when assessing the occurrence of telomere lengthening.
+Previously described approaches to distinguish telomere elongation from measurement error, based on assumptions about follow-up time between measurements [@Steenstrup2013], or based on measuring variance among measurements [@Simons2014], assume that telomere elongation within individuals is consistent over time.
 Such sporadic changes in telomere length over lifespans could occur due to changes in the composition of cell types within individual samples, or due to the actual elongation of telomeres.
@@
 Our data suggest that birth year is a highly important factor in shaping telomere dynamics, but also that age-related declines in telomere length vary among cohorts.
-This is consistent with telomere length being controlled by both genetic and environmental factors - something that is becoming apparent from quantitative genetic studies of telomeres.
+This is consistent with telomere length being controlled by both genetic and environmental factors - something that is becoming apparent from quantitative genetic studies of telomere dynamics.
 Moreover, our findings suggest that the telomere dynamics of a population at a given point in time represent a snapshot of a temporally varying process.
@@
 
-We found that temporal variation in food availability was positively related to telomere length.
-This is consistent with strong cohort effects, and suggests that temporal variation in conditions may be a key driver of costs in the Seychelles warbler.
-Although Cousin Island is relatively benign in comparison to other island systems [e.g.
+We found that temporal variation in population-level food availability was positively related to telomere length.
+This is consistent with the strong cohort effects we found, and suggests that temporal variation in conditions may be a key driver of costs in the Seychelles warbler.
+Although the environmental conditions on Cousin Island are relatively benign in comparison to other island systems [e.g.
 @Coulson2001], substantial annual variation in rainfall does occur, with associated changes in insect abundance, and it appears that this confers a cost to Seychelles warblers.
@@
 3B).
-One possibility is that the environment imposes differential costs on males and females: a recent study in captive zebra finches found that manipulation of dietary nutrients had sex-dependent affects on telomere dynamics [@Noguera2015].
+One possibility is that the environment imposes differential costs on males and females: a recent study in captive zebra finches found that manipulation of dietary nutrients had sex-dependent effects on telomere dynamics [@Noguera2015].
 Also worth noting is that the effect of telomere length on survival in strongest in male Seychelles warblers [@Barrett2013a], although the nature of the relationship between sex, telomeres and survival is not yet clear [@Barrett2011]
 
-Although we found clear associations between the environment and telomere dynamics, we should bear in mind that our social and ecological variables explained a modest proportion of the variance in RTL.
+Although we found clear associations between the environment and telomere dynamics, we should bear in mind that our social and ecological variables explained a only a small proportion of the variance in RTL.
 A poor social and ecological environment is known to be detrimental to Seychelles warblers, both in terms of oxidative stress and survival [@VandeCrommenacker2011a; @Brouwer2012], and it is therefore perhaps surprising that these variables do not explain more variance in RTL.
-While measurement error clearly accounts for some of this noise, it is also likely that early-life RTL in the Seychelles warbler is explained by a set of environmental and genetic variables not considered here [e.g.
+While measurement error clearly accounts for some of this lack of explanatory power, it is also likely that early-life RTL in the Seychelles warbler is explained by a set of environmental and genetic variables not considered here [e.g.
 @Bebbington2016].
@@
 @Becker2015; @Asghar2014; @Heidinger2016].
-Future research on the Seychelles warbler will examine, in a quantitative genetic framework, how genetic and environmental components, and their interactions, affect telomere dynamics and senescence.
-
-Finally, we emphasise once more that many of the findings here were dependent on long-term ecological data.
-Long-term ecological study systems are uniquely suited to addressing a wide range of problems in ecology [@Clutton-Brock2010], and understanding variation in telomere dynamics is one such problem.
+Long-term ecological study systems are uniquely suited to addressing such questions in natural systems [@Clutton-Brock2010].
 To gain a full understanding of telomere dynamics in natural systems, long-term studies combining ecological and genetic data will be required from a range of species.
@@
 Emma Barrett laid the foundations for this study, generating the original telomere qPCR protocol.
-We thank everyone who has helped in the field, and the Seychelles warbler research group for discussions.
+We thank everyone who has helped in the field, with lab work and with database management, and the Seychelles warbler research group for discussions.
 This work was funded by two Natural Environment Research Council (NERC) grants to DSR (NE/F02083X/1 and NE/K005502/1).
-LGS was also funded by a fellowship from the BBSRC, and HLD by a NERC fellowship.
+LGS was also funded by a fellowship from the BBSRC, and HLD by a NERC fellowship (NE/I021748/1).
+MH was funded by a VENI fellowship from the Netherlands Organisation for Scientific Research (863.15.020).
 

### 2017-01-16 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Home edits

@@
 nams[nams == 'Agemonths'] <- 'Age (linear)'
+nams[nams == 'MeanAge'] <- 'Age (mean)'
 nams[nams == 'LogAge'] <- 'Age (log)'

### 2017-01-16 Scripts/SW juvenile telomeres IPC clean.R

@@ -95,8 +95,8 @@ for(i in 1:nrow(dd))
                 dd$Tarsus[i] <- NA,
                 dd$Tarsus[i] <- dd$LeftTarsus[i]),
          dd$Tarsus[i] <- dd$RightTarsus[i])
-  
-  dd$DeltaAge[i] <- dd$Agemonths[i] - mean(subset(dd,BirdID == dd$BirdID[i])$Agemonths)
+  dd$MeanAge[i] <- mean(subset(dd,BirdID == dd$BirdID[i])$Agemonths)
+  dd$DeltaAge[i] <- dd$Agemonths[i] - dd$MeanAge[i]
   dd$DeltaLogAge[i] <- dd$LogAge[i] - mean(subset(dd,BirdID == dd$BirdID[i])$LogAge)
   dd$DeltaAge2[i] <- dd$Age2[i] - mean(subset(dd,BirdID == dd$BirdID[i])$Age2)
 }

### 2017-01-16 Scripts/SW juvenile telomeres IPC models.r

@@ -12,7 +12,7 @@ RTL_Age <- lmer(RTL ~ Cohort*(Agemonths + LogAge + FAge + Age2) + (1|BirdID) + (
 RTL_LogAge <- lmer(RTL ~ LogAge + (1|BirdID) + (1|CatchYear) + (1|PlateID),
                 data=dd,
                 REML = FALSE)
-RTL_DeltaAge <- lmer(RTL ~ Cohort*(DeltaAge + DeltaLogAge + DeltaAge2) + (1|BirdID) + (1|CatchYear) + (1|PlateID),
+RTL_DeltaAge <- lmer(RTL ~ MeanAge + Cohort*(DeltaAge + DeltaLogAge + DeltaAge2) + (1|BirdID) + (1|CatchYear) + (1|PlateID),
                 data=dd3,
                 REML = FALSE)
 RTL_DeltaLogAge <- lmer(RTL ~ DeltaLogAge + (1|BirdID) + (1|CatchYear) + (1|PlateID),

### 2017-01-24 Manuscript/SW_early_life_telomeres.Rmd
Home edits. Sent full draft to David.

@@
 1.
-Understanding the costs that individuals face throughout their lifetimes is fundamental to understanding life-history evolution and population dynamics.
+Understanding individual-level variation in response to the social and ecological environment is fundamental to understanding life-history evolution and population dynamics.
 Telomeres, the protective caps at the ends of chromosomes, shorten in response to oxidative stress, and telomere shortening is correlated with reduced survival.
@@
 Here we study spatio-temporal variation in lifelong telomere dynamics in the Seychelles warbler (*Acrocephalus sechellensis*).
-We combine long-term life history and ecological data with one the largest longitudinal telomere datasets to date, consisting of `r nrow(dd)` samples from `r length(table(dd$Cohort))` cohorts born between `r min(dd$LayYear)` and `r max(dd$LayYear)`.
+We combine long-term life history and ecological data a large longitudinal telomere datasets, consisting of `r nrow(dd)` samples from `r length(table(dd$Cohort))` cohorts born between `r min(dd$LayYear)` and `r max(dd$LayYear)`.
 We provide a detailed analysis of how telomere dynamics vary over individual lifespans and cohorts, and with spatio-temporal variation in the social and ecological environment.
@@
 @Schultner2014; @Asghar2015].
-There is also some evidence that telomeres can increase, as well as decrease, in length [@Simons2014; @Bateson2016], which has important ramifications for our understanding of how telomeres reflect costs.
-However, increases in telomere length are often attributed to measurement error [@Steenstrup2013], and as such its ecological significance is unknown.
+There is also some evidence that telomeres can increase, as well as decrease, in length [@Simons2014; @Bateson2016a], which has important ramifications for our understanding of how telomeres reflect costs.
+However, increases in telomere length are often attributed to measurement error [@Steenstrup2013; but see @Bateson2016a], and as such its ecological significance is unknown.
 
@@
 When examining the distribution of longitudinal telomere changes we observed increases in telomere length with age, so we used repeat measurements to test whether these increases could be explained by measurement error.
-We calculated the change in RTL between pairs of repeat measurements within samples (hereafter $\Delta$RTL[sample]; N = `r nrow(subset(ddL,Group == 'Within samples'))` measurements from `r length(unique(subset(ddL,Group == 'Within samples')$BirdID))` samples) in exactly the same way as for across samples (hereafter $\Delta$RTL[individual]), using completely separate reactions run on separate plates.
-To test whether greater changes in RTL were observed among individuals compared to among repeat samples, we compared the variance in $\Delta$RTL[sample] and $\Delta$RTL[individual] using a Levene's test.
-Then, to separately test whether the extent of telomere increases and decreases within individuals was greater than expected by measurement error, we split $\Delta$RTL measurements into groups in which RTL decreased ($\Delta$RTL < 0) and increased $\Delta$RTL ($\Delta$RTL > 0), and tested whether $\Delta$RTL[inidividual] values were significantly different from $\Delta$RTL[sample] values, using Wilcoxon tests.
+We calculated the change in RTL between pairs of repeat measurements within samples (hereafter $\Delta$RTL~*sample*~; N = `r nrow(subset(ddL,Group == 'Within samples'))` measurements from `r length(unique(subset(ddL,Group == 'Within samples')$BirdID))` samples) in exactly the same way as for across samples (hereafter $\Delta$RTL~*individual*~), using completely separate reactions run on separate plates.
+To test whether greater changes in RTL were observed among individuals compared to among repeat samples, we compared the variance in $\Delta$RTL~*sample*~ and $\Delta$RTL~*individual*~ using a Levene's test.
+Then, to separately test whether the extent of telomere increases and decreases within individuals was greater than expected by measurement error, we split $\Delta$RTL measurements into groups in which RTL decreased ($\Delta$RTL < 0) and increased $\Delta$RTL ($\Delta$RTL > 0), and tested whether $\Delta$RTL~*inidividual*~ values were significantly different from $\Delta$RTL~*sample*~ values, using Wilcoxon tests.
 
@@
 
-For individuals with longitudinal data we repeated the above analyses of telomere dynamics, replacing telomere length with $\Delta$RTL[sample] as the response variable, and including the environmental/social explanatory variables from the first of the two sampling points.
+For individuals with longitudinal data we repeated the above analyses of telomere dynamics, replacing telomere length with $\Delta$RTL~*individual*~ as the response variable, and including the environmental/social explanatory variables from the first of the two sampling points.
 We excluded the cohort ID random effect from this analysis, as longitudinal telomere dynamics did not differ among cohorts; see results.
@@
 2A), but this was very weak (marginal R^2^ = `r round2(r.squaredGLMM(RTL_RTL1)['R2m'])`), and not significant (estimate = `r round3(fixef(RTL_RTL1)['RTL'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_RTL1)['RTL',],lessthan = F),collapse = ', ')`).
-Although both cross-sectional and longitudinal data indicated a general trend of telomere shortening with age, we found that RTL within individuals increased in `r round((nrow(subset(dd3,DeltaRTL>0))/nrow(dd3))*100,0)`% of our `r nrow(dd3)` $\Delta$RTL[individual] measurements (Fig.
+Although both cross-sectional and longitudinal data indicated a general trend of telomere shortening with age, we found that RTL within individuals increased in `r round((nrow(subset(dd3,DeltaRTL>0))/nrow(dd3))*100,0)`% of our `r nrow(dd3)` $\Delta$RTL~*individual*~ measurements (Fig.
 2A).
@@
 The most commonly invoked explanation for increases in telomere length is measurement error, which can be a particular problem in qPCR-based telomere studies [@Nussey2014; @Steenstrup2013; @Verhulst2015].
-To our knowledge, ours is the first study to explicitly compare intra-individual variation among samples to variation among sample replicates, on a large scale.
+However, recent modelling work suggests that longitudinal telomere dynamics in humans are indeed consistent with apparent lenghtening, and that dismissing apparent telomere lengthening as solely measurment error is "too strong" without additional data [@Bateson2016a].
+Here, we explicitly compare intra-individual variation among samples to variation among sample replicates, on a large scale.
 Our results suggest that qPCR measurement error alone cannot explain observed increases in RTL observed within individuals.

### 2017-01-24 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Home edits. Sent full draft to David.

--- 2016-11-25 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
@@
                            !(Agemonths & FAge) &
-                           !(Agemonths & Age2) &
-                           !(FAge & Age2))
+                           !(FAge & Age2) &
+                           !(Age2 & !(Agemonths)))
 
@@
 nams[nams == 'Cohort:FAge'] <- 'Age (factor)*cohort'
-nams[nams == 'Agemonths:Cohort'] <- 'Age (continuous)*cohort'
+nams[nams == 'Agemonths:Cohort'] <- 'Age (linear)*cohort'
 nams[nams == 'Cohort:LogAge'] <- 'Age (log)*cohort'
@@
 RTL_DeltaAge_Dredge <- dredge(RTL_DeltaAge,
-                         subset = !(DeltaLogAge & DeltaAge) &
+                         subset = MeanAge &
+                           !(DeltaLogAge & DeltaAge) &
                            !(DeltaLogAge & DeltaAge2) &
-                           !(DeltaAge & DeltaAge2))
+                           !(!(DeltaAge) & DeltaAge2))
 
@@
 The thick line and shaded area represent the fitted values and 95% confidence limits of a linear regression of RTL and log-transformed age.**B** Boxplot of variation in RTL among cohorts.
-For visualisation purposes a selection of cohorts with large sample sizes across a wide age range was chosen, but all cohorts were included in models.
 **C** RTL and age among cohorts.
@@
 The dotted line represents parity, and thus points above and below the line represent increases and decreases in RTL, respectively.
-**B** Scaled density plots of repeated RTL measurements among individual samples, and among different samples taken from the same individual.
+**B** Scaled density plots of repeated RTL measurements among individual samples (grey), and among different samples taken from the same individual (yellow).
+Areas of the density plot to the left of the dotted line represent decreases in RTL, while areas to the right represent increases.
 

### 2017-02-01 Manuscript/SW_early_life_telomeres.Rmd
Home edits

@@
 
-**Spatiotemporal variation in lifelong telomere dynamics in a long-term ecological study**
+**Spatio-temporal variation in lifelong telomere dynamics in a long-term ecological study**
 
@@
 
-**Running head:** Early-life telomeres
+**Running head:** Lifelong telomere dynamics in warblers
 
@@
 1.
-Understanding individual-level variation in response to the social and ecological environment is fundamental to understanding life-history evolution and population dynamics.
-Telomeres, the protective caps at the ends of chromosomes, shorten in response to oxidative stress, and telomere shortening is correlated with reduced survival.
-Investigating telomere dynamics may help us quantify individual variation in the costs of the social and ecological environment, and enhance our understanding of the dynamics of natural populations.
+Understanding individual-level variation in response to the environment is fundamental to understanding life-history evolution and population dynamics.
+Telomeres, the protective caps at the ends of chromosomes, shorten in response to oxidative stress, and telomere shortening is correlated with reduced survival and lifespan.
+Investigating telomere dynamics may help us quantify individual variation in the costs experienced as a result of different social and ecological environmental factors, and enhance our understanding of the dynamics of natural populations.
 
 2.
-Here we study spatio-temporal variation in lifelong telomere dynamics in the Seychelles warbler (*Acrocephalus sechellensis*).
-We combine long-term life history and ecological data a large longitudinal telomere datasets, consisting of `r nrow(dd)` samples from `r length(table(dd$Cohort))` cohorts born between `r min(dd$LayYear)` and `r max(dd$LayYear)`.
+Here we study spatio-temporal variation in lifelong telomere dynamics in the Seychelles warbler, *Acrocephalus sechellensis*.
+We combine long-term life history and ecological data with a large longitudinal telomere datasets, consisting of `r nrow(dd)` samples from `r length(table(dd$Cohort))` cohorts born between `r min(dd$LayYear)` and `r max(dd$LayYear)`.
 We provide a detailed analysis of how telomere dynamics vary over individual lifespans and cohorts, and with spatio-temporal variation in the social and ecological environment.
@@
 3.
-We found that telomere length decreases with cross-sectional and longitudinal measures of age, and that telomere length decreases most rapidly early in life.
-However, both cross-sectional and longitudinal data suggested that against this overall pattern of shortening, bouts of telomere length increases occur in some individuals.
+We found that telomere length decreases with cross-sectional and longitudinal measures of age, and most rapidly very early in life.
+However, both cross-sectional and longitudinal data suggested that against this overall pattern of shortening, bouts of telomere length increase occur in some individuals.
 Using a large number of repeated measurements we show statistically that these increases are unlikely to be explained solely by qPCR measurement error.
@@
 4.
-Telomere length decreased with age in almost all cohorts studied, but telomere length and the slope of the relation ship with age varied markedly among cohorts.
-Telomere length was positively associated with island-wide temporal insect abundance, suggesting that the costs associated with living in a harsher environment can be studied by investigating telomere dynamics.
-We also found evidence for sex-specific relationships between telomeres and tarsus length, potentially reflecting differential costs of growth.
+Telomere length and the slope of the relationship with age varied markedly among cohorts.
+Telomere length was positively associated with island-wide temporal insect abundance - a key resource for the insectivorous Seychelles warbler - suggesting that the costs associated with living in harsher environments can be studied by investigating telomere dynamics.
+We also found evidence for sex-specific relationships between telomeres and tarsus length, potentially reflecting differential costs of growth, which leads to adult females having significantly shorter telomeres than males.
 
@@
 ###Introduction
-A major aim of ecologists is to quantify and understand why individuals vary in their response to different environments.
-Identifying the costs imposed on populations is central to understanding variation in fitness [@Lindstrom1999], and thus for understanding population and community dynamics [@Bolnick2011].
+A major aim of ecologists and evolutionary biologists is to quantify and understand why individuals vary in their response to different environmental factors.
+Identifying this variation in environmental costs is central to understanding variation in fitness [@Lindstrom1999], and thus for understanding population and community dynamics [@Bolnick2011].
+Furthermore, knowledge of the relative impact that different environmental factors exert on individuals, and why individuals may differ in mitigating these costs, is important to understanding evolutionary trade-offs and life-history strategies [@Stearns1992].
 However, fully quantifying individual-level variation in costs is impossible in wild systems, and thus effective biomarkers that reflect the physiological consequences of individual-level experiences are required.
@@
 @Frenck1998; @Haussmann2003].
-Telomere shortening occurs in response to oxidative stress, which can be elevated due to environmental factors [@VonZglinicki2002].
-There is evidence from both humans and wild animal populations that telomere shortening is influenced by the conditions experienced during both early life and adulthood [@Price2013; @Monaghan2014; @Nettle2015; @Reichert2015].
+Telomere shortening is accelerated by oxidative stress, which can be elevated due to many environmental factors [@VonZglinicki2002].
+There is evidence from humans, and from captive and wild animal populations, that telomere shortening is influenced by the conditions experienced during both early life and adulthood  [@Price2013; @Monaghan2014; @Nettle2015; @Reichert2015].
 Importantly, the extent of telomere shortening is directly linked to senescence and survival.
-When telomeres become critically short cells senesce [@Campisi2003], and the accumulation of these cells can result in organismal senescence and death [@Wong2003].
+When telomeres become critically short cells senesce [@Campisi2003], and the accumulation of these cells has been suggested to result in organismal senescence and death [@Wong2003].
 
-The association between senescence and telomere length has inspired a great deal of recent research into telomere evolutionary ecology, and relationships between telomere dynamics and survival have been documented in several wild organisms [reviewed in @Horn2010; @Haussmann2010; @Simons2015].
-While there is as yet little direct evidence that the relationship between telomere dynamics and survival is causal [@Simons2015], there is mounting evidence that telomeres can act as biomarkers of cost in wild populations, providing a signature of the ecological stress that has been experienced and is otherwise difficult to detect [e.g.
-@Schultner2014; @Asghar2015].
-There is also some evidence that telomeres can increase, as well as decrease, in length [@Simons2014; @Bateson2016a], which has important ramifications for our understanding of how telomeres reflect costs.
-However, increases in telomere length are often attributed to measurement error [@Steenstrup2013; but see @Bateson2016a], and as such its ecological significance is unknown.
+The association between senescence and telomere length has inspired a great deal of recent research into telomere evolutionary ecology and relationships between telomere dynamics and survival or lifespan have been documented in several wild organisms [@Horn2010; @Haussmann2010; @Barrett2013a; @Simons2015].
+As yet there is little direct evidence that the relationship between telomere dynamics and survival is causal [@Simons2015].
+However, there is mounting evidence that telomeres can act as biomarkers of individual condition and ageing in wild populations, providing a measure of the ecological stress that an individual has been experienced; a signature that can otherwise be difficult to detect [e.g.
+@Schultner2014; @Asghar2015; @Bebbington2016].
+There is also some evidence that telomere length, measured longitudinally from individuals, can increase as well as decrease [@Simons2014; @Bateson2016a], which has important ramifications for our understanding of how telomeres reflect costs.
+However, such increases in telomere length are often attributed to measurement error [@Steenstrup2013; but see @Bateson2016a], and as such its ecological significance is unknown.
 
@@
 Understanding how different factors shape telomere length variation is important, as before we can use telomeres as a measure of the costs experienced by individuals, we need to know how different developmental, genetic and ecological variables interact to affect telomeres.
-Telomere length and rates of shortening can vary according to parental characteristics [Njajour2007; @Heidinger2016], among sexes [@Barrett2011], and with a whole host of environmental conditions experienced at different life-history stages [@Monaghan2014].
+Telomere length and rates of shortening can vary according to parental characteristics [Njajour2007; @Heidinger2016], among sexes [@Barrett2011; @Watson2017], and with a whole host of environmental conditions experienced at different life-history stages [@Monaghan2014].
 Recent evidence suggests that telomere dynamics are indeed highly variable over individual lifespans, and that even the relationship between telomeres and age can vary markedly among cohorts [@Fairlie2016].
@@
 The longitudinal study (since 1986) of the Seychelles warbler (*Acrocephalus sechellensis*) population on Cousin Island provides an excellent system for studying telomere dynamics and senescence patterns in the wild [reviewed in @Hammers2015].
-Due to the isolated nature of the study population and intensive field monitoring, we have comprehensive ecological and survival data spanning many years (see Methods, below).
-Environmental conditions and population density on Cousin Island vary across space and time due to weather-induced changes in foliage cover and food availability [@VandeCrommenacker2011a].
+Due to the isolated nature of the study population [@Komdeur2004a] and intensive field monitoring, we have comprehensive ecological and survival data spanning many years (see Methods, below).
+Environmental conditions and population density on Cousin Island vary across space and time due to weather-induced changes in foliage cover and insect prey availability [@VandeCrommenacker2011a].
 Variation in oxidative stress experienced by individuals is associated with territory quality [@VandeCrommenacker2011a].
-However, neither early-life nor adult survival appear to be associated with territory quality or local density [@Brouwer2006; @Hammers2013].
+However, the evidence that individual survival and lifespan is associated with spatial variation in early life territory quality or local density is equivocal and confounded by variation in subsequent life-history parameters [@Brouwer2006; @Hammers2013].
+There is also variation in the social environment that individual Seychelles warblers experience.
+Facultative cooperative breeding occurs in this species [@Komdeur1994a; @Richardson2003a; @Richardson2007], and the presence of helpers (but not other resident non-helpers) in the natal territory is associated with increased survival of offspring later in life [@Brouwer2012].
 
- As well as territory quality variation, there is also variation in the social environment.
-Facultative cooperative breeding occurs in the Seychelles warbler [@Komdeur1994a; @Richardson2003a], and the presence of helpers (but not non-helpers) in the natal territory is associated with increased survival of offspring later in life [@Brouwer2012].
-Lastly, we have an established protocol for assessing telomere length in this species [@Barrett2012; @Bebbington2016], and telomere length predicts survival independently of age, suggesting that individual variation in telomere length is ecologically relevant in this species [@Barrett2013a; @Hammers2015; @Bebbington2016].
-Thus, we have an excellent system in which to assess the costs of different social and environmental conditions experienced by individuals, and to assess how these costs vary over space and time.
+Importantly, we have an established protocol for assessing telomere length in the Seychelles warbler [@Barrett2012; @Bebbington2016].
+Furthermore, telomere dynamics predict survival independently of age [@Barrett2013a] and are negatively associated with inbreeding [@Bebbington2016], suggesting that individual variation in telomere length is ecologically relevant in this species.
+Thus, we have an excellent system in which to determine the impact of different social and environmental conditions experienced by individuals, and to assess how these costs vary over space and time.
 
-In this study, we test how lifelong telomere dynamics are related to environmental variation across `r length(table(dd$Cohort))` Seychelles warbler cohorts.
-We first study how telomere length and rates of shortening are related to age, and how this relationship varies among cohorts, in order to gain an in-depth understanding of the temporal dynamics of telomere changes.
-We then examine, within individuals, how telomere length changes with age, and statistically test whether observed increases in telomere length are larger than expected based on measurement error.
+In this study, we test how lifelong telomere dynamics are related to environmental variation across 22 Seychelles warbler cohorts (years).
+We first study how telomere length and rates of shortening are related to age and sex across all life stages, and how this relationship varies among cohorts, in order to gain an in-depth understanding of the temporal dynamics of telomere changes.
+We then examine, within individuals, how telomere length changes with age, and statistically test whether observed increases in telomere length across longitudinal samples are larger than expected based on measurement error.
 Finally, we test how a wide range of social and environmental variables are related to telomere length and shortening in order gain a fuller understanding of the forces driving telomere dynamics in natural populations.

### 2017-02-02 Manuscript/SW_early_life_telomeres.Rmd
Edits

@@
 The population of *ca*.
-320 birds on Cousin Island (04'20'S, 55'40'E) has been intensively studied since 1986 [@Richardson2003b; @Spurgin2014a].
-This species' main breeding season runs from June-September, when the breeding females on each of the *ca*.
-115 territories lay one or, rarely, two or three eggs [@Komdeur1991].
-As a result of this low reproductive output, combined with higher mortality in first-year birds [39% in first-year birds versus 16% in adults; @Brouwer2006], cohort sizes in the Seychelles warbler are typically small (< 50).
+320 adult birds on Cousin Island (04'20'S, 55'40'E) has been intensively studied since 1986 [@Komdeur1992; @Richardson2003b; @Spurgin2014a].
+).
+This species' main breeding season runs from June-September (though a small number of territories do sometimes try to breed between January-March) when the breeding females on many of the ca.
+110 territories will attempt to breed, laying one or, rarely, two or three eggs [@Komdeur1991].
+Breeding attempts are often unsuccessful, and as a result of this low reproductive output, and higher mortality in first-year birds [39% in first-year birds versus 16% in adults; @Brouwer2006], cohort sizes in the Seychelles warbler are typically small (< 50).
+The 22 birth year cohorts used in this study cover 1993 to 2014 – the time period during which our data and sampling is most complete.
 
-Individuals are usually ringed (with an individually numbered metal ring and unique combination of colour rings) in their first year of life, and so are of known age.
-We aged all birds using information on eye colour [@Komdeur1991a] and previous captures [@Richardson2003b].
-As Seychelles warblers are non-migratory endemics naturally confined to the island [@Komdeur2004a], a biannual census of birds on Cousin during each breeding season gives accurate measures of local density, social status (e.g.
-breedier, helper, non-helper) and individual survival [@VandeCrommenacker2011; @Barrett2013a].
+The majority of individuals are ringed (with an individually numbered metal ring and unique combination of colour rings) within the first year of life, and so are of known age.
+We aged all birds using information on eye colour at first capture [@Komdeur1991a] and previous capture history [@Richardson2003b].
+As Seychelles warblers are non-migratory endemics naturally confined to the island [@Komdeur2004a], an extensive biannual census of birds on Cousin during each breeding season gives accurate measures of local density, social status (e.g.
+breeder, helper, non-helper) and individual survival [@VandeCrommenacker2011; @Barrett2013a].
 Full details of monitoring methods can be found in Brouwer *et al.* [-@Brouwer2012].
 
-Seychelles warblers are highly territorial and all territories were mapped during each main breeding season using detailed observational data of foraging and territorial defence behaviour, and surveyed for territory quality [@Richardson2003b].
+Seychelles warblers are highly territorial and all territories were mapped during each main breeding season using detailed observational data of foraging and territorial defence behaviour [@Richardson2003b].
 Territory quality is calculated based on territory size, foliage cover and insect abundance [@Komdeur1992].
 Where territory quality estimates were not available for a specific year we used the average value for that territory across years [@Hammers2013].
-Cousin is subject to intra- and inter-annual variation in rainfall and food availability, and such island-wide temporal variation may override the effects of individual territory quality.
-As an estimate of seasonal variation in food availability, we calculated an index of the number of insects across the entire island during each main breeding season (referred to hereafter as 'insect abundance').
-This index is calculated as the mean number of insects found per unit leaf area over all surveys carried out on the island in a main breeding season.
+Cousin is subject to considerable intra- and inter-annual variation in rainfall and, consequently, insect availability.
+Such island-wide temporal variation may override the effects of variation in individual territory quality across the island.
+As an estimate of seasonal variation in food availability, we calculated an index of the abundance of insects across the entire island during each main breeding season (referred to hereafter as 'insect abundance').
+This index is calculated as the mean number of insects found per unit leaf area over all monthly surveys carried out on the island in a main breeding season.
 
-Each time a bird is caught on Cousin body mass and tarsus length are measured (to the nearest 0.1g and 0.1mm, respectively).
-A blood sample (*ca* 25 $\mu$l) is taken from each bird captured via brachial venipuncture, and stored at room temperature in 1 ml of absolute ethanol in a 1.5 ml screw-cap microfuge tube.
+Each time a bird is caught on Cousin a range of morphometric measurements are taken, including body mass and tarsus length (to the nearest 0.1g and 0.1mm, respectively).
+A blood sample (ca 25 l) is taken via brachial venipuncture, and stored at room temperature in 1 ml of absolute ethanol in a 1.5 ml screw-cap microfuge tube.
 
@@
 Sex was determined using the PCR-based method outlined by Griffiths *et al.* [-@Griffiths1998].
-Prior to telomere analysis, DNA concentration and purity were quantified using a NanoDrop 8000 Spectrophotometer (ThermoScientific), and the following thresholds were applied before samples were included for further analysis: i) DNA concentration must be at least 15 ng $\mu$l^-1^ (based on a mean of three measurements), ii) the 260/280 absorbance ratio has to be between 1.8 and 2 for acceptable DNA purity and, iii) the 260/230 absorbance ratio must be higher than 1.8.
+Prior to telomere analysis, DNA concentration and purity were quantified using a NanoDrop 8000 Spectrophotometer (ThermoScientific).
+The following thresholds were applied before samples were included for further analysis: i) DNA concentration must be at least 15 ng $\mu$l^-1^ (based on a mean of three measurements), ii) the 260/280 absorbance ratio has to be between 1.8 and 2 for acceptable DNA purity and, iii) the 260/230 absorbance ratio must be higher than 1.8.
 DNA integrity was further validated by visualization with ethidium bromide after electrophoresis on a 1.2% agarose gel, and all samples with evidence of DNA degradation were re-extracted or excluded.
@@
 We measured relative telomere length (RTL) for all samples using a quantitative PCR (qPCR) assay of telomeres and a GAPDH control gene, following Bebbington *et al.* [-@Bebbington2016].
+Based on the distribution of observed cq values, we excluded outlier samples with extremely large cq values (cq values > 25 and 26 were excluded for the telomere and GAPDH reactions, respectively), which were assumed to be failed reactions.
 
-For a subset of birds we had longitudinal data, with one or more additional samples taken at different ages (n = `r length(unique(c(dd3$BloodID,dd3$BloodID1)))` measurements from `r length(unique(dd3$BirdID))` birds).
+For a large subset of birds we had longitudinal data, with two or more samples taken at different ages (n = `r length(unique(c(dd3$BloodID,dd3$BloodID1)))` measurements from `r length(unique(dd3$BirdID))` birds).
 For these individuals we calculated the within-individual change in RTL by subtracting RTL at time point *t* from RTL at time point *t* + 1 (hereafter $\Delta$RTL, n = `r nrow(dd3)` measurements).
-Negative values of $\Delta$RTL reflect decreases in telomere length with age, while positive values increases.
+Negative values of RTL reflect decreases in telomere length with age, while positive values reflect increases.
 
@@
 RTL was square root transformed to improve linear model fits, and we assessed repeatability of RTL using the rptR package.
+
 We explored the cross-sectional relationship between RTL and age among cohorts using linear mixed models (LMMs) carried out in the lme4 package [@Bates2014].
@@
 We created models where the relationship between RTL and age was linear, quadratic, log-linear, and where age was fitted as a factor.
-For each age term, we fitted additional models including birth year (cohort) as a factor, and an interactions between cohort age.
+For each age term, we fitted additional models including birth year (cohort) as a factor, and an interaction between cohort age.
 All fitted models are included in Table 1.
+Note that we do not carry out full model selection or model averaging here, as our aim was to compare a set of specifically defined models.
 For random effects we included individual ID, and catch year.
-Models were compared using AIC with correction for finite sample size (AICc).
+Models were compared using AIC with correction for finite sample size [AICc; @Hurvich1989].
 
@@
 
-When examining the distribution of longitudinal telomere changes we observed increases in telomere length with age, so we used repeat measurements to test whether these increases could be explained by measurement error.
-We calculated the change in RTL between pairs of repeat measurements within samples (hereafter $\Delta$RTL~*sample*~; N = `r nrow(subset(ddL,Group == 'Within samples'))` measurements from `r length(unique(subset(ddL,Group == 'Within samples')$BirdID))` samples) in exactly the same way as for across samples (hereafter $\Delta$RTL~*individual*~), using completely separate reactions run on separate plates.
+When examining the distribution of longitudinal telomere changes we observed some increases in telomere length with age in individuals.
+We then used repeat measurements to test whether these increases could be explained by measurement error.
+We calculated the change in RTL between pairs of repeat measurements within the same samples (hereafter $\Delta$RTL~*sample*~; N = `r nrow(subset(ddL,Group == 'Within samples'))` measurements from `r length(unique(subset(ddL,Group == 'Within samples')$BirdID))` samples) in exactly the same way as for across samples (hereafter $\Delta$RTL~*individual*~), using completely separate reactions run on separate plates.
 To test whether greater changes in RTL were observed among individuals compared to among repeat samples, we compared the variance in $\Delta$RTL~*sample*~ and $\Delta$RTL~*individual*~ using a Levene's test.
@@
 We created a full model with RTL as a response variable, alongside the following explanatory variables: log age (based on the RTL and age analysis; see results), tarsus length, body mass, sex, insect abundance, territory quality, island-wide population density (an annual measure estimated from the summer breeding census), territory group size, and the number of helping subordinate birds present in the territory.
-The random effects structure was informed by the analysis of telomere dynamics and age (see results): we included individual ID, cohort ID, and a random slope of log age among cohorts.
+The random effects structure was informed by the analysis of telomere dynamics and age (see results): we included individual ID, cohort ID, and a random slope of log age among cohorts (to allow the effect of age on RTL to vary among cohorts).
 We report model estimates and confidence intervals for all effects included in the full model.
@@
 Efficiencies (mean $\pm$ s.d.) for our telomere and GAPDH reactions were (`r round2(mean(dd$EffTelomere))` $\pm$ `r round2(sd(dd$EffTelomere))`) and (`r round2(mean(dd$EffGAPDH))` $\pm$ `r round2(sd(dd$EffGAPDH))`) respectively.
-Inter-plate repeatability of RTL, based on `r nrow(subset(ddL,Group == 'Within samples'))` samples measured at least twice, was `r round2(rptall$R)` (CI = `r round2(rptall$CI_emp[1])`, `r round2(rptall$CI_emp[2])`).
+Inter-plate repeatability of RTL, based on `r nrow(subset(ddL,Group == 'Within samples'))` samples measured at least twice at different timepoints, was `r round2(rptall$R)` (CI = `r round2(rptall$CI_emp[1])`, `r round2(rptall$CI_emp[2])`).
 
-#####Telomere dynamics and age amnong cohorts
+#####Telomere dynamics and age among cohorts
 We first tested how RTL was related to age among cohorts using a model selection approach.
@@
 There was a negative relationship between RTL and age in 21 of the `r length(table(dd$Cohort))` cohorts, but the slope of the relationship between log age and RTL varied substantially among cohorts (Fig.1C).
-In the one year in which RTL increased with age (2013), 17 of the 18 birds were fledglings or subadults, suggesting that the observed pattern was an artifact of the sampling in this season rather than a real relationship.
+In the one year in which RTL increased with age (2013), 17 of the 18 birds sampled were fledglings or subadults, suggesting that the observed pattern was an artifact of the sampling in this season (i.e.
+a lack of variation in age among sampled birds) rather than a real relationship.
 
@@
 
-There was positive correlation between telomere measurements taken at different time points (Fig.
+There was positive correlation between RTL measured from different samples taken at different time points during an individual’s life (Fig.
 2A), but this was very weak (marginal R^2^ = `r round2(r.squaredGLMM(RTL_RTL1)['R2m'])`), and not significant (estimate = `r round3(fixef(RTL_RTL1)['RTL'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_RTL1)['RTL',],lessthan = F),collapse = ', ')`).
-Although both cross-sectional and longitudinal data indicated a general trend of telomere shortening with age, we found that RTL within individuals increased in `r round((nrow(subset(dd3,DeltaRTL>0))/nrow(dd3))*100,0)`% of our `r nrow(dd3)` $\Delta$RTL~*individual*~ measurements (Fig.
+Although both cross-sectional and longitudinal data indicated a general trend of telomere shortening with age, we found that RTL - measured across two samples taken from the same individuals over time - increased with age in `r round((nrow(subset(dd3,DeltaRTL>0))/nrow(dd3))*100,0)`% of our `r nrow(dd3)` $\Delta$RTL~*individual*~ measurements (Fig.
 2A).
-To test whether increases in telomere length in our dataset could be explained by measurement error, we compared variance in telomere length among repeat measurements of samples to the variance observed among different samples of the same individual.
+To test whether increases in telomere length in our dataset could be explained by measurement error, we compared variance in telomere length among repeat measurements of the same samples to the variance observed among different samples of the same individual.
 We found significantly higher variance in telomere length over individual lifetimes compared to among sample replicates (Levene's test: F = `r round2(leveneTest(DeltaRTL~Group,data=ddL)['F value'][1,1])`; P `r round3(leveneTest(DeltaRTL~Group,data=ddL)['Pr(>F)'][1,1])`; Fig.

### 2017-02-02 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Home edits

@@
 
-table1A <- data.frame(Model = apply(t1,1,getterms,c(2:10)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
+table1A <- data.frame(Model = apply(t1,1,getterms,c(2:6)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
 colnames(table1A) <- c('Model','df','AICc', 'Delta AICc','Weight')
@@
 
-table1B <- data.frame(Model = apply(t1,1,getterms,c(2:8)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
+table1B <- data.frame(Model = apply(t1,1,getterms,c(2:6)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
 colnames(table1B) <- c('Model','df','AICc', 'Delta AICc','Weight')
@@
 
+dd3$Gainer <- ifelse(dd3$DeltaRTL>0,1,0)
+
+Fig2C <- ddply(dd3,
+           .(Agemonths),
+           summarise,
+           PropGainers = mean(Gainer),
+           n = length(Gainer)) %>%
+  subset(n > 20) %>%
+  ggplot(aes(x = Agemonths,
+           y = PropGainers)) +
+  ylab("Proportion gainers") + 
+  xlab("Age") +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        legend.position = c('none'),
+        legend.title = element_text(size = 12),
+        axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black')) +
+    geom_point(col = 'grey')+
+  geom_line(col = 'grey')+
+  annotate('text',x = 0,y = 0.6,label = 'A')
+
 pdf('Figures/Figure 2.pdf',
-    width = 4,
-    height = 8,
+    width = 7,
+    height = 7,
     paper = 'a4')
-multiplot(Fig2A,Fig2B)
+multiplot(Fig2A,Fig2B,Fig2C,layout = matrix(c(1,2,3,3),nrow = 2,byrow = T))
 dev.off()
@@
 CIs <- CIwald(RTL_Full2)
-t1a <- data.frame(myx = row.names(CIs)[7:nrow(CIs)],
+t1a <- data.frame(myx = row.names(CIs)[6:nrow(CIs)],
                   Estimate = fixef(RTL_Full2),
-                  X2.5. = CIs[7:nrow(CIs),1],
-                  X97.5. = CIs[7:nrow(CIs),2],
+                  X2.5. = CIs[6:nrow(CIs),1],
+                  X97.5. = CIs[6:nrow(CIs),2],
                   row.names = NULL,
@@
 nams <- t1a$myx
-nams[nams == 'z.LogAge'] <- 'Age'
+nams[nams == 'z.LogAge'] <- 'Log age'
 nams[nams == 'z.Tarsus'] <- 'Tarsus length'

### 2017-02-02 Scripts/SW juvenile telomeres IPC models.r

@@ -6,13 +6,13 @@
 
 # Telomere length, age and cohort -----------------------------------------
 
-RTL_Age <- lmer(RTL ~ Cohort*(Agemonths + LogAge + FAge + Age2) + (1|BirdID) + (1|CatchYear),
+RTL_Age <- lmer(RTL ~ Cohort + Agemonths + LogAge + FAge + Age2 + (1|BirdID) + (1|CatchYear),
                 data=dd,
                 REML = FALSE)
 RTL_LogAge <- lmer(RTL ~ LogAge + (1|BirdID) + (1|CatchYear),
                 data=dd,
                 REML = FALSE)
-RTL_DeltaAge <- lmer(RTL ~ MeanAge + Cohort*(DeltaAge + DeltaLogAge + DeltaAge2) + (1|BirdID) + (1|CatchYear),
+RTL_DeltaAge <- lmer(RTL ~ MeanAge + Cohort + DeltaAge + DeltaLogAge + DeltaAge2 + (1|BirdID) + (1|CatchYear),
                 data=dd3,
                 REML = FALSE)
 RTL_DeltaLogAge <- lmer(RTL ~ DeltaLogAge + (1|BirdID) + (1|CatchYear),

### 2017-02-03 Manuscript/SW_early_life_telomeres.Rmd
Edits

@@
 Telomere length was positively associated with island-wide temporal insect abundance - a key resource for the insectivorous Seychelles warbler - suggesting that the costs associated with living in harsher environments can be studied by investigating telomere dynamics.
-We also found evidence for sex-specific relationships between telomeres and tarsus length, potentially reflecting differential costs of growth, which leads to adult females having significantly shorter telomeres than males.
+We also found evidence for sex-specific relationships between telomeres and tarsus length, potentially reflecting differential costs of growth.
 

### 2017-02-03 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Edits

@@
 
-dd3$Gainer <- ifelse(dd3$DeltaRTL>0,1,0)
+temp <- subset(dd3,TimeDiff < 4)
+temp$MidAge <- (temp$Agemonths+temp$Agemonths1)/2
+temp$Gainer <- ifelse(temp$DeltaRTL>0,1,0)
 
-Fig2C <- ddply(dd3,
-           .(Agemonths),
-           summarise,
-           PropGainers = mean(Gainer),
-           n = length(Gainer)) %>%
-  subset(n > 20) %>%
-  ggplot(aes(x = Agemonths,
-           y = PropGainers)) +
+Fig2C <- ggplot(temp, aes(x = Agemonths,
+           y = Gainer)) +
   ylab("Proportion gainers") + 
@@
   geom_point(col = 'grey')+
-  geom_line(col = 'grey')+
-  annotate('text',x = 0,y = 0.6,label = 'A')
+  geom_count(alpha = 0.2)+
+  geom_smooth()+
+  annotate('text',x = -0.2,y = 1,label = 'C')
 
@@
 ```
+
+ggplot(temp,aes(x = TimeDiff,y = DeltaRTL))+
+geom_point()+
+geom_smooth()
 

### 2017-02-06 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Edits

@@
 
-Fig2A <- ggplot(dd3,
+temp <- dd3
+temp$Gainer <- ifelse(temp$DeltaRTL>0,"1","0")
+
+Fig2A <- ggplot(temp,
        aes(x = RTL,
-           y = RTL1)) +
+           y = RTL1,
+           col = Gainer)) +
   ylab(expression('RTL ('*italic(t)*' + 1)')) + 
@@
         axis.line.y = element_line(colour = 'black')) +
-    geom_point(col = 'grey')+
+    geom_point()+
   geom_abline(slope = 1,intercept = 0,lty = 2) +
+  scale_colour_manual(values = c("navy","red"))+
   annotate('text',x = 0.3,y = 1.7,label = 'A')
@@
 
-temp <- subset(dd3,TimeDiff < 4)
-temp$MidAge <- (temp$Agemonths+temp$Agemonths1)/2
-temp$Gainer <- ifelse(temp$DeltaRTL>0,1,0)
-
 Fig2C <- ggplot(temp, aes(x = Agemonths,
-           y = Gainer)) +
+           y = DeltaRTL,col = Gainer)) +
   ylab("Proportion gainers") + 
@@
         axis.line.y = element_line(colour = 'black')) +
-  geom_point(col = 'grey')+
-  geom_count(alpha = 0.2)+
-  geom_smooth()+
+  geom_point()+
+  geom_smooth(col = "darkgrey",method = "lm",formula = y~log(x))+
+  geom_abline(slope = 0,intercept = 0,lty = 2)+
+  scale_colour_manual(values = c("navy","red"))+
+  annotate('text',x = -0.2,y = 1,label = 'C')
+
+Fig2D <- ddply()
+ggplot(temp, aes(x = Agemonths,
+           y = DeltaRTL,col = Gainer)) +
+  ylab("Proportion gainers") + 
+  xlab("Age") +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        legend.position = c('none'),
+        legend.title = element_text(size = 12),
+        axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black')) +
+  geom_point()+
+  geom_smooth(col = "darkgrey",method = "lm",formula = y~log(x))+
+  geom_abline(slope = 0,intercept = 0,lty = 2)+
+  scale_colour_manual(values = c("navy","red"))+
   annotate('text',x = -0.2,y = 1,label = 'C')
@@
     paper = 'a4')
-multiplot(Fig2A,Fig2B,Fig2C,layout = matrix(c(1,2,3,3),nrow = 2,byrow = T))
+multiplot(Fig2A,Fig2B,Fig2C,Fig2D,layout = matrix(c(1,2,3,4),nrow = 2,byrow = T))
 dev.off()

### 2017-02-06 Scripts/SW juvenile telomeres IPC models.r

@@ -18,6 +18,9 @@ RTL_DeltaAge <- lmer(RTL ~ MeanAge + Cohort + DeltaAge + DeltaLogAge + DeltaAge2
 RTL_DeltaLogAge <- lmer(RTL ~ DeltaLogAge + (1|BirdID) + (1|CatchYear),
                    data=dd3,
                    REML = FALSE)
+DeltaRTL_LogAge <- lmer(DeltaRTL ~ LogAge + (1|BirdID) + (1|CatchYear),
+                        data=subset(dd3,TimeDiff < 2),
+                        REML = FALSE)
 RTL_RTL1 <- lmer(RTL1 ~ RTL + LogAge + (1|BirdID) + (1|LayYear),
                         data=dd3,
                         REML = FALSE)

### 2017-02-06 Manuscript/SW_early_life_telomeres.Rmd
Eds

@@
 2B).
+To better visualise how longitudinal telomere dynamics, including the likelihood of lengthening, varied with age, we tested how $\Delta$RTL changed with (log) age.
+We found that $\Delta$RTL increased with log age (estimate = `r round3(fixef(DeltaRTL_LogAge)['LogAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(DeltaRTL_LogAge)['LogAge',],lessthan = F),collapse = ', ')`), again showing that
 

### 2017-02-07 Manuscript/SW_early_life_telomeres.Rmd
Edits

--- 2017-02-03 Manuscript/SW_early_life_telomeres.Rmd
@@
 2B).
+To better visualise how longitudinal telomere dynamics, including the likelihood of lengthening, varied with age, we tested how $\Delta$RTL changed with (log) age.
+To account for the potential confounding effect of follow-up time, we only included pairs of samples taken within two years of one another.
+We found that $\Delta$RTL increased with log age (estimate = `r round3(fixef(DeltaRTL_LogAge)['LogAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(DeltaRTL_LogAge)['LogAge',],lessthan = F),collapse = ', ')`), and raw data indicated that both the degree of telomere shortening, and the likelihood of shortening, declined as individuals became older (Fig.
+2C).
 

### 2017-02-09 Manuscript/SW_early_life_telomeres.Rmd
Edits

--- 2017-02-03 Manuscript/SW_early_life_telomeres.Rmd
@@
 2B).
+To better understand how longitudinal telomere dynamics vary with age, we exmained patterns of short-term telomere change, including only pairs of samples taken within two-years of each other.
+We found that tongitudinal telomere length increased with log age both when considered as the absolute degree of change (LMM with $\Delta$RTL~*individual*~ as response; estimate = `r round3(fixef(DeltaRTL_LogAge)['LogAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(DeltaRTL_LogAge)['LogAge',],lessthan = F),collapse = ', ')`; Fig.
+2C), and the likelihood of lengthening (GLMM with lengthened yes/no as binomial response; estimate = `r round3(fixef(DeltaRTLF_LogAge)['LogAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(DeltaRTLF_LogAge)['LogAge',],lessthan = F),collapse = ', ')`; Fig.
+2D).
+Increases in telomere length were most likely to be obsevrved shortly after the juvenile period, at around four years of age, and later in life (although sample sizes for older birds are much smaller; Fig.
+2C,D).
 

### 2017-02-09 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Edits

--- 2017-02-03 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
@@
 
+temp <- subset(dd3,TimeDiff < 3.5)
+
 Fig2A <- ggplot(dd3,
@@
         axis.line.y = element_line(colour = 'black')) +
-    geom_point(col = 'grey')+
+    geom_point(col = "grey")+
   geom_abline(slope = 1,intercept = 0,lty = 2) +
@@
 
-temp <- subset(dd3,TimeDiff < 4)
-temp$MidAge <- (temp$Agemonths+temp$Agemonths1)/2
-temp$Gainer <- ifelse(temp$DeltaRTL>0,1,0)
-
 Fig2C <- ggplot(temp, aes(x = Agemonths,
-           y = Gainer)) +
-  ylab("Proportion gainers") + 
-  xlab("Age") +
+           y = DeltaRTL)) +
+  ylab(expression(Delta*'RTL'))+
+  xlab("Age (years)") +
   theme_classic() +
@@
         axis.line.y = element_line(colour = 'black')) +
-  geom_point(col = 'grey')+
-  geom_count(alpha = 0.2)+
-  geom_smooth()+
+  geom_point(col = "grey")+
+  geom_smooth(col = 'black',fill = 'black',method = "lm",formula = y~log(x))+
+  geom_abline(slope = 0,intercept = 0,lty = 2)+
   annotate('text',x = -0.2,y = 1,label = 'C')
+
+Fig2D <- ddply(temp,
+               .(Agemonths),
+               summarise,
+               PropGainer = mean(DeltaRTLF),
+               n = length(DeltaRTLF)) %>%
+ggplot(aes(x = Agemonths,
+           y = PropGainer)) +
+  ylab("Proportion gainers") + 
+  xlab("Age (years)") +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        legend.position = c('none'),
+        legend.title = element_text(size = 12),
+        axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black')) +
+  geom_line(col="darkgrey")+
+  geom_count(data = temp,aes(y = DeltaRTLF,x = Agemonths),col = "grey")+
+  annotate('text',x = 0,y = 1.1,label = 'D')
 

### 2017-02-09 Scripts/SW juvenile telomeres IPC models.r

@@ -21,6 +21,9 @@ RTL_DeltaLogAge <- lmer(RTL ~ DeltaLogAge + (1|BirdID) + (1|CatchYear),
 DeltaRTL_LogAge <- lmer(DeltaRTL ~ LogAge + (1|BirdID) + (1|CatchYear),
                         data=subset(dd3,TimeDiff < 3.5),
                         REML = FALSE)
+DeltaRTLF_LogAge <- glmer(DeltaRTLF ~ LogAge + (1|BirdID) + (1|CatchYear),
+                        data=subset(dd3,TimeDiff < 3.5),
+                        family = "binomial")
 RTL_RTL1 <- lmer(RTL1 ~ RTL + LogAge + (1|BirdID) + (1|LayYear),
                         data=dd3,
                         REML = FALSE)

### 2017-02-09 Manuscript/SW_early_life_telomeres.Rmd
Edits

@@
 ###Discussion
-Here we use a long-term, multi-cohort dataset to assess the relationships between spatio-temporal variation in the ecological environment and lifelong telomere dynamics in a contained population of Seychelles warblers.
-We found that telomere length decreases with age, and that this decrease is greatest in early life.
-Telomere length decreased with age in almost all of the `r length(table(dd$Cohort))` cohorts studied, but telomere length, and the rate of decrease of telomere length with age, varied substantially among cohorts.
-Despite an overall trend for telomere shortening in the Seychelles warbler, we found that telomere length increased within some individuals at particular life stages, and that the extent of these increases cannot be explained solely by qPCR measurement error.
+Here we use a long-term, multi-cohort dataset to assess lifelong telomere dynamics and the relationship between these and spatio-temporal variation in the ecological environment in a contained population of Seychelles warblers.
+We found that telomere length decreases with age, and that this decrease is greatest very early in life.
+Telomere length decreased with age in almost all of the `r length(table(dd$Cohort))` cohorts studied, but the rate of decrease of telomere length with age, and consequently the telomere length of adults, varied substantially among cohorts.
+Despite an overall pattern of telomere shortening with age in the Seychelles warbler, we found that our measure of telomere length increased within some individuals at particular life stages, and that the extent of these increases could not be explained solely by qPCR measurement error.
 Finally, we found that telomeres are related to tarsus length in a sex-specific manner, and that telomere length is positively associated with temporal fluctuations in food availability.
@@
 However, we also found that despite an overall trend for shortening, telomere length both increased and decreased, especially after the juvenile period.
-Importantly, these increases were observed in longitudinal as well as cross-sectional data, suggesting that selective disappearance of individuals with shorter telomeres does not explain this pattern entirely.
-Longitudinal increases in measured telomere length have been observed in humans and [reviewed in @Steenstrup2013] wild animals [@Kotrschal2007; @Fairlie2016].
+Importantly, these increases were observed in longitudinal as well as cross-sectional data, indicating that selective disappearance of individuals with shorter telomeres does not explain this pattern.
+Longitudinal increases in measured telomere length have been observed in humans and wild animals  [@Steenstrup2013;@Kotrschal2007; @Fairlie2016; @Hoelzl2016; @Hoelzl2016a].
 The most commonly invoked explanation for increases in telomere length is measurement error, which can be a particular problem in qPCR-based telomere studies [@Nussey2014; @Steenstrup2013; @Verhulst2015].
@@
 
-Increases in telomere length were not consistent over individual lifespans, but increases occured at specific, short periods, against a backdrop of overall telomere shortening.
+Increases in telomere length were not consistent over individual lifespans, but occurred during specific, periods, against a backdrop of overall lifelong telomere shortening.
+This is consistent with recent findings in edible dormice *Glis glis*, in which telomere elongation was observed only later in life [@Hoelzl2016a].
 Consistent with a pattern of sporadic changes in telomere length with age, we found that within-individual telomere measurements were only weakly correlated.
@@
 
-The finding that increases in telomere length may be sporadic and overlaid onto an overall pattern of shortening is an important point when assessing the occurrence of telomere lengthening.
+The finding that increases in telomere length may be sporadic and overlaid on an overall pattern of shortening with age is an important point when assessing the occurrence of telomere lengthening.
 Previously described approaches to distinguish telomere elongation from measurement error, based on assumptions about follow-up time between measurements [@Steenstrup2013], or based on measuring variance among measurements [@Simons2014], assume that telomere elongation within individuals is consistent over time.
-Such sporadic changes in telomere length over lifespans could occur due to changes in the composition of cell types within individual samples, or due to the actual elongation of telomeres.
+Our data, and that of others [@Fairlie2016;@Hoelzl2016; @Hoelzl2016a] suggest that this is not the case.
+Such inconsistent changes in telomere length over lifespans could occur due to changes in the composition of cell types within individual samples, or due to the actual elongation of telomeres [@Blackburn1989].
 Determining the mechanism of these changes is essential for how we view telomeres as biomarkers of costs.
@@
 
-A few studies have shown that temporal variation in telomere dynamics occurs in natural populations, although these have been limited in the number of seasons [@Mizutani2013; @Watson2015; @Fairlie2016].
-The long-term Seychelles warbler dataset has allowed us to show that temporal variation in telomere dynamics can occur over substantial time periods.
-Our data suggest that birth year is a highly important factor in shaping telomere dynamics, but also that age-related declines in telomere length vary among cohorts.
-This is consistent with telomere length being controlled by both genetic and environmental factors - something that is becoming apparent from quantitative genetic studies of telomere dynamics.
+Measurement of cohorts across seasons or years is required if we are to understand how the environment impacts telomere dynamics.
+Although a few studies have shown that temporal variation in telomere dynamics occurs in natural populations, these have been limited in the number of seasons they cover [@Mizutani2013; @Watson2015; @Fairlie2016].
+TThe long-term Seychelles warbler dataset has allowed us to show that temporal variation in telomere dynamics can occur over substantial time periods.
+Our data suggest that conditions during the hatch year are a very important factor in shaping telomere dynamics, and that age-related declines in telomere length vary among cohorts.
+The rate of early-life telomere loss for a cohort was highly variable and had a persistent effect of the mean adult telomere length of the cohort.
+This effect is consistent with telomere length being strongly influenced by environmental, as well as genetic, factors - something that is becoming apparent from the, as yet, limited number of quantitative genetic studies of telomere dynamics in wild populations [e.g.
+@Olsson2011; @Atema2015; @Becker2015].
 Moreover, our findings suggest that the telomere dynamics of a population at a given point in time represent a snapshot of a temporally varying process.
-Research of telomere dynamics within and across multiple cohorts and populations will enable us to better understand how how and why population-level telomere dynamics vary over space and time.
+Research of telomere dynamics within and across multiple cohorts and populations will enable us to better understand how and why population-level telomere dynamics vary over space and time.
 
-We found that temporal variation in population-level food availability was positively related to telomere length.
-This is consistent with the strong cohort effects we found, and suggests that temporal variation in conditions may be a key driver of costs in the Seychelles warbler.
+We found that temporal variation in insect prey availability was positively related to telomere length.
+This is consistent with the strong cohort effects we found, and suggests that temporal variation in environmental conditions may be a key driver of costs in the Seychelles warbler.
 Although the environmental conditions on Cousin Island are relatively benign in comparison to other island systems [e.g.
-@Coulson2001], substantial annual variation in rainfall does occur, with associated changes in insect abundance, and it appears that this confers a cost to Seychelles warblers.
-We also found evidence for sex-specific telomere dynamics, with males having longer-telomeres than females, and that this sex-difference interacts with tarsus length.
-If the sex-dependent relationship between telomere and tarsus lengths was due to differential growth alone we would expect the opposite pattern to that observed, as male Seychelles warblers are larger than females (Fig.
+@Coulson2001], substantial annual variation in rainfall does occur, with associated changes in insect abundance [@Komdeur1996], and it appears that this confers a cost - in terms of intrinsic biological condition - to Seychelles warblers.
+Our results concur with other studies which show that early life  conditions / food availability can have a very significant and long term impacts on telomere length (and intrinsic biological condition) in captive and wild animals [e.g.
+@Nettle2015; @Heidinger2012; @Watson2015].
+
+We also found evidence for sex-specific telomere dynamics: males had longer-telomeres than females.
+Interestingly this sex-difference interacts with tarsus length: telomere length was negatively correlated with tarsus length in both sexes, but this effect was stronger in females than males.
+If the sex-dependent relationship between telomere and tarsus length was due to differential growth alone we would expect the opposite pattern to that observed, as male Seychelles warblers are larger than females (Fig.
 3B).
@@
 
-Although we found clear associations between the environment and telomere dynamics, we should bear in mind that our social and ecological variables explained a only a small proportion of the variance in RTL.
+Although we found clear associations between the environment and telomere dynamics, we should bear in mind that the social and ecological variables we tested here explained only a small proportion of the variance in RTL.
 A poor social and ecological environment is known to be detrimental to Seychelles warblers, both in terms of oxidative stress and survival [@VandeCrommenacker2011a; @Brouwer2012], and it is therefore perhaps surprising that these variables do not explain more variance in RTL.
-While measurement error clearly accounts for some of this lack of explanatory power, it is also likely that early-life RTL in the Seychelles warbler is explained by a set of environmental and genetic variables not considered here [e.g.
+While telomere measurement error will account for some of this lack of explanatory power, it is also likely that early-life RTL in the Seychelles warbler is explained by a set of environmental and genetic variables not considered here [e.g.
 @Bebbington2016].
@@
 @Becker2015; @Asghar2014; @Heidinger2016].
+For example, parental age and quality may key variables that impact the telomere dynamics of offspring in the Seychelles warbler, and will be addressed in future studies.
 Long-term ecological study systems are uniquely suited to addressing such questions in natural systems [@Clutton-Brock2010].

### 2017-02-09 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Edits

@@
 **A** RTL and age across all individuals.
-Points and  thin grey lines represent individual samples and birds, respectively.
-The thick line and shaded area represent the fitted values and 95% confidence limits of a linear regression of RTL and log-transformed age.**B** Boxplot of variation in RTL among cohorts.
+Points and  connecting thin grey lines represent individual samples and birds, respectively.
+The thick line and shaded area represent the fitted values and 95% confidence limits of a linear regression of RTL and log-transformed age.
+**B** Boxplot of variation in RTL among cohorts.
 **C** RTL and age among cohorts.
-Lines represent fitted values from a linear regression and log-transformed age, and colours correspond to **B**.
+Lines represent fitted values from a linear regression of RTL and log-transformed age, and colours correspond to **B**.
 **D** RTL in relation to and $\Delta$Log age (i.e.
@@
 The dotted line represents parity, and thus points above and below the line represent increases and decreases in RTL, respectively.
-**B** Scaled density plots of repeated RTL measurements among individual samples (grey), and among different samples taken from the same individual (yellow).
+**B** Scaled density plots of repeated RTL measurements among individual samples, and among different samples taken from the same individual.
 Areas of the density plot to the left of the dotted line represent decreases in RTL, while areas to the right represent increases.
+**C** $\Delta$RTL in relation to age in pairs of samples taken within two years.
+Black line and shaded area represent fitted values and 05% confidence limits from a linear regression of RTL and log-transformed age.
+**D** Telomere lengthening and shortening in relation to age.
+Points at zero and one represent pairs of samples where RTL has decreased and increased, respectively, with point size scaled by the number of overlapping values.
+The grey line represents the proportion of samples in which i ncreases in RTL where observed at each age.
 
@@
         axis.title.x = element_text(size = 14),
-        legend.position = 'none',
+        legend.position = "top",
         legend.title = element_blank(),
@@
         axis.title.x = element_text(size = 14),
-        legend.position = c('none'),
-        legend.title = element_text(size = 12),
         axis.line.x = element_line(colour = 'black'),
@@
         axis.line.y = element_line(colour = 'black'),
-        legend.position = 'none') +
+        legend.position = 'top',
+        legend.title = element_blank()) +
   stat_smooth(method = 'lm') +

### 2017-02-10 Manuscript/SW_early_life_telomeres.Rmd
Edits

@@
 Here we study spatio-temporal variation in lifelong telomere dynamics in the Seychelles warbler, *Acrocephalus sechellensis*.
-We combine long-term life history and ecological data with a large longitudinal telomere datasets, consisting of `r nrow(dd)` samples from `r length(table(dd$Cohort))` cohorts born between `r min(dd$LayYear)` and `r max(dd$LayYear)`.
+We combine long-term life-history and ecological data with a large longitudinal telomere dataset, consisting of `r nrow(dd)` samples from `r length(table(dd$Cohort))` cohorts born between `r min(dd$LayYear)` and `r max(dd$LayYear)`.
 We provide a detailed analysis of how telomere dynamics vary over individual lifespans and cohorts, and with spatio-temporal variation in the social and ecological environment.
@@
 @Schultner2014; @Asghar2015; @Bebbington2016].
-There is also some evidence that telomere length, measured longitudinally from individuals, can increase as well as decrease [@Simons2014; @Bateson2016a], which has important ramifications for our understanding of how telomeres reflect costs.
+There is also evidence that telomere length, measured longitudinally from individuals, can increase as well as decrease [@Muezzinler2013; @Simons2014; @Bateson2016a], which has important ramifications for our understanding of how telomeres reflect costs.
 However, such increases in telomere length are often attributed to measurement error [@Steenstrup2013; but see @Bateson2016a], and as such its ecological significance is unknown.
@@
 The Seychelles warbler is a small (~15 g), insectivorous passerine bird with a mean life expectancy of 5.5 years at fledging [@Hammers2013].
-The population of *ca*.
-320 adult birds on Cousin Island (04'20'S, 55'40'E) has been intensively studied since 1986 [@Komdeur1992; @Richardson2003b; @Spurgin2014a].
+The population of *ca* 320 adult birds on Cousin Island (04'20'S, 55'40'E) has been intensively studied since 1986 [@Komdeur1992; @Richardson2003b; @Spurgin2014a].
 ).
-This species' main breeding season runs from June-September (though a small number of territories do sometimes try to breed between January-March) when the breeding females on many of the ca.
-110 territories will attempt to breed, laying one or, rarely, two or three eggs [@Komdeur1991].
+This species' main breeding season runs from June-September (though a small number of territories do sometimes try to breed between January-March) when the breeding females on many of the *ca* 110 territories will attempt to breed, laying one or, rarely, two or three eggs [@Komdeur1991].
 Breeding attempts are often unsuccessful, and as a result of this low reproductive output, and higher mortality in first-year birds [39% in first-year birds versus 16% in adults; @Brouwer2006], cohort sizes in the Seychelles warbler are typically small (< 50).
@@
 Each time a bird is caught on Cousin a range of morphometric measurements are taken, including body mass and tarsus length (to the nearest 0.1g and 0.1mm, respectively).
-A blood sample (ca 25 l) is taken via brachial venipuncture, and stored at room temperature in 1 ml of absolute ethanol in a 1.5 ml screw-cap microfuge tube.
+A blood sample (*ca* 25 l) is taken via brachial venipuncture, and stored at room temperature in 1 ml of absolute ethanol in a 1.5 ml screw-cap microfuge tube.
 
@@
 We created models where the relationship between RTL and age was linear, quadratic, log-linear, and where age was fitted as a factor.
-For each age term, we fitted additional models including birth year (cohort) as a factor, and an interaction between cohort age.
+For each age term, we fitted additional models including birth year (cohort) as a factor.
 All fitted models are included in Table 1.
@@
 Using the longitudinal data, we then tested how telomeres change with age in individuals, using LMMs of RTL as a response and $\Delta$age [a longitudinal measure based on within-subject centring; @VandePol2009] as an explanatory variable.
-We calculated $\Delta$age using log and polynomial transformed age data, and carried out model selection as above, with the exception that we did not model $\Delta$age as a factor (due to a lack of discreet groupings), and mean age was also included in models to partition within-individual vs cross-sectional effects [@VandePol2009].
+We calculated $\Delta$age using log and polynomial transformed age data, and carried out model selection as above, with the exception that we did not model $\Delta$age as a factor (due to a lack of discreet groupings), and mean age was also included in models to partition within-individual *vs* cross-sectional effects [@VandePol2009].
 
@@
 Efficiencies (mean $\pm$ s.d.) for our telomere and GAPDH reactions were (`r round2(mean(dd$EffTelomere))` $\pm$ `r round2(sd(dd$EffTelomere))`) and (`r round2(mean(dd$EffGAPDH))` $\pm$ `r round2(sd(dd$EffGAPDH))`) respectively.
-Inter-plate repeatability of RTL, based on `r nrow(subset(ddL,Group == 'Within samples'))` samples measured at least twice at different timepoints, was `r round2(rptall$R)` (CI = `r round2(rptall$CI_emp[1])`, `r round2(rptall$CI_emp[2])`).
+Inter-plate repeatability of RTL, based on `r nrow(subset(ddL,Group == 'Within samples'))` samples measured at least twice at different time points, was `r round2(rptall$R)` (CI = `r round2(rptall$CI_emp[1])`, `r round2(rptall$CI_emp[2])`).
 
@@
 We first tested how RTL was related to age among cohorts using a model selection approach.
-The top model contained a loglinear relationship between RTL and age, as well as a log age x cohort interaction (Table 1A).
-The second-best model contained log age and cohort ID, with no interaction term, although this model was substantially poorer in terms of model fit ($\Delta$AICc > 10).
-All other models fitted the data much less well ($\Delta$AICc > 35; Table 1).
+The top model contained a loglinear relationship between RTL and age (Table 1A).
+All other models fitted the data much less well ($\Delta$AICc > 15; Table 1).
 The log-linear relationship between RTL and age could be seen clearly in the raw data; RTL decreased with age (estimate = `r round3(fixef(RTL_LogAge)['LogAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_LogAge)['LogAge',],lessthan = F),collapse = ', ')`), with the greatest decrease occurring in the first year of life (Fig.
@@
 There was a negative relationship between RTL and age in 21 of the `r length(table(dd$Cohort))` cohorts, but the slope of the relationship between log age and RTL varied substantially among cohorts (Fig.1C).
-In the one year in which RTL increased with age (2013), 17 of the 18 birds sampled were fledglings or subadults, suggesting that the observed pattern was an artifact of the sampling in this season (i.e.
+To test whether this was significant we fitted a model including log age x cohort interaction term, and found that this was a substantially better fit than a model only including main effects ($\Delta$AICc = `r round2(AICc(RTL_LogAge) - AICc(RTL_LogAgeXcohort))` ).
+In the one year in which RTL increased with age (2013), 17 of the 18 birds sampled were fledglings or subadults, suggesting that the observed pattern was an artefact of the sampling in this season (i.e.
 a lack of variation in age among sampled birds) rather than a real relationship.
@@
 A within-individual analysis of RTL and age revealed that the top model explaining RTL contained $\Delta$log age, which reflects within-individual changes in log-transformed age (Table 1B).
-Models including cohort ID and cohort x age interactions were substantially poorer fits than models only containing age (Table 1B).
+Models including cohort ID were substantially poorer fits than a model only containing age (Table 1B).
 RTL decreased with $\Delta$log age (estimate = `r round3(fixef(RTL_DeltaLogAge)['DeltaLogAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_DeltaLogAge)['DeltaLogAge',],lessthan = F),collapse = ', ')`), confirming that within-individual telomere shortening occurs across the Seychelles warbler dataset.
@@
 2B).
-To better understand how longitudinal telomere dynamics vary with age, we exmained patterns of short-term telomere change, including only pairs of samples taken within two-years of each other.
-We found that tongitudinal telomere length increased with log age both when considered as the absolute degree of change (LMM with $\Delta$RTL~*individual*~ as response; estimate = `r round3(fixef(DeltaRTL_LogAge)['LogAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(DeltaRTL_LogAge)['LogAge',],lessthan = F),collapse = ', ')`; Fig.
+To better understand how longitudinal telomere dynamics vary with age, we examined patterns of short-term telomere change, including only pairs of samples taken within two-years of each other.
+We found that longitudinal telomere length increased with log age both when considered as the absolute degree of change (LMM with $\Delta$RTL~*individual*~ as response; estimate = `r round3(fixef(DeltaRTL_LogAge)['LogAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(DeltaRTL_LogAge)['LogAge',],lessthan = F),collapse = ', ')`; Fig.
 2C), and the likelihood of lengthening (GLMM with lengthened yes/no as binomial response; estimate = `r round3(fixef(DeltaRTLF_LogAge)['LogAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(DeltaRTLF_LogAge)['LogAge',],lessthan = F),collapse = ', ')`; Fig.
 2D).
-Increases in telomere length were most likely to be obsevrved shortly after the juvenile period, at around four years of age, and later in life (although sample sizes for older birds are much smaller; Fig.
+Increases in telomere length were most likely to be observed shortly after the juvenile period, at around four years of age, and later in life (although sample sizes for older birds are much smaller; Fig.
 2C,D).
@@
 The most commonly invoked explanation for increases in telomere length is measurement error, which can be a particular problem in qPCR-based telomere studies [@Nussey2014; @Steenstrup2013; @Verhulst2015].
-However, recent modelling work suggests that longitudinal telomere dynamics in humans are indeed consistent with apparent lenghtening, and that dismissing apparent telomere lengthening as solely measurment error is "too strong" without additional data [@Bateson2016a].
+However, recent modelling work suggests that longitudinal telomere dynamics in humans are indeed consistent with apparent lengthening, and that dismissing apparent telomere lengthening as solely measurement error is "too strong" without additional data [@Bateson2016a].
 Here, we explicitly compare intra-individual variation among samples to variation among sample replicates, on a large scale.
@@
 Although a few studies have shown that temporal variation in telomere dynamics occurs in natural populations, these have been limited in the number of seasons they cover [@Mizutani2013; @Watson2015; @Fairlie2016].
-TThe long-term Seychelles warbler dataset has allowed us to show that temporal variation in telomere dynamics can occur over substantial time periods.
+The long-term Seychelles warbler dataset has allowed us to show that temporal variation in telomere dynamics can occur over substantial time periods.
 Our data suggest that conditions during the hatch year are a very important factor in shaping telomere dynamics, and that age-related declines in telomere length vary among cohorts.

### 2017-02-10 Scripts/SW juvenile telomeres IPC models.r

@@ -9,9 +9,12 @@
 RTL_Age <- lmer(RTL ~ Cohort + Agemonths + LogAge + FAge + Age2 + (1|BirdID) + (1|CatchYear),
                 data=dd,
                 REML = FALSE)
-RTL_LogAge <- lmer(RTL ~ LogAge + (1|BirdID) + (1|CatchYear),
+RTL_LogAge <- lmer(RTL ~ LogAge + Cohort + (1|BirdID) + (1|CatchYear),
                 data=dd,
                 REML = FALSE)
+RTL_LogAgeXcohort <- lmer(RTL ~ LogAge*Cohort + (1|BirdID) + (1|CatchYear),
+                   data=dd,
+                   REML = FALSE)
 RTL_DeltaAge <- lmer(RTL ~ MeanAge + Cohort + DeltaAge + DeltaLogAge + DeltaAge2 + (1|BirdID) + (1|CatchYear),
                 data=dd3,
                 REML = FALSE)

### 2017-02-15 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Edits

@@
 **Table 1** Telomere dynamics and age in Seychelles warbler cohorts.
-Linear mixed models were created with RTL as the response variable, and different measures of age, cohort ID, and cohort x age interactions as explanatory variables (see methods for details).
+Linear mixed models were created with RTL as the response variable, and different measures of age, along with cohort ID, were included as explanatory variables (see methods for details).
 Models are ranked by AICc, with best models at the top of the table.
@@
 **D** RTL in relation to and $\Delta$Log age (i.e.
-within indiviual variation in log age).
+within individual variation in log age).
 
@@
 Black line and shaded area represent fitted values and 05% confidence limits from a linear regression of RTL and log-transformed age.
-**D** Telomere lengthening and shortening in relation to age.
+**D** Probability of telomere lengthening occurring in relation to age.
 Points at zero and one represent pairs of samples where RTL has decreased and increased, respectively, with point size scaled by the number of overlapping values.
-The grey line represents the proportion of samples in which i ncreases in RTL where observed at each age.
+The black line represents the proportion of samples in which increases in RTL where observed at each age category.
 
@@
         axis.line.y = element_line(colour = 'black')) +
-  geom_line(col="darkgrey")+
+  geom_line(col="black")+
   geom_count(data = temp,aes(y = DeltaRTLF,x = Agemonths),col = "grey")+
@@
 ```
-
-ggplot(temp,aes(x = TimeDiff,y = DeltaRTL))+
-geom_point()+
-geom_smooth()
 

### 2017-02-21 Manuscript/SW_early_life_telomeres.Rmd
Edits

@@
 Telomere length and the slope of the relationship with age varied markedly among cohorts.
-Telomere length was positively associated with island-wide temporal insect abundance - a key resource for the insectivorous Seychelles warbler - suggesting that the costs associated with living in harsher environments can be studied by investigating telomere dynamics.
+Telomere length was positively associated with temporal variation in island-wide insect abundance - a key resource for the insectivorous Seychelles warbler - suggesting that the costs associated with living in harsher environments can be studied by investigating telomere dynamics.
 We also found evidence for sex-specific relationships between telomeres and tarsus length, potentially reflecting differential costs of growth.
@@
 ###Introduction
-A major aim of ecologists and evolutionary biologists is to quantify and understand why individuals vary in their response to different environmental factors.
+A major aim of ecologists and evolutionary biologists is to understand why individuals vary in their response to different environmental factors.
 Identifying this variation in environmental costs is central to understanding variation in fitness [@Lindstrom1999], and thus for understanding population and community dynamics [@Bolnick2011].

### 2017-03-16 Manuscript/SW_early_life_telomeres.Rmd
Home edits

@@
 
-Lewis G. Spurgin^1^, Kat Bebbington^1^, Eleanor A. Fairfield^1^, Martijn Hammers^2^, Jan Komdeur^2^, Terry Burke^3^, Hannah, L. Dugdale^2,4^, and David S. Richardson^1,5,^.
+Lewis G. Spurgin^1^, Kat Bebbington^1^, Eleanor A. Fairfield^1^, Martijn Hammers^2^, Jan Komdeur^2^, Terry Burke^3^, Hannah L. Dugdale^2,4^ and David S. Richardson^1,5,^.
 
@@
 Telomeres, the protective caps at the ends of chromosomes, shorten in response to oxidative stress, and telomere shortening is correlated with reduced survival and lifespan.
-Investigating telomere dynamics may help us quantify individual variation in the costs experienced as a result of different social and ecological environmental factors, and enhance our understanding of the dynamics of natural populations.
+Investigating telomere dynamics may help us quantify individual variation in the costs experienced from social and ecological environmental factors, and enhance our understanding of the dynamics of natural populations.
 
@@
 Here we study spatio-temporal variation in lifelong telomere dynamics in the Seychelles warbler, *Acrocephalus sechellensis*.
-We combine long-term life-history and ecological data with a large longitudinal telomere dataset, consisting of `r nrow(dd)` samples from `r length(table(dd$Cohort))` cohorts born between `r min(dd$LayYear)` and `r max(dd$LayYear)`.
+We combine long-term life-history and ecological data with a large longitudinal dataset of mean telomere lengths, consisting of `r nrow(dd)` samples from `r length(table(dd$Cohort))` cohorts born between `r min(dd$LayYear)` and `r max(dd$LayYear)`.
 We provide a detailed analysis of how telomere dynamics vary over individual lifespans and cohorts, and with spatio-temporal variation in the social and ecological environment.
@@
 4.
-Telomere length and the slope of the relationship with age varied markedly among cohorts.
+Telomere length and the slope of its relationship with age varied markedly among cohorts.
 Telomere length was positively associated with temporal variation in island-wide insect abundance - a key resource for the insectivorous Seychelles warbler - suggesting that the costs associated with living in harsher environments can be studied by investigating telomere dynamics.
@@
 5.
-Our long-term data show that in a natural population, telomere dynamics vary in a highly complex manner over individual lifespans, and across space and time.
-Variance in telomere dynamics among individuals is the product of wide array of genetic, parental and environmental factors.
+Our long-term data show that in a natural population, telomere dynamics vary in a complex manner over individual lifespans, and across space and time.
+Variance in telomere dynamics among individuals is the product of a wide array of genetic, parental and environmental factors.
 Explaining this variation more fully will require the integration of comprehensive long-term ecological and genetic data from multiple populations and species.
 
-**Keywords:** Biomarkers; Intra- and inter-individual variation; Life-history; Telomere; Seychelles warbler; Senescence
+**Keywords:** Biomarkers; Intra- and inter-individual variation; Life history; Telomere; Seychelles warbler; Senescence
 
@@
 Telomeres have been proposed to be a potential biomarker of such costs [@Monaghan2014].
-Telomeres are repetitive DNA sequences on the ends of linear chromosomes that protect against DNA damage.
-Telomeres generally shorten with age [@Monaghan2006], and there is evidence from a range of taxa that the greatest rate of telomere shortening occurs in early life [e.g.
+Telomeres are repetitive DNA sequences at the ends of linear chromosomes that protect against DNA damage.
+Telomeres generally shorten with age [@Monaghan2006], and there is evidence from a range of taxa that telomere shortening is fastest in early life [e.g.
 @Frenck1998; @Haussmann2003].
@@
 There is evidence from humans, and from captive and wild animal populations, that telomere shortening is influenced by the conditions experienced during both early life and adulthood  [@Price2013; @Monaghan2014; @Nettle2015; @Reichert2015].
-Importantly, the extent of telomere shortening is directly linked to senescence and survival.
-When telomeres become critically short cells senesce [@Campisi2003], and the accumulation of these cells has been suggested to result in organismal senescence and death [@Wong2003].
+Importantly, the extent of telomere shortening is linked to senescence and survival.
+When telomeres become critically short, cells senesce [@Campisi2003], and the accumulation of these cells has been suggested to result in organismal senescence and death [@Wong2003].
 
-The association between senescence and telomere length has inspired a great deal of recent research into telomere evolutionary ecology and relationships between telomere dynamics and survival or lifespan have been documented in several wild organisms [@Horn2010; @Haussmann2010; @Barrett2013a; @Simons2015].
-As yet there is little direct evidence that the relationship between telomere dynamics and survival is causal [@Simons2015].
-However, there is mounting evidence that telomeres can act as biomarkers of individual condition and ageing in wild populations, providing a measure of the ecological stress that an individual has been experienced; a signature that can otherwise be difficult to detect [e.g.
+The association between senescence and telomere length has inspired a great deal of recent research into telomere evolutionary ecology and relationships between telomere dynamics and survival or lifespan have been documented in wild population of several species [@Horn2010; @Haussmann2010; @Barrett2013a; @Simons2015].
+As yet, there is little direct evidence that the relationship between telomere dynamics and survival is causal [@Simons2015].
+However, there is mounting evidence that telomeres can act as biomarkers of individual condition and ageing in wild populations, providing a measure of the ecological stress that an individual has experienced, a signature that can otherwise be difficult to detect [e.g.
 @Schultner2014; @Asghar2015; @Bebbington2016].
-There is also evidence that telomere length, measured longitudinally from individuals, can increase as well as decrease [@Muezzinler2013; @Simons2014; @Bateson2016a], which has important ramifications for our understanding of how telomeres reflect costs.
-However, such increases in telomere length are often attributed to measurement error [@Steenstrup2013; but see @Bateson2016a], and as such its ecological significance is unknown.
+There is also evidence that telomere length, measured longitudinally in individuals, can increase as well as decrease [@Muezzinler2013; @Simons2014; @Bateson2016a], which has important ramifications for our understanding of how telomeres reflect costs.
+However, such increases in telomere length are often attributed to measurement error [@Steenstrup2013; but see @Bateson2016a], and as such their ecological significance is unknown.
 
@@
 Variation in oxidative stress experienced by individuals is associated with territory quality [@VandeCrommenacker2011a].
-However, the evidence that individual survival and lifespan is associated with spatial variation in early life territory quality or local density is equivocal and confounded by variation in subsequent life-history parameters [@Brouwer2006; @Hammers2013].
+However, the evidence that individual survival and lifespan is associated with spatial variation in early-life territory quality or local density is equivocal and confounded by variation in subsequent life-history parameters [@Brouwer2006; @Hammers2013].
 There is also variation in the social environment that individual Seychelles warblers experience.
@@
 Importantly, we have an established protocol for assessing telomere length in the Seychelles warbler [@Barrett2012; @Bebbington2016].
-Furthermore, telomere dynamics predict survival independently of age [@Barrett2013a] and are negatively associated with inbreeding [@Bebbington2016], suggesting that individual variation in telomere length is ecologically relevant in this species.
+Furthermore, telomere dynamics predict survival independently of age [@Barrett2013a] and telomere length is negatively associated with inbreeding [@Bebbington2016], suggesting that individual variation in telomere length is ecologically relevant in this species.
 Thus, we have an excellent system in which to determine the impact of different social and environmental conditions experienced by individuals, and to assess how these costs vary over space and time.
@@
 We first study how telomere length and rates of shortening are related to age and sex across all life stages, and how this relationship varies among cohorts, in order to gain an in-depth understanding of the temporal dynamics of telomere changes.
-We then examine, within individuals, how telomere length changes with age, and statistically test whether observed increases in telomere length across longitudinal samples are larger than expected based on measurement error.
-Finally, we test how a wide range of social and environmental variables are related to telomere length and shortening in order gain a fuller understanding of the forces driving telomere dynamics in natural populations.
+We then examine, within individuals, how telomere length changes with age, and statistically test whether observed increases in telomere length across temporally longitudinal samples within individuals are larger than can be accounted for by measurement error.
+Finally, we test how telomere length and shortening are related to a wide range of social and environmental variables in order to gain a fuller understanding of the forces driving telomere dynamics in natural populations.
 
@@
 The population of *ca* 320 adult birds on Cousin Island (04'20'S, 55'40'E) has been intensively studied since 1986 [@Komdeur1992; @Richardson2003b; @Spurgin2014a].
-).
-This species' main breeding season runs from June-September (though a small number of territories do sometimes try to breed between January-March) when the breeding females on many of the *ca* 110 territories will attempt to breed, laying one or, rarely, two or three eggs [@Komdeur1991].
+This species' main breeding season runs from June-September (though a small proportion of pairs also breed between January–March), when the breeding females on many of the *ca* 110 territories will attempt to breed, laying one or, rarely, two or three eggs [@Komdeur1991].
 Breeding attempts are often unsuccessful, and as a result of this low reproductive output, and higher mortality in first-year birds [39% in first-year birds versus 16% in adults; @Brouwer2006], cohort sizes in the Seychelles warbler are typically small (< 50).
-The 22 birth year cohorts used in this study cover 1993 to 2014 – the time period during which our data and sampling is most complete.
+The 22 birth year cohorts used in this study cover 1993 to 2014 – the time period during which our data and sampling are most complete.
 
@@
 
-Seychelles warblers are highly territorial and all territories were mapped during each main breeding season using detailed observational data of foraging and territorial defence behaviour [@Richardson2003b].
+Seychelles warblers are highly territorial and all territories were mapped during each main breeding season using detailed observational data on foraging and territorial defence behaviour [@Richardson2003b].
 Territory quality is calculated based on territory size, foliage cover and insect abundance [@Komdeur1992].
@@
 Prior to telomere analysis, DNA concentration and purity were quantified using a NanoDrop 8000 Spectrophotometer (ThermoScientific).
-The following thresholds were applied before samples were included for further analysis: i) DNA concentration must be at least 15 ng $\mu$l^-1^ (based on a mean of three measurements), ii) the 260/280 absorbance ratio has to be between 1.8 and 2 for acceptable DNA purity and, iii) the 260/230 absorbance ratio must be higher than 1.8.
+The following thresholds were applied before samples were included for further analysis: i) DNA concentration must be at least 15 ng $\mu$l^-1^ (based on a mean of three measurements), ii) the 260/280 absorbance ratio has to be between 1.8 and 2.0 for acceptable DNA purity, and iii) the 260/230 absorbance ratio must be higher than 1.8.
 DNA integrity was further validated by visualization with ethidium bromide after electrophoresis on a 1.2% agarose gel, and all samples with evidence of DNA degradation were re-extracted or excluded.
@@
 When examining the distribution of longitudinal telomere changes we observed some increases in telomere length with age in individuals.
-We then used repeat measurements to test whether these increases could be explained by measurement error.
-We calculated the change in RTL between pairs of repeat measurements within the same samples (hereafter $\Delta$RTL~*sample*~; N = `r nrow(subset(ddL,Group == 'Within samples'))` measurements from `r length(unique(subset(ddL,Group == 'Within samples')$BirdID))` samples) in exactly the same way as for across samples (hereafter $\Delta$RTL~*individual*~), using completely separate reactions run on separate plates.
+We therefore repeated the qPCR on each sample  using completely separate reactions run on separate plates and used these repeat measurements to test whether these increases could be explained by measurement error.
+We calculated the change in RTL between pairs of repeat measurements within the same samples (hereafter $\Delta$RTL~*sample*~; N = `r nrow(subset(ddL,Group == 'Within samples'))` pairs of measurements from `r length(unique(subset(ddL,Group == 'Within samples')$BirdID))` birds) in exactly the same way as for across samples (hereafter $\Delta$RTL~*individual*~).
 To test whether greater changes in RTL were observed among individuals compared to among repeat samples, we compared the variance in $\Delta$RTL~*sample*~ and $\Delta$RTL~*individual*~ using a Levene's test.
-Then, to separately test whether the extent of telomere increases and decreases within individuals was greater than expected by measurement error, we split $\Delta$RTL measurements into groups in which RTL decreased ($\Delta$RTL < 0) and increased $\Delta$RTL ($\Delta$RTL > 0), and tested whether $\Delta$RTL~*inidividual*~ values were significantly different from $\Delta$RTL~*sample*~ values, using Wilcoxon tests.
+Then, to separately test whether the extent of telomere increases and decreases within individuals were greater than expected by measurement error, we split $\Delta$RTL measurements into groups in which RTL decreased ($\Delta$RTL < 0) and increased $\Delta$RTL ($\Delta$RTL > 0), and tested whether $\Delta$RTL~*individual*~ values were significantly different from $\Delta$RTL~*sample*~ values, using Wilcoxon tests.
 
-We then used LMMs to test how variation in environmental and social conditions influenced telomere length within cohorts.
-We created a full model with RTL as a response variable, alongside the following explanatory variables: log age (based on the RTL and age analysis; see results), tarsus length, body mass, sex, insect abundance, territory quality, island-wide population density (an annual measure estimated from the summer breeding census), territory group size, and the number of helping subordinate birds present in the territory.
+We also tested whether consistent telomere lengthening acrossed out dataset using a modified version of the approach recently developed by Simons *et al.* [-@Simons2014].
+Briefly, this approach utilises samples with at least three telomere measurements to compare residual variance in telomere change over time with the overall change in telomere length between the first and least telomere measurements [@Simons2014].
+If, in samples that exhibit lengthening, the overall increase in telomere length exceeds the residual variance, then telomere lengthening is predicted to be occurring [@Simons2014].
+If, on the other hand, increases in telomere length are due to measurement error, within-individual residual variance in telomere length is expected to be similar to overall observed increases in telomere length.
+
+We used LMMs to explore how variation in environmental and social conditions influenced telomere length and dynamics within cohorts.
+We first created a full model with RTL as a response variable, alongside the following explanatory variables: log age (based on the RTL and age analysis; see results), tarsus length, body mass, sex, insect abundance, territory quality, island-wide population density (an annual measure estimated from the summer breeding census), territory group size, and the number of helping subordinate birds present in the territory.
 The random effects structure was informed by the analysis of telomere dynamics and age (see results): we included individual ID, cohort ID, and a random slope of log age among cohorts (to allow the effect of age on RTL to vary among cohorts).
@@
 As a complementary approach, we also performed model averaging, using the MuMIn package in R [@Barton2012].
-Model selection was performed using the full model described above and a top model set defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011].
+Model selection was performed using the full model described above.
+A top model set was then defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011].
 We report model-averaged coefficients, confidence intervals and 'relative importance', which reflects the relative weights of each predictor variable across the top model set.
@@
 We measured telomere lengths using a total of `r length(unique(dd$BloodID))` unique samples from juvenile and adult Seychelles warblers from `r length(table(dd$Cohort))` cohorts born between `r min(dd$LayYear)` and `r max(dd$LayYear)`.
-Efficiencies (mean $\pm$ s.d.) for our telomere and GAPDH reactions were (`r round2(mean(dd$EffTelomere))` $\pm$ `r round2(sd(dd$EffTelomere))`) and (`r round2(mean(dd$EffGAPDH))` $\pm$ `r round2(sd(dd$EffGAPDH))`) respectively.
+Efficiencies (mean $\pm$ s.d.) for our telomere and GAPDH reactions were `r round2(mean(dd$EffTelomere))` $\pm$ `r round2(sd(dd$EffTelomere))` and `r round2(mean(dd$EffGAPDH))` $\pm$ `r round2(sd(dd$EffGAPDH))` respectively.
 Inter-plate repeatability of RTL, based on `r nrow(subset(ddL,Group == 'Within samples'))` samples measured at least twice at different time points, was `r round2(rptall$R)` (CI = `r round2(rptall$CI_emp[1])`, `r round2(rptall$CI_emp[2])`).
@@
 We first tested how RTL was related to age among cohorts using a model selection approach.
-The top model contained a loglinear relationship between RTL and age (Table 1A).
+The top model contained a log-linear relationship between RTL and age (Table 1A).
 All other models fitted the data much less well ($\Delta$AICc > 15; Table 1).
@@
 1B).
-There was a negative relationship between RTL and age in 21 of the `r length(table(dd$Cohort))` cohorts, but the slope of the relationship between log age and RTL varied substantially among cohorts (Fig.1C).
-To test whether this was significant we fitted a model including log age x cohort interaction term, and found that this was a substantially better fit than a model only including main effects ($\Delta$AICc = `r round2(AICc(RTL_LogAge) - AICc(RTL_LogAgeXcohort))` ).
+There was a negative relationship between RTL and age in 21 of the `r length(table(dd$Cohort))` cohorts, but the slope of the relationship between RTL and log age varied substantially among cohorts (Fig.
+1C).
+To test whether this variation was significant we fitted a model including the log age x cohort interaction term, and found that this was a marginally better fit than a model including only main effects ($\Delta$AICc = `r round2(AICc(RTL_LogAge) - AICc(RTL_LogAgeXcohort))`).
 In the one year in which RTL increased with age (2013), 17 of the 18 birds sampled were fledglings or subadults, suggesting that the observed pattern was an artefact of the sampling in this season (i.e.
-a lack of variation in age among sampled birds) rather than a real relationship.
+a lack of variation in age among sampled birds), rather than a real relationship.
 
@@
 2B).
+
 To better understand how longitudinal telomere dynamics vary with age, we examined patterns of short-term telomere change, including only pairs of samples taken within two-years of each other.
@@
 2C,D).
+
+Using the approach outlined by Simons *et al.* we tested whether, statistically, overall increases in RTL over lifespans could be detected statistically in our dataset.
+We found no evidence that this was the case - overall increases in RTL within individuals did not exceed residual variance.
+In fact, the opposite was the case - residual variance in RTL was significantly greater than observed RTL increases over lifespans (P = 0.02).
+This suggests that increases in RTL within individuals are sporadic, and not consistent over individual lifespans.
 

### 2017-03-16 Scripts/SimonsTest.R

@@ -52,3 +52,57 @@ print(indexTLincreases[outsideconfindex]) #initial row of individual in the data
 
 
 
+
+
+
+
+
+
+
+
+
+
+
+
+
+ss <- table(subset(dd,Ageclass!="CH")$BirdID)
+tokeep <- names(ss[ss>2])
+data <- subset(dd,BirdID %in% tokeep)
+
+#Estimating error using individual regressions, Equation 4 in main manuscript
+
+
+birds <- unique(data$BirdID)
+output <- rep(NA,length(birds))
+deltaTL <- rep(NA,length(birds))
+  
+  #loop the individual regressions for the amount of individuals in the dataset
+
+for(i in 1:length(birds))
+{
+  cd <- subset(data,BirdID == birds[i])
+  fit<-lm(RTL~order(cd$Agemonths),data=cd)  #individual linear regression
+  output[i]<-sum((residuals(fit))^2)/(nrow(cd)-2) #residual sum of squares
+  deltaTL[i] <- cd[order(cd$Agemonths),"RTL"][length(cd[order(cd$Agemonths),"RTL"])] - cd[order(cd$Agemonths),"RTL"][1]
+  }
+sigma1<-mean(c(output)) #average residual sum of squares across the individuals
+
+
+#Estimating error under the assumption that TL cannot increase over time, Equation 5 in main manuscript
+
+#Next we determine which individuals increase in TL between the first and last timepoint
+indexTLincreases=which(deltaTL>0)
+
+#We create a new variable including only the data of individual increases
+TLincreases=deltaTL[indexTLincreases]
+#sigma2
+sigma2=0.5*sum(TLincreases^2)/(length(TLincreases))
+
+
+#compare both estimates of error variance (sigma1 and sigma2), Equation 6 in main manuscript
+vratio=sigma2/sigma1
+pvalue<-pf(vratio,length(TLincreases)-1,dim(data)[1]-1,lower.tail=T)
+print(pvalue)
+
+
+

### 2017-03-17 Manuscript/SW_early_life_telomeres.Rmd
Edits

@@
 
-We also used longitudinal data to determine individual-level consistency in RTL.
-We constructed a LMM with RTL at time *t* + 1 as the response variable, RTL at time *t* and age at time *t* as fixed effects, and individual ID and cohort as random effects.
-To obtain an estimate of explanatory power of the fixed effects, we calculated marginal R^2^ following Nakagawa & Schielzeth [-@Nakagawa2013].
+We used two approaches to determine individual-level consistency in RTL.
+We first calculated individual-level repeatability in RTL by dividing the random avariance explained by individual ID by the total random variance, in a model of that accounted for age and cohort effects.
+Second, we constructed a LMM with RTL at time *t* + 1 as the response variable, RTL at time *t* and age at time *t* as fixed effects, and individual ID and cohort as random effects.
+We estimated the slope of the relationship between within-individual telomere measurements, as well as the variance explained, by calculating the marginal R^2^ [@Nakagawa2013] of the model.
 
@@
 We report model estimates and confidence intervals for all effects included in the full model.
-We also calculated marginal R^2^ and conditional R^2^ [incorporating fixed and random effects, respectively; @Johnson2014] to assess the explanatory power of these models.
+We also calculated marginal R^2^ [incorporating only fixed effects; @Nakagawa2013] and conditional R^2^ [incorporating fixed and random effects; @Johnson2014] to assess the explanatory power of these models.
 As a complementary approach, we also performed model averaging, using the MuMIn package in R [@Barton2012].
@@
 
-There was positive correlation between RTL measured from different samples taken at different time points during an individual’s life (Fig.
-2A), but this was very weak (marginal R^2^ = `r round2(r.squaredGLMM(RTL_RTL1)['R2m'])`), and not significant (estimate = `r round3(fixef(RTL_RTL1)['RTL'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_RTL1)['RTL',],lessthan = F),collapse = ', ')`).
+Individual repeatability in RTL was `r round3(as.data.frame(VarCorr(RTL_LogAge))[1,"vcov"]/sum(as.data.frame(VarCorr(RTL_LogAge))[,"vcov"]))`, meaning that `r paste(round((as.data.frame(VarCorr(RTL_LogAge))[1,"vcov"]/sum(as.data.frame(VarCorr(RTL_LogAge))[,"vcov"]))*100,0))`% of variance in RTL could be explained by within-individual consistency.
+Accordingly, there was a positive correlation between RTL measured from different samples taken at different time points during an individual’s life (Fig.
+2A), but this was very weak (marginal R2 = `r round2(r.squaredGLMM(RTL_RTL1)['R2m'])`), and not significant (estimate = `r round3(fixef(RTL_RTL1)['RTL'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_RTL1)['RTL',],lessthan = F),collapse = ', ')`).
+
 Although both cross-sectional and longitudinal data indicated a general trend of telomere shortening with age, we found that RTL - measured across two samples taken from the same individuals over time - increased with age in `r round((nrow(subset(dd3,DeltaRTL>0))/nrow(dd3))*100,0)`% of our `r nrow(dd3)` $\Delta$RTL~*individual*~ measurements (Fig.
@@
 
-To better understand how longitudinal telomere dynamics vary with age, we examined patterns of short-term telomere change, including only pairs of samples taken within two-years of each other.
+To better understand how longitudinal telomere dynamics vary with age, we examined patterns of short-term telomere change, including only pairs of samples taken within two years of each other.
 We found that longitudinal telomere length increased with log age both when considered as the absolute degree of change (LMM with $\Delta$RTL~*individual*~ as response; estimate = `r round3(fixef(DeltaRTL_LogAge)['LogAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(DeltaRTL_LogAge)['LogAge',],lessthan = F),collapse = ', ')`; Fig.
@@
 3A).
-Tarsus length was negatively related to RTL and males had longer telomeres than females (Fig.
+RTL was negatively related to tarsus length and males had longer telomeres than females (Fig.
 3B), while insect abundance was positively related to RTL (Fig.
@@
 @Frenck1998; @Haussmann2003; @Heidinger2012].
-However, we also found that despite an overall trend for shortening, telomere length both increased and decreased, especially after the juvenile period.
-Importantly, these increases were observed in longitudinal as well as cross-sectional data, indicating that selective disappearance of individuals with shorter telomeres does not explain this pattern.
+However, we also found that, despite an overall trend for shortening, telomere length both increased and decreased, especially after the juvenile period.
+Importantly, these increases were observed in longitudinal as well as cross-sectional data, indicating that selective disappearance of individuals with shorter telomeres is not sufficient to explain this pattern.
 Longitudinal increases in measured telomere length have been observed in humans and wild animals  [@Steenstrup2013;@Kotrschal2007; @Fairlie2016; @Hoelzl2016; @Hoelzl2016a].
 The most commonly invoked explanation for increases in telomere length is measurement error, which can be a particular problem in qPCR-based telomere studies [@Nussey2014; @Steenstrup2013; @Verhulst2015].
-However, recent modelling work suggests that longitudinal telomere dynamics in humans are indeed consistent with apparent lengthening, and that dismissing apparent telomere lengthening as solely measurement error is "too strong" without additional data [@Bateson2016a].
+However, recent modelling work suggests that longitudinal telomere dynamics in humans are indeed consistent with instances of lengthening, and that dismissing apparent telomere lengthening as solely measurement error is "too strong" without additional data [@Bateson2016a].
 Here, we explicitly compare intra-individual variation among samples to variation among sample replicates, on a large scale.
@@
 
-Increases in telomere length were not consistent over individual lifespans, but occurred during specific, periods, against a backdrop of overall lifelong telomere shortening.
+Increases in telomere length were not consistent over individual lifespans, but occurred during specific periods, against a backdrop of overall lifelong telomere shortening.
 This is consistent with recent findings in edible dormice *Glis glis*, in which telomere elongation was observed only later in life [@Hoelzl2016a].
@@
 These findings are in contrast to other avian studies where birds were reared in laboratory [@Heidinger2012] or nest-box [@Boonekamp2014] conditions, in which within-individual telomere length measurements were highly consistent, and individual-level telomere shortening occurred throughout the juvenile period and into adulthood.
-However, the lifelong telomere dynamics found in Seychelles warblers are strikingly similar to those found in Soay sheep [@Fairlie2016], and we anticipate that a similar pattern may be found in other wild populations.
+However, the lifelong telomere dynamics found in Seychelles warblers are strikingly similar to those found in Soay sheep [@Fairlie2016], and we anticipate that a similar pattern may be found in other populations.
 
@@
 
-We also found evidence for sex-specific telomere dynamics: males had longer-telomeres than females.
-Interestingly this sex-difference interacts with tarsus length: telomere length was negatively correlated with tarsus length in both sexes, but this effect was stronger in females than males.
-If the sex-dependent relationship between telomere and tarsus length was due to differential growth alone we would expect the opposite pattern to that observed, as male Seychelles warblers are larger than females (Fig.
+We also found evidence for sex-specific telomere dynamics: males had longer telomeres than females.
+Interestingly this sex difference interacts with tarsus length: telomere length was negatively correlated with tarsus length in both sexes, but this effect was stronger in females than males.
+If the sex-dependent relationship between telomere and tarsus length was due to differential growth alone then we would expect the opposite pattern to that observed, as male Seychelles warblers are larger than females (Fig.
 3B).
 One possibility is that the environment imposes differential costs on males and females: a recent study in captive zebra finches found that manipulation of dietary nutrients had sex-dependent effects on telomere dynamics [@Noguera2015].
-Also worth noting is that the effect of telomere length on survival in strongest in male Seychelles warblers [@Barrett2013a], although the nature of the relationship between sex, telomeres and survival is not yet clear [@Barrett2011]
+Also worth noting is that the effect of telomere length on survival is strongest in male Seychelles warblers [@Barrett2013a], although the nature of the relationship between sex, telomeres and survival is not yet clear [@Barrett2011]
 
@@
 This work was funded by two Natural Environment Research Council (NERC) grants to DSR (NE/F02083X/1 and NE/K005502/1).
-LGS was also funded by a fellowship from the BBSRC, and HLD by a NERC fellowship (NE/I021748/1).
-MH was funded by a VENI fellowship from the Netherlands Organisation for Scientific Research (863.15.020).
+LGS was also funded by a fellowship from the BBSRC (BB/N011759/1), HLD by a NERC fellowship (NE/I021748/1), MH was funded by a VENI fellowship from the Netherlands Organisation for Scientific Research (863.15.020), and TB by a Leverhulme fellowship.
 

### 2017-03-19 Manuscript/Cover letter.Rmd
Home edits

---
output: pdf_document
---

Dr Lewis G. Spurgin
School of Biological Sciences
University of East Anglia
Norwich
Norfolk NR4 7TJ
United Kingdom

Email: L.Spurgin@uea.ac.uk
Web: lewisspurgin.wordpress.com

To the Editor,

Please find enclosed our manuscript *Spatio-temporal variation in lifelong telomere dynamics in a long-term ecological study*, for consideration as an article to be published in *Journal of Animal Ecology*. We use telomeres as biomarkers to understand how individuals vary in their response to changes in the environment, over different life history stages. Specifically, utilising the many benefits of the long-term Seychelles warbler study system, we study telomere dynamics in relation to age and the ecological environment in 22 cohorts from a single wild population.

Two of our main faindings are particularly striking and will, we hope, be of interest to your readers. Firstly, we find that cohort effects are the most important drivers of variation in telomere length in the Seychelles warbler population. This has important ramifications for how studies of natural populations interpret variaiton in teloemre length, as these are often based on a single timepoint. Our data suggest that telomere dynamics at a single point in time may in fact be a snapshot of a temporally dynamic process.

Secondly, using a large number of repeated telomere measurements we show that within-individual increases in telomere length cannot be solely explained by measurement error. This is in accordance with recent theoretical work, and data from humans, suggesting that telomere elongation may be an ecologically important process. We hope that our data will spur further research into when, where any why telomere elongation occurs.

In addition to these two main findings, we provide a broad and comprehensive analysis of telomere dynamics in the long-term Seychelles warbler study. I hope you are in agreement that our thorough analysis, and main findings, are of of broad ecological interest, and suited to the readership of *Journal of Animal Ecology*. I look forward to hearing from you.

Yours faithfully,

Lewis Spurgin (on behalf of the authors)

### 2017-03-19 Manuscript/SW_early_life_telomeres.Rmd
Home edits

@@
 Each time a bird is caught on Cousin a range of morphometric measurements are taken, including body mass and tarsus length (to the nearest 0.1g and 0.1mm, respectively).
-A blood sample (*ca* 25 l) is taken via brachial venipuncture, and stored at room temperature in 1 ml of absolute ethanol in a 1.5 ml screw-cap microfuge tube.
+A blood sample (*ca* 25 $\mu$l) is taken via brachial venipuncture, and stored at room temperature in 1 ml of absolute ethanol in a 1.5 ml screw-cap microfuge tube.
 
@@
 We used two approaches to determine individual-level consistency in RTL.
-We first calculated individual-level repeatability in RTL by dividing the random avariance explained by individual ID by the total random variance, in a model of that accounted for age and cohort effects.
+We first calculated individual-level repeatability in RTL by dividing the random variance explained by individual ID by the total random variance, in a model of that accounted for age and cohort effects.
 Second, we constructed a LMM with RTL at time *t* + 1 as the response variable, RTL at time *t* and age at time *t* as fixed effects, and individual ID and cohort as random effects.
@@
 
-We also tested whether consistent telomere lengthening acrossed out dataset using a modified version of the approach recently developed by Simons *et al.* [-@Simons2014].
+We also tested whether consistent telomere lengthening across our dataset using a modified version of the approach developed by Simons *et al.* [-@Simons2014].
 Briefly, this approach utilises samples with at least three telomere measurements to compare residual variance in telomere change over time with the overall change in telomere length between the first and least telomere measurements [@Simons2014].
-If, in samples that exhibit lengthening, the overall increase in telomere length exceeds the residual variance, then telomere lengthening is predicted to be occurring [@Simons2014].
+If, in samples that increase in length, the overall increase in telomere length exceeds the residual variance, then telomere lengthening cannot be explained by error [@Simons2014].
 If, on the other hand, increases in telomere length are due to measurement error, within-individual residual variance in telomere length is expected to be similar to overall observed increases in telomere length.
@@
 1B).
-There was a negative relationship between RTL and age in 21 of the `r length(table(dd$Cohort))` cohorts, but the slope of the relationship between RTL and log age varied substantially among cohorts (Fig.
+There was a negative relationship between RTL and log age in 21 of the `r length(table(dd$Cohort))` cohorts, but the slope the relationship varied substantially among cohorts (Fig.
 1C).
@@
 
-Using the approach outlined by Simons *et al.* we tested whether, statistically, overall increases in RTL over lifespans could be detected statistically in our dataset.
-We found no evidence that this was the case - overall increases in RTL within individuals did not exceed residual variance.
-In fact, the opposite was the case - residual variance in RTL was significantly greater than observed RTL increases over lifespans (P = 0.02).
+Using the approach outlined by Simons *et al.* we tested whether overall increases in RTL over lifespans could be detected statistically in our dataset.
+We found no evidence that this was the case: overall increases in RTL within individuals did not exceed residual variance; in fact, residual variance in RTL was significantly greater than observed RTL increases over lifespans (P = 0.02).
 This suggests that increases in RTL within individuals are sporadic, and not consistent over individual lifespans.

### 2017-03-19 Manuscript/Cover letter.Rmd
Home edits

@@
 ---
-output: pdf_document
+output: 
+  pdf_document:
+header-includes: \usepackage{graphicx}
 ---
 
-Dr Lewis G. Spurgin
-School of Biological Sciences
-University of East Anglia
-Norwich
-Norfolk NR4 7TJ
+\begin{flushright}
+
+\includegraphics{UEAlogo}
+
+School of Biological Sciences\\*
+University of East Anglia\\*
+Norwich\\*
+Norfolk NR4 7TJ\\* 
 United Kingdom
 
-Email: L.Spurgin@uea.ac.uk
-Web: lewisspurgin.wordpress.com
+L.Spurgin@uea.ac.uk\\*
+lewisspurgin.wordpress.com
+
+\end{flushright}
+
+\bigskip
+
+**19th March 2017**
+
+\bigskip
 
@@
 
-Two of our main faindings are particularly striking and will, we hope, be of interest to your readers.
+Two of our main findings are particularly striking and will, we hope, be of interest to your readers.
 Firstly, we find that cohort effects are the most important drivers of variation in telomere length in the Seychelles warbler population.
-This has important ramifications for how studies of natural populations interpret variaiton in teloemre length, as these are often based on a single timepoint.
+This has important ramifications for how studies of natural populations interpret variation in telomere length, as these are often based on a single time point.
 Our data suggest that telomere dynamics at a single point in time may in fact be a snapshot of a temporally dynamic process.
@@
 
+\bigskip
+
 Yours faithfully,
 
+\bigskip
+
 Lewis Spurgin (on behalf of the authors)

### 2017-03-20 Manuscript/SW_early_life_telomeres_FULL.Rmd
MAJOR EDITS
Changed to pdf format

--- 2017-03-19 Manuscript/SW_early_life_telomeres.Rmd
@@
 output:
-  word_document:
-    reference_docx: myref.docx
+  pdf_document:
 bibliography: MyBib.bib
-csl: Evolution.csl
-
+csl: JAE.csl
+header-includes: \usepackage{graphicx}
+    \usepackage{setspace}
+    \doublespacing
+    \usepackage{lineno}
+    \linenumbers
 ---
@@
                warning=FALSE,
-               message=FALSE)
+               message=FALSE,
+               fig.width=8,
+               fig.height=8)
 
@@
 
-**Spatio-temporal variation in lifelong telomere dynamics in a long-term ecological study**
+#Spatio-temporal variation in lifelong telomere dynamics in a long-term ecological study
 
@@
 
-###Abstract
+\pagebreak
+
+##Abstract
 1.
@@
 
-**Data archival location:** This manuscript was written in R Markdown (http://rmarkdown.rstudio.com/).
-All data and scripts required to reproduce the manuscript, figures and analyses will be made available on GitHub.
+\pagebreak
 
-###Introduction
+##Introduction
 A major aim of ecologists and evolutionary biologists is to understand why individuals vary in their response to different environmental factors.
@@
 
-###Methods
+##Methods
 
-#####Study species and sampling
+*Study species and sampling*
 The Seychelles warbler is a small (~15 g), insectivorous passerine bird with a mean life expectancy of 5.5 years at fledging [@Hammers2013].
@@
 
-#####Molecular methods
+*Molecular methods*
 For each sample, genomic DNA was extracted from a ~2 mm^2^ flake of preserved blood using the DNeasy Blood and Tissue Kit (Qiagen), following the manufacturer's protocol, with the modification of overnight lysis at 37^o^C and a final DNA elution volume of 80 $\mu$l.
@@
 
-#####Statistical analyses
+*Statistical analyses*
 We performed all statistical analyses using R version 3.2.2 [@RDevelopmentCoreTeam2011].
@@
 
-###Results
+##Results
 We measured telomere lengths using a total of `r length(unique(dd$BloodID))` unique samples from juvenile and adult Seychelles warblers from `r length(table(dd$Cohort))` cohorts born between `r min(dd$LayYear)` and `r max(dd$LayYear)`.
@@
 
-#####Telomere dynamics and age among cohorts
+*Telomere dynamics and age among cohorts*
 We first tested how RTL was related to age among cohorts using a model selection approach.
@@
 
-#####Telomere dynamics and the environment
+*Telomere dynamics and the environment*
 In addition to age, RTL was associated with tarsus length, sex and insect abundance (Fig.
@@
 
-###Discussion
+##Discussion
 Here we use a long-term, multi-cohort dataset to assess lifelong telomere dynamics and the relationship between these and spatio-temporal variation in the ecological environment in a contained population of Seychelles warblers.
@@
 
-###Acknowledgements
+##Acknowledgements
 We thank Nature Seychelles for facilitating the long-term Seychelles warbler project.
@@
 
-###References
+##Data accessibility
+This manuscript was written in R Markdown (http://rmarkdown.rstudio.com/). All data and scripts required to reproduce the manuscript, figures and analyses will be made available on GitHub.
+
+##Author contributions
+DSR, HLD, JK and TB manage the long-term Seychelles warbler project.
+DSR conceived and obtained funding for the telomere research.
+EAF and KB performed the molecular work.
+LGS processed the telomere data, with input from EAF, KB, MH, HLD and DSR.
+LGS analysed the data and wrote the manuscript, with input from DSR and all authors.
+
+##References
+
+<div id="refs"></div>
+
+\pagebreak
+
+**Table 1** Telomere dynamics and age in Seychelles warbler cohorts.
+Linear mixed models were created with RTL as the response variable, and different measures of age, along with cohort ID, were included as explanatory variables (see methods for details).
+Models are ranked by AICc, with best models at the top of the table.
+
+```{r Table 1,include = T,echo = F,warning = F}
+RTL_Age_Dredge <- dredge(RTL_Age,
+                         subset = !(LogAge & FAge) & 
+                           !(LogAge & Agemonths) &
+                           !(LogAge & Age2) &
+                           !(Agemonths & FAge) &
+                           !(FAge & Age2) &
+                           !(Age2 & !(Agemonths)))
+
+t1 <- as.matrix(RTL_Age_Dredge)
+
+#Sort out row/col names
+nams <- colnames(t1)
+
+nams[nams == 'Agemonths'] <- 'Age (linear)'
+nams[nams == 'MeanAge'] <- 'Age (mean)'
+nams[nams == 'LogAge'] <- 'Age (log)'
+nams[nams == 'FAge'] <- 'Age (factor)'
+nams[nams == 'Cohort:FAge'] <- 'Age (factor)*cohort'
+nams[nams == 'Agemonths:Cohort'] <- 'Age (linear)*cohort'
+nams[nams == 'Cohort:LogAge'] <- 'Age (log)*cohort'
+nams[nams == 'Age2:Cohort'] <- 'Age (quadratic)*cohort'
+nams[nams == 'Age2'] <- 'Age (quadratic)'
+
+colnames(t1) <- nams
+
+table1A <- data.frame(Model = apply(t1,1,getterms,c(2:6)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
+colnames(table1A) <- c('Model','df','AICc', 'Delta AICc','Weight')
+rownames(table1A) <- NULL
+table1A$Model[table1A$Model == ''] <- 'Null model'
+table1A[,3:5] <- lapply(table1A[,3:5],as.numeric)
+table1A[,3:5] <- round(table1A[,3:5],3)
+
+RTL_DeltaAge_Dredge <- dredge(RTL_DeltaAge,
+                         subset = MeanAge &
+                           !(DeltaLogAge & DeltaAge) &
+                           !(DeltaLogAge & DeltaAge2) &
+                           !(!(DeltaAge) & DeltaAge2))
+
+t1 <- as.matrix(RTL_DeltaAge_Dredge)
+
+#Sort out row/col names
+nams <- colnames(t1)
+
+nams[nams == 'DeltaAge'] <- 'Delta age (linear)'
+nams[nams == 'DeltaLogAge'] <- 'Delta age (log)'
+nams[nams == 'Cohort:DeltaAge'] <- 'Delta age (continuous)*cohort'
+nams[nams == 'Cohort:DeltaLogAge'] <- 'Delta age (log)*cohort'
+nams[nams == 'Cohort:DeltaAge2'] <- 'Delta age (quadratic)*cohort'
+nams[nams == 'DeltaAge2'] <- 'Delta age (quadratic)'
+
+colnames(t1) <- nams
+
+table1B <- data.frame(Model = apply(t1,1,getterms,c(2:6)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
+colnames(table1B) <- c('Model','df','AICc', 'Delta AICc','Weight')
+rownames(table1B) <- NULL
+table1B$Model[table1B$Model == ''] <- 'Null model'
+table1B[,3:5] <- lapply(table1B[,3:5],as.numeric)
+table1B[,3:5] <- round(table1B[,3:5],3)
+
+#Sort out rounding and exlcude NAs from blank spaces. Clunky but works
+table1 <- rbind(c('A',rep('-',4)),table1A,c('B',rep('-',4)),table1B)
+
+kable(table1,digits = 3)
+```
+
+\pagebreak
+
+##Figure Legends
+
+**Figure 1** Telomere dynamics in relation to age in Seychelles warbler cohorts.
+**A** RTL and age across all individuals.
+Points and  connecting thin grey lines represent individual samples and birds, respectively.
+The thick line and shaded area represent the fitted values and 95% confidence limits of a linear regression of RTL and log-transformed age.
+**B** Boxplot of variation in RTL among cohorts.
+**C** RTL and age among cohorts.
+Lines represent fitted values from a linear regression of RTL and log-transformed age, and colours correspond to **B**.
+**D** RTL in relation to and $\Delta$Log age (i.e.
+within individual variation in log age).
+
+**Figure 2** Longitudinal telomere dynamics in the Seychelles warbler.
+**A** Variation in RTL within individuals sampled at different time points.
+The dotted line represents parity, and thus points above and below the line represent increases and decreases in RTL, respectively.
+**B** Scaled density plots of repeated RTL measurements among individual samples, and among different samples taken from the same individual.
+Areas of the density plot to the left of the dotted line represent decreases in RTL, while areas to the right represent increases.
+**C** $\Delta$RTL in relation to age in pairs of samples taken within two years.
+Black line and shaded area represent fitted values and 95% confidence limits from a linear regression of RTL and log-transformed age.
+**D** Probability of telomere lengthening occurring in relation to age.
+Points at zero and one represent pairs of samples where RTL has decreased and increased, respectively, with point size scaled by the number of overlapping values.
+The black line represents the proportion of samples in which increases in RTL where observed at each age category.
+
+**Figure 3** Telomere length in relation to the social and ecological environment in the Seychelles warbler.
+**A** Estimates and 95% confidence intervals for all explanatory variables fitted in a linear mixed model (see methods for details).
+**B** RTL in relation to tarsus length and sex.
+**C** RTL in relation to variation in annual food availability.
+Lines and shaded areas represent the fitted values and 95% confidence limits from linear regressions.
+
+\pagebreak
+
+**Figure 1**
+
+```{r Figure 1}
+
+Fig1A <- ddply(dd,
+               .(BirdID),
+        mutate,
+        nreps = length(RTL)) %>%
+  subset(nreps>0) %>%
+  subset(!duplicated(paste0(BirdID,Agemonths))) %>%
+  ggplot(aes(x = Agemonths,y = RTL,group = BirdID,col=factor(BirdID)))+
+  geom_point(alpha = 1,col='grey')+
+geom_line(col = 'darkgrey',alpha = 0.2)+
+  xlab('Age (years)') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'),
+        legend.position = 'none')+
+  geom_smooth(col = 'black',fill = 'black',lwd = 0.5,aes(group = NULL),method = 'lm',formula = y~log(x))+
+  annotate('text',x = 0,y = 1.7,label = 'A')
+
+c15 <- c("dodgerblue2","#E31A1C", # red
+                "green4",
+                "#FF7F00", # orange
+                "gold1",
+                "skyblue2", # lt pink
+                "palegreen2",
+                "#FDBF6F", # lt orange
+                "gray70", "khaki2","blue1","steelblue4",
+                "darkturquoise",
+                "darkorange4","brown","navy",
+         'yellow','green','khaki3','blue','pink','orange')
+
+Fig1B <- transform(dd,Cohort = factor(LayYear)) %>%
+  ggplot(aes(y = RTL,x = Cohort, fill = Cohort))+
+  xlab('Birth year') +
+  geom_boxplot(outlier.shape = NA) +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12,angle = 90), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'),
+        legend.position = 'none') +
+        scale_fill_manual(values = c15)+
+  annotate('text',x = 1,y = 1.6,label = 'B')
+
+Fig1C <- transform(dd,Cohort = factor(LayYear)) %>%
+  ggplot(aes(y = RTL,x = Agemonths, col = Cohort,fill = Cohort))+
+  xlab('Age (years)') +
+  geom_smooth(method = 'lm', formula = y~log(x),se = F) +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'),
+        legend.position = 'none') +
+    scale_color_manual(values = c15)+
+        scale_fill_manual(values = c15)+
+  annotate('text',x = 0,y = 1.3,label = 'C')
+
+  Fig1D <- ggplot(dd3, aes(x = DeltaLogAge,y = RTL))+
+  geom_point(alpha = 1,col='grey')+
+  xlab(expression(Delta*'Log age (years)')) +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'),
+        legend.position = 'none')+
+  geom_smooth(col = 'black',method = 'lm')+
+  annotate('text',x = -1.5,y = 1.7,label = 'D')
+
+multiplot(Fig1A,Fig1B,Fig1C,Fig1D,layout = matrix(1:4,2,2,byrow = T))
+
+```
+
+\pagebreak
+
+**Figure 2**
+
+```{r Figure 2}
+
+temp <- subset(dd3,TimeDiff < 3.5)
+
+Fig2A <- ggplot(dd3,
+       aes(x = RTL,
+           y = RTL1)) +
+  ylab(expression('RTL ('*italic(t)*' + 1)')) + 
+  xlab(expression('RTL ('*italic(t)*')')) +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        legend.position = c('none'),
+        legend.title = element_text(size = 12),
+        axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black')) +
+    geom_point(col = "grey")+
+  geom_abline(slope = 1,intercept = 0,lty = 2) +
+  annotate('text',x = 0.3,y = 1.7,label = 'A')
+
+Fig2B <- ggplot(ddL,aes(col=Group,fill = Group,y = ..scaled..,x = DeltaRTL))+
+    geom_density(alpha = 0.2)+
+  theme_classic()+
+  ylab ('Scaled density')+
+  xlab(expression(Delta*'RTL'))+
+    theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        legend.position = "top",
+        legend.title = element_blank(),
+        axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black')) +
+  scale_colour_manual(values = c('darkgrey','gold'))+
+  scale_fill_manual(values = c('darkgrey','gold'))+
+  geom_vline(xintercept = 0,lty = 2) +
+  annotate('text',x = -1,y = 1,label = 'B')
+
+Fig2C <- ggplot(temp, aes(x = Agemonths,
+           y = DeltaRTL)) +
+  ylab(expression(Delta*'RTL'))+
+  xlab("Age (years)") +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black')) +
+  geom_point(col = "grey")+
+  geom_smooth(col = 'black',fill = 'black',method = "lm",formula = y~log(x))+
+  geom_abline(slope = 0,intercept = 0,lty = 2)+
+  annotate('text',x = -0.2,y = 1,label = 'C')
+
+Fig2D <- ddply(temp,
+               .(Agemonths),
+               summarise,
+               PropGainer = mean(DeltaRTLF),
+               n = length(DeltaRTLF)) %>%
+ggplot(aes(x = Agemonths,
+           y = PropGainer)) +
+  ylab("Proportion gainers") + 
+  xlab("Age (years)") +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+        legend.position = c('none'),
+        legend.title = element_text(size = 12),
+        axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black')) +
+  geom_line(col="black")+
+  geom_count(data = temp,aes(y = DeltaRTLF,x = Agemonths),col = "grey")+
+  annotate('text',x = 0,y = 1.1,label = 'D')
+
+multiplot(Fig2A,Fig2B,Fig2C,Fig2D,layout = matrix(c(1,2,3,4),nrow = 2,byrow = T))
+
+```
+
+\pagebreak
+
+**Figure 3**
+
+```{r Figure 3}
+
+RTL_Full2 <- standardize(RTL_Full)
+
+CIs <- CIwald(RTL_Full2)
+t1a <- data.frame(myx = row.names(CIs)[6:nrow(CIs)],
+                  Estimate = fixef(RTL_Full2),
+                  X2.5. = CIs[6:nrow(CIs),1],
+                  X97.5. = CIs[6:nrow(CIs),2],
+                  row.names = NULL,
+                  stringsAsFactors = F)
+
+nams <- t1a$myx
+nams[nams == 'z.LogAge'] <- 'Log age'
+nams[nams == 'z.Tarsus'] <- 'Tarsus length'
+nams[nams == 'c.Helper'] <- 'Number helpers'
+nams[nams == 'z.GroupSize'] <- 'Group size'
+nams[nams == 'c.Sex'] <- 'Sex (males)'
+nams[nams == 'z.BodyMass'] <- 'Body mass'
+nams[nams == 'z.Insect'] <- 'Insect abundance'
+nams[nams == 'z.TQ'] <- 'Territory quality'
+nams[nams == 'z.Density'] <- 'Density'
+
+t1a$myx <- nams
+
+t1a$myx <- factor(t1a$myx,levels = t1a$myx[order(t1a$Estimate,decreasing = T)])
+t1a <- subset(t1a,myx != "(Intercept)")
+
+Fig3A <- forestplot(t1a,RI = F)+
+  annotate('text',y = 0.1,x = 0.6,label = 'A')
+
+Fig3B <- ggplot(dd,
+       aes(x = Tarsus,
+           y = RTL,
+           col = Sex)) +
+  geom_point()+
+  ylab('RTL') + 
+  xlab('Tarsus length (mm)') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+                axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'),
+        legend.position = 'top',
+        legend.title = element_blank()) +
+  stat_smooth(method = 'lm') +
+  scale_color_manual(values = c('gold2','red'))+
+  annotate('text',x = 17,y = 1.7,label = 'B')
+
+Fig3C <- ggplot(dd,
+       aes(y = RTL,
+           x=Insect)) +
+    ylab('RTL') + 
+  xlab('Insect abundance') +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12), 
+        axis.text.y = element_text(size = 12), 
+        axis.title.y = element_text(size = 14,vjust=0.8),
+        axis.title.x = element_text(size = 14),
+                axis.line.x = element_line(colour = 'black'),
+        axis.line.y = element_line(colour = 'black'),
+        legend.position = 'none') +
+  geom_boxplot(aes(group = Insect),outlier.shape = NA)+
+   stat_smooth(method = 'lm',col = 'darkgrey') +
+  annotate('text',x = 2.5,y = 1.6,label = 'C')
+
+multiplot(Fig3A,Fig3B,Fig3C,layout = matrix(c(1,1,2,3),byrow = T,nrow = 2))
+
+```

### 2017-03-20 Manuscript/SW_early_life_telomeres_SOM.Rmd
MAJOR EDITS
Changed to pdf format

@@
 ---
-output:
-  word_document:
-    reference_docx: myref.docx
+output: pdf_document
+classoption: landscape
 ---
 
-###Supplementary information for Spurgin *et al.* "Spatiotemporal variation in lifelong telomere dynamics in a long-term ecological study"
+#Supplementary information for Spurgin *et al.* "Spatiotemporal variation in lifelong telomere dynamics in a long-term ecological study"
 
@@
                fig.width= 5,
-               fig.height= 5,
-               fig.show='asis',
-               dev=c('png'))
+               fig.height= 5)
 
@@
 
+\pagebreak
+
 **Table S2** Full model of factors affecting $\Delta$RTL in Seychelles warblers, using longitudinal data.
+
 ```{r Table S2}
@@
 ```
+
+\pagebreak
 
@@
 
-**Figure S1** Model averaged estimates and 95% cofidence intervals for social and ecological environmental variables in relation to RTL in Seychelles warblers.
+\pagebreak
 
@@
 
-**Figure S2** Model averaged estimates and 95% cofidence intervals for social and ecological environmental variables in relation to $\Delta$RTL in Seychelles warblers, using longitudinal data.
+**Figure S1** Model averaged estimates and 95% cofidence intervals for social and ecological environmental variables in relation to RTL in Seychelles warblers.
+
+\pagebreak
 
@@
 ```
+
+**Figure S2** Model averaged estimates and 95% cofidence intervals for social and ecological environmental variables in relation to $\Delta$RTL in Seychelles warblers, using longitudinal data.

### 2017-03-20 Manuscript/Cover letter.Rmd
MAJOR EDITS
Changed to pdf format
Edits - submitted!

@@
 
+\bigskip
+
+Dr Lewis G. Spurgin\\*
+BBSRC Research Fellow\\*
 School of Biological Sciences\\*
@@
 Norwich\\*
-Norfolk NR4 7TJ\\* 
+Norfolk, NR4 7TJ\\* 
 United Kingdom
@@
 
+\bigskip
 \bigskip
@@
 
-Secondly, using a large number of repeated telomere measurements we show that within-individual increases in telomere length cannot be solely explained by measurement error.
+Secondly, using a large number of repeated telomere measurements we show that within-individual increases in telomere length occure, and that these cannot be solely explained by measurement error.
 This is in accordance with recent theoretical work, and data from humans, suggesting that telomere elongation may be an ecologically important process.
@@
 \bigskip
+\bigskip
 

### 2017-03-20 Manuscript/SW_early_life_telomeres_FULL.Rmd
Edits - submitted!

@@
 *Study species and sampling*
+
 The Seychelles warbler is a small (~15 g), insectivorous passerine bird with a mean life expectancy of 5.5 years at fledging [@Hammers2013].
@@
 *Molecular methods*
+
 For each sample, genomic DNA was extracted from a ~2 mm^2^ flake of preserved blood using the DNeasy Blood and Tissue Kit (Qiagen), following the manufacturer's protocol, with the modification of overnight lysis at 37^o^C and a final DNA elution volume of 80 $\mu$l.
@@
 *Statistical analyses*
+
 We performed all statistical analyses using R version 3.2.2 [@RDevelopmentCoreTeam2011].
@@
 *Telomere dynamics and age among cohorts*
+
 We first tested how RTL was related to age among cohorts using a model selection approach.
@@
 *Telomere dynamics and the environment*
+
 In addition to age, RTL was associated with tarsus length, sex and insect abundance (Fig.
@@
 nams[nams == 'DeltaAge2'] <- 'Delta age (quadratic)'
+nams[nams == 'MeanAge'] <- 'Mean age'
 

### 2017-05-23 Manuscript/Response_to_reviewers.Rmd
Edits

---
output: 
  pdf_document:
header-includes:
  \usepackage{graphicx}
  \usepackage{changepage}
  \usepackage{color}
---

\begin{flushright}

\includegraphics{UEAlogo}

\bigskip

Dr Lewis G. Spurgin\\*
BBSRC Research Fellow\\*
School of Biological Sciences\\*
University of East Anglia\\*
Norwich\\*
Norfolk, NR4 7TJ\\* 
United Kingdom

L.Spurgin@uea.ac.uk\\*
lewisspurgin.wordpress.com

\end{flushright}

\bigskip
\bigskip

**June 2017**

\bigskip

Dear Ben,

Thank you for taking the time to consider our manuscript for publication in *Journal of Animal Ecology*. We were very pleased that the editors and reviewers were all so enthusiastic about the quality of our paper, and were impressed by the detailed and constructive nature of the reviews. We have now been through the Associate Editor and reviewer comments in detail, and made substantial changes to our manuscript in response. Below we outline, point-by-point, how we have responded to each of the reviewer comments. We also reproduce the sections of our manuscript we have changed in response to the reviewer comments. Reviewer comments are in bold, and reproduced manuscript sections are indented and italicised.

Please do not hesitate to let me know if you require any further information or clarifications. I look forward to hearing from you.

\bigskip
\bigskip

Kind Regards,

\bigskip
\bigskip

Lewis Spurgin (on behalf of all authors)

\newpage

#Reviewer 1

---

\bigskip

**This paper reports interesting findings on lifelong telomere dynamics and its environmental correlates in a wild bird species. This is a very interesting study, but a moderate/major revision would be needed before considering this paper for publication in my opinion. With such revision, I am convinced that this paper will be an important contribution in the field of telomeres in ecology and evolution. I have some comments/criticism about the methodology presented by the authors, especially about the poor repeatability found between qPCR runs (see major comment 1) and the low efficiency for the telomere amplification. In addition, one main conclusion reached by the authors (i.e. that lifelong telomere dynamics is different between cohort, see comment 6) is not supported by the data in my opinion, or at least by the stats presented in the paper. Despite such limitations, the research question is novel and merits to be addressed and published in high-quality journal such as Journal of Animal Ecology. I really hope that my comments will be useful to the authors in revising their manuscript.**

RESPONSE: We thank Dr Stier for his constructive comments, and are pleased that he believes our work to be novel and worthy of publication. We agree that the main issues he raises were not properly addressed/explained, and respond to these points in detail below.

\bigskip

---

\bigskip

**I am not sure to understand how you calculated the inter-plate repeatability of 0.68 mentioned line 234. Assuming that it is more or less equivalent to the intraclass coefficient of correlation r, it would mean that approximately 50% of the variation in your measurement is linked to the error of measurement. If my understanding is correct, it would mean that your qPCR methodology is quite poorly repeatable. Can you please clarify this point?**

RESPONSE: The reviewer is correct that the estimate of repeatability is the intraclass correlation coefficient (this is made clear in the rptR documentation, which we cite). We agree with the reviewer in the sense that repeatability is not as high as we would have hoped, although we note that in the context of the telomere literature, a repeatability estimate of 0.68 is not especially low. However, because of how the samples were run, we are confident that any issues of repeatability have not led to any kind of bias in our results, and in fact make our analyses conservative. We respond to this point in detail, and outline how we clarified our discussion, in our response to the next comment.

\bigskip

---

\bigskip

**You should mention in the methods how the samples were assigned to the different plates (i.e. totally random, balanced between years/age, or all samples of 1 individual on the same plate?). If you allocated more or less randomly your samples to the different plates, can you check (and present as supplementary) that you do not have a “plate effect” on the average TL measurement. If you do observe such plate effect (something not that unusual with qPCR...), it might contribute to explain your low repeatability, and maybe part of the telomere lengthening that you describe.**

RESPONSE: We are confident that plate effects are not a source of systematic bias in our data. Firstly, we assigned samples completely randomly (i.e. using a random number generator) to plates prior to running the telomere assays, minimising the chance that plate effects would introduce systematic differences among ages, sexes or cohorts. Secondly, our method of estimating RTL normalises samples across plates, as each RTL value is obtained by subtracting sample Cq values from Cq values obtained from a "calibrator" sample, which is run on every plate (information contained within Bebbington et al. 2016, cited). Thirdly, in response to a request by reviewer 3 we have now also included plate ID as a random effect in the models (see below), and our results are virtually identical.

With regards to the observed telomere lenghtening, it is very possible that inter-plate variability accounts some of the observed lengthening in our dataset, but it cannot account for the amount and extent of lengthening we find. Indeed, this is exactly what Figure 2B shows. The lengthening observed "within" samples can be attributed to error across plates, but the additional lengthening observed "among" samples cannot be explained by measurement error.

We now include a brief discussion of the error rate on lines XX:

\begin{adjustwidth}{2cm}{2cm}

\textit{
"Put in the quoted text here..."
}

\end{adjustwidth}

\bigskip

---

\bigskip

**1.78 (78%) would not be considered appropriate by many researchers for qPCR methodology (the commonly accepted limit is more or less + or minus 10%, so here between 90% (1.90) and 110% (2.10)). Your efficiency is more than 20% below the expected 100%, which raises some questions about the validity of your methodology. How can you be confident that this low efficiency is not affecting the quality of your data?**

\bigskip

---

\bigskip

**Important information is missing in the methods section regarding the age and life stage at which birds have been sampled. Telomeres might shorten in a few days during the nestling period (e.g. Stier et al. 2015b, 2016), and might be affected by breeding effort (e.g. Reichert et al. 2014, Sudyka et al. 2014), so it is important in my opinion to know more about the age and status of the birds used in this study.**

\bigskip

---

Lines 145-147: Storage and extraction methods are known to potentially affect qPCR results (e.g. Reichert et al. 2017 in press Oecologia). Storing samples in 100% ethanol at RT is not very common among people measuring telomeres as far as I know. Do you have any papers supporting the relevance of such storage method? Can you provide some DNA integrity gel images as supplementary material? Were older samples more likely to be degraded? Such information would be helpful for the reader to assess the quality of your DNA and to have an idea of the appearance of normal vs. degraded DNA samples.

How can you exclude that the cohort effects observed in your study are not linked to differences in storage conditions / DNA quality-integrity rather than true biological variation? Can you try for instance to disentangle/quantify the proportion of variation linked to the year of sampling (I would not expect a strong biological year effect on adults) vs. the proportion of variation linked to the year of birth (cohort effect)?

Would it be more relevant to present Figure 1B only with telomere data from fledglings individuals of the cohort in question. I am assuming from data presentation than it is an average of TL of all the individuals from these cohorts throughout their life, but I might be wrong. Presenting average TL data over the life of individuals might be somewhat misleading here in my opinion since some cohorts might experience poorer survival than other, etc... In addition, the main cohort effect seems to be on telomere length per se, rather than on telomere dynamics with age (see comment 6 below).

Cohort effects (year of birth in analyses of nestling TL) have been reported without being clearly discussed (e.g. Stier et al. 2014, Becker et al. 2015). It might be worth mentioning it in the discussion, and pointing out the difficulties in distinguishing between real cohort effects and potential methodological artefacts linked to sampling, storage, extraction and/or qPCR methods. Were the different cohorts well randomized between qPCR plates?

You mention in several places that telomere dynamics with age is different between cohorts (e.g. lines 241-244; 299-300; 350-54), but in my opinion this is not supported by your data; so such conclusion should be avoided. Indeed, in the first analysis of RTL, including the log age * cohort interaction only marginally improved the fit of the model (delta AIC of 1.40, lines 242-244), which is not sufficient to support your conclusion. In addition, you mention that including cohort ID in the within-individual analyse of TL was even decreasing the fit of the model to your data, again suggesting that telomere dynamics is not significantly impacted by the cohort.

Please remove any reference to such effect or try to provide convincing analysis justifying your conclusion.

While I agree that your results are very interesting and biologically meaningful (age effect, sex effet, link with body size and food abundance), and that they are very similar to those on Soay sheep from Dan Nussey’s team, an intra-individual repeatability of 8% is a bit worrying in my opinion. This is especially worrying for using telomere length/dynamics as a “biomarker” as you suggest it in several places in the manuscript. A true biomarker should be at least to some extent consistent within an individual to be useful. You should raise such limitation in your discussion.

You mention that in other bird studies such within-individual consistency was much higher because they were done in the lab/nest-boxes. Another major difference with these studies is the shorter time-frame on which they were conducted, in addition to the more controlled conditions to store the samples. Therefore you would expect less noise in theses studies linked to methodological artefact due to sample storage/integrity, which might contribute to explain your low within-individual consistency.

Body mass is known to be less affected than tarsus length by bias linked to different experimenters in long-term studies like yours. Do you have any body mass data, and if yes why not including it in this paper?

Line 16: Telomeres shorten in response to oxidative stress and cellular division. It is important to mention that here since you find higher telomere shortening very early in life that might be linked to cellular division rather than oxidative stress.

Line 30: “Telomere length and the slope of its relationship with age” is not a straightforward formulation in my opinion. You could use “Telomere length and the rate of telomere change with age”, it has more biological meaning in my opinion.

Line 44: “environmental costs” is not clearly defined, so please rephrase or define clearly what you mean by environmental costs here.

Lines 55-56: While I acknowledged that Von Zglinicki 2002 is the common citation used by people to support the role of oxidative stress in telomere shortening, this is in vitro work with cultured cells using supra-physiological doses of ROS. I would suggest to mention that this relationship is observed “in vitro” or provide a reference supporting such assumption in vivo (if you find one, because I am still looking for solid evidence of this relationship..).

Lines 64-65: I am not sure that Horn 2010 and Simons 2015 are appropriate here, however you could cite Stier et al. 2015 Exp. Gerontol. reviewing relationships between age and TL as well as TL and survival in non-mammalian vertebrates species.

Line 82: To help a bit the naïve reader, I would give here some recent examples of “environmental conditions” that are known to affect telomere length; for instance anthropogenic noise (Meillère et al. 2016), altitude (Stier et al. 2016), heat stress (Simide et al. 2016) or infection (Asghar et al. 2016).

Lines 231-234: information on sample size, efficiencies and repeatability would be better placed in the methods section in my opinion, since it is not results per se.

Lines 267-271: You mention that “longitudinal telomere length increased with log age...”, and cite fig 2C, 2D to support your statement. As far as I understand fig 2C, it seems that telomere length decrease with age before more or less 2 years of age, and then is not changing overall with age later in life. Yet, such figure does not fit with the estimate you give in the text, so the presentation of your results is quite confusing here.. I do not see how your results are supporting this particular statement. In my opinion, you should probably state that the probability of detecting a within-individual telomere shortening is decreasing with age // the probability of detecting telomere elongation is increasing with age.

Line 302: the sentence is too vague, especially “particular life stage”.

Line 321: same comment as above “specific periods”

Lines 337-338: Such change in blood cell types would be quite convincing for mammals since “mammalian people” are measuring telomeres from different kind of WBCs and also immature nucleated RBCs.. However, in birds more than 99% of the cells in the blood are RBCs, therefore a change in blood cell types is quite unlikely. However, cellular turnover of RBCs might change with age/life stage, and you can expect that RBCs that have been in the blood stream for a longer time have shorter telomeres due to oxidative stress, though this remain to be tested properly.. Therefore you should probably mention the “cellular composition of the blood (e.g. age of RBCs, proportion of WBCs)” instead of “composition of cell types”.

Lines 348-349: you might cite Becker 2015 and Stier 2014 here too. Lines 350-54: Please reformulate according to major comment 6 above. Line 356: Atema 2015 is not in the wild

Lines 367-369: In my opinion, Heidinger et al. 2012 does not fit here (they do not look at differences in early-life conditions), but you might cite Reichert 2015, Stier 2014, 2015a, 2015b, Meillère 2016.

Lines 378-379: It is unclear here that you speak about a comparative study of the sex effect on TL.

---

\newpage

#Reviewer 2

---

\bigskip

**I congratulate the authors on this fantastic paper. It's very rare for me to read a manuscript and not be able to come up with a single meaningful criticism, but I really can't fault this effort. The authors tackle important emerging questions about telomere dynamics and the importance of telomere length as a biomarker in ecology using a fantastic long term data set. Importantly, and unlike many recent studies of telomere length in wild animals, this study is based on a high quality longitudinal data set spanning a long time series. I particularly like the great care the authors have taken to address the question of within-individual lengthening of telomere length, applying a variety of published methods to test whether observed lengthening is greater than that expected due to measurement error. These results, combined with the thorough analyses of ageing patterns and cohort effects, represent an important addition to our current understanding of the complexity of telomere dynamics in the wild. They also show interesting associations between telomere length, environmental quality, sex and size which offer important insights into the ecological and physiological factors responsible for telomere dynamics. All in all, I really enjoyed reading this paper and have no meaningful suggestions for improvements. I expect this excellent manuscript will make a very strong and widely cited addition to the emerging literature on telomere dynamics in wild populations.**

RESPONSE: We thank Dr Nussey for his extremely kind words. It is not often that one receives such a positive review!

\bigskip

---

\newpage

#Reviewer 3

---

\bigskip

Dear authors,
You have an impressive dataset of longitudinal data and your work on telomere dynamics can be an important contribution to the current knowledge. However, I have some major concerns regarding your manuscript.

Your introduction could be improved, first of all, by citing correctly. Also, there is some more recent literature on the topic, which you could cite. See more in comments on the introduction below. Further, I have some methodological questions. I would like to see more information on how samples were distributed over qPCR plates. Were samples of the same individual on the same plate? If not, how did you take plate ID into account in your statistical analyses? How exactly did you relate territory quality to RTL and delta RTL change? Only for birth year? Reading your abstract and aims in the introduction, I expected more. In the discussion, you speculate on effects of environmental conditions on telomere dynamics. Is there a relationship with telomere lengthening between years and insect abundance/territory quality in the year of lengthening? I guess you could test this with the data you have? This could add to your discussion on potential telomere lengthening, which remained rather methodological. In my opinion, the ecological context in the discussion could be improved. More comments see below.
I apologise for a delayed submission of my review, but hope that you find my comments helpful!

L 53: In Monaghan & Haussmann 2006, data for 4 species are presented whereof two show telomere shortening. I therefore don’t consider this the best citation for your statement. Also, given that by now there are plenty of others that show this relationship - including other Haussmann or Monaghan papers.

L 54: The statement is correct but reference Haussmann, Vleck & Nisbet 2003 is wrong! In this paper the result of a cross-sectional analysis is presented and discussed that one can classify birds into age classes according to their telomere length. Please cite appropriate papers.

L 56: If you use the reference von Zglinicki 2002, please state that this is based on cell culture research. Or site a convincing reference where it has been shown in vivo. As far as I know the relation between oxidative stress and telomere shortening in vivo is less clear.

L 65: Please either cite a meta-analysis or some more like Barrett et al. There are quite some studies with original data showing a relationship between telomere length and survival. Given all the literature available, I do not consider to cite Simons 2015 a good choice here.

L 65-66: Here you state there is little evidence of a direct relationship. Only a few lines above, in sentence line 59-61, you write about a potential causal relationship. Please consider to rewrite this, so that it is in one paragraph.

L 80: Njajour 2007: Where is this reference?

Methods:

L136: You describe how you assess territory quality etc. But how do you relate it to RTL or delta RTL in your statistical analyses? Do you only use territory quality in the hatching year? You could relate territory quality between RTL measurements to telomere dynamics. Is the variation of territory quality within-individually over time smaller than among-individually? Is the difference between years much larger than between territories?

L 165: “time point t + 1” Is the time span between repeated samples always the same? If not, please add a sentence to make it clear. Are samples taken in the same month every year?
Some more information on the sampled individuals your study is based on would be helpful and could be added to the supplement. E.g. cohort sizes.

L 168: How do you explain a non-normal distribution of telomere lengths? Data selection or methodological effect? How large is the variation of samples among cohorts?

L 180: Van de Pol & Wright 2009

L 205: first and least?

L 234: Please provide intra-plate repeatabilities as well. Do you have repeats that are not directly next to each other on the plate?

Given an inter-plate repeatability of 0.68, how do you take plate ID into account in your statistical analyses? Random effect plate ID?

For the samples analysed repeatedly, which measurement did you use? Or average?

L 238-239: If you would add a translation into base pairs lost per year, it would become easier to interpret your results. As the relationship is not linear give examples for a few ages.

L 247: Please compare cross-sectional and longitudinal telomere shortening and discuss it.

L 252: It is always the same time span between sampling, right? Could the low repeatability/correlation be partly due to inter-plate differences? Did you run all samples of the same individuals on the same plates or partly on different plates?

L 307: Again the citation Haussmann et al. is not correct here.

L 340: Does telomere lengthening or shortening relate to environmental conditions in your study?

Generally, the discussion of this part of potential causes of telomere dynamics remains very methodological. The ecological context could be improved by adding own results as mentioned above or more citations.

L 389: Sentence incomplete

### 2017-05-23 Scripts/SW juvenile telomeres IPC models.r

@@ -6,28 +6,28 @@
 
 # Telomere length, age and cohort -----------------------------------------
 
-RTL_Age <- lmer(RTL ~ Cohort + Agemonths + LogAge + FAge + Age2 + (1|BirdID) + (1|CatchYear),
+RTL_Age <- lmer(RTL ~ Cohort + Agemonths + LogAge + FAge + Age2 + (1|BirdID) + (1|CatchYear) + (1|PlateID),
                 data=dd,
                 REML = FALSE)
-RTL_LogAge <- lmer(RTL ~ LogAge + Cohort + (1|BirdID) + (1|CatchYear),
+RTL_LogAge <- lmer(RTL ~ LogAge + Cohort + (1|BirdID) + (1|CatchYear) + (1|PlateID),
                 data=dd,
                 REML = FALSE)
-RTL_LogAgeXcohort <- lmer(RTL ~ LogAge*Cohort + (1|BirdID) + (1|CatchYear),
+RTL_LogAgeXcohort <- lmer(RTL ~ LogAge*Cohort + (1|BirdID) + (1|CatchYear) + (1|PlateID),
                    data=dd,
                    REML = FALSE)
-RTL_DeltaAge <- lmer(RTL ~ MeanAge + Cohort + DeltaAge + DeltaLogAge + DeltaAge2 + (1|BirdID) + (1|CatchYear),
+RTL_DeltaAge <- lmer(RTL ~ MeanAge + Cohort + DeltaAge + DeltaLogAge + DeltaAge2 + (1|BirdID) + (1|CatchYear) + (1|PlateID),
                 data=dd3,
                 REML = FALSE)
-RTL_DeltaLogAge <- lmer(RTL ~ DeltaLogAge + (1|BirdID) + (1|CatchYear),
+RTL_DeltaLogAge <- lmer(RTL ~ DeltaLogAge + (1|BirdID) + (1|CatchYear) + (1|PlateID),
                    data=dd3,
                    REML = FALSE)
-DeltaRTL_LogAge <- lmer(DeltaRTL ~ LogAge + (1|BirdID) + (1|CatchYear),
+DeltaRTL_LogAge <- lmer(DeltaRTL ~ LogAge + (1|BirdID) + (1|CatchYear) + (1|PlateID),
                         data=subset(dd3,TimeDiff < 3.5),
                         REML = FALSE)
[…]

### 2017-05-23 Manuscript/Response_to_reviewers.Rmd
Home edits

@@
 
-With regards to the observed telomere lenghtening, it is very possible that inter-plate variability accounts some of the observed lengthening in our dataset, but it cannot account for the amount and extent of lengthening we find.
+With regards to the observed telomere lengthening, it is very possible that inter-plate variability accounts some of the observed lengthening in our dataset, but it cannot account for the amount and extent of lengthening we find.
 Indeed, this is exactly what Figure 2B shows.
@@
 
-We now include a brief discussion of the error rate on lines XX:
+We now explain our randomisation approach in the methods, line XX:
 
@@
 \textit{
-"Put in the quoted text here..."
+"Prior to qPCR, we used a random number generator to assign samples to qPCR plates, to ensure that no systematic bias could occur with regards to age, sex, cohort or ecological environment."
+}
+
+\end{adjustwidth}
+
+We also include a discussion on error rate on lines XX:
+
+\begin{adjustwidth}{2cm}{2cm}
+
+\textit{
+"Measurement error is one  factor that is an issue in our study, and other studies that use qPCR to measure telomere length (Nussey et al.
+2014).
+Measurement error is unlikely to be a cause of type I error in our study, because we were careful to randomise all samples across qPCR plates, and normalise RTL estimates across plates.
+However, the noise associated with within and among plate measure may have resulted in a decrease in explanatory power, and possibly in a degree of type II error.
+Techniques for measuring telomere length with a greater degree of precision may probe helpful in future ecological studies of teloere dynamics, but at present there is still atrade-off between obtaining precise telomere measurements, and utilising the large sample sizes necessary for ecological study."
 }
@@
 How can you be confident that this low efficiency is not affecting the quality of your data?**
+
+RESPONSE: Again, the level of efficiency is lower than we would have liked, although again suts fairly squarely within that found the published literature (see Fairlie et al.
+2015, for a recent example), and often it is not even reported.
+We also note that this problem is not explected to be too drastic, because variaiton in efficiency is taken into account when calculating RTL from the Cq values.
+And again, we suggest that because of our study design, random error of this kind makes the majority of our analyses conservative, and is accounted for in the analysis of telomere lengthening.
 
@@
 
+RESPONSE: We now provide more clarification about how we determined and analyse age, lines XX:
+
+\begin{adjustwidth}{2cm}{2cm}
+
+\textit{
+"The majority of individuals are ringed (with an individually numbered metal ring and unique combination of colour rings) within the first year of life, and so are of known age.
+We aged all birds using information on eye colour at first capture (Kondeur et al.
+1991) and previous capture history (Richardson et al.
+2003).
+Within the first year of life, birds are classified as nestlings less than one month old (rounded to one month for analyses), fledglings less than six month olds (rounded to six months) or subadults up to one year old (rounded to 10 months).
+Ages for adult birds were rounded to the nearest year."
+}
+
+\end{adjustwidth}
+
+We do not have more detailed data on age for the majority of our individuals, but given that we are studying broad, lifelong patterns of telomere length and dynamics, we feel that the resolution we have is sufficient.
+
 \bigskip
@@
 
-Lines 145-147: Storage and extraction methods are known to potentially affect qPCR results (e.g.
+\bigskip
+
+**Lines 145-147: Storage and extraction methods are known to potentially affect qPCR results (e.g.
 Reichert et al.
@@
 Such information would be helpful for the reader to assess the quality of your DNA and to have an idea of the appearance of normal vs.
-degraded DNA samples.
+degraded DNA samples.**
 
-How can you exclude that the cohort effects observed in your study are not linked to differences in storage conditions / DNA quality-integrity rather than true biological variation?
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**How can you exclude that the cohort effects observed in your study are not linked to differences in storage conditions / DNA quality-integrity rather than true biological variation?
 Can you try for instance to disentangle/quantify the proportion of variation linked to the year of sampling (I would not expect a strong biological year effect on adults) vs.
-the proportion of variation linked to the year of birth (cohort effect)?
+the proportion of variation linked to the year of birth (cohort effect)?**
 
-Would it be more relevant to present Figure 1B only with telomere data from fledglings individuals of the cohort in question.
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**Would it be more relevant to present Figure 1B only with telomere data from fledglings individuals of the cohort in question.
 I am assuming from data presentation than it is an average of TL of all the individuals from these cohorts throughout their life, but I might be wrong.
 Presenting average TL data over the life of individuals might be somewhat misleading here in my opinion since some cohorts might experience poorer survival than other, etc...
-In addition, the main cohort effect seems to be on telomere length per se, rather than on telomere dynamics with age (see comment 6 below).
+In addition, the main cohort effect seems to be on telomere length per se, rather than on telomere dynamics with age (see comment 6 below).**
 
-Cohort effects (year of birth in analyses of nestling TL) have been reported without being clearly discussed (e.g.
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**Cohort effects (year of birth in analyses of nestling TL) have been reported without being clearly discussed (e.g.
 Stier et al.
@@
 It might be worth mentioning it in the discussion, and pointing out the difficulties in distinguishing between real cohort effects and potential methodological artefacts linked to sampling, storage, extraction and/or qPCR methods.
-Were the different cohorts well randomized between qPCR plates?
+Were the different cohorts well randomized between qPCR plates?**
+
+RESPONSE:
+
+\bigskip
+
+---
 

### 2017-05-23 Manuscript/SW_early_life_telomeres_FULL.Rmd
Home edits

@@
 Breeding attempts are often unsuccessful, and as a result of this low reproductive output, and higher mortality in first-year birds [39% in first-year birds versus 16% in adults; @Brouwer2006], cohort sizes in the Seychelles warbler are typically small (< 50).
-The 22 birth year cohorts used in this study cover 1993 to 2014 – the time period during which our data and sampling are most complete.
+The 22 birth year cohorts used in this study cover `r min(dd$LayYear)` to `r max(dd$LayYear)` – the time period during which our data and sampling are most complete.
 
@@
 We aged all birds using information on eye colour at first capture [@Komdeur1991a] and previous capture history [@Richardson2003b].
+Within the first year of life, birds are classified as nestlings less than one month old (rounded to one month for analyses), fledglings less than six month olds (rounded to six months) or subadults up to one year old (rounded to 10 months).
+Ages for adult birds were rounded to the nearest year.
 As Seychelles warblers are non-migratory endemics naturally confined to the island [@Komdeur2004a], an extensive biannual census of birds on Cousin during each breeding season gives accurate measures of local density, social status (e.g.
@@
 We measured relative telomere length (RTL) for all samples using a quantitative PCR (qPCR) assay of telomeres and a GAPDH control gene, following Bebbington *et al.* [-@Bebbington2016].
+Prior to qPCR, we used a random number generator to assign samples to qPCR plates, to ensure that no systematic bias could occur with regards to age, sex, cohort or ecological environment.
 Based on the distribution of observed cq values, we excluded outlier samples with extremely large cq values (cq values > 25 and 26 were excluded for the telomere and GAPDH reactions, respectively), which were assumed to be failed reactions.
@@
 A poor social and ecological environment is known to be detrimental to Seychelles warblers, both in terms of oxidative stress and survival [@VandeCrommenacker2011a; @Brouwer2012], and it is therefore perhaps surprising that these variables do not explain more variance in RTL.
-While telomere measurement error will account for some of this lack of explanatory power, it is also likely that early-life RTL in the Seychelles warbler is explained by a set of environmental and genetic variables not considered here [e.g.
+Measurement error is one  factor that is an issue in our study, and other studies that use qPCR to measure telomere length [@Nussey2014].
+Measurement error is unlikely to be a cause of type I error in our study, because we were careful to randomise all samples across qPCR plates, and normalise RTL estimates across plates.
+However, the noise associated with within and among plate measure may have resulted in a decrease in explanatory power, and possibly in a degree of type II error.
+Techniques for measuring telomere length with a greater degree of precision may probe helpful in future ecological studies of teloere dynamics, but at present there is still atrade-off between obtaining precise telomere measurements, and utilising the large sample sizes necessary for ecological study.
+
+Sampling error notwithstanding, we also predict that lifelong variation in RTL in the Seychelles warbler is explained by a set of environmental and genetic variables not considered here [e.g.
 @Bebbington2016].

### 2017-05-24 Manuscript/Response_to_reviewers.Rmd
Edits

@@
 
-RESPONSE:
-
-\bigskip
-
----
-
-\bigskip
-
 **How can you exclude that the cohort effects observed in your study are not linked to differences in storage conditions / DNA quality-integrity rather than true biological variation?
@@
 
-RESPONSE:
+RESPONSE: Storage in ethanol is standardd procedure for avian blood samples, and allows DNA to be maintained at high quality for many years.
+We checked DNA integrity for a very large number of samples, and found no evidence of DNA degration in older samples.
+We now include two gel images to illustrate this in the supplementary material (Fig.
+S1).
+We refer to this in the main text, lines XX.
+
+\begin{adjustwidth}{2cm}{2cm}
+
+\textit{
+We found no evidence of DNA degradation in older samples (Fig. S1). 
+}
+
+\end{adjustwidth}
+
+We also performed an analysis of RTL and sample storage time, using adult birds, and found no evidence of a relationship - lines XX.
+
+\begin{adjustwidth}{2cm}{2cm}
+
+\textit{
+Using adult samples greater than one year old, we checked whether RTL was related to sample storage time, and found no evidence of such a relationship (estimate = -0.002, CIs = -0.007, 0.002).
+}
+
+\end{adjustwidth}
 
@@
 
-RESPONSE:
+RESPONSE: Figure 1B did not show averages for individuals, but rather all measures for all individuals from each cohort (i.e.
+including repeats from individuals).
+The aim here was to show the full distribution of telomere lengths for each cohort.
+However, on balance we agree that restricting this plot to juveniles is probably less biased.
+We have now altered the figure accordingly.
 
@@
 
-You mention in several places that telomere dynamics with age is different between cohorts (e.g.
+\bigskip
+
+**You mention in several places that telomere dynamics with age is different between cohorts (e.g.
 lines 241-244; 299-300; 350-54), but in my opinion this is not supported by your data; so such conclusion should be avoided.
@@
 In addition, you mention that including cohort ID in the within-individual analyse of TL was even decreasing the fit of the model to your data, again suggesting that telomere dynamics is not significantly impacted by the cohort.
+Please remove any reference to such effect or try to provide convincing analysis justifying your conclusion.**
 
-Please remove any reference to such effect or try to provide convincing analysis justifying your conclusion.
+RESPONSE:
 
-While I agree that your results are very interesting and biologically meaningful (age effect, sex effet, link with body size and food abundance), and that they are very similar to those on Soay sheep from Dan Nussey’s team, an intra-individual repeatability of 8% is a bit worrying in my opinion.
+\bigskip
+
+---
+
+\bigskip
+
+**While I agree that your results are very interesting and biologically meaningful (age effect, sex effet, link with body size and food abundance), and that they are very similar to those on Soay sheep from Dan Nussey’s team, an intra-individual repeatability of 8% is a bit worrying in my opinion.
 This is especially worrying for using telomere length/dynamics as a “biomarker” as you suggest it in several places in the manuscript.
 A true biomarker should be at least to some extent consistent within an individual to be useful.
-You should raise such limitation in your discussion.
+You should raise such limitation in your discussion.**
 
-You mention that in other bird studies such within-individual consistency was much higher because they were done in the lab/nest-boxes.
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**You mention that in other bird studies such within-individual consistency was much higher because they were done in the lab/nest-boxes.
 Another major difference with these studies is the shorter time-frame on which they were conducted, in addition to the more controlled conditions to store the samples.
-Therefore you would expect less noise in theses studies linked to methodological artefact due to sample storage/integrity, which might contribute to explain your low within-individual consistency.
+Therefore you would expect less noise in theses studies linked to methodological artefact due to sample storage/integrity, which might contribute to explain your low within-individual consistency.**
 
-Body mass is known to be less affected than tarsus length by bias linked to different experimenters in long-term studies like yours.
-Do you have any body mass data, and if yes why not including it in this paper?
+RESPONSE:
 
-Line 16: Telomeres shorten in response to oxidative stress and cellular division.
-It is important to mention that here since you find higher telomere shortening very early in life that might be linked to cellular division rather than oxidative stress.
+\bigskip
 
-Line 30: “Telomere length and the slope of its relationship with age” is not a straightforward formulation in my opinion.
-You could use “Telomere length and the rate of telomere change with age”, it has more biological meaning in my opinion.
+---
 
-Line 44: “environmental costs” is not clearly defined, so please rephrase or define clearly what you mean by environmental costs here.
+\bigskip
 
-Lines 55-56: While I acknowledged that Von Zglinicki 2002 is the common citation used by people to support the role of oxidative stress in telomere shortening, this is in vitro work with cultured cells using supra-physiological doses of ROS.
-I would suggest to mention that this relationship is observed “in vitro” or provide a reference supporting such assumption in vivo (if you find one, because I am still looking for solid evidence of this relationship..).
+**Body mass is known to be less affected than tarsus length by bias linked to different experimenters in long-term studies like yours.
+Do you have any body mass data, and if yes why not including it in this paper?**
 
-Lines 64-65: I am not sure that Horn 2010 and Simons 2015 are appropriate here, however you could cite Stier et al.
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**Line 16: Telomeres shorten in response to oxidative stress and cellular division.
+It is important to mention that here since you find higher telomere shortening very early in life that might be linked to cellular division rather than oxidative stress.**
+
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**Line 30: “Telomere length and the slope of its relationship with age” is not a straightforward formulation in my opinion.
+You could use “Telomere length and the rate of telomere change with age”, it has more biological meaning in my opinion.**
+
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**Line 44: “environmental costs” is not clearly defined, so please rephrase or define clearly what you mean by environmental costs here.**
+
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**Lines 55-56: While I acknowledged that Von Zglinicki 2002 is the common citation used by people to support the role of oxidative stress in telomere shortening, this is in vitro work with cultured cells using supra-physiological doses of ROS.
+I would suggest to mention that this relationship is observed “in vitro” or provide a reference supporting such assumption in vivo (if you find one, because I am still looking for solid evidence of this relationship..).**
+
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**Lines 64-65: I am not sure that Horn 2010 and Simons 2015 are appropriate here, however you could cite Stier et al.
 2015 Exp.
 Gerontol.
-reviewing relationships between age and TL as well as TL and survival in non-mammalian vertebrates species.
+reviewing relationships between age and TL as well as TL and survival in non-mammalian vertebrates species.**
 
-Line 82: To help a bit the naïve reader, I would give here some recent examples of “environmental conditions” that are known to affect telomere length; for instance anthropogenic noise (Meillère et al.
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**Line 82: To help a bit the naïve reader, I would give here some recent examples of “environmental conditions” that are known to affect telomere length; for instance anthropogenic noise (Meillère et al.
 2016), altitude (Stier et al.
@@
 2016) or infection (Asghar et al.
-2016).
+2016).**
 
-Lines 231-234: information on sample size, efficiencies and repeatability would be better placed in the methods section in my opinion, since it is not results per se.
+RESPONSE:
 
-Lines 267-271: You mention that “longitudinal telomere length increased with log age...”, and cite fig 2C, 2D to support your statement.
+\bigskip
+
+---
+
+\bigskip
+
+**Lines 231-234: information on sample size, efficiencies and repeatability would be better placed in the methods section in my opinion, since it is not results per se.**
+
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**Lines 267-271: You mention that “longitudinal telomere length increased with log age...”, and cite fig 2C, 2D to support your statement.
 As far as I understand fig 2C, it seems that telomere length decrease with age before more or less 2 years of age, and then is not changing overall with age later in life.
@@
 I do not see how your results are supporting this particular statement.
-In my opinion, you should probably state that the probability of detecting a within-individual telomere shortening is decreasing with age // the probability of detecting telomere elongation is increasing with age.
+In my opinion, you should probably state that the probability of detecting a within-individual telomere shortening is decreasing with age // the probability of detecting telomere elongation is increasing with age.**
 
-Line 302: the sentence is too vague, especially “particular life stage”.
+RESPONSE:
 
-Line 321: same comment as above “specific periods”
+\bigskip
 
-Lines 337-338: Such change in blood cell types would be quite convincing for mammals since “mammalian people” are measuring telomeres from different kind of WBCs and also immature nucleated RBCs..
+---
+
+\bigskip
+
+**Line 302: the sentence is too vague, especially “particular life stage”.**
+
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**Line 321: same comment as above “specific periods”.**
+
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**Lines 337-338: Such change in blood cell types would be quite convincing for mammals since “mammalian people” are measuring telomeres from different kind of WBCs and also immature nucleated RBCs..
 However, in birds more than 99% of the cells in the blood are RBCs, therefore a change in blood cell types is quite unlikely.
@@
 Therefore you should probably mention the “cellular composition of the blood (e.g.
-age of RBCs, proportion of WBCs)” instead of “composition of cell types”.
+age of RBCs, proportion of WBCs)” instead of “composition of cell types”.**
 
-Lines 348-349: you might cite Becker 2015 and Stier 2014 here too. Lines 350-54: Please reformulate according to major comment 6 above. Line 356: Atema 2015 is not in the wild
+RESPONSE:
 
-Lines 367-369: In my opinion, Heidinger et al.
-2012 does not fit here (they do not look at differences in early-life conditions), but you might cite Reichert 2015, Stier 2014, 2015a, 2015b, Meillère 2016.
+\bigskip
 
-Lines 378-379: It is unclear here that you speak about a comparative study of the sex effect on TL.
+---
+
+\bigskip
+
+**Lines 348-349: you might cite Becker 2015 and Stier 2014 here too.**
+
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**Lines 350-54: Please reformulate according to major comment 6 above.**
+
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**Line 356: Atema 2015 is not in the wild**
+
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**Lines 367-369: In my opinion, Heidinger et al.
+2012 does not fit here (they do not look at differences in early-life conditions), but you might cite Reichert 2015, Stier 2014, 2015a, 2015b, Meillère 2016.**
+
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**Lines 378-379: It is unclear here that you speak about a comparative study of the sex effect on TL.**
+
+RESPONSE:
+
+\bigskip
 

### 2017-05-24 Manuscript/SW_early_life_telomeres_FULL.Rmd
Edits

@@
 DNA integrity was further validated by visualization with ethidium bromide after electrophoresis on a 1.2% agarose gel, and all samples with evidence of DNA degradation were re-extracted or excluded.
+We found no evidence of DNA degradation in older samples (Fig.
+S1).
 All DNA extractions that passed the above criteria were diluted to 3.3 ng $\mu$l^-1^ before telomere measurement.
@@
 Inter-plate repeatability of RTL, based on `r nrow(subset(ddL,Group == 'Within samples'))` samples measured at least twice at different time points, was `r round2(rptall$R)` (CI = `r round2(rptall$CI_emp[1])`, `r round2(rptall$CI_emp[2])`).
+Using adult samples greater than one year old, we checked whether RTL was related to sample storage time, and found no evidence of such a relationship (estimate = `r round3(fixef(RTL_ST)['StorageTime'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_ST)['StorageTime',],lessthan = F),collapse = ', ')`).
 
@@
 The model averaging approach yielded qualitatively identical results to the full LMM, with the same explanatory variables 'significant' in terms of being retained in top models, and having model-averaged confidence intervals not overlapping zero (Table S1; Fig.
-S1).
+S2).
 One interesting finding from the model selection was that sex only appeared in top models where tarsus length was also present (Table S1).
@@
 $\Delta$RTL was positively related to age, consistent with telomere shortening being highest in early life, and negatively related to population density; however, in both instances model averaged confidence intervals overlapped zero (Fig.
-S2).
+S3).
 
@@
 The thick line and shaded area represent the fitted values and 95% confidence limits of a linear regression of RTL and log-transformed age.
-**B** Boxplot of variation in RTL among cohorts.
+**B** Boxplot of variation in RTL among juvenile individuals from all cohorts.
 **C** RTL and age among cohorts.
@@
   subset(!duplicated(paste0(BirdID,Agemonths))) %>%
+  subset(Agemonths > 2) %>%
   ggplot(aes(x = Agemonths,y = RTL,group = BirdID,col=factor(BirdID)))+

### 2017-05-24 Manuscript/SW_early_life_telomeres_SOM.Rmd
Edits

@@
 
+**Figure S1** Examples of DNA extractions from older and more recent Seychelles warbler samples. We found no evidence for differences in DNA integrity with sample age.
+
+\pagebreak
+
 ```{r Figure S1}
@@
 
-**Figure S1** Model averaged estimates and 95% cofidence intervals for social and ecological environmental variables in relation to RTL in Seychelles warblers.
+**Figure S2** Model averaged estimates and 95% cofidence intervals for social and ecological environmental variables in relation to RTL in Seychelles warblers.
 
@@
 
-**Figure S2** Model averaged estimates and 95% cofidence intervals for social and ecological environmental variables in relation to $\Delta$RTL in Seychelles warblers, using longitudinal data.
+**Figure S3** Model averaged estimates and 95% cofidence intervals for social and ecological environmental variables in relation to $\Delta$RTL in Seychelles warblers, using longitudinal data.

### 2017-05-24 Scripts/SW juvenile telomeres IPC clean.R

@@ -82,6 +82,7 @@ dd <- subset(dd,!is.na(LogAge))
 dd$Cohort <- factor(dd$LayYear)
 dd$FAge <- factor(dd$Age)
 dd$Age2 <- poly(dd$Agemonths,2)[,2]
+dd$StorageTime <- 2017-dd$CatchYear
 
 # Tarsus and delta age ------------------------------------------------------------------
 

### 2017-05-24 Manuscript/Response_to_reviewers.Rmd
Edits

@@
 
-RESPONSE:
+RESPONSE: See response above re.
+randomisation.
+While not an issue for our study, we agree that it is a point worth raising.
+We have added a brief point in the discussion to highlight the issue, lines XX:
+
+\begin{adjustwidth}{2cm}{2cm}
+
+\textit{"Although a few studies have shown that temporal variation in telomere dynamicsoccurs in natural populations, these have been limited in the number of seasons they cover (Mizutani et al.
+2013; Watson, Bolton and Monaghan 2015; Fairlie et al.
+2016).
+Other studies have found chort effects but not discussed them in an ecological context (Stier et al.
+2014; Becker et al.2015).
+One problem with studying cohort effects is that it can be difficult to tease apart true cohort effects from effects that arise due to sample degradation with aim, and/or batch effects in telomere assays, although neither of these factors were a problem in our study.
+Indeed, the long-term Seychelles warbler dataset has allowed us to show that temporal variation in telomere dynamics can occur over substantial time periods."}
+
+\end{adjustwidth}
 
@@
 
-Your introduction could be improved, first of all, by citing correctly.
-Also, there is some more recent literature on the topic, which you could cite.
-See more in comments on the introduction below.
+Your introduction could be improved, first of all, by citing correctly. Also, there is some more recent literature on the topic, which you could cite. See more in comments on the introduction below.
+
 Further, I have some methodological questions.
@@
 Reading your abstract and aims in the introduction, I expected more.
+
 In the discussion, you speculate on effects of environmental conditions on telomere dynamics.
@@
 In my opinion, the ecological context in the discussion could be improved.
-More comments see below.
-I apologise for a delayed submission of my review, but hope that you find my comments helpful!
 
@@
 L 80: Njajour 2007: Where is this reference?
-
-Methods:
 

### 2017-05-24 Manuscript/SW_early_life_telomeres_FULL.Rmd
Edits

@@
 Although a few studies have shown that temporal variation in telomere dynamics occurs in natural populations, these have been limited in the number of seasons they cover [@Mizutani2013; @Watson2015; @Fairlie2016].
-The long-term Seychelles warbler dataset has allowed us to show that temporal variation in telomere dynamics can occur over substantial time periods.
+Other studies have found chort effects but not discussed them in an ecological context [@Stier2014; @Becker2015].
+One problem with studying cohort effects is that it can be difficult to tease apart true cohort effects from effects that arise due to sample degradation with aim, and/or batch effects in telomere assays, although neither of these factors were a problem in our study.
+Indeed, the long-term Seychelles warbler dataset has allowed us to show that temporal variation in telomere dynamics can occur over substantial time periods.
 Our data suggest that conditions during the hatch year are a very important factor in shaping telomere dynamics, and that age-related declines in telomere length vary among cohorts.

### 2017-05-26 Manuscript/Response_to_reviewers.Rmd
Edits

@@
 
-RESPONSE: See response above re.
-randomisation.
-While not an issue for our study, we agree that it is a point worth raising.
-We have added a brief point in the discussion to highlight the issue, lines XX:
+RESPONSE: See response above re. randomisation. While not an issue for our study, we agree that it is a point worth raising. We now discuss this issue on lines XX:
 
@@
 
-RESPONSE:
+RESPONSE: We agree that while there is convincing evidence for cohort effects on telomere length, the evidence for cohort effects on telomere dynamics is much weaker.
+We have therefore removed references to this effect throughout.
 
@@
 
-RESPONSE:
+RESPONSE: We are not sure that "worrying" is the correct phrase to use, as we anticpiate that other studies in natural population will find similar patterns.
+Further, we disagree with the sentiment that a biomarker must be consistent over individual lifespans to be useful.
+We feel that we cover this point in the discussion, lines XX:
+
+\begin{adjustwidth}{2cm}{2cm}
+
+\textit{"Such inconsistent changes in telomere length over lifespans could occur due to changes in the composition of cell types within individual samples, or due to the actual elongation of telomeres (Blackburn et al.
+1999(.
+Determining the mechanism of these changes is essential for how we view telomeres as biomarkers of costs.
+For example, if telomeres can be lengthened in response to improvements in environmental conditions, this would suggest that they reflect short- to medium-term costs, rather than the cumulative costs that an individual has faced over its lifespan (Bateson et al.
+2016).
+New statistical and technical approaches are therefore now required to determine the mechanisms behind increases in telomere length within individuals, so that biologically informed hypotheses about the ecological causes and consequences of these increases can be generated and tested."}
+
+\end{adjustwidth}
 
@@
 
-RESPONSE:
+RESPONSE: It is true that the shorter timescales of these studies may be the reason for the higher within-individual consistency, and we have now added a note to this effect, lines XX:
+
+\begin{adjustwidth}{2cm}{2cm}
+
+\textit{"Consistent with a pattern of sporadic changes in telomere length with age, we found that within-individual telomere measurements were only weakly correlated.
+These findings are in contrast to other avian studies in which within-individual telomere length measurements were highly consistent, and individual-level telomere shortening occurred throughout the juvenile period and into adulthood (Heidinger et al.
+2012; Boonekamp et al.
+2014).
+However, the lifelong telomere dynamics found in Seychelles warblers are strikingly similar to those found in Soay sheep (Fairlie et al.
+2016).
+This discrepancy in results may be because in our study, and that of Fairlie et al.
+(2016), individuals were born and reared in the wild, as opposed to in nestbox or laboratory conditions.
+Alternatively it may be because our longitudinal telomere measurements have been taken over longer time periods."}
+
+\end{adjustwidth}
+
+However, as we argue above, we do not think it is likely that the pattern is a result of sample storage/DNA integrity.
+Past discussions with the Nussey group suggests that they are also confident their telomere assay is not negatively affected by long-term sample storage.
 
@@
 
-RESPONSE:
+RESPONSE: We do have body mass data, but decided to exclude it for a number of reasons.
+Firstly, with tarsus length we wanted to control for variation in structural size - even with measurer effects, tarsus length is a better indicator of structural size than body mass.
+Body mass is more of a measure of (short-term) condition, and varies according to the time of day, season, etc.
+In some respects body mass is likely to respond in a similar way to the environment as telomere length (especially if telomeres can get longer).
+There is thus a circularity issue with including bodymass as a predictor of telomere length.
+Finally, from apractical perspective, including body mass introduces collinearity issues, particularly with tarsus length, but also with other ecological variables.
 
@@
 
-RESPONSE:
+RESPONSE: Yes, this is a good point. We now clarify in the discussion, lines XX
+
+\begin{adjustwidth}{2cm}{2cm}
+
+\textit{"Our study adds to the substantial body of literature from humans and wild animals showing that telomere length decreases with age, and that this decrease is most rapid in early life - most likely as a consequence of cellular division..."}
+
+\end{adjustwidth}
 
@@
 
-RESPONSE:
+RESPONSE: We have now removed this sentence, as suggested in one of the previous comments.
 
@@
 
-RESPONSE:
+RESPONSE: We have changed the phrase to "individual responses to the environment".
 
@@
 
-RESPONSE:
+RESPONSE: We have clarified that this is in vitro.
 
@@
 
-RESPONSE:
+RESPONSE: We have changed the citations as suggested.
 
@@
 
-RESPONSE:
+RESPONSE: We have added in the suggested references.
 
@@
 
-RESPONSE:
+RESPONSE: We disagree here - it is the outcome of the molecular methods. We tried moving it but this disrupted the flow of the manuscript se have left it as it was.
 
@@
 
-RESPONSE:
+RESPONSE: We agree that this was confusing, and have simplified along the suggested lines - now lines XX:
+
+\begin{adjustwidth}{2cm}{2cm}
+
+\textit{"We found that the likelihood of telomere lengthening increased with log age...
+Increases in telomere length were most likely to be observed shortly after the juvenile period, at around four years of age, and later in life (although sample sizes for older birds are much smaller; Fig.
+2C,D)"}
+
+\end{adjustwidth}
 
@@
 
-RESPONSE:
+RESPONSE: We have now simplified the sentence to make it clearer, lines XX:
+
+\begin{adjustwidth}{2cm}{2cm}
+
+\textit{"Despite an overall pattern of telomere shortening with age in the Seychelles warbler, we found evidence of within-individual increases in telomere length, and that the extent of these increases could not be explained solely by qPCR measurement error."}
+
+\end{adjustwidth}
 
@@
 
-RESPONSE:
+RESPONSE: We are not completely sure what is unclear about this, but have tried rephrasing. Now lines XX:
+
+\begin{adjustwidth}{2cm}{2cm}
+
+\textit{"Increases in telomere length were not consistent over individual lifespans, but occurred in bouts, against a backdrop of overall lifelong telomere shortening.
+"}
+
+\end{adjustwidth}
 
@@
 
-RESPONSE:
+RESPONSE: This is a good point and we have made the suggested edit.
 
@@
 
-RESPONSE:
+RESPONSE: These bave been added in response to an earlier comment.
 
@@
 
-RESPONSE:
+RESPONSE: We have now removed this section.
 
@@
 
-RESPONSE:
+RESPONSE: We have removed this citation as part of a previous comment.
 
@@
 
-RESPONSE:
+RESPONSE: We have removed the Heidinger reference, leaving two citations in - we felt this was sufficient.
 
@@
 
-RESPONSE:
+RESPONSE: We rephrased accordingly, lines XX:
+
+\begin{adjustwidth}{2cm}{2cm}
+
+\textit{"Also worth noting is that the effect of telomere length on survival is strongest in male Seychelles warblers (Barrett et al.
+2013), although comparative research suggests that the nature of the relationship between sex, telomeres and survival is not yet clear (Barrett and Richardson 2011)."}
+
+\end{adjustwidth}
 
@@
 
-Dear authors,
-You have an impressive dataset of longitudinal data and your work on telomere dynamics can be an important contribution to the current knowledge.
-However, I have some major concerns regarding your manuscript.
+**Dear authors,**
+**You have an impressive dataset of longitudinal data and your work on telomere dynamics can be an important contribution to the current knowledge.
+However, I have some major concerns regarding your manuscript.**
 
-Your introduction could be improved, first of all, by citing correctly. Also, there is some more recent literature on the topic, which you could cite. See more in comments on the introduction below.
+**Your introduction could be improved, first of all, by citing correctly.
+Also, there is some more recent literature on the topic, which you could cite.
+See more in comments on the introduction below.**
 
-Further, I have some methodological questions.
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**Further, I have some methodological questions.
 I would like to see more information on how samples were distributed over qPCR plates.
@@
 Only for birth year?
-Reading your abstract and aims in the introduction, I expected more.
+Reading your abstract and aims in the introduction, I expected more.**
 
-In the discussion, you speculate on effects of environmental conditions on telomere dynamics.
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**In the discussion, you speculate on effects of environmental conditions on telomere dynamics.
 Is there a relationship with telomere lengthening between years and insect abundance/territory quality in the year of lengthening?
@@
 This could add to your discussion on potential telomere lengthening, which remained rather methodological.
-In my opinion, the ecological context in the discussion could be improved.
+In my opinion, the ecological context in the discussion could be improved.**
 
-L 53: In Monaghan & Haussmann 2006, data for 4 species are presented whereof two show telomere shortening.
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**L 53: In Monaghan & Haussmann 2006, data for 4 species are presented whereof two show telomere shortening.
 I therefore don’t consider this the best citation for your statement.
-Also, given that by now there are plenty of others that show this relationship - including other Haussmann or Monaghan papers.
+Also, given that by now there are plenty of others that show this relationship - including other Haussmann or Monaghan papers.**
 
-L 54: The statement is correct but reference Haussmann, Vleck & Nisbet 2003 is wrong!
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**L 54: The statement is correct but reference Haussmann, Vleck & Nisbet 2003 is wrong!
 In this paper the result of a cross-sectional analysis is presented and discussed that one can classify birds into age classes according to their telomere length.
-Please cite appropriate papers.
+Please cite appropriate papers.**
 
-L 56: If you use the reference von Zglinicki 2002, please state that this is based on cell culture research.
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**L 56: If you use the reference von Zglinicki 2002, please state that this is based on cell culture research.
 Or site a convincing reference where it has been shown in vivo.
-As far as I know the relation between oxidative stress and telomere shortening in vivo is less clear.
+As far as I know the relation between oxidative stress and telomere shortening in vivo is less clear.**
 
-L 65: Please either cite a meta-analysis or some more like Barrett et al.
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**L 65: Please either cite a meta-analysis or some more like Barrett et al.
 There are quite some studies with original data showing a relationship between telomere length and survival.
-Given all the literature available, I do not consider to cite Simons 2015 a good choice here.
+Given all the literature available, I do not consider to cite Simons 2015 a good choice here.**
 
-L 65-66: Here you state there is little evidence of a direct relationship.
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**L 65-66: Here you state there is little evidence of a direct relationship.
 Only a few lines above, in sentence line 59-61, you write about a potential causal relationship.
-Please consider to rewrite this, so that it is in one paragraph.
+Please consider to rewrite this, so that it is in one paragraph.**
 
-L 80: Njajour 2007: Where is this reference?
+RESPONSE:
 
-L136: You describe how you assess territory quality etc.
+\bigskip
+
+---
+
+\bigskip
+
+**L 80: Njajour 2007: Where is this reference?**
+
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**L136: You describe how you assess territory quality etc.
 But how do you relate it to RTL or delta RTL in your statistical analyses?
@@
 Is the variation of territory quality within-individually over time smaller than among-individually?
-Is the difference between years much larger than between territories?
+Is the difference between years much larger than between territories?**
 
-L 165: “time point t + 1” Is the time span between repeated samples always the same? If not, please add a sentence to make it clear. Are samples taken in the same month every year?
-Some more information on the sampled individuals your study is based on would be helpful and could be added to the supplement. E.g. cohort sizes.
+RESPONSE:
 
-L 168: How do you explain a non-normal distribution of telomere lengths? Data selection or methodological effect? How large is the variation of samples among cohorts?
+\bigskip
 
-L 180: Van de Pol & Wright 2009
+---
 
-L 205: first and least?
+\bigskip
 
-L 234: Please provide intra-plate repeatabilities as well. Do you have repeats that are not directly next to each other on the plate?
+**L 165: “time point t + 1” Is the time span between repeated samples always the same? If not, please add a sentence to make it clear. Are samples taken in the same month every year?**
 
-Given an inter-plate repeatability of 0.68, how do you take plate ID into account in your statistical analyses? Random effect plate ID?
+RESPONSE:
 
-For the samples analysed repeatedly, which measurement did you use? Or average?
+\bigskip
 
-L 238-239: If you would add a translation into base pairs lost per year, it would become easier to interpret your results. As the relationship is not linear give examples for a few ages.
+---
 
-L 247: Please compare cross-sectional and longitudinal telomere shortening and discuss it.
+\bigskip
 
-L 252: It is always the same time span between sampling, right?
+**Some more information on the sampled individuals your study is based on would be helpful and could be added to the supplement. E.g. cohort sizes.**
+
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**L 168: How do you explain a non-normal distribution of telomere lengths? Data selection or methodological effect? How large is the variation of samples among cohorts?**
+
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**L 180: Van de Pol & Wright 2009**
+
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**L 205: first and least?**
+
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**L 234: Please provide intra-plate repeatabilities as well. Do you have repeats that are not directly next to each other on the plate?**
+
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**Given an inter-plate repeatability of 0.68, how do you take plate ID into account in your statistical analyses? Random effect plate ID?**
+
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**For the samples analysed repeatedly, which measurement did you use? Or average?**
+
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**L 238-239: If you would add a translation into base pairs lost per year, it would become easier to interpret your results. As the relationship is not linear give examples for a few ages.**
+
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**L 247: Please compare cross-sectional and longitudinal telomere shortening and discuss it.**
+
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**L 252: It is always the same time span between sampling, right?
 Could the low repeatability/correlation be partly due to inter-plate differences?
-Did you run all samples of the same individuals on the same plates or partly on different plates?
+Did you run all samples of the same individuals on the same plates or partly on different plates?**
 
-L 307: Again the citation Haussmann et al. is not correct here.
+RESPONSE:
 
-L 340: Does telomere lengthening or shortening relate to environmental conditions in your study?
+\bigskip
 
-Generally, the discussion of this part of potential causes of telomere dynamics remains very methodological.
-The ecological context could be improved by adding own results as mentioned above or more citations.
+---
 
-L 389: Sentence incomplete
+\bigskip
+
+**L 307: Again the citation Haussmann et al. is not correct here.**
+
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**L 340: Does telomere lengthening or shortening relate to environmental conditions in your study?**
+
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**Generally, the discussion of this part of potential causes of telomere dynamics remains very methodological.
+The ecological context could be improved by adding own results as mentioned above or more citations.**
+
+RESPONSE:
+
+\bigskip
+
+---
+
+\bigskip
+
+**L 389: Sentence incomplete**
+
+RESPONSE:
+
+\bigskip
+
+---

### 2017-05-26 Manuscript/SW_early_life_telomeres_FULL.Rmd
Edits

@@
 bibliography: MyBib.bib
-csl: JAE.csl
 header-includes: \usepackage{graphicx}
@@
 4.
-Telomere length and the slope of its relationship with age varied markedly among cohorts.
+Telomere length varied markedly among cohorts.
 Telomere length was positively associated with temporal variation in island-wide insect abundance - a key resource for the insectivorous Seychelles warbler - suggesting that the costs associated with living in harsher environments can be studied by investigating telomere dynamics.
@@
 A major aim of ecologists and evolutionary biologists is to understand why individuals vary in their response to different environmental factors.
-Identifying this variation in environmental costs is central to understanding variation in fitness [@Lindstrom1999], and thus for understanding population and community dynamics [@Bolnick2011].
+Identifying this variation in individual responses to the environment is central to understanding variation in fitness [@Lindstrom1999], and thus for understanding population and community dynamics [@Bolnick2011].
 Furthermore, knowledge of the relative impact that different environmental factors exert on individuals, and why individuals may differ in mitigating these costs, is important to understanding evolutionary trade-offs and life-history strategies [@Stearns1992].
@@
 @Frenck1998; @Haussmann2003].
-Telomere shortening is accelerated by oxidative stress, which can be elevated due to many environmental factors [@VonZglinicki2002].
+*In vitro* research has shown tha telomere shortening can be accelerated by oxidative stress  [@VonZglinicki2002], which can be elevated due to many environmental factors.
 There is evidence from humans, and from captive and wild animal populations, that telomere shortening is influenced by the conditions experienced during both early life and adulthood  [@Price2013; @Monaghan2014; @Nettle2015; @Reichert2015].
@@
 
-The association between senescence and telomere length has inspired a great deal of recent research into telomere evolutionary ecology and relationships between telomere dynamics and survival or lifespan have been documented in wild population of several species [@Horn2010; @Haussmann2010; @Barrett2013a; @Simons2015].
+The association between senescence and telomere length has inspired a great deal of recent research into telomere evolutionary ecology and relationships between telomere dynamics and survival or lifespan have been documented in wild population of several species [@Haussmann2010; @Barrett2013a; @Stier2015].
 As yet, there is little direct evidence that the relationship between telomere dynamics and survival is causal [@Simons2015].
@@
 Understanding how different factors shape telomere length variation is important, as before we can use telomeres as a measure of the costs experienced by individuals, we need to know how different developmental, genetic and ecological variables interact to affect telomeres.
-Telomere length and rates of shortening can vary according to parental characteristics [Njajour2007; @Heidinger2016], among sexes [@Barrett2011; @Watson2017], and with a whole host of environmental conditions experienced at different life-history stages [@Monaghan2014].
+Telomere length and rates of shortening can vary according to parental characteristics [Njajour2007; @Heidinger2016], among sexes [@Barrett2011; @Watson2017], and with a whole host of environmental conditions, including altitude [@Stier2016], heat stress [@Simide2016] or infection [@Asghar2016].
 Recent evidence suggests that telomere dynamics are indeed highly variable over individual lifespans, and that even the relationship between telomeres and age can vary markedly among cohorts [@Fairlie2016].
@@
 To better understand how longitudinal telomere dynamics vary with age, we examined patterns of short-term telomere change, including only pairs of samples taken within two years of each other.
-We found that longitudinal telomere length increased with log age both when considered as the absolute degree of change (LMM with $\Delta$RTL~*individual*~ as response; estimate = `r round3(fixef(DeltaRTL_LogAge)['LogAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(DeltaRTL_LogAge)['LogAge',],lessthan = F),collapse = ', ')`; Fig.
-2C), and the likelihood of lengthening (GLMM with lengthened yes/no as binomial response; estimate = `r round3(fixef(DeltaRTLF_LogAge)['LogAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(DeltaRTLF_LogAge)['LogAge',],lessthan = F),collapse = ', ')`; Fig.
-2D).
+We found that the likelihood of telomere lengthening increased with log age (GLMM with lengthened yes/no as binomial response; estimate = `r round3(fixef(DeltaRTLF_LogAge)['LogAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(DeltaRTLF_LogAge)['LogAge',],lessthan = F),collapse = ', ')`).
 Increases in telomere length were most likely to be observed shortly after the juvenile period, at around four years of age, and later in life (although sample sizes for older birds are much smaller; Fig.
@@
 We found that telomere length decreases with age, and that this decrease is greatest very early in life.
-Telomere length decreased with age in almost all of the `r length(table(dd$Cohort))` cohorts studied, but the rate of decrease of telomere length with age, and consequently the telomere length of adults, varied substantially among cohorts.
-Despite an overall pattern of telomere shortening with age in the Seychelles warbler, we found that our measure of telomere length increased within some individuals at particular life stages, and that the extent of these increases could not be explained solely by qPCR measurement error.
+Telomere length decreased with age in almost all of the `r length(table(dd$Cohort))` cohorts studied, but also that telomere length varied substantially among cohorts.
+Despite an overall pattern of telomere shortening with age in the Seychelles warbler, we found evidence of within-individual increases in telomere length, and that the extent of these increases could not be explained solely by qPCR measurement error.
 Finally, we found that telomeres are related to tarsus length in a sex-specific manner, and that telomere length is positively associated with temporal fluctuations in food availability.
 
-Our study adds to the substantial body of literature from humans and wild animals showing that telomere length decreases with age, and that this decrease is most rapid in early life [e.g.
+Our study adds to the substantial body of literature from humans and wild animals showing that telomere length decreases with age, and that this decrease is most rapid in early life - most likely as a consequence of cellular division [e.g.
 @Frenck1998; @Haussmann2003; @Heidinger2012].
@@
 
-Increases in telomere length were not consistent over individual lifespans, but occurred during specific periods, against a backdrop of overall lifelong telomere shortening.
+Increases in telomere length were not consistent over individual lifespans, but occurred in bouts, against a backdrop of overall lifelong telomere shortening.
 This is consistent with recent findings in edible dormice *Glis glis*, in which telomere elongation was observed only later in life [@Hoelzl2016a].
 Consistent with a pattern of sporadic changes in telomere length with age, we found that within-individual telomere measurements were only weakly correlated.
-These findings are in contrast to other avian studies where birds were reared in laboratory [@Heidinger2012] or nest-box [@Boonekamp2014] conditions, in which within-individual telomere length measurements were highly consistent, and individual-level telomere shortening occurred throughout the juvenile period and into adulthood.
-However, the lifelong telomere dynamics found in Seychelles warblers are strikingly similar to those found in Soay sheep [@Fairlie2016], and we anticipate that a similar pattern may be found in other populations.
+These findings are in contrast to other avian studies in which within-individual telomere length measurements were highly consistent, and individual-level telomere shortening occurred throughout the juvenile period and into adulthood [@Heidinger2012; @Boonekamp2014].
+However, the lifelong telomere dynamics found in Seychelles warblers are strikingly similar to those found in Soay sheep [@Fairlie2016].
+This discrepancy in results may be because in our study, and that of Fairlie et al.
+[-@Fairlie2016], individuals were born and reared in the wild, as opposed to in nestbox or laboratory conditions.
+Alternatively it may be because our longitudinal telomere measurements have been taken over longer time periods.
 
@@
 Our data, and that of others [@Fairlie2016;@Hoelzl2016; @Hoelzl2016a] suggest that this is not the case.
-Such inconsistent changes in telomere length over lifespans could occur due to changes in the composition of cell types within individual samples, or due to the actual elongation of telomeres [@Blackburn1989].
+Such inconsistent changes in telomere length over lifespans could occur due to changes in the cellular composition of of the blood within individual samples, or due to the actual elongation of telomeres [@Blackburn1989].
 Determining the mechanism of these changes is essential for how we view telomeres as biomarkers of costs.
@@
 Indeed, the long-term Seychelles warbler dataset has allowed us to show that temporal variation in telomere dynamics can occur over substantial time periods.
-Our data suggest that conditions during the hatch year are a very important factor in shaping telomere dynamics, and that age-related declines in telomere length vary among cohorts.
-The rate of early-life telomere loss for a cohort was highly variable and had a persistent effect of the mean adult telomere length of the cohort.
-This effect is consistent with telomere length being strongly influenced by environmental, as well as genetic, factors - something that is becoming apparent from the, as yet, limited number of quantitative genetic studies of telomere dynamics in wild populations [e.g.
-@Olsson2011; @Atema2015; @Becker2015].
-Moreover, our findings suggest that the telomere dynamics of a population at a given point in time represent a snapshot of a temporally varying process.
+Our data suggest that conditions during the hatch year are a very important factor in shaping telomere dynamics throughout lifespan.
+Thus, our findings suggest that the telomere dynamics of a population at a given point in time represent a snapshot of a temporally varying process.
 Research of telomere dynamics within and across multiple cohorts and populations will enable us to better understand how and why population-level telomere dynamics vary over space and time.
@@
 One possibility is that the environment imposes differential costs on males and females: a recent study in captive zebra finches found that manipulation of dietary nutrients had sex-dependent effects on telomere dynamics [@Noguera2015].
-Also worth noting is that the effect of telomere length on survival is strongest in male Seychelles warblers [@Barrett2013a], although the nature of the relationship between sex, telomeres and survival is not yet clear [@Barrett2011]
+Also worth noting is that the effect of telomere length on survival is strongest in male Seychelles warblers [@Barrett2013a], although comparative research suggests that the nature of the relationship between sex, telomeres and survival is not yet clear [@Barrett2011].
 

### 2017-05-26 Scripts/SW juvenile telomeres IPC models.r

@@ -13,10 +13,10 @@ RTL_ST <- lmer(RTL ~ StorageTime + (1|BirdID) + (1|PlateID),
 RTL_Age <- lmer(RTL ~ Cohort + Agemonths + LogAge + FAge + Age2 + (1|BirdID) + (1|CatchYear) + (1|PlateID),
                 data=dd,
                 REML = FALSE)
-RTL_LogAge <- lmer(RTL ~ LogAge + Cohort + (1|BirdID) + (1|CatchYear) + (1|PlateID),
+RTL_LogAge <- lmer(RTL ~ LogAge + Cohort + (1|BirdID) + (1|CatchYear),
                 data=dd,
                 REML = FALSE)
-RTL_LogAgeXcohort <- lmer(RTL ~ LogAge*Cohort + (1|BirdID) + (1|CatchYear) + (1|PlateID),
+RTL_LogAgeXcohort <- lmer(RTL ~ LogAge*Cohort + (1|BirdID) + (1|CatchYear),
                    data=dd,
                    REML = FALSE)
 RTL_DeltaAge <- lmer(RTL ~ MeanAge + Cohort + DeltaAge + DeltaLogAge + DeltaAge2 + (1|BirdID) + (1|CatchYear),

### 2017-05-31 Manuscript/Response_to_reviewers.Rmd
Edits

@@
 
-RESPONSE: We do have body mass data, but decided to exclude it for a number of reasons.
-Firstly, with tarsus length we wanted to control for variation in structural size - even with measurer effects, tarsus length is a better indicator of structural size than body mass.
-Body mass is more of a measure of (short-term) condition, and varies according to the time of day, season, etc.
-In some respects body mass is likely to respond in a similar way to the environment as telomere length (especially if telomeres can get longer).
-There is thus a circularity issue with including bodymass as a predictor of telomere length.
-Finally, from apractical perspective, including body mass introduces collinearity issues, particularly with tarsus length, but also with other ecological variables.
+RESPONSE: As stated in the methods, we do include body mass data in our analyses. It is not related to RTL - see figure 3.
 
@@
 Also, there is some more recent literature on the topic, which you could cite.
-See more in comments on the introduction below.**
-
-RESPONSE:
-
-\bigskip
-
----
-
-\bigskip
-
-**Further, I have some methodological questions.
+See more in comments on the introduction below.
+Further, I have some methodological questions.
 I would like to see more information on how samples were distributed over qPCR plates.
@@
 Only for birth year?
-Reading your abstract and aims in the introduction, I expected more.**
-
-RESPONSE:
-
-\bigskip
-
----
-
-\bigskip
-
-**In the discussion, you speculate on effects of environmental conditions on telomere dynamics.
+Reading your abstract and aims in the introduction, I expected more.
+Your introduction could be improved, first of all, by citing correctly.
+Also, there is some more recent literature on the topic, which you could cite.
+See more in comments on the introduction below.
+Further, I have some methodological questions.
+I would like to see more information on how samples were distributed over qPCR plates.
+Were samples of the same individual on the same plate?
+If not, how did you take plate ID into account in your statistical analyses?
+How exactly did you relate territory quality to RTL and delta RTL change?
+Only for birth year?
+Reading your abstract and aims in the introduction, I expected more.
+In the discussion, you speculate on effects of environmental conditions on telomere dynamics.
 Is there a relationship with telomere lengthening between years and insect abundance/territory quality in the year of lengthening?
@@
 This could add to your discussion on potential telomere lengthening, which remained rather methodological.
-In my opinion, the ecological context in the discussion could be improved.**
+In my opinion, the ecological context in the discussion could be improved.
+More comments see below.
+I apologise for a delayed submission of my review, but hope that you find my comments helpful!**
+**
 
-RESPONSE:
+RESPONSE: We are pleased that the reviewer is generally positive about are manuscript.
+We think it a little harsh to imply that we have not "cited correctly", but are grateful for the suggested changes.
+We respond to each of these questions/comments below.
 
@@
 **L 53: In Monaghan & Haussmann 2006, data for 4 species are presented whereof two show telomere shortening.
-I therefore don’t consider this the best citation for your statement.
+I therefore don't consider this the best citation for your statement.
 Also, given that by now there are plenty of others that show this relationship - including other Haussmann or Monaghan papers.**
 
-RESPONSE:
+RESPONSE: We have changed the citations here.
 
@@
 
-RESPONSE:
+RESPONSE: We have replaced the Haussmann reference with Heidinger 2012, PNAS.
 
@@
 
-RESPONSE:
+RESPONSE: Done - see response to reviewer 1, who made the same point.
 
@@
 
-RESPONSE:
+RESPONSE: We disagree here - Simons 2015 *is* a meta-analysis that shows that there is little evidence that the relationship between telomeres and survival is causal.
 
@@
 
-RESPONSE:
+RESPONSE: We have re-written the text so that the contrast is clearer here. Lines XX:
+
+\begin{adjustwidth}{2cm}{2cm}
+
+\textit{"When telomeres become critically short, cells senesce (Campisi 2003), and the accumulation of these cells has been suggested to result in organismal senescence and death (Wong et al.
+2003).
+The association between senescence and telomere length has inspired a great deal of recent research into telomere evolutionary ecology, and relationships between telomere dynamics and survival or lifespan have been documented in wild population of several species (Haussmann and Marchetto 2010; Barrett et al.
+2013; Stier et al.
+2015).
+As yet, however, there is little direct evidence that the relationship between telomere dynamics and survival is causal (Simons 2015).}
+
+\textit{"Although the causal role of telomeres in senescence and survival is not yet clear, there is mounting evidence that telomeres can act as biomarkers of individual condition and ageing in wild populations, providing a measure of the ecological stress that an individual has experienced, a signature that can otherwise be difficult to detect (e.g.
+Schultner et al.
+2014; Asghar et al.
+2015; Bebbington et al.
+2016)."}
+
+\end{adjustwidth}
 
@@
 
-RESPONSE:
+RESPONSE: Apologies - this is now in the bibliography (there was also a typo - the reference is actually Njajou et al. 2007).
 
@@
 
-RESPONSE:
+RESPONSE: We now clarify (lines XX) that the territory quality estimates that we use in our statistical models are taken from the field period in which the telomere sample was taken.
+The same applies to insect abundance, group size etc.
+We feel that an in-depth explanation of how and why territory quality varies over space and time is outside the scope of this manuscript, especially given i) the word limit, and ii) the fact that territory quality is not related to RTL.
 
@@
 
-RESPONSE:
+RESPONSE: Samples are taken as and when we can catch the birds, which is not always in the same month. We now clarify this on lines XX:
 

### 2017-05-31 Manuscript/SW_early_life_telomeres_FULL.Rmd
Edits

@@
 bibliography: MyBib.bib
+csl: JAE.csl
 header-includes: \usepackage{graphicx}
@@
 Telomeres are repetitive DNA sequences at the ends of linear chromosomes that protect against DNA damage.
-Telomeres generally shorten with age [@Monaghan2006], and there is evidence from a range of taxa that telomere shortening is fastest in early life [e.g.
-@Frenck1998; @Haussmann2003].
+Telomeres generally shorten with age [@Muezzinler2013; @Barrett2013a], and there is evidence from a range of taxa that telomere shortening is fastest in early life [e.g.
+@Frenck1998; @Heidinger2012].
 *In vitro* research has shown tha telomere shortening can be accelerated by oxidative stress  [@VonZglinicki2002], which can be elevated due to many environmental factors.
@@
 When telomeres become critically short, cells senesce [@Campisi2003], and the accumulation of these cells has been suggested to result in organismal senescence and death [@Wong2003].
+The association between senescence and telomere length has inspired a great deal of recent research into telomere evolutionary ecology, and relationships between telomere dynamics and survival or lifespan have been documented in wild population of several species [@Haussmann2010; @Barrett2013a; @Stier2015].
+As yet, however, there is little direct evidence that the relationship between telomere dynamics and survival is causal [@Simons2015].
 
-The association between senescence and telomere length has inspired a great deal of recent research into telomere evolutionary ecology and relationships between telomere dynamics and survival or lifespan have been documented in wild population of several species [@Haussmann2010; @Barrett2013a; @Stier2015].
-As yet, there is little direct evidence that the relationship between telomere dynamics and survival is causal [@Simons2015].
-However, there is mounting evidence that telomeres can act as biomarkers of individual condition and ageing in wild populations, providing a measure of the ecological stress that an individual has experienced, a signature that can otherwise be difficult to detect [e.g.
+Although the causal role of telomeres in senescence and survival is not yet clear, there is mounting evidence that telomeres can act as biomarkers of individual condition and ageing in wild populations, providing a measure of the ecological stress that an individual has experienced, a signature that can otherwise be difficult to detect [e.g.
 @Schultner2014; @Asghar2015; @Bebbington2016].
-There is also evidence that telomere length, measured longitudinally in individuals, can increase as well as decrease [@Muezzinler2013; @Simons2014; @Bateson2016a], which has important ramifications for our understanding of how telomeres reflect costs.
+There is also evidence that telomere length, measured longitudinally in individuals, can increase as well as decrease [@Simons2014; @Bateson2016a], which has important ramifications for our understanding of how telomeres reflect costs.
 However, such increases in telomere length are often attributed to measurement error [@Steenstrup2013; but see @Bateson2016a], and as such their ecological significance is unknown.
@@
 Understanding how different factors shape telomere length variation is important, as before we can use telomeres as a measure of the costs experienced by individuals, we need to know how different developmental, genetic and ecological variables interact to affect telomeres.
-Telomere length and rates of shortening can vary according to parental characteristics [Njajour2007; @Heidinger2016], among sexes [@Barrett2011; @Watson2017], and with a whole host of environmental conditions, including altitude [@Stier2016], heat stress [@Simide2016] or infection [@Asghar2016].
+Telomere length and rates of shortening can vary according to parental characteristics [@Njajou2007; @Heidinger2016], among sexes [@Barrett2011; @Watson2017], and with a whole host of environmental conditions, including altitude [@Stier2016], heat stress [@Simide2016] or infection [@Asghar2015].
 Recent evidence suggests that telomere dynamics are indeed highly variable over individual lifespans, and that even the relationship between telomeres and age can vary markedly among cohorts [@Fairlie2016].
@@
 We used LMMs to explore how variation in environmental and social conditions influenced telomere length and dynamics within cohorts.
-We first created a full model with RTL as a response variable, alongside the following explanatory variables: log age (based on the RTL and age analysis; see results), tarsus length, body mass, sex, insect abundance, territory quality, island-wide population density (an annual measure estimated from the summer breeding census), territory group size, and the number of helping subordinate birds present in the territory.
+We first created a full model with RTL as a response variable, alongside the following explanatory variables: log age (based on the RTL and age analysis; see results), tarsus length at capture, body mass at capture, sex, insect abundance in sampling season, territory quality in sampling season, island-wide population density in sampling season (an annual measure estimated from the summer breeding census), territory group size in sampling season, and the number of helping subordinate birds present in the territory in the sampling season.
 The random effects structure was informed by the analysis of telomere dynamics and age (see results): we included individual ID, cohort ID, and a random slope of log age among cohorts (to allow the effect of age on RTL to vary among cohorts).

### 2017-06-01 Manuscript/Response_to_reviewers.Rmd
Edits

@@
 2016, cited).
-Thirdly, in response to a request by reviewer 3 we have now also included plate ID as a random effect in the models (see below), and our results are virtually identical.
 
@@
 
+\begin{adjustwidth}{2cm}{2cm}
+
+\textit{"Note that the follow-up time between samples varied among individuals, and that individuals are not always caught at exactly the same time of year."}
+
+\end{adjustwidth}
+
 \bigskip
@@
 
-RESPONSE:
+RESPONSE: We have now added a table of cohort sizes to the supplement (Table S1), and refer to this in the main text on Line XX.
 
@@
 
-RESPONSE:
+RESPONSE: Positive skew in telomere length distribution is a widespread and uncontroversial phenomenon, and the causes are likely biological (see for example Golubev et al.
+2003 J.
+Theor.
+Biol.; Hirt et al.
+2014; J.
+Math.
+Biol.; ).
+Investigating the causes of this are outside the scope of our study, and we feel that a discussion of this would be an unnecessary distraction.
 

### 2017-06-01 Manuscript/SW_early_life_telomeres_FULL.Rmd
Edits

@@
 Negative values of RTL reflect decreases in telomere length with age, while positive values reflect increases.
+Note that the follow-up time between samples varied among individuals, and that individuals are not always caught at exactly the same time of year.
 
@@
 ##Results
-We measured telomere lengths using a total of `r length(unique(dd$BloodID))` unique samples from juvenile and adult Seychelles warblers from `r length(table(dd$Cohort))` cohorts born between `r min(dd$LayYear)` and `r max(dd$LayYear)`.
+We measured telomere lengths using a total of `r length(unique(dd$BloodID))` unique samples from juvenile and adult Seychelles warblers from `r length(table(dd$Cohort))` cohorts born between `r min(dd$LayYear)` and `r max(dd$LayYear)` (Table S1).
 Efficiencies (mean $\pm$ s.d.) for our telomere and GAPDH reactions were `r round2(mean(dd$EffTelomere))` $\pm$ `r round2(sd(dd$EffTelomere))` and `r round2(mean(dd$EffGAPDH))` $\pm$ `r round2(sd(dd$EffGAPDH))` respectively.
@@
 The full model was weak in terms of explanatory power of fixed effects (marginal R^2^ = `r round2(r.squaredGLMM(RTL_Full)['R2m'])`), although including the random effect terms increased this substantially (conditional R^2^ = `r round2(r.squaredGLMM(RTL_Full)['R2c'])`).
-The model averaging approach yielded qualitatively identical results to the full LMM, with the same explanatory variables 'significant' in terms of being retained in top models, and having model-averaged confidence intervals not overlapping zero (Table S1; Fig.
+The model averaging approach yielded qualitatively identical results to the full LMM, with the same explanatory variables 'significant' in terms of being retained in top models, and having model-averaged confidence intervals not overlapping zero (Table S2; Fig.
 S2).
-One interesting finding from the model selection was that sex only appeared in top models where tarsus length was also present (Table S1).
+One interesting finding from the model selection was that sex only appeared in top models where tarsus length was also present (Table S2).
 In accordance with this, when tarsus length was removed from the full model sex was no longer significant (estimate = `r round3(fixef(RTL_Full_NoTarsus)['SexMales'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Full_NoTarsus)['SexMales',],lessthan = F),collapse = ', ')`), and a sex x tarsus interaction was significant when included (estimate = `r round3(fixef(RTL_Full_SexTarsus)['SexMales:Tarsus'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Full_SexTarsus)['SexMales:Tarsus',],lessthan = F),collapse = ', ')`); RTL decreased with tarsus length in both sexes, but this decrease was stronger in females (Fig.
 3B).
-No social or ecological environmental variables were significant predictors of $\Delta$RTL using the full model approach (Table S2).
-Using model selection, we found that the top model explaining $\Delta$RTL contained age and population density (Table S3).
+No social or ecological environmental variables were significant predictors of $\Delta$RTL using the full model approach (Table S3).
+Using model selection, we found that the top model explaining $\Delta$RTL contained age and population density (Table S4).
 $\Delta$RTL was positively related to age, consistent with telomere shortening being highest in early life, and negatively related to population density; however, in both instances model averaged confidence intervals overlapped zero (Fig.

### 2017-06-01 Manuscript/SW_early_life_telomeres_SOM.Rmd
Edits

@@
 
-**Table S1** Model selection table for analysis of factors affecting early life telomere length in juvenile Seychelles warblers.
+**Table S1** Cohort sizes for each birth year of the Seychelles warbler dataset used in this study.
+```{r Table S1}
+x <- subset(dd,!duplicated(BirdID)) %>%
+  ddply(.(LayYear),
+        summarise,
+        N = length(RTL)) %>%
+  `colnames<-`(c("Cohort", "N"))
+kable(x)
+```
+
+\pagebreak
+
+**Table S2** Model selection table for analysis of factors affecting early life telomere length in juvenile Seychelles warblers.
 Only models that remained in the top model set for model averaging (AICc $\leq$ 6 compared to the top model) are included here.
 
-```{r Table S1}
+```{r Table S2}
 
@@
 
-tableS1 <- data.frame(Model = apply(t1,1,getterms,c(2:10)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
-colnames(tableS1) <- c('Model','df','AICc', 'Delta AICc','Weight')
-rownames(tableS1) <- NULL
-tableS1$Model[tableS1$Model == ''] <- 'Null model'
-tableS1[,3:5] <- lapply(tableS1[,3:5],as.numeric)
-kable(tableS1,digits=2)
+tableS2 <- data.frame(Model = apply(t1,1,getterms,c(2:10)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
+colnames(tableS2) <- c('Model','df','AICc', 'Delta AICc','Weight')
+rownames(tableS2) <- NULL
+tableS2$Model[tableS2$Model == ''] <- 'Null model'
+tableS2[,3:5] <- lapply(tableS2[,3:5],as.numeric)
+kable(tableS2,digits=2)
 
@@
 
-**Table S2** Full model of factors affecting $\Delta$RTL in Seychelles warblers, using longitudinal data.
+**Table S3** Full model of factors affecting $\Delta$RTL in Seychelles warblers, using longitudinal data.
 
-```{r Table S2}
+```{r Table S3}
 
@@
 
-**Table S3** Model selection table for analysis of factors affecting $\Delta$RTL in Seychelles warblers, using longitudinal data.
+**Table S4** Model selection table for analysis of factors affecting $\Delta$RTL in Seychelles warblers, using longitudinal data.
 Only models that remained in the top model set for model averaging (AICc $\leq$ 6 compared to the top model) are included here.
@@
 
-```{r Table S3}
+```{r Table S4}
 
@@
 
-tableS3 <- data.frame(Model = apply(t1,1,getterms,c(2:10)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
-colnames(tableS3) <- c('Model','df','AICc', 'Delta AICc','Weight')
-rownames(tableS3) <- NULL
-tableS3$Model[tableS3$Model == ''] <- 'Null model'
-tableS3[,3:5] <- lapply(tableS3[,3:5],as.numeric)
-kable(tableS3,digits=2)
+tableS4 <- data.frame(Model = apply(t1,1,getterms,c(2:10)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
+colnames(tableS4) <- c('Model','df','AICc', 'Delta AICc','Weight')
+rownames(tableS4) <- NULL
+tableS4$Model[tableS4$Model == ''] <- 'Null model'
+tableS4[,3:5] <- lapply(tableS4[,3:5],as.numeric)
+kable(tableS4,digits=2)
 

### 2017-06-01 Scripts/SW juvenile telomeres IPC clean.R

@@ -11,7 +11,7 @@ dd0 <- subset(dd0,CqTelomere < 25) %>%
   subset(DiffTelomere < 0.5) %>%
   subset(RTL < 3) %>%
   subset(BloodID>6)
-
+dd0$RTL0 <- dd0$RTL
 dd0$RTL <- sqrt(dd0$RTL)
 
 

### 2017-06-02 Manuscript/Response_to_reviewers.Rmd
Edits

@@
 
-RESPONSE: Positive skew in telomere length distribution is a widespread and uncontroversial phenomenon, and the causes are likely biological (see for example Golubev et al.
-2003 J.
-Theor.
-Biol.; Hirt et al.
-2014; J.
-Math.
-Biol.; ).
+RESPONSE: Positive skew in telomere length distribution is a widespread and uncontroversial phenomenon (see for example Heidinger et al.
+2012 PNAS; Barrett et al.
+2013 Mol Ecol., plus many others), and we suspect that the causes are biological.
 Investigating the causes of this are outside the scope of our study, and we feel that a discussion of this would be an unnecessary distraction.
@@
 
-RESPONSE:
+RESPONSE: Yes, thanks - this was a reference manager error that we missed.
 
@@
 
-RESPONSE:
+RESPONSE: This was a typo - we meant to type "first and last". Now fixed.
 
@@
 
-RESPONSE:
+RESPONSE: NEED TO GO BACK TO RAW DATA TO DO
 
@@
 
-RESPONSE:
+RESPONSE: As samples are normalised within plates, including a plate ID random effect would not make sense. The error across plates is random. See also response to reviewer 1.
 
@@
 
-RESPONSE:
+RESPONSE: We used a mean. We have now clarified this on line XX.
 
@@
 
-RESPONSE:
+RESPONSE: We are not convinced that this would be accurate enough to be useful, and very few ecological studies do this.
+See Bebbington et al.
+2016 Mol Ecol for a discussion of why were using relative telomere length.
 
@@
 
-RESPONSE:
+RESPONSE: We are not sure exactly what the reviewer wants us to do here.
+We do compare cross-sectional and longitudinal telomere shortening, in the sense that we demonstrate how both measures shorten with log age.
+We are happy to add more if there is space, and if we are given a clarification.
 
@@
 
-RESPONSE:
+RESPONSE: No, it is not always the same time between sampling.
+And yes, part of the low repeatability could be explained by inter-plate error, as individual samples were fully randomised across plates.
+However, the within-individual repeatability is still far lower than the within-sample (across-plate) repeatability.
+We now add a caveat to the discussion to highlight this - lines XX:
 
@@
 
-RESPONSE:
+RESPONSE: Now removed.
 
@@
 
-RESPONSE:
+RESPONSE: As stated in the discussion, we feel that this is an important area for future research, and we hope that it will be a whole manuscript on its own.
+We are already pushing the page limit for Journal of Animal Ecology, so adding in an entire new set of analyses is just not possible.
+But watch this space!
 
@@
 
-RESPONSE:
+RESPONSE: See response above regarding the analysis of lengthening.
+Although this section is somewhat methodological, we feel that i) it is a very important point to be able to separate telomere elongation from sampling error, as few studies have done so in ecological systems, and ii) we make it clear (lines XX) that our findings are likely to be of ecological importance.
 
@@
 
-RESPONSE:
+RESPONSE: Now fixed.
 

### 2017-06-02 Manuscript/SW_early_life_telomeres_FULL.Rmd
Edits

@@
 We performed all statistical analyses using R version 3.2.2 [@RDevelopmentCoreTeam2011].
-RTL was square root transformed to improve linear model fits, and we assessed repeatability of RTL using the rptR package.
+RTL was square root transformed to improve linear model fits, and we calculated mean values for samples with repeat measurements.
+We assessed repeatability of RTL using the rptR package.
 
@@
 We also tested whether consistent telomere lengthening across our dataset using a modified version of the approach developed by Simons *et al.* [-@Simons2014].
-Briefly, this approach utilises samples with at least three telomere measurements to compare residual variance in telomere change over time with the overall change in telomere length between the first and least telomere measurements [@Simons2014].
+Briefly, this approach utilises samples with at least three telomere measurements to compare residual variance in telomere change over time with the overall change in telomere length between the first and last telomere measurements [@Simons2014].
 If, in samples that increase in length, the overall increase in telomere length exceeds the residual variance, then telomere lengthening cannot be explained by error [@Simons2014].
@@
 However, the noise associated with within and among plate measure may have resulted in a decrease in explanatory power, and possibly in a degree of type II error.
-Techniques for measuring telomere length with a greater degree of precision may probe helpful in future ecological studies of teloere dynamics, but at present there is still atrade-off between obtaining precise telomere measurements, and utilising the large sample sizes necessary for ecological study.
+Techniques for measuring telomere length with a greater degree of precision may probe helpful in future ecological studies of telomere dynamics, but at present there is still atrade-off between obtaining precise telomere measurements, and utilising the large sample sizes necessary for ecological study.
 

### 2017-06-05 Manuscript/Response_to_reviewers.Rmd
Edits

@@
 
-We now explain our randomisation approach in the methods, line XX:
+We now explain our randomisation approach in the methods, lines XX:
 
@@
 However, the noise associated with within and among plate measure may have resulted in a decrease in explanatory power, and possibly in a degree of type II error.
-Techniques for measuring telomere length with a greater degree of precision may probe helpful in future ecological studies of teloere dynamics, but at present there is still atrade-off between obtaining precise telomere measurements, and utilising the large sample sizes necessary for ecological study."
+Techniques for measuring telomere length with a greater degree of precision may probe helpful in future ecological studies of teloere dynamics, but at present there is still a trade-off between obtaining precise telomere measurements, and utilising the large sample sizes necessary for ecological study."
 }
@@
 
-RESPONSE: Again, the level of efficiency is lower than we would have liked, although again suts fairly squarely within that found the published literature (see Fairlie et al.
+RESPONSE: Again, the level of efficiency is lower than we would have liked, although again this value sits fairly squarely within that found in the published telomere literature (see Fairlie et al.
 2015, for a recent example), and often it is not even reported.
-We also note that this problem is not explected to be too drastic, because variaiton in efficiency is taken into account when calculating RTL from the Cq values.
+We also note that this problem is not explected to be too drastic, because variation in efficiency is taken into account when calculating RTL from the Cq values.
 And again, we suggest that because of our study design, random error of this kind makes the majority of our analyses conservative, and is accounted for in the analysis of telomere lengthening.
@@
 \textit{
-Using adult samples greater than one year old, we checked whether RTL was related to sample storage time, and found no evidence of such a relationship (estimate = -0.002, CIs = -0.007, 0.002).
+Using samples taken from adults greater than one year old, we checked whether RTL was related to sample storage time, and found no evidence of such a relationship (estimate = -0.002, CIs = -0.007, 0.002).
 }
@@
 However, on balance we agree that restricting this plot to juveniles is probably less biased.
-We have now altered the figure accordingly.
+We have now altered the figure and legend accordingly.
 
@@
 2014; Becker et al.2015).
-One problem with studying cohort effects is that it can be difficult to tease apart true cohort effects from effects that arise due to sample degradation with aim, and/or batch effects in telomere assays, although neither of these factors were a problem in our study.
+One problem with studying cohort effects is that it can be difficult to tease apart true cohort effects from effects that arise due to sample degradation with storage time, and/or batch effects in telomere assays, although neither of these factors were a problem in our study.
 Indeed, the long-term Seychelles warbler dataset has allowed us to show that temporal variation in telomere dynamics can occur over substantial time periods."}
@@
 \textit{"Such inconsistent changes in telomere length over lifespans could occur due to changes in the composition of cell types within individual samples, or due to the actual elongation of telomeres (Blackburn et al.
-1999(.
+1999).
 Determining the mechanism of these changes is essential for how we view telomeres as biomarkers of costs.
@@
 
-RESPONSE: We disagree here - it is the outcome of the molecular methods. We tried moving it but this disrupted the flow of the manuscript se have left it as it was.
+RESPONSE: We disagree here - it is the outcome of the molecular methods. We tried moving it but this disrupted the flow of the manuscript, so have left it as it was.
 
@@
 
-RESPONSE: These bave been added in response to an earlier comment.
+RESPONSE: These have been added in response to an earlier comment.
 
@@
 **Dear authors,**
+
 **You have an impressive dataset of longitudinal data and your work on telomere dynamics can be an important contribution to the current knowledge.
@@
 **Your introduction could be improved, first of all, by citing correctly.
-Also, there is some more recent literature on the topic, which you could cite.
-See more in comments on the introduction below.
-Further, I have some methodological questions.
-I would like to see more information on how samples were distributed over qPCR plates.
-Were samples of the same individual on the same plate?
-If not, how did you take plate ID into account in your statistical analyses?
-How exactly did you relate territory quality to RTL and delta RTL change?
-Only for birth year?
-Reading your abstract and aims in the introduction, I expected more.
-Your introduction could be improved, first of all, by citing correctly.
 Also, there is some more recent literature on the topic, which you could cite.
@@
 In my opinion, the ecological context in the discussion could be improved.
-More comments see below.
-I apologise for a delayed submission of my review, but hope that you find my comments helpful!**
-**
+More comments see below.**
+
+**I apologise for a delayed submission of my review, but hope that you find my comments helpful!**
 

### 2017-06-05 Manuscript/SW_early_life_telomeres_FULL.Rmd
Edits

@@
 Inter-plate repeatability of RTL, based on `r nrow(subset(ddL,Group == 'Within samples'))` samples measured at least twice at different time points, was `r round2(rptall$R)` (CI = `r round2(rptall$CI_emp[1])`, `r round2(rptall$CI_emp[2])`).
-Using adult samples greater than one year old, we checked whether RTL was related to sample storage time, and found no evidence of such a relationship (estimate = `r round3(fixef(RTL_ST)['StorageTime'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_ST)['StorageTime',],lessthan = F),collapse = ', ')`).
+Using samples taken from adults greater than one year old, we checked whether RTL was related to sample storage time, and found no evidence of such a relationship (estimate = `r round3(fixef(RTL_ST)['StorageTime'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_ST)['StorageTime',],lessthan = F),collapse = ', ')`).
 
@@
 Other studies have found chort effects but not discussed them in an ecological context [@Stier2014; @Becker2015].
-One problem with studying cohort effects is that it can be difficult to tease apart true cohort effects from effects that arise due to sample degradation with aim, and/or batch effects in telomere assays, although neither of these factors were a problem in our study.
+One problem with studying cohort effects is that it can be difficult to tease apart true cohort effects from effects that arise due to sample degradation with storage time, and/or batch effects in telomere assays, although neither of these factors were a problem in our study.
 Indeed, the long-term Seychelles warbler dataset has allowed us to show that temporal variation in telomere dynamics can occur over substantial time periods.

### 2017-06-05 Manuscript/Response_to_reviewers.Rmd
Edits

@@
 
-RESPONSE: NEED TO GO BACK TO RAW DATA TO DO
+RESPONSE: We have now included intra-plate repeatability estimates on lines XX. Samples were generally placed next to one another on the plate.
 
@@
 
-RESPONSE: No, it is not always the same time between sampling.
+RESPONSE: No, it is not always the same time between sampling, as we now make clear in the manuscript (see response to previous comment).
 And yes, part of the low repeatability could be explained by inter-plate error, as individual samples were fully randomised across plates.
 However, the within-individual repeatability is still far lower than the within-sample (across-plate) repeatability.
-We now add a caveat to the discussion to highlight this - lines XX:
+We now add a caveat to this effect, on lines XX:
+
+\begin{adjustwidth}{2cm}{2cm}
+
+\textit{"Consistent with a pattern of sporadic changes in telomere length with age, we found that within-individual telomere measurements were only weakly correlated.
+Although some of this low within-individual repeatability will occur die to measurement error, our within-sample repeatability was still much higher than our within-individual repeatability."}
+
+\end{adjustwidth}
 

### 2017-06-05 Manuscript/SW_early_life_telomeres_FULL.Rmd
Edits

@@
 Efficiencies (mean $\pm$ s.d.) for our telomere and GAPDH reactions were `r round2(mean(dd$EffTelomere))` $\pm$ `r round2(sd(dd$EffTelomere))` and `r round2(mean(dd$EffGAPDH))` $\pm$ `r round2(sd(dd$EffGAPDH))` respectively.
+Intra-plate repeatability was 0.74 (CI = 0.74,0.75) and 0.73 (CI = 0.71,0.74) for the GAPDH and Telomere Cq values, respecitvely.
 Inter-plate repeatability of RTL, based on `r nrow(subset(ddL,Group == 'Within samples'))` samples measured at least twice at different time points, was `r round2(rptall$R)` (CI = `r round2(rptall$CI_emp[1])`, `r round2(rptall$CI_emp[2])`).
@@
 Here, we explicitly compare intra-individual variation among samples to variation among sample replicates, on a large scale.
-Our results suggest that qPCR measurement error alone cannot explain observed increases in RTL observed within individuals.
+Our results suggest that despite the substantial levels of qPCR measurement error in our study, error alone cannot explain observed increases in RTL observed within individuals.
 
@@
 Consistent with a pattern of sporadic changes in telomere length with age, we found that within-individual telomere measurements were only weakly correlated.
-These findings are in contrast to other avian studies in which within-individual telomere length measurements were highly consistent, and individual-level telomere shortening occurred throughout the juvenile period and into adulthood [@Heidinger2012; @Boonekamp2014].
+Although some of this low within-individual repeatability will occur die to measurement error, our within-sample repeatability was still much higher than our within-individual repeatability.
+Such a low value of within-individual repeatability in telomere length is in contrast to other avian studies in which within-individual telomere length measurements were highly consistent, and individual-level telomere shortening occurred throughout the juvenile period and into adulthood [@Heidinger2012; @Boonekamp2014].
 However, the lifelong telomere dynamics found in Seychelles warblers are strikingly similar to those found in Soay sheep [@Fairlie2016].

### 2017-06-06 Manuscript/Response_to_reviewers.Rmd
Edits

@@
 However, the noise associated with within and among plate measure may have resulted in a decrease in explanatory power, and possibly in a degree of type II error.
-Techniques for measuring telomere length with a greater degree of precision may probe helpful in future ecological studies of teloere dynamics, but at present there is still a trade-off between obtaining precise telomere measurements, and utilising the large sample sizes necessary for ecological study."
+Techniques for measuring telomere length with a greater degree of precision may probe helpful in future ecological studies of telomere dynamics, but at present there is still a trade-off between obtaining precise telomere measurements, and utilising the large sample sizes necessary for ecological study."
 }
@@
 2015, for a recent example), and often it is not even reported.
-We also note that this problem is not explected to be too drastic, because variation in efficiency is taken into account when calculating RTL from the Cq values.
+We also note that this problem is not expected to be too drastic, because variation in efficiency is taken into account when calculating RTL from the Cq values.
 And again, we suggest that because of our study design, random error of this kind makes the majority of our analyses conservative, and is accounted for in the analysis of telomere lengthening.
@@
 "The majority of individuals are ringed (with an individually numbered metal ring and unique combination of colour rings) within the first year of life, and so are of known age.
-We aged all birds using information on eye colour at first capture (Kondeur et al.
+We aged all birds using information on eye colour at first capture (Komdeur et al.
 1991) and previous capture history (Richardson et al.
 2003).
-Within the first year of life, birds are classified as nestlings less than one month old (rounded to one month for analyses), fledglings less than six month olds (rounded to six months) or subadults up to one year old (rounded to 10 months).
+Within the first year of life, birds are classified as nestlings less than one month old (rounded to one month for analyses), fledglings less than six month old (rounded to six months) or subadults up to one year old (rounded to 10 months).
 Ages for adult birds were rounded to the nearest year."
@@
 
-RESPONSE: Storage in ethanol is standardd procedure for avian blood samples, and allows DNA to be maintained at high quality for many years.
-We checked DNA integrity for a very large number of samples, and found no evidence of DNA degration in older samples.
+RESPONSE: Storage in ethanol is standard procedure for avian blood samples, and allows DNA to be maintained at high quality for many years.
+We checked DNA integrity for a very large number of samples, and found no evidence of DNA degradation in older samples.
 We now include two gel images to illustrate this in the supplementary material (Fig.
@@
 
-\textit{"Although a few studies have shown that temporal variation in telomere dynamicsoccurs in natural populations, these have been limited in the number of seasons they cover (Mizutani et al.
+\textit{"Although a few studies have shown that temporal variation in telomere dynamics occurs in natural populations, these have been limited in the number of seasons they cover (Mizutani et al.
 2013; Watson, Bolton and Monaghan 2015; Fairlie et al.
 2016).
-Other studies have found chort effects but not discussed them in an ecological context (Stier et al.
+Other studies have found cohort effects but not discussed them in an ecological context (Stier et al.
 2014; Becker et al.2015).
@@
 
-RESPONSE: We are not sure that "worrying" is the correct phrase to use, as we anticpiate that other studies in natural population will find similar patterns.
+RESPONSE: We are not sure that "worrying" is the correct phrase to use, as we anticipate that other studies in natural population will find similar patterns.
 Further, we disagree with the sentiment that a biomarker must be consistent over individual lifespans to be useful.

### 2017-06-07 Manuscript/Response_to_reviewers.Rmd
Edits

@@
 2016, cited).
+However, we appreciate that plate effects could still occur (e.g.
+because of an effect on the calibrator sample), so we have now included plate ID as a random effect in our models.
+Our results are almost identical, confirming that plate effects have not biased our results.
+We have updated the methods sections in the relevant places to describe the addition of this random effect.
 
@@
 
-RESPONSE: We agree that while there is convincing evidence for cohort effects on telomere length, the evidence for cohort effects on telomere dynamics is much weaker.
-We have therefore removed references to this effect throughout.
+RESPONSE: Interestingly, this is the one effect that has changed from  including a plate ID random effect in our analyses, with delta AIC now ~11.
+However, we do not wish to place any strong conclusions on this analysis.
+We agree that while there is convincing evidence for cohort effects on telomere length, the evidence for cohort effects on telomere dynamics is weaker.
+We have therefore removed references to this effect throughout the discussion, and allow readers to interpret the effect as they wish.
 
@@
 
-RESPONSE: As samples are normalised within plates, including a plate ID random effect would not make sense. The error across plates is random. See also response to reviewer 1.
+RESPONSE: We have now included plate ID as a random effect, and this has not changed any of our results. See also response to reviewer 1.
 

### 2017-06-07 Manuscript/SW_early_life_telomeres_FULL.Rmd
Edits

@@
 @Frenck1998; @Heidinger2012].
-*In vitro* research has shown tha telomere shortening can be accelerated by oxidative stress  [@VonZglinicki2002], which can be elevated due to many environmental factors.
+*In vitro* research has shown that telomere shortening can be accelerated by oxidative stress  [@VonZglinicki2002], which can be elevated due to many environmental factors.
 There is evidence from humans, and from captive and wild animal populations, that telomere shortening is influenced by the conditions experienced during both early life and adulthood  [@Price2013; @Monaghan2014; @Nettle2015; @Reichert2015].
@@
 We aged all birds using information on eye colour at first capture [@Komdeur1991a] and previous capture history [@Richardson2003b].
-Within the first year of life, birds are classified as nestlings less than one month old (rounded to one month for analyses), fledglings less than six month olds (rounded to six months) or subadults up to one year old (rounded to 10 months).
+Within the first year of life, birds are classified as nestlings less than one month old (rounded to one month for analyses), fledglings less than six month old (rounded to six months) or subadults up to one year old (rounded to 10 months).
 Ages for adult birds were rounded to the nearest year.
@@
 Note that we do not carry out full model selection or model averaging here, as our aim was to compare a set of specifically defined models.
-For random effects we included individual ID, and catch year.
+For random effects we included individual ID, catch year and qPCR plate ID.
 Models were compared using AIC with correction for finite sample size [AICc; @Hurvich1989].
@@
 We first created a full model with RTL as a response variable, alongside the following explanatory variables: log age (based on the RTL and age analysis; see results), tarsus length at capture, body mass at capture, sex, insect abundance in sampling season, territory quality in sampling season, island-wide population density in sampling season (an annual measure estimated from the summer breeding census), territory group size in sampling season, and the number of helping subordinate birds present in the territory in the sampling season.
-The random effects structure was informed by the analysis of telomere dynamics and age (see results): we included individual ID, cohort ID, and a random slope of log age among cohorts (to allow the effect of age on RTL to vary among cohorts).
+The random effects structure was informed by the analysis of telomere dynamics and age (see results): we included individual ID, qPCR plate ID, cohort ID, and a random slope of log age among cohorts (to allow the effect of age on RTL to vary among cohorts).
 We report model estimates and confidence intervals for all effects included in the full model.
@@
 Efficiencies (mean $\pm$ s.d.) for our telomere and GAPDH reactions were `r round2(mean(dd$EffTelomere))` $\pm$ `r round2(sd(dd$EffTelomere))` and `r round2(mean(dd$EffGAPDH))` $\pm$ `r round2(sd(dd$EffGAPDH))` respectively.
-Intra-plate repeatability was 0.74 (CI = 0.74,0.75) and 0.73 (CI = 0.71,0.74) for the GAPDH and Telomere Cq values, respecitvely.
+Intra-plate repeatability was 0.74 (CI = 0.74,0.75) and 0.73 (CI = 0.71,0.74) for the GAPDH and Telomere Cq values, respectively.
 Inter-plate repeatability of RTL, based on `r nrow(subset(ddL,Group == 'Within samples'))` samples measured at least twice at different time points, was `r round2(rptall$R)` (CI = `r round2(rptall$CI_emp[1])`, `r round2(rptall$CI_emp[2])`).
@@
 Although a few studies have shown that temporal variation in telomere dynamics occurs in natural populations, these have been limited in the number of seasons they cover [@Mizutani2013; @Watson2015; @Fairlie2016].
-Other studies have found chort effects but not discussed them in an ecological context [@Stier2014; @Becker2015].
+Other studies have found cohort effects but not discussed them in an ecological context [@Stier2014; @Becker2015].
 One problem with studying cohort effects is that it can be difficult to tease apart true cohort effects from effects that arise due to sample degradation with storage time, and/or batch effects in telomere assays, although neither of these factors were a problem in our study.
@@
 However, the noise associated with within and among plate measure may have resulted in a decrease in explanatory power, and possibly in a degree of type II error.
-Techniques for measuring telomere length with a greater degree of precision may probe helpful in future ecological studies of telomere dynamics, but at present there is still atrade-off between obtaining precise telomere measurements, and utilising the large sample sizes necessary for ecological study.
+Techniques for measuring telomere length with a greater degree of precision may probe helpful in future ecological studies of telomere dynamics, but at present there is still a trade-off between obtaining precise telomere measurements, and utilising the large sample sizes necessary for ecological study.
 
@@
 CIs <- CIwald(RTL_Full2)
-t1a <- data.frame(myx = row.names(CIs)[6:nrow(CIs)],
+t1a <- data.frame(myx = row.names(CIs)[7:nrow(CIs)],
                   Estimate = fixef(RTL_Full2),
-                  X2.5. = CIs[6:nrow(CIs),1],
-                  X97.5. = CIs[6:nrow(CIs),2],
+                  X2.5. = CIs[7:nrow(CIs),1],
+                  X97.5. = CIs[7:nrow(CIs),2],
                   row.names = NULL,

### 2017-06-07 Manuscript/SW_early_life_telomeres_SOM.Rmd
Edits

@@
 classoption: landscape
+header-includes: \usepackage{graphicx}
 ---
@@
                message=FALSE,
+               cache = TRUE,
                fig.path='../Figures/',
@@
 
-**Figure S1** Examples of DNA extractions from older and more recent Seychelles warbler samples. We found no evidence for differences in DNA integrity with sample age.
+\includegraphics[width = 10cm]{Fig_S1.png}
+
+**Figure S1** Examples of DNA extractions from older Seychelles warbler samples (sample year in brackets). We found no evidence for differences in DNA integrity with sample age.
 

### 2017-06-07 Scripts/SW juvenile telomeres IPC models.r

@@ -13,16 +13,16 @@ RTL_ST <- lmer(RTL ~ StorageTime + (1|BirdID) + (1|PlateID),
 RTL_Age <- lmer(RTL ~ Cohort + Agemonths + LogAge + FAge + Age2 + (1|BirdID) + (1|CatchYear) + (1|PlateID),
                 data=dd,
                 REML = FALSE)
-RTL_LogAge <- lmer(RTL ~ LogAge + Cohort + (1|BirdID) + (1|CatchYear),
+RTL_LogAge <- lmer(RTL ~ LogAge + Cohort + (1|BirdID) + (1|CatchYear)+ (1|PlateID),
                 data=dd,
                 REML = FALSE)
-RTL_LogAgeXcohort <- lmer(RTL ~ LogAge*Cohort + (1|BirdID) + (1|CatchYear),
+RTL_LogAgeXcohort <- lmer(RTL ~ LogAge*Cohort + (1|BirdID) + (1|CatchYear)+ (1|PlateID),
                    data=dd,
                    REML = FALSE)
-RTL_DeltaAge <- lmer(RTL ~ MeanAge + Cohort + DeltaAge + DeltaLogAge + DeltaAge2 + (1|BirdID) + (1|CatchYear),
+RTL_DeltaAge <- lmer(RTL ~ MeanAge + Cohort + DeltaAge + DeltaLogAge + DeltaAge2 + (1|BirdID) + (1|CatchYear)+ (1|PlateID),
                 data=dd3,
                 REML = FALSE)
-RTL_DeltaLogAge <- lmer(RTL ~ DeltaLogAge + (1|BirdID) + (1|CatchYear),
+RTL_DeltaLogAge <- lmer(RTL ~ DeltaLogAge + (1|BirdID) + (1|CatchYear)+ (1|PlateID),
                    data=dd3,
                    REML = FALSE)
 DeltaRTL_LogAge <- lmer(DeltaRTL ~ LogAge + (1|BirdID) + (1|CatchYear),
@@ -40,7 +40,7 @@ RTL_RTL1 <- lmer(RTL1 ~ RTL + LogAge + (1|BirdID) + (1|LayYear),
 ddNA <- subset(subset(dd,!is.na(Insect)),!is.na(TQ))
 dd3NA <- subset(subset(dd3,!is.na(Insect)),!is.na(TQ))
 
[…]

### 2017-06-07 Manuscript/SW_early_life_telomeres_FULL.Rmd
Edits

@@
                message=FALSE,
+               cache = TRUE,
                fig.width=8,

### 2017-06-15 Manuscript/Response_to_reviewers.Rmd
Edits

@@
 RESPONSE: The reviewer is correct that the estimate of repeatability is the intraclass correlation coefficient (this is made clear in the rptR documentation, which we cite).
-We agree with the reviewer in the sense that repeatability is not as high as we would have hoped, although we note that in the context of the telomere literature, a repeatability estimate of 0.68 is not especially low.
+Our understanding of the ICC is that it is a measurement of the proportion of variance explained by sample - so 32% (rather than 50%) of the vatiation is linked to measurement error.
+However, we agree with the reviewer in the sense that repeatability is not as high as we would have hoped, although we note that in the context of the telomere literature, a repeatability estimate of 0.68 is not especially low.
 However, because of how the samples were run, we are confident that any issues of repeatability have not led to any kind of bias in our results, and in fact make our analyses conservative.
@@
 because of an effect on the calibrator sample), so we have now included plate ID as a random effect in our models.
-Our results are almost identical, confirming that plate effects have not biased our results.
-We have updated the methods sections in the relevant places to describe the addition of this random effect.
+Our results are almost identical, with all the reported significant effects remaining.
+The only change is that the cohort x age effect on telomere length is now stronger (see response to reviewer 3).
+Overall, we can be very confident that plate effects have not biased our results.
+We have updated the methods sections in the relevant places to describe the addition of the plate ID random effect.
 
-With regards to the observed telomere lengthening, it is very possible that inter-plate variability accounts some of the observed lengthening in our dataset, but it cannot account for the amount and extent of lengthening we find.
+With regards to the observed telomere lengthening, it is very possible that inter-plate variability accounts for some of the observed lengthening in our dataset, but it cannot account for the amount and extent of lengthening we find.
 Indeed, this is exactly what Figure 2B shows.
@@
 However, the noise associated with within and among plate measure may have resulted in a decrease in explanatory power, and possibly in a degree of type II error.
-Techniques for measuring telomere length with a greater degree of precision may probe helpful in future ecological studies of telomere dynamics, but at present there is still a trade-off between obtaining precise telomere measurements, and utilising the large sample sizes necessary for ecological study."
+Techniques for measuring telomere length with a greater degree of precision may prove helpful in future ecological studies of telomere dynamics, but at present there is still a trade-off between obtaining precise telomere measurements, and utilising the large sample sizes necessary for ecological study."
 }
@@
 \textit{
-"The majority of individuals are ringed (with an individually numbered metal ring and unique combination of colour rings) within the first year of life, and so are of known age.
+"The majority (96%) of individuals are ringed (with an individually numbered metal ring and unique combination of colour rings) within the first year of life, and so are of known age.
 We aged all birds using information on eye colour at first capture (Komdeur et al.
@@
 RESPONSE: We are not sure that "worrying" is the correct phrase to use, as we anticipate that other studies in natural population will find similar patterns.
-Further, we disagree with the sentiment that a biomarker must be consistent over individual lifespans to be useful.
+Further, we disagree with the sentiment that a biomarker must be consistent over individual lifespans to be useful - rather, a useful barker needs to change with ecological context.
 We feel that we cover this point in the discussion, lines XX:
@@
 
-RESPONSE: We are pleased that the reviewer is generally positive about are manuscript.
-We think it a little harsh to imply that we have not "cited correctly", but are grateful for the suggested changes.
-We respond to each of these questions/comments below.
+RESPONSE: We are pleased that the reviewer is generally positive about our manuscript.
+We thank the reviewer for pointing out areas where our citation of previous work was incomplete - although we believe our original citations were generally correct, we agree that there were one or two mistakes, and that it is important to provide the reader with a complete context for our work.
+We respond to each of the individual questions/comments below.
 
@@
 
-RESPONSE: We have changed the citations here.
+RESPONSE: We have removed the Monaghan and Haussmann reference and added in Muezlinner et al. 2013 as a replacement.
 
@@
 2003).
-The association between senescence and telomere length has inspired a great deal of recent research into telomere evolutionary ecology, and relationships between telomere dynamics and survival or lifespan have been documented in wild population of several species (Haussmann and Marchetto 2010; Barrett et al.
+The association between telomere length and senescence has inspired a great deal of recent research into telomere evolutionary ecology, and relationships between telomere dynamics and survival or lifespan have been documented in wild populations of several species (Haussmann and Marchetto 2010; Barrett et al.
 2013; Stier et al.
@@
 We feel that an in-depth explanation of how and why territory quality varies over space and time is outside the scope of this manuscript, especially given i) the word limit, and ii) the fact that territory quality is not related to RTL.
+However we do now include a reference to recent book chapter (Komdeur et al.
+2016) which explains how and why territory quality varies on Cousin Island.
+Our updated section on territory quality and indect abundance is now on lines XX:
+
+\begin{adjustwidth}{2cm}{2cm}
+
+\textit{"Where territory quality estimates were not available for a specific year we used the average value for that territory across years (Hammers et al.
+2013; see Komdeur et al.
+2016 for an explanation of how territory quality varies on Cousin Island).
+Cousin is subject to considerable intra- and inter-annual variation in rainfall and, consequently, insect availability (Komdeur et al.
+2016).
+Such island-wide temporal variation may override the effects of variation in individual territory quality across the island.
+As an estimate of seasonal variation in food availability, we calculated an index of the abundance of insects across the entire island during each main breeding season (referred to hereafter as 'insect abundance').
+This index is calculated as the mean number of insects found per unit leaf area over all monthly surveys carried out on the island in a main breeding season.
+"}
+
+\end{adjustwidth}
 
@@
 
-RESPONSE: We have now included plate ID as a random effect, and this has not changed any of our results. See also response to reviewer 1.
+RESPONSE: We have now included plate ID as a random effect, and this has had almost no effect on our results. See response to reviewer 1.
 
@@
 See Bebbington et al.
-2016 Mol Ecol for a discussion of why were using relative telomere length.
+2016 Mol Ecol for a discussion of why we are using relative telomere length.
 
@@
 
-RESPONSE: We are not sure exactly what the reviewer wants us to do here.
-We do compare cross-sectional and longitudinal telomere shortening, in the sense that we demonstrate how both measures shorten with log age.
-We are happy to add more if there is space, and if we are given a clarification.
+RESPONSE: We now include a formal test of the within and between individual slopes, as described in van de Pol and Wright (2009), and find that there is no difference between cross-sectional and longitudinal telomere shortening - lines XX:
 
@@
 \textit{"Consistent with a pattern of sporadic changes in telomere length with age, we found that within-individual telomere measurements were only weakly correlated.
-Although some of this low within-individual repeatability will occur die to measurement error, our within-sample repeatability was still much higher than our within-individual repeatability."}
+Although some of this low within-individual repeatability will occur due to measurement error, our within-sample repeatability was still much higher than our within-individual repeatability."}
 

### 2017-06-15 Manuscript/SW_early_life_telomeres_FULL.Rmd
Edits

--- 2017-06-07 Manuscript/SW_early_life_telomeres_FULL.Rmd
@@
 source('Scripts/SW juvenile telomeres IPC models.r')
-
 ```
@@
 When telomeres become critically short, cells senesce [@Campisi2003], and the accumulation of these cells has been suggested to result in organismal senescence and death [@Wong2003].
-The association between senescence and telomere length has inspired a great deal of recent research into telomere evolutionary ecology, and relationships between telomere dynamics and survival or lifespan have been documented in wild population of several species [@Haussmann2010; @Barrett2013a; @Stier2015].
+The association between telomere length and senescence has inspired a great deal of recent research into telomere evolutionary ecology, and relationships between telomere dynamics and survival or lifespan have been documented in wild populations of several species [@Haussmann2010; @Barrett2013a; @Stier2015].
 As yet, however, there is little direct evidence that the relationship between telomere dynamics and survival is causal [@Simons2015].
@@
 
-The majority of individuals are ringed (with an individually numbered metal ring and unique combination of colour rings) within the first year of life, and so are of known age.
+The majority (96%) of individuals are ringed (with an individually numbered metal ring and unique combination of colour rings) within the first year of life, and so are of known age.
 We aged all birds using information on eye colour at first capture [@Komdeur1991a] and previous capture history [@Richardson2003b].
@@
 Territory quality is calculated based on territory size, foliage cover and insect abundance [@Komdeur1992].
-Where territory quality estimates were not available for a specific year we used the average value for that territory across years [@Hammers2013].
-Cousin is subject to considerable intra- and inter-annual variation in rainfall and, consequently, insect availability.
+Where territory quality estimates were not available for a specific year we used the average value for that territory across years [@Hammers2013; see @Komdeur2016 for an explanation of how territory quality varies on Cousin Island].
+Cousin is subject to considerable intra- and inter-annual variation in rainfall and, consequently, insect availability [@Komdeur2016].
 Such island-wide temporal variation may override the effects of variation in individual territory quality across the island.
@@
 RTL decreased with $\Delta$log age (estimate = `r round3(fixef(RTL_DeltaLogAge)['DeltaLogAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_DeltaLogAge)['DeltaLogAge',],lessthan = F),collapse = ', ')`), confirming that within-individual telomere shortening occurs across the Seychelles warbler dataset.
+Further, we found no evidence that within and between individual slopes of telomere shortening varied (``r round3(fixef(RTL_LogAge2)['MeanAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_LogAge2)['MeanAge',],lessthan = F),collapse = ', ')`; see Van de Pol and Wright 2009)
 
@@
 Consistent with a pattern of sporadic changes in telomere length with age, we found that within-individual telomere measurements were only weakly correlated.
-Although some of this low within-individual repeatability will occur die to measurement error, our within-sample repeatability was still much higher than our within-individual repeatability.
+Although some of this low within-individual repeatability will occur due to measurement error, our within-sample repeatability was still much higher than our within-individual repeatability.
 Such a low value of within-individual repeatability in telomere length is in contrast to other avian studies in which within-individual telomere length measurements were highly consistent, and individual-level telomere shortening occurred throughout the juvenile period and into adulthood [@Heidinger2012; @Boonekamp2014].
@@
 However, the noise associated with within and among plate measure may have resulted in a decrease in explanatory power, and possibly in a degree of type II error.
-Techniques for measuring telomere length with a greater degree of precision may probe helpful in future ecological studies of telomere dynamics, but at present there is still a trade-off between obtaining precise telomere measurements, and utilising the large sample sizes necessary for ecological study.
+Techniques for measuring telomere length with a greater degree of precision may prove helpful in future ecological studies of telomere dynamics, but at present there is still a trade-off between obtaining precise telomere measurements, and utilising the large sample sizes necessary for ecological study.
 

### 2017-06-15 Scripts/SW juvenile telomeres IPC models.r

@@ -16,6 +16,9 @@ RTL_Age <- lmer(RTL ~ Cohort + Agemonths + LogAge + FAge + Age2 + (1|BirdID) + (
 RTL_LogAge <- lmer(RTL ~ LogAge + Cohort + (1|BirdID) + (1|CatchYear)+ (1|PlateID),
                 data=dd,
                 REML = FALSE)
+RTL_LogAge2 <- lmer(RTL ~ LogAge + MeanAge + Cohort + (1|BirdID) + (1|CatchYear)+ (1|PlateID),
+                   data=dd,
+                   REML = FALSE)
 RTL_LogAgeXcohort <- lmer(RTL ~ LogAge*Cohort + (1|BirdID) + (1|CatchYear)+ (1|PlateID),
                    data=dd,
                    REML = FALSE)

### 2017-06-15 Manuscript/Response_to_reviewers.Rmd
Edits

@@
 
+\begin{adjustwidth}{2cm}{2cm}
+
+\textit{"Further, we found no evidence that within and between individual slopes of telomere shortening varied (estimate = -0.003, CIs = -0.008, 0.003; see Van de Pol and Wright 2009), suggesting that there was no difference between cross-sectional and longitudinal telomere shortening with age."}
+
+\end{adjustwidth}
+
 \bigskip
@@
 
-RESPONSE: As stated in the discussion, we feel that this is an important area for future research, and we hope that it will be a whole manuscript on its own.
+RESPONSE: As stated in the discussion (lines XX), we feel that this is an important area for future research, and we hope that it will be a whole manuscript on its own.
 We are already pushing the page limit for Journal of Animal Ecology, so adding in an entire new set of analyses is just not possible.
-But watch this space!
 

### 2017-06-15 Manuscript/SW_early_life_telomeres_FULL.Rmd
Edits

@@
 RTL decreased with $\Delta$log age (estimate = `r round3(fixef(RTL_DeltaLogAge)['DeltaLogAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_DeltaLogAge)['DeltaLogAge',],lessthan = F),collapse = ', ')`), confirming that within-individual telomere shortening occurs across the Seychelles warbler dataset.
-Further, we found no evidence that within and between individual slopes of telomere shortening varied (``r round3(fixef(RTL_LogAge2)['MeanAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_LogAge2)['MeanAge',],lessthan = F),collapse = ', ')`; see Van de Pol and Wright 2009)
+Further, we found no evidence that within and between individual slopes of telomere shortening varied (estimate = `r round3(fixef(RTL_LogAge2)['MeanAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_LogAge2)['MeanAge',],lessthan = F),collapse = ', ')`; see Van de Pol and Wright 2009), suggesting that there was no difference between cross-sectional and longitudinal telomere shortening with age.
 

### 2017-06-23 Manuscript/Response_to_reviewers.Rmd
Edits

@@
 Our understanding of the ICC is that it is a measurement of the proportion of variance explained by sample - so 32% (rather than 50%) of the vatiation is linked to measurement error.
-However, we agree with the reviewer in the sense that repeatability is not as high as we would have hoped, although we note that in the context of the telomere literature, a repeatability estimate of 0.68 is not especially low.
-However, because of how the samples were run, we are confident that any issues of repeatability have not led to any kind of bias in our results, and in fact make our analyses conservative.
+However, we agree with the reviewer in the sense that repeatability is not as high as we would have hoped.
+The reduction in repeatability has arisen because of the long-term and large-scale nature of our study.
+The samples have been run over a period of many years, over which time reagents and materials change, machines break, and lab conditions vary.
+This has never been an issue in qPCR studies to date, becuase most have run a small number of samples, all at the same time, and all with the exact same set of reagents.
+However, we strongly believe that the benefits of having a long-term, large-scale telomere study far outweigh the costs of a slight reduction in repeatability.
+Further, because of how the samples were run, we are confident that any issues of repeatability have not led to any kind of bias in our results, and in fact make our analyses conservative.
 We respond to this point in detail, and outline how we clarified our discussion, in our response to the next comment.
@@
 2014).
-Measurement error is unlikely to be a cause of type I error in our study, because we were careful to randomise all samples across qPCR plates, and normalise RTL estimates across plates.
-However, the noise associated with within and among plate measure may have resulted in a decrease in explanatory power, and possibly in a degree of type II error.
-Techniques for measuring telomere length with a greater degree of precision may prove helpful in future ecological studies of telomere dynamics, but at present there is still a trade-off between obtaining precise telomere measurements, and utilising the large sample sizes necessary for ecological study."
+The measurement error in our study has most likely arisen because samples were run aver a period of several years, during which reagents, consumables and personell all change.
+Compared to a set-up where a small amount of samples are all run at the same time, it is unsurprising that we have a higher error rate.
+Techniques for measuring telomere length with a greater degree of precision may prove helpful in future ecological studies of telomere dynamics, but at present there is still a trade-off between obtaining precise telomere measurements, and utilising the large sample sizes necessary for ecological study.
+Further, measurement error is unlikely to be a cause of type I error in our study, because we were careful to randomise all samples across qPCR plates, and normalise RTL estimates across plates.
+However, the noise associated with within and among plate measure may have resulted in a decrease in explanatory power, and possibly in a degree of type II error."
 }
@@
 \textit{
-"The majority (96%) of individuals are ringed (with an individually numbered metal ring and unique combination of colour rings) within the first year of life, and so are of known age.
+"The majority (96\%) of individuals are ringed (with an individually numbered metal ring and unique combination of colour rings) within the first year of life, and so are of known age.
 We aged all birds using information on eye colour at first capture (Komdeur et al.
@@
 2014; Becker et al.2015).
-One problem with studying cohort effects is that it can be difficult to tease apart true cohort effects from effects that arise due to sample degradation with storage time, and/or batch effects in telomere assays, although neither of these factors were a problem in our study.
+One problem with studying cohort effects is that it can be difficult to tease apart true cohort effects from effects that may arise if samples degrade with storage time, and/or batch effects in telomere assays, although neither of these factors were a problem in our study.
 Indeed, the long-term Seychelles warbler dataset has allowed us to show that temporal variation in telomere dynamics can occur over substantial time periods."}
@@
 RESPONSE: We are not sure that "worrying" is the correct phrase to use, as we anticipate that other studies in natural population will find similar patterns.
-Further, we disagree with the sentiment that a biomarker must be consistent over individual lifespans to be useful - rather, a useful barker needs to change with ecological context.
+Further, we disagree with the sentiment that a biomarker must be consistent over individual lifespans to be useful - rather, a useful marker needs to change with ecological context.
 We feel that we cover this point in the discussion, lines XX:
@@
 
-**Lines 337-338: Such change in blood cell types would be quite convincing for mammals since “mammalian people” are measuring telomeres from different kind of WBCs and also immature nucleated RBCs..
+**Lines 337-338: Such change in blood cell types would be quite convincing for mammals since “mammalian people” are measuring telomeres from different kind of WBCs and also immature nucleated RBCs.
 However, in birds more than 99% of the cells in the blood are RBCs, therefore a change in blood cell types is quite unlikely.

### 2017-06-23 Manuscript/SW_early_life_telomeres_FULL.Rmd
Edits

@@
 Other studies have found cohort effects but not discussed them in an ecological context [@Stier2014; @Becker2015].
-One problem with studying cohort effects is that it can be difficult to tease apart true cohort effects from effects that arise due to sample degradation with storage time, and/or batch effects in telomere assays, although neither of these factors were a problem in our study.
+One problem with studying cohort effects is that it can be difficult to tease apart true cohort effects from effects that may arise if samples degrade with storage time, and/or batch effects in telomere assays, although neither of these factors were a problem in our study.
 Indeed, the long-term Seychelles warbler dataset has allowed us to show that temporal variation in telomere dynamics can occur over substantial time periods.
@@
 Measurement error is one factor that is an issue in our study, and other studies that use qPCR to measure telomere length [@Nussey2014].
-Measurement error is unlikely to be a cause of type I error in our study, because we were careful to randomise all samples across qPCR plates, and normalise RTL estimates across plates.
+The measurement error in our study has most likely arisen because samples were run aver a period of several years, during which reagents, consumables and personell all change.
+Compared to a set-up where a small amount of samples are all run at the same time, it is unsurprising that we have a higher error rate.
+Techniques for measuring telomere length with a greater degree of precision may prove helpful in future ecological studies of telomere dynamics, but at present there is still a trade-off between obtaining precise telomere measurements, and utilising the large sample sizes necessary for ecological study.
+Further, measurement error is unlikely to be a cause of type I error in our study, because we were careful to randomise all samples across qPCR plates, and normalise RTL estimates across plates.
 However, the noise associated with within and among plate measure may have resulted in a decrease in explanatory power, and possibly in a degree of type II error.
-Techniques for measuring telomere length with a greater degree of precision may prove helpful in future ecological studies of telomere dynamics, but at present there is still a trade-off between obtaining precise telomere measurements, and utilising the large sample sizes necessary for ecological study.
 

### 2017-06-24 Manuscript/Response_to_reviewers.Rmd
Home edits

@@
 
+#Associate editor comments
+
+---
+
+\bigskip
+
+**This manuscript describes spatio-temporal variation in telomere dynamics in the cooperatively breeding Seychelles warbler.
+It uses an impressive long-term longitudinal dataset of telomere lengths (obtained by qPCR) of >400 individuals and relates telomere length and dynamics to age, sex and various ecological variables, such as yearly insect availability and density, local territory quality and group size and the number of helpers during reproduction.**
+
+**The manuscript has now been read by three reviewers and myself.
+All of us are very impressed with the dataset and questions asked and answered in this manuscript.
+Two of the reviewers, however, have some important methodological questions, and some suggestions for updating the bibliography of the manuscript.
+I agree with their assessment, and believe that when the authors thoroughly clarify these issues, their paper will make an important contribution to the field and will be widely read and cited.**
+
+RESPONSE: We thank the editor and reviewers for their positive and constructive comments.
+We have now addressed the methodological questions raised by reviewers one and three, and have updated several of our analyses in line with their suggestions.
+We have also performed additional analyses to show that sample storage time does not affect telomere length, and included a discussion on repeatability.
+While almost none of our results have changed, we believe that incorporating the reviewer comments has substantially improved the clarity, readability and rigour of our work.
+Finally, we have updated our bibliography, in accordance with the reviewer comments.
+
+\newpage
+
 #Reviewer 1
@@
 
-RESPONSE: We thank Dr Stier for his constructive comments, and are pleased that he believes our work to be novel and worthy of publication.
+RESPONSE: We thank Dr Stier for his comments, and are pleased that he believes our work to be novel and worthy of publication.
 We agree that the main issues he raises were not properly addressed/explained, and respond to these points in detail below.
@@
 However, as we argue above, we do not think it is likely that the pattern is a result of sample storage/DNA integrity.
-Past discussions with the Nussey group suggests that they are also confident their telomere assay is not negatively affected by long-term sample storage.
+We have no evidence from our samples that there is sample degradiation with respect to DNA integrity or telomere length (see above) - something confirmed by other studies (see Reichert et al.
+2017; Oecologia)
 
@@
 
-\textit{"Our study adds to the substantial body of literature from humans and wild animals showing that telomere length decreases with age, and that this decrease is most rapid in early life - most likely as a consequence of cellular division..."}
+\textit{"Our study adds to the substantial body of literature from humans and wild animals showing that telomere length decreases with age, and that this decrease is most rapid in early life - most likely as a consequence of the much more rapid rate of cellular division that occurs during the growth phase, but also perhaps higher levels of cellular stress during development..."}
 
@@
 
-RESPONSE: We have removed the Heidinger reference, leaving two citations in - we felt this was sufficient.
+RESPONSE: We have now removed the Heidinger reference and added in Stier et al. 2014.
 
@@
 
-RESPONSE: We disagree here - Simons 2015 *is* a meta-analysis that shows that there is little evidence that the relationship between telomeres and survival is causal.
+RESPONSE: We agree that many such studies exist, and we do first give three cited examples of the relationship between telomere length and survival.
+We then cite the Simons paper (note that this is a meta-analysis) to highlight the point that we are not yet sure whether the relationship between telomeres and survival is causal.
 
@@
 
-\textit{"Note that the follow-up time between samples varied among individuals, and that individuals are not always caught at exactly the same time of year."}
+\textit{"Individuals were not always caught at the same time of year, but within a 3 month breeding season window."}
 
@@
 
+Not also that this analysis was done in one of the original Seychelles warbler telomere studies (Barrett et al. 2013; Mol. Ecol.)
+
 \end{adjustwidth}
@@
 RESPONSE: No, it is not always the same time between sampling, as we now make clear in the manuscript (see response to previous comment).
-And yes, part of the low repeatability could be explained by inter-plate error, as individual samples were fully randomised across plates.
+We agree that part of the low repeatability could be explained by inter-plate error, as individual samples were fully randomised across plates.
 However, the within-individual repeatability is still far lower than the within-sample (across-plate) repeatability.
@@
 
-RESPONSE: As stated in the discussion (lines XX), we feel that this is an important area for future research, and we hope that it will be a whole manuscript on its own.
-We are already pushing the page limit for Journal of Animal Ecology, so adding in an entire new set of analyses is just not possible.
+RESPONSE: As stated in the discussion (lines XX), we feel that this is an important area for future research.
+However, this would involve a comprehensive, complex set of new analyses.
+Difficulties associated with such an analysis are incorporating temporal changes in territory quality/conditions, incorporating the effects of individuals moving territories, and moving beyong detecting telomere lenghthening at a population level, to distinguishing legthening from error at the individual level.
+Thus, we hope that addressing the relationship between telomere lengthening and enviroinment will form a whole manuscript on its own - but we need to do a lot more work before such a manuscript can be written.
+From a practical perspective, we are already pushing the page limit for Journal of Animal Ecology, so adding in an entire new set of analyses is just not possible.
 

### 2017-06-24 Manuscript/SW_early_life_telomeres_FULL.Rmd
Home edits

@@
 Breeding attempts are often unsuccessful, and as a result of this low reproductive output, and higher mortality in first-year birds [39% in first-year birds versus 16% in adults; @Brouwer2006], cohort sizes in the Seychelles warbler are typically small (< 50).
-The 22 birth year cohorts used in this study cover `r min(dd$LayYear)` to `r max(dd$LayYear)` – the time period during which our data and sampling are most complete.
+The 22 hatch year cohorts used in this study cover `r min(dd$LayYear)` to `r max(dd$LayYear)` – the time period during which our data and sampling are most complete.
 
@@
 Negative values of RTL reflect decreases in telomere length with age, while positive values reflect increases.
-Note that the follow-up time between samples varied among individuals, and that individuals are not always caught at exactly the same time of year.
+Individuals were not always caught at the same time of year, but within a 3 month breeding season window.
 
@@
 We created models where the relationship between RTL and age was linear, quadratic, log-linear, and where age was fitted as a factor.
-For each age term, we fitted additional models including birth year (cohort) as a factor.
+For each age term, we fitted additional models including hatch year (cohort) as a factor.
 All fitted models are included in Table 1.
@@
 
-Our study adds to the substantial body of literature from humans and wild animals showing that telomere length decreases with age, and that this decrease is most rapid in early life - most likely as a consequence of cellular division [e.g.
+Our study adds to the substantial body of literature from humans and wild animals showing that telomere length decreases with age, and that this decrease is most rapid in early life - most likely as a consequence of the much more rapid rate of cellular division that occurs during the growth phase, but also perhaps higher levels of cellular stress during development [e.g.
 @Frenck1998; @Haussmann2003; @Heidinger2012].
@@
 Our results concur with other studies which show that early life  conditions / food availability can have a very significant and long term impacts on telomere length (and intrinsic biological condition) in captive and wild animals [e.g.
-@Nettle2015; @Heidinger2012; @Watson2015].
+@Stier2014; @Nettle2015; @Watson2015].
 
@@
 
-**Table 1** Telomere dynamics and age in Seychelles warbler cohorts.
-Linear mixed models were created with RTL as the response variable, and different measures of age, along with cohort ID, were included as explanatory variables (see methods for details).
+**Table 1** Cross-sectional (**A**) and longitudinal (**B**) telomere dynamics and age in Seychelles warbler cohorts.
+Linear mixed models were created with RTL (**A**) or $\Delta$RTL (**B**) as the response variable, and different measures of age, along with cohort ID, were included as explanatory variables (see methods for details).
 Models are ranked by AICc, with best models at the top of the table.
@@
   ggplot(aes(y = RTL,x = Cohort, fill = Cohort))+
-  xlab('Birth year') +
+  xlab('Hatch year') +
   geom_boxplot(outlier.shape = NA) +

### 2017-06-25 Manuscript/Response_to_reviewers.Rmd
Edits

@@
 We have now been through the Associate Editor and reviewer comments in detail, and made substantial changes to our manuscript in response.
-Below we outline, point-by-point, how we have responded to each of the reviewer comments.
-We also reproduce the sections of our manuscript we have changed in response to the reviewer comments.
-Reviewer comments are in bold, and reproduced manuscript sections are indented and italicised.
+Below we outline, point-by-point, how we have responded to each of the AE and reviewer comments.
+We also reproduce the sections of our manuscript we have changed in response to the comments.
+AE/Reviewer comments are in bold, and reproduced manuscript sections are indented and italicised.
 
@@
 
-RESPONSE: We thank the editor and reviewers for their positive and constructive comments.
+RESPONSE: We thank the editor and for these comments.
 We have now addressed the methodological questions raised by reviewers one and three, and have updated several of our analyses in line with their suggestions.
-We have also performed additional analyses to show that sample storage time does not affect telomere length, and included a discussion on repeatability.
-While almost none of our results have changed, we believe that incorporating the reviewer comments has substantially improved the clarity, readability and rigour of our work.
+In particular, we have now included a qPCR plate ID random effect term into all relevant analyses, which has not changed our results.
+We have also performed additional analyses to show that sample storage time does not affect telomere length, included more information on the DNA quality of our samples, including a new supplementary figure.
+We have also updated our discussion to address the methodological concerns of reviewers one and three, and we have toned down our conclusions on one of the specific analyses, as reqwuested by reviewer one.
 Finally, we have updated our bibliography, in accordance with the reviewer comments.
+We believe that incorporating the reviewer comments has substantially improved the clarity, readability and rigour of our work.
 
@@
 Our results are almost identical, with all the reported significant effects remaining.
-The only change is that the cohort x age effect on telomere length is now stronger (see response to reviewer 3).
+The only change is that the cohort x age effect on telomere length is now stronger (see response to later comment).
 Overall, we can be very confident that plate effects have not biased our results.
@@
 
-We now explain our randomisation approach in the methods, lines XX:
+We now explain our randomisation approach in the methods, lines 168-170:
 
@@
 
-We also include a discussion on error rate on lines XX:
+We also include a discussion on error rate on lines 413-423:
 
@@
 2014).
-The measurement error in our study has most likely arisen because samples were run aver a period of several years, during which reagents, consumables and personell all change.
+The measurement error in our study has most likely arisen because samples were run aver a period of several years, during which reagents, consumables and personnel all change.
 Compared to a set-up where a small amount of samples are all run at the same time, it is unsurprising that we have a higher error rate.
@@
 
-RESPONSE: Again, the level of efficiency is lower than we would have liked, although again this value sits fairly squarely within that found in the published telomere literature (see Fairlie et al.
+RESPONSE: Again, the level of efficiency is lower than we would have liked, although this value sits fairly squarely within that found in the published telomere literature (see Fairlie et al.
 2015, for a recent example), and often it is not even reported.
@@
 
-RESPONSE: We now provide more clarification about how we determined and analyse age, lines XX:
+RESPONSE: We now provide more clarification about how we determined and analyse age, lines 128-133:
 
@@
 2003).
-Within the first year of life, birds are classified as nestlings less than one month old (rounded to one month for analyses), fledglings less than six month old (rounded to six months) or subadults up to one year old (rounded to 10 months).
+Within the first year of life, birds are classified as nestlings less than one month old (rounded to one month for analyses), fledglings less than six months old (rounded to six months) or subadults up to one year old (rounded to 10 months).
 Ages for adult birds were rounded to the nearest year."
@@
 S1).
-We refer to this in the main text, lines XX.
+We refer to this in the main text, lines 165-166.
 
@@
 
-We also performed an analysis of RTL and sample storage time, using adult birds, and found no evidence of a relationship - lines XX.
+We also performed an analysis of RTL and sample storage time, using adult birds, and found no evidence of a relationship - lines 254-255.
 
@@
 
-RESPONSE: See response above re. randomisation. While not an issue for our study, we agree that it is a point worth raising. We now discuss this issue on lines XX:
+RESPONSE: See response above re. randomisation. While not an issue for our study, we agree that it is a point worth raising. We now discuss this issue on lines 376-383:
 
@@
 \textit{"Although a few studies have shown that temporal variation in telomere dynamics occurs in natural populations, these have been limited in the number of seasons they cover (Mizutani et al.
-2013; Watson, Bolton and Monaghan 2015; Fairlie et al.
+2013; Watson, Bolton, and Monaghan 2015; Fairlie et al.
 2016).
 Other studies have found cohort effects but not discussed them in an ecological context (Stier et al.
-2014; Becker et al.2015).
+2014; Becker et al.
+2015).
 One problem with studying cohort effects is that it can be difficult to tease apart true cohort effects from effects that may arise if samples degrade with storage time, and/or batch effects in telomere assays, although neither of these factors were a problem in our study.
@@
 RESPONSE: Interestingly, this is the one effect that has changed from  including a plate ID random effect in our analyses, with delta AIC now ~11.
-However, we do not wish to place any strong conclusions on this analysis.
-We agree that while there is convincing evidence for cohort effects on telomere length, the evidence for cohort effects on telomere dynamics is weaker.
+However, we do not wish to place any strong conclusions on this analysis, given that the significance of the effect appears to be dependent on the inlcusion of a plate ID random effect that we had not previously considered.
+Thus, we agree that while there is convincing evidence for cohort effects on telomere length, the evidence for cohort effects on telomere dynamics is weaker.
 We have therefore removed references to this effect throughout the discussion, and allow readers to interpret the effect as they wish.
@@
 Further, we disagree with the sentiment that a biomarker must be consistent over individual lifespans to be useful - rather, a useful marker needs to change with ecological context.
-We feel that we cover this point in the discussion, lines XX:
+We feel that we cover this point in the discussion, lines 366-374:
 
@@
 
-\textit{"Such inconsistent changes in telomere length over lifespans could occur due to changes in the composition of cell types within individual samples, or due to the actual elongation of telomeres (Blackburn et al.
-1999).
+\textit{"Such inconsistent changes in telomere length over lifespans could occur due to changes in the cellular composition of of the blood within individual samples, or due to the actual elongation of telomeres (Blackburn et al.
+1989).
 Determining the mechanism of these changes is essential for how we view telomeres as biomarkers of costs.
-For example, if telomeres can be lengthened in response to improvements in environmental conditions, this would suggest that they reflect short- to medium-term costs, rather than the cumulative costs that an individual has faced over its lifespan (Bateson et al.
-2016).
+For example, if telomeres can be lengthened in response to improvements in environmental conditions, this would suggest that they reflect short- to medium-term costs, rather than the cumulative costs that an individual has faced over its lifespan (Bateson 2016).
 New statistical and technical approaches are therefore now required to determine the mechanisms behind increases in telomere length within individuals, so that biologically informed hypotheses about the ecological causes and consequences of these increases can be generated and tested."}
@@
 
-RESPONSE: It is true that the shorter timescales of these studies may be the reason for the higher within-individual consistency, and we have now added a note to this effect, lines XX:
+RESPONSE: It is true that the shorter timescales of these studies may be the reason for the higher within-individual consistency, and we have now added a note to this effect, lines 348-359:
 
@@
 \textit{"Consistent with a pattern of sporadic changes in telomere length with age, we found that within-individual telomere measurements were only weakly correlated.
-These findings are in contrast to other avian studies in which within-individual telomere length measurements were highly consistent, and individual-level telomere shortening occurred throughout the juvenile period and into adulthood (Heidinger et al.
+Although some of this low within-individual repeatability will occur due to measurement error, our within-sample repeatability was still much higher than our within- individual repeatability.
+Such a low value of within-individual repeatability in telomere length is in contrast to other avian studies in which within-individual telomere length measurements were highly consistent, and individual-level telomere shortening occurred throughout the juvenile period and into adulthood (Heidinger et al.
 2012; Boonekamp et al.
@@
 However, as we argue above, we do not think it is likely that the pattern is a result of sample storage/DNA integrity.
-We have no evidence from our samples that there is sample degradiation with respect to DNA integrity or telomere length (see above) - something confirmed by other studies (see Reichert et al.
-2017; Oecologia)
+We have no evidence from our samples that there is sample degradiation with respect to DNA integrity or telomere length (see above).
 
@@
 
-RESPONSE: Yes, this is a good point. We now clarify in the discussion, lines XX
+RESPONSE: Yes, this is a good point. We now clarify in the discussion, lines 328-332:
 
@@
 
-\textit{"Our study adds to the substantial body of literature from humans and wild animals showing that telomere length decreases with age, and that this decrease is most rapid in early life - most likely as a consequence of the much more rapid rate of cellular division that occurs during the growth phase, but also perhaps higher levels of cellular stress during development..."}
+\textit{"Our study adds to the substantial body of literature from humans and wild animals showing that telomere length decreases with age, and that this decrease is most rapid in early life - most likely as a consequence of the much more rapid rate of cellular division that occurs during the growth phase, but also perhaps higher levels of cellular stress during development (e.g.
+Frenck, Blackburn, and Shannon 1998; Haussmann, Vleck, and Nisbet 2003; Heidinger et al.
+2012)."}
 
@@
 
-RESPONSE: We have clarified that this is in vitro.
+RESPONSE: Agreed. We now have clarified that this is in vitro.
 
@@
 
-RESPONSE: We disagree here - it is the outcome of the molecular methods. We tried moving it but this disrupted the flow of the manuscript, so have left it as it was.
+RESPONSE: We tried moving this section as suggested, but could not do so without significantly disrupting the flow and readability of the manuscript.
+We have therefore left it as it was, but can change if this is something viewed as essential.
 
@@
 
-RESPONSE: We agree that this was confusing, and have simplified along the suggested lines - now lines XX:
+RESPONSE: We agree that this was confusing, and have simplified along the suggested lines - now lines 291-295:
 
@@
 
-\textit{"We found that the likelihood of telomere lengthening increased with log age...
+\textit{"We found that the likelihood of telomere lengthening increased with log age (GLMM with lengthened yes/no as binomial response; estimate = 0.296, CIs = 0.005, 0.588).
 Increases in telomere length were most likely to be observed shortly after the juvenile period, at around four years of age, and later in life (although sample sizes for older birds are much smaller; Fig.
-2C,D)"}
+2C,D)."}
 
@@
 
-RESPONSE: We have now simplified the sentence to make it clearer, lines XX:
+RESPONSE: We have now simplified the sentence to make it clearer, lines 323-325:
 
@@
 
-RESPONSE: We are not completely sure what is unclear about this, but have tried rephrasing. Now lines XX:
+RESPONSE: We are not completely sure what is unclear about this, but have tried rephrasing. Now lines 346-347:
 
@@
 
-\textit{"Increases in telomere length were not consistent over individual lifespans, but occurred in bouts, against a backdrop of overall lifelong telomere shortening.
-"}
+\textit{"Increases in telomere length were not consistent over individual lifespans, but occurred in bouts, against a backdrop of overall lifelong telomere shortening."}
 
@@
 
-RESPONSE: We rephrased accordingly, lines XX:
+RESPONSE: We rephrased accordingly, lines 405-407:
 
@@
 We thank the reviewer for pointing out areas where our citation of previous work was incomplete - although we believe our original citations were generally correct, we agree that there were one or two mistakes, and that it is important to provide the reader with a complete context for our work.
-We respond to each of the individual questions/comments below.
+We respond this point, and to each of the individual questions/comments, below.
 
@@
 
-RESPONSE: We have re-written the text so that the contrast is clearer here. Lines XX:
+RESPONSE: We have re-written the text so that the contrast is clearer here. Lines 60-71:
 
@@
 2003).
-The association between telomere length and senescence has inspired a great deal of recent research into telomere evolutionary ecology, and relationships between telomere dynamics and survival or lifespan have been documented in wild populations of several species (Haussmann and Marchetto 2010; Barrett et al.
+The association between telomere length and senescence has inspired a great deal of recent research into telomere evolutionary ecology, and relationships between telomere dynamics and survival or lifespan have been documented in wild populations of several species (Barrett et al.
 2013; Stier et al.
@@
 
-\textit{"Although the causal role of telomeres in senescence and survival is not yet clear, there is mounting evidence that telomeres can act as biomarkers of individual condition and ageing in wild populations, providing a measure of the ecological stress that an individual has experienced, a signature that can otherwise be difficult to detect (e.g.
+\textit{"Although the causal role of telomeres in senescence and survival is not yet clear, there is mounting evidence that telomeres can act as biomarkers of individual condition and ageing in wild populations.
+Specifically, telomeres may be able to provide a measure of the ecological stress that an individual has experienced - a signature that can otherwise be difficult to detect (e.g.
 Schultner et al.
@@
 2015; Bebbington et al.
-2016)."}
+2016)"}
 
@@
 
-RESPONSE: We now clarify (lines XX) that the territory quality estimates that we use in our statistical models are taken from the field period in which the telomere sample was taken.
+RESPONSE: We now clarify (lines 227-232) that the territory quality estimates that we use in our statistical models are taken from the field period in which the telomere sample was taken.
 The same applies to insect abundance, group size etc.
 We feel that an in-depth explanation of how and why territory quality varies over space and time is outside the scope of this manuscript, especially given i) the word limit, and ii) the fact that territory quality is not related to RTL.
-However we do now include a reference to recent book chapter (Komdeur et al.
+However we do now include a reference to a recent book chapter (Komdeur et al.
 2016) which explains how and why territory quality varies on Cousin Island.
-Our updated section on territory quality and indect abundance is now on lines XX:
+Our updated section on territory quality and indect abundance is now on lines 142-150:
 
@@
 Such island-wide temporal variation may override the effects of variation in individual territory quality across the island.
-As an estimate of seasonal variation in food availability, we calculated an index of the abundance of insects across the entire island during each main breeding season (referred to hereafter as 'insect abundance').
-This index is calculated as the mean number of insects found per unit leaf area over all monthly surveys carried out on the island in a main breeding season.
-"}
+As an estimate of seasonal variation in food availability, we calculated an index of the abundance of insects across the entire island during each main breeding season (referred to hereafter as ‘insect abundance’).
+This index is calculated as the mean number of insects found per unit leaf area over all monthly surveys carried out on the island in a main breeding season."}
 
@@
 
-RESPONSE: Samples are taken as and when we can catch the birds, which is not always in the same month. We now clarify this on lines XX:
+RESPONSE: Samples are taken as and when we can catch the birds, which is not always in the same month, but samples are generally taken at the same time of year. We now clarify this on lines 177-178:
 
@@
 
-\textit{"Individuals were not always caught at the same time of year, but within a 3 month breeding season window."}
+\textit{"Individuals were not always caught in the same month, but were generally caught within a 3 month breeding season window."}
 
@@
 
-RESPONSE: We have now added a table of cohort sizes to the supplement (Table S1), and refer to this in the main text on Line XX.
+RESPONSE: We have now added a table of cohort sizes to the supplement (Table S1), and refer to this in the main text (e.g. on Line 126).
 
@@
 
-RESPONSE: We have now included intra-plate repeatability estimates on lines XX. Samples were generally placed next to one another on the plate.
+RESPONSE: We have now included intra-plate repeatability estimates on lines 251-252. Samples were generally placed next to one another on the plate.
 
@@
 
-RESPONSE: We used a mean. We have now clarified this on line XX.
+RESPONSE: We used a mean. We have now clarified this on lines 181-182.
 
@@
 
-RESPONSE: We now include a formal test of the within and between individual slopes, as described in van de Pol and Wright (2009), and find that there is no difference between cross-sectional and longitudinal telomere shortening - lines XX:
+RESPONSE: We now include a formal test of the within and between individual slopes, as described in van de Pol and Wright (2009), and find that there is no difference between cross-sectional and longitudinal telomere shortening - lines 273-274:
 
@@
 
-\textit{"Further, we found no evidence that within and between individual slopes of telomere shortening varied (estimate = -0.003, CIs = -0.008, 0.003; see Van de Pol and Wright 2009), suggesting that there was no difference between cross-sectional and longitudinal telomere shortening with age."}
-
-Not also that this analysis was done in one of the original Seychelles warbler telomere studies (Barrett et al. 2013; Mol. Ecol.)
+\textit{"Further, we found no evidence that within and between individual slopes of telomere shortening varied (estimate = -0.003, CIs = -0.008, 0.003), suggesting that there was no difference"}
 
@@
 However, the within-individual repeatability is still far lower than the within-sample (across-plate) repeatability.
-We now add a caveat to this effect, on lines XX:
+We now add a caveat to this effect, on lines 348-352:
 
@@
 
-RESPONSE: As stated in the discussion (lines XX), we feel that this is an important area for future research.
+RESPONSE: As stated in the discussion (lines 360-374), we feel that this is an important area for future research.
 However, this would involve a comprehensive, complex set of new analyses.
@@
 Thus, we hope that addressing the relationship between telomere lengthening and enviroinment will form a whole manuscript on its own - but we need to do a lot more work before such a manuscript can be written.
-From a practical perspective, we are already pushing the page limit for Journal of Animal Ecology, so adding in an entire new set of analyses is just not possible.
+From a practical perspective, we are already pushing the word limit for Journal of Animal Ecology, so adding in an entire new set of analyses is just not possible.
 
@@
 RESPONSE: See response above regarding the analysis of lengthening.
-Although this section is somewhat methodological, we feel that i) it is a very important point to be able to separate telomere elongation from sampling error, as few studies have done so in ecological systems, and ii) we make it clear (lines XX) that our findings are likely to be of ecological importance.
+Although this section is somewhat methodological, we feel that i) it is a very important point to be able to separate telomere elongation from sampling error, as few studies have done so in ecological systems, and ii) we make it clear (lines 360-374) that our findings are likely to be of ecological importance.
 

### 2017-06-25 Manuscript/SW_early_life_telomeres_FULL.Rmd
Edits

@@
 Telomeres, the protective caps at the ends of chromosomes, shorten in response to oxidative stress, and telomere shortening is correlated with reduced survival and lifespan.
-Investigating telomere dynamics may help us quantify individual variation in the costs experienced from social and ecological environmental factors, and enhance our understanding of the dynamics of natural populations.
+Investigating telomere dynamics may help us quantify individual variation in the costs experienced from social and ecological factors, and enhance our understanding of the dynamics of natural populations.
 
@@
 When telomeres become critically short, cells senesce [@Campisi2003], and the accumulation of these cells has been suggested to result in organismal senescence and death [@Wong2003].
-The association between telomere length and senescence has inspired a great deal of recent research into telomere evolutionary ecology, and relationships between telomere dynamics and survival or lifespan have been documented in wild populations of several species [@Haussmann2010; @Barrett2013a; @Stier2015].
+The association between telomere length and senescence has inspired a great deal of recent research into telomere evolutionary ecology, and relationships between telomere dynamics and survival or lifespan have been documented in wild populations of several species [@Barrett2013a; @Stier2015].
 As yet, however, there is little direct evidence that the relationship between telomere dynamics and survival is causal [@Simons2015].
 
-Although the causal role of telomeres in senescence and survival is not yet clear, there is mounting evidence that telomeres can act as biomarkers of individual condition and ageing in wild populations, providing a measure of the ecological stress that an individual has experienced, a signature that can otherwise be difficult to detect [e.g.
+Although the causal role of telomeres in senescence and survival is not yet clear, there is mounting evidence that telomeres can act as biomarkers of individual condition and ageing in wild populations.
+Specifically, telomeres may be able to provide a measure of the ecological stress that an individual has experienced - a signature that can otherwise be difficult to detect [e.g.
 @Schultner2014; @Asghar2015; @Bebbington2016].
@@
 
-In this study, we test how lifelong telomere dynamics are related to environmental variation across 22 Seychelles warbler cohorts (years).
+In this study, we test how lifelong telomere dynamics are related to environmental variation across 22 Seychelles warbler cohorts.
 We first study how telomere length and rates of shortening are related to age and sex across all life stages, and how this relationship varies among cohorts, in order to gain an in-depth understanding of the temporal dynamics of telomere changes.
-We then examine, within individuals, how telomere length changes with age, and statistically test whether observed increases in telomere length across temporally longitudinal samples within individuals are larger than can be accounted for by measurement error.
+We then examine, within individuals, how telomere length changes with age, and statistically test whether observed increases in telomere length over individual lifespans are larger than can be accounted for by measurement error.
 Finally, we test how telomere length and shortening are related to a wide range of social and environmental variables in order to gain a fuller understanding of the forces driving telomere dynamics in natural populations.
@@
 This species' main breeding season runs from June-September (though a small proportion of pairs also breed between January–March), when the breeding females on many of the *ca* 110 territories will attempt to breed, laying one or, rarely, two or three eggs [@Komdeur1991].
-Breeding attempts are often unsuccessful, and as a result of this low reproductive output, and higher mortality in first-year birds [39% in first-year birds versus 16% in adults; @Brouwer2006], cohort sizes in the Seychelles warbler are typically small (< 50).
+Breeding attempts are often unsuccessful, and as a result of this low reproductive output, and higher mortality in first-year birds [39% in first-year birds versus 16% in adults; @Brouwer2006], cohort sizes in the Seychelles warbler are typically small (< 50; Table S1).
 The 22 hatch year cohorts used in this study cover `r min(dd$LayYear)` to `r max(dd$LayYear)` – the time period during which our data and sampling are most complete.
@@
 We aged all birds using information on eye colour at first capture [@Komdeur1991a] and previous capture history [@Richardson2003b].
-Within the first year of life, birds are classified as nestlings less than one month old (rounded to one month for analyses), fledglings less than six month old (rounded to six months) or subadults up to one year old (rounded to 10 months).
+Within the first year of life, birds are classified as nestlings less than one month old (rounded to one month for analyses), fledglings less than six months old (rounded to six months) or subadults up to one year old (rounded to 10 months).
 Ages for adult birds were rounded to the nearest year.
@@
 Negative values of RTL reflect decreases in telomere length with age, while positive values reflect increases.
-Individuals were not always caught at the same time of year, but within a 3 month breeding season window.
+Individuals were not always caught in the same month, but were generally caught within a 3 month breeding season window.
 
@@
 We calculated $\Delta$age using log and polynomial transformed age data, and carried out model selection as above, with the exception that we did not model $\Delta$age as a factor (due to a lack of discreet groupings), and mean age was also included in models to partition within-individual *vs* cross-sectional effects [@VandePol2009].
+We also excluded the qPCR plate ID random effect from the longitudinal analyses, as each longitudinal measurement was obtained from reactions run on separate plates.
 
@@
 When examining the distribution of longitudinal telomere changes we observed some increases in telomere length with age in individuals.
-We therefore repeated the qPCR on each sample  using completely separate reactions run on separate plates and used these repeat measurements to test whether these increases could be explained by measurement error.
+We therefore repeated the qPCR on a large number of samples, using completely separate reactions run on separate plates.
+We used these repeat measurements to test whether these increases could be explained by measurement error.
 We calculated the change in RTL between pairs of repeat measurements within the same samples (hereafter $\Delta$RTL~*sample*~; N = `r nrow(subset(ddL,Group == 'Within samples'))` pairs of measurements from `r length(unique(subset(ddL,Group == 'Within samples')$BirdID))` birds) in exactly the same way as for across samples (hereafter $\Delta$RTL~*individual*~).
@@
 
-We also tested whether consistent telomere lengthening across our dataset using a modified version of the approach developed by Simons *et al.* [-@Simons2014].
+We also tested whether consistent telomere lengthening occurred across our dataset using a modified version of the approach developed by Simons *et al.* [-@Simons2014].
 Briefly, this approach utilises samples with at least three telomere measurements to compare residual variance in telomere change over time with the overall change in telomere length between the first and last telomere measurements [@Simons2014].
@@
 For individuals with longitudinal data we repeated the above analyses of telomere dynamics, replacing telomere length with $\Delta$RTL~*individual*~ as the response variable, and including the environmental/social explanatory variables from the first of the two sampling points.
-We excluded the cohort ID random effect from this analysis, as longitudinal telomere dynamics did not differ among cohorts; see results.
+We excluded the qPCR plate ID random effect from this analysis (see above), and excluded the cohort ID random effect, as longitudinal telomere dynamics did not differ among cohorts; see results.
 
 ##Results
+
 We measured telomere lengths using a total of `r length(unique(dd$BloodID))` unique samples from juvenile and adult Seychelles warblers from `r length(table(dd$Cohort))` cohorts born between `r min(dd$LayYear)` and `r max(dd$LayYear)` (Table S1).
@@
 We first tested how RTL was related to age among cohorts using a model selection approach.
-The top model contained a log-linear relationship between RTL and age (Table 1A).
-All other models fitted the data much less well ($\Delta$AICc > 15; Table 1).
+The top model contained cohort ID and a log-linear relationship between RTL and age (Table 1A).
+All other models fitted the data much less well ($\Delta$AICc > 20; Table 1A).
 The log-linear relationship between RTL and age could be seen clearly in the raw data; RTL decreased with age (estimate = `r round3(fixef(RTL_LogAge)['LogAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_LogAge)['LogAge',],lessthan = F),collapse = ', ')`), with the greatest decrease occurring in the first year of life (Fig.
@@
 RTL decreased with $\Delta$log age (estimate = `r round3(fixef(RTL_DeltaLogAge)['DeltaLogAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_DeltaLogAge)['DeltaLogAge',],lessthan = F),collapse = ', ')`), confirming that within-individual telomere shortening occurs across the Seychelles warbler dataset.
-Further, we found no evidence that within and between individual slopes of telomere shortening varied (estimate = `r round3(fixef(RTL_LogAge2)['MeanAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_LogAge2)['MeanAge',],lessthan = F),collapse = ', ')`; see Van de Pol and Wright 2009), suggesting that there was no difference between cross-sectional and longitudinal telomere shortening with age.
+Further, we found no evidence that within and between individual slopes of telomere shortening varied (estimate = `r round3(fixef(RTL_LogAge2)['MeanAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_LogAge2)['MeanAge',],lessthan = F),collapse = ', ')`), suggesting that there was no difference between cross-sectional and longitudinal telomere shortening with age [see @VandePol2009].
 
@@
 Accordingly, there was a positive correlation between RTL measured from different samples taken at different time points during an individual’s life (Fig.
-2A), but this was very weak (marginal R2 = `r round2(r.squaredGLMM(RTL_RTL1)['R2m'])`), and not significant (estimate = `r round3(fixef(RTL_RTL1)['RTL'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_RTL1)['RTL',],lessthan = F),collapse = ', ')`).
+2A), but this was very weak (marginal R^2^ = `r round2(r.squaredGLMM(RTL_RTL1)['R2m'])`), and not significant (estimate = `r round3(fixef(RTL_RTL1)['RTL'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_RTL1)['RTL',],lessthan = F),collapse = ', ')`).
 
@@
 ##Discussion
+
 Here we use a long-term, multi-cohort dataset to assess lifelong telomere dynamics and the relationship between these and spatio-temporal variation in the ecological environment in a contained population of Seychelles warblers.
 We found that telomere length decreases with age, and that this decrease is greatest very early in life.
-Telomere length decreased with age in almost all of the `r length(table(dd$Cohort))` cohorts studied, but also that telomere length varied substantially among cohorts.
+Telomere length decreased with age in almost all of the `r length(table(dd$Cohort))` cohorts studied, but telomere length varied substantially among cohorts.
 Despite an overall pattern of telomere shortening with age in the Seychelles warbler, we found evidence of within-individual increases in telomere length, and that the extent of these increases could not be explained solely by qPCR measurement error.
@@
 Measurement error is one factor that is an issue in our study, and other studies that use qPCR to measure telomere length [@Nussey2014].
-The measurement error in our study has most likely arisen because samples were run aver a period of several years, during which reagents, consumables and personell all change.
+The measurement error in our study has most likely arisen because samples were run aver a period of several years, during which reagents, consumables and personnel all change.
 Compared to a set-up where a small amount of samples are all run at the same time, it is unsurprising that we have a higher error rate.
@@
 LGS was also funded by a fellowship from the BBSRC (BB/N011759/1), HLD by a NERC fellowship (NE/I021748/1), MH was funded by a VENI fellowship from the Netherlands Organisation for Scientific Research (863.15.020), and TB by a Leverhulme fellowship.
+Eric Stier, Dan Nussey and one anonymous reviewer provided helpful comments on the manuscript.
 
@@
   subset(!duplicated(paste0(BirdID,Agemonths))) %>%
-  subset(Agemonths > 2) %>%
   ggplot(aes(x = Agemonths,y = RTL,group = BirdID,col=factor(BirdID)))+
@@
 Fig1B <- transform(dd,Cohort = factor(LayYear)) %>%
+    subset(Agemonths > 2) %>%
   ggplot(aes(y = RTL,x = Cohort, fill = Cohort))+

### 2017-07-10 Manuscript/Response_to_reviewers_2.Rmd
Reviews round 2

---
output: 
  pdf_document:
header-includes:
  \usepackage{graphicx}
  \usepackage{changepage}
  \usepackage{color}
---

\begin{flushright}

\includegraphics{UEAlogo}

\bigskip

Dr Lewis G. Spurgin\\*
BBSRC Research Fellow\\*
School of Biological Sciences\\*
University of East Anglia\\*
Norwich\\*
Norfolk, NR4 7TJ\\* 
United Kingdom

L.Spurgin@uea.ac.uk\\*
lewisspurgin.wordpress.com

\end{flushright}

\bigskip
\bigskip

**July 2017**

\bigskip

Dear Ben,

Thank you for taking the time once again to read our manuscript and response letter. Once again the reviews have been robust but fair, and we have been very impressed at the quality of the reviewing process. We have now been through the second round of Associate Editor and Reviewer comments, and adjusted our manuscript accordingly. In particular, we have added further caveats to our discussion with regards to repeatability, and we agree that this does not take away from the positive aspects of the manuscript.

In addition to the revised manuscript, below we provide a point-by-point response to the Associate Editor and Reviewer comments. As with our previous submission, the editor/reviewer comments are in bold, and edited manuscript sections are reproduced in italics and indented.

Please let us know if you need anything else.

\bigskip
\bigskip

Kind Regards,

\bigskip
\bigskip

Lewis Spurgin (on behalf of all authors)

\newpage

#Associate editor comments

---

\bigskip

**One of the original reviewers and myself have now read your revised manuscript and we both think that it has improved substantially. The reviewer, however, very nicely describes some points that I strongly agree would need more attention in the discussion of the paper to more honestly balance/describe the strong AND weak points of your study. These weaker points do not decrease the impact, novelty or relevance of your study, but discussing them in more detail could rather inspire advances to the methodology used in your field of research.**

RESPONSE: We are glad that you both agree that our manuscript has improved, and thank you again for your constructive comments. We agree with the central point made by reviewer and yourself, and have now adjusted our discussion accordingly. We hope you agree that this has further improved our manuscript, and we are hopful that this improved version will generate discussion about both the ecological and technical aspects of telomere dynamics in wild populations.

\newpage

#Reviewer 1

---

\bigskip

**Line 441: It is very kind of the authors to acknowledge me by name, but my first name is Antoine, not Eric...**

RESPONSE: We are very sorry about this, and we do of course know that your name is Antoine! We have now changed the ackowledgements.

\bigskip

---

\bigskip

**As far as I understand ICC, it is relatively close to a pearson r with some adjustments. Therefore, to quantify the proportion % of variance explained, you should square transform it to obtain something looking like a R2. Therefore, the repeatability of 0.68 mentioned by the authors would mean that 46.2% of their measurement is explained by the sample and that 53.8% is explained by measurement error.**

RESPONSE: We feel that we will have to agree to disagree here. ICC is calculated as:

$variation between groups / (variation between groups + variation within groups)$

Thus if within-group variation is extremely small (i.e. no measurement error), the ICC would be 1, so 0% of the variation would be explained by measurement error. Thus we would interpret an ICC of 0.68 as 68% of the variation being explained by sample identity and 32% by variation other than sample identity (i.e. measurement error).

However, this feels like a debate over semantics. In the manuscript, we are careful not to phrase things in terms of repeatability, and nowhere do we state that 68% of the variance can be explained by sample identity.

\bigskip

---

\bigskip

**While I understand that measurement error is higher in their study due to the long-term nature of the data, such measurement error (if I am calculating/understanding the numbers properly) is still a major issue to underline with more emphasis in the paper. I agree that it will not lead to false positive, but still it is reducing the power to detect any relationship with other factors. In Bebbington 2016, the repeatability given is 0.94 despite having 1064 samples from the same population/dataset. What is explaining such a loss of repeatability between the two datasets?**

**qPCR efficiency: An efficiency of 78% means that the qPCR conditions have not been optimized properly. Correcting with LinReg for differences in efficiency is minimizing noise related to such problem, but not fixing the problem when efficiency is so different from 100%. Again, this is likely to increase measurement error, but not to create false positive. Such limitation should be discussed in the paper too. The fact that others are publishing with poor efficiency is not a good answer in my opinion, and I can cite you many papers reporting efficiencies being in the range 90-110%, including most (all?) papers of Criscuolo’s lab and Monaghan’s lab. In LinReg, are you correcting for sample specific efficiency or per the average plate efficiency? Correcting for sample specific efficiency has been criticized in the past : "Sample specific PCR efficiency estimation has its usefulness, but currently only for outlier detection", Hellemans et al. 2007 qBase relative quantification framework and software for management and automated analysis of real-time quantitative PCR data.**

RESPONSE: We are glad that the reviewer agrees that our observed repeatability/efficiency values will not lead to false positives. We agree that this issue deserves more discussion space, and have now addressed the issue of repeatability, and the specific issue of qPCR efficiency, in the discussion on lines XX:

\bigskip

---

\bigskip

**Information on the age of birds: While I acknowledge that the authors now provide better information on the age of individuals, the resolution they have for TL at the nestling age is not optimal in my opinion since many research groups (Criscuolo, Monaghan, Verhulst) have identified telomere shortening within 1-4weeks in nestlings, and the data the authors present here seems to indicate that most TL shortening occur early (very early?) in life.**

RESPONSE: We would love to have had fine-scale longitudinal samples from within the nest, but this is not possible in the Seychelles warbler system, in which the nests are very high and difficult to access safely. Systems in which birds can be sample repeatedly early in life (e.g. from nest boxes) are much better for addressing the specific issue of early life telomere loss. The point that we tried to make was that the (lack of) sampling resolution in early life should be offset against have broad-scale, longitudinal samples froma  large number of cohorts. Nonetheless, we agree that we are likely be missing a degree of early-life telomere shortening, and have a added a section to the discussion on this, lines XX:

\bigskip

---

\bigskip

**Intra-individual repeatability of 8%: Maybe “worrying” was not the proper word; please excuse my approximate usage sometimes of English. Yet, I am still ‘puzzled’ by the very low intra-individual repeatability for telomere found here and by Dan Nussey’s team. I don’t have similar long-term dataset, but overall in my datasets, I find an intra-individual repeatability ca. 65% over a period of 9 days during growth in tits (see Fig below) for great and coal tits from Stier et al. 2016 J Avian Biol), and 28% over a period of 105 days in king penguin chicks (unpublished). Verhulst’s team find similar/stronger intra-individual repeatability (e.g. Boonekamp 2014 both for nestling and between nestling and adulthood; Bauch 2013 Mol Ecol in adult seabirds).**

**Do you have any idea about the presence of interstitial telomeric sequence in your study species? Because presence of ITS would even artificially increase intra-individual repeatability when telomeres are measured with qPCR.
If there is only 8% of intra-individual repeatability, it is quite surprising that telomere length at a given point can predict survival, as shown by many studies including in your own study system. With such a low intra-individual repeatability, I agree that telomeres can be seen as a biomarker of short-term “costs”, but we would have to re-think our way of using it as a biomarker of long-term costs and survival prospects. This idea should be better highlighted in the discussion in my opinion.**

RESPONSE: We thank Dr Stier for clarifying this point. Unfortunately we do not have any data on the presence of ITS. However, we have now added further discussion on what may cause the difference between our study and the ones mentioned above. We have also now discussed what low intra-individual repeatability may mean for the relationship between telomere length and long-term survival - lines XX:

### 2017-07-11 Manuscript/Response_to_reviewers_2.Rmd
Edits

@@
 no measurement error), the ICC would be 1, so 0% of the variation would be explained by measurement error.
-Thus we would interpret an ICC of 0.68 as 68% of the variation being explained by sample identity and 32% by variation other than sample identity (i.e.
+We would therefore interpret an ICC of 0.68 as 68% of the variation being explained by sample identity and 32% by variation other than sample identity (i.e.
 measurement error).
 
-However, this feels like a debate over semantics.
-In the manuscript, we are careful not to phrase things in terms of repeatability, and nowhere do we state that 68% of the variance can be explained by sample identity.
+However, this is a debate over semantics.
+In the manuscript, we are careful to phrase things in terms of repeatability and avoid using pearson R and R2 values as these are generally considered inappropriate for measuring repeatability.
+Nowhere do we state that 68% of the variance can be explained by sample identity.
 
@@
 RESPONSE: We are glad that the reviewer agrees that our observed repeatability/efficiency values will not lead to false positives.
-We agree that this issue deserves more discussion space, and have now addressed the issue of repeatability, and the specific issue of qPCR efficiency, in the discussion on lines XX:
+Our dataset includes repeats that have run several years apart, and these repeats are included in our repeatability estimates.
+We suspect that our dataset is fairly unique in this respect at pressent, but other studies will certainly have such data available in the near future.
+Thew Bebbington et al.
+paper only included the set of most recently run telomere samples, hence the higher repeatability.
+We agree that this issue deserves more discussion space, and have now devoted an entire paragraph to the issue of repeatability, and the specific issue of qPCR efficiency, in the discussion on lines XX:
 
@@
 from nest boxes) are much better for addressing the specific issue of early life telomere loss.
-The point that we tried to make was that the (lack of) sampling resolution in early life should be offset against have broad-scale, longitudinal samples froma  large number of cohorts.
-Nonetheless, we agree that we are likely be missing a degree of early-life telomere shortening, and have a added a section to the discussion on this, lines XX:
+The point that we tried to make was that the (lack of) sampling resolution in early life should be offset against have broad-scale, longitudinal samples from a  large number of cohorts.
+Nonetheless, we agree that our set-up is not optimal for looking at early-life telomere shortening, and have added a section to the discussion on this, lines XX:
 
@@
 
-RESPONSE: We thank Dr Stier for clarifying this point.
+RESPONSE: We thank the reviewer for clarifying this point.
 Unfortunately we do not have any data on the presence of ITS.

### 2017-07-11 Manuscript/SW_early_life_telomeres_FULL.Rmd
Reviews round 2
Edits

@@
 
-Our study adds to the substantial body of literature from humans and wild animals showing that telomere length decreases with age, and that this decrease is most rapid in early life - most likely as a consequence of the much more rapid rate of cellular division that occurs during the growth phase, but also perhaps higher levels of cellular stress during development [e.g.
+Our study adds to the substantial body of literature from humans and wild animals showing that telomere length decreases with age, and that this decrease is most rapid in early life.
+Rapid telomere shortening in early life occurs as a consequence of the much more rapid rate of cellular division that occurs during the growth phase, but also perhaps higher levels of cellular stress during development [e.g.
 @Frenck1998; @Haussmann2003; @Heidinger2012].
-However, we also found that, despite an overall trend for shortening, telomere length both increased and decreased, especially after the juvenile period.
+Because we are rarely able to sample Seychelles warblers more than once in the nest, our system is not the best suited for looking in detail at the reasons behind telomere shortening during the growth phase.
+In contrast, we have good longitudinal and cross-sectional sampling from across individual lifespans, and using this data we have shown that, despite an overall trend for shortening, telomere length both increased and decreased, especially after the juvenile period.
 Importantly, these increases were observed in longitudinal as well as cross-sectional data, indicating that selective disappearance of individuals with shorter telomeres is not sufficient to explain this pattern.
@@
 For example, if telomeres can be lengthened in response to improvements in environmental conditions, this would suggest that they reflect short- to medium-term costs, rather than the cumulative costs that an individual has faced over its lifespan [@Bateson2016].
-New statistical and technical approaches are therefore now required to determine the mechanisms behind increases in telomere length within individuals, so that biologically informed hypotheses about the ecological causes and consequences of these increases can be generated and tested.
+It is clear that telomeres can be a marker of long-term as well as short-term costs, as telomere length has been associated with both survival and lifespan in wild populations [@Barrett2013a; @Stier2015], but we do not yet know how biologically meaningful within-individual fluctuations in telomere length are.
+New research is therefore  required to determine when and why telomere length increases within individuals, so that biologically informed hypotheses about the nature of telomeres as biomarkers in wild populations.
 
@@
 
-Although we found clear associations between the environment and telomere dynamics, we should bear in mind that the social and ecological variables we tested here explained only a small proportion of the variance in RTL.
-A poor social and ecological environment is known to be detrimental to Seychelles warblers, both in terms of oxidative stress and survival [@VandeCrommenacker2011a; @Brouwer2012], and it is therefore perhaps surprising that these variables do not explain more variance in RTL.
-Measurement error is one factor that is an issue in our study, and other studies that use qPCR to measure telomere length [@Nussey2014].
-The measurement error in our study has most likely arisen because samples were run aver a period of several years, during which reagents, consumables and personnel all change.
+It is worth considering the fact that the social and ecological variables we tested here explained only a small proportion of the variance in RTL.
+Furthermore, some factors such as territory quality and social group size were not related to telomere dynamics when we may have expected them to be [@VandeCrommenacker2011a; @Brouwer2012].
+Measurement error is clearly an issue in our study, and has almost certainly decreased the explanatory power of our models, and elevated levels of Type II error.
+The low repeatability we observed is the product of i) lower levels of efficiency in the telomere qPCR reaction than we would have liked, ii) the long-term nature of the study.
+Samples for this study were run aver a period of several years, during which reagents, consumables and personnel all change.
+Our repeatability estimate includes samples that were run several years apart, and reflect all the sources of error that accumulated over that time.
 Compared to a set-up where a small amount of samples are all run at the same time, it is unsurprising that we have a higher error rate.
-Techniques for measuring telomere length with a greater degree of precision may prove helpful in future ecological studies of telomere dynamics, but at present there is still a trade-off between obtaining precise telomere measurements, and utilising the large sample sizes necessary for ecological study.
-Further, measurement error is unlikely to be a cause of type I error in our study, because we were careful to randomise all samples across qPCR plates, and normalise RTL estimates across plates.
-However, the noise associated with within and among plate measure may have resulted in a decrease in explanatory power, and possibly in a degree of type II error.
+Techniques for measuring telomere length with a greater degree of precision are likely to prove helpful in future ecological studies of telomere dynamics [@Nussey2014], and discussions are now required on how to best optimise measuring telomere length for long-term studies.
+A central issue to resolve is how best to balance the trade-off between obtaining precise telomere measurements, and utilising the large sample sizes necessary for ecological study.
 
-Sampling error notwithstanding, we also predict that lifelong variation in RTL in the Seychelles warbler is explained by a set of environmental and genetic variables not considered here [e.g.
-@Bebbington2016].
+While sampling error is a problem in our study, we are confident that because of our study design, and plate randomisation in particular, sampling error is highly unlikely to have resulted in a high false positive rate.
+This is clearly a problem that needs to be considered, however, when designing long-term studies of telomere dynamics.
+And, clearly, sampling error is not the only factor contributing to the unexplained variance in telomere length in our study.
+It should always be borne in  mind that, in any system, unmeasured environmental and genetic variables will contribute to unexplained variance in telomere dynamics.
 A key question to be addressed is the extent to which RTL, especially in early life, reflects inheritance and parental effects [e.g.
@@
 LGS was also funded by a fellowship from the BBSRC (BB/N011759/1), HLD by a NERC fellowship (NE/I021748/1), MH was funded by a VENI fellowship from the Netherlands Organisation for Scientific Research (863.15.020), and TB by a Leverhulme fellowship.
-Eric Stier, Dan Nussey and one anonymous reviewer provided helpful comments on the manuscript.
+Antoine Stier, Dan Nussey and one anonymous reviewer provided helpful comments on the manuscript.
 

### 2017-07-11 Scripts/SW juvenile telomeres IPC clean.R

@@ -3,6 +3,7 @@
 ## CLEAN THE DATA
 #################################################################################
 
+
 #Subset bad samples
 dd0 <- subset(dd0,CqTelomere < 25) %>%
   subset(CqGAPDH < 26) %>%

### 2017-07-11 Scripts/SW juvenile telomeres IPC load.r

@@ -28,3 +28,4 @@ insects <- read.csv('Data/Insects.csv')
 dens <- read.table('Data/PsizeFP.txt')
 chickinfo <- read.csv('Data/Table chick info.csv')
 mal <- read.csv('Data/Malaria.csv')
+

### 2017-07-12 Manuscript/Response_to_reviewers_2.Rmd
Edits

@@
 output: 
-  pdf_document:
+  word_document:
 header-includes:
@@
 RESPONSE: We are glad that the reviewer agrees that our observed repeatability/efficiency values will not lead to false positives.
-Our dataset includes repeats that have run several years apart, and these repeats are included in our repeatability estimates.
-We suspect that our dataset is fairly unique in this respect at pressent, but other studies will certainly have such data available in the near future.
-Thew Bebbington et al.
-paper only included the set of most recently run telomere samples, hence the higher repeatability.
-We agree that this issue deserves more discussion space, and have now devoted an entire paragraph to the issue of repeatability, and the specific issue of qPCR efficiency, in the discussion on lines XX:
+Our dataset includes repeats that have run as much as six years apart, and these repeats are included in our repeatability estimates.
+We suspect that our dataset is fairly unique in this respect at present, but other studies will certainly have such data available in the near future.
+The Bebbington et al.
+paper only included a set of more recently run telomere samples, hence the higher repeatability.
+In this paper, the aim was to cpature variation (and repeatability) among as money cohorts and samples as possible.
+We agree that this issue deserves more discussion space, and have now devoted an entire paragraph to the issue of repeatability, and the specific issue of qPCR efficiency, in the discussion on lines 414-428:
+
+\begin{adjustwidth}{2cm}{2cm}
+
+\textit{
+It is worth considering the fact that the social and ecological variables we tested here explained only a small proportion of the variance in RTL.
+Furthermore, some factors such as territory quality and social group size were not related to telomere dynamics when we may have expected them to be (Van de Crommenacker et al.
+2011; Brouwer, Richardson, and Komdeur 2012).
+Measurement error is clearly an issue in our study, and has almost certainly decreased the explanatory power of our models, and elevated levels of Type II error.
+The low repeatability we observed is the product of i) lower levels of efficiency in the telomere qPCR reaction than we would have liked, ii) the long-term nature of the study.
+Samples for this study were run aver a period of several years, during which reagents, consumables and personnel all change.
+Our repeatability estimate includes samples that were run several years apart, and reflect all the sources of error that accumulated over that time.
+Compared to a set-up where a small amount of samples are all run at the same time, it is unsurprising that we have a higher error rate.
+Techniques for measuring telomere length with a greater degree of precision are likely to prove helpful in future ecological studies of telomere dynamics (Nussey et al.
+2014), and discussions are now required on how to best optimise measuring telomere length for long-term studies.
+A central issue to resolve is how best to balance the trade-off between obtaining precise telomere measurements, and utilising the large sample sizes necessary for ecological study.
+}
+
+\end{adjustwidth}
 
@@
 The point that we tried to make was that the (lack of) sampling resolution in early life should be offset against have broad-scale, longitudinal samples from a  large number of cohorts.
-Nonetheless, we agree that our set-up is not optimal for looking at early-life telomere shortening, and have added a section to the discussion on this, lines XX:
+Nonetheless, we agree that our set-up is not optimal for looking at early-life telomere shortening, and have added a section to the discussion on this, lines 328-337:
+
+\begin{adjustwidth}{2cm}{2cm}
+
+\textit{
+Our study adds to the substantial body of literature from humans and wild animals showing that telomere length decreases with age, and that this decrease is most rapid in early life.
+Rapid telomere shortening in early life occurs as a consequence of the much more rapid rate of cellular division that occurs during the growth phase, but also perhaps higher levels of cellular stress during development (e.g.
+Frenck, Blackburn, and Shannon 1998; Haussmann, Vleck, and Nisbet 2003; Heidinger et al.
+2012).
+Because we are rarely able to sample Seychelles warblers more than once in the nest, our system is not the best suited for looking in detail at the reasons behind telomere shortening during the growth phase.
+In contrast, we have good longitudinal and cross-sectional sampling from across individual lifespans, and using this data we have shown that, despite an overall trend for shortening, telomere length both increased and decreased, especially after the juvenile period.
+}
+
+\end{adjustwidth}
 
@@
 However, we have now added further discussion on what may cause the difference between our study and the ones mentioned above.
-We have also now discussed what low intra-individual repeatability may mean for the relationship between telomere length and long-term survival - lines XX:
+We have also now discussed what low intra-individual repeatability may mean for the relationship between telomere length and long-term survival - lines 364-380:
+
+\begin{adjustwidth}{2cm}{2cm}
+
+\textit{
+The finding that increases in telomere length may be sporadic and overlaid on an overall pattern of shortening with age is an important point when assessing the occurrence of telomere lengthening.
+Previously described approaches to distinguish telomere elongation from measurement error, based on assumptions about follow-up time between measurements (Steenstrup et al.
+2013), or based on measuring variance among measurements (Simons, Stulp, and Nakagawa 2014), assume that telomere elongation within individuals is consistent over time.
+Our data, and that of others (Fairlie et al.
+2016; Hoelzl, Cornils, et al.
+2016; Hoelzl, Smith, et al.
+2016) suggest that this is not the case.
+Such inconsistent changes in telomere length over lifespans could occur due to changes in the cellular composition of of the blood within individual samples, or due to the actual elongation of telomeres (Blackburn et al.
+1989).
+Determining the mechanism of these changes is essential for how we view telomeres as biomarkers of costs.
+For example, if telomeres can be lengthened in response to improvements in environmental conditions, this would suggest that they reflect short- to medium-term costs, rather than the cumulative costs that an individual has faced over its lifespan (Bateson 2016).
+It is clear that telomeres can be a marker of long-term as well as short-term costs, as telomere length has been associated with both survival and lifespan in wild populations (Barrett et al.
+2013; Stier et al.
+2015), but we do not yet know how biologically meaningful within-individual fluctuations in telomere length are.
+New research is therefore required to determine when and why telomere length increases within individuals, so that biologically informed hypotheses about the nature of telomeres as biomarkers in wild populations.
+}
+
+\end{adjustwidth}

### 2017-07-13 Manuscript/Response_to_reviewers_2.Rmd
Final edits before resubmission

@@
 output: 
-  word_document:
+  pdf_document:
 header-includes:
@@
 We agree with the central point made by reviewer and yourself, and have now adjusted our discussion accordingly.
-We hope you agree that this has further improved our manuscript, and we are hopful that this improved version will generate discussion about both the ecological and technical aspects of telomere dynamics in wild populations.
+We hope you agree that this has further improved our manuscript, and we are hopeful that this improved version will generate discussion about both the ecological and technical aspects of telomere dynamics in wild populations.
 
@@
 
-RESPONSE: We are very sorry about this, and we do of course know that your name is Antoine! We have now changed the ackowledgements.
+RESPONSE: We are very sorry about this, and we do of course know that your name is Antoine! We have now changed the acknowledgements.
 
@@
 However, this is a debate over semantics.
-In the manuscript, we are careful to phrase things in terms of repeatability and avoid using pearson R and R2 values as these are generally considered inappropriate for measuring repeatability.
+In the manuscript, we are careful to phrase things in terms of repeatability and avoid using Pearson R and R2 values as these are generally considered inappropriate for measuring repeatability.
 Nowhere do we state that 68% of the variance can be explained by sample identity.
@@
 paper only included a set of more recently run telomere samples, hence the higher repeatability.
-In this paper, the aim was to cpature variation (and repeatability) among as money cohorts and samples as possible.
+In this paper, the aim was to capture variation (and repeatability) among as money cohorts and samples as possible.
 We agree that this issue deserves more discussion space, and have now devoted an entire paragraph to the issue of repeatability, and the specific issue of qPCR efficiency, in the discussion on lines 414-428:
@@
 The low repeatability we observed is the product of i) lower levels of efficiency in the telomere qPCR reaction than we would have liked, ii) the long-term nature of the study.
-Samples for this study were run aver a period of several years, during which reagents, consumables and personnel all change.
+Samples for this study were run over a period of up to six years, during which reagents, consumables and personnel all change.
 Our repeatability estimate includes samples that were run several years apart, and reflect all the sources of error that accumulated over that time.
-Compared to a set-up where a small amount of samples are all run at the same time, it is unsurprising that we have a higher error rate.
-Techniques for measuring telomere length with a greater degree of precision are likely to prove helpful in future ecological studies of telomere dynamics (Nussey et al.
+Compared to a set-up where a small amount of samples are all run at the same time, it is unsurprising that we have a higher error rate - indeed, in our own studies where we have run all the samples within a short timeframe we have observed higher repeatability (Barrett et al.
+2012; Bebbington et al.
+2016).
+Techniques for measuring telomere length with a greater degree of precision are likely to prove helpful in future long-term ecological studies of telomere dynamics (Nussey et al.
 2014), and discussions are now required on how to best optimise measuring telomere length for long-term studies.
@@
 RESPONSE: We thank the reviewer for clarifying this point.
-Unfortunately we do not have any data on the presence of ITS.
+We have addressed the issue of ITS in a previous paper (Barrett et al.
+2012; J.
+Avian.
+Biol.) - essentially we are reasonably confident that they are not an issue in our dataset, because the intra-individual changes in telomere length are too great.
 However, we have now added further discussion on what may cause the difference between our study and the ones mentioned above.
@@
 2016) suggest that this is not the case.
-Such inconsistent changes in telomere length over lifespans could occur due to changes in the cellular composition of of the blood within individual samples, or due to the actual elongation of telomeres (Blackburn et al.
+Such inconsistent changes in telomere length over lifespans could occur due to changes in the cellular composition of of the blood within individual samples, due to variation in the presence of interstitial telomeric sequence, or due to the actual elongation of telomeres (Blackburn et al.
 1989).

### 2017-07-13 Manuscript/SW_early_life_telomeres_FULL.Rmd
Final edits before resubmission

@@
 Our data, and that of others [@Fairlie2016;@Hoelzl2016; @Hoelzl2016a] suggest that this is not the case.
-Such inconsistent changes in telomere length over lifespans could occur due to changes in the cellular composition of of the blood within individual samples, or due to the actual elongation of telomeres [@Blackburn1989].
+Such inconsistent changes in telomere length over lifespans could occur due to changes in the cellular composition of of the blood within individual samples, due to variation in the presence of interstitial telomeric sequence, or due to the actual elongation of telomeres [@Blackburn1989].
 Determining the mechanism of these changes is essential for how we view telomeres as biomarkers of costs.
@@
 The low repeatability we observed is the product of i) lower levels of efficiency in the telomere qPCR reaction than we would have liked, ii) the long-term nature of the study.
-Samples for this study were run aver a period of several years, during which reagents, consumables and personnel all change.
+Samples for this study were run aver a period of up to six years, during which reagents, consumables and personnel all change.
 Our repeatability estimate includes samples that were run several years apart, and reflect all the sources of error that accumulated over that time.
-Compared to a set-up where a small amount of samples are all run at the same time, it is unsurprising that we have a higher error rate.
-Techniques for measuring telomere length with a greater degree of precision are likely to prove helpful in future ecological studies of telomere dynamics [@Nussey2014], and discussions are now required on how to best optimise measuring telomere length for long-term studies.
+Compared to a set-up where a small amount of samples are all run at the same time, it is unsurprising that we have a higher error rate - indeed, in our own studies where we have run all the samples within a short timeframe we have observed higher repeatability [Barrett2012;@Bebbington2016].
+Techniques for measuring telomere length with a greater degree of precision are likely to prove helpful in future long-term ecological studies of telomere dynamics [@Nussey2014], and discussions are now required on how to best optimise measuring telomere length for long-term studies.
 A central issue to resolve is how best to balance the trade-off between obtaining precise telomere measurements, and utilising the large sample sizes necessary for ecological study.
@@
         axis.line.y = element_line(colour = 'black')) +
-  scale_colour_manual(values = c('darkgrey','gold'))+
-  scale_fill_manual(values = c('darkgrey','gold'))+
+  scale_colour_manual(values = c('darkgrey','black'))+
+  scale_fill_manual(values = c('darkgrey','black'))+
   geom_vline(xintercept = 0,lty = 2) +
@@
   stat_smooth(method = 'lm') +
-  scale_color_manual(values = c('gold2','red'))+
+  scale_color_manual(values = c('gold2','black'))+
   annotate('text',x = 17,y = 1.7,label = 'B')

### 2017-07-28 Manuscript/SW_early_life_telomeres_FULL.Rmd
Edits for publication

@@
 output:
-  pdf_document:
+  word_document:
 bibliography: MyBib.bib
@@
 ##Data accessibility
-This manuscript was written in R Markdown (http://rmarkdown.rstudio.com/). All data and scripts required to reproduce the manuscript, figures and analyses will be made available on GitHub.
+This manuscript was written in R Markdown (http://rmarkdown.rstudio.com/).
+All data and scripts required to reproduce the manuscript, figures and analyses are available on GitHub (doi: 10.5281/zenodo.835844).
 
@@
 
-**Figure 1**
-
-```{r Figure 1}
+```{r Figure 1,include = F}
 
@@
 
+pdf("Figures/Figure 1.pdf",
+    width = 10,
+    height = 8,
+    paper = "a4")
+
 multiplot(Fig1A,Fig1B,Fig1C,Fig1D,layout = matrix(1:4,2,2,byrow = T))
 
+dev.off()
 ```
@@
 
-**Figure 2**
-
-```{r Figure 2}
+```{r Figure 2,include = F}
 
@@
 
+pdf("Figures/Figure 2.pdf",
+    width = 10,
+    height = 8,
+    paper = "a4")
 multiplot(Fig2A,Fig2B,Fig2C,Fig2D,layout = matrix(c(1,2,3,4),nrow = 2,byrow = T))
+dev.off()
 
@@
 
-**Figure 3**
-
-```{r Figure 3}
+```{r Figure 3,include = F}
 
@@
 
+pdf("Figures/Figure 3.pdf",
+    width = 10,
+    height = 8,
+    paper = "a4")
 multiplot(Fig3A,Fig3B,Fig3C,layout = matrix(c(1,1,2,3),byrow = T,nrow = 2))
+dev.off()
 

### 2017-07-28 Manuscript/SW_early_life_telomeres_SOM.Rmd
Home edits
Final edits before resub
Edits for publication

@@
 ---
-output: pdf_document
+output: word_document
 classoption: landscape
@@
 
-**Table S1** Cohort sizes for each birth year of the Seychelles warbler dataset used in this study.
+**Table S1** Cohort sizes for each hatch year of the Seychelles warbler dataset used in this study. Note that cohort sizes on Cousin are typically small due to the saturated nature of the population.
 ```{r Table S1}
